### 2022-01-13 paper/paper-jsdse/paper-jsdse.tex
Add the paper and the data.

\documentclass[12pt]{article}
\usepackage{amsmath}
%\usepackage{longtable}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{tcolorbox}
\usepackage{fontawesome}
\usepackage{booktabs}
\usepackage{float}
\usepackage{url} % not crucial - just used below for the URL
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{1}

% need this for pandoc %
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
  % set entry spacing
  \ifnum #2 > 0
  \setlength{\parskip}{#2\baselineskip}
  \fi
 }%
 {}
\usepackage{calc} % for \widthof, \maxof
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\maxof{\widthof{#1}}{\csllabelwidth}}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth}{#1}}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}

% DON'T change margins - should be 1 inch all around.
\addtolength{\oddsidemargin}{-.5in}%
\addtolength{\evensidemargin}{-.5in}%
\addtolength{\textwidth}{1in}%
\addtolength{\textheight}{1.3in}%
\addtolength{\topmargin}{-.8in}%

%-------------------------------------------------------------------
% R MARKDOWN
%-------------------------------------------------------------------
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

\begin{document}

%\bibliographystyle{natbib}

\def\spacingset#1{\renewcommand{\baselinestretch}%
{#1}\small\normalsize} \spacingset{1}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\if0\blind
{
  \title{\bf The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth}
  \author{
            Dewi Amaliah\thanks{Corresponding author, \href{mailto:dlamaleeah@gmail.com}{\nolinkurl{dlamaleeah@gmail.com}}}\\
      Dept of Econometrics and Business Statistics, Monash University, \\
            Dianne Cook\\
      Dept of Econometrics and Business Statistics, Monash University, \\
            Emi Tanaka\\
      Dept of Econometrics and Business Statistics, Monash University, \\
            Kate Hyde\\
      Dept of Econometrics and Business Statistics, Monash University, \\
            Nicholas Tierney\\
      Telethon Kids Institute, Perth, WA, }
        \maketitle
} \fi

\if1\blind
{
  \bigskip
  \bigskip
  \bigskip
  \begin{center}
    {\LARGE\bf The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth}
\end{center}
  \medskip
} \fi

\bigskip
\begin{abstract}
The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data. Subsets of this data can be found in numerous textbooks and research articles. However, the steps and decisions taken to get from the raw data to the textbook data are never clearly articulated. This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly. Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research. Three new data sets and the code to produce them are provided in an accompanying open source R package, called {[}CENSORED{]}. As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks and providing more convenient samples of the data to potential users.
\end{abstract}

\noindent%
{\it Keywords:}  Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;
\vfill

\newpage
\spacingset{1.45} % DON'T change the spacing!

\hypertarget{intro}{%
\section{Introduction}\label{intro}}

``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021). This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape. Huebner, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (Huebner et al. 2020).
Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985). In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data. EDA can be considered to encompass IDA. Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.

Despite it's importance, this IDA stage is often undervalued and neglected (Chatfield 1985). There are few research papers that document the data cleaning (Wickham 2014). Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (Huebner et al. 2020).

This paper demonstrates the steps of IDA, and documents the process, for a prominent open data source, National Longitudinal Survey of Youth 1979, henceforth referred to NLSY79. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al. 2001). In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research (Pergamit et al. 2001; Cooksey 2017). According to Cooksey (2017), thousands of articles, and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al. 2001).

Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
Our aim is to refresh this textbook data with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003). We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the purposes of studying racism, not race (Fullilove 1998).

Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al. 2019). The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011). (Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

This paper is structured in the following way. Section \ref{database} describes the NLSY79 data source. Section \ref{cleaning} presents the steps of cleaning the data, including getting and tidying the data from the NLSY79 and IDA to find and repair anomalies. Our final subset is compared to the old textbook subset in Section \ref{compare}. Finally, Section \ref{summary} summarizes the contribution and makes recommendations for the NLSY79 data curators.

\hypertarget{database}{%
\section{The NLSY79}\label{database}}

\hypertarget{database-1}{%
\subsection{Database}\label{database-1}}

The NLSY79 is a longitudinal survey administered by the U.S Bureau of Labor Statistics that follows the lives of a sample of American youth born between 1957-1964 (Bureau of Labor Statistics, U.S. Department of Labor 2021). The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979. It comprised of Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military. In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members of the economically disadvantaged non-Black non-Hispanics, respectively. Hence, 9,964 respondents remain in the eligible samples. The surveys were conducted annually from 1979 to 1994 and biennially thereafter. Data are currently available from Round 1 (1979 survey year) to Round 28 (2018 survey year).

Although the main focus area of the NLSY is labor and employment, the NLSY also covers several other topics, including education; training and achievement; household, geography, and contextual variables; dating, marriage, cohabitation; sexual activity, pregnancy, and fertility; children; income, assets and program participation; health; attitudes and expectations; and crime and substance use.

There are two ways to conduct the interview of the NLSY79, which are face-to-face or by telephone interviews. In recent survey years, more than 90 percent of respondents were interviewed by telephone (Cooksey 2017).

\hypertarget{target}{%
\subsection{Target data}\label{target}}

The NLSY79 data used in Singer and Willett (2003) contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994. In addition, the cohort is restricted to male high-school dropouts who first participated in the study at age 14-17 years.
Thus the target data set is to collect the same variables for the extended time frame of 1994 through to 2018, the most recent year reported.

\hypertarget{cleaning}{%
\section{Data cleaning}\label{cleaning}}

van der Loo and de Jonge (2018) describe the notion of a ``statistical value chain'' where the production stages of the data cleaning process are earmarked as raw data (data as arrived to the desk of an analyst), input data (data organised with correct type and identified variables) and valid data (data that faithfully represent the variables). In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida}.

\hypertarget{getdata}{%
\subsection{Getting the data}\label{getdata}}

The NLSY79 data contains a large number of variables but for our aim, the scope required is limited to demographic profiles and wages data. More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data and navigated as described in Figure \ref{fig:source-nav}.

\begin{figure}[t]

\begin{tcolorbox}[title = Navigating the data source]
\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
\vspace{1mm}
\faCheck\ The CASEID will be always be selected.  \\
\vspace{1mm}
\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
\vspace{1mm}
For the remaining variables, we went to the "Variable Search" tab and select variables as follows
\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCheck] All 80 variables in Highest grade completed were selected.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked  
\begin{itemize}
\item[\faCheck] All 442 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected. 
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 686 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download". We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename "NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA. 
\end{itemize}
\end{tcolorbox}
\caption{The above documents the steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
\end{figure}

The downloaded data set comes as a zip file, containing the following set of files:

\begin{itemize}
\tightlist
\item
  \texttt{NLSY79.csv}: Comma Separated Value format of the response data,
\item
  \texttt{NLSY79.dat}: .dat format of the response data,
\item
  \texttt{NLSY79.NLSY79}: Tagset of variables that can be uploaded to the website to recreate the data set, and
\item
  \texttt{NLSY79.R}: R script for reading the data into R and converting the variables' names and label into something more sensible.
\end{itemize}

We alter only the file path in \texttt{NLSY79.R} and run the script without any other alteration. This results in an initial processing of the raw data into two data sets, \texttt{categories\_qnames} (where the observations are stored in categorical/interval values) and \texttt{new\_data\_qnames} (the observations are stored in integer form).

The raw data, \texttt{new\_data\_qnames}, is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year. The raw data consequently has a large number of columns (686 to be specific). The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars. A glimpse of this data output is shown below.

\begin{verbatim}
#> 'data.frame':    12686 obs. of  686 variables:
#>  $ CASEID_1979                  : int  1 2 3 4 5 6 7 8 9 10 ...
#>  $ HRP1_1979                    : int  328 385 365 NA 310 NA NA NA 214 NA ...
#>  $ HRP2_1979                    : int  NA NA NA NA 375 NA NA NA NA NA ...
#>  $ HRP3_1979                    : int  NA NA 275 NA NA NA NA NA NA NA ...
#>  $ HRP4_1979                    : int  NA NA NA NA NA 250 NA NA NA NA ...
#>  $ HRP5_1979                    : int  NA NA NA NA NA NA NA NA NA NA ...
#>  $ HRP1_1980                    : int  NA 457 397 NA 333 275 300 394 200 318 ...
#>  $ HRP2_1980                    : int  NA NA 367 NA NA NA NA NA NA NA ...
#>  $ HRP3_1980                    : int  NA NA 380 NA NA NA 290 NA NA NA ...
#>  $ HRP4_1980                    : int  NA NA NA NA NA NA NA NA NA NA ...
#>   [list output truncated]
\end{verbatim}

According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables. This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al. 2019), \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham, François, et al. 2020), and \texttt{stringr} (Wickham 2019) for creating new variables, and levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{CENSORED/articles/raw-to-input-data.html} but also described in the subsequent subsections.

\hypertarget{tidydemog}{%
\subsubsection{Tidying demographic variables}\label{tidydemog}}

In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}) and the corresponding year this grade was completed (\texttt{yr\_hgc}).

For gender and race, we only rename the column names. It is worth noting that using these two variable needs special attention. Gender, as reported in the data, only has two categories, which is recognised today as inadequate. Gender is not binary. Further, race is as reported in the database. When doing analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.

The \texttt{new\_data\_qnames} contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981} which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}). The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981. We checked for consistency between the two versions and found no discrepancy where the responses were recorded in both 1979 and 1981. The age was then calculated using the birth year.

The next step is tidying to obtain \texttt{hgc} and \texttt{yr\_hgc}. The highest grade completed are recorded in \texttt{new\_data\_qnames} as variables beginning with \texttt{Q3-4} and \texttt{HGC} with suffix of the year it was recorded. In addition the variables beginning with \texttt{HGCREV} contain the revised data. We choose to use this revised May data because it seemed to have less missing and presumably has been checked. However, there is no revised May data for 2012, 2014, 2016, and 2018, thus, we use the ordinary May data for these years.

The \texttt{hgc} is measured and could be updated in each period of the survey. We chose to only retain the highest grade completed for each individual and derived the year when they completed it (\texttt{yr\_hgc}) by finding the minimum year with the highest completed grade.

Finally, we get all of the demographic profiles of the NLSY79 cohort. We then save this data as \texttt{demog\_nlsy79}.

\hypertarget{tidyemp}{%
\subsubsection{Tidying employment variables}\label{tidyemp}}

Our target variables for the employment are to obtain respondent's mean hourly wage (\texttt{mean\_hourly\_wage}), the number of jobs (\texttt{number\_of\_jobs}) and the total hours of work per week (\texttt{total\_hours}) for each survey year. As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.

From 1979 to 1987, \texttt{new\_data\_qnames} only contains one version of hours worked per week for each job (in the variables with names starting with \texttt{QES-52A}). From 1988 onward, we selected the total hours worked per week, including hours working from home (\texttt{QES-52D}). However, in 1993, this variable was missing for the first and last job so we selected to use \texttt{QES-52A} instead. In addition, 2008 only had jobs 1-4 for the \texttt{QES-52D} variable, so we use only these.

The hourly wages are in the variables beginning with \texttt{HRP} in \texttt{new\_data\_qnames}. As a respondent may have multiple jobs, the \texttt{mean\_hourly\_wage} is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on number of hours is available); if number of hours worked for any job is missing, then the \texttt{mean\_hourly\_wage} is computed as a simple average of all available hourly wages. Prior to computing the mean hourly wage, we undertook a number of steps to treat unusual observations as described below:

\begin{itemize}
\tightlist
\item
  If the hourly rate is recorded as 0, we set wage as missing.
\item
  If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hour worked as missing.
\end{itemize}

The number of jobs (\texttt{number\_of\_jobs}) for each respondent per year is computed from the number of non-missing values of hourly wage \emph{and} hours worked. If either the hourly wage or hours worked is missing, we do not tally this.

\begin{verbatim}
#> # A tibble: 10 x 6
#>       id  year mean_hourly_wage total_hours number_of_jobs is_wm
#>    <int> <dbl>            <dbl>       <int>          <dbl> <lgl>
#>  1     1  1979             3.28          38              1 FALSE
#>  2     1  1981             3.61          NA              1 FALSE
#>  3     2  1979             3.85          35              1 FALSE
#>  4     2  1980             4.57          NA              1 FALSE
#>  5     2  1981             5.14          NA              1 FALSE
#>  6     2  1982             5.71          35              1 FALSE
#>  7     2  1983             5.71          NA              1 FALSE
#>  8     2  1984             5.14          NA              1 FALSE
#>  9     2  1985             7.71          NA              1 FALSE
#> 10     2  1986             7.69          NA              1 FALSE
\end{verbatim}

The employment and demographic variables are then joined. These data are further filtered to the cohort who completed education up to 12th grade and participated in at least five rounds in the survey. We save the resultant wage data on this cohort as \texttt{wages}.

\hypertarget{ida}{%
\subsection{Initial data analysis}\label{ida}}

According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis. Moreover, Chatfield (1985) argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.

In this paper, we conduct an IDA or a preliminary data analysis to assess the consistency of the data with the cohort information that the NLSY provides. In addition, we also aim to find the anomaly in the wages values using this approach. We mainly use graphical summaries to do the IDA using \texttt{ggplot2} (Wickham 2016) and \texttt{brolgar} (Tierney, Cook, and Prvan 2020).

As stated previously, the respondents' ages ranged from 12 to 22 when first interviewed in 1979. Hence, we validate whether all of the respondents were in this range. Additionally, the \href{https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction}{NLSY} also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic). To validate this, we used the \texttt{demog\_nlsy79}, i.e., the data with the survey years 1979 sample.
Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest that the demographic data we had is consistent with the sample information in the database.

\begin{table}

\caption{\label{tab:age-table}The frequency table of the age at the start of the survey (?CHECK) in the full NSLY79 data}
\centering
\begin{tabular}[t]{rr}
\toprule
Age & Number of individuals\\
\midrule
15 & 1,265\\
16 & 1,550\\
17 & 1,600\\
18 & 1,530\\
19 & 1,662\\
20 & 1,722\\
21 & 1,677\\
22 & 1,680\\
\bottomrule
\end{tabular}
\end{table}

\begin{table}

\caption{\label{tab:gender-race-table}The contingency table for gender and race for the full NLSY79 data.}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Race} & \multicolumn{1}{c}{ } \\
\cmidrule(l{3pt}r{3pt}){2-4}
Gender & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
\midrule
Male & 1,000 (15.62\%) & 1,613 (25.19\%) & 3,790 (59.19\%) & 6,403 (100.00\%)\\
Female & 1,002 (15.95\%) & 1,561 (24.84\%) & 3,720 (59.21\%) & 6,283 (100.00\%)\\
\midrule
Total & 2,002 (15.78\%) & 3,174 (25.02\%) & 7,510 (59.20\%) & 12,686 (100.00\%)\\
\bottomrule
\end{tabular}
\end{table}

In the next step, we explore the mean hourly wage data. In this case, we only explore the wages data of respondents that have the highest completed grade of up to \(12^{th}\) grade. We employ visualization techniques to perform the IDA as described next.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figures/sample-plot-1}

}

\caption{Longitudinal profiles of wages for a random sample of 20 individuals in the refreshed data. Most, but not all, individuals here experienced increasing wages over time, and several have experienced considerable fluctuation in wages. Some individuals are only measured for a short period.}\label{fig:sample-plot}
\end{figure}

We randomly take 20 samples from the data and plot them, as shown in Figure \ref{fig:sample-plot}. It shows that these respondents have a lot of variability in wages, for example, the IDs in panel numbers 5, 7, and 11. The plot also implies that the samples have a different pattern of mean hourly wages. Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.
However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages. Some of them, for example respondents in Figure \ref {fig:feature-plot} C had experienced unusual wages only in certain year.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figures/feature-plot-1}

}

\caption{Summary plots to check the cleaned data reveal more cleaning is necessary: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual (id=39) with an unusual wage relative to their years of data (C). Some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.}\label{fig:feature-plot}
\end{figure}

The anomalies are also found in the total hours of work, where some observations reported as having worked for 420 hours a week in total.
According to Pergamit et al. (2001), one of the flaws of the NLSY79 employment data is that the NLSY79 collects the information of the working hours since the last interview. Thus, it might be challenging for the respondents to track the within-job hours' changes between survey years, especially for the respondents with fluctuating working hours or seasonal jobs. It even has been more challenging since 1994, where the respondents had to recall two year periods. This shortcoming might also contribute to the fluctuation of one's wages data.

\hypertarget{censor}{%
\subsubsection{Replacing extreme values}\label{censor}}

As part of the IDA, which is the model formulation, we build a robust linear regression model to treat the extreme values in the data. Robust linear regression yields an estimation robust to the influence of noise or contamination (Koller 2016). It also aims to detect the contamination by weighting each observation based on how ``well-behaved'' they are, known as robustness weight. Observations with lower robustness weight are suggested as an outlier by this method (Koller 2016).

Since we work with longitudinal data, we build the model for each ID instead of the overall data. The robust mixed model could be the best model to be employed in this case. However, this method is too computationally and memory expensive, especially for a large data set, like the NLSY79 data. Therefore, the model for each ID is built utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively. The full code for this is shown at \url{CENSORED/articles/input-to-valid-data.html} but also described in detail next.

We build the model using the \texttt{rlm} function from \texttt{MASS} package (Venables and Ripley 2002). We set the \texttt{mean\_hourly\_wage} and \texttt{year} as the dependent and predictor, respectively. Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021). However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers. Moreover, it should be noted that not all the outliers are due to an error. Instead, it might be that one had reasonably increasing or decreasing wages in a particular period.

To minimize the risk of mistakenly identifying an outlier as an ``erroneous outlier'', we simulate some thresholds and study how they affect the data. We find that 0.12 is the most reasonable value to be the threshold to minimize that drawback's risk because it still captures the sensible spikes in the data. In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording. After deciding the threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value. We then flag those observations in a new variable called \texttt{is\_pred}.

Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced. It implies that the fluctuation can still be observed in the data after the treatment. However, the large spikes, which are considered ``erroneous outliers'', are already eliminated from the data. Hence, the model produces a data set with a more reasonable degree of fluctuation.

\begin{figure}

{\centering \includegraphics[width=0.9\linewidth]{figures/compare-plot-1}

}

\caption{Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for a sample of individuals. A robust linear model prediction was used to correct mean hourly wages value. We can see that some extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed. Some spikes might remain when wage vaues are plausible.}\label{fig:compare-plot}
\end{figure}

Further, Figure \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around \$350. The spikes are still observed but not as extreme as the original data set. In Figure \ref{fig:fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value.
We still can see some extreme values in maximum wages, but consider it as a natural variability of the data. In Figure \ref{fig:fixed-feature-plot} C, we can see that after imputing the extreme value in ID=39, we can see how the the wages change over the years more clearly.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figures/fixed-feature-plot-1}

}

\caption{Re-make of the summary plots of the fully processed data suggest that it is now in a reasonable state: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual with an unusual wage relative to their years of data (C). }\label{fig:fixed-feature-plot}
\end{figure}

Finally, we save the imputed data and set the appropriate data type for the variables.

\hypertarget{recap}{%
\subsection{Recap}\label{recap}}

Figure \ref{fig:flow-chart-blind} summarizes the steps taken to go from raw to input to valid data (van der Loo and de Jonge 2018) to create a refreshed wages data set.

\begin{figure}

{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie/paper/figures/flowchart_blind}

}

\caption{The stages of data cleaning from the raw data to get three datasets contained in \texttt{[CENSORED]}. ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data. The color represents the stage of data cleaning in statistical value chain (van der Loo and de Jonge 2021). Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart-blind}
\end{figure}

\hypertarget{compare}{%
\section{Comparison of refreshed with the original data}\label{compare}}

Now is the time to see how close we have come to the original textbook data. The original set contains wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994. The refreshed data set for comparison is also wages on high school drop outs from 1979 to 2018. The original set is available in the R package \texttt{brolgar} and the refreshed data is available in the R package \texttt{[CENSORED]}.

\begin{figure}

{\centering \includegraphics[width=1\linewidth]{figures/plotting-sw-do-1}

}

\caption{Comparison of original textbook example (A) with refreshed data (B). The original data was inflation-adjusted to 1990 prices and the individual's time of collection was converted to a length of experience in the workforce, which makes it difficult to precisely compare the two sets.}\label{fig:plotting-sw-do}
\end{figure}

There are two aspects of the original data that make a direct comparison difficult. The time variable provided is ``experience in the workforce''. It is not clear how this is calculated. One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection. We have not been able to find this information in the database. Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.

The treatment of unlikely wages differ in the refreshed data. In the original data by Singer and Willett (2003), wages greater than \$75 are set to be missing. However, in recent context, this value is too low to be set as the maximum threshold. We opted to use the weights from a robust linear regression to determine what should be set as missing value as described in Section \ref{censor}.

Figure \ref{fig:plotting-sw-do} shows a comparison of the two sets. (A direct ID matching is not possible.) There are 888 individuals in the original and 1,188 individuals in the refreshed data. This suggests some individuals were removed in the original data.
In addition, in the original set the racial breakdown was 246 Black, 204 Hispanic and 438 White participants, while in the refreshed data there are 346 Black, 219 Hispanic, and 623 White participants, so the proportions have not been exactly reproduced. On education, in the refreshed data there are very few individuals with less than a 12th grade education, which is different from the original. In the original data there were 366 individuals with up to 8th grade and 522 9th-12th grade, as compared to 967 with 12th grade in the refreshed data.

\hypertarget{summary}{%
\section{Summary}\label{summary}}

This paper has described the stages to take a particular open data set and make it a textbook data set, ready for the classroom or research. In the first stage, we showed the steps performed to get the data from the NLSY79 database. The data format needed conversion to tidy format, and this was described. After that, an initial data analysis was conducted to investigate and screen the quality of the data. We found and fixed the anomalous observations in wages using a robust linear regression model. Finally a comparison is made between the original and refreshed data sets.

The data cleaning process is documented and the code has been made available. These provide the opportunity to again refresh the textbook data as updated data is published in the NLSY79 database.
Determining the appropriate robustness weight to threshold the anomalous observations is documented, and a \texttt{shiny} (Chang et al. 2020) app is provided to assist.
The current subset is made available in a new R package, called \texttt{[CENSORED]}.

Various difficulties were encountered in trying to refresh the data, which include:

\begin{itemize}
\tightlist
\item
  determining which records should be downloaded from the database.
\item
  there are many errors in the data, e.g.~hourly wages greater than \$30,000 per hour.
\item
  there is no explicit variable in the database recording high school dropout, which means we needed to compare date of 12th grade with the individual's age.
\item
  calculating experience in the workforce would require knowing the time of first job within the first year the individual was recorded.
\end{itemize}

Ultimately, the refreshed data is reasonably similar to the original, but unsatisfactorily far from it. The last step required would be to inflation-adjust wages, but this is better to do with each wave of new data added, so that it is relative to the last date in the data.

Some readers may disagree with our decisions made to produce the refreshed textbook data and may have better insight than us in producing a more appropriate textbook data. We do not assert that we have produced the best textbook data, but rather we describe our journey to provide a reasonable textbook data set. All code and documentation are provided for transparency, and future updates of the \texttt{[CENSORED]} package may contain additional variables, or filters of the full set, if it is deemed important.

Finally, for the data providers we recommend that a validation system with clear rules is added on data entry, and that alternative output formats, such as a tidy format would help users make better use of the resource. The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs. These values can be validated with simple checks at data entry. Providing an open data resource also is accompanied with the responsibility that the data, especially data that is as valuable as this, is reliable. Users need to be able to trust the data.

\hypertarget{acknowledgements}{%
\section{Acknowledgements}\label{acknowledgements}}

We would like to thank Aarathy Babu for the insight and discussion during the writing of this paper.

The entire analysis is conducted using \texttt{R} (R Core Team 2020) in RStudio IDE using these packages: \texttt{tidyverse} (Wickham, Averick, et al. 2019), \texttt{ggplot2} (Wickham 2016), \texttt{dplyr} (Wickham, François, et al. 2020), \texttt{readr} (Wickham and Hester 2020), \texttt{tidyr} (Wickham 2020), \texttt{stringr} (Wickham 2019), \texttt{purrr} (Henry and Wickham 2020), \texttt{brolgar} (Tierney, Cook, and Prvan 2020), \texttt{patchwork} (Pedersen 2020), \texttt{kableExtra} (Zhu 2019), \texttt{MASS} (Venables and Ripley 2002), \texttt{janitor} (Firke 2020), and \texttt{tsibble} (Wang, Cook, and Hyndman 2020). The paper was generated using \texttt{knitr} (Xie 2014) and \texttt{rmarkdown} (Xie, Dervieux, and Riederer 2020).

\hypertarget{supplementary-materials}{%
\section{Supplementary Materials}\label{supplementary-materials}}

\begin{itemize}
\item
  \textbf{Codes}: R script to reproduce data tidying and cleaning is available in this \href{CENSORED/articles/process-data.html}{page}.
\item
  \textbf{R Package}:\texttt{[CENSORED]} is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort. This package can be accessed \href{CENSORED}{here}.
\item
  \textbf{shiny app}: An interactive \texttt{shiny} web app to visualise the effect of selecting different weight threshold for substituting the wages data to its predicted value from a fit of the robust linear regression model. This app can be accessed \href{CENSORED}{here} with the source code provided \href{CENSORED/tree/master/app}{here}.
\end{itemize}

\hypertarget{data-availability-statement}{%
\section{Data Availability Statement}\label{data-availability-statement}}

The authors confirm that the data supporting the findings of this study are available within the supplementary materials.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-nlsy79}{}%
Bureau of Labor Statistics, U.S. Department of Labor. 2021. ``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University. Columbus, OH.

\leavevmode\hypertarget{ref-shiny}{}%
Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.

\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.

\leavevmode\hypertarget{ref-eliznlsy}{}%
Cooksey, Elizabeth C. 2017. ``Using the National Longitudinal Surveys of Youth (Nlsy) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M. Lerner Neal Halfon Christoper B. Forrest, 561--77. Cham: Springer. \url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.

\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.

\leavevmode\hypertarget{ref-janitor}{}%
Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.

\leavevmode\hypertarget{ref-racismnotrace}{}%
Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.

\leavevmode\hypertarget{ref-grimshaw}{}%
Grimshaw, Scott D. 2015. ``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14. \url{https://doi.org/10.1080/00031305.2015.1081106}.

\leavevmode\hypertarget{ref-purrr}{}%
Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.

\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
Huebner, Marianne, Werner Vach, and Saskia le Cessie. 2016. ``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.

\leavevmode\hypertarget{ref-HuebnerMarianne2020Haar}{}%
Huebner, Marianne, Werner Vach, Saskia le Cessie, Carsten Oliver Schmidt, and Lara Lusa. 2020. ``Hidden Analyses: A Review of Reporting Practice and Recommendations for More Transparent Reporting of Initial Data Analyses.'' \emph{BMC Medical Research Methodology} 20 (1): 61--61.

\leavevmode\hypertarget{ref-tamedata}{}%
Kim, A. Y, C. Ismay, and J. Chunn. 2018. ``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1). \url{https://doi.org/10.5070/T511103589}.

\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.

\leavevmode\hypertarget{ref-opendata}{}%
Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.

\leavevmode\hypertarget{ref-patchwork}{}%
Pedersen, Thomas Lin. 2020. \emph{patchwork: The Composer of Plots}. \url{https://CRAN.R-project.org/package=patchwork}.

\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
Pergamit, Michael R., Charles R. Pierret, Donna S. Rothstein, and Jonathan R. Veum. 2001. ``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.

\leavevmode\hypertarget{ref-R}{}%
R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.

\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.

\leavevmode\hypertarget{ref-brolgar}{}%
Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.

\leavevmode\hypertarget{ref-tukey}{}%
Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.

\leavevmode\hypertarget{ref-rlm}{}%
UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.

\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.

\leavevmode\hypertarget{ref-validate}{}%
van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.

\leavevmode\hypertarget{ref-mass}{}%
Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.

\leavevmode\hypertarget{ref-tsibble}{}%
Wang, Earo, Dianne Cook, and Rob J Hyndman. 2020. ``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78. \url{https://doi.org/10.1080/10618600.2019.1695624}.

\leavevmode\hypertarget{ref-plyr}{}%
Wickham, Hadley. 2011. ``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.

\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
---------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.

\leavevmode\hypertarget{ref-ggplot2}{}%
---------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.

\leavevmode\hypertarget{ref-stringr}{}%
---------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.

\leavevmode\hypertarget{ref-tidyr}{}%
---------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.

\leavevmode\hypertarget{ref-tidyverse}{}%
Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al. 2019. ``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686. \url{https://doi.org/10.21105/joss.01686}.

\leavevmode\hypertarget{ref-dplyr}{}%
Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.

\leavevmode\hypertarget{ref-readr}{}%
Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.

\leavevmode\hypertarget{ref-knitr}{}%
Xie, Yihui. 2014. ``Knitr: A Comprehensive Tool for Reproducible Research in R.'' In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D. Peng. Chapman; Hall/CRC. \url{http://www.crcpress.com/product/isbn/9781466561595}.

\leavevmode\hypertarget{ref-rmarkdown}{}%
Xie, Yihui, Christophe Dervieux, and Emily Riederer. 2020. \emph{R Markdown Cookbook}. Boca Raton, Florida: Chapman; Hall/CRC. \url{https://bookdown.org/yihui/rmarkdown-cookbook}.

\leavevmode\hypertarget{ref-kableExtra}{}%
Zhu, Hao. 2019. \emph{kableExtra: Construct Complex Table with 'kable' and Pipe Syntax}. \url{https://CRAN.R-project.org/package=kableExtra}.

\bibliographystyle{Chicago}

\end{document}

### 2022-01-13 paper/paper.tex
Add the paper and the data.

--- 2022-01-13 paper/paper-jsdse/paper-jsdse.tex
@@
-\documentclass[12pt]{article}
-\usepackage{amsmath}
-%\usepackage{longtable}
-\usepackage{graphicx,psfrag,epsf}
-\usepackage{enumerate}
-\usepackage{natbib}
+\documentclass{article}
+
+\usepackage{arxiv}
+
+\usepackage[utf8]{inputenc} % allow utf-8 input
+\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
+\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
+\usepackage{hyperref}       % hyperlinks
+\usepackage{url}            % simple URL typesetting
+\usepackage{booktabs}       % professional-quality tables
+\usepackage{amsfonts}       % blackboard math symbols
+\usepackage{nicefrac}       % compact symbols for 1/2, etc.
+\usepackage{microtype}      % microtypography
+\usepackage{lipsum}
+\usepackage{graphicx}
+
+\title{The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth}
+
+\author{
+    Dewi Amaliah
+   \\
+    Department of Econometrics and Business Statistics \\
+    Monash University \\
+  Clayton, VIC 3800 \\
+  \texttt{\href{mailto:dama0007@student.monash.edu}{\nolinkurl{dama0007@student.monash.edu}}} \\
+   \And
+    Dianne Cook
+   \\
+    Department of Econometrics and Business Statistics \\
+    Monash University \\
+  Clayton, VIC 3800 \\
+  \texttt{\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}} \\
+   \And
+    Emi Tanaka
+   \\
+    Department of Econometrics and Business Statistics \\
+    Monash University \\
+  Clayton, VIC 3800 \\
+  \texttt{\href{mailto:emi.tanaka@monash.edu}{\nolinkurl{emi.tanaka@monash.edu}}} \\
+   \And
+    Kate Hyde
+   \\
+    Department of Econometrics and Business Statistics \\
+    Monash University \\
+  Clayton, VIC 3800 \\
+  \texttt{} \\
+   \And
+    Nicholas Tierney
+   \\
+     \\
+   \\
+  \texttt{} \\
+  }
+
+% Pandoc citation processing
+
 \usepackage{tcolorbox}
 \usepackage{fontawesome}
+\usepackage{color, colortbl}
 \usepackage{booktabs}
+\usepackage{longtable}
+\usepackage{array}
+\usepackage{multirow}
+\usepackage{wrapfig}
 \usepackage{float}
-\usepackage{url} % not crucial - just used below for the URL
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-%\pdfminorversion=4
-% NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{1}
-
-% need this for pandoc %
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
-  % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
- }%
- {}
-\usepackage{calc} % for \widthof, \maxof
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\maxof{\widthof{#1}}{\csllabelwidth}}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth}{#1}}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-
-% DON'T change margins - should be 1 inch all around.
-\addtolength{\oddsidemargin}{-.5in}%
-\addtolength{\evensidemargin}{-.5in}%
-\addtolength{\textwidth}{1in}%
-\addtolength{\textheight}{1.3in}%
-\addtolength{\topmargin}{-.8in}%
-
-%-------------------------------------------------------------------
-% R MARKDOWN
-%-------------------------------------------------------------------
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-{#1}\small\normalsize} \spacingset{1}
+\usepackage{colortbl}
+\usepackage{pdflscape}
+\usepackage{tabu}
+\usepackage{threeparttable}
+\usepackage{threeparttablex}
+\usepackage[normalem]{ulem}
+\usepackage{makecell}
+\usepackage{xcolor}
 
 \begin{document}
+\maketitle
 
-%\bibliographystyle{natbib}
+\def\tightlist{}
 
-\def\spacingset#1{\renewcommand{\baselinestretch}%
-{#1}\small\normalsize} \spacingset{1}
-
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-\if0\blind
-{
-  \title{\bf The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth}
-  \author{
-            Dewi Amaliah\thanks{Corresponding author, \href{mailto:dlamaleeah@gmail.com}{\nolinkurl{dlamaleeah@gmail.com}}}\\
-      Dept of Econometrics and Business Statistics, Monash University, \\
-            Dianne Cook\\
-      Dept of Econometrics and Business Statistics, Monash University, \\
-            Emi Tanaka\\
-      Dept of Econometrics and Business Statistics, Monash University, \\
-            Kate Hyde\\
-      Dept of Econometrics and Business Statistics, Monash University, \\
-            Nicholas Tierney\\
-      Telethon Kids Institute, Perth, WA, }
-        \maketitle
-} \fi
-
-\if1\blind
-{
-  \bigskip
-  \bigskip
-  \bigskip
-  \begin{center}
-    {\LARGE\bf The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth}
-\end{center}
-  \medskip
-} \fi
-
-\bigskip
 \begin{abstract}
@@
 
-\noindent%
-{\it Keywords:}  Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;
-\vfill
-
-\newpage
-\spacingset{1.45} % DON'T change the spacing!
+\keywords{
+    Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;
+  }
 
@@
 
-\bibliographystyle{Chicago}
+\bibliographystyle{unsrt}
+\bibliography{references.bib}
 

### 2022-01-13 data-raw/data-preprocessing.R

@@ -1,0 +1,234 @@
+
+
+## ---- load-pkgs
+library(tidyverse)
+select <- dplyr::select
+
+## ---- raw-data
+source(here::here("data-raw/NLSY79/NLSY79.R"))
+
+## ---- untidy-data
+options(width=70)
+new_data_qnames %>%
+  select(CASEID_1979,
+         starts_with("HRP") &
+           ends_with(c("1979", "1980")),
+         everything()) %>%
+  str(list.len = 10)
+
+
+## ---- dob-tidy
+dob_tidy <- new_data_qnames %>%
+  select(id = CASEID_1979,
+         year_1979 = `Q1-3_A~Y_1979`,
+         year_1981 = `Q1-3_A~Y_1981`,
+         month_1979 = `Q1-3_A~M_1979`,
+         month_1981 = `Q1-3_A~M_1981`) %>%
+  mutate(dob_year = case_when(
+                      # if the years recorded in both sets match, take 79 data
+                      year_1979 == year_1981 ~ year_1979,
+                      # if the year in the 81 set is missing, take the 79 data
+                      is.na(year_1981) ~ year_1979,
+                      # if the sets don't match for dob year, take the 79 data
+                      year_1979 != year_1981 ~ year_1979),
+         dob_month = case_when(
+                      # if months recorded in both sets match, take 79 data
+                      month_1979 == month_1981 ~ month_1979,
+                      # if month in 81 set is missing, take the 79 data
+                      is.na(month_1981) ~ month_1979,
+                      # if sets don't match for dob month, take the 79 data
+                      month_1979 != month_1981 ~ month_1979),
+         # flag if there is a conflict between dob recorded in 79 and 81
+         dob_conflict = case_when(
+                      !is.na(month_1981) && (month_1979 != month_1981) ~ TRUE,
+                      !is.na(year_1981) && (year_1979 != year_1981) ~ TRUE,
+                      (year_1979 == year_1981) & (month_1979 == month_1981) ~ FALSE,
+                      is.na(month_1981) | is.na(year_1981) ~ FALSE)) %>%
+  select(id,
+         dob_month,
+         dob_year,
+         dob_conflict)
+
[…]

### 2022-01-13 data-raw/data-valid.R

@@ -1,0 +1,49 @@
+
+## ---- load-pkgs-valid
+library(tidyverse)
+library(MASS)
+select <- dplyr::select
+wages_before <- readRDS(here::here("paper/results/wages_before.rds"))
+
+## ---- rlm
+
+# nest the data by id to build a robust linear model
+by_id <- wages_before %>%
+  select(id, year, mean_hourly_wage) %>%
+  group_by(id) %>%
+  nest()
+
+# build a robust linear model
+id_rlm <- by_id %>%
+  mutate(model = map(data, ~rlm(mean_hourly_wage ~ year, data = .x)))
+
+# extract the properties of regression model and weight of each observation
+
+id_aug_w <- id_rlm %>%
+  mutate(.fitted = map(model, fitted),
+         .resid = map(model, resid),
+         .hat = map(model, hatvalues),
+         .sigma = map(model, sigma),
+         w = map(model, ~.x$w)) %>%
+  select(-model) %>%
+  unnest(data:w)
+
+# if the weight < 0.12, the mean_hourly_wage is replaced by the model's fitted/predicted value.
+# and add the flag whether the observation is predicted value or not.
+# since the fitted value is sometimes <0, and wages value could never be negative,
+# we keep the mean hourly wage value even its weight < 0.12.
+
+wages_rlm_dat <- id_aug_w %>%
+  mutate(wages_rlm = ifelse(w < 0.12  & .fitted >= 0, .fitted,
+                            mean_hourly_wage)) %>%
+  mutate(is_pred = ifelse(w < 0.12 & .fitted >= 0, TRUE,
+                          FALSE)) %>%
+  select(id, year, wages_rlm, is_pred)
+
+# join back the `wages_rlm_dat` to `wages_demog_hs`
+
+wages_after <- left_join(wages_before, wages_rlm_dat, by = c("id", "year"))
+
+## ---- save-wages-after
+# save it to rds file so it would faster the knitting process
+saveRDS(wages_after, here::here("paper/results/wages_after.rds"))

### 2022-01-13 data-raw/eda.R

@@ -1,0 +1,198 @@
+## ---- sw_wages
+
+sw <- brolgar::wages
+
+sw_wages <- sw %>%
+  ggplot(aes(x = xp,
+             y = ln_wages)) +
+  geom_line(aes(group = id), alpha = 0.1) +
+  geom_smooth(se = FALSE, colour = "pink") +
+  labs(tag = "(A)", title = "Original") +
+  theme_bw() +
+  ylab("Hourly wage ($, natural log)") +
+  xlab("Experience in years") +
+  #theme(plot.title = element_text(size = 10)) +
+  ylim(0, 5)
+
+sw_wages_mod <- sw %>%
+  as_tibble() %>%
+  mutate(hgc = ifelse(high_grade < 9, "8TH", "12TH")) %>%
+  mutate(race = case_when(black == 1 ~ "black",
+                          hispanic == 1 ~ "hispanic",
+                          TRUE ~ "white")) %>%
+  mutate(race = factor(race)) %>%
+  ggplot(aes(x = xp,
+             y = ln_wages,
+             linetype = hgc)) +
+  geom_smooth(method = "lm", se = FALSE,
+              colour = "black") +
+  labs(tag = "(A)", title = "Original") +
+  scale_linetype("") +
+  theme_bw() +
+  ylab("Hourly wage ($, natural log)") +
+  xlab("Experience in years") +
+  theme(legend.position = "bottom",
+        legend.direction = "horizontal")
+
+## ---- do_refreshed
+
+# singer-willet did not stated how they calculate experience
+# in database, there is work experience topic
+# (https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience)
+# but there is no variable explicitly mention about experience.
+# in this case, we approximate it with year-1979
+
+# filter do data up to 1994 since the textbook data only covers that period.
+
+do <- yowie::wages_hs_do %>%
+  mutate(lnwage = log(wage)) #%>%
+  #filter(year < 1995)
+
+do_ref <- do %>%
+  ggplot(aes(x = year,
+             y = lnwage)) +
+  geom_line(aes(group = id), alpha = 0.1) +
+  geom_smooth(se = FALSE, colour = "pink") +
+  labs(tag = "(B)", title = "Refreshed data") +
+  theme_bw() +
+  ylab("Hourly wage ($, natural log)") +
+  xlab("Year of data collection") +
+  theme(plot.title = element_text(size = 10)) +
+  ylim(0, 5)
+
+do_ref_mod <- do %>%
[…]

### 2022-01-13 data-raw/ida.R

@@ -1,0 +1,362 @@
+## ---- load-pkgs2
+library(kableExtra)
+library(janitor)
+library(tsibble)
+library(brolgar)
+library(patchwork)
+
+## ---- age-table
+age_table <- full_demographics %>%
+  group_by(age_1979) %>%
+  count(age_1979)
+
+age_table %>%
+  mutate(n = scales::comma(n)) %>%
+  kable(
+      caption = "The frequency table of the age at the start of the survey (?CHECK) in the full NSLY79 data",
+      col.names = c("Age", "Number of individuals"),
+      booktabs = TRUE,
+      linesep = "",
+      align = "r") %>%
+  kable_styling()
+
+## ---- gender-race-table
+#' @param x a tabyl with frequency table in "core" attribute
+freq_formatting <- function(x) {
+  attr(x, "core") %>%
+    adorn_totals(c("row", "col")) %>%
+    mutate(across(where(is.numeric), ~scales::comma(.x, 1)))
+}
+
+gender_race_table <- full_demographics %>%
+  tabyl(gender, race) %>%
+  adorn_totals(c("row", "col")) %>%
+  adorn_percentages("row") %>%
+  adorn_pct_formatting(digits = 2) %>%
+  adorn_ns(., position = "front", ns = freq_formatting(.)) %>%
+  mutate(gender = str_to_title(gender))
+
+kable(gender_race_table,
+      caption = "The contingency table for gender and race for the full NLSY79 data.",
+      col.names = c("Gender", "Hispanic", "Black", "Non-Black, Non-Hispanic", "Total"),
+      booktabs = TRUE,
+      linesep = "",
+      align = "lrrrr") %>%
+  kable_classic() %>%
+  row_spec(2, hline_after = TRUE) %>%
+  add_header_above(c(" " = 1, "Race" = 3, " " = 1))
+
+## ---- summarytable
+kable(as.array(summary(wages_before$mean_hourly_wage)),
+      caption = "Summary statistics of the hourly wages from the high school dropout subcohort of NLSY79 data",
+      col.names = c("Statistics", "Hourly wage ($)"),
+      booktabs = TRUE,
+      linesep = "") %>%
+  kable_styling()
+
+## ---- sample-plot
+wages_before_tsibble <- as_tsibble(x = wages_before,
+                                     key = id,
+                                     index = year,
[…]

### 2022-01-13 paper/paper.Rmd
outline of changes to intro

---
title: "The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth"
authors:
  - name: Dewi Amaliah
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dama0007@student.monash.edu
  - name: Dianne Cook
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dicook@monash.edu
  - name: Emi Tanaka
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Kate Hyde
    thanks: ""
    department: Department of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: ""
  - name: Nicholas Tierney
    thanks: ""
    department: ""
    affiliation: ""
    location: ""
    email: ""
bibliography: references.bib
biblio-style: unsrt
#preamble: >
header-includes:
  - \usepackage{tcolorbox}
  - \usepackage{fontawesome}
  - \usepackage{color, colortbl}
output: 
  #rticles::arxiv_article:
    #keep_tex: false
  bookdown::pdf_book:
    base_format: rticles::arxiv_article
date: "24/02/2021"
always_allow_html: true
keywords: "Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;"
abstract: > 
 The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data. Subsets of this data can be found in numerous textbooks and research articles. However, the steps and decisions taken to get from the raw data to the textbook data are never clearly articulated. This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly. Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research. Three new data sets and the code to produce them are provided in an accompanying open source R package, called [CENSORED]. As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks and providing more convenient samples of the data to potential users. 
---

```{r censor, include = FALSE}
CENSOR <- TRUE
yowie <- ifelse(CENSOR, "\\texttt{[CENSORED]}", "\\texttt{yowie}")
yowie_repo <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie")
yowie_pkgdown <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie")
yowie_shiny <- ifelse(CENSOR, "CENSORED", "https://ebsmonash.shinyapps.io/yowie_app/")
```

```{r setup, echo = FALSE, cache = FALSE, include = FALSE}
library(knitr)
opts_chunk$set(echo = FALSE, 
               warning = FALSE, 
               message = FALSE,
               cache = TRUE, 
               cache.path = "cache/",
               fig.retina = 2,
               fig.path = "figures/",
               comment = "#>",
               fig.align = 'center')
read_chunk(here::here("data-raw/data-preprocessing.R"))
read_chunk(here::here("data-raw/ida.R"))
read_chunk(here::here("data-raw/eda.R"))
```

```{r ref.label=c("load-pkgs", "load-pkgs2"), cache = FALSE}
```

```{r theme, eval = FALSE}
theme_set(theme_bw(base_size = 18) +
            theme(plot.background = element_rect(fill = 'transparent', colour = NA), axis.line.y = element_line(color = "black", linetype = "solid"),
                  plot.title.position = "plot",
                  plot.title = element_text(size = 24),
                  panel.background  = element_rect(fill = 'transparent', colour = NA),
                  legend.background = element_rect(fill = 'transparent', colour = NA),
                  legend.key        = element_rect(fill = 'transparent', colour = NA)
                  ) )
```

# Introduction {#intro}

<!--
- how does this particular data relate to education
- textbook data, why is this important for education
- being able to refresh textbook data is important
- what can we learn about the data that is different what is usually taught
-->

"Open data" is data that is freely accessible, modifiable, and shareable by anyone for any purpose [@opendata]. This type of data can be useful as example data in statistical textbooks and for research purposes. However, open data is often referred to as what we might call "wild data" because it requires substantial cleaning and tidying to tame it into textbook shape. @HuebnerMariannePhD2016Asat emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building <!-- AE comment: It's unclear from context what is being referred to from the Huebner et al 2020 citation -->  [@HuebnerMarianne2020Haar]. 
Data cleaning can be considered to be a part of what is called "initial data analysis" (IDA) [@Chatfield1985TIEo]. In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling. This is also related to exploratory data analysis (EDA), coined by @tukey with a focus on learning from data. EDA can be considered to encompass IDA. @DasuTamraparni2003Edma say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80% of the data mining task.

Despite it's importance, this IDA stage is often undervalued and neglected [@Chatfield1985TIEo]. There are few research papers that document the data cleaning [@WickhamHadley2014TD]. Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties [@HuebnerMarianne2020Haar].

This paper demonstrates the steps of IDA, and documents the process, for a prominent open data source, National Longitudinal Survey of Youth 1979, henceforth referred to NLSY79. This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN]. In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research [@MichaelRPergamit2001DWTN; @eliznlsy]. According to @eliznlsy, thousands of articles, and hundreds of book chapters and monographs have utilized this data. Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data [@MichaelRPergamit2001DWTN].

@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race. <!--AE comment: For example, I would like more discussion about the textbook selected as the exemplar. It sounds like this dataset is used in many books, why choose that one? The text should include the name of the book, so readers do not need to flip to the citations in order to learn what it was. From the name, I am guessing that the book focuses heavily on this dataset, which is why it makes sense to reproduce their subset. Be explicit about this.-->
Our aim is to refresh this textbook data with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences. Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda. We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the purposes of studying racism, not race [@racismnotrace].  <!--However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.-->

Our process of cleaning builds heavily on the `tidyverse` approach [@tidyverse]. The data is first organised into "tidy data" [@WickhamHadley2014TD] and then further wrangled using the data pipeline and split-apply-combine approach [@plyr]. <!--AE comment: this is an important point that shouldn't be relegated to a parenthetical statement. --> (Tidy data shouldn't be confused with "tame data" which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called `yowie` which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.

This paper is structured in the following way. Section \@ref(database) describes the NLSY79 data source. Section \@ref(cleaning) presents the steps of cleaning the data, including getting and tidying the data from the NLSY79 and IDA to find and repair anomalies. Our final subset is compared to the old textbook subset in Section \@ref(compare). Finally, Section \@ref(summary) summarizes the contribution and makes recommendations for the NLSY79 data curators.

# The NLSY79 {#database}

## Database

The NLSY79 is a longitudinal survey administered by the U.S Bureau of Labor Statistics that follows the lives of a sample of American youth born between 1957-1964 [@nlsy79]. The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979. <!--AE comment: “It comprised of Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military.” Definitely needs to remove the term “Blacks” as that is considered to be offensive. Perhaps “It comprised of people sampled because they were Black, Hispanic, economically disadvantaged, or in the military.” --> It comprised of Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military. <!--AE comment: “1,643 members of the economically disadvantaged non-Black non-Hispanics, respectively” could perhaps be rephrased to “1,643 members who were selected because of their economic disadvantage, but were not Black or Hispanic.”-->In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members of the economically disadvantaged non-Black non-Hispanics, respectively. Hence, 9,964 respondents remain in the eligible samples. The surveys were conducted annually from 1979 to 1994 and biennially thereafter. Data are currently available from Round 1 (1979 survey year) to Round 28 (2018 survey year).

<!--AE comment: I found the semicolons distracting. Perhaps replace them with commas and use parenthetical for sub-categories, or use a bulleted list.-->
Although the main focus area of the NLSY is labor and employment, the NLSY also covers several other topics, including education; training and achievement; household, geography, and contextual variables; dating, marriage, cohabitation; sexual activity, pregnancy, and fertility; children; income, assets and program participation; health; attitudes and expectations; and crime and substance use.

There are two ways to conduct the interview of the NLSY79, which are face-to-face or by telephone interviews. In recent survey years, more than 90 percent of respondents were interviewed by telephone [@eliznlsy].

## Target data {#target}

<!--AE comment: What does “yearly mean hourly wages with years of workforce experience” mean? Is that one variable or two?-->
<!--AE comment: Was your goal to collect data from 1994-2018 or 1979-2018? Page 4 states 1994 onward, but from the figures later it appears that you collected data starting in 1979 (which makes more sense to me).-->
<!--AE comment: “The NLSY79 data used in Singer and Willett (2003) contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994.” Might work better split into two sentences. -->
The NLSY79 data used in @SingerJudithD2003Alda contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994. In addition, the cohort is restricted to male high-school dropouts who first participated in the study at age 14-17 years. 
<!--AE comment: Thus the target of our study is to collect the same variables for the same time period and extending through to 2018, the most recent year reported.-->Thus the target data set is to collect the same variables for the extended time frame of 1994 through to 2018, the most recent year reported.

<!-- Since no specific criteria on high school dropouts was defined
in @SingerJudithD2003Alda, we define high school dropouts as respondents who only completed either 9th to 11th grade, or who completed 12th when their age is at least 19 years old (meaning that the individual may have dropped out of high school but returned at a later time to complete high school education). XXX This should probably go further down in the flow. -->

# Data cleaning {#cleaning}

@LooMarkvander2018Sdcw describe the notion of a "statistical value chain" where the production stages of the data cleaning process are earmarked as raw data (data as arrived to the desk of an analyst), input data (data organised with correct type and identified variables) and valid data (data that faithfully represent the variables).  In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidy the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so that the resulting input data can be used downstream for validating the data as described in Section \@ref(ida).

## Getting the data {#getdata}

The NLSY79 data contains a large number of variables but for our aim, the scope required is limited to demographic profiles and wages data. More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data and navigated as described in Figure \ref{fig:source-nav}.

\begin{figure}[t]

\begin{tcolorbox}[title = Navigating the data source]
\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
\vspace{1mm}
\faCheck\ The CASEID will be always be selected.  \\
\vspace{1mm}
\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
\vspace{1mm}
For the remaining variables, we went to the "Variable Search" tab and select variables as follows
\begin{itemize}
\item[$\triangleright$] Education, Training and Achievement Scores
\begin{itemize}
\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
\begin{itemize}
\item[\faCheck] All 80 variables in Highest grade completed were selected.
\end{itemize}
\end{itemize}
\item[$\triangleright$] Employment
\begin{itemize}
\item[$\triangleright$] Summary measures $\triangleright$ By job
\begin{itemize}
\item[$\triangleright$] Hours worked  
\begin{itemize}
\item[\faCheck] All 442 primary variables in Hours worked were selected.
\end{itemize}
\item[$\triangleright$] Hourly wages
\begin{itemize}
\item[\faCheck] All 156 variables in Hourly wages were selected.
\end{itemize}
\end{itemize}
\end{itemize}
\item[$\triangleright$] Household, Geography and Contextual Variables
\begin{itemize}
\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics $\triangleright$ Date of birth
\begin{itemize}
\item[\faCheck] All 4 variables in Date of birth were selected. 
\end{itemize}
\end{itemize}
\end{itemize}
\begin{itemize}
\item[\faCloudDownload] To download all 686 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download". We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers. We name the filename "NLSY79" and press the download button. There are also options to get control or dictionary files for SAS, SPSS and STATA. 
\end{itemize}
\end{tcolorbox}
\caption{The above documents the steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
\end{figure}

The downloaded data set comes as a zip file, containing the following set of files:

- `NLSY79.csv`: Comma Separated Value format of the response data,
- `NLSY79.dat`: .dat format of the response data,
- `NLSY79.NLSY79`: Tagset of variables that can be uploaded to the website to recreate the data set, and
- `NLSY79.R`: R script for reading the data into R and converting the variables' names and label into something more sensible.

We alter only the file path in `NLSY79.R` and run the script without any other alteration. This results in an initial processing of the raw data into two data sets,  `categories_qnames` (where the observations are stored in categorical/interval values) and `new_data_qnames` (the observations are stored in integer form).

```{r raw-data}
```

The raw data, `new_data_qnames`, is organised such that each row corresponds to an individual. As respondents can have multiple jobs at specific years, the column names, such as `HRP1_1979`, `HRP2_1979`, `HRP1_1980` and `HRP2_1980`, contain the information about the job number up to 5 (`HRP1` = job 1, `HRP2` = job 2) and the year. The raw data consequently has a large number of columns (686 to be specific). The values in the cell under the variables that begin with `HRP` correspond to the hourly wage in dollars. A glimpse of this data output is shown below.

```{r untidy-data}
```

According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table. The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.<!-- AE comment: This sentence is unclear. What is dplyr used for? Are there tidyverse packages used but not mentioned? If not, splitting this sentence into two may make it clearer which package is used to what purpose. Also, I haven't tried it in code, but I imagine that the data could be tidied as described just using pivot_longer, at least for the job number, year, and wage data. In other words, I don't see why dplyr or stringr would be needed for the data described on page 6.--> This is done using the `tidyverse` suite of packages [@tidyverse], `tidyr` [@tidyr] to pivot the data into long form, `dplyr` [@dplyr], and `stringr` [@stringr] for creating new variables, and levels of factors by text wrangling. The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data. The code for tidying the data are demonstrated at \url{`r yowie_pkgdown`/articles/raw-to-input-data.html} but also described in the subsequent subsections.

<!-- required by making use of the `tidyverse` suite of packages [@tidyverse], which include the main set of packages needed, namely `tidyr` [@tidyr], `dplyr` [@dplyr], and `stringr` [@stringr]. The tidy data form, where we have a variable that indicates the individual ID, year, job number, wage in dollars and demographic variables, is important for downstream analysis, such as visualisation and modelling that will be performed later.

[REMOVE] Unfortunately, the `new_data_qnames` did not meet these requirements in the way that data for each year and job for the same respondent is stored as one variable. Hence, the data contains a huge number of columns (686 columns). The example of the untidy data set is displayed in Table \ref{tab:untidy-data}. The table intended to display the hourly rate of each respondent by job (HRP1 to HRP5) and by year (1979 and 1980). The table implies that the column headers are values of the year and job, not variable names. Consequently, the data should be tidied and wrangled first to extract the demographic and employment variables we want to put in the final data set. We mainly used `tidyr` [@tidyr], `dplyr` [@dplyr], and `stringr` [@stringr] to do this job.-->

### Tidying demographic variables {#tidydemog}

In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (`gender`), race (`race`), age (`age_1979`), highest grade completed (`hgc`), highest grade completed in terms of years, e.g. 9th grade = 9, 3rd year college = 15, (`hgc_i`) and the corresponding year this grade was completed (`yr_hgc`).

For gender and race, we only rename the column names. It is worth noting that using these two variable needs special attention. Gender, as reported in the data, only has two categories, which is recognised today as inadequate. Gender is not binary. Further, race is as reported in the database. When doing analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.

The `new_data_qnames` contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981` which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`). The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981. We checked for consistency between the two versions and found no discrepancy where the responses were recorded in both 1979 and 1981. The age was then calculated using the birth year.

```{r dob-tidy}
```
```{r demog-tidy}
```
```{r demog-ed}
```

<!--AE comment: I was confused by the following section: “We choose to use this revised May data because it seemed to have less missing and presumably has been checked. However, there is no revised May data for 2012, 2014, 2016, and 2018, thus, we use the ordinary May data for these years.” I was assuming the data was collected on a yearly basis, but this section suggests it was collected each month. Please clarify somewhere in the paper. 
-->
The next step is tidying to obtain `hgc` and `yr_hgc`. The highest grade completed are recorded in `new_data_qnames` as variables beginning with `Q3-4` and `HGC` with suffix of the year it was recorded. In addition the variables beginning with `HGCREV` contain the revised data. We choose to use this revised May data because it seemed to have less missing and presumably has been checked. However, there is no revised May data for 2012, 2014, 2016, and 2018, thus, we use the ordinary May data for these years.

```{r tidy-grade}
```

The `hgc` is measured and could be updated in each period of the survey. We chose to only retain the highest grade completed for each individual and derived the year when they completed it (`yr_hgc`) by finding the minimum year with the highest completed grade.

```{r tidy-hgc}
```

Finally, we get all of the demographic profiles of the NLSY79 cohort. We then save this data as `demog_nlsy79`.

```{r full-demog}
```

### Tidying employment variables {#tidyemp}

Our target variables for the employment are to obtain respondent's mean hourly wage (`mean_hourly_wage`), the number of jobs (`number_of_jobs`) and the total hours of work per week (`total_hours`) for each survey year. As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.

From 1979 to 1987, `new_data_qnames` only contains one version of hours worked per week for each job (in the variables with names starting with `QES-52A`). From 1988 onward, we selected the total hours worked per week, including hours working from home (`QES-52D`). However, in 1993, this variable was missing for the first and last job so we selected to use `QES-52A` instead. In addition, 2008 only had jobs 1-4 for the `QES-52D` variable, so we use only these.

```{r tidy-hours}
```

The hourly wages are in the variables beginning with `HRP` in `new_data_qnames`. As a respondent may have multiple jobs, the `mean_hourly_wage` is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on number of hours is available); if number of hours worked for any job is missing, then the `mean_hourly_wage` is computed as a simple average of all available hourly wages. Prior to computing the mean hourly wage, we undertook a number of steps to treat unusual observations as described below:

* If the hourly rate is recorded as 0, we set wage as missing.
* If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hour worked as missing.

The number of jobs (`number_of_jobs`) for each respondent per year is computed from the number of non-missing values of hourly wage *and* hours worked. If either the hourly wage or hours worked is missing, we do not tally this. <!-- AE comment: "If either the hourly wage or hours worked is missing, we do not tally this." I take "this" to mean "number_of_jobs". But in row 2, total_hours is missing, yet number_of_jobs is 1. The number of jobs (1) was tallied even though hours worked was missing. -->

```{r tidy-rate}
```

```{r tidy-nojob}
```

The employment and demographic variables are then joined. These data are further filtered to the cohort who completed education up to 12th grade and participated in at least five rounds in the survey. We save the resultant wage data on this cohort as `wages`.

```{r wages-demog-hs}
```

## Initial data analysis {#ida}

According to @HuebnerMariannePhD2016Asat, initial data analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis. Moreover, @Chatfield1985TIEo argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.

In this paper, we conduct an IDA or a preliminary data analysis to assess the consistency of the data with the cohort information that the NLSY provides. <!-- AE comments: This comes as a surprise to the reader. What anomaly? The wage anomalies should be mentioned in the introduction so as not to be a surprise at this point in the manuscript.
 -->In addition, we also aim to find the anomaly in the wages values using this approach. We mainly use graphical summaries to do the IDA using `ggplot2` [@ggplot2] and `brolgar` [@brolgar].

As stated previously, the respondents' ages ranged from 12 to 22 when first interviewed in 1979. Hence, we validate whether all of the respondents were in this range. Additionally, the [NLSY](https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction) also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic). To validate this, we used the `demog_nlsy79`, i.e., the data with the survey years 1979 sample. <!--AE comment:  I don't know that Table 1 is necessary as I don't have anything to compare it to. If the numbers check out, a simple statement to that effect would suffice. Also, what does "(?CHECK)" mean? This needs to be resolved.--> 
<!--AE comment: Table 1 includes the note ?CHECK which I assume was a note to self. -->
Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest that the demographic data we had is consistent with the sample information in the database.

```{r age-table}
```

<!--AE comment: Table 2 seems to have more precision than is necessary. Certainly 100% does not need to be reported as 100.00%, and probably all the percentages could be rounded to the nearest 1 percent. The caption could explain this, as there might be some rounding errors. -->

```{r gender-race-table}
```

In the next step, we explore the mean hourly wage data. In this case, we only explore the wages data of respondents that have the highest completed grade of up to $12^{th}$ grade.  We employ visualization techniques to perform the IDA as described next.

```{r summarytable, eval=FALSE}
```

<!--AE comment: Figure 2’s caption should include something about how that data is the raw values from the survey, before inputing anomalous values so the figure and caption can stand alone.-->

```{r sample-plot, echo = FALSE, fig.cap="Longitudinal profiles of wages for a random sample of 20 individuals in the refreshed data. Most, but not all, individuals here experienced increasing wages over time, and several have experienced considerable fluctuation in wages. Some individuals are only measured for a short period.", fig.height=8, fig.width=10, out.width="100%"}
```

<!--AE comment: With the IDA about wage data, I would appreciate a bit more context. Why sample 36 observations? I am not a longitudinal data expert, so I was a bit surprised when I flipped to the plot and found it to be small-multiple time series, although upon further consideration it makes sense. In that section it would be good to describe what the figure shows in a bit more detail. Perhaps “We randomly sample 36 respondents from the data, and plot their average wage as a time series. Looking at the patterns over time, we see a lot of variability in wages. For example, the people shown in panels 5, 7, and 11 have [explain what is interesting here, again I’m not an expert].” The next sentence here, “Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.” does not seem to provide much additional information. What were you looking for here? Overall trends over time, increasing/decreasing? Any places where one year’s wage looked really different?-->

We randomly take 20 samples from the data and plot them, as shown in Figure \ref{fig:sample-plot}.<!--AE comment: "It" implies... Please specify what "it" refers to. This could be done by adding onto the first sentence... "as shown in Figure 2, which shows that some respondents have high variability in wages, particularly 5, 7, and 11." However, 7 and 11 don't seem to have particularly high variability, especially when compared to 18. This whole paragraph needs to be fixed/updated.--> It shows that these respondents have a lot of variability in wages, for example, the IDs in panel numbers 5, 7, and 11. The plot also implies that the samples have a different pattern of mean hourly wages. Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, <!--AE comment: Not panel 9 --> the IDs in panel 9. 
<!--I think there is a narrative step missing here, which is that looking at those samples led you to look at plots that show values for the wage variable over all participants and all years. The phrase “the summary plots” was not enough description for me to know what to expect when I flipped to Figure 3. Please be explicit about what the three plots were, why they were made, and what you were looking for. I think there is another narrative step missing about Figure 3C, which is that after you saw there were outliers you tried to identify which respondents had the extreme values. Again, please spell this out in the text. When you return to the plots for the entire dataset in Figure 5, please make parallel comments. -->However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages. Some of them, for example respondents in Figure \ref {fig:feature-plot} C had experienced unusual wages only in certain year.

<!--For both Figure 3 and Figure 5, I would prefer the letter label before the description, but that’s a matter of personal preference.-->

```{r feature-plot, fig.cap = "Summary plots to check the cleaned data reveal more cleaning is necessary: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual (id=39) with an unusual wage relative to their years of data (C). Some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.", fig.width=10, fig.height=4, out.width="100%", cache = FALSE}
```

```{r high-wages}
```

```{r high-wages, fig.cap= "6 out of 45 IDs with extremely high mean hourly wage. Most of the IDs only have one point of high wage.", eval=FALSE}
```

<!--The next paragraph, which begins “The anomalies are also found” should perhaps be “Similar anomalies were found.” Was it the same respondents with the extremely high wages that showed extremely high number of hours of work? I suspect not, but please be explicit.-->
<!--AE comment: Rewrite this sentence. "Observations" aren't working 420 hours. A "respondent" might have been reported as having worked 420 hours per week.--> The anomalies are also found in the total hours of work, where some observations reported as having worked for 420 hours a week in total. 
According to @MichaelRPergamit2001DWTN, one of the flaws of the NLSY79 employment data is that the NLSY79 collects the information of the working hours since the last interview. Thus, it might be challenging for the respondents to track the within-job hours' changes between survey years, especially for the respondents with fluctuating working hours or seasonal jobs. It even has been more challenging since 1994, where the respondents had to recall two year periods. This shortcoming might also contribute to the fluctuation of one's wages data.

### Replacing extreme values {#censor}

<!--AE comment: Section 3.2.1 begins “As part of the IDA, which is the model formulation, we build a robust linear regression model to treat the extreme values in the data.” This could use clarification. In 3.2, you said that part of IDA is “ model formulation without any formal statistical inference.” To me, that sounds like using modeling as a descriptive technique, and it’s something I would consider to be more a part of EDA. But then in 3.2.1, while it is perhaps not using modeling in an inferential way, the model is being used for more than just description. I might rephrase the first sentence as follows: “In order to treat the extreme values in the data, we built a robust linear regression model.” -->
As part of the IDA, <!--AE comment: Awkward sentence. "which is the model formulation"?--> which is the model formulation, we build a robust linear regression model to treat the extreme values in the data. Robust linear regression yields an estimation robust to the influence of noise or contamination [@KollerManuel2016rARP]. It also aims to detect the contamination by weighting each observation based on how "well-behaved" they are, known as robustness weight. Observations with lower robustness weight are suggested as an outlier by this method [@KollerManuel2016rARP].

Since we work with longitudinal data, we build the model for each ID instead of the overall data. The robust mixed model could be the best model to be employed in this case. However, this method is too computationally and memory expensive, especially for a large data set, like the NLSY79 data. <!--I would pull the tidy and purrr citations to the end of the sentence to make it read more smoothly-->Therefore, the model for each ID is built utilizing the `nest` and `map` function from `tidyr` [@tidyr] and `purrr` [@purrr], respectively. The full code for this is shown at \url{`r yowie_pkgdown`/articles/input-to-valid-data.html} but also described in detail next.

We build the model using the `rlm` function from `MASS` package [@mass]. We set the `mean_hourly_wage` and `year` as the dependent and predictor, respectively. Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm]. <!--AE comment: “However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.” I don’t think this sentence needs a “However” and probably should say “past which” rather than “in which”-->However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers. Moreover, it should be noted that not all the outliers are due to an error. Instead, it might be that one had reasonably increasing or decreasing wages in a particular period.

<!--AE comment: At the bottom of page 11, where you describe the model formation, more detail would be appreciated. You are modeling the mean hourly wage based on year. Why? Is this standard practice for robust linear regression? What is a “slight” residual? Is it defined based on absolute magnitude, or standard deviations, or something else? Later, you say “To minimize the risk of mistakenly identifying an outlier as an “erroneous outlier”.” Is an erroneous outlier a technical term? If not, I think you are trying not to identify erroneous outliers, not mistakenly identifying outliers as erroneous outliers (that’s a double negative). The value of 0.12 feels very specific and out of place in a paper that has had few numerals. Is this a number an expert in RLR would be able to understand? Again, is that a raw value? Is it a standard deviation? Is there literature about how to choose a threshold? Overall, the sentence “We find that 0.12 is the most reasonable value to be the threshold to minimize that drawback’s risk because it still captures the sensible spikes in the data.” is vague and should be expanded upon. What is “that drawback’s risk”? What are “the sensible spikes in the data?” Be more explicit.-->

To minimize the risk of mistakenly identifying an outlier as an "erroneous outlier", we simulate some thresholds and study how they affect the data. We find that 0.12 is the most reasonable value to be the threshold to minimize that drawback's risk because it still captures the sensible spikes in the data. In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording. After deciding the threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value. We then flag those observations in a new variable called `is_pred`.

```{r rlm, eval=FALSE}
```

Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced. It implies that the fluctuation can still be observed in the data after the treatment. However, the large spikes, which are considered "erroneous outliers", are already eliminated from the data. Hence, the model produces a data set with a more reasonable degree of fluctuation.

```{r compare-data}
```

```{r compare-plot, fig.cap="Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for a sample of individuals. A robust linear model prediction was used to correct mean hourly wages value. We can see that some extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed. Some spikes might remain when wage vaues are plausible.", fig.height=8, fig.width=10, out.width="90%"}
```

Further, Figure \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around $350. The spikes are still observed but not as extreme as the original data set. In Figure \ref{fig:fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value. 
<!--AE comment: What does “It implies that the fluctuation can still be observed in the data after the treatment.” mean? I suspect this sentence should be replaced with something along the lines of “The figure shows that fluctuations in wages still exist even in the imputed data.”-->
We still can see some extreme values in maximum wages, but consider it as a natural variability of the data. In Figure \ref{fig:fixed-feature-plot} C, we can see that after imputing the extreme value in ID=39, we can see how the the wages change over the years more clearly.

```{r fixed-feature-plot, fig.cap = "Re-make of the summary plots of the fully processed data suggest that it is now in a reasonable state: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual with an unusual wage relative to their years of data (C). ", fig.width=10, fig.height=4, out.width="100%", cache = FALSE}
```

Finally, we save the imputed data and set the appropriate data type for the variables.

```{r save-data}
```

```{r nrow}
```

## Recap

Figure \ref{fig:flow-chart-blind} summarizes the steps taken to go from raw to input to valid data [@LooMarkvander2018Sdcw] to create a refreshed wages data set.

<!--To summarise, the  a workflow from getting the data from the NSLY79 database until a ready-analysed data. Referring to @LooMarkvander2018Sdcw, this section shows  part of statistical value chain, which is data cleaning from raw, input, and valid data. The raw demographic and employment data from the database are not in a tidy format. Hence, we perform data wrangling to get the input data. We inspect the input data to find eligible observations to the target data. One of the target data is wages data of respondents who completed up to 12th grade and participated at least in 5 rounds of survey. This needs us to join the extracted demographic and employment data.

Further, using IDA, we inspect on the existence of anomalies/extreme values in the data using the weight from robust linear model. We then replace these extreme values using their predicted value. 
From this stage forward, we assume that we already have valid data. We also subset the data to get the wages of high school dropouts. We then make these three data sets and the cleaning workflow documentation publicly available through an R data container package called `yowie`. A visualisation of these process is displayed in  Figure \ref{fig:flow-chart}.-->

(ref:flow-chart) The stages of data cleaning from the raw data to get three datasets contained in `r yowie`. "# of individuals" means the number of respondents included in each stage, while "# of observations" means the number of rows in the data. The color represents the stage of data cleaning in statistical value chain [@validate]. Pink, blue, and green represent the raw, input, and valid data, respectively.

```{r flow-chart-blind, fig.cap="(ref:flow-chart)", out.width="85%", eval = CENSOR}
```

```{r flow-chart, fig.cap="(ref:flow-chart)", out.width="85%", eval = !CENSOR}
```

# Comparison of refreshed with the original data {#compare}

```{r sw_wages}
```

```{r do_refreshed}
```

```{r compare-sw-do}
```

<!--AE comment: first sentence can be removed or rephrased. Next sentence includes the word wages too many times. Next sentence should say “dropouts” rather than “drop outs” to be consistent with the previous sentence. -->
Now is the time to see how close we have come to the original textbook data. The original set contains wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994. The refreshed data set for comparison is also wages on high school drop outs from 1979 to 2018. The original set is available in the R package `brolgar` and the refreshed data is available in the R package `r yowie`.

<!--AE comment: what does the natural log have to do with the y-axis?-->
<!--AE comment: For Figure 7, why did you choose to show the entire dataset here? It would be easier to make a comparison if the plot was cut off at 1994 on the x-axis.-->

```{r plotting-sw-do, fig.cap = "Comparison of original textbook example (A) with refreshed data (B). The original data was inflation-adjusted to 1990 prices and the individual's time of collection was converted to a length of experience in the workforce, which makes it difficult to precisely compare the two sets.", fig.width = 9, fig.height = 5, out.width = "100%", cache = FALSE}
```

<!--AE comment: In Section 4, you talk about the variables in the original data. “One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection.” I’m not sure what this means. Please clarify.-->
<!--AE comment: “experience in the workforce” should have the period inside the quotations.-->
There are two aspects of the original data that make a direct comparison difficult. The time variable provided is "experience in the workforce". It is not clear how this is calculated. One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection. We have not been able to find this information in the database. Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.

<!--AE comment: “The treatment of unlikely wages differ in the refreshed data.” I think this means “The treatment of anomalous wages differs between the original and refreshed data sets.” Then you say “We opted to use the weights from a robust linear regression to determine what should be set as missing value as described in Section 3.2.1.” I don’t think you were setting missing values though, were you? It was used to determine which values should be imputed, correct? Either way, please clarify.-->
The treatment of unlikely wages differ in the refreshed data. In the original data by @SingerJudithD2003Alda, wages greater than $75 are set to be missing. However, in recent context, this value is too low to be set as the maximum threshold. We opted to use the weights from a robust linear regression to determine what should be set as missing value as described in Section \@ref(censor).

Figure \ref{fig:plotting-sw-do} shows a comparison of the two sets. (A direct ID matching is not possible.) There are 888 individuals in the original and 1,188 individuals in the refreshed data. This suggests some individuals were removed in the original data.
In addition, in the original set the racial breakdown was 246 Black, 204 Hispanic and 438 White participants, while in the refreshed data there are 346 Black, 219 Hispanic, and 623 White participants, so the proportions have not been exactly reproduced. On education, in the refreshed data there are very few individuals with less than a 12th grade education, which is different from the original. <!--AE comment: How many individuals up to 8th grade are in the refreshed dataset? What are plausible reasons for these differences?--> In the original data there were 366 individuals with up to 8th grade and 522 9th-12th grade, as compared to 967 with 12th grade in the refreshed data.

```{r summaries, eval=FALSE}
```

<!-- Hence, in this section, we will compare how close the `wages_hs_do`, henceforth refreshed data, to the textbook data. It is also worth noting that the textbook data is reused from @MurnaneRichardJ1999DMDB’s paper. Hence, it is not directly extracted from the NLSY79 database.  @SingerJudithD2003Alda do not state the time frame covered by the data. However, from @MurnaneRichardJ1999DMDB’s paper, we find that the textbook data cover the survey period from 1979-1994. Hence, for comparison purposes, we filter the refreshed data only until 1994.

In the absence of highs chool dropouts criteria in the textbook data, we create a clearer criteria of high school dropouts as mentioned in Section \@ref(#target). Consequently, we do not have the same ID of high-school dropouts as in the textbook data. The textbook data has `r nrow(sw_id)` IDs included, while the refreshed data has `r nrow(do_id)` IDs. Furthermore, there are only `r nrow(sw_do)` IDs included both in the textbook and the refreshed data. After observing @MurnaneRichardJ1999DMDB’s paper, it is mentioned that the data is randomly drawn from the low income subsample of the NLSY. This is probably the reason why, even using some criteria, we cannot match the dropouts in the textbook and the refreshed data.

Further, the temporal variable used in the textbook data is experience, instead of year. The only explanation we found regarding this variable is that it is derived from the years of IDs’ experience since their first job. @SingerJudithD2003Alda stated that this variable reflects the specific moment in each ID labor force history associated with each wage observation. 
Since it is not clearly articulated, we decide to keep the temporal variable in the refreshed data as year. Accordingly, in this comparison, we only compare the same ID but not the same temporal variable.

Regarding the wages, the textbook data express the wages as inflation adjusted with 1990 dollars as constant [@SingerJudithD2003Alda]. However, it is not stated how exactly the wages are adjusted. Note that in the refreshed data we publish in `yowie`, we do not adjust the data with inflation because the time frame covered ranged from 1979-2018. Hence, we allow flexibility to the data user to adjust, not adjust,  or adjust it with different base years. For comparison purposes, we adjust the wages with the average Consumer Price Index For All Urban Consumers (CPI-U) we got from @fed. -->

<!-- Figure \ref{fig:plotting-sw-do} shows the comparison between the textbook and the refreshed data. Even though it has different temporal variables and possibly different CPI indicators used in inflation adjustment, we can still capture some of the patterns in the refreshed data.

From this comparison, we learn that we cannot make an apple to apple comparison and produce the exact same data as the textbook due to the absence of clearly articulated data extraction and derivation process. Hence, the takeaway is that it is imperative to disclose how the data derived from the initial source to make data analysis reproducible. This is why in this paper, we show the practice of reproducible and responsible data cleaning workflow to make sure the longitudinal data from a continuous survey can be refreshed.-->

# Summary {#summary}

This paper has described the stages to take a particular open data set and make it a textbook data set, ready for the classroom or research. In the first stage, we showed the steps performed to get the data from the NLSY79 database. The data format needed conversion to tidy format, and this was described. After that, an initial data analysis was conducted to investigate and screen the quality of the data. We found and fixed the anomalous observations in wages using a robust linear regression model. Finally a comparison is made between the original and refreshed data sets.

The data cleaning process is documented and the code has been made available. These provide the opportunity to again refresh the textbook data as updated data is published in the NLSY79 database.
Determining the appropriate robustness weight to threshold the anomalous observations is documented, and a `shiny`  [@shiny] app is provided to assist. 
The current subset is made available in a new R package, called `r yowie`.

<!--AE comment: The bulleted difficulties in Section 5 are a bit jarring. Make them parallel (i.e. every bullet should start with a verb -ing so it is determining, calculating, etc) or turn them into a paragraph. I liked the specificity of $30,000/hour but it would be better up in the section where you discuss the anomalous values.-->

Various difficulties were encountered in trying to refresh the data, which include:

- determining which records should be downloaded from the database.
- there are many errors in the data, e.g. hourly wages greater than $30,000 per hour.
- there is no explicit variable in the database recording high school dropout, which means we needed to compare date of 12th grade with the individual's age.
- calculating experience in the workforce would require knowing the time of first job within the first year the individual was recorded.

Ultimately, the refreshed data is reasonably similar to the original, but unsatisfactorily far from it. The last step required would be to inflation-adjust wages, but this is better to do with each wave of new data added, so that it is relative to the last date in the data.

<!--AE comment: At the end of page 14, it would be good to explicitly say that if people don’t like your decisions, they can grab the code and adjust the parameters themselves! It’s implied, but should be made explicit. -->
Some readers may disagree with our decisions made to produce the refreshed textbook data and may have better insight than us in producing a more appropriate textbook data. We do not assert that we have produced the best textbook data, but rather we describe our journey to provide a reasonable textbook data set. All code and documentation are provided for transparency, and future updates of the `r yowie` package may contain additional variables, or filters of the full set, if it is deemed important.

Finally, for the data providers we recommend that a validation system with clear rules is added on data entry, and that alternative output formats, such as a tidy format would help users make better use of the resource. The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs. These values can be validated with simple checks at data entry. Providing an open data resource also is accompanied with the responsibility that the data, especially data that is as valuable as this, is reliable. Users need to be able to trust the data.

<!--this paper implies that data providers should design a database that is able to produce tidy data sets. A data provider should also check for data anomalies before the data publishing or at least provides a set of rules or threshold values. For example, in this case, is the threshold of reasonable wages. This will greatly support the data users to validate and set the same understanding of which data are considered outliers. Moreover, providing validation rules would facilitate any established data validation tool, such as `validate` [@validate] package. In this case, we cannot use this handy validation package due to the absence of validation rules. -->

# Acknowledgements

We would like to thank Aarathy Babu for the insight and discussion during the writing of this paper.

The entire analysis is conducted using `R` [@R] in RStudio IDE using these packages: `tidyverse` [@tidyverse], `ggplot2` [@ggplot2], `dplyr` [@dplyr], `readr` [@readr], `tidyr` [@tidyr], `stringr` [@stringr], `purrr` [@purrr], `brolgar` [@brolgar], `patchwork` [@patchwork], `kableExtra` [@kableExtra], `MASS` [@mass], `janitor` [@janitor], and `tsibble` [@tsibble]. The paper was generated using `knitr` [@knitr] and `rmarkdown` [@rmarkdown].

# Supplementary Materials

- **Codes**: R script to reproduce data tidying and cleaning is available in this [page](`r yowie_pkgdown`/articles/process-data.html).

- **R Package**:`r yowie` is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort. This package can be accessed [here](`r yowie_repo`).

- **shiny app**: An interactive `shiny` web app to visualise the effect of selecting different weight threshold for substituting the wages data to its predicted value from a fit of the robust linear regression model. This app can be accessed [here](`r yowie_shiny`) with the source code provided [here](`r yowie_repo`/tree/master/app).

# Data Availability Statement

The authors confirm that the data supporting the findings of this study are available within the supplementary materials.

# References

### 2022-01-14 paper/paper-jsdse/paper-jsdse.Rmd
revised introduction

---
authors:
  - name: Dewi Amaliah
    thanks: "Corresponding author, dlamaleeah@gmail.com"
    department: Dept of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dlamaleeah@gmail.com
  - name: Dianne Cook
    department: Dept of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: dicook@monash.edu
  - name: Emi Tanaka
    department: Dept of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: emi.tanaka@monash.edu
  - name: Kate Hyde
    thanks: ""
    department: Dept of Econometrics and Business Statistics
    affiliation: Monash University
    location: Clayton, VIC 3800
    email: ""
  - name: Nicholas Tierney
    thanks: ""
    department: 
    affiliation: Telethon Kids Institute
    location: Nedlands, WA 6009
    email: ""
blind: 1 # 0 for no
bibliography: ../references.bib
#preamble: >
output: 
  bookdown::pdf_document2:
    template: TAS-template-for-rmd.tex
    keep_tex: true
date: "`r Sys.Date()`"
always_allow_html: true
---

```{r, include = FALSE}
paper <- here::here("paper/paper.Rmd")
meta <- rmarkdown::yaml_front_matter(paper)
```

```{r, child = "../paper.Rmd"}
```

---
title: "`r meta$title`"
abstract: >
  `r meta$abstract`
keywords: `r meta$keywords`
---

### 2022-01-14 paper/paper-jsdse/paper-jsdse.tex
revised introduction

@@
             Nicholas Tierney\\
-      Telethon Kids Institute, Perth, WA, }
+      , Telethon Kids Institute, }
         \maketitle
@@
 
+Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
+An example of this the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
+The freely available data contains records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79).
+
+This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was developed into a case study for use in teaching exploratory data analysis at Iowa State University.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+
 ``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021).
@@
 However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape.
-Huebner, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (Huebner et al.
+Huebner Marianne, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (M.
+Huebner et al.
 2020).
@@
 There are few research papers that document the data cleaning (Wickham 2014).
-Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (Huebner et al.
+Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (M.
+Huebner et al.
 2020).
@@
 
-Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
+Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al.
 2019).
@@
 
+M.
 van der Loo and de Jonge (2018) describe the notion of a ``statistical value chain'' where the production stages of the data cleaning process are earmarked as raw data (data as arrived to the desk of an analyst), input data (data organised with correct type and identified variables) and valid data (data that faithfully represent the variables).
@@
 The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
-This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
-2019), \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham, François, et al.
+This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
+2019), \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham et al.
 2020), and \texttt{stringr} (Wickham 2019) for creating new variables, and levels of factors by text wrangling.
@@
 
-According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis.
+According to Huebner Marianne, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis.
 Moreover, Chatfield (1985) argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
@@
 
-Figure \ref{fig:flow-chart-blind} summarizes the steps taken to go from raw to input to valid data (van der Loo and de Jonge 2018) to create a refreshed wages data set.
+Figure \ref{fig:flow-chart-blind} summarizes the steps taken to go from raw to input to valid data (M. van der Loo and de Jonge 2018) to create a refreshed wages data set.
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie/paper/figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{/Users/cookd/numbats/yowie-paper/paper/figures/flowchart_blind}
 
@@
 ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data.
-The color represents the stage of data cleaning in statistical value chain (van der Loo and de Jonge 2021).
+The color represents the stage of data cleaning in statistical value chain (M.
+P.
+J.
+van der Loo and de Jonge 2021).
 Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart-blind}
@@
 
-The entire analysis is conducted using \texttt{R} (R Core Team 2020) in RStudio IDE using these packages: \texttt{tidyverse} (Wickham, Averick, et al.
-2019), \texttt{ggplot2} (Wickham 2016), \texttt{dplyr} (Wickham, François, et al.
+The entire analysis is conducted using \texttt{R} (R Core Team 2020) in RStudio IDE using these packages: \texttt{tidyverse} (Wickham et al.
+2019), \texttt{ggplot2} (Wickham 2016), \texttt{dplyr} (Wickham et al.
 2020), \texttt{readr} (Wickham and Hester 2020), \texttt{tidyr} (Wickham 2020), \texttt{stringr} (Wickham 2019), \texttt{purrr} (Henry and Wickham 2020), \texttt{brolgar} (Tierney, Cook, and Prvan 2020), \texttt{patchwork} (Pedersen 2020), \texttt{kableExtra} (Zhu 2019), \texttt{MASS} (Venables and Ripley 2002), \texttt{janitor} (Firke 2020), and \texttt{tsibble} (Wang, Cook, and Hyndman 2020).
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-nlsy79}{}%
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
 Bureau of Labor Statistics, U.S.
@@
 2021.
-``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
+{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
 Columbus, OH.
 
-\leavevmode\hypertarget{ref-shiny}{}%
-Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.
+\leavevmode\vadjust pre{\hypertarget{ref-shiny}{}}%
+Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{{shiny: Web Application Framework for R}}. \url{https://CRAN.R-project.org/package=shiny}.
 
-\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
-Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
+\leavevmode\vadjust pre{\hypertarget{ref-Chatfield1985TIEo}{}}%
+Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
 
-\leavevmode\hypertarget{ref-eliznlsy}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-eliznlsy}{}}%
 Cooksey, Elizabeth C.
 2017.
-``Using the National Longitudinal Surveys of Youth (Nlsy) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
+{``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M.
 Lerner Neal Halfon Christoper B.
@@
 Cham: Springer.
-\url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.
+https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.
 
-\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-DasuTamraparni2003Edma}{}}%
 Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.
 
-\leavevmode\hypertarget{ref-janitor}{}%
-Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.
+\leavevmode\vadjust pre{\hypertarget{ref-janitor}{}}%
+Firke, Sam. 2020. \emph{{janitor: Simple Tools for Examining and Cleaning Dirty Data}}. \url{https://CRAN.R-project.org/package=janitor}.
 
-\leavevmode\hypertarget{ref-racismnotrace}{}%
-Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.
+\leavevmode\vadjust pre{\hypertarget{ref-racismnotrace}{}}%
+Fullilove, M. T. 1998. {``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.
 
-\leavevmode\hypertarget{ref-grimshaw}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
 Grimshaw, Scott D.
 2015.
-``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14.
+{``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14.
 \url{https://doi.org/10.1080/00031305.2015.1081106}.
 
-\leavevmode\hypertarget{ref-purrr}{}%
-Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.
+\leavevmode\vadjust pre{\hypertarget{ref-purrr}{}}%
+Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.
 
-\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
 Huebner, Marianne, Werner Vach, and Saskia le Cessie.
 2016.
-``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
+{``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
 
-\leavevmode\hypertarget{ref-HuebnerMarianne2020Haar}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMarianne2020Haar}{}}%
 Huebner, Marianne, Werner Vach, Saskia le Cessie, Carsten Oliver Schmidt, and Lara Lusa.
 2020.
-``Hidden Analyses: A Review of Reporting Practice and Recommendations for More Transparent Reporting of Initial Data Analyses.'' \emph{BMC Medical Research Methodology} 20 (1): 61--61.
+{``Hidden Analyses: A Review of Reporting Practice and Recommendations for More Transparent Reporting of Initial Data Analyses.''} \emph{BMC Medical Research Methodology} 20 (1): 61--61.
 
-\leavevmode\hypertarget{ref-tamedata}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-ilk2004}{}}%
+Ilk, Ozlem.
+2004.
+{``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.''} PhD thesis, Iowa State University.
+\url{https://doi.org/10.31274/rtd-180813-11012}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
 Kim, A.
@@
 2018.
-``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1).
+{``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1).
 \url{https://doi.org/10.5070/T511103589}.
 
-\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
-Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.
+\leavevmode\vadjust pre{\hypertarget{ref-KollerManuel2016rARP}{}}%
+Koller, Manuel. 2016. {``{robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models}.''} \emph{Journal of Statistical Software} 75 (6): 1--24.
 
-\leavevmode\hypertarget{ref-opendata}{}%
-Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.
+\leavevmode\vadjust pre{\hypertarget{ref-opendata}{}}%
+Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.
 
-\leavevmode\hypertarget{ref-patchwork}{}%
-Pedersen, Thomas Lin. 2020. \emph{patchwork: The Composer of Plots}. \url{https://CRAN.R-project.org/package=patchwork}.
+\leavevmode\vadjust pre{\hypertarget{ref-patchwork}{}}%
+Pedersen, Thomas Lin. 2020. \emph{{patchwork: The Composer of Plots}}. \url{https://CRAN.R-project.org/package=patchwork}.
 
-\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-MichaelRPergamit2001DWTN}{}}%
 Pergamit, Michael R., Charles R.
@@
 2001.
-``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
+{``Data Watch: The National Longitudinal Surveys.''} \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
 
-\leavevmode\hypertarget{ref-R}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
 R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
 
-\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-SingerJudithD2003Alda}{}}%
 Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.
 
-\leavevmode\hypertarget{ref-brolgar}{}%
-Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.
+\leavevmode\vadjust pre{\hypertarget{ref-brolgar}{}}%
+Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}}. \url{https://github.com/njtierney/brolgar}.
 
-\leavevmode\hypertarget{ref-tukey}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tukey}{}}%
 Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.
 
-\leavevmode\hypertarget{ref-rlm}{}%
-UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
+\leavevmode\vadjust pre{\hypertarget{ref-rlm}{}}%
+UCLA: Statistical Consulting Group. 2021. {``{Robust Regression \textbar{} R Data Analysis Examples}.''} February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
 
-\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
-van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.
+\leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%
+van der Loo, Mark P. J., and Edwin de Jonge. 2021. {``Data Validation Infrastructure for {R}.''} \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
 
-\leavevmode\hypertarget{ref-validate}{}%
-van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
+\leavevmode\vadjust pre{\hypertarget{ref-LooMarkvander2018Sdcw}{}}%
+van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in r}.
 
-\leavevmode\hypertarget{ref-mass}{}%
-Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
+\leavevmode\vadjust pre{\hypertarget{ref-mass}{}}%
+Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with s}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
 
-\leavevmode\hypertarget{ref-tsibble}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tsibble}{}}%
 Wang, Earo, Dianne Cook, and Rob J Hyndman.
 2020.
-``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
+{``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.''} \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
 \url{https://doi.org/10.1080/10618600.2019.1695624}.
 
-\leavevmode\hypertarget{ref-plyr}{}%
-Wickham, Hadley. 2011. ``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
+\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
+Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
 
-\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
----------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.
+\leavevmode\vadjust pre{\hypertarget{ref-WickhamHadley2014TD}{}}%
+---------. 2014. {``Tidy Data.''} \emph{Journal of Statistical Software} 59 (10): 1--23.
 
-\leavevmode\hypertarget{ref-ggplot2}{}%
----------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
+\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
+---------. 2016. \emph{{ggplot2: Elegant Graphics for Data Analysis}}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
 
-\leavevmode\hypertarget{ref-stringr}{}%
----------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.
+\leavevmode\vadjust pre{\hypertarget{ref-stringr}{}}%
+---------. 2019. \emph{{stringr: Simple, Consistent Wrappers for Common String Operations}}. \url{https://CRAN.R-project.org/package=stringr}.
 
-\leavevmode\hypertarget{ref-tidyr}{}%
----------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.
+\leavevmode\vadjust pre{\hypertarget{ref-tidyr}{}}%
+---------. 2020. \emph{{tidyr: Tidy Messy Data}}. \url{https://CRAN.R-project.org/package=tidyr}.
 
-\leavevmode\hypertarget{ref-tidyverse}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al.
 2019.
-``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686.
+{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686.
 \url{https://doi.org/10.21105/joss.01686}.
 
-\leavevmode\hypertarget{ref-dplyr}{}%
-Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.
+\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
+Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{{dplyr: A Grammar of Data Manipulation}}. \url{https://CRAN.R-project.org/package=dplyr}.
 
-\leavevmode\hypertarget{ref-readr}{}%
-Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.
+\leavevmode\vadjust pre{\hypertarget{ref-readr}{}}%
+Wickham, Hadley, and Jim Hester. 2020. \emph{{readr: Read Rectangular Text Data}}. \url{https://CRAN.R-project.org/package=readr}.
 
-\leavevmode\hypertarget{ref-knitr}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-knitr}{}}%
 Xie, Yihui.
 2014.
-``Knitr: A Comprehensive Tool for Reproducible Research in R.'' In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
+{``Knitr: A Comprehensive Tool for Reproducible Research in {R}.''} In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
 Peng.
@@
 
-\leavevmode\hypertarget{ref-rmarkdown}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-rmarkdown}{}}%
 Xie, Yihui, Christophe Dervieux, and Emily Riederer. 2020. \emph{R Markdown Cookbook}. Boca Raton, Florida: Chapman; Hall/CRC. \url{https://bookdown.org/yihui/rmarkdown-cookbook}.
 
-\leavevmode\hypertarget{ref-kableExtra}{}%
-Zhu, Hao. 2019. \emph{kableExtra: Construct Complex Table with 'kable' and Pipe Syntax}. \url{https://CRAN.R-project.org/package=kableExtra}.
+\leavevmode\vadjust pre{\hypertarget{ref-kableExtra}{}}%
+Zhu, Hao. 2019. \emph{{kableExtra: Construct Complex Table with 'kable' and Pipe Syntax}}. \url{https://CRAN.R-project.org/package=kableExtra}.
+
+\end{CSLReferences}
 

### 2022-01-14 paper/paper.Rmd
Add the missing code chunk.
revised introduction

@@
     department: ""
-    affiliation: ""
-    location: ""
+    affiliation: Telethon Kids Institute
+    location: Nedlands, WA 6009
     email: ""
@@
 
+Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
+An example of this the wages data made public by @SingerJudithD2003Alda which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
+The freely available data contains records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
+
+This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook [@ilk2004], and was developed into a case study for use in teaching exploratory data analysis at Iowa State University.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+The story from an exploratory analysis is very different: the individual experience varies a lot from the overall average.
+Some individuals experience a decline in wages the longer they are in the workforce, and most experience volatility in their wages.
+
+This disparity between the average and the individual is a part of statistics that is given less attention.
+It is this disparity which makes this particular data set very important as a textbook example.
+Textbook data sets have staying power if it has a challenge that has never been completely solved.
+The iris data (XXX REF) is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
+(A side note: the iris data is best replaced today with the penguins data (XXX REF), which has similar qualities, is new and does not suffer from a connection with eugenics [@notiris].) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
+
+For the field of statistics, and data science by association, it is increasingly important to reach the individual.
+One might describe this as a divergence of purpose, statistics for public policy or statistics for the public.
+The two are not the same.
+As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual.
+For example, with the wages data, the message for the individual is that you are more than your demographic.
+The majority of people tracked are not like the average, the average is an anomaly.
+If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
+
+As a textbook data though, the dates are tired.
+The most recent year in the data is 1994, 10 years prior to when @SingerJudithD2003Alda was published.
+Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students.
+Using tired old textbook data sets sets up a misconception that the field is not current.
+The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated.
+It should be possible to continuously refresh the textbook data from the data repository.
+This is our story of the journey from open [@opendata] but wild data to textbook data, it's not entirely pretty.
+
+@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
+This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN].
+In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research [@MichaelRPergamit2001DWTN; @eliznlsy].
+According to @eliznlsy, thousands of articles, and hundreds of book chapters and monographs have utilized this data.
+Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data [@MichaelRPergamit2001DWTN].
+<!--AE comment: For example, I would like more discussion about the textbook selected as the exemplar.
+It sounds like this dataset is used in many books, why choose that one?
+The text should include the name of the book, so readers do not need to flip to the citations in order to learn what it was.
+From the name, I am guessing that the book focuses heavily on this dataset, which is why it makes sense to reproduce their subset.
+Be explicit about this.-->
+Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences.
+Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda.
+We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the *purposes of studying racism, not race* [@racismnotrace].
+<!--However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.-->
+
+<!-- 
 "Open data" is data that is freely accessible, modifiable, and shareable by anyone for any purpose [@opendata].
@@
 However, open data is often referred to as what we might call "wild data" because it requires substantial cleaning and tidying to tame it into textbook shape.
-@HuebnerMariannePhD2016Asat emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building <!-- AE comment: It's unclear from context what is being referred to from the Huebner et al 2020 citation -->  [@HuebnerMarianne2020Haar].
+@HuebnerMariannePhD2016Asat emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building --> <!-- AE comment: It's unclear from context what is being referred to from the Huebner et al 2020 citation  [@HuebnerMarianne2020Haar].
+-->
+<!-- Data cleaning can be considered to be a part of what is called "initial data analysis" (IDA) [@Chatfield1985TIEo].
+In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
+This is also related to exploratory data analysis (EDA), coined by @tukey with a focus on learning from data.
+EDA can be considered to encompass IDA.
+@DasuTamraparni2003Edma say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80% of the data mining task.
+-->
+
+<!-- Despite it's importance, this IDA stage is often undervalued and neglected [@Chatfield1985TIEo].
+There are few research papers that document the data cleaning [@WickhamHadley2014TD].
+Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties [@HuebnerMarianne2020Haar].
+-->
+
+This paper demonstrates the steps of cleaning data, and documents the process, as recommended by @HuebnerMariannePhD2016Asat.
+They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
 Data cleaning can be considered to be a part of what is called "initial data analysis" (IDA) [@Chatfield1985TIEo].
@@
 @DasuTamraparni2003Edma say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80% of the data mining task.
-
-Despite it's importance, this IDA stage is often undervalued and neglected [@Chatfield1985TIEo].
-There are few research papers that document the data cleaning [@WickhamHadley2014TD].
-Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties [@HuebnerMarianne2020Haar].
-
-This paper demonstrates the steps of IDA, and documents the process, for a prominent open data source, National Longitudinal Survey of Youth 1979, henceforth referred to NLSY79.
-This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN].
-In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research [@MichaelRPergamit2001DWTN; @eliznlsy].
-According to @eliznlsy, thousands of articles, and hundreds of book chapters and monographs have utilized this data.
-Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data [@MichaelRPergamit2001DWTN].
-
-@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
-<!--AE comment: For example, I would like more discussion about the textbook selected as the exemplar.
-It sounds like this dataset is used in many books, why choose that one?
-The text should include the name of the book, so readers do not need to flip to the citations in order to learn what it was.
-From the name, I am guessing that the book focuses heavily on this dataset, which is why it makes sense to reproduce their subset.
-Be explicit about this.-->
-Our aim is to refresh this textbook data with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences.
-Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda.
-We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the purposes of studying racism, not race [@racismnotrace].
-<!--However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.-->
 
@@
 ```{r tidy-rate}
+```
+
+```{r tidy-rate-hour}
 ```

### 2022-01-14 paper/paper.tex
revised introduction

@@
 \usepackage{microtype}      % microtypography
-\usepackage{lipsum}
 \usepackage{graphicx}
@@
    \\
-     \\
-   \\
+    Telethon Kids Institute \\
+  Nedlands, WA 6009 \\
   \texttt{} \\
@@
 
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% From pandoc table feature
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+
 % Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
+  % set entry spacing
+  \setlength{\parskip}{#2\cslentryspacingunit}
+ }%
+ {}
+\usepackage{calc}
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 \usepackage{xcolor}
-
 \begin{document}
 \maketitle
-
-\def\tightlist{}
 
@@
 
-``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021).
-This type of data can be useful as example data in statistical textbooks and for research purposes.
-However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape.
-Huebner, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (Huebner et al.
-2020).
-Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985).
-In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
-This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
-EDA can be considered to encompass IDA.
-Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.
+Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
+An example of this the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
+The freely available data contains records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
+Department of Labor 2021).
 
-Despite it's importance, this IDA stage is often undervalued and neglected (Chatfield 1985).
-There are few research papers that document the data cleaning (Wickham 2014).
-Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (Huebner et al.
-2020).
+This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was developed into a case study for use in teaching exploratory data analysis at Iowa State University.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+The story from an exploratory analysis is very different: the individual experience varies a lot from the overall average.
+Some individuals experience a decline in wages the longer they are in the workforce, and most experience volatility in their wages.
 
-This paper demonstrates the steps of IDA, and documents the process, for a prominent open data source, National Longitudinal Survey of Youth 1979, henceforth referred to NLSY79.
+This disparity between the average and the individual is a part of statistics that is given less attention.
+It is this disparity which makes this particular data set very important as a textbook example.
+Textbook data sets have staying power if it has a challenge that has never been completely solved.
+The iris data (XXX REF) is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
+(A side note: the iris data is best replaced today with the penguins data (XXX REF), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
+
+For the field of statistics, and data science by association, it is increasingly important to reach the individual.
+One might describe this as a divergence of purpose, statistics for public policy or statistics for the public.
+The two are not the same.
+As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual.
+For example, with the wages data, the message for the individual is that you are more than your demographic.
+The majority of people tracked are not like the average, the average is an anomaly.
+If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
+
+As a textbook data though, the dates are tired.
+The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
+Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students.
+Using tired old textbook data sets sets up a misconception that the field is not current.
+The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated.
+It should be possible to continuously refresh the textbook data from the data repository.
+This is our story of the journey from open (Open Knowledge Foundation 2021) but wild data to textbook data, it's not entirely pretty.
+
+Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
 This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al.
@@
 2001).
+Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
+Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).
+We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the \emph{purposes of studying racism, not race} (Fullilove 1998).
 
-Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
-Our aim is to refresh this textbook data with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
-Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).
-We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the purposes of studying racism, not race (Fullilove 1998).
+This paper demonstrates the steps of cleaning data, and documents the process, as recommended by Huebner, Vach, and Cessie (2016).
+They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
+Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985).
+In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
+This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
+EDA can be considered to encompass IDA.
+Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.
 
-Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
+Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al.
 2019).
@@
 Cham: Springer.
-\url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.
+https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.
 
-\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-DasuTamraparni2003Edma}{}}%
 Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.
 
-\leavevmode\hypertarget{ref-janitor}{}%
-Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.
+\leavevmode\vadjust pre{\hypertarget{ref-janitor}{}}%
+Firke, Sam. 2020. \emph{{janitor: Simple Tools for Examining and Cleaning Dirty Data}}. \url{https://CRAN.R-project.org/package=janitor}.
 
-\leavevmode\hypertarget{ref-racismnotrace}{}%
-Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.
+\leavevmode\vadjust pre{\hypertarget{ref-racismnotrace}{}}%
+Fullilove, M. T. 1998. {``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.
 
-\leavevmode\hypertarget{ref-grimshaw}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
 Grimshaw, Scott D.
 2015.
-``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14.
+{``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14.
 \url{https://doi.org/10.1080/00031305.2015.1081106}.
 
-\leavevmode\hypertarget{ref-purrr}{}%
-Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.
+\leavevmode\vadjust pre{\hypertarget{ref-purrr}{}}%
+Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.
 
-\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
 Huebner, Marianne, Werner Vach, and Saskia le Cessie.
 2016.
-``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
+{``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
 
-\leavevmode\hypertarget{ref-HuebnerMarianne2020Haar}{}%
-Huebner, Marianne, Werner Vach, Saskia le Cessie, Carsten Oliver Schmidt, and Lara Lusa.
-2020.
-``Hidden Analyses: A Review of Reporting Practice and Recommendations for More Transparent Reporting of Initial Data Analyses.'' \emph{BMC Medical Research Methodology} 20 (1): 61--61.
+\leavevmode\vadjust pre{\hypertarget{ref-ilk2004}{}}%
+Ilk, Ozlem.
+2004.
+{``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.''} PhD thesis, Iowa State University.
+\url{https://doi.org/10.31274/rtd-180813-11012}.
 
-\leavevmode\hypertarget{ref-tamedata}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
 Kim, A.
@@
 2001.
-``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
+{``Data Watch: The National Longitudinal Surveys.''} \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
 
-\leavevmode\hypertarget{ref-R}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
 R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
 
-\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-SingerJudithD2003Alda}{}}%
 Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.
 
-\leavevmode\hypertarget{ref-brolgar}{}%
-Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.
+\leavevmode\vadjust pre{\hypertarget{ref-notiris}{}}%
+Stodel, Megan. 2020. {``Stop Using Iris.''} \url{https://www.meganstodel.com/posts/no-to-iris/}.
 
-\leavevmode\hypertarget{ref-tukey}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-brolgar}{}}%
+Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}}. \url{https://github.com/njtierney/brolgar}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tukey}{}}%
 Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.
 
-\leavevmode\hypertarget{ref-rlm}{}%
-UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
+\leavevmode\vadjust pre{\hypertarget{ref-rlm}{}}%
+UCLA: Statistical Consulting Group. 2021. {``{Robust Regression \textbar{} R Data Analysis Examples}.''} February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
 
-\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
-van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.
+\leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%
+van der Loo, Mark P. J., and Edwin de Jonge. 2021. {``Data Validation Infrastructure for {R}.''} \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
 
-\leavevmode\hypertarget{ref-validate}{}%
-van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
+\leavevmode\vadjust pre{\hypertarget{ref-LooMarkvander2018Sdcw}{}}%
+van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in r}.
 
-\leavevmode\hypertarget{ref-mass}{}%
-Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
+\leavevmode\vadjust pre{\hypertarget{ref-mass}{}}%
+Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with s}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
 
-\leavevmode\hypertarget{ref-tsibble}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tsibble}{}}%
 Wang, Earo, Dianne Cook, and Rob J Hyndman.
 2020.
-``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
+{``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.''} \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
 \url{https://doi.org/10.1080/10618600.2019.1695624}.
 
-\leavevmode\hypertarget{ref-plyr}{}%
-Wickham, Hadley. 2011. ``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
+\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
+Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
 
-\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
----------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.
+\leavevmode\vadjust pre{\hypertarget{ref-WickhamHadley2014TD}{}}%
+---------. 2014. {``Tidy Data.''} \emph{Journal of Statistical Software} 59 (10): 1--23.
 
-\leavevmode\hypertarget{ref-ggplot2}{}%
----------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
+\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
+---------. 2016. \emph{{ggplot2: Elegant Graphics for Data Analysis}}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
 
-\leavevmode\hypertarget{ref-stringr}{}%
----------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.
+\leavevmode\vadjust pre{\hypertarget{ref-stringr}{}}%
+---------. 2019. \emph{{stringr: Simple, Consistent Wrappers for Common String Operations}}. \url{https://CRAN.R-project.org/package=stringr}.
 
-\leavevmode\hypertarget{ref-tidyr}{}%
----------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.
+\leavevmode\vadjust pre{\hypertarget{ref-tidyr}{}}%
+---------. 2020. \emph{{tidyr: Tidy Messy Data}}. \url{https://CRAN.R-project.org/package=tidyr}.
 
-\leavevmode\hypertarget{ref-tidyverse}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al.
 2019.
-``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686.
+{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686.
 \url{https://doi.org/10.21105/joss.01686}.
 
-\leavevmode\hypertarget{ref-dplyr}{}%
-Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.
+\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
+Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{{dplyr: A Grammar of Data Manipulation}}. \url{https://CRAN.R-project.org/package=dplyr}.
 
-\leavevmode\hypertarget{ref-readr}{}%
-Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.
+\leavevmode\vadjust pre{\hypertarget{ref-readr}{}}%
+Wickham, Hadley, and Jim Hester. 2020. \emph{{readr: Read Rectangular Text Data}}. \url{https://CRAN.R-project.org/package=readr}.
 
-\leavevmode\hypertarget{ref-knitr}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-knitr}{}}%
 Xie, Yihui.
 2014.
-``Knitr: A Comprehensive Tool for Reproducible Research in R.'' In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
+{``Knitr: A Comprehensive Tool for Reproducible Research in {R}.''} In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
 Peng.

### 2022-01-23 paper/paper.Rmd
intro update

@@
 An example of this the wages data made public by @SingerJudithD2003Alda which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
-The freely available data contains records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
+The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
 
 This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook [@ilk2004], and was developed into a case study for use in teaching exploratory data analysis at Iowa State University.
-The story from modeling the data (and as reported by Singer and Willett) is that wages increase with length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
-The story from an exploratory analysis is very different: the individual experience varies a lot from the overall average.
-Some individuals experience a decline in wages the longer they are in the workforce, and most experience volatility in their wages.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+An exploratory analysis reveals however that the individual experience varies a great deal from the overall average.
+Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
 
 This disparity between the average and the individual is a part of statistics that is given less attention.
-It is this disparity which makes this particular data set very important as a textbook example.
-Textbook data sets have staying power if it has a challenge that has never been completely solved.
-The iris data (XXX REF) is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(A side note: the iris data is best replaced today with the penguins data (XXX REF), which has similar qualities, is new and does not suffer from a connection with eugenics [@notiris].) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
+It is this disparity which makes this particular data set important as a textbook example.
+Textbook data sets have longevity if there is a unresolved mystery.
+The iris data [@iris-data] is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
+(A side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new and does not suffer from a connection with eugenics [@notiris].) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
 
@@
 
-As a textbook data though, the dates are tired.
+As a textbook data though, the original data is outdated.
 The most recent year in the data is 1994, 10 years prior to when @SingerJudithD2003Alda was published.
@@
 It should be possible to continuously refresh the textbook data from the data repository.
-This is our story of the journey from open [@opendata] but wild data to textbook data, it's not entirely pretty.
+This paper describes our (non-glamorous) journey from open [@opendata] but wild data to textbook data.
 

### 2022-01-23 paper/paper.tex
updated compiled paper

@@
 
-% tightlist command for lists without linebreak
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-% From pandoc table feature
 \usepackage{longtable,booktabs,array}
@@
 % Pandoc citation processing
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
 \newlength{\cslhangindent}
 \setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
 % for Pandoc 2.8 to 2.10.1
@@
   % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
   \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
  }%
  {}
-\usepackage{calc}
+\usepackage{calc} % for calculating minipage widths
 \newcommand{\CSLBlock}[1]{#1\hfill\break}
@@
 \usepackage{xcolor}
+
 \begin{document}
 \maketitle
+
+\def\tightlist{}
 
@@
 An example of this the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
-The freely available data contains records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
+The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
 Department of Labor 2021).
@@
 This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was developed into a case study for use in teaching exploratory data analysis at Iowa State University.
-The story from modeling the data (and as reported by Singer and Willett) is that wages increase with length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
-The story from an exploratory analysis is very different: the individual experience varies a lot from the overall average.
-Some individuals experience a decline in wages the longer they are in the workforce, and most experience volatility in their wages.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+An exploratory analysis reveals however that the individual experience varies a great deal from the overall average.
+Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
 
 This disparity between the average and the individual is a part of statistics that is given less attention.
-It is this disparity which makes this particular data set very important as a textbook example.
-Textbook data sets have staying power if it has a challenge that has never been completely solved.
-The iris data (XXX REF) is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(A side note: the iris data is best replaced today with the penguins data (XXX REF), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
+It is this disparity which makes this particular data set important as a textbook example.
+Textbook data sets have longevity if there is a unresolved mystery.
+The iris data (Andereson 1935) is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
+(A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
 
@@
 
-As a textbook data though, the dates are tired.
+As a textbook data though, the original data is outdated.
 The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
@@
 It should be possible to continuously refresh the textbook data from the data repository.
-This is our story of the journey from open (Open Knowledge Foundation 2021) but wild data to textbook data, it's not entirely pretty.
+This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) but wild data to textbook data.
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/cookd/numbats/yowie-paper/paper/figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{/Users/etan0038/Dropbox/repo/yowie-paper/paper/figures/flowchart_blind}
 
@@
 \begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-iris-data}{}}%
+Andereson, Edgar. 1935. {``The Irises of the Gaspe Peninsula.''} \emph{Bulletin of the American Iris Society} 59: 2--5.
+
 \leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
@@
 Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-penguins-data}{}}%
+Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
 

### 2022-01-25 paper/paper.Rmd
Edited several sentences.

@@
 Thus, it might be challenging for the respondents to track the within-job hours' changes between survey years, especially for the respondents with fluctuating working hours or seasonal jobs.
-It even has been more challenging since 1994, where the respondents had to recall two year periods.
+It even has been more challenging since 1994, after which respondents were only surveyed every other year and thus had to recall two full years’ job history.
 This shortcoming might also contribute to the fluctuation of one's wages data.
@@
 It also aims to detect the contamination by weighting each observation based on how "well-behaved" they are, known as robustness weight.
-Observations with lower robustness weight are suggested as an outlier by this method [@KollerManuel2016rARP].
+An observation with a lower robustness weight would be suggested as an outlier [@KollerManuel2016rARP].
 

### 2022-01-28 paper/paper.Rmd
intro still needs a smoother rtansition from new paragraphs to the older paragraphs

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this the wages data made public by @SingerJudithD2003Alda which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
+An example of this, is the wages data made public by @SingerJudithD2003Alda which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
 The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
@@
 Textbook data sets have longevity if there is a unresolved mystery.
-The iris data [@iris-data] is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
+The iris data [@iris-data] is a prime example.
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
 (A side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new and does not suffer from a connection with eugenics [@notiris].) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
@@
 
-As a textbook data though, the original data is outdated.
+As textbook data though, the original data is outdated.
 The most recent year in the data is 1994, 10 years prior to when @SingerJudithD2003Alda was published.

### 2022-01-28 paper/paper.tex
intro still needs a smoother rtansition from new paragraphs to the older paragraphs

@@
 
+% tightlist command for lists without linebreak
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+
+% From pandoc table feature
 \usepackage{longtable,booktabs,array}
@@
 % Pandoc citation processing
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
 \newlength{\csllabelwidth}
 \setlength{\csllabelwidth}{3em}
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
 % for Pandoc 2.8 to 2.10.1
@@
   % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
   % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
+  \setlength{\parskip}{#2\cslentryspacingunit}
  }%
  {}
-\usepackage{calc} % for calculating minipage widths
+\usepackage{calc}
 \newcommand{\CSLBlock}[1]{#1\hfill\break}
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
+An example of this, is the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
 The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 Textbook data sets have longevity if there is a unresolved mystery.
-The iris data (Andereson 1935) is a prime example, it has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
+The iris data (Andereson 1935) is a prime example.
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
 (A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
@@
 
-As a textbook data though, the original data is outdated.
+As textbook data though, the original data is outdated.
 The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
@@
 Thus, it might be challenging for the respondents to track the within-job hours' changes between survey years, especially for the respondents with fluctuating working hours or seasonal jobs.
-It even has been more challenging since 1994, where the respondents had to recall two year periods.
+It even has been more challenging since 1994, after which respondents were only surveyed every other year and thus had to recall two full years' job history.
 This shortcoming might also contribute to the fluctuation of one's wages data.
@@
 It also aims to detect the contamination by weighting each observation based on how ``well-behaved'' they are, known as robustness weight.
-Observations with lower robustness weight are suggested as an outlier by this method (Koller 2016).
+An observation with a lower robustness weight would be suggested as an outlier (Koller 2016).
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/etan0038/Dropbox/repo/yowie-paper/paper/figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{/Users/cookd/numbats/yowie-paper/paper/figures/flowchart_blind}
 

### 2022-01-28 paper/paper.Rmd
change email address for Dewi to non-monash

@@
     location: Clayton, VIC 3800
-    email: dama0007@student.monash.edu
+    email: dlamaleeah@gmail.com
   - name: Dianne Cook

### 2022-01-29 data-raw/data-preprocessing.R
Added start work and work experience variable.

@@ -80,8 +80,8 @@ demog_education <- new_data_qnames %>%
 # Get the highest year completed
 highest_year <- demog_education %>%
   group_by(id) %>%
-  filter(grade == max(grade),
-         year == min(year)) %>%
+  filter(grade == max(grade)) %>%
+  filter(year == min(year)) %>%
   rename(yr_hgc = year,
          hgc_i = grade) %>%
   select(id, yr_hgc, hgc_i) %>%
@@ -163,6 +163,29 @@ get_rate <- function(year) {
 
 rates_all <- map_dfr(years, get_rate)
 
+## ---- tidy-start-work
+stwork <- new_data_qnames %>%
+  select(CASEID_1979,
+         `EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND`) %>%
+  rename(id = CASEID_1979,
+         stwork_year = `EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND`)
+
+## ---- tidy-work-experience
+
+exp <- new_data_qnames %>%
+  select(CASEID_1979,
+         starts_with("WKSWK")) %>%
+  pivot_longer(!CASEID_1979, names_to = "exp", values_to = "weeks") %>%
+  separate("exp", c("exp", "year"), sep = -4) %>%
+  group_by(CASEID_1979) %>%
+  mutate(cum_exp = cumsum(coalesce(weeks, 0)) + weeks*0,
+         exp_years = cum_exp/52.143,
+         year = as.numeric(year)) %>%
+  select(CASEID_1979, year, exp_years) %>%
+  rename(id = CASEID_1979,
+         exp = exp_years)
+
+
 ## ---- tidy-rate-hour
 
 # join hours and rates variable
@@ -215,11 +238,14 @@ head(mean_hourly_wage, n = 10)
 
 ## ---- wages-demog-hs
 # join the wages information and the demographic information by case id.
-wages_demog <- left_join(mean_hourly_wage, full_demographics, by="id")
+wages_demog <- left_join(mean_hourly_wage, full_demographics, by="id") %>%
+  left_join(stwork, by = "id") %>%
+  left_join(exp, by = c("id", "year"))
+
 # calculate the years in work force and the age of the subjects in 1979
 wages_demog <- wages_demog %>%
   mutate(yr_hgc = as.numeric(yr_hgc)) %>%
-  mutate(years_in_workforce = year - yr_hgc) %>%
[…]

### 2022-01-31 paper/paper.Rmd
added email addresses for Kate and Nick and changed license

@@
     location: Clayton, VIC 3800
-    email: ""
+    email: "hyde.kate.a@gmail.com"
   - name: Nicholas Tierney
@@
     location: Nedlands, WA 6009
-    email: ""
+    email: "nicholas.tierney@gmail.com"
 bibliography: references.bib
@@
 -->
+
+```{r tidy-rate}
+```
+
+```{r tidy-start-work}
+```
+
+```{r tidy-work-experience}
+```
 

### 2022-02-02 paper/paper-jsdse/paper-jsdse.tex
Push the latest change that could be knitted.

--- 2022-01-13 paper/paper-jsdse/paper-jsdse.tex
@@
 
-``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021).
-This type of data can be useful as example data in statistical textbooks and for research purposes.
-However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape.
-Huebner, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (Huebner et al.
-2020).
-Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985).
-In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
-This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
-EDA can be considered to encompass IDA.
-Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.
+Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
+An example of this, is the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
+The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
+Department of Labor 2021).
 
-Despite it's importance, this IDA stage is often undervalued and neglected (Chatfield 1985).
-There are few research papers that document the data cleaning (Wickham 2014).
-Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (Huebner et al.
-2020).
+This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was developed into a case study for use in teaching exploratory data analysis at Iowa State University.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+An exploratory analysis reveals however that the individual experience varies a great deal from the overall average.
+Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
 
-This paper demonstrates the steps of IDA, and documents the process, for a prominent open data source, National Longitudinal Survey of Youth 1979, henceforth referred to NLSY79.
+This disparity between the average and the individual is a part of statistics that is given less attention.
+It is this disparity which makes this particular data set important as a textbook example.
+Textbook data sets have longevity if there is a unresolved mystery.
+The iris data (Andereson 1935) is a prime example.
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
+(A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
+
+For the field of statistics, and data science by association, it is increasingly important to reach the individual.
+One might describe this as a divergence of purpose, statistics for public policy or statistics for the public.
+The two are not the same.
+As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual.
+For example, with the wages data, the message for the individual is that you are more than your demographic.
+The majority of people tracked are not like the average, the average is an anomaly.
+If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
+
+As textbook data though, the original data is outdated.
+The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
+Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students.
+Using tired old textbook data sets sets up a misconception that the field is not current.
+The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated.
+It should be possible to continuously refresh the textbook data from the data repository.
+This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) but wild data to textbook data.
+
+Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
 This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al.
@@
 2001).
+Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
+Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).
+We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the \emph{purposes of studying racism, not race} (Fullilove 1998).
 
-Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
-Our aim is to refresh this textbook data with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
-Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).
-We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the purposes of studying racism, not race (Fullilove 1998).
+This paper demonstrates the steps of cleaning data, and documents the process, as recommended by Huebner, Vach, and Cessie (2016).
+They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
+Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985).
+In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
+This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
+EDA can be considered to encompass IDA.
+Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.
 
@@
 \begin{verbatim}
-#> 'data.frame':    12686 obs. of  686 variables:
+#> 'data.frame':    12686 obs. of  716 variables:
 #>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 9 10 ...
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie/paper/figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart_blind}
 
@@
 \hypertarget{refs}{}
+\leavevmode\hypertarget{ref-iris-data}{}%
+Andereson, Edgar. 1935. ``The Irises of the Gaspe Peninsula.'' \emph{Bulletin of the American Iris Society} 59: 2--5.
+
 \leavevmode\hypertarget{ref-nlsy79}{}%
@@
 
+\leavevmode\hypertarget{ref-penguins-data}{}%
+Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
+
 \leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
@@
 
-\leavevmode\hypertarget{ref-HuebnerMarianne2020Haar}{}%
-Huebner, Marianne, Werner Vach, Saskia le Cessie, Carsten Oliver Schmidt, and Lara Lusa.
-2020.
-``Hidden Analyses: A Review of Reporting Practice and Recommendations for More Transparent Reporting of Initial Data Analyses.'' \emph{BMC Medical Research Methodology} 20 (1): 61--61.
+\leavevmode\hypertarget{ref-ilk2004}{}%
+Ilk, Ozlem.
+2004.
+``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.'' PhD thesis, Iowa State University.
+\url{https://doi.org/10.31274/rtd-180813-11012}.
 
@@
 Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.
+
+\leavevmode\hypertarget{ref-notiris}{}%
+Stodel, Megan. 2020. ``Stop Using Iris.'' \url{https://www.meganstodel.com/posts/no-to-iris/}.
 

### 2022-02-02 paper/paper.tex
change email address for Dewi to non-monash
Push the latest change that could be knitted.

@@
 \usepackage{microtype}      % microtypography
+\usepackage{lipsum}
 \usepackage{graphicx}
@@
   Clayton, VIC 3800 \\
-  \texttt{\href{mailto:dama0007@student.monash.edu}{\nolinkurl{dama0007@student.monash.edu}}} \\
+  \texttt{\href{mailto:dlamaleeah@gmail.com}{\nolinkurl{dlamaleeah@gmail.com}}} \\
    \And
@@
   Clayton, VIC 3800 \\
-  \texttt{} \\
+  \texttt{\href{mailto:hyde.kate.a@gmail.com}{\nolinkurl{hyde.kate.a@gmail.com}}} \\
    \And
@@
   Nedlands, WA 6009 \\
-  \texttt{} \\
+  \texttt{\href{mailto:nicholas.tierney@gmail.com}{\nolinkurl{nicholas.tierney@gmail.com}}} \\
   }
 
-% tightlist command for lists without linebreak
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-% From pandoc table feature
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-
 % Pandoc citation processing
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
-Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al.
+Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
 2019).
@@
 
-M.
 van der Loo and de Jonge (2018) describe the notion of a ``statistical value chain'' where the production stages of the data cleaning process are earmarked as raw data (data as arrived to the desk of an analyst), input data (data organised with correct type and identified variables) and valid data (data that faithfully represent the variables).
@@
 The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
-This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
-2019), \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham et al.
+This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
+2019), \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham, François, et al.
 2020), and \texttt{stringr} (Wickham 2019) for creating new variables, and levels of factors by text wrangling.
@@
 
-Figure \ref{fig:flow-chart-blind} summarizes the steps taken to go from raw to input to valid data (M. van der Loo and de Jonge 2018) to create a refreshed wages data set.
+Figure \ref{fig:flow-chart-blind} summarizes the steps taken to go from raw to input to valid data (van der Loo and de Jonge 2018) to create a refreshed wages data set.
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/cookd/numbats/yowie-paper/paper/figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart_blind}
 
@@
 ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data.
-The color represents the stage of data cleaning in statistical value chain (M.
-P.
-J.
-van der Loo and de Jonge 2021).
+The color represents the stage of data cleaning in statistical value chain (van der Loo and de Jonge 2021).
 Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart-blind}
@@
 
-The entire analysis is conducted using \texttt{R} (R Core Team 2020) in RStudio IDE using these packages: \texttt{tidyverse} (Wickham et al.
-2019), \texttt{ggplot2} (Wickham 2016), \texttt{dplyr} (Wickham et al.
+The entire analysis is conducted using \texttt{R} (R Core Team 2020) in RStudio IDE using these packages: \texttt{tidyverse} (Wickham, Averick, et al.
+2019), \texttt{ggplot2} (Wickham 2016), \texttt{dplyr} (Wickham, François, et al.
 2020), \texttt{readr} (Wickham and Hester 2020), \texttt{tidyr} (Wickham 2020), \texttt{stringr} (Wickham 2019), \texttt{purrr} (Henry and Wickham 2020), \texttt{brolgar} (Tierney, Cook, and Prvan 2020), \texttt{patchwork} (Pedersen 2020), \texttt{kableExtra} (Zhu 2019), \texttt{MASS} (Venables and Ripley 2002), \texttt{janitor} (Firke 2020), and \texttt{tsibble} (Wang, Cook, and Hyndman 2020).
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-iris-data}{}}%
-Andereson, Edgar. 1935. {``The Irises of the Gaspe Peninsula.''} \emph{Bulletin of the American Iris Society} 59: 2--5.
+\leavevmode\hypertarget{ref-iris-data}{}%
+Andereson, Edgar. 1935. ``The Irises of the Gaspe Peninsula.'' \emph{Bulletin of the American Iris Society} 59: 2--5.
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
+\leavevmode\hypertarget{ref-nlsy79}{}%
 Bureau of Labor Statistics, U.S.
@@
 2021.
-{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
+``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
 Columbus, OH.
 
-\leavevmode\vadjust pre{\hypertarget{ref-shiny}{}}%
-Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{{shiny: Web Application Framework for R}}. \url{https://CRAN.R-project.org/package=shiny}.
+\leavevmode\hypertarget{ref-shiny}{}%
+Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Chatfield1985TIEo}{}}%
-Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
+\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
+Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
 
-\leavevmode\vadjust pre{\hypertarget{ref-eliznlsy}{}}%
+\leavevmode\hypertarget{ref-eliznlsy}{}%
 Cooksey, Elizabeth C.
 2017.
-{``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M.
+``Using the National Longitudinal Surveys of Youth (Nlsy) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
 Lerner Neal Halfon Christoper B.
@@
 Cham: Springer.
-https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.
+\url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-DasuTamraparni2003Edma}{}}%
+\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
 Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.
 
-\leavevmode\vadjust pre{\hypertarget{ref-janitor}{}}%
-Firke, Sam. 2020. \emph{{janitor: Simple Tools for Examining and Cleaning Dirty Data}}. \url{https://CRAN.R-project.org/package=janitor}.
+\leavevmode\hypertarget{ref-janitor}{}%
+Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-racismnotrace}{}}%
-Fullilove, M. T. 1998. {``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.
+\leavevmode\hypertarget{ref-racismnotrace}{}%
+Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.
 
-\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
+\leavevmode\hypertarget{ref-grimshaw}{}%
 Grimshaw, Scott D.
 2015.
-{``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14.
+``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14.
 \url{https://doi.org/10.1080/00031305.2015.1081106}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-purrr}{}}%
-Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.
+\leavevmode\hypertarget{ref-purrr}{}%
+Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-penguins-data}{}}%
+\leavevmode\hypertarget{ref-penguins-data}{}%
 Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
+\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
 Huebner, Marianne, Werner Vach, and Saskia le Cessie.
 2016.
-{``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
+``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
 
-\leavevmode\vadjust pre{\hypertarget{ref-ilk2004}{}}%
+\leavevmode\hypertarget{ref-ilk2004}{}%
 Ilk, Ozlem.
 2004.
-{``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.''} PhD thesis, Iowa State University.
+``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.'' PhD thesis, Iowa State University.
 \url{https://doi.org/10.31274/rtd-180813-11012}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
+\leavevmode\hypertarget{ref-tamedata}{}%
 Kim, A.
@@
 2018.
-{``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1).
+``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1).
 \url{https://doi.org/10.5070/T511103589}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-KollerManuel2016rARP}{}}%
-Koller, Manuel. 2016. {``{robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models}.''} \emph{Journal of Statistical Software} 75 (6): 1--24.
+\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
+Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.
 
-\leavevmode\vadjust pre{\hypertarget{ref-opendata}{}}%
-Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.
+\leavevmode\hypertarget{ref-opendata}{}%
+Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-patchwork}{}}%
-Pedersen, Thomas Lin. 2020. \emph{{patchwork: The Composer of Plots}}. \url{https://CRAN.R-project.org/package=patchwork}.
+\leavevmode\hypertarget{ref-patchwork}{}%
+Pedersen, Thomas Lin. 2020. \emph{patchwork: The Composer of Plots}. \url{https://CRAN.R-project.org/package=patchwork}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-MichaelRPergamit2001DWTN}{}}%
+\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
 Pergamit, Michael R., Charles R.
@@
 2001.
-{``Data Watch: The National Longitudinal Surveys.''} \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
+``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
 
-\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
+\leavevmode\hypertarget{ref-R}{}%
 R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-SingerJudithD2003Alda}{}}%
+\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
 Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.
 
-\leavevmode\vadjust pre{\hypertarget{ref-notiris}{}}%
-Stodel, Megan. 2020. {``Stop Using Iris.''} \url{https://www.meganstodel.com/posts/no-to-iris/}.
+\leavevmode\hypertarget{ref-notiris}{}%
+Stodel, Megan. 2020. ``Stop Using Iris.'' \url{https://www.meganstodel.com/posts/no-to-iris/}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-brolgar}{}}%
-Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}}. \url{https://github.com/njtierney/brolgar}.
+\leavevmode\hypertarget{ref-brolgar}{}%
+Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-tukey}{}}%
+\leavevmode\hypertarget{ref-tukey}{}%
 Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.
 
-\leavevmode\vadjust pre{\hypertarget{ref-rlm}{}}%
-UCLA: Statistical Consulting Group. 2021. {``{Robust Regression \textbar{} R Data Analysis Examples}.''} February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
+\leavevmode\hypertarget{ref-rlm}{}%
+UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%
-van der Loo, Mark P. J., and Edwin de Jonge. 2021. {``Data Validation Infrastructure for {R}.''} \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
+\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
+van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-LooMarkvander2018Sdcw}{}}%
-van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in r}.
+\leavevmode\hypertarget{ref-validate}{}%
+van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-mass}{}}%
-Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with s}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
+\leavevmode\hypertarget{ref-mass}{}%
+Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-tsibble}{}}%
+\leavevmode\hypertarget{ref-tsibble}{}%
 Wang, Earo, Dianne Cook, and Rob J Hyndman.
 2020.
-{``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.''} \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
+``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
 \url{https://doi.org/10.1080/10618600.2019.1695624}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
-Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
+\leavevmode\hypertarget{ref-plyr}{}%
+Wickham, Hadley. 2011. ``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-WickhamHadley2014TD}{}}%
----------. 2014. {``Tidy Data.''} \emph{Journal of Statistical Software} 59 (10): 1--23.
+\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
+---------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.
 
-\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
----------. 2016. \emph{{ggplot2: Elegant Graphics for Data Analysis}}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
+\leavevmode\hypertarget{ref-ggplot2}{}%
+---------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-stringr}{}}%
----------. 2019. \emph{{stringr: Simple, Consistent Wrappers for Common String Operations}}. \url{https://CRAN.R-project.org/package=stringr}.
+\leavevmode\hypertarget{ref-stringr}{}%
+---------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-tidyr}{}}%
----------. 2020. \emph{{tidyr: Tidy Messy Data}}. \url{https://CRAN.R-project.org/package=tidyr}.
+\leavevmode\hypertarget{ref-tidyr}{}%
+---------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
+\leavevmode\hypertarget{ref-tidyverse}{}%
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al.
 2019.
-{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686.
+``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686.
 \url{https://doi.org/10.21105/joss.01686}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
-Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{{dplyr: A Grammar of Data Manipulation}}. \url{https://CRAN.R-project.org/package=dplyr}.
+\leavevmode\hypertarget{ref-dplyr}{}%
+Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-readr}{}}%
-Wickham, Hadley, and Jim Hester. 2020. \emph{{readr: Read Rectangular Text Data}}. \url{https://CRAN.R-project.org/package=readr}.
+\leavevmode\hypertarget{ref-readr}{}%
+Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-knitr}{}}%
+\leavevmode\hypertarget{ref-knitr}{}%
 Xie, Yihui.
 2014.
-{``Knitr: A Comprehensive Tool for Reproducible Research in {R}.''} In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
+``Knitr: A Comprehensive Tool for Reproducible Research in R.'' In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
 Peng.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rmarkdown}{}}%
+\leavevmode\hypertarget{ref-rmarkdown}{}%
 Xie, Yihui, Christophe Dervieux, and Emily Riederer. 2020. \emph{R Markdown Cookbook}. Boca Raton, Florida: Chapman; Hall/CRC. \url{https://bookdown.org/yihui/rmarkdown-cookbook}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-kableExtra}{}}%
-Zhu, Hao. 2019. \emph{{kableExtra: Construct Complex Table with 'kable' and Pipe Syntax}}. \url{https://CRAN.R-project.org/package=kableExtra}.
-
-\end{CSLReferences}
+\leavevmode\hypertarget{ref-kableExtra}{}%
+Zhu, Hao. 2019. \emph{kableExtra: Construct Complex Table with 'kable' and Pipe Syntax}. \url{https://CRAN.R-project.org/package=kableExtra}.
 

### 2022-02-02 data-raw/NLSY79/NLSY79.R

@@ -3,7 +3,7 @@
 # setwd()
 
 
-new_data <- read.table('data-raw/NLSY79/NLSY79.dat', sep=' ')
+new_data <- read.table(here::here('data-raw/NLSY79/NLSY79.dat'), sep=' ')
 names(new_data) <- c('E0010101',
   'E0010102',
   'R0000100',

### 2022-02-02 paper/paper.Rmd
Edited Figure 1.

@@
 
-The NLSY79 data contains a large number of variables but for our aim, the scope required is limited to demographic profiles and wages data.
+The NLSY79 data contains a large number of variables but for our aim, the scope required is limited to demographic profiles, wages data, and work experience.
 More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data and navigated as described in Figure \ref{fig:source-nav}.
@@
 \begin{itemize}
-\item[\faCheck] All 442 primary variables in Hours worked were selected.
+\item[\faCheck] All 447 primary variables in Hours worked were selected.
 \end{itemize}
@@
 \end{itemize}
+\end{itemize}
+\end{itemize}
+\begin{itemize}
+\item[$\triangleright$] Summary measures $\triangleright$ Since date of last interview $\triangleright$ Weeks worked
+\begin{itemize}
+\item[\faCheck] All 28 variables in Weeks worked were selected.
+\end{itemize}
+\end{itemize}
+\begin{itemize}
+\item[$\triangleright$] Employer Roster $\triangleright$ Job dates $\triangleright$ Original start date
+\begin{itemize}
+\item[\faCheck] Only selected the start date (Year) for the first job (E00101.02)
 \end{itemize}

### 2022-02-02 data-raw/data-preprocessing.R
Push the latest update.

@@ -118,7 +118,7 @@ demog_nlsy79 <- full_demographics %>%
 
 
 ## ---- save-demog-data
-usethis::use_data(demog_nlsy79, overwrite = TRUE)
+#usethis::use_data(demog_nlsy79, overwrite = TRUE)
 
 ## ---- tidy-hours
 # make a list for years where we used the "QES-52A"

### 2022-02-03 paper/paper.Rmd
removed some identifying information and added censorship for blinding

@@
 yowie_shiny <- ifelse(CENSOR, "CENSORED", "https://ebsmonash.shinyapps.io/yowie_app/")
+censor_word <- function(word) ifelse(CENSOR, "CENSORED", word)
 ```
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by @SingerJudithD2003Alda which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
+An example of this, is the wages data made public by @SingerJudithD2003Alda which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel models.
 The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
 
-This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook [@ilk2004], and was developed into a case study for use in teaching exploratory data analysis at Iowa State University.
+This textbook data was used as an example in @ilk2004, and was developed into a case study for use in teaching exploratory data analysis at `r censor_word("Iowa State University")`.
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.

### 2022-02-03 paper/paper.tex
removed some identifying information and added censorship for blinding

--- 2022-01-28 paper/paper.tex
@@
   Nedlands, WA 6009 \\
-  \texttt{} \\
+  \texttt{\href{mailto:nicholas.tierney@gmail.com}{\nolinkurl{nicholas.tierney@gmail.com}}} \\
   }
 
-% tightlist command for lists without linebreak
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-% From pandoc table feature
 \usepackage{longtable,booktabs,array}
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
+An example of this, is the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel models.
 The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 
-This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was developed into a case study for use in teaching exploratory data analysis at Iowa State University.
+This textbook data was used as an example in Ilk (2004), and was developed into a case study for use in teaching exploratory data analysis at CENSORED.
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.

### 2022-02-03 paper/paper-jsdse/paper-jsdse.Rmd
still tweaking intro

@@
     location: Nedlands, WA 6009
-    email: ""
+    email: "nicholas.tierney@gmail.com"
 blind: 1 # 0 for no

### 2022-02-03 paper/paper-jsdse/paper-jsdse.tex
still tweaking intro

--- 2022-01-14 paper/paper-jsdse/paper-jsdse.tex
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
-The freely available data contains records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79).
+An example of this, is the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
+The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
+Department of Labor 2021).
 
 This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was developed into a case study for use in teaching exploratory data analysis at Iowa State University.
-The story from modeling the data (and as reported by Singer and Willett) is that wages increase with length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+An exploratory analysis reveals however that the individual experience varies a great deal from the overall average.
+Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
 
-``Open data'' is data that is freely accessible, modifiable, and shareable by anyone for any purpose (Open Knowledge Foundation 2021).
-This type of data can be useful as example data in statistical textbooks and for research purposes.
-However, open data is often referred to as what we might call ``wild data'' because it requires substantial cleaning and tidying to tame it into textbook shape.
-Huebner Marianne, Vach, and Cessie (2016) emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building (M.
-Huebner et al.
-2020).
-Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985).
-In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
-This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
-EDA can be considered to encompass IDA.
-Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.
+This disparity between the average and the individual is a part of statistics that is given less attention.
+It is this disparity which makes this particular data set important as a textbook example.
+Textbook data sets have longevity if there is a unresolved mystery.
+The iris data (Andereson 1935) is a prime example.
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
+(A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
 
-Despite it's importance, this IDA stage is often undervalued and neglected (Chatfield 1985).
-There are few research papers that document the data cleaning (Wickham 2014).
-Furthermore, the decisions made in this stage often go unreported in the sense that IDA is often performed in an unplanned and unstructured way and is only shared among restricted parties (M.
-Huebner et al.
-2020).
+For the field of statistics, and data science by association, it is increasingly important to reach the individual.
+One might describe this as a divergence of purpose, statistics for public policy or statistics for the public.
+The two are not the same.
+As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual.
+For example, with the wages data, the message for the individual is that you are more than your demographic.
+The majority of people tracked are not like the average, the average is an anomaly.
+If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
 
-This paper demonstrates the steps of IDA, and documents the process, for a prominent open data source, National Longitudinal Survey of Youth 1979, henceforth referred to NLSY79.
+As textbook data though, the original data is outdated.
+The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
+Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students.
+Using tired old textbook data sets sets up a misconception that the field is not current.
+The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated.
+It should be possible to continuously refresh the textbook data from the data repository.
+This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) but wild data to textbook data.
+
+Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
 This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al.
@@
 
-According to Huebner Marianne, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis.
+According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis.
 Moreover, Chatfield (1985) argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-penguins-data}{}}%
+Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
@@
 {``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
-
-\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMarianne2020Haar}{}}%
-Huebner, Marianne, Werner Vach, Saskia le Cessie, Carsten Oliver Schmidt, and Lara Lusa.
-2020.
-{``Hidden Analyses: A Review of Reporting Practice and Recommendations for More Transparent Reporting of Initial Data Analyses.''} \emph{BMC Medical Research Methodology} 20 (1): 61--61.
 
@@
 Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.
+
+\leavevmode\vadjust pre{\hypertarget{ref-notiris}{}}%
+Stodel, Megan. 2020. {``Stop Using Iris.''} \url{https://www.meganstodel.com/posts/no-to-iris/}.
 

### 2022-02-03 paper/paper.Rmd
still tweaking intro

--- 2022-02-02 paper/paper.Rmd
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by @SingerJudithD2003Alda which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
+An example of this, is the wages data made public by @SingerJudithD2003Alda which can be used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
 The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
 
-This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook [@ilk2004], and was developed into a case study for use in teaching exploratory data analysis at Iowa State University.
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
+The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis supervised by author Cook [@ilk2004], and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
 
-This disparity between the average and the individual is a part of statistics that is given less attention.
+This disparity between the average and the individual is a part of statistics, as a discipline, that is given less attention.
 It is this disparity which makes this particular data set important as a textbook example.
@@
 It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(A side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new and does not suffer from a connection with eugenics [@notiris].) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
+(A side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new and does not suffer from a connection with eugenics [@notiris].) We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
 
@@
 As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual.
-For example, with the wages data, the message for the individual is that you are more than your demographic.
-The majority of people tracked are not like the average, the average is an anomaly.
+For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic.
+The majority of people in the study do not have a pattern that is similar to the average.
 If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
 
-As textbook data though, the original data is outdated.
+As a textbook data set though, it is outdated.
 The most recent year in the data is 1994, 10 years prior to when @SingerJudithD2003Alda was published.
 Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students.
-Using tired old textbook data sets sets up a misconception that the field is not current.
+Using tired old textbook data sets can imbue a misconception that the field is not current.
 The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated.
 It should be possible to continuously refresh the textbook data from the data repository.
-This paper describes our (non-glamorous) journey from open [@opendata] but wild data to textbook data.
-
-@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
-This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN].
-In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research [@MichaelRPergamit2001DWTN; @eliznlsy].
-According to @eliznlsy, thousands of articles, and hundreds of book chapters and monographs have utilized this data.
-Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data [@MichaelRPergamit2001DWTN].
-<!--AE comment: For example, I would like more discussion about the textbook selected as the exemplar.
-It sounds like this dataset is used in many books, why choose that one?
-The text should include the name of the book, so readers do not need to flip to the citations in order to learn what it was.
-From the name, I am guessing that the book focuses heavily on this dataset, which is why it makes sense to reproduce their subset.
-Be explicit about this.-->
-Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences.
-Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda.
-We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the *purposes of studying racism, not race* [@racismnotrace].
-<!--However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.-->
+This paper describes our (non-glamorous) journey from open [@opendata] wild data to textbook data.
 
@@
 # The NLSY79 {#database}
+
+@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
+This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN].
+In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research [@MichaelRPergamit2001DWTN; @eliznlsy].
+According to @eliznlsy, thousands of articles, and hundreds of book chapters and monographs have utilized this data.
+Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data [@MichaelRPergamit2001DWTN].
+<!--AE comment: For example, I would like more discussion about the textbook selected as the exemplar.
+It sounds like this dataset is used in many books, why choose that one?
+The text should include the name of the book, so readers do not need to flip to the citations in order to learn what it was.
+From the name, I am guessing that the book focuses heavily on this dataset, which is why it makes sense to reproduce their subset.
+Be explicit about this.-->
+Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences.
+Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda.
+We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the *purposes of studying racism, not race* [@racismnotrace].
+<!--However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.-->
 

### 2022-02-03 paper/paper.tex
still tweaking intro

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel models.
+An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
 The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 
-This textbook data was used as an example in Ilk (2004), and was developed into a case study for use in teaching exploratory data analysis at CENSORED.
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
+The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
 
-This disparity between the average and the individual is a part of statistics that is given less attention.
+This disparity between the average and the individual is a part of statistics, as a discipline, that is given less attention.
 It is this disparity which makes this particular data set important as a textbook example.
@@
 It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) The wages data is in this class of textbook data, because it presents this challenge for exploratory data analysis: how can we better summarize and explain the individual experience?
+(A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
 
@@
 As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual.
-For example, with the wages data, the message for the individual is that you are more than your demographic.
-The majority of people tracked are not like the average, the average is an anomaly.
+For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic.
+The majority of people in the study do not have a pattern that is similar to the average.
 If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
 
-As textbook data though, the original data is outdated.
+As a textbook data set though, it is outdated.
 The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
 Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students.
-Using tired old textbook data sets sets up a misconception that the field is not current.
+Using tired old textbook data sets can imbue a misconception that the field is not current.
 The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated.
 It should be possible to continuously refresh the textbook data from the data repository.
-This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) but wild data to textbook data.
-
-Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
-This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al.
-2001).
-In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research (Pergamit et al.
-2001; Cooksey 2017).
-According to Cooksey (2017), thousands of articles, and hundreds of book chapters and monographs have utilized this data.
-Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al.
-2001).
-Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
-Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).
-We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the \emph{purposes of studying racism, not race} (Fullilove 1998).
+This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) wild data to textbook data.
 
@@
 \section{The NLSY79}\label{database}}
+
+Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
+This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al.
+2001).
+In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research (Pergamit et al.
+2001; Cooksey 2017).
+According to Cooksey (2017), thousands of articles, and hundreds of book chapters and monographs have utilized this data.
+Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al.
+2001).
+Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
+Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).
+We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the \emph{purposes of studying racism, not race} (Fullilove 1998).
 

### 2022-02-03 paper/paper.Rmd
related article about "the average Australian doesn't exist" from the Census data analysis

@@
 The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis supervised by author Cook [@ilk2004], and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
+
+This textbook data was used as an example in @ilk2004, and was developed into a case study for use in teaching exploratory data analysis at `r censor_word("Iowa State University")`.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+An exploratory analysis reveals however that the individual experience varies a great deal from the overall average.
+Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
 

### 2022-02-03 paper/paper.tex

@@
 
+This textbook data was used as an example in Ilk (2004), and was developed into a case study for use in teaching exploratory data analysis at CENSORED.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+An exploratory analysis reveals however that the individual experience varies a great deal from the overall average.
+Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
+
 This disparity between the average and the individual is a part of statistics, as a discipline, that is given less attention.
@@
 If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
+{[}ET: a related article: \url{https://www.smh.com.au/national/by-the-numbers--the-average-australian-doesnt-exist--not-a-single-one-of-us-is-normal-20150124-12xlh0.html}{]}
 

### 2022-02-03 paper/paper.Rmd
censor yowie

--- 2022-02-03 paper/paper.Rmd
@@
 If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
+[ET: a related article: https://www.smh.com.au/national/by-the-numbers--the-average-australian-doesnt-exist--not-a-single-one-of-us-is-normal-20150124-12xlh0.html]
 
@@
 <!--AE comment: this is an important point that shouldn't be relegated to a parenthetical statement.
---> (Tidy data shouldn't be confused with "tame data" which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called `yowie` which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
+--> (Tidy data shouldn't be confused with "tame data" which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called `r yowie` which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
 

### 2022-02-03 response/response1.Rmd
template response

---
title: Response to reviewers
author: ""
fontsize: 12pt
date: "Submission ID 217815520"
output: monash::memo
# if you don't have the monash package install by:
# remotes::install_github("numbats/monash")
---

We thank the Editor and the three reviewers for their comments, which have helped substantially to improve the manuscript. In response, we have made considerable changes to the main manuscript as outlined below.

\section*{Minor changes:}

* We have corrected all minor grammatical errors pointed out by the reviewers.

\section*{Major changes:}

### 2022-02-03 paper/paper.Rmd
Edited section 2.1.

@@
 Our process of cleaning builds heavily on the `tidyverse` approach [@tidyverse].
-The data is first organised into "tidy data" [@WickhamHadley2014TD] and then further wrangled using the data pipeline and split-apply-combine approach [@plyr].
+The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using the data pipeline and split-apply-combine approach [@plyr].
 <!--AE comment: this is an important point that shouldn't be relegated to a parenthetical statement.
@@
 The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979.
-<!--AE comment: “It comprised of Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military.” Definitely needs to remove the term “Blacks” as that is considered to be offensive.
-Perhaps “It comprised of people sampled because they were Black, Hispanic, economically disadvantaged, or in the military.” --> It comprised of Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military.
-<!--AE comment: “1,643 members of the economically disadvantaged non-Black non-Hispanics, respectively” could perhaps be rephrased to “1,643 members who were selected because of their economic disadvantage, but were not Black or Hispanic.”-->In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members of the economically disadvantaged non-Black non-Hispanics, respectively.
+It comprised of people sampled because they were Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military.
+In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members who represents economically disadvantage people, but were not Black nor Hispanics, respectively.
+These subsamples were dropped because they are not eligible anymore for interview [@nlsy79].
 Hence, 9,964 respondents remain in the eligible samples.
@@
 
-<!--AE comment: I found the semicolons distracting. Perhaps replace them with commas and use parenthetical for sub-categories, or use a bulleted list.-->
-Although the main focus area of the NLSY is labor and employment, the NLSY also covers several other topics, including education; training and achievement; household, geography, and contextual variables; dating, marriage, cohabitation; sexual activity, pregnancy, and fertility; children; income, assets and program participation; health; attitudes and expectations; and crime and substance use.
+Although the main focus area of the NLSY is labor and employment, the NLSY also covers several other topics, including education, training, achievement, household, geography, dating, marriage, cohabitation, sexual activity, pregnancy, fertility, children, income, assets, health, attitudes and expectations, crime, and substance use.
 
@@
 -->
-Now is the time to see how close we have come to the original textbook data.
 The original set contains wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994.
-The refreshed data set for comparison is also wages on high school drop outs from 1979 to 2018.
+The refreshed data set for comparison is also wages on high school dropouts from 1979 to 2018.
 The original set is available in the R package `brolgar` and the refreshed data is available in the R package `r yowie`.

### 2022-02-03 response/response1.Rmd
response update

@@
 \section*{Major changes:}
+
+* Reviewers suggested reframing the problem space so there is more emphasis on the case study.
+
+\begin{description}
+\item[Reviewer 1:] \textcolor{violet}{The authors prominently cite (Chatfield 1985) for the definition of the term “IDA”.
+However, Chatfield fairly explicitly defines IDA to be a data summarizing and scrutinizing process; not a cleaning, scrubbing, or munging process.
+It seems curious to me that the whole paper is framed as being “IDA” based on the author’s unsupported statement that data cleaning should be considered party of IDA.}
+\end{description}
+
+* Supporting information for the lack of data cleaning in papers.
+
+\begin{description}
+\item[Reviewer 1:] \textcolor{violet}{The authors also emphasize that “There are few research papers that document the data cleaning”.
+While I agree that data cleaning is under-emphasized and under-documented, it’s hardly true that there are not papers on the topic.
+A quick Google Scholar search for “data cleaning” unearths many scholarly papers and books on the principles of this process, none of which appear to be cited in this work.}
+\end{description}
+
+* Outlier detection
+
+\begin{description}
+\item[Reviewer 1:] \textcolor{violet}{
+Section 3.2.1 goes into detail on the authors’ approach to characterizing erroneous outliers in
+the data. Their approach is reasonable, but it is not (as far as I can tell) based on any
+established or tested procedure. This section once again presents a tension between the paper
+as a case study versus a topical commentary: if it is a case study only, then the “common sense”
+justification for the approach is appropriate, but if it is meant to establish future norms, these
+modeling choices must be more formally supported. The reference to a “reasonable degree of
+fluctuation” particularly struck me as a subjective or “ad hoc” claim. I am particularly a bit concerned by the statement on page 13: “The robust mixed model could
+be the best model to be employed in this case. However, this method is too computationally
+and memory expensive, especially for a large data set, like the NLSY79 data.” Surely, fitting a 
+mixed-effect model to a dataset with only one predictor and 1188 individuals ought to be very
+feasible?
+}
+\end{description}
+
+* We thank the reviewer for pointing out to us that the data contains information on the work experience -- this additional data is added in the extraction of the data in Figure 1.
+This information was compared with Singer & Willet's original data.
+Comparison shown in Figure X shows that the information does not match up, however, there is a high correlation between the variables.

### 2022-02-03 response/response1.Rmd
response update

@@
 
-\section*{Minor changes:}
-
-* We have corrected all minor grammatical errors pointed out by the reviewers.
-
 \section*{Major changes:}
 
-* Reviewers suggested reframing the problem space so there is more emphasis on the case study.
+Reviewers suggested reframing the problem space so there is more emphasis on the case study.
 
 \begin{description}
-\item[Reviewer 1:] \textcolor{violet}{The authors prominently cite (Chatfield 1985) for the definition of the term “IDA”.
+\item[Reviewer 1:] \textcolor{violet}{How is this manuscript relevant for teachers and/or students in statistics and data science?}
+\item[Reviewer 3:] \textcolor{violet}{The authors prominently cite (Chatfield 1985) for the definition of the term “IDA”.
 However, Chatfield fairly explicitly defines IDA to be a data summarizing and scrutinizing process; not a cleaning, scrubbing, or munging process.
 It seems curious to me that the whole paper is framed as being “IDA” based on the author’s unsupported statement that data cleaning should be considered party of IDA.}
+\item[Reviewer 3:] \textcolor{violet}{My suggestion is that the authors reconfigure this paper to be presented as a pure case study.
+Ibelieve the work that has been done is valuable: the data cleaning process is often messy andunplanned, so a case study like this on a popular dataset would provide an excellent educational resource.}
 \end{description}
 
-* Supporting information for the lack of data cleaning in papers.
+Supporting information for the lack of data cleaning in papers.
 
 \begin{description}
-\item[Reviewer 1:] \textcolor{violet}{The authors also emphasize that “There are few research papers that document the data cleaning”.
+\item[Reviewer 3:] \textcolor{violet}{The authors also emphasize that “There are few research papers that document the data cleaning”.
 While I agree that data cleaning is under-emphasized and under-documented, it’s hardly true that there are not papers on the topic.
@@
 
-* Outlier detection
+Outlier detection
 
 \begin{description}
-\item[Reviewer 1:] \textcolor{violet}{
+\item[Reviewer 3:] \textcolor{violet}{
 Section 3.2.1 goes into detail on the authors’ approach to characterizing erroneous outliers in
@@
 
-* We thank the reviewer for pointing out to us that the data contains information on the work experience -- this additional data is added in the extraction of the data in Figure 1.
+We thank the reviewer for pointing out to us that the data contains information on the work experience -- this additional data is added in the extraction of the data in Figure 1.
 This information was compared with Singer & Willet's original data.
 Comparison shown in Figure X shows that the information does not match up, however, there is a high correlation between the variables.
+
+\section*{Minor changes:}
+
+* We have corrected all minor grammatical errors pointed out by the reviewers.  
+* We have made the descriptions of the steps more explicit.
+
+\begin{description}
+\item[Reviewer 2:] \textcolor{violet}{My primary comment is that many elements need to be more explicitly spelled out.}
+\end{description}

### 2022-02-03 data-raw/eda.R
Add comparison plots of SW and refreshed work experience.

@@ -1,6 +1,9 @@
 ## ---- sw_wages
 
-sw <- brolgar::wages
+sw <- brolgar::wages %>%
+  group_by(id) %>%
+  mutate(index = 1:n(),
+         id = as.factor(id))
 
 sw_wages <- sw %>%
   ggplot(aes(x = xp,
@@ -45,8 +48,9 @@ sw_wages_mod <- sw %>%
 # filter do data up to 1994 since the textbook data only covers that period.
 
 do <- yowie::wages_hs_do %>%
-  mutate(lnwage = log(wage)) #%>%
-  #filter(year < 1995)
+  group_by(id) %>%
+  mutate(lnwage = log(wage),
+         index = 1:n())
 
 do_ref <- do %>%
   ggplot(aes(x = year,
@@ -169,6 +173,26 @@ do2 <- do %>%
 do2 %>% select(id, race) %>% distinct() %>% count(race)
 do2 %>% select(id, hgc12) %>% distinct() %>% count(hgc12)
 
+# ---- compare_xp_to_yrworkforce
+
+ggplot(do, aes(x = yr_wforce,
+               y = exp,
+               color = id)) +
+  geom_point() +
+  guides(color = "none")
+
+# ---- compare_xp_sw
+
+do_sw_join <- left_join(sw, do, by = c("id", "index"))
+
+ggplot(do_sw_join, aes(x = xp,
+               y = exp,
+               color = id)) +
+  geom_point() +
+  geom_abline(intercept = 0, slope = 1) +
+  guides(color = "none") +
+  xlab("Work experience (Singer & Willet, 2008)") +
+  ylab("Work experience (refreshed)")
 
 
 # Takeaways:

### 2022-02-03 paper/paper.tex
Add the experience comparison to the paper.

@@
 
-% tightlist command for lists without linebreak
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-
-% From pandoc table feature
-\usepackage{longtable,booktabs,array}
-\usepackage{calc} % for calculating minipage widths
-% Correct order of tables after \paragraph or \subparagraph
-\usepackage{etoolbox}
-\makeatletter
-\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
-\makeatother
-% Allow footnotes in longtable head/foot
-\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
-\makesavenoteenv{longtable}
-
 % Pandoc citation processing
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-% for Pandoc 2.8 to 2.10.1
-\newenvironment{cslreferences}%
-  {}%
-  {\par}
-% For Pandoc 2.11+
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 
-Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al.
+Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
 2019).
-The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
-(Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
+The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
+(Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
 
@@
 The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979.
-It comprised of Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military.
-In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members of the economically disadvantaged non-Black non-Hispanics, respectively.
+It comprised of people sampled because they were Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military.
+In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members who represents economically disadvantage people, but were not Black nor Hispanics, respectively.
+These subsamples were dropped because they are not eligible anymore for interview (Bureau of Labor Statistics, U.S.
+Department of Labor 2021).
 Hence, 9,964 respondents remain in the eligible samples.
@@
 
-Although the main focus area of the NLSY is labor and employment, the NLSY also covers several other topics, including education; training and achievement; household, geography, and contextual variables; dating, marriage, cohabitation; sexual activity, pregnancy, and fertility; children; income, assets and program participation; health; attitudes and expectations; and crime and substance use.
+Although the main focus area of the NLSY is labor and employment, the NLSY also covers several other topics, including education, training, achievement, household, geography, dating, marriage, cohabitation, sexual activity, pregnancy, fertility, children, income, assets, health, attitudes and expectations, crime, and substance use.
 
@@
 
-Now is the time to see how close we have come to the original textbook data.
 The original set contains wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994.
-The refreshed data set for comparison is also wages on high school drop outs from 1979 to 2018.
+The refreshed data set for comparison is also wages on high school dropouts from 1979 to 2018.
 The original set is available in the R package \texttt{brolgar} and the refreshed data is available in the R package \texttt{[CENSORED]}.
@@
 There are two aspects of the original data that make a direct comparison difficult.
-The time variable provided is ``experience in the workforce''.
+Th first one is that The time variable provided is ``experience in the workforce''.
 It is not clear how this is calculated.
 One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection.
-We have not been able to find this information in the database.
 Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.
+
+\begin{center}\includegraphics[width=468px]{figures/compare_xp_to_yrworkforce-1} \end{center}
+
+\begin{center}\includegraphics[width=468px]{figures/compare_xp_sw-1} \end{center}
 

### 2022-02-03 data-raw/eda.R

@@ -93,7 +93,6 @@ sw_id <- as_tibble(sw) %>%
 
 do_id <- as_tibble(do) %>%
   group_by(id) %>%
-  mutate(id = as.numeric(id)) %>%
   count() %>%
   select(id)
 

### 2022-02-03 paper/paper-jsdse/paper-jsdse.tex
pdf paper for jsdse

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) which is commonly used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
+An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
 The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 
-This textbook data was used as an example in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was developed into a case study for use in teaching exploratory data analysis at Iowa State University.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+An exploratory analysis reveals however that the individual experience varies a great deal from the overall average.
+Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
+The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
+
+This textbook data was used as an example in Ilk (2004), and was developed into a case study for use in teaching exploratory data analysis at CENSORED.
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
@@
 
-This disparity between the average and the individual is a part of statistics that is given less attention.
+This disparity between the average and the individual is a part of statistics, as a discipline, that is given less attention.
 It is this disparity which makes this particular data set important as a textbook example.
@@
 As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual.
-For example, with the wages data, the message for the individual is that you are more than your demographic.
-The majority of people tracked are not like the average, the average is an anomaly.
+For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic.
+The majority of people in the study do not have a pattern that is similar to the average.
 If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
+{[}ET: a related article: \url{https://www.smh.com.au/national/by-the-numbers--the-average-australian-doesnt-exist--not-a-single-one-of-us-is-normal-20150124-12xlh0.html}{]}
 
-As textbook data though, the original data is outdated.
+As a textbook data set though, it is outdated.
 The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
 Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students.
-Using tired old textbook data sets sets up a misconception that the field is not current.
+Using tired old textbook data sets can imbue a misconception that the field is not current.
 The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated.
 It should be possible to continuously refresh the textbook data from the data repository.
-This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) but wild data to textbook data.
+This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) wild data to textbook data.
+
+This paper demonstrates the steps of cleaning data, and documents the process, as recommended by Huebner, Vach, and Cessie (2016).
+They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
+Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985).
+In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
+This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
+EDA can be considered to encompass IDA.
+Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.
+
+Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al.
+2019).
+The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
+(Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
+
+This paper is structured in the following way.
+Section \ref{database} describes the NLSY79 data source.
+Section \ref{cleaning} presents the steps of cleaning the data, including getting and tidying the data from the NLSY79 and IDA to find and repair anomalies.
+Our final subset is compared to the old textbook subset in Section \ref{compare}.
+Finally, Section \ref{summary} summarizes the contribution and makes recommendations for the NLSY79 data curators.
+
+\hypertarget{database}{%
+\section{The NLSY79}\label{database}}
 
@@
 
-This paper demonstrates the steps of cleaning data, and documents the process, as recommended by Huebner, Vach, and Cessie (2016).
-They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
-Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985).
-In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
-This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
-EDA can be considered to encompass IDA.
-Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.
-
-Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al.
-2019).
-The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
-(Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
-
-This paper is structured in the following way.
-Section \ref{database} describes the NLSY79 data source.
-Section \ref{cleaning} presents the steps of cleaning the data, including getting and tidying the data from the NLSY79 and IDA to find and repair anomalies.
-Our final subset is compared to the old textbook subset in Section \ref{compare}.
-Finally, Section \ref{summary} summarizes the contribution and makes recommendations for the NLSY79 data curators.
-
-\hypertarget{database}{%
-\section{The NLSY79}\label{database}}
-
 \hypertarget{database-1}{%

### 2022-02-03 paper/paper-jsdse/paper-jsdse.tex
Edited Figure 7.

--- 2022-02-02 paper/paper-jsdse/paper-jsdse.tex
@@
 As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual.
-For example, with the wages data, the message for the individual is that you are more than your demographic.
-The majority of people tracked are not like the average, the average is an anomaly.
+For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic.
+The majority of people in the study do not have a pattern that is similar to the average.
 If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
+{[}ET: a related article: \url{https://www.smh.com.au/national/by-the-numbers--the-average-australian-doesnt-exist--not-a-single-one-of-us-is-normal-20150124-12xlh0.html}{]}
 
-As textbook data though, the original data is outdated.
+As a textbook data set though, it is outdated.
 The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
 Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students.
-Using tired old textbook data sets sets up a misconception that the field is not current.
+Using tired old textbook data sets can imbue a misconception that the field is not current.
 The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated.
 It should be possible to continuously refresh the textbook data from the data repository.
-This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) but wild data to textbook data.
+This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) wild data to textbook data.
+
+This paper demonstrates the steps of cleaning data, and documents the process, as recommended by Huebner, Vach, and Cessie (2016).
+They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
+Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985).
+In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
+This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
+EDA can be considered to encompass IDA.
+Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.
+
+Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
+2019).
+The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
+(Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
+
+This paper is structured in the following way.
+Section \ref{database} describes the NLSY79 data source.
+Section \ref{cleaning} presents the steps of cleaning the data, including getting and tidying the data from the NLSY79 and IDA to find and repair anomalies.
+Our final subset is compared to the old textbook subset in Section \ref{compare}.
+Finally, Section \ref{summary} summarizes the contribution and makes recommendations for the NLSY79 data curators.
+
+\hypertarget{database}{%
+\section{The NLSY79}\label{database}}
 
@@
 
-This paper demonstrates the steps of cleaning data, and documents the process, as recommended by Huebner, Vach, and Cessie (2016).
-They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
-Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985).
-In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
-This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
-EDA can be considered to encompass IDA.
-Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.
-
-Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
-2019).
-The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
-(Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
-
-This paper is structured in the following way.
-Section \ref{database} describes the NLSY79 data source.
-Section \ref{cleaning} presents the steps of cleaning the data, including getting and tidying the data from the NLSY79 and IDA to find and repair anomalies.
-Our final subset is compared to the old textbook subset in Section \ref{compare}.
-Finally, Section \ref{summary} summarizes the contribution and makes recommendations for the NLSY79 data curators.
-
-\hypertarget{database}{%
-\section{The NLSY79}\label{database}}
-
 \hypertarget{database-1}{%
@@
 There are two aspects of the original data that make a direct comparison difficult.
-The time variable provided is ``experience in the workforce''.
+Th first one is that The time variable provided is ``experience in the workforce''.
 It is not clear how this is calculated.
 One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection.
-We have not been able to find this information in the database.
 Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.
+
+\textbackslash begin\{figure\}
+
+\{\centering \includegraphics[width=468px]{figures/compare_xp_to_yrworkforce-1}
+
+\}
+
+\caption{The association between the lenght of time in the workforce (years) and the work experience.
+The color represents the respondents (id).
+Generally, the longer someone were in the workforce, the higher the work experience.
+However, some of the respondents have much lower working experience compared their duration in the workforce.}
+
+(\#fig:compare\_xp\_to\_yrworkforce)
+\textbackslash end\{figure\}
+
+\begin{center}\includegraphics[width=468px]{figures/compare_xp_sw-1} \end{center}
 

### 2022-02-03 paper/paper.Rmd
Add the experience comparison to the paper.
Edited Figure 7.

@@
 There are two aspects of the original data that make a direct comparison difficult.
-The time variable provided is "experience in the workforce".
+Th first one is that The time variable provided is "experience in the workforce".
 It is not clear how this is calculated.
 One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection.
-We have not been able to find this information in the database.
 Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.
+
+```{r compare_xp_to_yrworkforce, fig.cap = "The association between the lenght of time in the workforce (years) and the work experience.
+The color represents the respondents (id).
+Generally, the longer someone were in the workforce, the higher the work experience.
+However, some of the respondents have much lower working experience compared their duration in the workforce."}
+```
+
+```{r compare_xp_sw}
+```
 

### 2022-02-03 data-raw/eda.R

@@ -66,7 +66,7 @@ do_ref <- do %>%
 
 do_ref_mod <- do %>%
   mutate(hgc12 = ifelse(hgc_i < 12, "BELOW 12TH", "12TH")) %>%
-  ggplot(aes(x = year,
+  ggplot(aes(x = exp,
              y = lnwage,
              linetype = hgc12)) +
   geom_smooth(method = "lm", se = FALSE,
@@ -75,7 +75,7 @@ do_ref_mod <- do %>%
   scale_linetype("") +
   theme_bw() +
   ylab("Hourly wage ($, natural log)") +
-  xlab("Year of data collection") +
+  xlab("Experience (years)") +
   theme(legend.position = "bottom",
         legend.direction = "horizontal")
 
@@ -140,18 +140,18 @@ sw_wages_agree <- sw_agree %>%
 
 
 do_ref_agree <- do_agree %>%
-  ggplot(aes(x = year,
+  ggplot(aes(x = exp,
              y = lnwage)) +
   geom_line(aes(group = id), alpha = 0.1) +
   geom_smooth(se = FALSE) +
   labs(tag = "B") +
   theme_bw() +
   ylab("ln(Hourly wage) ($)") +
-  xlab("Year") +
+  xlab("Experience (years)") +
   theme(plot.title = element_text(size = 10)) +
   ylim(-3, 5)
 
-# sw_wages_agree + do_ref_agree
+#sw_wages_agree + do_ref_agree
 # sw_wages + do_ref
 sw_wages_mod + do_ref_mod
 
@@ -175,19 +175,18 @@ do2 %>% select(id, hgc12) %>% distinct() %>% count(hgc12)
 # ---- compare_xp_to_yrworkforce
 
 ggplot(do, aes(x = yr_wforce,
-               y = exp,
-               color = id)) +
-  geom_point() +
-  guides(color = "none")
+               y = exp)) +
+  geom_point(alpha = 0.1) +
+  guides(color = "none") +
+  geom_abline(intercept = 0, slope = 1)
 
 # ---- compare_xp_sw
 
 do_sw_join <- left_join(sw, do, by = c("id", "index"))
 
 ggplot(do_sw_join, aes(x = xp,
-               y = exp,
-               color = id)) +
-  geom_point() +
+               y = exp)) +
+  geom_point(alpha = 0.1) +
   geom_abline(intercept = 0, slope = 1) +
   guides(color = "none") +
   xlab("Work experience (Singer & Willet, 2008)") +

### 2022-02-03 paper/paper.Rmd
Edited the paper and add the response.

@@
 According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
-The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.<!-- AE comment: This sentence is unclear.
-What is dplyr used for?
-Are there tidyverse packages used but not mentioned?
-If not, splitting this sentence into two may make it clearer which package is used to what purpose.
-Also, I haven't tried it in code, but I imagine that the data could be tidied as described just using pivot_longer, at least for the job number, year, and wage data.
-In other words, I don't see why dplyr or stringr would be needed for the data described on page 6.--> This is done using the `tidyverse` suite of packages [@tidyverse], `tidyr` [@tidyr] to pivot the data into long form, `dplyr` [@dplyr], and `stringr` [@stringr] for creating new variables, and levels of factors by text wrangling.
+The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the `tidyverse` suite of packages [@tidyverse], i.e., `tidyr` [@tidyr] to pivot the data into long form, `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and create levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 
-The number of jobs (`number_of_jobs`) for each respondent per year is computed from the number of non-missing values of hourly wage *and* hours worked.
-If either the hourly wage or hours worked is missing, we do not tally this.
-<!-- AE comment: "If either the hourly wage or hours worked is missing, we do not tally this." I take "this" to mean "number_of_jobs".
-But in row 2, total_hours is missing, yet number_of_jobs is 1.
-The number of jobs (1) was tallied even though hours worked was missing.
--->
+The number of jobs (`number_of_jobs`) for each respondent per year is computed from the number of non-missing values of hourly wage.
+Even if the information of hours worked exists for a particular observation, we do not tally when the hourly wage is missing.
 
@@
 I suspect not, but please be explicit.-->
-<!--AE comment: Rewrite this sentence.
-"Observations" aren't working 420 hours.
-A "respondent" might have been reported as having worked 420 hours per week.--> The anomalies are also found in the total hours of work, where some observations reported as having worked for 420 hours a week in total.
+The anomalies are also found in the total hours of work, where some observations reported as having worked for 420 hours a week in total.
 According to @MichaelRPergamit2001DWTN, one of the flaws of the NLSY79 employment data is that the NLSY79 collects the information of the working hours since the last interview.
@@
 
-```{r compare_xp_to_yrworkforce, fig.cap = "The association between the lenght of time in the workforce (years) and the work experience.
-The color represents the respondents (id).
+```{r compare_xp_to_yrworkforce, fig.cap = "The association between the lenght of time in the workforce and the work experience.
+The color represents the respondents.
 Generally, the longer someone were in the workforce, the higher the work experience.
-However, some of the respondents have much lower working experience compared their duration in the workforce."}
+However, some of the respondents have much lower working experience compared their duration in the workforce.", eval=FALSE}
 ```

### 2022-02-03 response/response1.Rmd
Edited the paper and add the response.

@@
 \begin{description}
+\item[Reviewer 1:]\textcolor{violet}{It's unclear from context what is being referred to from the Huebner et al 2020 citation.}
+\end{description}
+
+* We have fixed the sentences and the reference.
+
+\begin{description}
+\item[Reviewer 1:]\textcolor{violet}{What is dplyr used for?
+Are there tidyverse packages used but not mentioned?
+If not, splitting this sentence into two may make it clearer which package is used to what purpose.
+Also, I haven't tried it in code, but I imagine that the data could be tidied as described just using pivot_longer, at least for the job number, year, and wage data.
+In other words, I don't see why dplyr or stringr would be needed for the data described on page 6.}
+\end{description}
+
+* We have split the sentences and described the use of each package mentioned.
+
+\begin{description}
+\item[Reviewer 1:]\textcolor{violet}{please clarify.
+"If either the hourly wage or hours worked is missing, we do not tally this." I take "this" to mean "number_of_jobs".
+But in row 2, total_hours is missing, yet number_of_jobs is 1.
+The number of jobs (1) was tallied even though hours worked was missing.}
+\end{description}
+
+* We have clarified this that we only tally the number of jobs if the hourly wage is not missing.
+
+\begin{description}
+\item[Reviewer 1:]{"to find the anomaly in the wages values" This comes as a surprise to the reader.
+What anomaly?
+The wage anomalies should be mentioned in the introduction so as not to be a surprise at this point in the manuscript.}
+\end{description}
+
+\begin{description}
 \item[Reviewer 2:] \textcolor{violet}{My primary comment is that many elements need to be more explicitly spelled out.}

### 2022-02-03 response/response1.tex
Edited the paper and add the response.

\documentclass[12pt,a4paper,]{article}
\usepackage{float}
%\usepackage[sfdefault,lf,t]{carlito}
\usepackage[default,tabular,lf]{sourcesanspro}
\usepackage[cmintegrals]{newtxsf}
\usepackage[italic,eulergreek]{mathastext}
\usepackage{setspace}

\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{csquotes}
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
\usepackage{fancybox}
\usepackage[unicode=true]{hyperref}
\hypersetup{
            pdftitle={Response to reviewers},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage{geometry}
\geometry{left=2cm,right=2cm,top=2.5cm,bottom=2.5cm}
\usepackage[style=authoryear-comp,backend=biber, natbib=true,]{biblatex}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\title{Response to reviewers}

%% MONASH STUFF

%% CAPTIONS
\RequirePackage{caption}
\DeclareCaptionStyle{italic}[justification=centering]
 {labelfont={bf},textfont={it},labelsep=colon}
\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}

%% FONT
\usepackage{bm,url}

%% HEADERS AND FOOTERS
\RequirePackage{fancyhdr}
\pagestyle{fancy}
\lfoot{}\cfoot{}\rfoot{}
\lhead{\textsf{Response to reviewers}}
\rhead{\textsf{\thepage}}
\setlength{\headheight}{15pt}
\renewcommand{\headrulewidth}{0.4pt}
\fancypagestyle{plain}{%
\fancyhf{} % clear all header and footer fields
\fancyfoot[C]{\sffamily\thepage} % except the center
\renewcommand{\headrulewidth}{0pt}
\renewcommand{\footrulewidth}{0pt}}

%% MATHS
\RequirePackage{bm,amsmath}
\allowdisplaybreaks

%% GRAPHICS
\RequirePackage{graphicx}
\setcounter{topnumber}{2}
\setcounter{bottomnumber}{2}
\setcounter{totalnumber}{4}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.85}
\renewcommand{\textfraction}{0.15}
\renewcommand{\floatpagefraction}{0.8}

%\RequirePackage[section]{placeins}

%% SECTION TITLES
\RequirePackage[compact,sf,bf]{titlesec}
\titleformat{\section}[block]
  {\fontsize{15}{17}\bfseries\sffamily}
  {\thesection}
  {0.4em}{}
\titleformat{\subsection}[block]
  {\fontsize{12}{14}\bfseries\sffamily}
  {\thesubsection}
  {0.4em}{}
\titlespacing{\section}{0pt}{*3}{*1}
\titlespacing{\subsection}{0pt}{*1}{*0.5}

%% LINE AND PAGE BREAKING
\sloppy
\raggedbottom
\usepackage[bottom]{footmisc}
\clubpenalty = 10000
\widowpenalty = 10000
\brokenpenalty = 10000
\RequirePackage{microtype}

%% HYPERLINKS
\RequirePackage{xcolor} % Needed for links
\definecolor{darkblue}{rgb}{0,0,.6}
\RequirePackage{url}

\makeatletter
\@ifpackageloaded{hyperref}{}{\RequirePackage{hyperref}}
\makeatother
\hypersetup{
     citecolor=0 0 0,
     breaklinks=true,
     bookmarksopen=true,
     bookmarksnumbered=true,
     linkcolor=darkblue,
     urlcolor=blue,
     citecolor=darkblue,
     colorlinks=true}

\usepackage[showonlyrefs]{mathtools}

%% BIBLIOGRAPHY

\makeatletter
\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
\makeatother
\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false}
\DeclareFieldFormat{url}{\texttt{\url{#1}}}
\DeclareFieldFormat[article]{pages}{#1}
\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
\DeclareFieldFormat[article]{url}{}
%\DeclareFieldFormat[book]{url}{}
%\DeclareFieldFormat[inbook]{url}{}
%\DeclareFieldFormat[incollection]{url}{}
%\DeclareFieldFormat[inproceedings]{url}{}
\DeclareFieldFormat[inproceedings]{title}{#1}
\DeclareFieldFormat{shorthandwidth}{#1}
%\DeclareFieldFormat{extrayear}{}
% No dot before number of articles
\usepackage{xpatch}
\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
% Remove In: for an article.
\renewbibmacro{in:}{%
  \ifentrytype{article}{}{%
  \printtext{\bibstring{in}\intitlepunct}}}
\AtEveryBibitem{\clearfield{month}}
\AtEveryCitekey{\clearfield{month}}
\makeatletter
\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
\makeatother
\renewcommand*{\finalnamedelim}{%
  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
  \addspace\&\space}

%%% Change title format
\usepackage{color,titling,framed}
\usepackage[absolute,overlay]{textpos}
\setlength{\TPHorizModule}{1cm}
\setlength{\TPVertModule}{1cm}

\pretitle{%

\vspace*{-1.2cm}

\LARGE\bfseries}
\posttitle{\vspace*{0.3cm}\par}
\preauthor{\large}
\postauthor{\hfill}
\predate{\small}
\postdate{\vspace*{0.1cm}}

\raggedbottom

\usepackage[australian]{babel}
\date{Submission ID 217815520}

\begin{document}
\vspace*{-2cm}
\definecolor{shadecolor}{RGB}{210,210,210}
\begin{snugshade}\sffamily
\maketitle
\end{snugshade}\vspace*{0.5cm}
\definecolor{shadecolor}{RGB}{248,248,248}
\setstretch{1.2}

We thank the Editor and the three reviewers for their comments, which have helped substantially to improve the manuscript. In response, we have made considerable changes to the main manuscript as outlined below.

\section*{Major changes:}

Reviewers suggested reframing the problem space so there is more emphasis on the case study.

\begin{description}
\item[Reviewer 1:] \textcolor{violet}{How is this manuscript relevant for teachers and/or students in statistics and data science?}
\item[Reviewer 3:] \textcolor{violet}{The authors prominently cite (Chatfield 1985) for the definition of the term “IDA”. However, Chatfield fairly explicitly defines IDA to be a data summarizing and scrutinizing process; not a cleaning, scrubbing, or munging process. It seems curious to me that the whole paper is framed as being “IDA” based on the author’s unsupported statement that data cleaning should be considered party of IDA.}
\item[Reviewer 3:] \textcolor{violet}{My suggestion is that the authors reconfigure this paper to be presented as a pure case study. Ibelieve the work that has been done is valuable: the data cleaning process is often messy andunplanned, so a case study like this on a popular dataset would provide an excellent educational resource.}
\end{description}

Supporting information for the lack of data cleaning in papers.

\begin{description}
\item[Reviewer 3:] \textcolor{violet}{The authors also emphasize that “There are few research papers that document the data cleaning”. While I agree that data cleaning is under-emphasized and under-documented, it’s hardly true that there are not papers on the topic. A quick Google Scholar search for “data cleaning” unearths many scholarly papers and books on the principles of this process, none of which appear to be cited in this work.}
\end{description}

Outlier detection

\begin{description}
\item[Reviewer 3:] \textcolor{violet}{
Section 3.2.1 goes into detail on the authors’ approach to characterizing erroneous outliers in
the data. Their approach is reasonable, but it is not (as far as I can tell) based on any
established or tested procedure. This section once again presents a tension between the paper
as a case study versus a topical commentary: if it is a case study only, then the “common sense”
justification for the approach is appropriate, but if it is meant to establish future norms, these
modeling choices must be more formally supported. The reference to a “reasonable degree of
fluctuation” particularly struck me as a subjective or “ad hoc” claim. I am particularly a bit concerned by the statement on page 13: “The robust mixed model could
be the best model to be employed in this case. However, this method is too computationally
and memory expensive, especially for a large data set, like the NLSY79 data.” Surely, fitting a 
mixed-effect model to a dataset with only one predictor and 1188 individuals ought to be very
feasible?
}
\end{description}

We thank the reviewer for pointing out to us that the data contains information on the work experience -- this additional data is added in the extraction of the data in Figure 1. This information was compared with Singer \& Willet's original data. Comparison shown in Figure X shows that the information does not match up, however, there is a high correlation between the variables.

\section*{Minor changes:}

\begin{itemize}
\tightlist
\item
  We have corrected all minor grammatical errors pointed out by the reviewers.\\
\item
  We have made the descriptions of the steps more explicit.
\end{itemize}

\begin{description}
\item[Reviewer 1:]\textcolor{violet}{It's unclear from context what is being referred to from the Huebner et al 2020 citation.}
\end{description}

\begin{itemize}
\tightlist
\item
  We have fixed the sentences and the reference.
\end{itemize}

\textbackslash begin\{description\}

\item[Reviewer 1:]

\textbackslash textcolor\{violet\}\{What is dplyr used for? Are there tidyverse packages used but not mentioned? If not, splitting this sentence into two may make it clearer which package is used to what purpose. Also, I haven't tried it in code, but I imagine that the data could be tidied as described just using pivot\_longer, at least for the job number, year, and wage data. In other words, I don't see why dplyr or stringr would be needed for the data described on page 6.\}
\textbackslash end\{description\}

\begin{itemize}
\tightlist
\item
  We have split the sentences and described the use of each package mentioned.
\end{itemize}

\textbackslash begin\{description\}

\item[Reviewer 1:]

\textbackslash textcolor\{violet\}\{please clarify. \enquote{If either the hourly wage or hours worked is missing, we do not tally this.} I take \enquote{this} to mean \enquote{number\_of\_jobs}. But in row 2, total\_hours is missing, yet number\_of\_jobs is 1. The number of jobs (1) was tallied even though hours worked was missing.\}
\textbackslash end\{description\}

\begin{itemize}
\tightlist
\item
  We have clarified this that we only tally the number of jobs if the hourly wage is not missing.
\end{itemize}

\begin{description}
\item[Reviewer 2:] \textcolor{violet}{My primary comment is that many elements need to be more explicitly spelled out.}
\end{description}

\printbibliography

\end{document}

### 2022-02-04 paper/paper.Rmd
Edited and add response from Reviewer 1.

@@
 As stated previously, the respondents' ages ranged from 12 to 22 when first interviewed in 1979.
-Hence, we validate whether all of the respondents were in this range.
+Hence, we validate whether all of the respondents were in this range in the data we extracted.
 Additionally, the [NLSY](https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction) also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic).
 To validate this, we used the `demog_nlsy79`, i.e., the data with the survey years 1979 sample.
-<!--AE comment:  I don't know that Table 1 is necessary as I don't have anything to compare it to.
-If the numbers check out, a simple statement to that effect would suffice.
-Also, what does "(?CHECK)" mean?
-This needs to be resolved.-->
-<!--AE comment: Table 1 includes the note ?CHECK which I assume was a note to self. -->
 Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest that the demographic data we had is consistent with the sample information in the database.

### 2022-02-04 response/response1.Rmd
Edited and add response from Reviewer 1.

@@
 If not, splitting this sentence into two may make it clearer which package is used to what purpose.
-Also, I haven't tried it in code, but I imagine that the data could be tidied as described just using pivot_longer, at least for the job number, year, and wage data.
+Also, I haven't tried it in code, but I imagine that the data could be tidied as described just using pivot\_longer, at least for the job number, year, and wage data.
 In other words, I don't see why dplyr or stringr would be needed for the data described on page 6.}
@@
 \item[Reviewer 1:]\textcolor{violet}{please clarify.
-"If either the hourly wage or hours worked is missing, we do not tally this." I take "this" to mean "number_of_jobs".
-But in row 2, total_hours is missing, yet number_of_jobs is 1.
+"If either the hourly wage or hours worked is missing, we do not tally this." I take "this" to mean "number\_of\_jobs".
+But in row 2, total\_hours is missing, yet number\_of\_jobs is 1.
 The number of jobs (1) was tallied even though hours worked was missing.}
@@
 \begin{description}
-\item[Reviewer 1:]{"to find the anomaly in the wages values" This comes as a surprise to the reader.
+\item[Reviewer 1:]\textcolor{violet}{"to find the anomaly in the wages values" This comes as a surprise to the reader.
 What anomaly?
@@
 \begin{description}
-\item[Reviewer 2:] \textcolor{violet}{My primary comment is that many elements need to be more explicitly spelled out.}
+\item[Reviewer 1:]\textcolor{violet}{I don't know that Table 1 is necessary as I don't have anything to compare it to.
+If the numbers check out, a simple statement to that effect would suffice.
+Also, what does "(?CHECK)" mean?
+This needs to be resolved.}
 \end{description}
+
+* We have removed "(?CHECK)" in Table 1's caption. We also have stated in a sentences after the table that Table 1 reflects the consistency between the data we had and the database.
+
+\begin{description}
+\item[Reviewer 1:]\textcolor{violet}{for example, ID 39 experienced an unusual wage only in one year.' That's one way to fix this sentence, which needs to be rewritten.}
+\end{description}
+
+* We have rewritten this sentences.
+
+\begin{description}
+\item[Reviewer 1:]\textcolor{violet}{How many individuals up to 8th grade are in the refreshed dataset? What are plausible reasons for these differences?}
+\end{description}
+
+* There is no individuals up to 8th grade in the refreshed dataset as we only filtered the high school dropouts (started from 9th grade ,i.e., high school grade).
+The plausible reasons for the differences are we might have different definition of high school dropouts with the original data and the corresponding individual back to high school after they dropped out so that their highest grade completed is 12th grade.
+
+\begin{description}
+\item[Reviewer 1:]\textcolor{violet}{How many individuals up to 8th grade are in the refreshed dataset? What are plausible reasons for these differences?}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{My primary comment is that many elements need to be more explicitly spelled out.}
+\end{description}

### 2022-02-04 data-raw/ida.R

@@ -13,7 +13,7 @@ age_table <- full_demographics %>%
 age_table %>%
   mutate(n = scales::comma(n)) %>%
   kable(
-      caption = "The frequency table of the age at the start of the survey (?CHECK) in the full NSLY79 data",
+      caption = "The frequency table of the age at the start of the survey in NSLY79 cohort",
       col.names = c("Age", "Number of individuals"),
       booktabs = TRUE,
       linesep = "",

### 2022-02-07 paper/paper.Rmd
Modified Section 2.2. as per Reviewer 2's comment.

@@
                   legend.key        = element_rect(fill = 'transparent', colour = NA)
-                  ) )
+                  ))
 ```
@@
 
-<!--AE comment: What does “yearly mean hourly wages with years of workforce experience” mean? Is that one variable or two?-->
-<!--AE comment: Was your goal to collect data from 1994-2018 or 1979-2018?
-Page 4 states 1994 onward, but from the figures later it appears that you collected data starting in 1979 (which makes more sense to me).-->
-<!--AE comment: “The NLSY79 data used in Singer and Willett (2003) contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994.” Might work better split into two sentences.
--->
-The NLSY79 data used in @SingerJudithD2003Alda contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994.
-In addition, the cohort is restricted to male high-school dropouts who first participated in the study at age 14-17 years.
-<!--AE comment: Thus the target of our study is to collect the same variables for the same time period and extending through to 2018, the most recent year reported.-->Thus the target data set is to collect the same variables for the extended time frame of 1994 through to 2018, the most recent year reported.
+The NLSY79 data used in @SingerJudithD2003Alda contains the longitudinal measurements of the male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994.
+This dataset comprises of several variables as follows:
+
+1. ID: the respondents' ID. 
+2. EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
+3. LNW: natural logarithm of wages, adjusted with 1990's inflation rate. 
+4. BLACK: binary variables, the respondents are coded to 1 because they were Black and 0 otherwise. 
+5. HGC: the highest grade completed.
+6. UERATE: unemployment rate of the year of survey. When missing, the variable is set to be 7.875 (the average rate).
+
+From the database, we will refresh this data with a more up to date period, which is from survey year 1979 to survey year 2018 (the most recent year published).
+We will  also modify the variables (for example, we create a race variable, instead of the binary variable "Black") and add some variables so that more examples of data exploration can be displayed in teaching.
+The variables added are year of survey, age of individual in 1979, year when the highest grade completed, number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual starting to work, and the number of years the individual worked.
+
+In addition to the dropouts data, we will also make the up to high school graduates' wages available, including the females, and the demographic data of the NLSY79 cohort available in separate datasets.
+Thus, we will create three datasets as follows:
+
+1. The demographic data of the NLSY79 cohort.
+2. The wages data of the up to high school graduates.
+3. The wages data of the high school dropouts, which is the refreshed version of @SingerJudithD2003Alda's data.
 

### 2022-02-07 paper/paper.tex
Edited the paper and add the response.
Modified Section 2.2. as per Reviewer 2's comment.

@@
 
-The NLSY79 data used in Singer and Willett (2003) contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994.
-In addition, the cohort is restricted to male high-school dropouts who first participated in the study at age 14-17 years.
-Thus the target data set is to collect the same variables for the extended time frame of 1994 through to 2018, the most recent year reported.
+The NLSY79 data used in Singer and Willett (2003) contains the longitudinal measurements of the male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994.
+This dataset comprises of several variables as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  ID: the respondents' ID.
+\item
+  EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
+\item
+  LNW: natural logarithm of wages, adjusted with 1990's inflation rate.
+\item
+  BLACK: binary variables, the respondents are coded to 1 because they were Black and 0 otherwise.
+\item
+  HGC: the highest grade completed.
+\item
+  UERATE: unemployment rate of the year of survey. When missing, the variable is set to be 7.875 (the average rate).
+\end{enumerate}
+
+From the database, we will refresh this data with a more up to date period, which is from survey year 1979 to survey year 2018 (the most recent year published).
+We will also modify the variables (for example, we create a race variable, instead of the binary variable ``Black'') and add some variables so that more examples of data exploration can be displayed in teaching.
+The variables added are year of survey, age of individual in 1979, year when the highest grade completed, number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual starting to work, and the number of years the individual worked.
+
+In addition to the dropouts data, we will also make the up to high school graduates' wages available, including the females, and the demographic data of the NLSY79 cohort available in separate datasets.
+Thus, we will create three datasets as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The demographic data of the NLSY79 cohort.
+\item
+  The wages data of the up to high school graduates.
+\item
+  The wages data of the high school dropouts, which is the refreshed version of Singer and Willett (2003)'s data.
+\end{enumerate}
 
@@
 According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
-The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
-This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
-2019), \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham, François, et al.
-2020), and \texttt{stringr} (Wickham 2019) for creating new variables, and levels of factors by text wrangling.
+The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
+2019), i.e., \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham, François, et al.
+2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and create levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 
-The number of jobs (\texttt{number\_of\_jobs}) for each respondent per year is computed from the number of non-missing values of hourly wage \emph{and} hours worked.
-If either the hourly wage or hours worked is missing, we do not tally this.
+The number of jobs (\texttt{number\_of\_jobs}) for each respondent per year is computed from the number of non-missing values of hourly wage.
+Even if the information of hours worked exists for a particular observation, we do not tally when the hourly wage is missing.
 
@@
 As stated previously, the respondents' ages ranged from 12 to 22 when first interviewed in 1979.
-Hence, we validate whether all of the respondents were in this range.
+Hence, we validate whether all of the respondents were in this range in the data we extracted.
 Additionally, the \href{https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction}{NLSY} also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic).
@@
 
-\caption{\label{tab:age-table}The frequency table of the age at the start of the survey (?CHECK) in the full NSLY79 data}
+\caption{\label{tab:age-table}The frequency table of the age at the start of the survey in NSLY79 cohort}
 \centering
@@
 Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.
-
-\begin{center}\includegraphics[width=468px]{figures/compare_xp_to_yrworkforce-1} \end{center}
 

### 2022-02-07 response/response1.Rmd
Modified Section 2.2. as per Reviewer 2's comment.

@@
 \end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{I would like more discussion about the textbook selected as the exemplar.
+It sounds like this dataset is used in many books, why choose that one?
+The text should include the name of the book, so readers do not need to flip to the citations in order to learn what it was.
+From the name, I am guessing that the book focuses heavily on this dataset, which is why it makes sense to reproduce their subset.
+Be explicit about this.}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{When listing the files that come zipped from the data source, be more explicit about the file formats. Comma separated values instead of csv, etc.}
+\end{description}
+
+* We have made it explicit.
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Before the data cleaning section I would love a glimpse into the future to see what the target dataset would look like.
+What are the observations?
+What are the variables?
+This would allow me to follow the narrative of data wrangling more easily.
+Perhaps this could be put into section 2.2, which is titled Target data.
+Then, when starting to describe the raw data in 3,1, explain the same pieces and explicitly explain why the raw data it is not tidy to begin with.}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Before the data cleaning section I would love a glimpse into the future to see what the target dataset would look like.
+What are the observations?
+What are the variables?
+This would allow me to follow the narrative of data wrangling more easily.
+Perhaps this could be put into section 2.2, which is titled Target data.
+Then, when starting to describe the raw data in 3.1, explain the same pieces and explicitly explain why the raw data it is not tidy to begin with.}
+\end{description}
+
+* We have added the explanation of the observations and the variable in Section 2.2.

### 2022-02-07 paper/paper.Rmd
Added some responses to the reviewers' comments.

@@
 
-The raw data, `new_data_qnames`, is organised such that each row corresponds to an individual.
+According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
+The raw data, `new_data_qnames`, does not comply with these rules as it is organised such that each row corresponds to an individual.
 As respondents can have multiple jobs at specific years, the column names, such as `HRP1_1979`, `HRP2_1979`, `HRP1_1980` and `HRP2_1980`, contain the information about the job number up to 5 (`HRP1` = job 1, `HRP2` = job 2) and the year.
-The raw data consequently has a large number of columns (686 to be specific).
+The raw data consequently has a large number of columns (716 to be specific).
 The values in the cell under the variables that begin with `HRP` correspond to the hourly wage in dollars.
@@
 
-According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
-The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the `tidyverse` suite of packages [@tidyverse], i.e., `tidyr` [@tidyr] to pivot the data into long form, `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and create levels of factors by text wrangling.
+Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the `tidyverse` suite of packages [@tidyverse], i.e., `tidyr` [@tidyr] to pivot the data into long form, `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and create levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 The highest grade completed are recorded in `new_data_qnames` as variables beginning with `Q3-4` and `HGC` with suffix of the year it was recorded.
-In addition the variables beginning with `HGCREV` contain the revised data.
-We choose to use this revised May data because it seemed to have less missing and presumably has been checked.
-However, there is no revised May data for 2012, 2014, 2016, and 2018, thus, we use the ordinary May data for these years.
+In addition, the variables beginning with `HGCREV` contain the revised data.
+We choose to use this revised data because it seemed to have less missing and presumably has been checked.
+However, there is no revised data for 2012, 2014, 2016, and 2018, thus, we use the ordinary data for these years.
 
@@
 The number of jobs (`number_of_jobs`) for each respondent per year is computed from the number of non-missing values of hourly wage.
-Even if the information of hours worked exists for a particular observation, we do not tally when the hourly wage is missing.
+In other words, even if the information of hours worked exists for a particular observation, we do not tally when the hourly wage is missing.
 
@@
 There are two aspects of the original data that make a direct comparison difficult.
-Th first one is that The time variable provided is "experience in the workforce".
+The first one is that The time variable provided is "experience in the workforce".
 It is not clear how this is calculated.
@@
 However, in recent context, this value is too low to be set as the maximum threshold.
-We opted to use the weights from a robust linear regression to determine what should be set as missing value as described in Section \@ref(censor).
+We opted to use the weights from a robust linear regression to determine what should be regarded as outliers and imputed them with their predicted values as as described in Section \@ref(censor).
 

### 2022-02-07 response/response1.Rmd
Added some responses to the reviewers' comments.

@@
 \begin{description}
-\item[Reviewer 1:]\textcolor{violet}{How many individuals up to 8th grade are in the refreshed dataset? What are plausible reasons for these differences?}
+\item[Reviewer 1:]\textcolor{violet}{what does the natural log have to do with the y-axis?}
 \end{description}
@@
 * We have added the explanation of the observations and the variable in Section 2.2.
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{With the IDA about wage data, I would appreciate a bit more context.
+Why sample 20 observations?
+I am not a longitudinal data expert, so I was a bit surprised when I flipped to the plot and found it to be small-multiple time series, although upon further consideration it makes sense.
+In that section it would be good to describe what the figure shows in a bit more detail.
+Perhaps “We randomly sample 36 respondents from the data, and plot their average wage as a time series.
+Looking at the patterns over time, we see a lot of variability in wages.
+For example, the people shown in panels 5, 7, and 11 have [explain what is interesting here, again I’m not an expert].” The next sentence here, “Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.” does not seem to provide much additional information.
+What were you looking for here?
+Overall trends over time, increasing/decreasing?
+Any places where one year’s wage looked really different?}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{I think there is a narrative step missing here, which is that looking at those samples led you to look at plots that show values for the wage variable over all participants and all years.
+The phrase “the summary plots” was not enough description for me to know what to expect when I flipped to Figure 3.
+Please be explicit about what the three plots were, why they were made, and what you were looking for.
+I think there is another narrative step missing about Figure 3C, which is that after you saw there were outliers you tried to identify which respondents had the extreme values.
+Again, please spell this out in the text.
+When you return to the plots for the entire dataset in Figure 5, please make parallel comments.}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{
+The next paragraph, which begins “The anomalies are also found” should perhaps be “Similar anomalies were found.” Was it the same respondents with the extremely high wages that showed extremely high number of hours of work?
+I suspect not, but please be explicit.}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Section 3.2.1 begins “As part of the IDA, which is the model formulation, we build a robust linear regression model to treat the extreme values in the data.” This could use clarification.
+In 3.2, you said that part of IDA is “ model formulation without any formal statistical inference.” To me, that sounds like using modeling as a descriptive technique, and it’s something I would consider to be more a part of EDA.
+But then in 3.2.1, while it is perhaps not using modeling in an inferential way, the model is being used for more than just description.
+I might rephrase the first sentence as follows: “In order to treat the extreme values in the data, we built a robust linear regression model.”}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{
+At the bottom of page 11, where you describe the model formation, more detail would be appreciated.
+You are modeling the mean hourly wage based on year.
+Why?
+Is this standard practice for robust linear regression?
+What is a “slight” residual?
+Is it defined based on absolute magnitude, or standard deviations, or something else?
+Later, you say “To minimize the risk of mistakenly identifying an outlier as an “erroneous outlier”.” Is an erroneous outlier a technical term?
+If not, I think you are trying not to identify erroneous outliers, not mistakenly identifying outliers as erroneous outliers (that’s a double negative).
+The value of 0.12 feels very specific and out of place in a paper that has had few numerals.
+Is this a number an expert in RLR would be able to understand?
+Again, is that a raw value?
+Is it a standard deviation?
+Is there literature about how to choose a threshold?
+Overall, the sentence “We find that 0.12 is the most reasonable value to be the threshold to minimize that drawback’s risk because it still captures the sensible spikes in the data.” is vague and should be expanded upon.
+What is “that drawback’s risk”?
+What are “the sensible spikes in the data?” Be more explicit.}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{
+What does “It implies that the fluctuation can still be observed in the data after the treatment.” mean?
+I suspect this sentence should be replaced with something along the lines of “The figure shows that fluctuations in wages still exist even in the imputed data.”}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{In Section 4, you talk about the variables in the original data.
+“One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection.” I’m not sure what this means.
+Please clarify.}
+\end{description}
+
+* We removed this sentence as we already have the `yr_workforce` variable which is calculated from the year of the individual start working.
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{“The treatment of unlikely wages differ in the refreshed data.” I think this means “The treatment of anomalous wages differs between the original and refreshed data sets.” Then you say “We opted to use the weights from a robust linear regression to determine what should be set as missing value as described in Section 3.2.1.” I don’t think you were setting missing values though, were you?
+It was used to determine which values should be imputed, correct?
+Either way, please clarify.}
+\end{description}
+
+* Yes, you are correct. We imputed the extreme wages with its predicted values instead of setting it as missing values. We have clarified this in the paper.
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{The bulleted difficulties in Section 5 are a bit jarring.
+Make them parallel (i.e.
+every bullet should start with a verb -ing so it is determining, calculating, etc) or turn them into a paragraph.
+I liked the specificity of \$30,000/hour but it would be better up in the section where you discuss the anomalous values.}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{At the end of page 14, it would be good to explicitly say that if people don’t like your decisions, they can grab the code and adjust the parameters themselves!
+It’s implied, but should be made explicit.}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Figure 2’s caption should include something about how that data is the raw values from the survey, before inputing anomalous values so the figure and caption can stand alone.}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{For both Figure 3 and Figure 5, I would prefer the letter label before the description, but that’s a matter of personal preference.}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{For Figure 7, why did you choose to show the entire dataset here? It would be easier to make a comparison if the plot was cut off at 1994 on the x-axis.}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Table 1 includes the note ?CHECK which I assume was a note to self.}
+\end{description}
+
+* We have removed it from the caption.
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Table 2 seems to have more precision than is necessary.
+Certainly 100\% does not need to be reported as 100.00\%, and probably all the percentages could be rounded to the nearest 1 percent.
+The caption could explain this, as there might be some rounding errors.}
+\end{description}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{What does “yearly mean hourly wages with years of workforce experience” mean? Is that one variable or two?}
+\end{description}
+
+* They are two different variables. We have separated it in the revised version.
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Was your goal to collect data from 1994-2018 or 1979-2018?
+Page 4 states 1994 onward, but from the figures later it appears that you collected data starting in 1979 (which makes more sense to me).}
+\end{description}
+
+* The original data covers 1979-1994 period, while in the refreshed data, we extended the period covered until 2018 (so that the data is from 1979-2018). We have made this explicit in Section 2.2.
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{I was confused by the following section: “We choose to use this revised May data because it seemed to have less missing and presumably has been checked.
+However, there is no revised May data for 2012, 2014, 2016, and 2018, thus, we use the ordinary May data for these years.” I was assuming the data was collected on a yearly basis, but this section suggests it was collected each month.
+Please clarify somewhere in the paper.}
+\end{description}
+
+* Yes, the data is collected on a yearly basis.
+However, according to the database, the highest grade completed variable is revised as in May 1st of each survey year.
+Thus, we mentioned as the revised May data.
+In the revised version, we only mention it as revised data to avoid confusion.

### 2022-02-08 paper/paper.tex
tweaking response

--- 2022-02-03 paper/paper.tex
@@
 2019).
-The data is first organised into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
-(Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{yowie} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
+The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
+(Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
 
@@
 
-The raw data, \texttt{new\_data\_qnames}, is organised such that each row corresponds to an individual.
+According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
+The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organised such that each row corresponds to an individual.
 As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year.
-The raw data consequently has a large number of columns (686 to be specific).
+The raw data consequently has a large number of columns (716 to be specific).
 The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars.
@@
 
-According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
-The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
-This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
-2019), \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham et al.
-2020), and \texttt{stringr} (Wickham 2019) for creating new variables, and levels of factors by text wrangling.
+Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
+2019), i.e., \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham et al.
+2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and create levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 The highest grade completed are recorded in \texttt{new\_data\_qnames} as variables beginning with \texttt{Q3-4} and \texttt{HGC} with suffix of the year it was recorded.
-In addition the variables beginning with \texttt{HGCREV} contain the revised data.
-We choose to use this revised May data because it seemed to have less missing and presumably has been checked.
-However, there is no revised May data for 2012, 2014, 2016, and 2018, thus, we use the ordinary May data for these years.
+In addition, the variables beginning with \texttt{HGCREV} contain the revised data.
+We choose to use this revised data because it seemed to have less missing and presumably has been checked.
+However, there is no revised data for 2012, 2014, 2016, and 2018, thus, we use the ordinary data for these years.
 
@@
 
-The number of jobs (\texttt{number\_of\_jobs}) for each respondent per year is computed from the number of non-missing values of hourly wage \emph{and} hours worked.
-If either the hourly wage or hours worked is missing, we do not tally this.
+The number of jobs (\texttt{number\_of\_jobs}) for each respondent per year is computed from the number of non-missing values of hourly wage.
+In other words, even if the information of hours worked exists for a particular observation, we do not tally when the hourly wage is missing.
 
@@
 There are two aspects of the original data that make a direct comparison difficult.
-The time variable provided is ``experience in the workforce''.
+The first one is that The time variable provided is ``experience in the workforce''.
 It is not clear how this is calculated.
 One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection.
-We have not been able to find this information in the database.
 Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.
+
+\begin{center}\includegraphics[width=468px]{figures/compare_xp_sw-1} \end{center}
 
@@
 However, in recent context, this value is too low to be set as the maximum threshold.
-We opted to use the weights from a robust linear regression to determine what should be set as missing value as described in Section \ref{censor}.
+We opted to use the weights from a robust linear regression to determine what should be regarded as outliers and imputed them with their predicted values as as described in Section \ref{censor}.
 

### 2022-02-08 response/response1.Rmd
tweaking response

@@
 We thank the Editor and the three reviewers for their comments, which have helped substantially to improve the manuscript.
-In response, we have made considerable changes to the main manuscript as outlined below.
+In response, we have made a major revision of the main manuscript as outlined below.
 
@@
 
-Reviewers suggested reframing the problem space so there is more emphasis on the case study.
+The editor and reviewers suggest reframing so that readers can help their students to experience the journal from wild to textbook data and either emphasize the case study, for the paper to be a regular submission for JSDSE, or to better explain the relevance for  teachers and/or students in statistics and data science to be considered for the special issue.
+We have chosen to do the latter, because we hope that the paper fits into the goals of the special issue.
 

### 2022-02-08 paper/paper.Rmd
abstract and title revised

@@
 ---
-title: "The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth"
+title: "A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database"
 authors:
@@
 abstract: > 
- The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data.
-Subsets of this data can be found in numerous textbooks and research articles.
-However, the steps and decisions taken to get from the raw data to the textbook data are never clearly articulated.
-This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly.
-Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research.
-Three new data sets and the code to produce them are provided in an accompanying open source R package, called [CENSORED].
-As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks and providing more convenient samples of the data to potential users.
+ Textbook data is useful for teaching statistics and data science methods because it is highly processed and clean.
+Where possible textbook data sets should be refreshed regularly, because it is also important to use contemporary data for teaching.
+This paper describes the trials and tribulations of refreshing the wages data extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s.
+The data is useful for teaching modeling and exploratory analysis of longitudinal data.
+Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles.
+The NLSY79 database has been continuously updated through to 2021, so new records are available.
+Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future.
+The process was made more difficult because the steps and decisions taken to get from the raw data to the original textbook data were not clearly articulated, so we have been diligent to provide a reproducible workflow for others to follow.
+Three new data sets and the code to produce them are provided in the open source R package, called [CENSORED].
 ---

### 2022-02-08 paper/paper.tex
abstract and title revised

@@
 
-\title{The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth}
+\title{A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database}
 
@@
 \begin{abstract}
-The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data.
-Subsets of this data can be found in numerous textbooks and research articles.
-However, the steps and decisions taken to get from the raw data to the textbook data are never clearly articulated.
-This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly.
-Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research.
-Three new data sets and the code to produce them are provided in an accompanying open source R package, called {[}CENSORED{]}.
-As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks and providing more convenient samples of the data to potential users.
+Textbook data is useful for teaching statistics and data science methods because it is highly processed and clean.
+Where possible textbook data sets should be refreshed regularly, because it is also important to use contemporary data for teaching.
+This paper describes the trials and tribulations of refreshing the wages data extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s.
+The data is useful for teaching modeling and exploratory analysis of longitudinal data.
+Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles.
+The NLSY79 database has been continuously updated through to 2021, so new records are available.
+Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future.
+The process was made more difficult because the steps and decisions taken to get from the raw data to the original textbook data were not clearly articulated, so we have been diligent to provide a reproducible workflow for others to follow.
+Three new data sets and the code to produce them are provided in the open source R package, called {[}CENSORED{]}.
 \end{abstract}

### 2022-02-08 paper/paper.Rmd
revised up to database section

@@
 Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles.
-The NLSY79 database has been continuously updated through to 2021, so new records are available.
+The NLSY79 database has been continuously updated through to 2018, so new records are available.
 Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future.
@@
 If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
-[ET: a related article: https://www.smh.com.au/national/by-the-numbers--the-average-australian-doesnt-exist--not-a-single-one-of-us-is-normal-20150124-12xlh0.html]
+A similar tone is echoed occasionally in the public media, for example, an article published in the Sydney Morning Herald argues that there is no average Australian [@notaverage].
 
-As a textbook data set though, it is outdated.
+As a textbook data set though, the wages data is outdated.
 The most recent year in the data is 1994, 10 years prior to when @SingerJudithD2003Alda was published.
-Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students.
+Teachers of statistics need use contemporary data sets to show how techniques are relevant for today's students.
 Using tired old textbook data sets can imbue a misconception that the field is not current.
@@
 The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979.
-It comprised of people sampled because they were Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military.
-In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members who represents economically disadvantage people, but were not Black nor Hispanics, respectively.
-These subsamples were dropped because they are not eligible anymore for interview [@nlsy79].
-Hence, 9,964 respondents remain in the eligible samples.
-The surveys were conducted annually from 1979 to 1994 and biennially thereafter.
+For a variety of reasons, some structural, the number of respondents dropped to 9,964 after 1990.
+<!--It is comprised of people sampled because they were black, hispanic, economically disadvantaged non-black non-hispanics, and youth in the military.
+In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members who represents economically disadvantaged people, but were not black nor hispanics, respectively.
+These subsamples were dropped because they were no longer eligible for interviews [@nlsy79].
+Hence 9,964 respondents remain in the eligible samples.--> The surveys were conducted annually from 1979 to 1994 and biennially thereafter.
 Data are currently available from Round 1 (1979 survey year) to Round 28 (2018 survey year).

### 2022-02-08 paper/paper.tex
revised up to database section

@@
 If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
-{[}ET: a related article: \url{https://www.smh.com.au/national/by-the-numbers--the-average-australian-doesnt-exist--not-a-single-one-of-us-is-normal-20150124-12xlh0.html}{]}
+A similar tone is echoed occasionally in the public media, for example, an article published in the Sydney Morning Herald argues that there is no average Australian (Moncrief 2015).
 
-As a textbook data set though, it is outdated.
+As a textbook data set though, the wages data is outdated.
 The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
-Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students.
+Teachers of statistics need use contemporary data sets to show how techniques are relevant for today's students.
 Using tired old textbook data sets can imbue a misconception that the field is not current.
@@
 The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979.
-It comprised of people sampled because they were Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military.
-In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members who represents economically disadvantage people, but were not Black nor Hispanics, respectively.
-These subsamples were dropped because they are not eligible anymore for interview (Bureau of Labor Statistics, U.S.
-Department of Labor 2021).
-Hence, 9,964 respondents remain in the eligible samples.
+For a variety of reasons, some structural, the number of respondents dropped to 9,964 after 1990.
 The surveys were conducted annually from 1979 to 1994 and biennially thereafter.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-notaverage}{}}%
+Moncrief, Marc. 2015. {``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.''} \url{https://bit.ly/smh-not-normal}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-opendata}{}}%

### 2022-02-08 paper/paper.Rmd
revised target data section

@@
 
-The NLSY79 data used in @SingerJudithD2003Alda contains the longitudinal measurements of the male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994.
-This dataset comprises of several variables as follows:
+The NLSY79 data used in @SingerJudithD2003Alda contains the longitudinal records of male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994.
+This dataset contains several variables as follows:
 
@@
 3. LNW: natural logarithm of wages, adjusted with 1990's inflation rate. 
-4. BLACK: binary variables, the respondents are coded to 1 because they were Black and 0 otherwise. 
-5. HGC: the highest grade completed.
-6. UERATE: unemployment rate of the year of survey. When missing, the variable is set to be 7.875 (the average rate).
+4. BLACK: binary variable, 1 indicates black and 0 otherwise. 
+5. HISPANIC: binary variable, 1 indicates hispanic and 0 otherwise. 
+6. HGC: the highest grade completed.
+7. UERATE: unemployment rate of the year of survey. When missing, the variable is set to be 7.875 (the average rate).
 
-From the database, we will refresh this data with a more up to date period, which is from survey year 1979 to survey year 2018 (the most recent year published).
-We will  also modify the variables (for example, we create a race variable, instead of the binary variable "Black") and add some variables so that more examples of data exploration can be displayed in teaching.
-The variables added are year of survey, age of individual in 1979, year when the highest grade completed, number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual starting to work, and the number of years the individual worked.
+From the database, we will refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published).
+We also plan to make some modifications of the variables.
+For example, we create a categorical race variable instead of the binary variable "black" and "hispanic".
+We also include additional variables so that more interesting data exploration can be included in teaching examples: year of survey, age of individual in 1979, year when the highest grade was completed, number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
 
-In addition to the dropouts data, we will also make the up to high school graduates' wages available, including the females, and the demographic data of the NLSY79 cohort available in separate datasets.
-Thus, we will create three datasets as follows:
+We will also make all of the high school graduates data available, including females, in addition to high school dropouts, and separate the demographic data of the NLSY79 cohort.
+Thus, the plan is to create three datasets as follows:
 
@@
 2. The wages data of the up to high school graduates.
-3. The wages data of the high school dropouts, which is the refreshed version of @SingerJudithD2003Alda's data.
+3. The wages data of the high school dropouts, which is closest to a refreshed version of @SingerJudithD2003Alda's data.
 

### 2022-02-08 paper/paper.tex
revised target data section

@@
 
-The NLSY79 data used in Singer and Willett (2003) contains the longitudinal measurements of the male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994.
-This dataset comprises of several variables as follows:
+The NLSY79 data used in Singer and Willett (2003) contains the longitudinal records of male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994.
+This dataset contains several variables as follows:
 
@@
 \item
-  BLACK: binary variables, the respondents are coded to 1 because they were Black and 0 otherwise.
+  BLACK: binary variable, 1 indicates black and 0 otherwise.
+\item
+  HISPANIC: binary variable, 1 indicates hispanic and 0 otherwise.
 \item
@@
 
-From the database, we will refresh this data with a more up to date period, which is from survey year 1979 to survey year 2018 (the most recent year published).
-We will also modify the variables (for example, we create a race variable, instead of the binary variable ``Black'') and add some variables so that more examples of data exploration can be displayed in teaching.
-The variables added are year of survey, age of individual in 1979, year when the highest grade completed, number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual starting to work, and the number of years the individual worked.
+From the database, we will refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published).
+We also plan to make some modifications of the variables.
+For example, we create a categorical race variable instead of the binary variable ``black'' and ``hispanic''.
+We also include additional variables so that more interesting data exploration can be included in teaching examples: year of survey, age of individual in 1979, year when the highest grade was completed, number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
 
-In addition to the dropouts data, we will also make the up to high school graduates' wages available, including the females, and the demographic data of the NLSY79 cohort available in separate datasets.
-Thus, we will create three datasets as follows:
+We will also make all of the high school graduates data available, including females, in addition to high school dropouts, and separate the demographic data of the NLSY79 cohort.
+Thus, the plan is to create three datasets as follows:
 
@@
 \item
-  The wages data of the high school dropouts, which is the refreshed version of Singer and Willett (2003)'s data.
+  The wages data of the high school dropouts, which is closest to a refreshed version of Singer and Willett (2003)'s data.
 \end{enumerate}

### 2022-02-08 paper/paper.Rmd
addressing the first three of the reviewers comments

@@
 They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
-Data cleaning can be considered to be a part of what is called "initial data analysis" (IDA) [@Chatfield1985TIEo].
-In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
-This is also related to exploratory data analysis (EDA), coined by @tukey with a focus on learning from data.
+Clean data often then goes through an "initial data analysis" (IDA) [@Chatfield1985TIEo], where one would  summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling.
+This stage is related to exploratory data analysis (EDA), coined by @tukey with a focus on learning from data.
 EDA can be considered to encompass IDA.
-@DasuTamraparni2003Edma say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80% of the data mining task.
+In practice the three stages of cleaning, summarizing and exploring are cyclical, that one often needs to do more cleaning after scrutinizing.
+@DasuTamraparni2003Edma say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.
 

### 2022-02-08 paper/paper.tex
addressing the first three of the reviewers comments

@@
 They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
-Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985).
-In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
-This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
+Clean data often then goes through an ``initial data analysis'' (IDA) (Chatfield 1985), where one would summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling.
+This stage is related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
 EDA can be considered to encompass IDA.
-Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.
+In practice the three stages of cleaning, summarizing and exploring are cyclical, that one often needs to do more cleaning after scrutinizing.
+Dasu and Johnson (2003) say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.
 

### 2022-02-08 response/response1.Rmd
abstract and title revised
addressing the first three of the reviewers comments

@@
 \item[Reviewer 1:] \textcolor{violet}{How is this manuscript relevant for teachers and/or students in statistics and data science?}
+
+The introduction has been re-written with an emphasis on how this work should be interesting for teachers and students of data science and statistics.
+
 \item[Reviewer 3:] \textcolor{violet}{The authors prominently cite (Chatfield 1985) for the definition of the term “IDA”.
@@
 It seems curious to me that the whole paper is framed as being “IDA” based on the author’s unsupported statement that data cleaning should be considered party of IDA.}
+
+Yes, there is grey area in these three activities.
+Thanks for pointing out Chatfield's original viewpoint, which differs from Huebner et al's.
+We have revised this paragraph to de-emphasize cleaning as being part of IDA.
+We have kept the explanation of the terminology of IDA and EDA because we think this is important for clarifying for teaching purposes.
+
 \item[Reviewer 3:] \textcolor{violet}{My suggestion is that the authors reconfigure this paper to be presented as a pure case study.
-Ibelieve the work that has been done is valuable: the data cleaning process is often messy andunplanned, so a case study like this on a popular dataset would provide an excellent educational resource.}
+I believe the work that has been done is valuable: the data cleaning process is often messy and unplanned, so a case study like this on a popular dataset would provide an excellent educational resource.}
+
+We have kept the case study frame of mind but have re-cast the paper so that it focuses on the importance of this type of activity for teaching of statistics and data science. 
 \end{description}
-
-Supporting information for the lack of data cleaning in papers.
 

### 2022-02-08 paper/paper.Rmd
reconciling the statistical value chain with wild and textbook data terms

@@
 
-@LooMarkvander2018Sdcw describe the notion of a "statistical value chain" where the production stages of the data cleaning process are earmarked as raw data (data as arrived to the desk of an analyst), input data (data organised with correct type and identified variables) and valid data (data that faithfully represent the variables).
+@LooMarkvander2018Sdcw in the context of official statistics describe the "statistical value chain" which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organised with correct type and identified variables) and valid data (data that has been cleaned and more accurately represents the intent of variables).
+What we have colorfully named as wild data can be considered to be raw data, and valid data could be considered to be textbook data, in the above statistical value chain.
 In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidy the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so that the resulting input data can be used downstream for validating the data as described in Section \@ref(ida).

### 2022-02-08 paper/paper.tex
reconciling the statistical value chain with wild and textbook data terms

@@
 M.
-van der Loo and de Jonge (2018) describe the notion of a ``statistical value chain'' where the production stages of the data cleaning process are earmarked as raw data (data as arrived to the desk of an analyst), input data (data organised with correct type and identified variables) and valid data (data that faithfully represent the variables).
+van der Loo and de Jonge (2018) in the context of official statistics describe the ``statistical value chain'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organised with correct type and identified variables) and valid data (data that has been cleaned and more accurately represents the intent of variables).
+What we have colorfully named as wild data can be considered to be raw data, and valid data could be considered to be textbook data, in the above statistical value chain.
 In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida}.

### 2022-02-08 response/response1.Rmd
reconciling the statistical value chain with wild and textbook data terms

@@
 We have revised this paragraph to de-emphasize cleaning as being part of IDA.
-We have kept the explanation of the terminology of IDA and EDA because we think this is important for clarifying for teaching purposes.
+We have kept the explanation of the terminology of IDA and EDA because we think this is important to clarify for teaching purposes.
 
@@
 
-Outlier detection
+XXX We need to review this literature. It's not true that we have ignored it, the main reference is Dasu! But we need to decide whether there is more that we shoudl cite.
+
+Outlier detection ???
 
@@
 
-* We have fixed the sentences and the reference.
+* We have fixed the sentences and the reference. XXX Which one?
 

### 2022-02-08 paper/paper.Rmd
steps taken to get raw data

@@
 
-The NLSY79 data contains a large number of variables but for our aim, the scope required is limited to demographic profiles, wages data, and work experience.
+The NLSY79 data contains a large number of variables but for our purposes the scope required is limited to demographic profiles, wages data, and work experience.
 More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data and navigated as described in Figure \ref{fig:source-nav}.
@@
 \end{tcolorbox}
-\caption{The above documents the steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
+\caption{Documented steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
 \end{figure}
@@
 
-Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the `tidyverse` suite of packages [@tidyverse], i.e., `tidyr` [@tidyr] to pivot the data into long form, `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and create levels of factors by text wrangling.
+Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the `tidyverse` suite of packages [@tidyverse]:`tidyr` [@tidyr] to pivot the data into long form, with `dplyr` [@dplyr] and `stringr` [@stringr] to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.

### 2022-02-08 paper/paper.tex
steps taken to get raw data

@@
 Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
-2019), i.e., \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham et al.
-2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and create levels of factors by text wrangling.
+2019):\texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham et al.
+2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.

### 2022-02-08 paper/paper.Rmd
revised IDA section

@@
 
-According to @HuebnerMariannePhD2016Asat, initial data analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis.
-Moreover, @Chatfield1985TIEo argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
+According to @HuebnerMariannePhD2016Asat, initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure that the data is clean, valid, and ready to be deployed in the later analyses.
+This is supported by @Chatfield1985TIEo who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
 
-In this paper, we conduct an IDA or a preliminary data analysis to assess the consistency of the data with the cohort information that the NLSY provides.
+In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides.
 <!-- AE comments: This comes as a surprise to the reader.
@@
 The wage anomalies should be mentioned in the introduction so as not to be a surprise at this point in the manuscript.
- -->In addition, we also aim to find the anomaly in the wages values using this approach. We mainly use graphical summaries to do the IDA using `ggplot2` [@ggplot2] and `brolgar` [@brolgar].
+ --> A first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
+This is followed with graphical summaries using methods available in `ggplot2` [@ggplot2] and `brolgar` [@brolgar].
 
-As stated previously, the respondents' ages ranged from 12 to 22 when first interviewed in 1979.
+The respondents' ages ranged from 12 to 22 when first interviewed in 1979.
 Hence, we validate whether all of the respondents were in this range in the data we extracted.

### 2022-02-08 paper/paper.tex
revised IDA section

@@
 
-According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis.
-Moreover, Chatfield (1985) argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
+According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure that the data is clean, valid, and ready to be deployed in the later analyses.
+This is supported by Chatfield (1985) who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
 
-In this paper, we conduct an IDA or a preliminary data analysis to assess the consistency of the data with the cohort information that the NLSY provides.
-In addition, we also aim to find the anomaly in the wages values using this approach.
-We mainly use graphical summaries to do the IDA using \texttt{ggplot2} (Wickham 2016) and \texttt{brolgar} (Tierney, Cook, and Prvan 2020).
+In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides.
+A first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
+This is followed with graphical summaries using methods available in \texttt{ggplot2} (Wickham 2016) and \texttt{brolgar} (Tierney, Cook, and Prvan 2020).
 
-As stated previously, the respondents' ages ranged from 12 to 22 when first interviewed in 1979.
+The respondents' ages ranged from 12 to 22 when first interviewed in 1979.
 Hence, we validate whether all of the respondents were in this range in the data we extracted.

### 2022-02-08 data-raw/ida.R
modify table 2 (check required on my other computer to see if it compiles as expected)

@@ -30,9 +30,9 @@ freq_formatting <- function(x) {
 
 gender_race_table <- full_demographics %>%
   tabyl(gender, race) %>%
-  adorn_totals(c("row", "col")) %>%
   adorn_percentages("row") %>%
-  adorn_pct_formatting(digits = 2) %>%
+  adorn_totals(c("row", "col")) %>%
+  adorn_pct_formatting(digits = 0) %>%
   adorn_ns(., position = "front", ns = freq_formatting(.)) %>%
   mutate(gender = str_to_title(gender))
 

### 2022-02-08 response/response1.Rmd
minor changes to response letter

@@
 
-* We have corrected all minor grammatical errors pointed out by the reviewers.  
-* We have made the descriptions of the steps more explicit.
+* We have corrected all minor grammatical errors pointed out by the reviewers as outlined below. [ADD the changes]  
+* We have made the descriptions of the steps more explicit as follows [LIST OF THE CHANGES]
 
@@
 
-* We have made it explicit.
+* We have made the file format more explicit.
 
@@
 Then, when starting to describe the raw data in 3,1, explain the same pieces and explicitly explain why the raw data it is not tidy to begin with.}
-\end{description}
-
-\begin{description}
-\item[Reviewer 2:]\textcolor{violet}{Before the data cleaning section I would love a glimpse into the future to see what the target dataset would look like.
-What are the observations?
-What are the variables?
-This would allow me to follow the narrative of data wrangling more easily.
-Perhaps this could be put into section 2.2, which is titled Target data.
-Then, when starting to describe the raw data in 3.1, explain the same pieces and explicitly explain why the raw data it is not tidy to begin with.}
 \end{description}
@@
 
+We agree that this level of precision was not necessary. We have modified Table 2 in line with this.
+
 \begin{description}
@@
 
-* They are two different variables. We have separated it in the revised version.
+* These are two different variables. We have made this easier to follow in the manuscript as: [FILL]
 

### 2022-02-08 paper/paper.Rmd
placement of tables
width of printed data on page 5 fixed

@@
 The values in the cell under the variables that begin with `HRP` correspond to the hourly wage in dollars.
-A glimpse of this data output is shown below.
+A glimpse of this data shows:
 
@@
 
-Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the `tidyverse` suite of packages [@tidyverse]:`tidyr` [@tidyr] to pivot the data into long form, with `dplyr` [@dplyr] and `stringr` [@stringr] to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
+Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long form, with `dplyr` [@dplyr] and `stringr` [@stringr] to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 
-```{r age-table}
-```
-
-<!--AE comment: Table 2 seems to have more precision than is necessary.
-Certainly 100% does not need to be reported as 100.00%, and probably all the percentages could be rounded to the nearest 1 percent.
-The caption could explain this, as there might be some rounding errors.
--->
-
-```{r gender-race-table}
-```
-
 In the next step, we explore the mean hourly wage data.
@@
 We employ visualization techniques to perform the IDA as described next.
-
-```{r summarytable, eval=FALSE}
-```
-
-<!--AE comment: Figure 2’s caption should include something about how that data is the raw values from the survey, before inputing anomalous values so the figure and caption can stand alone.-->
 
@@
 Some of them, for example respondents in Figure \ref {fig:feature-plot} C had experienced unusual wages only in certain year.
+
+```{r age-table}
+```
+
+<!--AE comment: Table 2 seems to have more precision than is necessary.
+Certainly 100% does not need to be reported as 100.00%, and probably all the percentages could be rounded to the nearest 1 percent.
+The caption could explain this, as there might be some rounding errors.
+-->
+
+```{r gender-race-table}
+```
+
+```{r summarytable, eval=FALSE}
+```
+
+<!--AE comment: Figure 2’s caption should include something about how that data is the raw values from the survey, before inputing anomalous values so the figure and caption can stand alone.-->
 

### 2022-02-08 paper/paper.tex
placement of tables
width of printed data on page 5 fixed

@@
 The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars.
-A glimpse of this data output is shown below.
+A glimpse of this data shows:
 
@@
 #> 'data.frame':    12686 obs. of  716 variables:
-#>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 9 10 ...
-#>  $ HRP1_1979                                 : int  328 385 365 NA 310 NA NA NA 214 NA ...
-#>  $ HRP2_1979                                 : int  NA NA NA NA 375 NA NA NA NA NA ...
-#>  $ HRP3_1979                                 : int  NA NA 275 NA NA NA NA NA NA NA ...
-#>  $ HRP4_1979                                 : int  NA NA NA NA NA 250 NA NA NA NA ...
-#>  $ HRP5_1979                                 : int  NA NA NA NA NA NA NA NA NA NA ...
-#>  $ HRP1_1980                                 : int  NA 457 397 NA 333 275 300 394 200 318 ...
-#>  $ HRP2_1980                                 : int  NA NA 367 NA NA NA NA NA NA NA ...
-#>  $ HRP3_1980                                 : int  NA NA 380 NA NA NA 290 NA NA NA ...
-#>  $ HRP4_1980                                 : int  NA NA NA NA NA NA NA NA NA NA ...
+#>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 ...
+#>  $ HRP1_1979                                 : int  328 385 365 NA 310 NA NA NA ...
+#>  $ HRP2_1979                                 : int  NA NA NA NA 375 NA NA NA ...
+#>  $ HRP3_1979                                 : int  NA NA 275 NA NA NA NA NA ...
+#>  $ HRP4_1979                                 : int  NA NA NA NA NA 250 NA NA ...
+#>  $ HRP5_1979                                 : int  NA NA NA NA NA NA NA NA ...
+#>  $ HRP1_1980                                 : int  NA 457 397 NA 333 275 300 394 ...
+#>  $ HRP2_1980                                 : int  NA NA 367 NA NA NA NA NA ...
+#>  $ HRP3_1980                                 : int  NA NA 380 NA NA NA 290 NA ...
+#>  $ HRP4_1980                                 : int  NA NA NA NA NA NA NA NA ...
 #>   [list output truncated]
@@
 Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
-2019):\texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham et al.
+2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham et al.
 2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
@@
 
+In the next step, we explore the mean hourly wage data.
+In this case, we only explore the wages data of respondents that have the highest completed grade of up to \(12^{th}\) grade.
+We employ visualization techniques to perform the IDA as described next.
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{figures/sample-plot-1}
+
+}
+
+\caption{Longitudinal profiles of wages for a random sample of 20 individuals in the refreshed data.
+Most, but not all, individuals here experienced increasing wages over time, and several have experienced considerable fluctuation in wages.
+Some individuals are only measured for a short period.}\label{fig:sample-plot}
+\end{figure}
+
+We randomly take 20 samples from the data and plot them, as shown in Figure \ref{fig:sample-plot}.
+It shows that these respondents have a lot of variability in wages, for example, the IDs in panel numbers 5, 7, and 11.
+The plot also implies that the samples have a different pattern of mean hourly wages.
+Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.
+However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages.
+Some of them, for example respondents in Figure \ref {fig:feature-plot} C had experienced unusual wages only in certain year.
+
 \begin{table}
@@
 \midrule
-Male & 1,000 (15.62\%) & 1,613 (25.19\%) & 3,790 (59.19\%) & 6,403 (100.00\%)\\
-Female & 1,002 (15.95\%) & 1,561 (24.84\%) & 3,720 (59.21\%) & 6,283 (100.00\%)\\
+Male & 1,000 (16\%) & 1,613 (25\%) & 3,790  (59\%) & 6,403 (100\%)\\
+Female & 1,002 (16\%) & 1,561 (25\%) & 3,720  (59\%) & 6,283 (100\%)\\
 \midrule
-Total & 2,002 (15.78\%) & 3,174 (25.02\%) & 7,510 (59.20\%) & 12,686 (100.00\%)\\
+Total & 2,002 (32\%) & 3,174 (50\%) & 7,510 (118\%) & 12,686 (200\%)\\
 \bottomrule
@@
 \end{table}
-
-In the next step, we explore the mean hourly wage data.
-In this case, we only explore the wages data of respondents that have the highest completed grade of up to \(12^{th}\) grade.
-We employ visualization techniques to perform the IDA as described next.
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{figures/sample-plot-1}
-
-}
-
-\caption{Longitudinal profiles of wages for a random sample of 20 individuals in the refreshed data.
-Most, but not all, individuals here experienced increasing wages over time, and several have experienced considerable fluctuation in wages.
-Some individuals are only measured for a short period.}\label{fig:sample-plot}
-\end{figure}
-
-We randomly take 20 samples from the data and plot them, as shown in Figure \ref{fig:sample-plot}.
-It shows that these respondents have a lot of variability in wages, for example, the IDs in panel numbers 5, 7, and 11.
-The plot also implies that the samples have a different pattern of mean hourly wages.
-Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.
-However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages.
-Some of them, for example respondents in Figure \ref {fig:feature-plot} C had experienced unusual wages only in certain year.
 

### 2022-02-08 data-raw/data-preprocessing.R

@@ -8,13 +8,12 @@ select <- dplyr::select
 source(here::here("data-raw/NLSY79/NLSY79.R"))
 
 ## ---- untidy-data
-options(width=70)
 new_data_qnames %>%
   select(CASEID_1979,
          starts_with("HRP") &
            ends_with(c("1979", "1980")),
          everything()) %>%
-  str(list.len = 10)
+  str(list.len = 10, vec.len = 3)
 
 
 ## ---- dob-tidy

### 2022-02-08 paper/paper.Rmd
extending conclusions

@@
 - there are many errors in the data, e.g. hourly wages greater than $30,000 per hour.
-- there is no explicit variable in the database recording high school dropout, which means we needed to compare date of 12th grade with the individual's age.
-- calculating experience in the workforce would require knowing the time of first job within the first year the individual was recorded.
+- there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with the individual's age.
+<!-- calculating experience in the workforce would require knowing the time of first job within the first year the individual was recorded.-->
 
@@
 Users need to be able to trust the data.
+
+Why is this exercise important for teachers and students?
+The data package and software could be used for illustrating data cleaning in the classroom.
+The data provided in the package can be used as examples i nthe teaching of initial data analysis, and modeling, and exploratory data analysis of longitudinal data.
+
+XXX Importance of public policy vs for the public; and the importance of using relatable data examples in teaching that it needs to be acurate, and limitations on interpretation need to be available.
 

### 2022-02-08 paper/paper.tex
extending conclusions

@@
 \item
-  there is no explicit variable in the database recording high school dropout, which means we needed to compare date of 12th grade with the individual's age.
-\item
-  calculating experience in the workforce would require knowing the time of first job within the first year the individual was recorded.
+  there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with the individual's age.
 \end{itemize}

### 2022-02-08 paper/paper.Rmd
refining conclusions

@@
 
-Finally, for the data providers we recommend that a validation system with clear rules is added on data entry, and that alternative output formats, such as a tidy format would help users make better use of the resource.
+For the data providers, we recommend that a validation system with clear rules is added on data entry, and that alternative output formats, such as a tidy format would help users make better use of the resource.
 The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs.
@@
 The data package and software could be used for illustrating data cleaning in the classroom.
-The data provided in the package can be used as examples i nthe teaching of initial data analysis, and modeling, and exploratory data analysis of longitudinal data.
+The refreshed and contemporary data provided in the package can be used as examples in the teaching of initial data analysis and modeling, and exploratory data analysis of longitudinal data.
 
-XXX Importance of public policy vs for the public; and the importance of using relatable data examples in teaching that it needs to be acurate, and limitations on interpretation need to be available.
+The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual.
+From modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns.
+That more education leads to increasingly higher wages, which is a satisfying result for educators.
+It means that education makes a difference in the wage experience, and that public policy that encourages education is a data-supported action.
+We would also learn, although not presented here, that race matters and that being black leads to lower wages.
+This is a disturbing finding because there is no rational for such a difference in a fair society.
+This is would provide support for action in public policy to remove this overall average effect.
+It also provides an example for educators to explain the use of data to support public policy action.
+
+These are interpretations of results from analyzing the wages data, which rely on trusting that the data provided is accurate and valid.
+The data is collected from a reputable organization, but the data has obvious errors that need correcting.
+This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.
+
+Having trustworthy data is imperative for statistics and data science education.
+Students do interpret results and take seriously what is reported from example analyses.
+If the data in the examples are synthetic, with our own inherent prejudices (on fog e.g.
+gender and race) applied, or inaccurately processed leaving errors, we as educators are being irresponsible because the societal message taught to students is flawed.
+This paper demonstrates the process to produce a trustworthy data set for teaching.
 

### 2022-02-08 paper/paper.tex
refining conclusions

@@
 The data package and software could be used for illustrating data cleaning in the classroom.
-The data provided in the package can be used as examples i nthe teaching of initial data analysis, and modeling, and exploratory data analysis of longitudinal data.
+The refreshed and contemporary data provided in the package can be used as examples in the teaching of initial data analysis and modeling, and exploratory data analysis of longitudinal data.
 
-XXX Importance of public policy vs for the public; and the importance of using relatable data examples in teaching that it needs to be acurate, and limitations on interpretation need to be available.
+The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual.
+From modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns.
+That more education leads to increasingly higher wages, which is a satisfying result for educators.
+It means that education makes a difference in the wage experience, and that public policy that encourages education is a data-supported action.
+We would also learn, although not presented here, that race matters and that being black leads to lower wages.
+This is a disturbing finding because there is no rational for such a difference in a fair society.
+This is would provide support for action in public policy to remove this overall average effect.
+It also provides an example for educators to explain the use of data to support public policy action.
+
+These are interpretations of results from analyzing the wages data, which rely on trusting that the data provided is accurate and valid.
+The data is collected from a reputable organization, but the data has obvious errors that need correcting.
+This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.
+
+Having trustworthy data is imperative for statistics and data science education.
+Students do interpret results and take seriously what is reported from example analyses.
+If the data in the examples are synthetic, with our own inherent prejudices (on fog e.g.~gender and race) applied, or inaccurately processed leaving errors, we as educators are being irresponsible because the societal message taught to students is flawed.
+This paper demonstrates the process to produce a trustworthy data set for teaching.
 

### 2022-02-09 paper/paper.Rmd
Add explanation of how to obtain the work experience-reated variables.

@@
 
-Our target variables for the employment are to obtain respondent's mean hourly wage (`mean_hourly_wage`), the number of jobs (`number_of_jobs`) and the total hours of work per week (`total_hours`) for each survey year.
+Our target variables for the employment are to obtain respondent's mean hourly wage (`wage`), the number of jobs (`njobs`), the total hours of work per week (`hours`) for each survey year, the year when individual starting to work (`stwork`), the length of time (years) in workforce (`yr_workforce`), and work experience measured as the number of years worked (`exp`).
 As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.
@@
 ```
+
+For `st_work` variable,  we only rename the column names from the raw, NLSY79 data.
+This variable is then used to calculate the next variable, `yr_workforce` for each survey round, which is the year of survey (`year`) minus the year of individual start working (`st_work`).
+Finally, `exp` variable is derived from the variable started with `WKSWK` in `new_data_qnames`.
+This variable measured since the last interview.
+Thus, to obtain the work experience since 1979, we calculate the cumulative value.
+As the measurement unit is in week, we convert this to year.
 

### 2022-02-09 data-raw/ida.R
fix Table 2

@@ -34,7 +34,8 @@ gender_race_table <- full_demographics %>%
   adorn_totals(c("row", "col")) %>%
   adorn_pct_formatting(digits = 0) %>%
   adorn_ns(., position = "front", ns = freq_formatting(.)) %>%
-  mutate(gender = str_to_title(gender))
+  mutate(gender = str_to_title(gender),
+         Total = str_replace(Total, "\\([1|2]00\\%\\)", ""))
 
 kable(gender_race_table,
       caption = "The contingency table for gender and race for the full NLSY79 data.",

### 2022-02-09 data-raw/ida.R
Table 2 fix, added to response

@@ -30,15 +30,15 @@ freq_formatting <- function(x) {
 
 gender_race_table <- full_demographics %>%
   tabyl(gender, race) %>%
-  adorn_percentages("row") %>%
   adorn_totals(c("row", "col")) %>%
+  adorn_percentages("row") %>%
   adorn_pct_formatting(digits = 0) %>%
   adorn_ns(., position = "front", ns = freq_formatting(.)) %>%
   mutate(gender = str_to_title(gender),
          Total = str_replace(Total, "\\([1|2]00\\%\\)", ""))
 
 kable(gender_race_table,
-      caption = "The contingency table for gender and race for the full NLSY79 data.",
+      caption = "The contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1%) is out of the total corresponding to row.",
       col.names = c("Gender", "Hispanic", "Black", "Non-Black, Non-Hispanic", "Total"),
       booktabs = TRUE,
       linesep = "",

### 2022-02-09 paper/paper-jsdse/paper-jsdse.tex
fixed issue for the CSL hang! It was because of change in pandoc

--- 2022-02-03 paper/paper-jsdse/paper-jsdse.tex
@@
 
-% need this for pandoc %
+% Pandoc citation processing
 \newlength{\cslhangindent}
@@
 \setlength{\csllabelwidth}{3em}
-\newenvironment{CSLReferences}[3] % #1 hanging-ident, #2 entry spacing
+\newlength{\cslentryspacingunit} % times entry-spacing
+\setlength{\cslentryspacingunit}{\parskip}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
  {% don't indent paragraphs
@@
   % turn on hanging indent if param 1 is 1
-  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  \ifodd #1
+  \let\oldpar\par
+  \def\par{\hangindent=\cslhangindent\oldpar}
+  \fi
   % set entry spacing
-  \ifnum #2 > 0
-  \setlength{\parskip}{#2\baselineskip}
-  \fi
+  \setlength{\parskip}{#2\cslentryspacingunit}
  }%
  {}
-\usepackage{calc} % for \widthof, \maxof
+\usepackage{calc}
 \newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\maxof{\widthof{#1}}{\csllabelwidth}}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth}{#1}}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
 \newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
@@
 {
-  \title{\bf The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth}
+  \title{\bf A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database}
   \author{
@@
   \begin{center}
-    {\LARGE\bf The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth}
+    {\LARGE\bf A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database}
 \end{center}
@@
 \begin{abstract}
-The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data.
-Subsets of this data can be found in numerous textbooks and research articles.
-However, the steps and decisions taken to get from the raw data to the textbook data are never clearly articulated.
-This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly.
-Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research.
-Three new data sets and the code to produce them are provided in an accompanying open source R package, called {[}CENSORED{]}.
-As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks and providing more convenient samples of the data to potential users.
+Textbook data is useful for teaching statistics and data science methods because it is highly processed and clean.
+Where possible textbook data sets should be refreshed regularly, because it is also important to use contemporary data for teaching.
+This paper describes the trials and tribulations of refreshing the wages data extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s.
+The data is useful for teaching modeling and exploratory analysis of longitudinal data.
+Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles.
+The NLSY79 database has been continuously updated through to 2018, so new records are available.
+Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future.
+The process was made more difficult because the steps and decisions taken to get from the raw data to the original textbook data were not clearly articulated, so we have been diligent to provide a reproducible workflow for others to follow.
+Three new data sets and the code to produce them are provided in the open source R package, called {[}CENSORED{]}.
 \end{abstract}
@@
 
-The NLSY79 data used in Singer and Willett (2003) contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994.
-In addition, the cohort is restricted to male high-school dropouts who first participated in the study at age 14-17 years.
-Thus the target data set is to collect the same variables for the extended time frame of 1994 through to 2018, the most recent year reported.
+The NLSY79 data used in Singer and Willett (2003) contains the longitudinal records of male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994.
+This dataset contains several variables as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  ID: the respondents' ID.
+\item
+  EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
+\item
+  LNW: natural logarithm of wages, adjusted with 1990's inflation rate.
+\item
+  BLACK: binary variable, 1 indicates black and 0 otherwise.
+\item
+  HISPANIC: binary variable, 1 indicates hispanic and 0 otherwise.
+\item
+  HGC: the highest grade completed.
+\item
+  UERATE: unemployment rate of the year of survey. When missing, the variable is set to be 7.875 (the average rate).
+\end{enumerate}
+
+From the database, we will refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published).
+We also plan to make some modifications of the variables.
+For example, we create a categorical race variable instead of the binary variable ``black'' and ``hispanic''.
+We also include additional variables so that more interesting data exploration can be included in teaching examples: year of survey, age of individual in 1979, year when the highest grade was completed, number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
+
+We will also make all of the high school graduates data available, including females, in addition to high school dropouts, and separate the demographic data of the NLSY79 cohort.
+Thus, the plan is to create three datasets as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The demographic data of the NLSY79 cohort.
+\item
+  The wages data of the up to high school graduates.
+\item
+  The wages data of the high school dropouts, which is closest to a refreshed version of Singer and Willett (2003)'s data.
+\end{enumerate}
 
@@
 
-The raw data, \texttt{new\_data\_qnames}, is organised such that each row corresponds to an individual.
+According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
+The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organised such that each row corresponds to an individual.
 As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year.
-The raw data consequently has a large number of columns (686 to be specific).
+The raw data consequently has a large number of columns (716 to be specific).
 The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars.
-A glimpse of this data output is shown below.
+A glimpse of this data shows:
 
@@
 
-According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
-The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
-This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
-2019), \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham et al.
-2020), and \texttt{stringr} (Wickham 2019) for creating new variables, and levels of factors by text wrangling.
+Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
+2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham et al.
+2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 
-Our target variables for the employment are to obtain respondent's mean hourly wage (\texttt{mean\_hourly\_wage}), the number of jobs (\texttt{number\_of\_jobs}) and the total hours of work per week (\texttt{total\_hours}) for each survey year.
+Our target variables for the employment are to obtain respondent's mean hourly wage (\texttt{wage}), the number of jobs (\texttt{njobs}), the total hours of work per week (\texttt{hours}) for each survey year, the year when individual starting to work (\texttt{stwork}), the length of time (years) in workforce (\texttt{yr\_workforce}), and work experience measured as the number of years worked (\texttt{exp}).
 As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.
@@
 
+For \texttt{st\_work} variable, we only rename the column names from the raw, NLSY79 data.
+This variable is then used to calculate the next variable, \texttt{yr\_workforce} for each survey round, which is the year of survey (\texttt{year}) minus the year of individual start working (\texttt{st\_work}).
+Finally, \texttt{exp} variable is derived from the variable started with \texttt{WKSWK} in \texttt{new\_data\_qnames}.
+This variable measured since the last interview.
+Thus, to obtain the work experience since 1979, we calculate the cumulative value.
+As the measurement unit is in week, we convert this to year.
+
 The employment and demographic variables are then joined.
@@
 
-According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after being collected to ensure that the data is clean, valid, and ready to be deployed in the later formal statistical analysis.
-Moreover, Chatfield (1985) argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
+According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure that the data is clean, valid, and ready to be deployed in the later analyses.
+This is supported by Chatfield (1985) who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
 
-In this paper, we conduct an IDA or a preliminary data analysis to assess the consistency of the data with the cohort information that the NLSY provides.
-In addition, we also aim to find the anomaly in the wages values using this approach.
-We mainly use graphical summaries to do the IDA using \texttt{ggplot2} (Wickham 2016) and \texttt{brolgar} (Tierney, Cook, and Prvan 2020).
+In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides.
+A first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
+This is followed with graphical summaries using methods available in \texttt{ggplot2} (Wickham 2016) and \texttt{brolgar} (Tierney, Cook, and Prvan 2020).
 
-As stated previously, the respondents' ages ranged from 12 to 22 when first interviewed in 1979.
-Hence, we validate whether all of the respondents were in this range.
+The respondents' ages ranged from 12 to 22 when first interviewed in 1979.
+Hence, we validate whether all of the respondents were in this range in the data we extracted.
 Additionally, the \href{https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction}{NLSY} also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic).
@@
 Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest that the demographic data we had is consistent with the sample information in the database.
-
-\begin{table}
-
-\caption{\label{tab:age-table}The frequency table of the age at the start of the survey (?CHECK) in the full NSLY79 data}
-\centering
-\begin{tabular}[t]{rr}
-\toprule
-Age & Number of individuals\\
-\midrule
-15 & 1,265\\
-16 & 1,550\\
-17 & 1,600\\
-18 & 1,530\\
-19 & 1,662\\
-20 & 1,722\\
-21 & 1,677\\
-22 & 1,680\\
-\bottomrule
-\end{tabular}
-\end{table}
-
-\begin{table}
-
-\caption{\label{tab:gender-race-table}The contingency table for gender and race for the full NLSY79 data.}
-\centering
-\begin{tabular}[t]{lrrrr}
-\toprule
-\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Race} & \multicolumn{1}{c}{ } \\
-\cmidrule(l{3pt}r{3pt}){2-4}
-Gender & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
-\midrule
-Male & 1,000 (15.62\%) & 1,613 (25.19\%) & 3,790 (59.19\%) & 6,403 (100.00\%)\\
-Female & 1,002 (15.95\%) & 1,561 (24.84\%) & 3,720 (59.21\%) & 6,283 (100.00\%)\\
-\midrule
-Total & 2,002 (15.78\%) & 3,174 (25.02\%) & 7,510 (59.20\%) & 12,686 (100.00\%)\\
-\bottomrule
-\end{tabular}
-\end{table}
 
@@
 Some of them, for example respondents in Figure \ref {fig:feature-plot} C had experienced unusual wages only in certain year.
+
+\begin{table}
+
+\caption{\label{tab:age-table}The frequency table of the age at the start of the survey in NSLY79 cohort}
+\centering
+\begin{tabular}[t]{rr}
+\toprule
+Age & Number of individuals\\
+\midrule
+15 & 1,265\\
+16 & 1,550\\
+17 & 1,600\\
+18 & 1,530\\
+19 & 1,662\\
+20 & 1,722\\
+21 & 1,677\\
+22 & 1,680\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\textbackslash begin\{table\}
+
+\textbackslash caption\{\label{tab:gender-race-table}The contingency table for gender and race for the full NLSY79 data.
+The percentage (rounded to closest 1\%) is out of the total corresponding to row.\}
+\centering
+
+\begin{tabular}[t]{lrrrr}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Race} & \multicolumn{1}{c}{ } \\
+\cmidrule(l{3pt}r{3pt}){2-4}
+Gender & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
+\midrule
+Male & 1,000 (16\%) & 1,613 (25\%) & 3,790 (59\%) & 6,403\\
+Female & 1,002 (16\%) & 1,561 (25\%) & 3,720 (59\%) & 6,283\\
+\midrule
+Total & 2,002 (16\%) & 3,174 (25\%) & 7,510 (59\%) & 12,686\\
+\bottomrule
+\end{tabular}
+
+\textbackslash end\{table\}
 
@@
 Users need to be able to trust the data.
+
+Why is this exercise important for teachers and students?
+The data package and software could be used for illustrating data cleaning in the classroom.
+The refreshed and contemporary data provided in the package can be used as examples in the teaching of initial data analysis and modeling, and exploratory data analysis of longitudinal data.
+
+The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual.
+From modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns.
+That more education leads to increasingly higher wages, which is a satisfying result for educators.
+It means that education makes a difference in the wage experience, and that public policy that encourages education is a data-supported action.
+We would also learn, although not presented here, that race matters and that being black leads to lower wages.
+This is a disturbing finding because there is no rational for such a difference in a fair society.
+This is would provide support for action in public policy to remove this overall average effect.
+It also provides an example for educators to explain the use of data to support public policy action.
+
+These are interpretations of results from analyzing the wages data, which rely on trusting that the data provided is accurate and valid.
+The data is collected from a reputable organization, but the data has obvious errors that need correcting.
+This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.
+
+Having trustworthy data is imperative for statistics and data science education.
+Students do interpret results and take seriously what is reported from example analyses.
+If the data in the examples are synthetic, with our own inherent prejudices (on fog e.g.~gender and race) applied, or inaccurately processed leaving errors, we as educators are being irresponsible because the societal message taught to students is flawed.
+This paper demonstrates the process to produce a trustworthy data set for teaching.
 

### 2022-02-09 paper/paper.Rmd
Added the explanation of work experience in Section 4.

@@
 We also plan to make some modifications of the variables.
-For example, we create a categorical race variable instead of the binary variable "black" and "hispanic".
+For example, we create a categorical race variable instead of the binary variable "Black" and "Hispanic".
 We also include additional variables so that more interesting data exploration can be included in teaching examples: year of survey, age of individual in 1979, year when the highest grade was completed, number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
@@
 
-<!--AE comment: In Section 4, you talk about the variables in the original data.
-“One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection.” I’m not sure what this means.
-Please clarify.-->
-<!--AE comment: “experience in the workforce” should have the period inside the quotations.-->
-There are two aspects of the original data that make a direct comparison difficult.
-The first one is that The time variable provided is "experience in the workforce".
-It is not clear how this is calculated.
-One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection.
-Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.
+There are two aspects of the original data that make the replication with exactly the same result is not obtained.
+The first one is that the calculation of the time variable provided, which is the "work experience", is not clearly articulated.
+@SingerJudithD2003Alda mention that the temporal variable is the years of experience since the first day of work.
+However, this exact same variable is not explicitly available in the database.
+From the NLSY79 topical guide [@nlsy79guide], we find that several variables are tagged as work experience-related variables.
+One of them is the weeks of worked since the last interview.
+Hence, we use this to calculate the variable.
+Even though we got different result, the correlation between the original and refreshed work experience is reasonably high as shown in Figure \ref{fig:compare-xp-sw}.
 
-```{r compare_xp_to_yrworkforce, fig.cap = "The association between the lenght of time in the workforce and the work experience.
-The color represents the respondents.
-Generally, the longer someone were in the workforce, the higher the work experience.
-However, some of the respondents have much lower working experience compared their duration in the workforce.", eval=FALSE}
+```{r compare-xp-sw, fig.cap = "Males dropouts' work experience in the original and refreshed data.
+Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
+We can see that the correlation is reasonably high between them even though the exactly same figure could not be obtained.", out.width = "70%"}
 ```
 
-```{r compare_xp_sw}
+Secondly, in the original dataset, wages were inflation-adjusted to 1990 prices.
+This is not done for the refreshed data because we plan to keep refreshing it as the data from the newest survey period release.
+The 1990 prices might not be relevant anymore for the future context.
+Hence, we keep it as the original value to allow the users to treat it as their needs.
+
+```{r compare-xp-to-yrworkforce, fig.cap = "The association between the lenght of time in the workforce and the work experience.
+Generally, the longer someone were in the workforce, the higher the work experience.
+However, some of the respondents have much lower working experience compared to their duration in the workforce.
+This is possible when someone ", eval=FALSE}
 ```
@@
 Either way, please clarify.-->
-The treatment of unlikely wages differ in the refreshed data.
+It is also worth noting that the treatment of unlikely wages differ in the refreshed data.
 In the original data by @SingerJudithD2003Alda, wages greater than $75 are set to be missing.

### 2022-02-09 paper/paper.tex
Added the explanation of work experience in Section 4.

@@
 The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
-Department of Labor 2021).
+Department of Labor 2021a).
 
@@
 The NLSY79 is a longitudinal survey administered by the U.S Bureau of Labor Statistics that follows the lives of a sample of American youth born between 1957-1964 (Bureau of Labor Statistics, U.S.
-Department of Labor 2021).
+Department of Labor 2021a).
 The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979.
@@
 We also plan to make some modifications of the variables.
-For example, we create a categorical race variable instead of the binary variable ``black'' and ``hispanic''.
+For example, we create a categorical race variable instead of the binary variable ``Black'' and ``Hispanic''.
 We also include additional variables so that more interesting data exploration can be included in teaching examples: year of survey, age of individual in 1979, year when the highest grade was completed, number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
@@
 
-M.
 van der Loo and de Jonge (2018) in the context of official statistics describe the ``statistical value chain'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organised with correct type and identified variables) and valid data (data that has been cleaned and more accurately represents the intent of variables).
@@
 
-Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
-2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham et al.
+Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
+2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham, François, et al.
 2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
@@
 \end{verbatim}
+
+For \texttt{st\_work} variable, we only rename the column names from the raw, NLSY79 data.
+This variable is then used to calculate the next variable, \texttt{yr\_workforce} for each survey round, which is the year of survey (\texttt{year}) minus the year of individual start working (\texttt{st\_work}).
+Finally, \texttt{exp} variable is derived from the variable started with \texttt{WKSWK} in \texttt{new\_data\_qnames}.
+This variable measured since the last interview.
+Thus, to obtain the work experience since 1979, we calculate the cumulative value.
+As the measurement unit is in week, we convert this to year.
 
@@
 
-There are two aspects of the original data that make a direct comparison difficult.
-The first one is that The time variable provided is ``experience in the workforce''.
-It is not clear how this is calculated.
-One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection.
-Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.
+There are two aspects of the original data that make the replication with exactly the same result is not obtained.
+The first one is that the calculation of the time variable provided, which is the ``work experience'', is not clearly articulated.
+Singer and Willett (2003) mention that the temporal variable is the years of experience since the first day of work.
+However, this exact same variable is not explicitly available in the database.
+From the NLSY79 topical guide (Bureau of Labor Statistics, U.S.
+Department of Labor 2021b), we find that several variables are tagged as work experience-related variables.
+One of them is the weeks of worked since the last interview.
+Hence, we use this to calculate the variable.
+Even though we got different result, the correlation between the original and refreshed work experience is reasonably high as shown in Figure \ref{fig:compare_xp_sw}.
 
-\begin{center}\includegraphics[width=468px]{figures/compare_xp_sw-1} \end{center}
+\begin{figure}
 
-The treatment of unlikely wages differ in the refreshed data.
+{\centering \includegraphics[width=0.7\linewidth]{figures/compare-xp-sw-1}
+
+}
+
+\caption{Males dropouts' work experience in the original and refreshed data.
+Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
+We can see that the correlation is reasonably high between them even though the exactly same figure could not be obtained.}\label{fig:compare-xp-sw}
+\end{figure}
+
+Secondly, in the original dataset, wages were inflation-adjusted to 1990 prices.
+This is not done for the refreshed data because we plan to keep refreshing it as the data from the newest survey period release.
+The 1990 prices might not be relevant anymore for the future context.
+Hence, we keep it as the original value to allow the users to treat it as their needs.
+
+It is also worth noting that the treatment of unlikely wages differ in the refreshed data.
 In the original data by Singer and Willett (2003), wages greater than \$75 are set to be missing.
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-iris-data}{}}%
-Andereson, Edgar. 1935. {``The Irises of the Gaspe Peninsula.''} \emph{Bulletin of the American Iris Society} 59: 2--5.
+\leavevmode\hypertarget{ref-iris-data}{}%
+Andereson, Edgar. 1935. ``The Irises of the Gaspe Peninsula.'' \emph{Bulletin of the American Iris Society} 59: 2--5.
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
+\leavevmode\hypertarget{ref-nlsy79}{}%
 Bureau of Labor Statistics, U.S.
 Department of Labor.
-2021.
-{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
+2021a.
+``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
 Columbus, OH.
 
-\leavevmode\vadjust pre{\hypertarget{ref-shiny}{}}%
-Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{{shiny: Web Application Framework for R}}. \url{https://CRAN.R-project.org/package=shiny}.
+\leavevmode\hypertarget{ref-nlsy79guide}{}%
+---------.
+2021b.
+``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
+Columbus, OH.
+\url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Chatfield1985TIEo}{}}%
-Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
+\leavevmode\hypertarget{ref-shiny}{}%
+Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-eliznlsy}{}}%
+\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
+Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
+
+\leavevmode\hypertarget{ref-eliznlsy}{}%
 Cooksey, Elizabeth C.
 2017.
-{``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M.
+``Using the National Longitudinal Surveys of Youth (Nlsy) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
 Lerner Neal Halfon Christoper B.
@@
 2018.
-{``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1).
+``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1).
 \url{https://doi.org/10.5070/T511103589}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-KollerManuel2016rARP}{}}%
-Koller, Manuel. 2016. {``{robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models}.''} \emph{Journal of Statistical Software} 75 (6): 1--24.
+\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
+Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.
 
-\leavevmode\vadjust pre{\hypertarget{ref-notaverage}{}}%
-Moncrief, Marc. 2015. {``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.''} \url{https://bit.ly/smh-not-normal}.
+\leavevmode\hypertarget{ref-notaverage}{}%
+Moncrief, Marc. 2015. ``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.'' \url{https://bit.ly/smh-not-normal}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-opendata}{}}%
-Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.
+\leavevmode\hypertarget{ref-opendata}{}%
+Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-patchwork}{}}%
-Pedersen, Thomas Lin. 2020. \emph{{patchwork: The Composer of Plots}}. \url{https://CRAN.R-project.org/package=patchwork}.
+\leavevmode\hypertarget{ref-patchwork}{}%
+Pedersen, Thomas Lin. 2020. \emph{patchwork: The Composer of Plots}. \url{https://CRAN.R-project.org/package=patchwork}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-MichaelRPergamit2001DWTN}{}}%
+\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
 Pergamit, Michael R., Charles R.

### 2022-02-09 data-raw/eda.R

@@ -172,7 +172,7 @@ do2 <- do %>%
 do2 %>% select(id, race) %>% distinct() %>% count(race)
 do2 %>% select(id, hgc12) %>% distinct() %>% count(hgc12)
 
-# ---- compare_xp_to_yrworkforce
+# ---- compare-xp-to-yrworkforce
 
 ggplot(do, aes(x = yr_wforce,
                y = exp)) +
@@ -180,7 +180,7 @@ ggplot(do, aes(x = yr_wforce,
   guides(color = "none") +
   geom_abline(intercept = 0, slope = 1)
 
-# ---- compare_xp_sw
+# ---- compare-xp-sw
 
 do_sw_join <- left_join(sw, do, by = c("id", "index"))
 

### 2022-02-09 data-raw/eda.R
Fix the issue in table 2's caption (percentage symbol).

@@ -190,7 +190,8 @@ ggplot(do_sw_join, aes(x = xp,
   geom_abline(intercept = 0, slope = 1) +
   guides(color = "none") +
   xlab("Work experience (Singer & Willet, 2008)") +
-  ylab("Work experience (refreshed)")
+  ylab("Work experience (refreshed)") +
+  theme_bw()
 
 
 # Takeaways:

### 2022-02-09 data-raw/ida.R

@@ -38,7 +38,7 @@ gender_race_table <- full_demographics %>%
          Total = str_replace(Total, "\\([1|2]00\\%\\)", ""))
 
 kable(gender_race_table,
-      caption = "The contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1%) is out of the total corresponding to row.",
+      caption = "The contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1\\%) is out of the total corresponding to row.",
       col.names = c("Gender", "Hispanic", "Black", "Non-Black, Non-Hispanic", "Total"),
       booktabs = TRUE,
       linesep = "",

### 2022-02-09 response/response1.Rmd
Table 2 fix, added to response
update to response

@@
 
-We thank the Editor and the three reviewers for their comments, which have helped substantially to improve the manuscript.
+We thank the Editor, Associate Editor and the three reviewers for their extensive comments, which have helped substantially to improve the manuscript.
 In response, we have made a major revision of the main manuscript as outlined below.
@@
 
-The editor and reviewers suggest reframing so that readers can help their students to experience the journal from wild to textbook data and either emphasize the case study, for the paper to be a regular submission for JSDSE, or to better explain the relevance for  teachers and/or students in statistics and data science to be considered for the special issue.
+The editors and reviewers suggest reframing so that readers can help their students to experience the journal from wild to textbook data and either emphasize the case study, for the paper to be a regular submission for JSDSE, or to better explain the relevance for teachers and/or students in statistics and data science to be considered for the special issue.
 We have chosen to do the latter, because we hope that the paper fits into the goals of the special issue.
@@
 \begin{description}
-\item[Reviewer 1:] \textcolor{violet}{How is this manuscript relevant for teachers and/or students in statistics and data science?}
+\item[Editor and Reviewer 1:] \textcolor{violet}{How is this manuscript relevant for teachers and/or students in statistics and data science?}
 
 The introduction has been re-written with an emphasis on how this work should be interesting for teachers and students of data science and statistics.
+
+\item[Reviewer 1:] \textcolor{violet}{I don't see the direct relevance of this manuscript with the readership of JSDSE.
+I think this could be a valuable paper to describe the author(s) R package and would have value for researchers using the NLSY79 dataset.
+I am not convinced that JSDSE is the appropriate venue for this.
+I do not see this manuscript being engaging enough to be used in the classroom as a case study in data cleaning.}
+
+It is commonly said 80\% of a data analysts/scientist's work is on data cleaning/preperation so we believe that shedding more light in this aspect of work is an important for students to be aware of.
+We believe that there are often subjective decisions made in the process (as we have done) which can be used as a conversation for transparency and reproducibility in class based on a real case study.
 
@@
 
-* We have corrected all minor grammatical errors pointed out by the reviewers as outlined below. [ADD the changes]  
+* We have corrected all minor grammatical errors pointed out by the reviewers as outlined below:
+
+\begin{description}
+\item[Reviewer 1:] \textcolor{violet}{
+\begin{itemize}
+\item p2 l16: It's unclear from context what is being referred to from the Huebner et al 2020 citation.
+\item p3 l19: sentence ends with a comma
+\item p3 l24: this is an important point that shouldn't be relegated to a parenthetical statement
+\item p3 l38: compared "to" the old textbook subset
+\item p4 l51: might be better to add "the" before "data cleaning process"
+\item p5 l39: two data "sets"
+\item p6 l28: comply "with" these rules
+\end{itemize}}
+\item[Reviewer 1:] \textcolor{violet}{}
+\end{description}
+
 * We have made the descriptions of the steps more explicit as follows [LIST OF THE CHANGES]
@@
 
-We agree that this level of precision was not necessary. We have modified Table 2 in line with this.
+We agree that this level of precision was not necessary. We have modified Table 2 so percentages are to the nearest 1 percent and total percentage (=100%) is omitted.
 

### 2022-02-09 paper/paper.Rmd

@@
 
-For `st_work` variable,  we only rename the column names from the raw, NLSY79 data.
+For `st_work` variable,  we only rename the column names from `new_data_qnames`.
 This variable is then used to calculate the next variable, `yr_workforce` for each survey round, which is the year of survey (`year`) minus the year of individual start working (`st_work`).
-Finally, `exp` variable is derived from the variable started with `WKSWK` in `new_data_qnames`.
-This variable measured since the last interview.
-Thus, to obtain the work experience since 1979, we calculate the cumulative value.
+Finally, `exp` variable is derived from the number of weeks worked since the last interview indicated as variable started with `WKSWK` in `new_data_qnames`.
+To obtain the work experience since 1979, we calculate the cumulative value.
 As the measurement unit is in week, we convert this to year.
@@
 
+```{r compare-xp-sw, fig.cap = "Males dropouts' work experience in the original and refreshed data.
+Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
+We can see that the correlation is reasonably high between them even though the exactly same figure could not be obtained.", out.width = "70%"}
+```
+
 <!--AE comment: first sentence can be removed or rephrased.
@@
 
-```{r compare-xp-sw, fig.cap = "Males dropouts' work experience in the original and refreshed data.
-Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
-We can see that the correlation is reasonably high between them even though the exactly same figure could not be obtained.", out.width = "70%"}
-```
-
 Secondly, in the original dataset, wages were inflation-adjusted to 1990 prices.
 This is not done for the refreshed data because we plan to keep refreshing it as the data from the newest survey period  release.
-The 1990 prices might not be relevant anymore for the future context.
+Thus, the 1990 prices might not be relevant anymore for the future context.
 Hence, we keep it as the original value to allow the users to treat it as their needs.

### 2022-02-09 paper/paper.tex

@@
 
-For \texttt{st\_work} variable, we only rename the column names from the raw, NLSY79 data.
+For \texttt{st\_work} variable, we only rename the column names from \texttt{new\_data\_qnames}.
 This variable is then used to calculate the next variable, \texttt{yr\_workforce} for each survey round, which is the year of survey (\texttt{year}) minus the year of individual start working (\texttt{st\_work}).
-Finally, \texttt{exp} variable is derived from the variable started with \texttt{WKSWK} in \texttt{new\_data\_qnames}.
-This variable measured since the last interview.
-Thus, to obtain the work experience since 1979, we calculate the cumulative value.
+Finally, \texttt{exp} variable is derived from the number of weeks worked since the last interview indicated as variable started with \texttt{WKSWK} in \texttt{new\_data\_qnames}.
+To obtain the work experience since 1979, we calculate the cumulative value.
 As the measurement unit is in week, we convert this to year.
@@
 
-\caption{\label{tab:gender-race-table}The contingency table for gender and race for the full NLSY79 data.}
+\caption{\label{tab:gender-race-table}The contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
 \centering
@@
 \midrule
-Male & 1,000 (16\%) & 1,613 (25\%) & 3,790  (59\%) & 6,403 (100\%)\\
-Female & 1,002 (16\%) & 1,561 (25\%) & 3,720  (59\%) & 6,283 (100\%)\\
+Male & 1,000 (16\%) & 1,613 (25\%) & 3,790 (59\%) & 6,403\\
+Female & 1,002 (16\%) & 1,561 (25\%) & 3,720 (59\%) & 6,283\\
 \midrule
-Total & 2,002 (32\%) & 3,174 (50\%) & 7,510 (118\%) & 12,686 (200\%)\\
+Total & 2,002 (16\%) & 3,174 (25\%) & 7,510 (59\%) & 12,686\\
 \bottomrule
@@
 
+\begin{figure}
+
+{\centering \includegraphics[width=0.7\linewidth]{figures/compare-xp-sw-1}
+
+}
+
+\caption{Males dropouts' work experience in the original and refreshed data.
+Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
+We can see that the correlation is reasonably high between them even though the exactly same figure could not be obtained.}\label{fig:compare-xp-sw}
+\end{figure}
+
 The original set contains wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994.
@@
 Hence, we use this to calculate the variable.
-Even though we got different result, the correlation between the original and refreshed work experience is reasonably high as shown in Figure \ref{fig:compare_xp_sw}.
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.7\linewidth]{figures/compare-xp-sw-1}
-
-}
-
-\caption{Males dropouts' work experience in the original and refreshed data.
-Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
-We can see that the correlation is reasonably high between them even though the exactly same figure could not be obtained.}\label{fig:compare-xp-sw}
-\end{figure}
+Even though we got different result, the correlation between the original and refreshed work experience is reasonably high as shown in Figure \ref{fig:compare-xp-sw}.
 
@@
 This is not done for the refreshed data because we plan to keep refreshing it as the data from the newest survey period release.
-The 1990 prices might not be relevant anymore for the future context.
+Thus, the 1990 prices might not be relevant anymore for the future context.
 Hence, we keep it as the original value to allow the users to treat it as their needs.

### 2022-02-09 response/response1.Rmd
update to response

@@
 \end{itemize}}
-\item[Reviewer 1:] \textcolor{violet}{}
+\item[Reviewer 1:] \textcolor{violet}{This sentence is unclear.
+What is dplyr used for?
+Are there tidyverse packages used but not mentioned?
+If not, splitting this sentence into two may make it clearer which package is used to what purpose.
+Also, I haven't tried it in code, but I imagine that the data could be tidied as described just using pivot\_longer, at least for the job number, year, and wage data.
+In other words, I don't see why dplyr or stringr would be needed for the data described on page 6.}
+
+We did not go into great detail about the packages used as the whole code is available as a vignette within the package.
+We realise that this was not available to the reviewers due to the requirement of a blind review.
+To circumvent this, we have added the vignettes as supplementary materials.
 \end{description}

### 2022-02-09 data-raw/ida.R
modify color to be color blind friendly

@@ -124,7 +124,8 @@ feature_bp <- ggplot(wages_feat_long,
   labs(tag = "(B)")  +
   ylab("Hourly wage ($)") +
   xlab("Feature") +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  scale_color_viridis_d()
 
 plot_high <- ggplot(filter(wages_high2, id == 39)) +
   geom_line(aes(x = year,
@@ -229,7 +230,8 @@ feature2_bp <- ggplot(wages_feat_long_rlm,
   labs(x = "Feature",
        y = "Hourly wage ($)",
        tag = "(B)") +
-  theme(legend.position = "none")
+  theme(legend.position = "none") +
+  scale_color_viridis_d()
 
 plot_high_after <- ggplot(filter(wages_cleaned, id == 39)) +
   geom_line(aes(x = year,

### 2022-02-09 response/response1.Rmd
remove minor grammars

@@
 
-\section*{Minor changes:}
-
-* We have corrected all minor grammatical errors pointed out by the reviewers as outlined below:
-
 \begin{description}
-\item[Reviewer 1:] \textcolor{violet}{
-\begin{itemize}
-\item p2 l16: It's unclear from context what is being referred to from the Huebner et al 2020 citation.
-\item p3 l19: sentence ends with a comma
-\item p3 l24: this is an important point that shouldn't be relegated to a parenthetical statement
-\item p3 l38: compared "to" the old textbook subset
-\item p4 l51: might be better to add "the" before "data cleaning process"
-\item p5 l39: two data "sets"
-\item p6 l28: comply "with" these rules
-\end{itemize}}
 \item[Reviewer 1:] \textcolor{violet}{This sentence is unclear.
@@
 \end{description}
+
+\section*{Minor changes:}
+
+* We have corrected all minor grammatical errors pointed out by the reviewers.
 

### 2022-02-09 data-raw/eda.R
make x label same for swdata

@@ -13,7 +13,7 @@ sw_wages <- sw %>%
   labs(tag = "(A)", title = "Original") +
   theme_bw() +
   ylab("Hourly wage ($, natural log)") +
-  xlab("Experience in years") +
+  xlab("Experience (years)") +
   #theme(plot.title = element_text(size = 10)) +
   ylim(0, 5)
 

### 2022-02-13 data-raw/NLSY79/NLSY79.R
Edited the droputs.

@@ -4,12 +4,11 @@
 
 
 new_data <- read.table(here::here('data-raw/NLSY79/NLSY79.dat'), sep=' ')
-names(new_data) <- c('E0010101',
-  'E0010102',
+names(new_data) <- c('E0010102',
   'R0000100',
   'R0000300',
   'R0000500',
-  'R0017300',
+  'R0018300',
   'R0047010',
   'R0070500',
   'R0070600',
@@ -25,9 +24,7 @@ names(new_data) <- c('E0010101',
   'R0214700',
   'R0214800',
   'R0215300',
-  'R0216700',
-  'R0216701',
-  'R0229200',
+  'R0230000',
   'R0263710',
   'R0337800',
   'R0338910',
@@ -39,12 +36,10 @@ names(new_data) <- c('E0010101',
   'R0373410',
   'R0383800',
   'R0384910',
-  'R0406400',
-  'R0406401',
   'R0406700',
   'R0410100',
   'R0410300',
-  'R0417400',
+  'R0418200',
   'R0446810',
   'R0545600',
   'R0546610',
@@ -56,10 +51,8 @@ names(new_data) <- c('E0010101',
   'R0585910',
   'R0598000',
   'R0599010',
-  'R0618900',
-  'R0618901',
   'R0645800',
-  'R0664500',
+  'R0665300',
   'R0702510',
   'R0839900',
   'R0841010',
@@ -72,9 +65,7 @@ names(new_data) <- c('E0010101',
   'R0892300',
   'R0893410',
   'R0897300',
-  'R0898200',
-  'R0898201',
-  'R0905900',
+  'R0906700',
   'R0945610',
   'R1087300',
   'R1088210',
@@ -86,10 +77,8 @@ names(new_data) <- c('E0010101',
   'R1127810',
   'R1140100',
   'R1141010',
-  'R1145000',
-  'R1145001',
   'R1145800',
-  'R1205800',
+  'R1206600',
   'R1256010',
   'R1463000',
   'R1463910',
@@ -101,10 +90,8 @@ names(new_data) <- c('E0010101',
   'R1503210',
   'R1515400',
   'R1516310',
-  'R1520200',
-  'R1520201',
   'R1521000',
-  'R1605100',
+  'R1605900',
   'R1650810',
   'R1809800',
   'R1810710',
@@ -116,10 +103,8 @@ names(new_data) <- c('E0010101',
[…]

### 2022-02-13 data-raw/data-preprocessing.R

@@ -1,5 +1,3 @@
-
-
 ## ---- load-pkgs
 library(tidyverse)
 select <- dplyr::select
@@ -59,32 +57,25 @@ demog_tidy <- categories_qnames %>%
          gender = SAMPLE_SEX_1979)
 
 ## ---- demog-ed
-demog_education <- new_data_qnames %>%
-  # in 2018, the variable's name is Q3-4_2018, instead of HGC_2018
-  rename(HGC_2018 = `Q3-4_2018`) %>%
-  select(id = CASEID_1979,
-         starts_with("HGCREV"),
-         "HGC_2012",
-         "HGC_2014",
-         "HGC_2016",
-         "HGC_2018") %>%
-  pivot_longer(!id,
-               names_to = "var",
-               values_to = "grade") %>%
-  separate("var", c("var", "year"), sep = "_") %>%
-  filter(!is.na(grade)) %>%
-  select(-var)
+# demog_education <- new_data_qnames %>%
+#   # in 2018, the variable's name is Q3-4_2018, instead of HGC_2018
+#   rename(HGC_2018 = `Q3-4_2018`) %>%
+#   select(id = CASEID_1979,
+#          starts_with("HGCREV"),
+#          "HGC_2012",
+#          "HGC_2014",
+#          "HGC_2016",
+#          "HGC_2018") %>%
+#   pivot_longer(!id,
+#                names_to = "var",
+#                values_to = "grade") %>%
+#   separate("var", c("var", "year"), sep = "_") %>%
+#   filter(!is.na(grade)) %>%
+#   select(-var)
 
-## ---- tidy-hgc
-# Get the highest year completed
-highest_year <- demog_education %>%
-  group_by(id) %>%
-  filter(grade == max(grade)) %>%
-  filter(year == min(year)) %>%
-  rename(yr_hgc = year,
-         hgc_i = grade) %>%
-  select(id, yr_hgc, hgc_i) %>%
-  ungroup() %>%
+hgc_ever <- new_data_qnames %>%
+  select(id = CASEID_1979,
+         hgc_i = HGC_EVER_XRND) %>%
   mutate(hgc = case_when(hgc_i == 0 ~ "NONE",
                          hgc_i == 1 ~ "1ST GRADE",
                          hgc_i == 2 ~ "2ND GRADE",
@@ -93,10 +84,48 @@ highest_year <- demog_education %>%
                          hgc_i == 13 ~ "1ST YEAR COL",
                          hgc_i == 14 ~ "2ND YEAR COL",
[…]

### 2022-02-13 data-raw/eda.R

@@ -48,6 +48,7 @@ sw_wages_mod <- sw %>%
 # filter do data up to 1994 since the textbook data only covers that period.
 
 do <- yowie::wages_hs_do %>%
+  as_tibble() %>%
   group_by(id) %>%
   mutate(lnwage = log(wage),
          index = 1:n())
@@ -65,7 +66,7 @@ do_ref <- do %>%
   ylim(0, 5)
 
 do_ref_mod <- do %>%
-  mutate(hgc12 = ifelse(hgc_i < 12, "BELOW 12TH", "12TH")) %>%
+  mutate(hgc12 = ifelse(hgc_i < 9, "8TH", "12TH")) %>%
   ggplot(aes(x = exp,
              y = lnwage,
              linetype = hgc12)) +

### 2022-02-13 data-raw/ida.R

@@ -259,17 +259,19 @@ wages_clean <- wages_cleaned %>%
 
 # rename and select the wages in tidy
 wages <- wages_clean %>%
-  select(id, year, mean_hourly_wage, age_1979, gender, race, hgc, hgc_i, yr_hgc,
-                number_of_jobs, total_hours, is_wm, is_pred) %>%
+  select(id, year, mean_hourly_wage, age_1979, gender, race, hgc, hgc_i, ged = dip_or_ged,
+                number_of_jobs, total_hours, stwork_year, years_in_workforce, exp, is_wm, is_pred) %>%
   mutate(id = as.factor(id),
          hgc = as.factor(hgc),
          year = as.integer(year),
          age_1979 = as.integer(age_1979),
-         yr_hgc = as.integer(yr_hgc),
+         ged = as.factor(ged),
          number_of_jobs = as.integer(number_of_jobs)) %>%
   rename(wage = mean_hourly_wage,
          njobs = number_of_jobs,
-         hours = total_hours) %>%
+         hours = total_hours,
+         stwork = stwork_year,
+         yr_wforce = years_in_workforce) %>%
   as_tsibble(key = id,
              index = year,
              regular = FALSE)
@@ -282,28 +284,46 @@ demog_nlsy79 <- full_demographics %>%
          race,
          hgc,
          hgc_i,
-         yr_hgc) %>%
+         ged = dip_or_ged) %>%
   mutate(id = as.factor(id),
          age_1979 = as.integer(age_1979),
          hgc = as.factor(hgc),
-         yr_hgc = as.integer(yr_hgc))
+         ged = as.factor(ged))
 
 # Create a data set for the high school dropouts cohort
+# Create a data set for the high school dropouts cohort
 wages_hs_do <- wages %>%
-  mutate(dob = 1979 - age_1979,
-         age_hgc = yr_hgc - dob) %>%
-  filter((hgc %in% c("9TH GRADE",
-                     "10TH GRADE",
-                     "11TH GRADE")) |
-           (hgc == "12TH GRADE" &
-              age_hgc >= 19)) %>%
+  filter(hgc_i < 12 | (hgc_i >= 12 & ged != 1)) %>%
   filter(age_1979 <= 17,
          gender == "MALE") %>%
-  select(-dob,
-                -age_hgc) %>%
   as_tsibble(key = id,
              index = year,
              regular = FALSE)
+
+
[…]

### 2022-02-13 paper/paper.Rmd
Edited the paper with respect to the changes in dataset.

@@
 For example, we create a categorical race variable instead of the binary variable "Black" and "Hispanic".
-We also include additional variables so that more interesting data exploration can be included in teaching examples: year of survey, age of individual in 1979, year when the highest grade was completed, number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
+We also include additional variables so that more interesting data exploration can be included in teaching examples: year of survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
 
-We will also make all of the high school graduates data available, including females, in addition to high school dropouts, and separate the demographic data of the NLSY79 cohort.
-Thus, the plan is to create three datasets as follows:
+We will also make the wages and demographic data of the NLSY79 cohort available, including females, in addition to high school dropouts. Thus, the plan is to create three datasets as follows:
 
 1. The demographic data of the NLSY79 cohort.
-2. The wages data of the up to high school graduates.
+2. The wages data of NLSY79 cohort.
 3. The wages data of the high school dropouts, which is closest to a refreshed version of @SingerJudithD2003Alda's data.
@@
 \begin{itemize}
-\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year $\triangleright$ Highest grade completed
+\item[$\triangleright$] Education $\triangleright$ Summary measures $\triangleright$ All schools $\triangleright$ By year
 \begin{itemize}
-\item[\faCheck] All 80 variables in Highest grade completed were selected.
+\item[$\triangleright$] Dates of diploma or degree
+\begin{itemize}
+\item[\faCheck] All variables named Q3-8A were selected. 
+\end{itemize}
+\end{itemize}
+\begin{itemize}
+\item[$\triangleright$] Highest grade completed
+\begin{itemize}
+\item[\faCheck] Selected HGC\_EVER, which is the highest grade respondent has ever reported completed and getting credit for.
+\end{itemize}
 \end{itemize}
@@
 \end{itemize}
-\item[$\triangleright$] Household, Geography and Contextual Variables
+\item[$\triangleright$] Household, Geography \& Demographics
 \begin{itemize}
-\item[$\triangleright$] Context $\triangleright$ Summary measures $\triangleright$ Basic demographics $\triangleright$ Date of birth
+\item[$\triangleright$] Demographics $\triangleright$ Basic demographics $\triangleright$ Date of birth
 \begin{itemize}
@@
 \begin{itemize}
-\item[\faCloudDownload] To download all 686 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download".
+\item[\faCloudDownload] To download all 663 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download".
 We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers.
@@
 In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (`gender`), race (`race`), age (`age_1979`), highest grade completed (`hgc`), highest grade completed in terms of years, e.g.
-9th grade = 9, 3rd year college = 15, (`hgc_i`) and the corresponding year this grade was completed (`yr_hgc`).
+9th grade = 9, 3rd year college = 15, (`hgc_i`), and the whether the graduate equivalency diploma is obtained (`ged`).
 
@@
 -->
-The next step is tidying to obtain `hgc` and `yr_hgc`.
-The highest grade completed are recorded in `new_data_qnames` as variables beginning with `Q3-4` and `HGC` with suffix of the year it was recorded.
-In addition, the variables beginning with `HGCREV` contain the revised data.
-We choose to use this revised data because it seemed to have less missing and presumably has been checked.
-However, there is no revised data for 2012, 2014, 2016, and 2018, thus, we use the ordinary data for these years.
+The next step is tidying to obtain `hgc` and `ged`.
+As the highest grade completed is already exist in  `new_data_qnames` and stored in years unit, we only rename the column name to be `hgc_i` and recode it so we have the factor data type.
+For `ged`, we only separate the years and the GED status of the individual.
+Later on, this variable is useful to determine the refreshed high school dropouts data.
 
@@
 ```
-
-The `hgc` is measured and could be updated in each period of the survey.
-We chose to only retain the highest grade completed for each individual and derived the year when they completed it (`yr_hgc`) by finding the minimum year with the highest completed grade.
 
@@
 The employment and demographic variables are then joined.
-These data are further filtered to the cohort who completed education up to 12th grade and participated in at least five rounds in the survey.
+These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation articulated in @SingerJudithD2003Alda.
+We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when the data is used to teach longitudinal data.
 We save the resultant wage data on this cohort as `wages`.
@@
 ```
+
+## Highlighting the work experience
 
@@
 
-### Replacing extreme values {#censor}
+## Replacing extreme values {#censor}
 
@@
 # Comparison of refreshed with the original data {#compare}
+
+## Calculated variables: Determining who is a dropout
 

### 2022-02-13 paper/paper.tex
Edited the droputs.
Edited the paper with respect to the changes in dataset.

@@
 For example, we create a categorical race variable instead of the binary variable ``Black'' and ``Hispanic''.
-We also include additional variables so that more interesting data exploration can be included in teaching examples: year of survey, age of individual in 1979, year when the highest grade was completed, number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
+We also include additional variables so that more interesting data exploration can be included in teaching examples: year of survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
 
-We will also make all of the high school graduates data available, including females, in addition to high school dropouts, and separate the demographic data of the NLSY79 cohort.
-Thus, the plan is to create three datasets as follows:
+We will also make the wages and demographic data of the NLSY79 cohort available, including females, in addition to high school dropouts. Thus, the plan is to create three datasets as follows:
 
@@
 \item
-  The wages data of the up to high school graduates.
+  The wages data of NLSY79 cohort.
 \item
@@
 \begin{verbatim}
-#> 'data.frame':    12686 obs. of  716 variables:
+#> 'data.frame':    12686 obs. of  663 variables:
 #>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 ...
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}) and the corresponding year this grade was completed (\texttt{yr\_hgc}).
+In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), and the whether the graduate equivalency diploma is obtained (\texttt{ged}).
 
@@
 
-The next step is tidying to obtain \texttt{hgc} and \texttt{yr\_hgc}.
-The highest grade completed are recorded in \texttt{new\_data\_qnames} as variables beginning with \texttt{Q3-4} and \texttt{HGC} with suffix of the year it was recorded.
-In addition, the variables beginning with \texttt{HGCREV} contain the revised data.
-We choose to use this revised data because it seemed to have less missing and presumably has been checked.
-However, there is no revised data for 2012, 2014, 2016, and 2018, thus, we use the ordinary data for these years.
-
-The \texttt{hgc} is measured and could be updated in each period of the survey.
-We chose to only retain the highest grade completed for each individual and derived the year when they completed it (\texttt{yr\_hgc}) by finding the minimum year with the highest completed grade.
+The next step is tidying to obtain \texttt{hgc} and \texttt{ged}.
+As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in years unit, we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type.
+For \texttt{ged}, we only separate the years and the GED status of the individual.
+Later on, this variable is useful to determine the refreshed high school dropouts data.
 
@@
 The employment and demographic variables are then joined.
-These data are further filtered to the cohort who completed education up to 12th grade and participated in at least five rounds in the survey.
+These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation articulated in Singer and Willett (2003).
+We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when the data is used to teach longitudinal data.
 We save the resultant wage data on this cohort as \texttt{wages}.
+
+\hypertarget{highlighting-the-work-experience}{%
+\subsection{Highlighting the work experience}\label{highlighting-the-work-experience}}
 
@@
 \hypertarget{censor}{%
-\subsubsection{Replacing extreme values}\label{censor}}
+\subsection{Replacing extreme values}\label{censor}}
 
@@
 \section{Comparison of refreshed with the original data}\label{compare}}
+
+\hypertarget{calculated-variables-determining-who-is-a-dropout}{%
+\subsection{Calculated variables: Determining who is a dropout}\label{calculated-variables-determining-who-is-a-dropout}}
 

### 2022-02-13 paper/paper.Rmd
intro edits part 1/2

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by @SingerJudithD2003Alda which can be used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
-The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
+An example of this, is the wages data made public by @SingerJudithD2003Alda which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel.
+The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
 
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis supervised by author Cook [@ilk2004], and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
+The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis [@ilk2004], and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
 
@@
 
-This disparity between the average and the individual is a part of statistics, as a discipline, that is given less attention.
+This disparity between the average and the individual is a part of statistics, as a discipline, that require more attention.
 It is this disparity which makes this particular data set important as a textbook example.
@@
 The most recent year in the data is 1994, 10 years prior to when @SingerJudithD2003Alda was published.
-Teachers of statistics need use contemporary data sets to show how techniques are relevant for today's students.
+Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students.
 Using tired old textbook data sets can imbue a misconception that the field is not current.

### 2022-02-14 paper/paper.Rmd
intro update part 2/2

@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis [@ilk2004], and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
+The wages data was used to illustrate exploratory longitudinal data analysis in @ilk2004, and was further developed into a case study for use in the teaching of exploratory data analysis at `r censor_word("Iowa State University")`.
 
-This textbook data was used as an example in @ilk2004, and was developed into a case study for use in teaching exploratory data analysis at `r censor_word("Iowa State University")`.
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
@@
 This disparity between the average and the individual is a part of statistics, as a discipline, that require more attention.
-It is this disparity which makes this particular data set important as a textbook example.
+This particular data set is a prime example of discussing this disparity.
 Textbook data sets have longevity if there is a unresolved mystery.
@@
 
-This paper demonstrates the steps of cleaning data, and documents the process, as recommended by @HuebnerMariannePhD2016Asat.
+This paper demonstrates the steps of cleaning data, including subjective decisions made on dealing with anomalies, and documents the process, as recommended by @HuebnerMariannePhD2016Asat.
 They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
@@
 The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using the data pipeline and split-apply-combine approach [@plyr].
-<!--AE comment: this is an important point that shouldn't be relegated to a parenthetical statement.
---> (Tidy data shouldn't be confused with "tame data" which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called `r yowie` which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
+Tidy data shouldn't be confused with "tame data" which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
+The resulting (tame) data is provided in a new R package called `r yowie` which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
 

### 2022-02-14 data-raw/ida.R
error for `rbind` for `tsibble`

@@ -323,7 +323,7 @@ join_eligible <- join %>% filter(age_1979 <= 17) %>%
 also_do <- wages %>% filter(id %in% join_eligible$id)
 
 # bind the initial dropouts data and the eligible data that are not captured in the first filtering
-wages_hs_do <- rbind(wages_hs_do, also_do)
+wages_hs_do <- bind_rows(wages_hs_do, also_do)
 
 ### ---- nrow
 a <- nrow(categories_qnames)

### 2022-02-14 paper/paper.Rmd
Add explanation in dropouts section.

--- 2022-02-13 paper/paper.Rmd
@@
 In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (`gender`), race (`race`), age (`age_1979`), highest grade completed (`hgc`), highest grade completed in terms of years, e.g.
-9th grade = 9, 3rd year college = 15, (`hgc_i`), and the whether the graduate equivalency diploma is obtained (`ged`).
+9th grade = 9, 3rd year college = 15, (`hgc_i`), and whether the graduate equivalency diploma is obtained (`ged`).
 
@@
 The next step is tidying to obtain `hgc` and `ged`.
-As the highest grade completed is already exist in  `new_data_qnames` and stored in years unit, we only rename the column name to be `hgc_i` and recode it so we have the factor data type.
+As the highest grade completed is already exist in  `new_data_qnames` and stored in year unit, we only rename the column name to be `hgc_i` and recode it so we have the factor data type.
 For `ged`, we only separate the years and the GED status of the individual.
@@
 
-## Calculated variables: Determining who is a dropout
-
 ```{r sw_wages}
@@
 
+## Calculated variables: Determining who is a dropout
+
+In the original data, there is no explicit explanation of how the dropouts cohort is obtained from the database.
+Hence, we use the high school dropouts criteria mentioned in @nlsy79edu to determine who is the dropouts, which are:
+
+1. The individual whose highest grade completed (`hgc`) is reported to be less than 12th grade, **or**
+2. The individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and have received a GED (`ged` is code to 2).
+
+Further, @SingerJudithD2003Alda set additional criteria for the dropouts in the textbook data, which are male and aged between 14 and 17 years old in 1979.
+Hence, we also use this information to re-create the dropouts data.
+
+With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
+We then matched the individuals in the original and refreshed dataset to investigate why some individuals in the original data were not captured in the refreshed data.
+We found several reasons for this as follows:
+
+1.
+173 individuals were more than 17 years old in 1979.
+Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in their subset.
+Our decision is to also include them in the refreshed data as the data has age variable in it, so analyst could filter them later.
+2. 79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:
+
+  + 35 of them completed at least 12th grade and having a diploma instead of GED (`ged` variable is coded to 1).
+Thus, they are indeed are not dropouts and we decide to not include them in the refreshed data.
+  + The information of `ged` is missing in 38 individuals. We decide to include them in the refreshed data. 
+  + 3 individuals are having both diploma and GED (`ged` is coded to 3). They essentially have GED, so we decide to keep them in the refreshed data as well. 
+  + 12 individuals do not exist in `wages` data because they were participated in less than 3 rounds of survey.
+
+We then redo the filtering based on those decisions we made on those individuals who were not captured in the refreshed data.
+Finally, we have 863 dropouts in the refreshed data, the closest number of individuals we get to the original data.
+
 <!--AE comment: what does the natural log have to do with the y-axis?-->
@@
 ```
+
+## The takeaways
 

### 2022-02-14 paper/paper.tex
Add explanation in dropouts section.

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
-The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
+An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel.
+The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
 Department of Labor 2021a).
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
+The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis (Ilk 2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
 
@@
 The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
-Teachers of statistics need use contemporary data sets to show how techniques are relevant for today's students.
+Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students.
 Using tired old textbook data sets can imbue a misconception that the field is not current.
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), and the whether the graduate equivalency diploma is obtained (\texttt{ged}).
+In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), and whether the graduate equivalency diploma is obtained (\texttt{ged}).
 
@@
 The next step is tidying to obtain \texttt{hgc} and \texttt{ged}.
-As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in years unit, we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type.
+As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in year unit, we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type.
 For \texttt{ged}, we only separate the years and the GED status of the individual.
@@
 
-\hypertarget{calculated-variables-determining-who-is-a-dropout}{%
-\subsection{Calculated variables: Determining who is a dropout}\label{calculated-variables-determining-who-is-a-dropout}}
-
 \begin{figure}
@@
 
+\hypertarget{calculated-variables-determining-who-is-a-dropout}{%
+\subsection{Calculated variables: Determining who is a dropout}\label{calculated-variables-determining-who-is-a-dropout}}
+
+In the original data, there is no explicit explanation of how the dropouts cohort is obtained from the database.
+Hence, we use the high school dropouts criteria mentioned in Wolpin (2005) to determine who is the dropouts, which are:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The individual whose highest grade completed (\texttt{hgc}) is reported to be less than 12th grade, \textbf{or}
+\item
+  The individual whose highest grade completed (\texttt{hgc}) is reported to be at least 12th grade and have received a GED (\texttt{ged} is code to 2).
+\end{enumerate}
+
+Further, Singer and Willett (2003) set additional criteria for the dropouts in the textbook data, which are male and aged between 14 and 17 years old in 1979.
+Hence, we also use this information to re-create the dropouts data.
+
+With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
+We then matched the individuals in the original and refreshed dataset to investigate why some individuals in the original data were not captured in the refreshed data.
+We found several reasons for this as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  173 individuals were more than 17 years old in 1979.
+Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in their subset.
+Our decision is to also include them in the refreshed data as the data has age variable in it, so analyst could filter them later.
+\item
+  79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:
+\end{enumerate}
+
+\begin{itemize}
+\tightlist
+\item
+  35 of them completed at least 12th grade and having a diploma instead of GED (\texttt{ged} variable is coded to 1).
+Thus, they are indeed are not dropouts and we decide to not include them in the refreshed data.
+\item
+  The information of \texttt{ged} is missing in 38 individuals. We decide to include them in the refreshed data.
+\item
+  3 individuals are having both diploma and GED (\texttt{ged} is coded to 3). They essentially have GED, so we decide to keep them in the refreshed data as well.
+\item
+  12 individuals do not exist in \texttt{wages} data because they were participated in less than 3 rounds of survey.
+\end{itemize}
+
+We then redo the filtering based on those decisions we made on those individuals who were not captured in the refreshed data.
+Finally, we have 863 dropouts in the refreshed data, the closest number of individuals we get to the original data.
+
 \begin{figure}
@@
 \end{figure}
+
+\hypertarget{the-takeaways}{%
+\subsection{The takeaways}\label{the-takeaways}}
 
@@
 \leavevmode\hypertarget{ref-nlsy79guide}{}%
----------.
-2021b.
-``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
-Columbus, OH.
-\url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
+---------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.'' \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
 
@@
 
+\leavevmode\hypertarget{ref-nlsy79edu}{}%
+Wolpin, Kenneth I.
+2005.
+``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Published by Bureau of Labor Statistics, U.S.
+Department of Labor.
+\url{https://www.bls.gov/opub/mlr/2005/02/art3full.pdf}.
+
 \leavevmode\hypertarget{ref-knitr}{}%

### 2022-02-14 data-raw/ida.R

@@ -302,6 +302,9 @@ wages_hs_do <- wages %>%
 
 
 # investigate the dropouts IDs with the IDs in the original data
+
+ori_do_nrow <- wages_hs_do %>% as_tibble() %>% count(id) %>% nrow()
+
 `%!in%` <- Negate(`%in%`)
 
 sw <- brolgar::wages
@@ -311,13 +314,29 @@ not_in_sw <- sw %>% filter(id %!in% wages_hs_do$id) %>% distinct(id, .keep_all =
 join <- left_join(not_in_sw, full_demographics, by = "id")
 
 # some IDs are more than 17 y.o
-join_more_17 <- join %>% filter(age_1979 > 17) #this IDs are not eligible
+join_more_17 <- join %>% filter(age_1979 > 17)
+#this IDs are not eligible according to original data definition, bu we decide to include them later on
 
-# investigate ID less than 17 y.o (should be eligible)
-# if these IDs are in teh wages data, they will be also included in the dropouts database
-join_eligible <- join %>% filter(age_1979 <= 17) %>%
-  filter((hgc_i >= 12 & dip_or_ged != 1) |
-           is.na(dip_or_ged))
+# the following should be eligible by age rule but somehow not included
+join_less_17 <- join %>% filter(age_1979 <= 17)
+
+# reasons:
+# some ID are not eligible as their hgc_i >= 12 and GED is coded to 1
+join_not_elig <- join_less_17 %>% filter(hgc_i >= 12 & dip_or_ged == 1)
+
+# the information of dip_or_ged is missing in some IDs, decide to include them
+join_missing_ged <- join_less_17 %>% filter(is.na(dip_or_ged))
+
+## the ged is coded to 3, we decide to include them in the data
+join_ged_both <- join_less_17 %>% filter(dip_or_ged == 3)
+
+## the IDs do not exist in the complete wages data
+join_not_exist <- join_less_17 %>% filter(id %!in% wages$id)
+
+# based on those decisions, we refilter those who are not included in the refreshed  data
+not_elig <- join %>% filter(hgc_i >= 12 & dip_or_ged == 1)
+
+join_eligible <- join %>% filter(id %!in% not_elig$id)
 
 # filter the wages data
 also_do <- wages %>% filter(id %in% join_eligible$id)

### 2022-02-14 paper/paper.Rmd
updates up to stwork

@@
 
-@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
+The book "Applied longitudinal data analysis" [@SingerJudithD2003Alda] used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
 This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN].
@@
 
-From the database, we will refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published).
-We also plan to make some modifications of the variables.
+We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published).
+We also made some modifications of the variables.
 For example, we create a categorical race variable instead of the binary variable "Black" and "Hispanic".
@@
 
-1. The demographic data of the NLSY79 cohort.
-2. The wages data of NLSY79 cohort.
+1. The demographic data of the whole NLSY79 cohort.
+2. The wages data of the whole NLSY79 cohort.
 3. The wages data of the high school dropouts, which is closest to a refreshed version of @SingerJudithD2003Alda's data.
@@
 As respondents can have multiple jobs at specific years, the column names, such as `HRP1_1979`, `HRP2_1979`, `HRP1_1980` and `HRP2_1980`, contain the information about the job number up to 5 (`HRP1` = job 1, `HRP2` = job 2) and the year.
-The raw data consequently has a large number of columns (716 to be specific).
+The raw data consequently has a large number of columns (`r ncol(new_data_qnames)` to be specific).
 The values in the cell under the variables that begin with `HRP` correspond to the hourly wage in dollars.
@@
 
-Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long form, with `dplyr` [@dplyr] and `stringr` [@stringr] to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
+Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
+This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long form, with `dplyr` [@dplyr] and `stringr` [@stringr] to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 
-The `new_data_qnames` contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981` which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
+The raw data, `new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981` which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.

### 2022-02-14 paper/paper.Rmd
inconsistency check required
an attempt in fixing the outlier section

@@
 As the measurement unit is in week, we convert this to year.
+[ET: inconsistency in `st_work` and `yr_workforce`?]
 
@@
 I might rephrase the first sentence as follows: “In order to treat the extreme values in the data, we built a robust linear regression model.” -->
-As part of the IDA, <!--AE comment: Awkward sentence.
-"which is the model formulation"?--> which is the model formulation, we build a robust linear regression model to treat the extreme values in the data.
-Robust linear regression yields an estimation robust to the influence of noise or contamination [@KollerManuel2016rARP].
-It also aims to detect the contamination by weighting each observation based on how "well-behaved" they are, known as robustness weight.
-An observation with a lower robustness weight would be suggested as an outlier [@KollerManuel2016rARP].
-
-Since we work with longitudinal data, we build the model for each ID instead of the overall data.
-The robust mixed model could be the best model to be employed in this case.
-However, this method is too computationally and memory expensive, especially for a large data set, like the NLSY79 data.
-<!--I would pull the tidy and purrr citations to the end of the sentence to make it read more smoothly-->Therefore, the model for each ID is built utilizing the `nest` and `map` function from `tidyr` [@tidyr] and `purrr` [@purrr], respectively.
-The full code for this is shown at \url{`r yowie_pkgdown`/articles/input-to-valid-data.html} but also described in detail next.
+<!--AE comment: Awkward sentence.
+"which is the model formulation"?--> To treat the extreme values in the data, we build a robust linear regression model where we use the robustness weight to determine if a value should be replaced with the fitted value from the model.
+We build the model for each ID utilizing the `nest` and `map` function from `tidyr` [@tidyr] and `purrr` [@purrr], respectively.
+This fits a model separately for each individual, and a robust linear mixed model could be used as an alternative approach.
+We tried fitting a robust linear mixed model using `robustlmm` [@KollerManuel2016rARP], however perhaps because the data is large and imbalanced, the model was not converging.
+<!--I would pull the tidy and purrr citations to the end of the sentence to make it read more smoothly-->The full code for this is shown at \url{`r yowie_pkgdown`/articles/input-to-valid-data.html} but also described in detail next.
 
@@
 Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm].
-<!--AE comment: “However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.” I don’t think this sentence needs a “However” and probably should say “past which” rather than “in which”-->However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
+<!--AE comment: “However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.” I don’t think this sentence needs a “However” and probably should say “past which” rather than “in which”-->The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
 Moreover, it should be noted that not all the outliers are due to an error.
@@
 
-To minimize the risk of mistakenly identifying an outlier as an "erroneous outlier", we simulate some thresholds and study how they affect the data.
-We find that 0.12 is the most reasonable value to be the threshold to minimize that drawback's risk because it still captures the sensible spikes in the data.
+To minimize the risk of mistakenly identifying an outlier as an "erroneous outlier", we simulate some thresholds and study how they affect the data by eye balling the results.
+Based on this eye balling, we find that 0.12 is the most reasonable value to use as the threshold to minimize the risk of overly smoothing the wage by retaining sensible spikes in the data.
 In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording.

### 2022-02-14 paper/paper.tex
an attempt in fixing the outlier section

@@
 
+\usepackage{longtable,booktabs,array}
+\usepackage{calc} % for calculating minipage widths
+% Correct order of tables after \paragraph or \subparagraph
+\usepackage{etoolbox}
+\makeatletter
+\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
+\makeatother
+% Allow footnotes in longtable head/foot
+\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
+\makesavenoteenv{longtable}
+
 % Pandoc citation processing
+\newlength{\csllabelwidth}
+\setlength{\csllabelwidth}{3em}
+\newlength{\cslhangindent}
+\setlength{\cslhangindent}{1.5em}
+% for Pandoc 2.8 to 2.10.1
+\newenvironment{cslreferences}%
+  {}%
+  {\par}
+% For Pandoc 2.11+
+\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
+ {% don't indent paragraphs
+  \setlength{\parindent}{0pt}
+  % turn on hanging indent if param 1 is 1
+  \ifodd #1 \everypar{\setlength{\hangindent}{\cslhangindent}}\ignorespaces\fi
+  % set entry spacing
+  \ifnum #2 > 0
+  \setlength{\parskip}{#2\baselineskip}
+  \fi
+ }%
+ {}
+\usepackage{calc} % for calculating minipage widths
+\newcommand{\CSLBlock}[1]{#1\hfill\break}
+\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
+\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
+\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
 
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis (Ilk 2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
-
-This textbook data was used as an example in Ilk (2004), and was developed into a case study for use in teaching exploratory data analysis at CENSORED.
+The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at CENSORED.
+
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
@@
 
-This paper demonstrates the steps of cleaning data, and documents the process, as recommended by Huebner, Vach, and Cessie (2016).
+This paper demonstrates the steps of cleaning data, including subjective decisions made on dealing with anomalies, and documents the process, as recommended by Huebner, Vach, and Cessie (2016).
 They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
@@
 
-Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
+Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al.
 2019).
 The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
-(Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
+Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
+The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
 
@@
 
-Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
+The book ``Applied longitudinal data analysis'' (Singer and Willett 2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
 This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al.
@@
 
-From the database, we will refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published).
-We also plan to make some modifications of the variables.
+We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published).
+We also made some modifications of the variables.
 For example, we create a categorical race variable instead of the binary variable ``Black'' and ``Hispanic''.
@@
 \item
-  The demographic data of the NLSY79 cohort.
-\item
-  The wages data of NLSY79 cohort.
+  The demographic data of the whole NLSY79 cohort.
+\item
+  The wages data of the whole NLSY79 cohort.
 \item
@@
 
+M.
 van der Loo and de Jonge (2018) in the context of official statistics describe the ``statistical value chain'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organised with correct type and identified variables) and valid data (data that has been cleaned and more accurately represents the intent of variables).
@@
 As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year.
-The raw data consequently has a large number of columns (716 to be specific).
+The raw data consequently has a large number of columns (663 to be specific).
 The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars.
@@
 
-Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
-2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham, François, et al.
+Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
+This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
+2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham et al.
 2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
@@
 
-The \texttt{new\_data\_qnames} contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981} which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
+The raw data, \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981} which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 As the measurement unit is in week, we convert this to year.
+{[}ET: inconsistency in \texttt{st\_work} and \texttt{yr\_workforce}?{]}
 
@@
 
-As part of the IDA, which is the model formulation, we build a robust linear regression model to treat the extreme values in the data.
-Robust linear regression yields an estimation robust to the influence of noise or contamination (Koller 2016).
-It also aims to detect the contamination by weighting each observation based on how ``well-behaved'' they are, known as robustness weight.
-An observation with a lower robustness weight would be suggested as an outlier (Koller 2016).
-
-Since we work with longitudinal data, we build the model for each ID instead of the overall data.
-The robust mixed model could be the best model to be employed in this case.
-However, this method is too computationally and memory expensive, especially for a large data set, like the NLSY79 data.
-Therefore, the model for each ID is built utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively.
+To treat the extreme values in the data, we build a robust linear regression model where we use the robustness weight to determine if a value should be replaced with the fitted value from the model.
+We build the model for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively.
+This fits a model separately for each individual, and a robust linear mixed model could be used as an alternative approach.
+We tried fitting a robust linear mixed model using \texttt{robustlmm} (Koller 2016), however perhaps because the data is large and imbalanced, the model was not converging.
 The full code for this is shown at \url{CENSORED/articles/input-to-valid-data.html} but also described in detail next.
@@
 Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021).
-However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
+The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
 Moreover, it should be noted that not all the outliers are due to an error.
@@
 
-To minimize the risk of mistakenly identifying an outlier as an ``erroneous outlier'', we simulate some thresholds and study how they affect the data.
-We find that 0.12 is the most reasonable value to be the threshold to minimize that drawback's risk because it still captures the sensible spikes in the data.
+To minimize the risk of mistakenly identifying an outlier as an ``erroneous outlier'', we simulate some thresholds and study how they affect the data by eye balling the results.
+Based on this eye balling, we find that 0.12 is the most reasonable value to use as the threshold to minimize the risk of overly smoothing the wage by retaining sensible spikes in the data.
 In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording.
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{/Users/etan0038/Dropbox/repo/yowie-paper/paper/figures/flowchart_blind}
 
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-iris-data}{}%
-Andereson, Edgar. 1935. ``The Irises of the Gaspe Peninsula.'' \emph{Bulletin of the American Iris Society} 59: 2--5.
-
-\leavevmode\hypertarget{ref-nlsy79}{}%
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-iris-data}{}}%
+Andereson, Edgar. 1935. {``The Irises of the Gaspe Peninsula.''} \emph{Bulletin of the American Iris Society} 59: 2--5.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
 Bureau of Labor Statistics, U.S.
@@
 2021a.
-``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
+{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
 Columbus, OH.
 
-\leavevmode\hypertarget{ref-nlsy79guide}{}%
----------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.'' \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
-
-\leavevmode\hypertarget{ref-shiny}{}%
-Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.
-
-\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
-Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
-
-\leavevmode\hypertarget{ref-eliznlsy}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-nlsy79guide}{}}%
+---------. 2021b. {``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.''} \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-shiny}{}}%
+Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{{shiny: Web Application Framework for R}}. \url{https://CRAN.R-project.org/package=shiny}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chatfield1985TIEo}{}}%
+Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
+
+\leavevmode\vadjust pre{\hypertarget{ref-eliznlsy}{}}%
 Cooksey, Elizabeth C.
 2017.
-``Using the National Longitudinal Surveys of Youth (Nlsy) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
+{``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M.
 Lerner Neal Halfon Christoper B.
@@
 Cham: Springer.
-\url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.
-
-\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
+https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-DasuTamraparni2003Edma}{}}%
 Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.
 
-\leavevmode\hypertarget{ref-janitor}{}%
-Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.
-
-\leavevmode\hypertarget{ref-racismnotrace}{}%
-Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.
-
-\leavevmode\hypertarget{ref-grimshaw}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-janitor}{}}%
+Firke, Sam. 2020. \emph{{janitor: Simple Tools for Examining and Cleaning Dirty Data}}. \url{https://CRAN.R-project.org/package=janitor}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-racismnotrace}{}}%
+Fullilove, M. T. 1998. {``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.
+
+\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
 Grimshaw, Scott D.
 2015.
-``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14.
+{``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14.
 \url{https://doi.org/10.1080/00031305.2015.1081106}.
 
-\leavevmode\hypertarget{ref-purrr}{}%
-Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.
-
-\leavevmode\hypertarget{ref-penguins-data}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-purrr}{}}%
+Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-penguins-data}{}}%
 Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
 
-\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
 Huebner, Marianne, Werner Vach, and Saskia le Cessie.
 2016.
-``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
-
-\leavevmode\hypertarget{ref-ilk2004}{}%
+{``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ilk2004}{}}%
 Ilk, Ozlem.
 2004.
-``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.'' PhD thesis, Iowa State University.
+{``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.''} PhD thesis, Iowa State University.
 \url{https://doi.org/10.31274/rtd-180813-11012}.
 
-\leavevmode\hypertarget{ref-tamedata}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
 Kim, A.
@@
 2018.
-``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1).
+{``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1).
 \url{https://doi.org/10.5070/T511103589}.
 
-\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
-Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.
-
-\leavevmode\hypertarget{ref-notaverage}{}%
-Moncrief, Marc. 2015. ``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.'' \url{https://bit.ly/smh-not-normal}.
-
-\leavevmode\hypertarget{ref-opendata}{}%
-Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.
-
-\leavevmode\hypertarget{ref-patchwork}{}%
-Pedersen, Thomas Lin. 2020. \emph{patchwork: The Composer of Plots}. \url{https://CRAN.R-project.org/package=patchwork}.
-
-\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-KollerManuel2016rARP}{}}%
+Koller, Manuel. 2016. {``{robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models}.''} \emph{Journal of Statistical Software} 75 (6): 1--24.
+
+\leavevmode\vadjust pre{\hypertarget{ref-notaverage}{}}%
+Moncrief, Marc. 2015. {``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.''} \url{https://bit.ly/smh-not-normal}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-opendata}{}}%
+Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-patchwork}{}}%
+Pedersen, Thomas Lin. 2020. \emph{{patchwork: The Composer of Plots}}. \url{https://CRAN.R-project.org/package=patchwork}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-MichaelRPergamit2001DWTN}{}}%
 Pergamit, Michael R., Charles R.
@@
 2001.
-``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
-
-\leavevmode\hypertarget{ref-R}{}%
+{``Data Watch: The National Longitudinal Surveys.''} \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
 R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
 
-\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-SingerJudithD2003Alda}{}}%
 Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.
 
-\leavevmode\hypertarget{ref-notiris}{}%
-Stodel, Megan. 2020. ``Stop Using Iris.'' \url{https://www.meganstodel.com/posts/no-to-iris/}.
-
-\leavevmode\hypertarget{ref-brolgar}{}%
-Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.
-
-\leavevmode\hypertarget{ref-tukey}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-notiris}{}}%
+Stodel, Megan. 2020. {``Stop Using Iris.''} \url{https://www.meganstodel.com/posts/no-to-iris/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-brolgar}{}}%
+Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}}. \url{https://github.com/njtierney/brolgar}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tukey}{}}%
 Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.
 
-\leavevmode\hypertarget{ref-rlm}{}%
-UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
-
-\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
-van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.
-
-\leavevmode\hypertarget{ref-validate}{}%
-van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
-
-\leavevmode\hypertarget{ref-mass}{}%
-Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
-
-\leavevmode\hypertarget{ref-tsibble}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-rlm}{}}%
+UCLA: Statistical Consulting Group. 2021. {``{Robust Regression \textbar{} R Data Analysis Examples}.''} February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%
+van der Loo, Mark P. J., and Edwin de Jonge. 2021. {``Data Validation Infrastructure for {R}.''} \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-LooMarkvander2018Sdcw}{}}%
+van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in r}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mass}{}}%
+Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with s}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tsibble}{}}%
 Wang, Earo, Dianne Cook, and Rob J Hyndman.
 2020.
-``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
+{``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.''} \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
 \url{https://doi.org/10.1080/10618600.2019.1695624}.
 
-\leavevmode\hypertarget{ref-plyr}{}%
-Wickham, Hadley. 2011. ``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
-
-\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
----------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.
-
-\leavevmode\hypertarget{ref-ggplot2}{}%
----------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
-
-\leavevmode\hypertarget{ref-stringr}{}%
----------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.
-
-\leavevmode\hypertarget{ref-tidyr}{}%
----------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.
-
-\leavevmode\hypertarget{ref-tidyverse}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
+Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-WickhamHadley2014TD}{}}%
+---------. 2014. {``Tidy Data.''} \emph{Journal of Statistical Software} 59 (10): 1--23.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
+---------. 2016. \emph{{ggplot2: Elegant Graphics for Data Analysis}}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-stringr}{}}%
+---------. 2019. \emph{{stringr: Simple, Consistent Wrappers for Common String Operations}}. \url{https://CRAN.R-project.org/package=stringr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tidyr}{}}%
+---------. 2020. \emph{{tidyr: Tidy Messy Data}}. \url{https://CRAN.R-project.org/package=tidyr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al.
 2019.
-``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686.
+{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686.
 \url{https://doi.org/10.21105/joss.01686}.
 
-\leavevmode\hypertarget{ref-dplyr}{}%
-Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.
-
-\leavevmode\hypertarget{ref-readr}{}%
-Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.
-
-\leavevmode\hypertarget{ref-nlsy79edu}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
+Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{{dplyr: A Grammar of Data Manipulation}}. \url{https://CRAN.R-project.org/package=dplyr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-readr}{}}%
+Wickham, Hadley, and Jim Hester. 2020. \emph{{readr: Read Rectangular Text Data}}. \url{https://CRAN.R-project.org/package=readr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nlsy79edu}{}}%
 Wolpin, Kenneth I.
 2005.
-``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Published by Bureau of Labor Statistics, U.S.
+{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Published by Bureau of Labor Statistics, U.S.
 Department of Labor.
@@
 
-\leavevmode\hypertarget{ref-knitr}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-knitr}{}}%
 Xie, Yihui.
 2014.
-``Knitr: A Comprehensive Tool for Reproducible Research in R.'' In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
+{``Knitr: A Comprehensive Tool for Reproducible Research in {R}.''} In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
 Peng.

### 2022-02-14 paper/paper.Rmd
Edited the explanation of stwork and yr_wforce.

@@
 
-Our target variables for the employment are to obtain respondent's mean hourly wage (`wage`), the number of jobs (`njobs`), the total hours of work per week (`hours`) for each survey year, the year when individual starting to work (`stwork`), the length of time (years) in workforce (`yr_workforce`), and work experience measured as the number of years worked (`exp`).
+Our target variables for the employment are to obtain respondent's mean hourly wage (`wage`), the number of jobs (`njobs`), the total hours of work per week (`hours`) for each survey year, the year when individual starting to work (`stwork`), the length of time (years) in workforce (`yr_wforce`), and work experience measured as the number of years worked (`exp`).
 As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.
@@
 
-```{r tidy-rate}
-```
-
 ```{r tidy-rate-hour}
@@
 
-For `st_work` variable,  we only rename the column names from `new_data_qnames`.
-This variable is then used to calculate the next variable, `yr_workforce` for each survey round, which is the year of survey (`year`) minus the year of individual start working (`st_work`).
+For `stwork` variable,  we only rename the column names from `new_data_qnames`.
+This variable is then used to calculate the next variable, `yr_wforce` for each survey round, which is the year of survey (`year`) minus the year of individual start working (`stwork`).
 Finally, `exp` variable is derived from the number of weeks worked since the last interview indicated as variable started with `WKSWK` in `new_data_qnames`.
@@
 As the measurement unit is in week, we convert this to year.
-[ET: inconsistency in `st_work` and `yr_workforce`?]
 
@@
 These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation articulated in @SingerJudithD2003Alda.
-We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when the data is used to teach longitudinal data.
+We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data.
 We save the resultant wage data on this cohort as `wages`.
@@
 
-## Highlighting the work experience
+## Calculated variables: work experience
+
+It is conceptually motivated by the variables available in the original data and it also illustrates how one would create a new variable from the information provided by the database, a variable that is more useful than those provided.
 
@@
 
-## Calculated variables: Determining who is a dropout
+## Filtering: Determining who is a dropout
 

### 2022-02-14 paper/paper.tex
Edited the explanation of stwork and yr_wforce.

--- 2022-02-14 paper/paper.tex
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis (Ilk 2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
+The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at CENSORED.
 
-This textbook data was used as an example in Ilk (2004), and was developed into a case study for use in teaching exploratory data analysis at CENSORED.
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
@@
 The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
-(Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.) The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
+Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
+The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
 
@@
 \item
-  The demographic data of the NLSY79 cohort.
+  The demographic data of the whole NLSY79 cohort.
 \item
-  The wages data of NLSY79 cohort.
+  The wages data of the whole NLSY79 cohort.
 \item
@@
 
-Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
+Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
+This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham, François, et al.
@@
 
-Our target variables for the employment are to obtain respondent's mean hourly wage (\texttt{wage}), the number of jobs (\texttt{njobs}), the total hours of work per week (\texttt{hours}) for each survey year, the year when individual starting to work (\texttt{stwork}), the length of time (years) in workforce (\texttt{yr\_workforce}), and work experience measured as the number of years worked (\texttt{exp}).
+Our target variables for the employment are to obtain respondent's mean hourly wage (\texttt{wage}), the number of jobs (\texttt{njobs}), the total hours of work per week (\texttt{hours}) for each survey year, the year when individual starting to work (\texttt{stwork}), the length of time (years) in workforce (\texttt{yr\_wforce}), and work experience measured as the number of years worked (\texttt{exp}).
 As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.
@@
 
-For \texttt{st\_work} variable, we only rename the column names from \texttt{new\_data\_qnames}.
-This variable is then used to calculate the next variable, \texttt{yr\_workforce} for each survey round, which is the year of survey (\texttt{year}) minus the year of individual start working (\texttt{st\_work}).
+For \texttt{stwork} variable, we only rename the column names from \texttt{new\_data\_qnames}.
+This variable is then used to calculate the next variable, \texttt{yr\_wforce} for each survey round, which is the year of survey (\texttt{year}) minus the year of individual start working (\texttt{stwork}).
 Finally, \texttt{exp} variable is derived from the number of weeks worked since the last interview indicated as variable started with \texttt{WKSWK} in \texttt{new\_data\_qnames}.
@@
 As the measurement unit is in week, we convert this to year.
+{[}ET: inconsistency in \texttt{st\_work} and \texttt{yr\_wforce}?{]}
 
@@
 These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation articulated in Singer and Willett (2003).
-We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when the data is used to teach longitudinal data.
+We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data.
 We save the resultant wage data on this cohort as \texttt{wages}.
 
-\hypertarget{highlighting-the-work-experience}{%
-\subsection{Highlighting the work experience}\label{highlighting-the-work-experience}}
+\hypertarget{calculated-variables-work-experience}{%
+\subsection{Calculated variables: work experience}\label{calculated-variables-work-experience}}
+
+It is conceptually motivated by the variables available in the original data and it also illustrates how one would create a new variable from the information provided by the database, a variable that is more useful than those provided.
 
@@
 
-\hypertarget{calculated-variables-determining-who-is-a-dropout}{%
-\subsection{Calculated variables: Determining who is a dropout}\label{calculated-variables-determining-who-is-a-dropout}}
+\hypertarget{filtering-determining-who-is-a-dropout}{%
+\subsection{Filtering: Determining who is a dropout}\label{filtering-determining-who-is-a-dropout}}
 

### 2022-02-14 data-raw/data-preprocessing.R

@@ -210,11 +210,11 @@ get_rate <- function(year) {
 rates_all <- map_dfr(years, get_rate)
 
 ## ---- tidy-start-work
-stwork <- new_data_qnames %>%
+st_work <- new_data_qnames %>%
   select(CASEID_1979,
          `EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND`) %>%
   rename(id = CASEID_1979,
-         stwork_year = `EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND`)
+         stwork = `EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND`)
 
 ## ---- tidy-work-experience
 
@@ -305,9 +305,9 @@ head(mean_hourly_wage, n = 10)
 
 # join the wages information and the demographic information by case id.
 wages_demog <- left_join(mean_hourly_wage, full_demographics, by="id") %>%
-  left_join(stwork, by = "id") %>%
+  left_join(st_work, by = "id") %>%
   left_join(exp, by = c("id", "year")) %>%
-  mutate(years_in_workforce = year - stwork_year)
+  mutate(yr_wforce = year - stwork)
 
 # # filter only the id with high school education
 # wages_before <- wages_demog  %>% filter(grepl("GRADE", hgc))

### 2022-02-14 data-raw/ida.R

@@ -260,7 +260,7 @@ wages_clean <- wages_cleaned %>%
 # rename and select the wages in tidy
 wages <- wages_clean %>%
   select(id, year, mean_hourly_wage, age_1979, gender, race, hgc, hgc_i, ged = dip_or_ged,
-                number_of_jobs, total_hours, stwork_year, years_in_workforce, exp, is_wm, is_pred) %>%
+         number_of_jobs, total_hours, stwork, yr_wforce, exp, is_wm, is_pred) %>%
   mutate(id = as.factor(id),
          hgc = as.factor(hgc),
          year = as.integer(year),
@@ -269,9 +269,7 @@ wages <- wages_clean %>%
          number_of_jobs = as.integer(number_of_jobs)) %>%
   rename(wage = mean_hourly_wage,
          njobs = number_of_jobs,
-         hours = total_hours,
-         stwork = stwork_year,
-         yr_wforce = years_in_workforce) %>%
+         hours = total_hours) %>%
   as_tsibble(key = id,
              index = year,
              regular = FALSE)

### 2022-02-15 paper/paper.Rmd
Add subsection calculated experience.

@@
 
-It is conceptually motivated by the variables available in the original data and it also illustrates how one would create a new variable from the information provided by the database, a variable that is more useful than those provided.
+Work experience is one of the "key" variables in @SingerJudithD2003Alda as it indicates time variable, instead of using the survey year.
+Thus, in the spirit of refreshing the data to the newest round of the survey, we also want to include this variable.
+However,it is not straightforwardly available in the database.
+Thus, we take the definition of experience mentioned by @SingerJudithD2003Alda to recreate it with the information already provided in the database.
+
+According to @SingerJudithD2003Alda, experience is years after entering the labor force.
+It identifies the specific day in which individual entering the labor force associated with each wages.
+Based on that definition, one would assume that experience must be related to the date of individual starting to work.
+However, after calculating it (`yr_wforce`), we found that the number we obtained was not quite match with the one in @SingerJudithD2003Alda's data.
+On the other hand, the [topical guide of NLSY79 data](https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience), provided by @nlsy79guide, has the information of variable related to labor force experience.
+One of the variables mentioned is "number of weeks worked since last interview".
+We then calculated the experience variables using this information.
+Since it is only measured since the last interview, to adjust it with the experience since the individual entering the labor force, we calculate the cumulative value.
+We found a more appropriate results in terms of its closeness with the original data (see Subsection \@ref(takeaways)).
+
+<!--It is conceptually motivated by the variables available in the original data and it also illustrates how one would create a new variable from the information provided by the database, a variable that is more useful than those provided.-->
 
@@
 
-## The takeaways
+## The takeaways {#takeaways}
 

### 2022-02-15 paper/paper.tex
Add subsection calculated experience.

@@
 As the measurement unit is in week, we convert this to year.
-{[}ET: inconsistency in \texttt{st\_work} and \texttt{yr\_wforce}?{]}
 
@@
 
-It is conceptually motivated by the variables available in the original data and it also illustrates how one would create a new variable from the information provided by the database, a variable that is more useful than those provided.
+Work experience is one of the ``key'' variables in Singer and Willett (2003) as it indicates time variable, instead of using the survey year.
+Thus, in the spirit of refreshing the data to the newest round of the survey, we also want to include this variable.
+However,it is not straightforwardly available in the database.
+Thus, we take the definition of experience mentioned by Singer and Willett (2003) to recreate it with the information already provided in the database.
+
+According to Singer and Willett (2003), experience is years after entering the labor force.
+It identifies the specific day in which individual entering the labor force associated with each wages.
+Based on that definition, one would assume that experience must be related to the date of individual starting to work.
+However, after calculating it (\texttt{yr\_wforce}), we found that the number we obtained was not quite match with the one in Singer and Willett (2003)'s data.
+On the other hand, the \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{topical guide of NLSY79 data}, provided by Bureau of Labor Statistics, U.S.
+Department of Labor (2021b), has the information of variable related to labor force experience.
+One of the variables mentioned is ``number of weeks worked since last interview''.
+We then calculated the experience variables using this information.
+Since it is only measured since the last interview, to adjust it with the experience since the individual entering the labor force, we calculate the cumulative value.
+We found a more appropriate results in terms of its closeness with the original data (see Subsection \ref{takeaways}).
 
@@
 
-\hypertarget{the-takeaways}{%
-\subsection{The takeaways}\label{the-takeaways}}
+\hypertarget{takeaways}{%
+\subsection{The takeaways}\label{takeaways}}
 

### 2022-02-15 paper/paper.Rmd
abstract and conclusions

@@
 abstract: > 
- Textbook data is useful for teaching statistics and data science methods because it is highly processed and clean.
-Where possible textbook data sets should be refreshed regularly, because it is also important to use contemporary data for teaching.
-This paper describes the trials and tribulations of refreshing the wages data extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s.
+ Textbook data is essential for teaching statistics and data science methods because they are clean, allowing the instructor to  focus on methodology.
+Ideally textbook data sets are refreshed regularly, especially when they are subsets taken from an on-going data collection.
+It is also important to use contemporary data for teaching, to imbue the sense that the methodology is relevant today.
+This paper describes the trials and tribulations of refreshing a textbook data set on wages, extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s.
 The data is useful for teaching modeling and exploratory analysis of longitudinal data.
@@
 Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future.
-The process was made more difficult because the steps and decisions taken to get from the raw data to the original textbook data were not clearly articulated, so we have been diligent to provide a reproducible workflow for others to follow.
+Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset, have not been clearly articulated.
+We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching.
 Three new data sets and the code to produce them are provided in the open source R package, called [CENSORED].
@@
 
-This paper has described the stages to take a particular open data set and make it a textbook data set, ready for the classroom or research.
+This paper has illustrated the steps and decisions made to take a particular open data set and make it a textbook data set, ready for the classroom or research.
 In the first stage, we showed the steps performed to get the data from the NLSY79 database.
-The data format needed conversion to tidy format, and this was described.
-After that, an initial data analysis was conducted to investigate and screen the quality of the data.
-We found and fixed the anomalous observations in wages using a robust linear regression model.
-Finally a comparison is made between the original and refreshed data sets.
+The data format was converted to tidy format, for more flexibility in cleaning and exploring.
+An initial data analysis was conducted to investigate and screen the quality of the data.
+We found and provided a fix for many anomalous observations in wages using a robust linear regression model.
+The refreshed data is compared with the original set using a variety of numerical summaries and graphics.
+The current subset is made available in a new R package, called `r yowie`.
 
-The data cleaning process is documented and the code has been made available. These provide the opportunity to again refresh the textbook data as updated data is published in the NLSY79 database.
-Determining the appropriate robustness weight to threshold the anomalous observations is documented, and a `shiny`  [@shiny] app is provided to assist. 
-The current subset is made available in a new R package, called `r yowie`.
+The data cleaning process is documented and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
+Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a `shiny`  [@shiny] app, and the choices used in the refreshed data are documented.
 
@@
 - determining which records should be downloaded from the database.
-- there are many errors in the data, e.g. hourly wages greater than $30,000 per hour.
+- calculating experience in the workforce requires comparing the date of first job with the first year the individual was recorded, both of which are available in the database. 
+- there are many unusually high hourly wages, e.g. greater than $30,000 per hour.
 - there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with the individual's age.
-<!-- calculating experience in the workforce would require knowing the time of first job within the first year the individual was recorded.-->
+- matching id's from the original data with those in the refreshed data, do refer to the same person based on the demographic information available.
 
 Ultimately, the refreshed data is reasonably similar to the original, but unsatisfactorily far from it.
-The last step required would be to inflation-adjust wages, but this is better to do with each wave of new data added, so that it is relative to the last date in the data.
+The last step required would be to inflation-adjust wages.
+This is better to do with each wave of new data added, so that it is relative to the last date in the data.
+Our decision was to provide the raw wages, and include code to do the adjustment as part of the package.
 
@@
 We do not assert that we have produced the best textbook data, but rather we describe our journey to provide a reasonable textbook data set.
-All code and documentation are provided for transparency, and future updates of the `r yowie` package may contain additional variables, or filters of the full set, if it is deemed important.
+All code and documentation are provided for transparency.
+Readers could use this to make different decisions, or provide suggestions on through package for better choices.
+Future updates of the `r yowie` package may contain additional variables, or filters of the full set, if it is deemed important.
 
-For the data providers, we recommend that a validation system with clear rules is added on data entry, and that alternative output formats, such as a tidy format would help users make better use of the resource.
+For the data providers, we recommend that a better validation system with clear rules applied at data entry, and that alternative output formats, such as a tidy format would help users make better use of their resource.
 The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs.

### 2022-02-15 paper/paper.tex
abstract and conclusions

--- 2022-02-14 paper/paper.tex
@@
 \usepackage{xcolor}
-
 \begin{document}
@@
 
-\def\tightlist{}
-
 \begin{abstract}
-Textbook data is useful for teaching statistics and data science methods because it is highly processed and clean.
-Where possible textbook data sets should be refreshed regularly, because it is also important to use contemporary data for teaching.
-This paper describes the trials and tribulations of refreshing the wages data extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s.
+Textbook data is essential for teaching statistics and data science methods because they are clean, allowing the instructor to focus on methodology.
+Ideally textbook data sets are refreshed regularly, especially when they are subsets taken from an on-going data collection.
+It is also important to use contemporary data for teaching, to imbue the sense that the methodology is relevant today.
+This paper describes the trials and tribulations of refreshing a textbook data set on wages, extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s.
 The data is useful for teaching modeling and exploratory analysis of longitudinal data.
@@
 Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future.
-The process was made more difficult because the steps and decisions taken to get from the raw data to the original textbook data were not clearly articulated, so we have been diligent to provide a reproducible workflow for others to follow.
+Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset, have not been clearly articulated.
+We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching.
 Three new data sets and the code to produce them are provided in the open source R package, called {[}CENSORED{]}.
@@
 As the measurement unit is in week, we convert this to year.
-{[}ET: inconsistency in \texttt{st\_work} and \texttt{yr\_workforce}?{]}
 
@@
 These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation articulated in Singer and Willett (2003).
-We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when the data is used to teach longitudinal data.
+We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data.
 We save the resultant wage data on this cohort as \texttt{wages}.
 
-\hypertarget{highlighting-the-work-experience}{%
-\subsection{Highlighting the work experience}\label{highlighting-the-work-experience}}
+\hypertarget{calculated-variables-work-experience}{%
+\subsection{Calculated variables: work experience}\label{calculated-variables-work-experience}}
+
+Work experience is one of the ``key'' variables in Singer and Willett (2003) as it indicates time variable, instead of using the survey year.
+Thus, in the spirit of refreshing the data to the newest round of the survey, we also want to include this variable.
+However,it is not straightforwardly available in the database.
+Thus, we take the definition of experience mentioned by Singer and Willett (2003) to recreate it with the information already provided in the database.
+
+According to Singer and Willett (2003), experience is years after entering the labor force.
+It identifies the specific day in which individual entering the labor force associated with each wages.
+Based on that definition, one would assume that experience must be related to the date of individual starting to work.
+However, after calculating it (\texttt{yr\_wforce}), we found that the number we obtained was not quite match with the one in Singer and Willett (2003)'s data.
+On the other hand, the \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{topical guide of NLSY79 data}, provided by Bureau of Labor Statistics, U.S.
+Department of Labor (2021b), has the information of variable related to labor force experience.
+One of the variables mentioned is ``number of weeks worked since last interview''.
+We then calculated the experience variables using this information.
+Since it is only measured since the last interview, to adjust it with the experience since the individual entering the labor force, we calculate the cumulative value.
+We found a more appropriate results in terms of its closeness with the original data (see Subsection \ref{takeaways}).
 
@@
 
-This paper has described the stages to take a particular open data set and make it a textbook data set, ready for the classroom or research.
+This paper has illustrated the steps and decisions made to take a particular open data set and make it a textbook data set, ready for the classroom or research.
 In the first stage, we showed the steps performed to get the data from the NLSY79 database.
-The data format needed conversion to tidy format, and this was described.
-After that, an initial data analysis was conducted to investigate and screen the quality of the data.
-We found and fixed the anomalous observations in wages using a robust linear regression model.
-Finally a comparison is made between the original and refreshed data sets.
-
-The data cleaning process is documented and the code has been made available. These provide the opportunity to again refresh the textbook data as updated data is published in the NLSY79 database.
-Determining the appropriate robustness weight to threshold the anomalous observations is documented, and a \texttt{shiny} (Chang et al. 2020) app is provided to assist.
+The data format was converted to tidy format, for more flexibility in cleaning and exploring.
+An initial data analysis was conducted to investigate and screen the quality of the data.
+We found and provided a fix for many anomalous observations in wages using a robust linear regression model.
+The refreshed data is compared with the original set using a variety of numerical summaries and graphics.
 The current subset is made available in a new R package, called \texttt{[CENSORED]}.
 
+The data cleaning process is documented and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
+Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a \texttt{shiny} (Chang et al.
+2020) app, and the choices used in the refreshed data are documented.
+
 Various difficulties were encountered in trying to refresh the data, which include:
@@
 \item
-  there are many errors in the data, e.g.~hourly wages greater than \$30,000 per hour.
+  calculating experience in the workforce requires comparing the date of first job with the first year the individual was recorded, both of which are available in the database.
+\item
+  there are many unusually high hourly wages, e.g.~greater than \$30,000 per hour.
 \item
   there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with the individual's age.
+\item
+  matching id's from the original data with those in the refreshed data, do refer to the same person based on the demographic information available.
 \end{itemize}
@@
 Ultimately, the refreshed data is reasonably similar to the original, but unsatisfactorily far from it.
-The last step required would be to inflation-adjust wages, but this is better to do with each wave of new data added, so that it is relative to the last date in the data.
+The last step required would be to inflation-adjust wages.
+This is better to do with each wave of new data added, so that it is relative to the last date in the data.
+Our decision was to provide the raw wages, and include code to do the adjustment as part of the package.
 
@@
 We do not assert that we have produced the best textbook data, but rather we describe our journey to provide a reasonable textbook data set.
-All code and documentation are provided for transparency, and future updates of the \texttt{[CENSORED]} package may contain additional variables, or filters of the full set, if it is deemed important.
-
-For the data providers, we recommend that a validation system with clear rules is added on data entry, and that alternative output formats, such as a tidy format would help users make better use of the resource.
+All code and documentation are provided for transparency.
+Readers could use this to make different decisions, or provide suggestions on through package for better choices.
+Future updates of the \texttt{[CENSORED]} package may contain additional variables, or filters of the full set, if it is deemed important.
+
+For the data providers, we recommend that a better validation system with clear rules applied at data entry, and that alternative output formats, such as a tidy format would help users make better use of their resource.
 The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs.

### 2022-02-15 paper/paper.Rmd
target data

@@
 We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published).
-We also made some modifications of the variables.
-For example, we create a categorical race variable instead of the binary variable "Black" and "Hispanic".
-We also include additional variables so that more interesting data exploration can be included in teaching examples: year of survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
+We also modify some variables.
+For example, use a single categorical race variable instead of the two binary race variables.
+We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), the number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
+We do not attempt to re-create the unemployment rate variable.
 
-We will also make the wages and demographic data of the NLSY79 cohort available, including females, in addition to high school dropouts. Thus, the plan is to create three datasets as follows:
+The plan is to create three datasets as follows:
 
-1. The demographic data of the whole NLSY79 cohort.
-2. The wages data of the whole NLSY79 cohort.
+1. The wages data of the whole NLSY79 cohort, including females.
+2. A separate table of the demographic data of the whole NLSY79 cohort.
 3. The wages data of the high school dropouts, which is closest to a refreshed version of @SingerJudithD2003Alda's data.

### 2022-02-15 paper/paper.tex
target data

@@
 We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published).
-We also made some modifications of the variables.
-For example, we create a categorical race variable instead of the binary variable ``Black'' and ``Hispanic''.
-We also include additional variables so that more interesting data exploration can be included in teaching examples: year of survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
-
-We will also make the wages and demographic data of the NLSY79 cohort available, including females, in addition to high school dropouts. Thus, the plan is to create three datasets as follows:
+We also modify some variables.
+For example, use a single categorical race variable instead of the two binary race variables.
+We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), the number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
+We do not attempt to re-create the unemployment rate variable.
+
+The plan is to create three datasets as follows:
 
@@
 \item
-  The demographic data of the whole NLSY79 cohort.
-\item
-  The wages data of the whole NLSY79 cohort.
+  The wages data of the whole NLSY79 cohort, including females.
+\item
+  A separate table of the demographic data of the whole NLSY79 cohort.
 \item

### 2022-02-15 paper/paper.Rmd
work experience section

@@
 
-Work experience is one of the "key" variables in @SingerJudithD2003Alda as it indicates time variable, instead of using the survey year.
-Thus, in the spirit of refreshing the data to the newest round of the survey, we also want to include this variable.
-However,it is not straightforwardly available in the database.
-Thus, we take the definition of experience mentioned by @SingerJudithD2003Alda to recreate it with the information already provided in the database.
+Work experience is one of the most important variables in @SingerJudithD2003Alda as it indicates time, and makes longitudinal analysis possible.
+It is desirable to calculate this rather than using the survey year for time because it more accurately reflects a person's time in the workforce.
+Thus, in the spirit of refreshing the data to the newest round of the survey, this variable needs to be calculated from other variables provided.
+It is not straightforward.
+We start with the definition of experience in @SingerJudithD2003Alda.
 
-According to @SingerJudithD2003Alda, experience is years after entering the labor force.
-It identifies the specific day in which individual entering the labor force associated with each wages.
-Based on that definition, one would assume that experience must be related to the date of individual starting to work.
-However, after calculating it (`yr_wforce`), we found that the number we obtained was not quite match with the one in @SingerJudithD2003Alda's data.
-On the other hand, the [topical guide of NLSY79 data](https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience), provided by @nlsy79guide, has the information of variable related to labor force experience.
-One of the variables mentioned is "number of weeks worked since last interview".
-We then calculated the experience variables using this information.
-Since it is only measured since the last interview, to adjust it with the experience since the individual entering the labor force, we calculate the cumulative value.
-We found a more appropriate results in terms of its closeness with the original data (see Subsection \@ref(takeaways)).
+Experience is years after entering the labor force.
+It represents the difference between the day an individual enters the labor force (`EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND` in the database) relative to the date of the survey, which we call `yr_wforce`.
+However, using this calculation produced numbers which don't quite match with the original data.
+
+Reading the section titled ["Topical Guide to the Data"](https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience) in the guide, suggests that it should be calculated based on the variable "number of weeks worked since last interview".
+This would remove periods of unemployment which makes sense when measuring experience while actually working.
+Since it is only measured since the last interview, this needs to be cumulated for each survey year.
+This produced results more similar to the original data (see Subsection \@ref(takeaways)).
 

### 2022-02-15 paper/paper.tex
work experience section

@@
 
-Work experience is one of the ``key'' variables in Singer and Willett (2003) as it indicates time variable, instead of using the survey year.
-Thus, in the spirit of refreshing the data to the newest round of the survey, we also want to include this variable.
-However,it is not straightforwardly available in the database.
-Thus, we take the definition of experience mentioned by Singer and Willett (2003) to recreate it with the information already provided in the database.
-
-According to Singer and Willett (2003), experience is years after entering the labor force.
-It identifies the specific day in which individual entering the labor force associated with each wages.
-Based on that definition, one would assume that experience must be related to the date of individual starting to work.
-However, after calculating it (\texttt{yr\_wforce}), we found that the number we obtained was not quite match with the one in Singer and Willett (2003)'s data.
-On the other hand, the \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{topical guide of NLSY79 data}, provided by Bureau of Labor Statistics, U.S.
-Department of Labor (2021b), has the information of variable related to labor force experience.
-One of the variables mentioned is ``number of weeks worked since last interview''.
-We then calculated the experience variables using this information.
-Since it is only measured since the last interview, to adjust it with the experience since the individual entering the labor force, we calculate the cumulative value.
-We found a more appropriate results in terms of its closeness with the original data (see Subsection \ref{takeaways}).
+Work experience is one of the most important variables in Singer and Willett (2003) as it indicates time, and makes longitudinal analysis possible.
+It is desirable to calculate this rather than using the survey year for time because it more accurately reflects a person's time in the workforce.
+Thus, in the spirit of refreshing the data to the newest round of the survey, this variable needs to be calculated from other variables provided.
+It is not straightforward.
+We start with the definition of experience in Singer and Willett (2003).
+
+Experience is years after entering the labor force.
+It represents the difference between the day an individual enters the labor force (\texttt{EMPLOYERS\_ALL\_STARTDATE\_ORIGINAL.01\textasciitilde{}Y\_XRND} in the database) relative to the date of the survey, which we call \texttt{yr\_wforce}.
+However, using this calculation produced numbers which don't quite match with the original data.
+
+Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide, suggests that it should be calculated based on the variable ``number of weeks worked since last interview''.
+This would remove periods of unemployment which makes sense when measuring experience while actually working.
+Since it is only measured since the last interview, this needs to be cumulated for each survey year.
+This produced results more similar to the original data (see Subsection \ref{takeaways}).
 
@@
 
-\caption{\label{tab:age-table}The frequency table of the age at the start of the survey in NSLY79 cohort}
+\caption{\label{tab:age-table}Frequency table of the age at the start of the survey in NSLY79 cohort}
 \centering
@@
 
-\caption{\label{tab:gender-race-table}The contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
+\caption{\label{tab:gender-race-table}Contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
 \centering
@@
 {``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
-Columbus, OH.
+Columbus, OH, through \url{https://www.nlsinfo.org/bibliography-citing-nls-data}.
 

### 2022-02-15 data-raw/ida.R

@@ -13,7 +13,7 @@ age_table <- full_demographics %>%
 age_table %>%
   mutate(n = scales::comma(n)) %>%
   kable(
-      caption = "The frequency table of the age at the start of the survey in NSLY79 cohort",
+      caption = "Frequency table of the age at the start of the survey in NSLY79 cohort",
       col.names = c("Age", "Number of individuals"),
       booktabs = TRUE,
       linesep = "",
@@ -38,7 +38,7 @@ gender_race_table <- full_demographics %>%
          Total = str_replace(Total, "\\([1|2]00\\%\\)", ""))
 
 kable(gender_race_table,
-      caption = "The contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1\\%) is out of the total corresponding to row.",
+      caption = "Contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1\\%) is out of the total corresponding to row.",
       col.names = c("Gender", "Hispanic", "Black", "Non-Black, Non-Hispanic", "Total"),
       booktabs = TRUE,
       linesep = "",

### 2022-02-15 data-raw/eda.R
dropouts plot

@@ -190,9 +190,12 @@ ggplot(do_sw_join, aes(x = xp,
   geom_point(alpha = 0.1) +
   geom_abline(intercept = 0, slope = 1) +
   guides(color = "none") +
+  #scale_x_sqrt() +
+  #scale_y_sqrt() +
   xlab("Work experience (Singer & Willet, 2008)") +
   ylab("Work experience (refreshed)") +
-  theme_bw()
+  theme_bw() +
+  theme(aspect.ratio=1)
 
 
 # Takeaways:

### 2022-02-15 paper/paper.tex
dropouts plot
see XXX in eda.r for mismatch of hgc

@@
 
+\begin{verbatim}
+#> # A tibble: 2 x 3
+#>   hgc   hgc12     n
+#>   <chr> <chr> <int>
+#> 1 12TH  8TH       1
+#> 2 8TH   12TH    243
+\end{verbatim}
+
 \begin{figure}
@@
 Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
-We can see that the correlation is reasonably high between them even though the exactly same figure could not be obtained.}\label{fig:compare-xp-sw}
+We can see that the correlation is reasonably high between them even there are different values for many individuals.}\label{fig:compare-xp-sw}
 \end{figure}

### 2022-02-15 data-raw/data-preprocessing.R

@@ -258,6 +258,8 @@ eligible_wages <- hours_wages %>%
 # calculate the mean_hourly_wage
 # flag1 = code 1 for weighted mean
 # code 0 for arithmetic mean
+# XXX we lose 346 individuals here, should they be kept?
+# with NAs for mean hourly wage?
 mean_hourly_wage <- eligible_wages %>%
   group_by(id, year) %>%
   #calculate the weighted mean if the number of jobs > 1
@@ -314,7 +316,11 @@ wages_demog <- left_join(mean_hourly_wage, full_demographics, by="id") %>%
 # # calculate the number of observation
 keep_me <- wages_demog %>%
   count(id) %>%
-  filter(n >= 3)
+  filter(n >= 3) # XXX why do we need to filter here?
+# is it because this data is processed with rlm only
+# when there are 3 or more cases?
+# how does it get joined again with the individuals
+# with less than 3 measurements?
 
 wages_before <- wages_demog %>%
   filter(id %in% keep_me$id)

### 2022-02-15 data-raw/eda.R

@@ -35,7 +35,8 @@ sw_wages_mod <- sw %>%
   ylab("Hourly wage ($, natural log)") +
   xlab("Experience in years") +
   theme(legend.position = "bottom",
-        legend.direction = "horizontal")
+        legend.direction = "horizontal") +
+  xlim(c(0,15))
 
 ## ---- do_refreshed
 
@@ -47,7 +48,7 @@ sw_wages_mod <- sw %>%
 
 # filter do data up to 1994 since the textbook data only covers that period.
 
-do <- yowie::wages_hs_do %>%
+do <- wages_hs_do %>% #yowie::wages_hs_do %>%
   as_tibble() %>%
   group_by(id) %>%
   mutate(lnwage = log(wage),
@@ -78,13 +79,27 @@ do_ref_mod <- do %>%
   ylab("Hourly wage ($, natural log)") +
   xlab("Experience (years)") +
   theme(legend.position = "bottom",
-        legend.direction = "horizontal")
+        legend.direction = "horizontal") +
+  xlim(c(0,15))
 
 # sw_wages + do_ref
 
 # the two plots are not comparable
 # since the id who are in Singer-Willet and the id in do-refreshed might be different
 
+# XXX Code to check that the hgc matches
+do_check <- do %>%
+  mutate(hgc12 = ifelse(hgc_i < 9, "8TH", "12TH")) %>%
+  select(id, hgc12) %>%
+  distinct()
+sw_check <- sw %>%
+  as_tibble() %>%
+  mutate(hgc = ifelse(high_grade < 9, "8TH", "12TH")) %>%
+  select(id, hgc) %>%
+  distinct()
+
+sw_check <- sw_check %>% left_join(do_check, by="id")
+sw_check %>% filter(hgc != hgc12) %>% count(hgc, hgc12)
 
 ## --- compare-sw-do
 sw_id <- as_tibble(sw) %>%

### 2022-02-15 data-raw/ida.R

@@ -5,6 +5,8 @@ library(tsibble)
 library(brolgar)
 library(patchwork)
 
+# Note: full_demographics has 12686 individuals
+
 ## ---- age-table
 age_table <- full_demographics %>%
   group_by(age_1979) %>%

### 2022-02-15 paper/paper.Rmd
dropouts plot
revised Filtering: Determining who is a dropout

@@
 Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
-We can see that the correlation is reasonably high between them even though the exactly same figure could not be obtained.", out.width = "70%"}
+We can see that the correlation is reasonably high between them even there are different values for many individuals.", out.width = "70%"}
 ```
@@
 
-In the original data, there is no explicit explanation of how the dropouts cohort is obtained from the database.
-Hence, we use the high school dropouts criteria mentioned in @nlsy79edu to determine who is the dropouts, which are:
+There is no explicit explanation of how the dropouts cohort is determined in the original data. Hence, we use the high school dropouts criteria from @nlsy79edu, which are:
 
-1. The individual whose highest grade completed (`hgc`) is reported to be less than 12th grade, **or**
-2. The individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and have received a GED (`ged` is code to 2).
+1. An individual whose highest grade completed (`hgc`) is reported to be less than 12th grade, **or**
+2. An individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and have received a GED (`ged` is code to 2).
 
-Further, @SingerJudithD2003Alda set additional criteria for the dropouts in the textbook data, which are male and aged between 14 and 17 years old in 1979.
-Hence, we also use this information to re-create the dropouts data.
+An additional criteria from @SingerJudithD2003Alda is to only include males aged between 14 and 17 years old in 1979. Filters using these criteria used re-create the dropouts data.
 
 With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
-We then matched the individuals in the original and refreshed dataset to investigate why some individuals in the original data were not captured in the refreshed data.
-We found several reasons for this as follows:
+To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
+This revealed several reasons for for the disparity:
 
@@
 173 individuals were more than 17 years old in 1979.
-Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in their subset.
-Our decision is to also include them in the refreshed data as the data has age variable in it, so analyst could filter them later.
+Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in the subset.
+Our decision is to also include them in the refreshed data as the new data contains an age variable, so analyst could filter them later.
 2. 79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:
 
-  + 35 of them completed at least 12th grade and having a diploma instead of GED (`ged` variable is coded to 1).
-Thus, they are indeed are not dropouts and we decide to not include them in the refreshed data.
-  + The information of `ged` is missing in 38 individuals. We decide to include them in the refreshed data. 
-  + 3 individuals are having both diploma and GED (`ged` is coded to 3). They essentially have GED, so we decide to keep them in the refreshed data as well. 
+  + 35 of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded to 1). This suggests that they are not dropouts, and so we excluded them from the refreshed data. 
+  + The information about `ged` is missing in 38 individuals. We decided to include them in the refreshed data. 
+  + 3 individuals have both diploma and GED (`ged` is coded to 3). These were kept in the refreshed data. 
   + 12 individuals do not exist in `wages` data because they were participated in less than 3 rounds of survey.
 
-We then redo the filtering based on those decisions we made on those individuals who were not captured in the refreshed data.
-Finally, we have 863 dropouts in the refreshed data, the closest number of individuals we get to the original data.
+The filtering was re-applied using these decisions. The refreshed dropout data has 863 individuals.
 

### 2022-02-15 paper/paper.tex
revised Filtering: Determining who is a dropout

@@
 
-In the original data, there is no explicit explanation of how the dropouts cohort is obtained from the database.
-Hence, we use the high school dropouts criteria mentioned in Wolpin (2005) to determine who is the dropouts, which are:
+There is no explicit explanation of how the dropouts cohort is determined in the original data. Hence, we use the high school dropouts criteria from Wolpin (2005), which are:
 
@@
 \item
-  The individual whose highest grade completed (\texttt{hgc}) is reported to be less than 12th grade, \textbf{or}
-\item
-  The individual whose highest grade completed (\texttt{hgc}) is reported to be at least 12th grade and have received a GED (\texttt{ged} is code to 2).
+  An individual whose highest grade completed (\texttt{hgc}) is reported to be less than 12th grade, \textbf{or}
+\item
+  An individual whose highest grade completed (\texttt{hgc}) is reported to be at least 12th grade and have received a GED (\texttt{ged} is code to 2).
 \end{enumerate}
 
-Further, Singer and Willett (2003) set additional criteria for the dropouts in the textbook data, which are male and aged between 14 and 17 years old in 1979.
-Hence, we also use this information to re-create the dropouts data.
+An additional criteria from Singer and Willett (2003) is to only include males aged between 14 and 17 years old in 1979. Filters using these criteria used re-create the dropouts data.
 
 With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
-We then matched the individuals in the original and refreshed dataset to investigate why some individuals in the original data were not captured in the refreshed data.
-We found several reasons for this as follows:
+To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
+This revealed several reasons for for the disparity:
 
@@
   173 individuals were more than 17 years old in 1979.
-Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in their subset.
-Our decision is to also include them in the refreshed data as the data has age variable in it, so analyst could filter them later.
+Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in the subset.
+Our decision is to also include them in the refreshed data as the new data contains an age variable, so analyst could filter them later.
 \item
@@
 \item
-  35 of them completed at least 12th grade and having a diploma instead of GED (\texttt{ged} variable is coded to 1).
-Thus, they are indeed are not dropouts and we decide to not include them in the refreshed data.
-\item
-  The information of \texttt{ged} is missing in 38 individuals. We decide to include them in the refreshed data.
-\item
-  3 individuals are having both diploma and GED (\texttt{ged} is coded to 3). They essentially have GED, so we decide to keep them in the refreshed data as well.
+  35 of them completed at least 12th grade with a diploma instead of GED (\texttt{ged} variable is coded to 1).
+This suggests that they are not dropouts, and so we excluded them from the refreshed data.
+\item
+  The information about \texttt{ged} is missing in 38 individuals. We decided to include them in the refreshed data.
+\item
+  3 individuals have both diploma and GED (\texttt{ged} is coded to 3). These were kept in the refreshed data.
 \item
@@
 
-We then redo the filtering based on those decisions we made on those individuals who were not captured in the refreshed data.
-Finally, we have 863 dropouts in the refreshed data, the closest number of individuals we get to the original data.
+The filtering was re-applied using these decisions. The refreshed dropout data has 863 individuals.
 

### 2022-02-16 paper/paper.Rmd
conclusions

@@
 Why is this exercise important for teachers and students?
-The data package and software could be used for illustrating data cleaning in the classroom.
-The refreshed and contemporary data provided in the package can be used as examples in the teaching of initial data analysis and modeling, and exploratory data analysis of longitudinal data.
+This work illustrates the steps in cleaning and processing data in the preparation for it to be a textbook data set.
+Choices made during the cleaning and processing can affect findings made with the data, and these should be transparent.
+Along with the textbook data, which is now provided as an R package, the documentation of the process provide some examples for teaching data cleaning, initial data analysis and exploratory data analysis.
+The refreshed textbook data provides a resource for teaching longitudinal data analysis.
 
 The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual.
-From modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns.
+Public policy is based on models, yielding averages that might vary across strata.
+Modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns.
 That more education leads to increasingly higher wages, which is a satisfying result for educators.
@@
 This is a disturbing finding because there is no rational for such a difference in a fair society.
-This is would provide support for action in public policy to remove this overall average effect.
+This would provide support for action in public policy to remove this overall average effect.
 It also provides an example for educators to explain the use of data to support public policy action.
 
-These are interpretations of results from analyzing the wages data, which rely on trusting that the data provided is accurate and valid.
-The data is collected from a reputable organization, but the data has obvious errors that need correcting.
+On an individual level, one needs to know where am I in this data, and does this data relate to me.
+To do this, the individual profiles need to be explored.
+Pre-dominantly, we would learn that the variation from one individual to another is far more than the variation between demographic strata.
+For example, many individuals with lower educational attainment earn very high wages.
+From a statistics and data science educator perspective, more focus and more methodology for this type of statistics needs to be included in the curriculum.
+
+The above interpretations of results from analyzing the wages data, rely on trusting that the data provided is accurate and valid.
+The wages data is collected by a reputable organization, but we found that the data has obvious errors that should be corrected.
 This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.
@@
 Having trustworthy data is imperative for statistics and data science education.
-Students do interpret results and take seriously what is reported from example analyses.
-If the data in the examples are synthetic, with our own inherent prejudices (on fog e.g.
-gender and race) applied, or inaccurately processed leaving errors, we as educators are being irresponsible because the societal message taught to students is flawed.
+Whenever one uses a textbook data set that is perceived as relating to the students lives, there will be interpretations made.
+The wages data is an example of this.
+Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race.
+If one uses data examples that are synthetic, instilled with our own inherent prejudices (for e.g.
+gender and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
 This paper demonstrates the process to produce a trustworthy data set for teaching.

### 2022-02-16 paper/paper.tex
conclusions

@@
 Having trustworthy data is imperative for statistics and data science education.
-Students do interpret results and take seriously what is reported from example analyses.
-If the data in the examples are synthetic, with our own inherent prejudices (on fog e.g.~gender and race) applied, or inaccurately processed leaving errors, we as educators are being irresponsible because the societal message taught to students is flawed.
+Whenever one uses a textbook data set that is perceived as relating to the students lives, there will be interpretations made.
+The wages data is an example of this.
+Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race.
+If one uses data examples that are synthetic, instilled with our own inherent prejudices (for e.g.~gender and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
 This paper demonstrates the process to produce a trustworthy data set for teaching.

### 2022-02-16 response/response1.Rmd
response

@@
 
-XXX We need to review this literature. It's not true that we have ignored it, the main reference is Dasu! But we need to decide whether there is more that we shoudl cite.
-
-Outlier detection ???
+We agree that there is substantial literature on data cleaning.
+We have referenced one of the main works, Dasu and Johnson (2003) but don't feel it is necessary to cite more than this.
+We have similarly only pointed to one main reference for IDA and EDA.
 
@@
 
-* We have fixed the sentences and the reference. XXX Which one?
+* We have fixed the sentences and the reference.
 

### 2022-02-16 paper/paper-jsdse/paper-jsdse.tex
Add longitudinal version of .

--- 2022-02-03 paper/paper-jsdse/paper-jsdse.tex
@@
 \begin{abstract}
-The National Longitudinal Survey of Youth (NLSY79) is a prominent open data source that has been important for educational purposes and multidisciplinary research on longitudinal data.
-Subsets of this data can be found in numerous textbooks and research articles.
-However, the steps and decisions taken to get from the raw data to the textbook data are never clearly articulated.
-This article describes our journey when trying to re-create a textbook example data set from the original database, with the goal being to refresh the textbook data more regularly.
-Thus, this paper demonstrates the process -- extracting, tidying, cleaning, and exploring with documentation -- to make refreshed data available for education or research.
-Three new data sets and the code to produce them are provided in an accompanying open source R package, called {[}CENSORED{]}.
-As a result of this process, some recommendations are also made for the NLSY79 curators for incorporating data quality checks and providing more convenient samples of the data to potential users.
+Textbook data is essential for teaching statistics and data science methods because they are clean, allowing the instructor to focus on methodology.
+Ideally textbook data sets are refreshed regularly, especially when they are subsets taken from an on-going data collection.
+It is also important to use contemporary data for teaching, to imbue the sense that the methodology is relevant today.
+This paper describes the trials and tribulations of refreshing a textbook data set on wages, extracted from the National Longitudinal Survey of Youth (NLSY79) in the early 1990s.
+The data is useful for teaching modeling and exploratory analysis of longitudinal data.
+Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles.
+The NLSY79 database has been continuously updated through to 2018, so new records are available.
+Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future.
+Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset, have not been clearly articulated.
+We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching.
+Three new data sets and the code to produce them are provided in the open source R package, called {[}CENSORED{]}.
 \end{abstract}
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, including hierarchical, mixed effects and multilevel.
-The freely available data records hourly wages of a sample of high school dropouts, from 1979-1994, along with demographic variables including education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
-Department of Labor 2021).
+An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel.
+The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
+Department of Labor 2021a).
 
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-The wages data was used to illustrate exploratory longitudinal data analysis in Ozlem Ilk's PhD thesis supervised by author Cook (Ilk 2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
+The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at CENSORED.
 
-This textbook data was used as an example in Ilk (2004), and was developed into a case study for use in teaching exploratory data analysis at CENSORED.
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
@@
 
-This disparity between the average and the individual is a part of statistics, as a discipline, that is given less attention.
-It is this disparity which makes this particular data set important as a textbook example.
+This disparity between the average and the individual is a part of statistics, as a discipline, that require more attention.
+This particular data set is a prime example of discussing this disparity.
 Textbook data sets have longevity if there is a unresolved mystery.
@@
 It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).) We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
+(A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).
+We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
 
@@
 If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
-{[}ET: a related article: \url{https://www.smh.com.au/national/by-the-numbers--the-average-australian-doesnt-exist--not-a-single-one-of-us-is-normal-20150124-12xlh0.html}{]}
+A similar tone is echoed occasionally in the public media, for example, an article published in the Sydney Morning Herald argues that there is no average Australian (Moncrief 2015).
 
-As a textbook data set though, it is outdated.
+As a textbook data set though, the wages data is outdated.
 The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
-Teachers of statistics need use contemporary data sets to show how techniques are relevant to today's students.
+Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students.
 Using tired old textbook data sets can imbue a misconception that the field is not current.
@@
 
-This paper demonstrates the steps of cleaning data, and documents the process, as recommended by Huebner, Vach, and Cessie (2016).
+This paper demonstrates the steps of cleaning data, including subjective decisions made on dealing with anomalies, and documents the process, as recommended by Huebner, Vach, and Cessie (2016).
 They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
-Data cleaning can be considered to be a part of what is called ``initial data analysis'' (IDA) (Chatfield 1985).
-In IDA one would also explore the data, especially to check if the data is consistent with assumptions required for modeling.
-This is also related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
+Clean data often then goes through an ``initial data analysis'' (IDA) (Chatfield 1985), where one would summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling.
+This stage is related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
 EDA can be considered to encompass IDA.
-Dasu and Johnson (2003) say that data cleaning and exploration, without naming it as IDA, is a difficult task and consumes 80\% of the data mining task.
+In practice the three stages of cleaning, summarizing and exploring are cyclical, that one often needs to do more cleaning after scrutinizing.
+Dasu and Johnson (2003) say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.
 
@@
 The NLSY79 is a longitudinal survey administered by the U.S Bureau of Labor Statistics that follows the lives of a sample of American youth born between 1957-1964 (Bureau of Labor Statistics, U.S.
-Department of Labor 2021).
+Department of Labor 2021a).
 The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979.
-It comprised of people sampled because they were Blacks, Hispanics, economically disadvantaged non-Black non-Hispanics, and youth in the military.
-In 1984 and 1990, two sub-samples were dropped from the interview; the dropped subjects were the 1,079 members of the military sample and 1,643 members who represents economically disadvantage people, but were not Black nor Hispanics, respectively.
-These subsamples were dropped because they are not eligible anymore for interview (Bureau of Labor Statistics, U.S.
-Department of Labor 2021).
-Hence, 9,964 respondents remain in the eligible samples.
+For a variety of reasons, some structural, the number of respondents dropped to 9,964 after 1990.
 The surveys were conducted annually from 1979 to 1994 and biennially thereafter.
@@
 
-The NLSY79 data used in Singer and Willett (2003) contains the longitudinal measurements on yearly mean hourly wages with years of workforce experience, and demographic variables education and race, from 1979 through to 1994.
-In addition, the cohort is restricted to male high-school dropouts who first participated in the study at age 14-17 years.
-Thus the target data set is to collect the same variables for the extended time frame of 1994 through to 2018, the most recent year reported.
+The NLSY79 data used in Singer and Willett (2003) contains the longitudinal records of male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994.
+This dataset contains several variables as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  ID: the respondents' ID.
+\item
+  EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
+\item
+  LNW: natural logarithm of wages, adjusted with 1990's inflation rate.
+\item
+  BLACK: binary variable, 1 indicates black and 0 otherwise.
+\item
+  HISPANIC: binary variable, 1 indicates hispanic and 0 otherwise.
+\item
+  HGC: the highest grade completed.
+\item
+  UERATE: unemployment rate of the year of survey. When missing, the variable is set to be 7.875 (the average rate).
+\end{enumerate}
+
+We refresh this data by re-creating the full data with records from survey years 1979 through to 2018 (the most recent year published).
+We also modify some variables.
+For example, use a single categorical race variable instead of the two binary race variables.
+We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), the number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
+We do not attempt to re-create the unemployment rate variable.
+
+The plan is to create three datasets as follows:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  The wages data of the whole NLSY79 cohort, including females.
+\item
+  A separate table of the demographic data of the whole NLSY79 cohort.
+\item
+  The wages data of the high school dropouts, which is closest to a refreshed version of Singer and Willett (2003)'s data.
+\end{enumerate}
 
@@
 
-van der Loo and de Jonge (2018) describe the notion of a ``statistical value chain'' where the production stages of the data cleaning process are earmarked as raw data (data as arrived to the desk of an analyst), input data (data organised with correct type and identified variables) and valid data (data that faithfully represent the variables).
+van der Loo and de Jonge (2018) in the context of official statistics describe the ``statistical value chain'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organised with correct type and identified variables) and valid data (data that has been cleaned and more accurately represents the intent of variables).
+What we have colorfully named as wild data can be considered to be raw data, and valid data could be considered to be textbook data, in the above statistical value chain.
 In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida}.
@@
 
-The raw data, \texttt{new\_data\_qnames}, is organised such that each row corresponds to an individual.
+According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
+The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organised such that each row corresponds to an individual.
 As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year.
-The raw data consequently has a large number of columns (686 to be specific).
+The raw data consequently has a large number of columns (663 to be specific).
 The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars.
-A glimpse of this data output is shown below.
+A glimpse of this data shows:
 
 \begin{verbatim}
-#> 'data.frame':    12686 obs. of  716 variables:
-#>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 9 10 ...
-#>  $ HRP1_1979                                 : int  328 385 365 NA 310 NA NA NA 214 NA ...
-#>  $ HRP2_1979                                 : int  NA NA NA NA 375 NA NA NA NA NA ...
-#>  $ HRP3_1979                                 : int  NA NA 275 NA NA NA NA NA NA NA ...
-#>  $ HRP4_1979                                 : int  NA NA NA NA NA 250 NA NA NA NA ...
-#>  $ HRP5_1979                                 : int  NA NA NA NA NA NA NA NA NA NA ...
-#>  $ HRP1_1980                                 : int  NA 457 397 NA 333 275 300 394 200 318 ...
-#>  $ HRP2_1980                                 : int  NA NA 367 NA NA NA NA NA NA NA ...
-#>  $ HRP3_1980                                 : int  NA NA 380 NA NA NA 290 NA NA NA ...
-#>  $ HRP4_1980                                 : int  NA NA NA NA NA NA NA NA NA NA ...
+#> 'data.frame':    12686 obs. of  663 variables:
+#>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 ...
+#>  $ HRP1_1979                                 : int  328 385 365 NA 310 NA NA NA ...
+#>  $ HRP2_1979                                 : int  NA NA NA NA 375 NA NA NA ...
+#>  $ HRP3_1979                                 : int  NA NA 275 NA NA NA NA NA ...
+#>  $ HRP4_1979                                 : int  NA NA NA NA NA 250 NA NA ...
+#>  $ HRP5_1979                                 : int  NA NA NA NA NA NA NA NA ...
+#>  $ HRP1_1980                                 : int  NA 457 397 NA 333 275 300 394 ...
+#>  $ HRP2_1980                                 : int  NA NA 367 NA NA NA NA NA ...
+#>  $ HRP3_1980                                 : int  NA NA 380 NA NA NA 290 NA ...
+#>  $ HRP4_1980                                 : int  NA NA NA NA NA NA NA NA ...
 #>   [list output truncated]
@@
 
-According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
-The raw data does not comply with these rules, thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
+Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
 This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
-2019), \texttt{tidyr} (Wickham 2020) to pivot the data into long form, \texttt{dplyr} (Wickham, François, et al.
-2020), and \texttt{stringr} (Wickham 2019) for creating new variables, and levels of factors by text wrangling.
+2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham, François, et al.
+2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}) and the corresponding year this grade was completed (\texttt{yr\_hgc}).
+In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), and whether the graduate equivalency diploma is obtained (\texttt{ged}).
 
@@
 
-The next step is tidying to obtain \texttt{hgc} and \texttt{yr\_hgc}.
-The highest grade completed are recorded in \texttt{new\_data\_qnames} as variables beginning with \texttt{Q3-4} and \texttt{HGC} with suffix of the year it was recorded.
-In addition the variables beginning with \texttt{HGCREV} contain the revised data.
-We choose to use this revised May data because it seemed to have less missing and presumably has been checked.
-However, there is no revised May data for 2012, 2014, 2016, and 2018, thus, we use the ordinary May data for these years.
-
-The \texttt{hgc} is measured and could be updated in each period of the survey.
-We chose to only retain the highest grade completed for each individual and derived the year when they completed it (\texttt{yr\_hgc}) by finding the minimum year with the highest completed grade.
+The next step is tidying to obtain \texttt{hgc} and \texttt{ged}.
+As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
+For \texttt{ged}, we only separate the year of survey and the GED status of the individual.
+Later on, this variable is useful to determine the refreshed high school dropouts data.
 
@@
 
-Our target variables for the employment are to obtain respondent's mean hourly wage (\texttt{mean\_hourly\_wage}), the number of jobs (\texttt{number\_of\_jobs}) and the total hours of work per week (\texttt{total\_hours}) for each survey year.
+Our target variables for the employment are to obtain respondent's mean hourly wage (\texttt{wage}), the number of jobs (\texttt{njobs}), the total hours of work per week (\texttt{hours}) for each survey year, the year when individual starting to work (\texttt{stwork}), the length of time (years) in workforce (\texttt{yr\_wforce}), and work experience measured as the number of years worked (\texttt{exp}).
 As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.
@@
 
+For \texttt{stwork} variable, we only rename the column names from \texttt{new\_data\_qnames}.
+This variable is then used to calculate the next variable, \texttt{yr\_wforce} for each survey round, which is the year of survey (\texttt{year}) minus the year of individual start working (\texttt{stwork}).
+Finally, \texttt{exp} variable is derived from the number of weeks worked since the last interview indicated as variable started with \texttt{WKSWK} in \texttt{new\_data\_qnames}.
+To obtain the work experience since 1979, we calculate the cumulative value.
+As the measurement unit is in week, we convert this to year.
+
 The employment and demographic variables are then joined.
-These data are further filtered to the cohort who completed education up to 12th grade and participated in at least five rounds in the survey.
-We save the resultant wage data on this cohort as \texttt{wages}.
+These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in Singer and Willett (2003).
+We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data.
+However, it is worth noting that the original data does not restrict the number of observation for each individual, i.e, there are individuals with only one and two observations.
+
+Finally, we save the resultant wage data on this cohort as \texttt{wages}.
+
+\hypertarget{calculated-variables-work-experience}{%
+\subsection{Calculated variables: work experience}\label{calculated-variables-work-experience}}
+
+Work experience is one of the most important variables in Singer and Willett (2003) as it indicates time, and makes longitudinal analysis possible.
+It is desirable to calculate this rather than using the survey year for time because it more accurately reflects a person's time in the workforce.
+Thus, in the spirit of refreshing the data to the newest round of the survey, this variable needs to be calculated from other variables provided.
+It is not straightforward.
+We start with the definition of experience in Singer and Willett (2003).
+
+Experience is years after entering the labor force.
+It represents the difference between the day an individual enters the labor force (\texttt{EMPLOYERS\_ALL\_STARTDATE\_ORIGINAL.01\textasciitilde{}Y\_XRND} in the database) relative to the date of the survey, which we call \texttt{yr\_wforce}.
+However, using this calculation produced numbers which don't quite match with the original data.
+
+Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide, suggests that it should be calculated based on the variable ``number of weeks worked since last interview''.
+This would remove periods of unemployment which makes sense when measuring experience while actually working.
+Since it is only measured since the last interview, this needs to be cumulated for each survey year.
+This produced results more similar to the original data (see Subsection \ref{takeaways}).
 
@@
 
+\begin{table}
+
+\caption{\label{tab:age-table}Frequency table of the age at the start of the survey in NSLY79 cohort}
+\centering
+\begin{tabular}[t]{rr}
+\toprule
+Age & Number of individuals\\
+\midrule
+15 & 1,265\\
+16 & 1,550\\
+17 & 1,600\\
+18 & 1,530\\
+19 & 1,662\\
+20 & 1,722\\
+21 & 1,677\\
+22 & 1,680\\
+\bottomrule
+\end{tabular}
+\end{table}
+
+\begin{table}
+
+\caption{\label{tab:gender-race-table}Contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
+\centering
+\begin{tabular}[t]{lrrrr}
+\toprule
+\multicolumn{1}{c}{ } & \multicolumn{3}{c}{Race} & \multicolumn{1}{c}{ } \\
+\cmidrule(l{3pt}r{3pt}){2-4}
+Gender & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
+\midrule
+Male & 1,000 (16\%) & 1,613 (25\%) & 3,790 (59\%) & 6,403\\
+Female & 1,002 (16\%) & 1,561 (25\%) & 3,720 (59\%) & 6,283\\
+\midrule
+Total & 2,002 (16\%) & 3,174 (25\%) & 7,510 (59\%) & 12,686\\
+\bottomrule
+\end{tabular}
+\end{table}
+
 \begin{figure}
@@
 \hypertarget{censor}{%
-\subsubsection{Replacing extreme values}\label{censor}}
+\subsection{Replacing extreme values}\label{censor}}
 
-As part of the IDA, which is the model formulation, we build a robust linear regression model to treat the extreme values in the data.
-Robust linear regression yields an estimation robust to the influence of noise or contamination (Koller 2016).
-It also aims to detect the contamination by weighting each observation based on how ``well-behaved'' they are, known as robustness weight.
-An observation with a lower robustness weight would be suggested as an outlier (Koller 2016).
-
-Since we work with longitudinal data, we build the model for each ID instead of the overall data.
-The robust mixed model could be the best model to be employed in this case.
-However, this method is too computationally and memory expensive, especially for a large data set, like the NLSY79 data.
-Therefore, the model for each ID is built utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively.
+To treat the extreme values in the data, we build a robust linear regression model where we use the robustness weight to determine if a value should be replaced with the fitted value from the model.
+We build the model for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively.
+This fits a model separately for each individual, and a robust linear mixed model could be used as an alternative approach.
+We tried fitting a robust linear mixed model using \texttt{robustlmm} (Koller 2016), however perhaps because the data is large and imbalanced, the model was not converging.
 The full code for this is shown at \url{CENSORED/articles/input-to-valid-data.html} but also described in detail next.
@@
 
+\begin{verbatim}
+#> # A tibble: 2 x 3
+#>   hgc   hgc12     n
+#>   <chr> <chr> <int>
+#> 1 12TH  8TH       1
+#> 2 8TH   12TH    243
+\end{verbatim}
+
+\begin{figure}
+
+{\centering \includegraphics[width=0.7\linewidth]{figures/compare-xp-sw-1}
+
+}
+
+\caption{Males dropouts' work experience in the original and refreshed data.
+Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
+We can see that the correlation is reasonably high between them even there are different values for many individuals.}\label{fig:compare-xp-sw}
+\end{figure}
+
 The original set contains wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994.
@@
 The original set is available in the R package \texttt{brolgar} and the refreshed data is available in the R package \texttt{[CENSORED]}.
+
+\hypertarget{filtering-determining-who-is-a-dropout}{%
+\subsection{Filtering: Determining who is a dropout}\label{filtering-determining-who-is-a-dropout}}
+
+There is no explicit explanation of how the dropouts cohort is determined in the original data. Hence, we use the high school dropouts criteria from Wolpin (2005), which are:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  An individual whose highest grade completed (\texttt{hgc}) is reported to be less than 12th grade, \textbf{or}
+\item
+  An individual whose highest grade completed (\texttt{hgc}) is reported to be at least 12th grade and have received a GED (\texttt{ged} is code to 2).
+\end{enumerate}
+
+An additional criteria from Singer and Willett (2003) is to only include males aged between 14 and 17 years old in 1979. Filters using these criteria used re-create the dropouts data.
+
+With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
+To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
+This revealed several reasons for for the disparity:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  173 individuals were more than 17 years old in 1979.
+Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in the subset.
+Our decision is to also include them in the refreshed data as the new data contains an age variable, so analyst could filter them later.
+\item
+  79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:
+\end{enumerate}
+
+\begin{itemize}
+\tightlist
+\item
+  35 of them completed at least 12th grade with a diploma instead of GED (\texttt{ged} variable is coded to 1).
+This suggests that they are not dropouts, and so we excluded them from the refreshed data.
+\item
+  The information about \texttt{ged} is missing in 38 individuals. We decided to include them in the refreshed data.
+\item
+  3 individuals have both diploma and GED (\texttt{ged} is coded to 3). These were kept in the refreshed data.
+\item
+  12 individuals do not exist in \texttt{wages} data because they were participated in less than 3 rounds of survey.
+\end{itemize}
+
+The filtering was re-applied using these decisions. The refreshed dropout data has 863 individuals.
 
@@
 
-There are two aspects of the original data that make a direct comparison difficult.
-Th first one is that The time variable provided is ``experience in the workforce''.
-It is not clear how this is calculated.
-One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection.
-Secondly, wages were inflation-adjusted to 1990 prices, which is not done for the refreshed data.
+\hypertarget{takeaways}{%
+\subsection{The takeaways}\label{takeaways}}
 
-\textbackslash begin\{figure\}
+There are two aspects of the original data that make the replication with exactly the same result is not obtained.
+The first one is that the calculation of the time variable provided, which is the ``work experience'', is not clearly articulated.
+Singer and Willett (2003) mention that the temporal variable is the years of experience since the first day of work.
+However, this exact same variable is not explicitly available in the database.
+From the NLSY79 topical guide (Bureau of Labor Statistics, U.S.
+Department of Labor 2021b), we find that several variables are tagged as work experience-related variables.
+One of them is the weeks of worked since the last interview.
+Hence, we use this to calculate the variable.
+Even though we got different result, the correlation between the original and refreshed work experience is reasonably high as shown in Figure \ref{fig:compare-xp-sw}.
 
-\{\centering \includegraphics[width=468px]{figures/compare_xp_to_yrworkforce-1}
+Secondly, in the original dataset, wages were inflation-adjusted to 1990 prices.
+This is not done for the refreshed data because we plan to keep refreshing it as the data from the newest survey period release.
+Thus, the 1990 prices might not be relevant anymore for the future context.
+Hence, we keep it as the original value to allow the users to treat it as their needs.
 
-\}
-
-\caption{The association between the lenght of time in the workforce (years) and the work experience.
-The color represents the respondents (id).
-Generally, the longer someone were in the workforce, the higher the work experience.
-However, some of the respondents have much lower working experience compared their duration in the workforce.}
-
-(\#fig:compare\_xp\_to\_yrworkforce)
-\textbackslash end\{figure\}
-
-\begin{center}\includegraphics[width=468px]{figures/compare_xp_sw-1} \end{center}
-
-The treatment of unlikely wages differ in the refreshed data.
+It is also worth noting that the treatment of unlikely wages differ in the refreshed data.
 In the original data by Singer and Willett (2003), wages greater than \$75 are set to be missing.
 However, in recent context, this value is too low to be set as the maximum threshold.
-We opted to use the weights from a robust linear regression to determine what should be set as missing value as described in Section \ref{censor}.
+We opted to use the weights from a robust linear regression to determine what should be regarded as outliers and imputed them with their predicted values as as described in Section \ref{censor}.
 
@@
 
-This paper has described the stages to take a particular open data set and make it a textbook data set, ready for the classroom or research.
+This paper has illustrated the steps and decisions made to take a particular open data set and make it a textbook data set, ready for the classroom or research.
 In the first stage, we showed the steps performed to get the data from the NLSY79 database.
-The data format needed conversion to tidy format, and this was described.
-After that, an initial data analysis was conducted to investigate and screen the quality of the data.
-We found and fixed the anomalous observations in wages using a robust linear regression model.
-Finally a comparison is made between the original and refreshed data sets.
+The data format was converted to tidy format, for more flexibility in cleaning and exploring.
+An initial data analysis was conducted to investigate and screen the quality of the data.
+We found and provided a fix for many anomalous observations in wages using a robust linear regression model.
+The refreshed data is compared with the original set using a variety of numerical summaries and graphics.
+The current subset is made available in a new R package, called \texttt{[CENSORED]}.
 
-The data cleaning process is documented and the code has been made available. These provide the opportunity to again refresh the textbook data as updated data is published in the NLSY79 database.
-Determining the appropriate robustness weight to threshold the anomalous observations is documented, and a \texttt{shiny} (Chang et al. 2020) app is provided to assist.
-The current subset is made available in a new R package, called \texttt{[CENSORED]}.
+The data cleaning process is documented and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
+Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a \texttt{shiny} (Chang et al.
+2020) app, and the choices used in the refreshed data are documented.
 
@@
 \item
-  there are many errors in the data, e.g.~hourly wages greater than \$30,000 per hour.
+  calculating experience in the workforce requires comparing the date of first job with the first year the individual was recorded, both of which are available in the database.
 \item
-  there is no explicit variable in the database recording high school dropout, which means we needed to compare date of 12th grade with the individual's age.
+  there are many unusually high hourly wages, e.g.~greater than \$30,000 per hour.
 \item
-  calculating experience in the workforce would require knowing the time of first job within the first year the individual was recorded.
+  there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with the individual's age.
+\item
+  matching id's from the original data with those in the refreshed data, do refer to the same person based on the demographic information available.
 \end{itemize}
@@
 We do not assert that we have produced the best textbook data, but rather we describe our journey to provide a reasonable textbook data set.
-All code and documentation are provided for transparency, and future updates of the \texttt{[CENSORED]} package may contain additional variables, or filters of the full set, if it is deemed important.
+All code and documentation are provided for transparency.
+Readers could use this to make different decisions, or provide suggestions on through package for better choices.
+Future updates of the \texttt{[CENSORED]} package may contain additional variables, or filters of the full set, if it is deemed important.
 
-Finally, for the data providers we recommend that a validation system with clear rules is added on data entry, and that alternative output formats, such as a tidy format would help users make better use of the resource.
+For the data providers, we recommend that a better validation system with clear rules applied at data entry, and that alternative output formats, such as a tidy format would help users make better use of their resource.
 The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs.
@@
 Users need to be able to trust the data.
+
+Why is this exercise important for teachers and students?
+This work illustrates the steps in cleaning and processing data in the preparation for it to be a textbook data set.
+Choices made during the cleaning and processing can affect findings made with the data, and these should be transparent.
+Along with the textbook data, which is now provided as an R package, the documentation of the process provide some examples for teaching data cleaning, initial data analysis and exploratory data analysis.
+The refreshed textbook data provides a resource for teaching longitudinal data analysis.
+
+The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual.
+Public policy is based on models, yielding averages that might vary across strata.
+Modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns.
+That more education leads to increasingly higher wages, which is a satisfying result for educators.
+It means that education makes a difference in the wage experience, and that public policy that encourages education is a data-supported action.
+We would also learn, although not presented here, that race matters and that being black leads to lower wages.
+This is a disturbing finding because there is no rational for such a difference in a fair society.
+This would provide support for action in public policy to remove this overall average effect.
+It also provides an example for educators to explain the use of data to support public policy action.
+
+On an individual level, one needs to know where am I in this data, and does this data relate to me.
+To do this, the individual profiles need to be explored.
+Pre-dominantly, we would learn that the variation from one individual to another is far more than the variation between demographic strata.
+For example, many individuals with lower educational attainment earn very high wages.
+From a statistics and data science educator perspective, more focus and more methodology for this type of statistics needs to be included in the curriculum.
+
+The above interpretations of results from analyzing the wages data, rely on trusting that the data provided is accurate and valid.
+The wages data is collected by a reputable organization, but we found that the data has obvious errors that should be corrected.
+This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.
+
+Having trustworthy data is imperative for statistics and data science education.
+Whenever one uses a textbook data set that is perceived as relating to the students lives, there will be interpretations made.
+The wages data is an example of this.
+Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race.
+If one uses data examples that are synthetic, instilled with our own inherent prejudices (for e.g.~gender and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
+This paper demonstrates the process to produce a trustworthy data set for teaching.
 
@@
 Department of Labor.
-2021.
+2021a.
 ``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
-Columbus, OH.
+Columbus, OH, through \url{https://www.nlsinfo.org/bibliography-citing-nls-data}.
+
+\leavevmode\hypertarget{ref-nlsy79guide}{}%
+---------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.'' \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
 
@@
 Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.
+
+\leavevmode\hypertarget{ref-notaverage}{}%
+Moncrief, Marc. 2015. ``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.'' \url{https://bit.ly/smh-not-normal}.
 

### 2022-02-16 paper/paper.Rmd
Add longitudinal version of .

@@
 It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(A side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new and does not suffer from a connection with eugenics [@notiris].) We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
+(A side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new and does not suffer from a connection with eugenics [@notiris].
+We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
 
@@
 The next step is tidying to obtain `hgc` and `ged`.
-As the highest grade completed is already exist in  `new_data_qnames` and stored in year unit, we only rename the column name to be `hgc_i` and recode it so we have the factor data type.
-For `ged`, we only separate the years and the GED status of the individual.
+As the highest grade completed is already exist in  `new_data_qnames` and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be `hgc_i` and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
+For `ged`, we only separate the year of survey and the GED status of the individual.
 Later on, this variable is useful to determine the refreshed high school dropouts data.
@@
 The employment and demographic variables are then joined.
-These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation articulated in @SingerJudithD2003Alda.
+These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in @SingerJudithD2003Alda.
 We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data.
-We save the resultant wage data on this cohort as `wages`.
+However, it is worth noting that the original data does not restrict the number of observation for each individual, i.e, there are individuals with only one and two observations.
+
+Finally, we save the resultant wage data on this cohort as `wages`.
 

### 2022-02-16 paper/paper.tex
Add longitudinal version of .

@@
 Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
-This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
-2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham et al.
+This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
+2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham, François, et al.
 2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
@@
 The next step is tidying to obtain \texttt{hgc} and \texttt{ged}.
-As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in year unit, we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type.
-For \texttt{ged}, we only separate the years and the GED status of the individual.
+As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
+For \texttt{ged}, we only separate the year of survey and the GED status of the individual.
 Later on, this variable is useful to determine the refreshed high school dropouts data.
@@
 The employment and demographic variables are then joined.
-These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation articulated in Singer and Willett (2003).
+These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in Singer and Willett (2003).
 We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data.
-We save the resultant wage data on this cohort as \texttt{wages}.
+However, it is worth noting that the original data does not restrict the number of observation for each individual, i.e, there are individuals with only one and two observations.
+
+Finally, we save the resultant wage data on this cohort as \texttt{wages}.
 
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-iris-data}{}}%
-Andereson, Edgar. 1935. {``The Irises of the Gaspe Peninsula.''} \emph{Bulletin of the American Iris Society} 59: 2--5.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
+\leavevmode\hypertarget{ref-iris-data}{}%
+Andereson, Edgar. 1935. ``The Irises of the Gaspe Peninsula.'' \emph{Bulletin of the American Iris Society} 59: 2--5.
+
+\leavevmode\hypertarget{ref-nlsy79}{}%
 Bureau of Labor Statistics, U.S.
@@
 2021a.
-{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
+``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
 Columbus, OH, through \url{https://www.nlsinfo.org/bibliography-citing-nls-data}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlsy79guide}{}}%
----------. 2021b. {``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.''} \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-shiny}{}}%
-Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{{shiny: Web Application Framework for R}}. \url{https://CRAN.R-project.org/package=shiny}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chatfield1985TIEo}{}}%
-Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
-
-\leavevmode\vadjust pre{\hypertarget{ref-eliznlsy}{}}%
+\leavevmode\hypertarget{ref-nlsy79guide}{}%
+---------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.'' \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
+
+\leavevmode\hypertarget{ref-shiny}{}%
+Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.
+
+\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
+Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
+
+\leavevmode\hypertarget{ref-eliznlsy}{}%
 Cooksey, Elizabeth C.
 2017.
-{``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M.
+``Using the National Longitudinal Surveys of Youth (Nlsy) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
 Lerner Neal Halfon Christoper B.
@@
 Cham: Springer.
-https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-DasuTamraparni2003Edma}{}}%
+\url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.
+
+\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
 Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.
 
-\leavevmode\vadjust pre{\hypertarget{ref-janitor}{}}%
-Firke, Sam. 2020. \emph{{janitor: Simple Tools for Examining and Cleaning Dirty Data}}. \url{https://CRAN.R-project.org/package=janitor}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-racismnotrace}{}}%
-Fullilove, M. T. 1998. {``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.
-
-\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
+\leavevmode\hypertarget{ref-janitor}{}%
+Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.
+
+\leavevmode\hypertarget{ref-racismnotrace}{}%
+Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.
+
+\leavevmode\hypertarget{ref-grimshaw}{}%
 Grimshaw, Scott D.
 2015.
-{``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14.
+``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14.
 \url{https://doi.org/10.1080/00031305.2015.1081106}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-purrr}{}}%
-Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-penguins-data}{}}%
+\leavevmode\hypertarget{ref-purrr}{}%
+Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.
+
+\leavevmode\hypertarget{ref-penguins-data}{}%
 Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
+\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
 Huebner, Marianne, Werner Vach, and Saskia le Cessie.
 2016.
-{``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ilk2004}{}}%
+``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
+
+\leavevmode\hypertarget{ref-ilk2004}{}%
 Ilk, Ozlem.
 2004.
-{``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.''} PhD thesis, Iowa State University.
+``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.'' PhD thesis, Iowa State University.
 \url{https://doi.org/10.31274/rtd-180813-11012}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
+\leavevmode\hypertarget{ref-tamedata}{}%
 Kim, A.
@@
 2018.
-{``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1).
+``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1).
 \url{https://doi.org/10.5070/T511103589}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-KollerManuel2016rARP}{}}%
-Koller, Manuel. 2016. {``{robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models}.''} \emph{Journal of Statistical Software} 75 (6): 1--24.
-
-\leavevmode\vadjust pre{\hypertarget{ref-notaverage}{}}%
-Moncrief, Marc. 2015. {``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.''} \url{https://bit.ly/smh-not-normal}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-opendata}{}}%
-Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-patchwork}{}}%
-Pedersen, Thomas Lin. 2020. \emph{{patchwork: The Composer of Plots}}. \url{https://CRAN.R-project.org/package=patchwork}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-MichaelRPergamit2001DWTN}{}}%
+\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
+Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.
+
+\leavevmode\hypertarget{ref-notaverage}{}%
+Moncrief, Marc. 2015. ``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.'' \url{https://bit.ly/smh-not-normal}.
+
+\leavevmode\hypertarget{ref-opendata}{}%
+Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.
+
+\leavevmode\hypertarget{ref-patchwork}{}%
+Pedersen, Thomas Lin. 2020. \emph{patchwork: The Composer of Plots}. \url{https://CRAN.R-project.org/package=patchwork}.
+
+\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
 Pergamit, Michael R., Charles R.
@@
 2001.
-{``Data Watch: The National Longitudinal Surveys.''} \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
-
-\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
+``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
+
+\leavevmode\hypertarget{ref-R}{}%
 R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-SingerJudithD2003Alda}{}}%
+\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
 Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.
 
-\leavevmode\vadjust pre{\hypertarget{ref-notiris}{}}%
-Stodel, Megan. 2020. {``Stop Using Iris.''} \url{https://www.meganstodel.com/posts/no-to-iris/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-brolgar}{}}%
-Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}}. \url{https://github.com/njtierney/brolgar}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tukey}{}}%
+\leavevmode\hypertarget{ref-notiris}{}%
+Stodel, Megan. 2020. ``Stop Using Iris.'' \url{https://www.meganstodel.com/posts/no-to-iris/}.
+
+\leavevmode\hypertarget{ref-brolgar}{}%
+Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.
+
+\leavevmode\hypertarget{ref-tukey}{}%
 Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.
 
-\leavevmode\vadjust pre{\hypertarget{ref-rlm}{}}%
-UCLA: Statistical Consulting Group. 2021. {``{Robust Regression \textbar{} R Data Analysis Examples}.''} February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%
-van der Loo, Mark P. J., and Edwin de Jonge. 2021. {``Data Validation Infrastructure for {R}.''} \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-LooMarkvander2018Sdcw}{}}%
-van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in r}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mass}{}}%
-Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with s}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tsibble}{}}%
+\leavevmode\hypertarget{ref-rlm}{}%
+UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
+
+\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
+van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.
+
+\leavevmode\hypertarget{ref-validate}{}%
+van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
+
+\leavevmode\hypertarget{ref-mass}{}%
+Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
+
+\leavevmode\hypertarget{ref-tsibble}{}%
 Wang, Earo, Dianne Cook, and Rob J Hyndman.
 2020.
-{``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.''} \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
+``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
 \url{https://doi.org/10.1080/10618600.2019.1695624}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
-Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-WickhamHadley2014TD}{}}%
----------. 2014. {``Tidy Data.''} \emph{Journal of Statistical Software} 59 (10): 1--23.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
----------. 2016. \emph{{ggplot2: Elegant Graphics for Data Analysis}}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-stringr}{}}%
----------. 2019. \emph{{stringr: Simple, Consistent Wrappers for Common String Operations}}. \url{https://CRAN.R-project.org/package=stringr}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tidyr}{}}%
----------. 2020. \emph{{tidyr: Tidy Messy Data}}. \url{https://CRAN.R-project.org/package=tidyr}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
+\leavevmode\hypertarget{ref-plyr}{}%
+Wickham, Hadley. 2011. ``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
+
+\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
+---------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.
+
+\leavevmode\hypertarget{ref-ggplot2}{}%
+---------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
+
+\leavevmode\hypertarget{ref-stringr}{}%
+---------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.
+
+\leavevmode\hypertarget{ref-tidyr}{}%
+---------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.
+
+\leavevmode\hypertarget{ref-tidyverse}{}%
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al.
 2019.
-{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686.
+``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686.
 \url{https://doi.org/10.21105/joss.01686}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
-Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{{dplyr: A Grammar of Data Manipulation}}. \url{https://CRAN.R-project.org/package=dplyr}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-readr}{}}%
-Wickham, Hadley, and Jim Hester. 2020. \emph{{readr: Read Rectangular Text Data}}. \url{https://CRAN.R-project.org/package=readr}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlsy79edu}{}}%
+\leavevmode\hypertarget{ref-dplyr}{}%
+Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.
+
+\leavevmode\hypertarget{ref-readr}{}%
+Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.
+
+\leavevmode\hypertarget{ref-nlsy79edu}{}%
 Wolpin, Kenneth I.
 2005.
-{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Published by Bureau of Labor Statistics, U.S.
+``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Published by Bureau of Labor Statistics, U.S.
 Department of Labor.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-knitr}{}}%
+\leavevmode\hypertarget{ref-knitr}{}%
 Xie, Yihui.
 2014.
-{``Knitr: A Comprehensive Tool for Reproducible Research in {R}.''} In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
+``Knitr: A Comprehensive Tool for Reproducible Research in R.'' In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
 Peng.

### 2022-02-16 response/response1.Rmd
Add longitudinal version of .

--- 2022-02-09 response/response1.Rmd
@@
 \item[Editor and Reviewer 1:] \textcolor{violet}{How is this manuscript relevant for teachers and/or students in statistics and data science?}
+\end{description}
 
-The introduction has been re-written with an emphasis on how this work should be interesting for teachers and students of data science and statistics.
+* The introduction has been re-written with an emphasis on how this work should be interesting for teachers and students of data science and statistics.
 
+\begin{description}
 \item[Reviewer 1:] \textcolor{violet}{I don't see the direct relevance of this manuscript with the readership of JSDSE.
@@
 I do not see this manuscript being engaging enough to be used in the classroom as a case study in data cleaning.}
+\end{description}
 
-It is commonly said 80\% of a data analysts/scientist's work is on data cleaning/preperation so we believe that shedding more light in this aspect of work is an important for students to be aware of.
+* It is commonly said 80\% of a data analysts/scientist's work is on data cleaning/preperation so we believe that shedding more light in this aspect of work is an important for students to be aware of.
 We believe that there are often subjective decisions made in the process (as we have done) which can be used as a conversation for transparency and reproducibility in class based on a real case study.
 
+\begin{description}
 \item[Reviewer 3:] \textcolor{violet}{The authors prominently cite (Chatfield 1985) for the definition of the term “IDA”.
@@
 It seems curious to me that the whole paper is framed as being “IDA” based on the author’s unsupported statement that data cleaning should be considered party of IDA.}
+\end{description}
 
-Yes, there is grey area in these three activities.
+* Yes, there is grey area in these three activities.
 Thanks for pointing out Chatfield's original viewpoint, which differs from Huebner et al's.
@@
 We have kept the explanation of the terminology of IDA and EDA because we think this is important to clarify for teaching purposes.
-
-\item[Reviewer 3:] \textcolor{violet}{My suggestion is that the authors reconfigure this paper to be presented as a pure case study.
-I believe the work that has been done is valuable: the data cleaning process is often messy and unplanned, so a case study like this on a popular dataset would provide an excellent educational resource.}
-
-We have kept the case study frame of mind but have re-cast the paper so that it focuses on the importance of this type of activity for teaching of statistics and data science. 
-\end{description}
 
@@
 
-XXX We need to review this literature. It's not true that we have ignored it, the main reference is Dasu! But we need to decide whether there is more that we shoudl cite.
-
-Outlier detection ???
+* **Need help** XXX We need to review this literature. It's not true that we have ignored it, the main reference is Dasu! But we need to decide whether there is more that we should cite.
 
@@
 
-We thank the reviewer for pointing out to us that the data contains information on the work experience -- this additional data is added in the extraction of the data in Figure 1.
-This information was compared with Singer & Willet's original data.
-Comparison shown in Figure X shows that the information does not match up, however, there is a high correlation between the variables.
+* **NEED HELP**
 
 \begin{description}
-\item[Reviewer 1:] \textcolor{violet}{This sentence is unclear.
-What is dplyr used for?
-Are there tidyverse packages used but not mentioned?
-If not, splitting this sentence into two may make it clearer which package is used to what purpose.
-Also, I haven't tried it in code, but I imagine that the data could be tidied as described just using pivot\_longer, at least for the job number, year, and wage data.
-In other words, I don't see why dplyr or stringr would be needed for the data described on page 6.}
+\item[Reviewer 3:]\textcolor{violet}{The authors explain the complication of the “years of job experience” variable being unavailable in the raw data.
+This information is available in the NLSY79 data accessor; it seems to simply have been not selected in the initial data download by the authors.
+Given that the stated goal is to recreate the textbook data, this seems like a glaring oversight.
+It also leads to Fig 7 (Now Figure 8) being essentially meaningless: we are unsurprised that salaries are higher with more years of experience, and we are unsurprised that salaries go up over the years with both age of employee and inflation.
+The positive slope of these two different plots does not mean they are related to each other in any way.}
+\end{description}
 
-We did not go into great detail about the packages used as the whole code is available as a vignette within the package.
-We realise that this was not available to the reviewers due to the requirement of a blind review.
-To circumvent this, we have added the vignettes as supplementary materials.
+* We thank the reviewer for pointing out to us that the data contains information on the work experience -- this additional data is added in the extraction of the data in Figure 1.
+This information was compared with Singer & Willet's original data.
+Comparison shown in Figure 7 shows that the information does not match up, however, there is a high correlation between the variables.
+
+* **NEED HELP in responding comments on Figure 8.**
+
+\begin{description}
+\item[Reviewer 3:] \textcolor{violet}{My suggestion is that the authors reconfigure this paper to be presented as a pure case study.
+I believe the work that has been done is valuable: the data cleaning process is often messy and unplanned, so a case study like this on a popular dataset would provide an excellent educational resource.}
 \end{description}
+
+* We have kept the case study frame of mind but have re-cast the paper so that it focuses on the importance of this type of activity for teaching of statistics and data science.
 
@@
 
-* We have fixed the sentences and the reference. XXX Which one?
+* We have fixed the sentences and the reference, it is reflected on the last paragraph in page 3.
 
@@
 
-* We have split the sentences and described the use of each package mentioned.
+* We did not go into great detail about the packages used as the whole code is available as a vignette within the package.
+Also, for the blind review purpose, we have made all of the code and the raw dataset available to reviewers and it is not depend on the R package we created.
+The reviewers can see the data processing code and run it to also see the reproducibility of the code.
+Besides, in the paper, we have split the sentences and described the use of each package mentioned.
 
@@
 \end{description}
+
+* **NEED HELP**
 
@@
 
-* There is no individuals up to 8th grade in the refreshed dataset as we only filtered the high school dropouts (started from 9th grade ,i.e., high school grade).
-The plausible reasons for the differences are we might have different definition of high school dropouts with the original data and the corresponding individual back to high school after they dropped out so that their highest grade completed is 12th grade.
+* There are 108 individuals up to 8th grade in the refreshed dataset, compared to 366 individuals in the original dataset.
+The plausible reason for this difference is that some people have had GED after the period covered in the original dataset.
 
@@
 
+* We have it more explicit as per the reviewers' comments.
+
 \begin{description}
@@
 
+* **NEED HELP**
+
 \begin{description}
@@
 \end{description}
+* **NEED HELP**
 
@@
 
-* We removed this sentence as we already have the `yr_workforce` variable which is calculated from the year of the individual start working.
+* We removed this sentence as we already have the `yr_wforce` variable which is calculated from the year of the individual start working.
 
@@
 \end{description}
+
+* We have filter the x-axis of the refreshed data set until 1994, so they are comparable.
 
@@
 
-We agree that this level of precision was not necessary. We have modified Table 2 so percentages are to the nearest 1 percent and total percentage (=100%) is omitted.
+* We agree that this level of precision was not necessary. We have modified Table 2 so percentages are to the nearest 1 percent and total percentage (=100%) is omitted.
 
@@
 
-* These are two different variables. We have made this easier to follow in the manuscript as: [FILL]
+* These are two different variables. We have made this easier to follow in the manuscript as we add explanation on what the target data is (See Subsection 2.2: Target Data).
 
@@
 
-* Yes, the data is collected on a yearly basis.
+<!--* Yes, the data is collected on a yearly basis.
 However, according to the database, the highest grade completed variable is revised as in May 1st of each survey year.
 Thus, we mentioned as the revised May data.
+In the revised version, we only mention it as revised data to avoid confusion.-->
+
+*  The database has the variable of `the highest grade completed ever`.
+Thus, we use this variable for `hgc` instead of calculating it from the`hgc` the respondents answer for each round of the survey.
+Besides, we also add new variable `grade`, which is the highest grade completed for each round of survey, so the value changes corresponding to the year of the survey.
+According to the database, the `hgc` variable is revised as in May 1st of each survey year.
+Thus, we mentioned as the revised May data.
 In the revised version, we only mention it as revised data to avoid confusion.
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Abstract: The abstract portrays this work as ”re-creating a textbook example data set.” This doesn’t appear to be quite accurate because there are several notable differences between the end result and the textbook data.
+Currently, it’s more about demonstrating a procedure for “initial data analysis.”}
+\end{description}
+
+* **NEED HELP**

### 2022-02-16 data-raw/NLSY79/NLSY79.R

@@ -8,6 +8,7 @@ names(new_data) <- c('E0010102',
   'R0000100',
   'R0000300',
   'R0000500',
+  'R0017300',
   'R0018300',
   'R0047010',
   'R0070500',
@@ -24,6 +25,9 @@ names(new_data) <- c('E0010102',
   'R0214700',
   'R0214800',
   'R0215300',
+  'R0216700',
+  'R0216701',
+  'R0229200',
   'R0230000',
   'R0263710',
   'R0337800',
@@ -36,9 +40,12 @@ names(new_data) <- c('E0010102',
   'R0373410',
   'R0383800',
   'R0384910',
+  'R0406400',
+  'R0406401',
   'R0406700',
   'R0410100',
   'R0410300',
+  'R0417400',
   'R0418200',
   'R0446810',
   'R0545600',
@@ -51,7 +58,10 @@ names(new_data) <- c('E0010102',
   'R0585910',
   'R0598000',
   'R0599010',
+  'R0618900',
+  'R0618901',
   'R0645800',
+  'R0664500',
   'R0665300',
   'R0702510',
   'R0839900',
@@ -65,6 +75,9 @@ names(new_data) <- c('E0010102',
   'R0892300',
   'R0893410',
   'R0897300',
+  'R0898200',
+  'R0898201',
+  'R0905900',
   'R0906700',
   'R0945610',
   'R1087300',
@@ -77,7 +90,10 @@ names(new_data) <- c('E0010102',
   'R1127810',
   'R1140100',
   'R1141010',
+  'R1145000',
+  'R1145001',
   'R1145800',
+  'R1205800',
   'R1206600',
   'R1256010',
   'R1463000',
@@ -90,7 +106,10 @@ names(new_data) <- c('E0010102',
   'R1503210',
   'R1515400',
   'R1516310',
+  'R1520200',
+  'R1520201',
   'R1521000',
+  'R1605100',
   'R1605900',
   'R1650810',
   'R1809800',
@@ -103,7 +122,10 @@ names(new_data) <- c('E0010102',
   'R1848810',
   'R1860600',
   'R1861510',
+  'R1890900',
+  'R1890901',
   'R1891700',
+  'R1905600',
   'R1906100',
   'R1923410',
   'R2171500',
@@ -116,7 +138,10 @@ names(new_data) <- c('E0010102',
   'R2213210',
[…]

### 2022-02-16 data-raw/data-preprocessing.R

@@ -57,21 +57,22 @@ demog_tidy <- categories_qnames %>%
          gender = SAMPLE_SEX_1979)
 
 ## ---- demog-ed
-# demog_education <- new_data_qnames %>%
-#   # in 2018, the variable's name is Q3-4_2018, instead of HGC_2018
-#   rename(HGC_2018 = `Q3-4_2018`) %>%
-#   select(id = CASEID_1979,
-#          starts_with("HGCREV"),
-#          "HGC_2012",
-#          "HGC_2014",
-#          "HGC_2016",
-#          "HGC_2018") %>%
-#   pivot_longer(!id,
-#                names_to = "var",
-#                values_to = "grade") %>%
-#   separate("var", c("var", "year"), sep = "_") %>%
-#   filter(!is.na(grade)) %>%
-#   select(-var)
+demog_education <- new_data_qnames %>%
+  # in 2018, the variable's name is Q3-4_2018, instead of HGC_2018
+  rename(HGC_2018 = `Q3-4_2018`) %>%
+  select(id = CASEID_1979,
+         starts_with("HGCREV"),
+         "HGC_2012",
+         "HGC_2014",
+         "HGC_2016",
+         "HGC_2018") %>%
+  pivot_longer(!id,
+               names_to = "var",
+               values_to = "grade") %>%
+  separate("var", c("var", "year"), sep = "_") %>%
+  mutate(year = as.numeric(year)) %>%
+  #filter(!is.na(grade)) %>%
+  select(-var)
 
 hgc_ever <- new_data_qnames %>%
   select(id = CASEID_1979,
@@ -89,6 +90,60 @@ hgc_ever <- new_data_qnames %>%
 
 ## ---- tidy-hgc
 
+# Get the highest year completed
+hgc_calc <- demog_education %>%
+  filter(!is.na(grade)) %>%
+  group_by(id) %>%
+  filter(grade == max(grade)) %>%
+  filter(year == min(year)) %>%
+  rename(yr_hgc = year,
+         hgc_i = grade) %>%
+  select(id, yr_hgc, hgc_i) %>%
+  ungroup() %>%
+  mutate(hgc = case_when(hgc_i == 0 ~ "NONE",
+                         hgc_i == 1 ~ "1ST GRADE",
+                         hgc_i == 2 ~ "2ND GRADE",
+                         hgc_i == 3 ~ "3RD GRADE",
[…]

### 2022-02-16 data-raw/ida.R

@@ -261,7 +261,7 @@ wages_clean <- wages_cleaned %>%
 
 # rename and select the wages in tidy
 wages <- wages_clean %>%
-  select(id, year, mean_hourly_wage, age_1979, gender, race, hgc, hgc_i, ged = dip_or_ged,
+  select(id, year, mean_hourly_wage, age_1979, gender, race, grade, hgc, hgc_i, ged = dip_or_ged,
          number_of_jobs, total_hours, stwork, yr_wforce, exp, is_wm, is_pred) %>%
   mutate(id = as.factor(id),
          hgc = as.factor(hgc),

### 2022-02-16 response/response1.Rmd

@@
 
-* **Need help** XXX We need to review this literature. It's not true that we have ignored it, the main reference is Dasu! But we need to decide whether there is more that we should cite.
+We agree that there is substantial literature on data cleaning.
+We have referenced one of the main works, Dasu and Johnson (2003) but don't feel it is necessary to cite more than this.
+We have similarly only pointed to one main reference for IDA and EDA.
 

### 2022-02-16 paper/paper-jsdse/paper-jsdse.tex
Add all of the reviewers' comments in the response file, add some of the response as well.

@@
 EDA can be considered to encompass IDA.
-In practice the three stages of cleaning, summarizing and exploring are cyclical, that one often needs to do more cleaning after scrutinizing.
+In practice, the three stages of cleaning, summarizing, and exploring are cyclical, that one often needs to do more cleaning after scrutinizing.
 Dasu and Johnson (2003) say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.
@@
 As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year.
-The raw data consequently has a large number of columns (663 to be specific).
+The raw data consequently has a large number of columns (742 to be specific).
 The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars.
@@
 \begin{verbatim}
-#> 'data.frame':    12686 obs. of  663 variables:
+#> 'data.frame':    12686 obs. of  742 variables:
 #>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 ...

### 2022-02-16 paper/paper.Rmd
Add all of the reviewers' comments in the response file, add some of the response as well.

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by @SingerJudithD2003Alda which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel.
+An example of this, is the wages data made public by @SingerJudithD2003Alda in their book, Applied longitudinal data analysis: Modeling change and event occurrence,  which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
@@
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
-An exploratory analysis reveals however that the individual experience varies a great deal from the overall average.
+An exploratory analysis reveals, however, that the individual experience varies a great deal from the overall average.
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
@@
 
-The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+<!--The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
 An exploratory analysis reveals however that the individual experience varies a great deal from the overall average.
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
+[DA: It seems that this paragraph is repetion form the previous paragraph.]-->
 
@@
 EDA can be considered to encompass IDA.
-In practice the three stages of cleaning, summarizing and exploring are cyclical, that one often needs to do more cleaning after scrutinizing.
+In practice, the three stages of cleaning, summarizing, and exploring are cyclical, that one often needs to do more cleaning after scrutinizing.
 @DasuTamraparni2003Edma say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.
@@
 The next step is tidying to obtain `hgc` and `ged`.
+Besides that `hgc` is available in the original data, along with `ged`, it is used to subset the data to the high school dropouts as in the original data.
 As the highest grade completed is already exist in  `new_data_qnames` and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be `hgc_i` and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
 For `ged`, we only separate the year of survey and the GED status of the individual.
-Later on, this variable is useful to determine the refreshed high school dropouts data.
 

### 2022-02-16 paper/paper.tex
Add all of the reviewers' comments in the response file, add some of the response as well.

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel.
+An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
-An exploratory analysis reveals however that the individual experience varies a great deal from the overall average.
+An exploratory analysis reveals, however, that the individual experience varies a great deal from the overall average.
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
 The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at CENSORED.
-
-The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
-An exploratory analysis reveals however that the individual experience varies a great deal from the overall average.
-Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
 

### 2022-02-16 response/response1.Rmd
Add all of the reviewers' comments in the response file, add some of the response as well.

@@
 
-We agree that there is substantial literature on data cleaning.
+* We agree that there is substantial literature on data cleaning.
 We have referenced one of the main works, Dasu and Johnson (2003) but don't feel it is necessary to cite more than this.
@@
 * We thank the reviewer for pointing out to us that the data contains information on the work experience -- this additional data is added in the extraction of the data in Figure 1.
+We also add a Subsection dedicated to explain this variable, i.e., Subsection 3.2.
+Calculated variables: work experience.
 This information was compared with Singer & Willet's original data.
@@
 
+* Sometimes, the natural log is used to rescale the variable in the graph to adjust with the skewness in the data.
+Singer and Willet (2003) use the natural log to store the wages data by default, although without mention their motivation of doing so.
+Thus, we also use the natural log in Figure 8 to compare the refreshed data with the original data.
+
 \begin{description}
@@
 
-* **NEED HELP**
+* We have mentioned the book's title in the Introduction. The motivation of refreshing this text book data and how it is used for teaching purpose are discussed in the Introduction of the paper.
 
@@
 \end{description}
+
 * **NEED HELP**
@@
 \begin{description}
-\item[Reviewer 2:]\textcolor{violet}{
-What does “It implies that the fluctuation can still be observed in the data after the treatment.” mean?
+\item[Reviewer 2:]\textcolor{violet}{What does “It implies that the fluctuation can still be observed in the data after the treatment.” mean?
 I suspect this sentence should be replaced with something along the lines of “The figure shows that fluctuations in wages still exist even in the imputed data.”}
@@
 
-* We removed this sentence as we already have the `yr_wforce` variable which is calculated from the year of the individual start working.
+* We removed this sentence as we already have the `stwork` and `yr_wforce`  variables which are the year of individual start working and length of the years of individual entering the labor force, respectively.
 
@@
 * **NEED HELP**
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Where is IDA undervalued and neglected? Research? Teaching? Both?}
+\end{description}
+
+* We removed this sentences, and edited the explanation about IDA in Paragraph 6 of the Introduction.
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{The authors discuss refreshing an example from Singer and Willet (2003), but this isn’t ever completed.
+They compare the data, but not the longitudinal model.
+Again, it’s more about demonstrating a procedure for “initial data analysis” than it is reproducing Singer and Willet (2003).}
+\end{description}
+
+* Our main aim in this paper is to refresh the data, which is used as one of examples of longitudinal data and its modelling in Singer and Willet (2003).
+In other words, we only aims to refresh the data  until the most recent survey period, and not the longitudinal model, so that it is more relevant for today's context, especially, if its is used for teaching.
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Tidy and tame data are discussed in passing and the utility of tidy and tame data are not obvious.
+The authors later describe tidy data rather clearly (page 7) but tame data never receives the same treatment.}
+\end{description}
+
+* We put explanation on the difference between tidy and tame data in the Introduction (Paragraph 7). Tame data is sort of a result product, in which, in order to produce it, we need to tidy the data.
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{A historical context of NLSY79 would be helpful.
+Specifically, the dropped subjects are mentioned but there’s no context of why.
+By the argument of the paper, these steps should be transparent and well-considered.}
+\end{description}
+
+* We have added the reason of why some subject were dropped from the interview in Section 2.1.
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{The author switches between gender (man/woman/etc) and sex (male/female/etc.) throughout the manuscript.
+Further, the explanation here covers important ideas in teaching how to use data on gender/sex and race, but it’s just the beginning of ideas.
+It might be better to discuss the questions on the survey and discuss possible shortcomings, rather than make a somewhat empty gesture towards today’s standard.}
+\end{description}
+
+* **NEED HELP**
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{The discussion around highest grade completed is mechanical in what was done, but lacks substantive reasoning for those decisions and possible downstream consequences.}
+\end{description}
+
+* The variable of `highest grade completed`, besides it is available in the original data, it is also use to filter the high school dropouts (the subset of NLSY79 cohort that is contained in the original data).
+We explain how we use this variable, along with a newly added variable, GED, and possible downstream consequences in Subsection 4.1 Filtering: Determining who is a dropout.
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{It appears the goal of the paper switches to just IDA and dealing with problematic points here. This is completed in the paper.} 
+\end{description}
+
+* **NEED HELP**
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Gender (man/woman/etc). Sex (male/female/etc).} 
+\end{description}
+
+* **NEED HELP**
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{It is not clear why participants without a high school degree are not part of the analysis.} 
+\end{description}
+
+* In the current version of the manuscript, we have included all of the cohort, regardless of their school degree in a dataset called `wages`.
+However, we also subset the data to be high school dropouts only as we aims to refresh Singer and Willet's data (only contains male high school dropouts aged 14-17 years old when first interviewed).
+Hence, for comparability, we only analyse the high school dropouts in Section 4.
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Figure 2 is helpful for observing the data there is a lot of variability in both hourly wage and trends.
+A discussion about censored data might be helpful here – this is not something you impute from the data.}
+\end{description}
+
+* **NEED HELP**
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{You’re using iteratively reweighted least squares with Huber weighting.} 
+\end{description}
+
+* **NEED HELP**
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{I’m not sure this is accurate.
+The robustlmm package in R fits these models rather quickly with thousands of observations.
+What are the authors using?
+Does it run out of memory (requires more RAM) or does it take too much time?}
+\end{description}
+
+* We use `MASS` package instead of `robustlmm` package.
+We have tried to use `robustlmm`, but perhaps because the data is large and imbalanced, the model was not converging.
+Also, for reproducibility reason, we made it more affordable (in terms of memory and time), especially when people want to perform the stages in cleaning the data in class/teaching.
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{The choice of 0.12 seems arbitrary at best.
+I don’t see any justification for this approach (e.g., literature, simulations, etc).
+Why not use bi-square, instead of huber, and simply impute for observations with weight 0.
+While still arbitrary, it’s decided by the bisquare model and isn’t a guesstimate.
+Further, it would be sensible for any dataset and not just this one.
+(This would correspond to a smaller threshold, which may catch too many strange outcomes).
+In any case, it’s hard to justify something so adhoc.}
+\end{description}
+
+* **NEED HELP**
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{It’s unclear about what fluctuation or treatment the author is referring to.}
+\end{description}
+
+* Fluctuation is the variability in the data (up and down in the wages), while treatment is imputing the extremely high wages (anomaly) with their RLM's predicted value.
+The detail explanation of fluctation is discussed in Subsection 3.3 Initial Data Analysis, while the treatment is explained in Subsection 3.4 (Paragraph 3).
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{If the goal was to reproduce the textbook data, then why weren’t wages inflation-adjusted?}
+\end{description}
+
+* This is not done for the refreshed data because we plan to keep refreshing it as the data from the newest survey period  release.
+Thus, the 1990 prices might not be relevant anymore for the future context.
+Hence, we keep it as the original value to allow the users to treat it as their needs.
+In other words, the inflation-adjustment of the wages is better to do with each wave of new data added, so that it is relative to the last date in the data.
+
+* This explanation exists in the manuscript in Section 4.2 and Section 5.
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Figure 7 (Figure 8 in the current manuscript) is interesting, but it wasn’t until my second time through that I realized this was the textbook example for replicating.
+There is a clear difference in analyses (different question really), but is there anything to justify a similarity, or improvement of the analysis?}
+\end{description}
+
+* **NEED HELP**
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{(page 14, line 46) Thank you for this acknowledgment.
+I think this is true for most analyses.
+It doesn’t mean that you shouldn’t transparently and fully investigate your own choices.}
+\end{description}
+
+* **NEED HELP**

### 2022-02-16 paper/paper-jsdse/paper-jsdse.tex
Added the response and mark 'Need Help'.

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel.
+An example of this, is the wages data made public by Singer and Willett (2003) in their book, Applied longitudinal data analysis: Modeling change and event occurrence, which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 The next step is tidying to obtain \texttt{hgc} and \texttt{ged}.
+Besides that \texttt{hgc} is available in the original data, along with \texttt{ged}, it is used to subset the data to the high school dropouts as in the original data.
 As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
 For \texttt{ged}, we only separate the year of survey and the GED status of the individual.
-Later on, this variable is useful to determine the refreshed high school dropouts data.
 
@@
 
-\caption{Summary plots to check the cleaned data reveal more cleaning is necessary: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual (id=39) with an unusual wage relative to their years of data (C).
+\caption{(A) Summary plots to check the cleaned data reveal more cleaning is necessary: longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, and maximum wages of each individual, (C) and one individual (id=39) with an unusual wage relative to their years of data.
 Some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.}\label{fig:feature-plot}
@@
 
-\caption{Re-make of the summary plots of the fully processed data suggest that it is now in a reasonable state: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual with an unusual wage relative to their years of data (C).
+\caption{Re-make of the summary plots of the fully processed data suggest that it is now in a reasonable state: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, (C) and maximum wages of each individual, and one individual with an unusual wage relative to their years of data.
 }\label{fig:fixed-feature-plot}

### 2022-02-16 paper/paper.Rmd
Added the response and mark 'Need Help'.

@@
 When you return to the plots for the entire dataset in Figure 5, please make parallel comments.
--->However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages.
-Some of them, for example respondents in Figure \ref {fig:feature-plot} C had experienced unusual wages only in certain year.
+-->
+However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages, the maximum value of the wages unbelievably reached about $60,000/hour.
+Besides, some of them, for example respondents in Figure \ref {fig:feature-plot} C, had experienced unusual wages only in certain year.
 
@@
 
-```{r feature-plot, fig.cap = "Summary plots to check the cleaned data reveal more cleaning is necessary: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual (id=39) with an unusual wage relative to their years of data (C).
-Some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.", fig.width=10, fig.height=4, out.width="100%", cache = FALSE}
+```{r feature-plot, fig.cap = "Summary plots to check the data after the tidying stage: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, and maximum wages of each individual, (C) and one individual (id=39) with an unusual wage relative to their years of data.
+It reveals that some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.
+Accordingly, more cleaning is necessary to treat these extreme values.", fig.width=10, fig.height=4, out.width="100%", cache = FALSE}
 ```
@@
 I suspect not, but please be explicit.-->
-The anomalies are also found in the total hours of work, where some observations reported as having worked for 420 hours a week in total.
+Extremely high values were also found in the total hours of work, where some observations reported as having worked for 420 hours a week in total.
 According to @MichaelRPergamit2001DWTN, one of the flaws of the NLSY79 employment data is that the NLSY79 collects the information of the working hours since the last interview.
@@
 Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
-It implies that the fluctuation can still be observed in the data after the treatment.
+The figure shows that fluctuations in wages still exist even in the imputed data.
 However, the large spikes, which are considered "erroneous outliers", are already eliminated from the data.
@@
 
-```{r fixed-feature-plot, fig.cap = "Re-make of the summary plots of the fully processed data suggest that it is now in a reasonable state: longitudinal profiles of wages for all individuals 1979-2018 (A), boxplots of minimum, median, and maximum wages of each individual (B), and one individual with an unusual wage relative to their years of data (C).
+```{r fixed-feature-plot, fig.cap = "Re-make of the summary plots of the fully processed data suggest that it is now in a reasonable state: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, (C) and maximum wages of each individual, and one individual with an unusual wage relative to their years of data.
 ", fig.width=10, fig.height=4, out.width="100%", cache = FALSE}
@@
 - calculating experience in the workforce requires comparing the date of first job with the first year the individual was recorded, both of which are available in the database. 
-- there are many unusually high hourly wages, e.g. greater than $30,000 per hour.
-- there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with the individual's age.
-- matching id's from the original data with those in the refreshed data, do refer to the same person based on the demographic information available.
+- treating the extreme values since there are many unusually high hourly wages, e.g. greater than $60,000 per hour.
+- determining the dropouts subset as there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with their GED status.
+- matching ID's from the original data with those in the refreshed data, do refer to the same person based on the demographic information available.
 

### 2022-02-16 paper/paper.tex
Added the response and mark 'Need Help'.

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
+An example of this, is the wages data made public by Singer and Willett (2003) in their book, Applied longitudinal data analysis: Modeling change and event occurrence, which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.

### 2022-02-16 response/response1.Rmd
Added the response and mark 'Need Help'.

@@
 
+* We have made this sentences more explicit that the extremely high values were also observed in the `total hours of work`.
+
 \begin{description}
@@
 
+* We have replaced this sentences as per the suggestion.
+
 \begin{description}
@@
 
+* We have added the specifity of the extremely high wages in Subsection 3.3. Initial Data Analysis. We have also edited every bullet so it started with verb -ing.
+
 \begin{description}
@@
 
+* We have edited the caption.
+
 \begin{description}
@@
 \end{description}
+
+* We have moved the letter label before the description.
 
@@
 
-* We have filter the x-axis of the refreshed data set until 1994, so they are comparable.
+* We have filtered the x-axis of the refreshed data set until 1994, so they are comparable.
 
@@
 
-* Fluctuation is the variability in the data (up and down in the wages), while treatment is imputing the extremely high wages (anomaly) with their RLM's predicted value.
+* Fluctuation is the variability in the data (up and down) in the wages, while treatment is imputing the extremely high wages (anomaly) with their RLM's predicted value.
 The detail explanation of fluctation is discussed in Subsection 3.3 Initial Data Analysis, while the treatment is explained in Subsection 3.4 (Paragraph 3).
+We also have edited this sentences to avoid implicity.
 

### 2022-02-16 paper/paper-jsdse/paper-jsdse.tex
Edited the paper, add information about grade.

@@
 Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.
-However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages.
-Some of them, for example respondents in Figure \ref {fig:feature-plot} C had experienced unusual wages only in certain year.
+However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages, the maximum value of the wages unbelievably reached about \$60,000/hour.
+Besides, some of them, for example respondents in Figure \ref {fig:feature-plot} C, had experienced unusual wages only in certain year.
 
@@
 
-\caption{(A) Summary plots to check the cleaned data reveal more cleaning is necessary: longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, and maximum wages of each individual, (C) and one individual (id=39) with an unusual wage relative to their years of data.
-Some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.}\label{fig:feature-plot}
+\caption{Summary plots to check the data after the tidying stage: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, and maximum wages of each individual, (C) and one individual (id=39) with an unusual wage relative to their years of data.
+It reveals that some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.
+Accordingly, more cleaning is necessary to treat these extreme values.}\label{fig:feature-plot}
 \end{figure}
 
-The anomalies are also found in the total hours of work, where some observations reported as having worked for 420 hours a week in total.
+Extremely high values were also found in the total hours of work, where some observations reported as having worked for 420 hours a week in total.
 According to Pergamit et al.
@@
 Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
-It implies that the fluctuation can still be observed in the data after the treatment.
+The figure shows that fluctuations in wages still exist even in the imputed data.
 However, the large spikes, which are considered ``erroneous outliers'', are already eliminated from the data.
@@
 \item
-  there are many unusually high hourly wages, e.g.~greater than \$30,000 per hour.
-\item
-  there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with the individual's age.
-\item
-  matching id's from the original data with those in the refreshed data, do refer to the same person based on the demographic information available.
+  treating the extreme values since there are many unusually high hourly wages, e.g.~greater than \$60,000 per hour.
+\item
+  determining the dropouts subset as there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with their GED status.
+\item
+  matching ID's from the original data with those in the refreshed data, do refer to the same person based on the demographic information available.
 \end{itemize}

### 2022-02-16 paper/paper.Rmd
Edited the paper, add information about grade.

@@
 For example, use a single categorical race variable instead of the two binary race variables.
-We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), the number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
+We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
 We do not attempt to re-create the unemployment rate variable.
@@
 \begin{itemize}
+\item[$\triangleright$] Highest grade completed
+\begin{itemize}
+\item[\faCheck] All 80 variables in Highest grade completed were selected.
+\end{itemize}
+\end{itemize}
+\begin{itemize}
 \item[$\triangleright$] Dates of diploma or degree
@@
 \item[\faCheck] All variables named Q3-8A were selected.
-\end{itemize}
-\end{itemize}
-\begin{itemize}
-\item[$\triangleright$] Highest grade completed
-\begin{itemize}
-\item[\faCheck] Selected HGC\_EVER, which is the highest grade respondent has ever reported completed and getting credit for.
 \end{itemize}
@@
 \begin{itemize}
-\item[\faCloudDownload] To download all 663 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download".
+\item[\faCloudDownload] To download all 742 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download".
 We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers.
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (`gender`), race (`race`), age (`age_1979`), highest grade completed (`hgc`), highest grade completed in terms of years, e.g.
+In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (`gender`), race (`race`), age (`age_1979`), highest grade completed reported in each round of survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g.
 9th grade = 9, 3rd year college = 15, (`hgc_i`), and whether the graduate equivalency diploma is obtained (`ged`).
@@
 -->
-The next step is tidying to obtain `hgc` and `ged`.
-Besides that `hgc` is available in the original data, along with `ged`, it is used to subset the data to the high school dropouts as in the original data.
-As the highest grade completed is already exist in  `new_data_qnames` and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be `hgc_i` and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
-For `ged`, we only separate the year of survey and the GED status of the individual.
+
+The refreshed dataset has two types of highest grade completed.
+The first one is the highest grade completed ever reported in the database (`hgc` and `hgc_i`, for the factor and integer type, respectively).
+For each ID, there is only one value of `hgc` and `hgc_i`.
+This variable is obtained from `new_data_qnames` with name `HGC_EVER_XRND` and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be `hgc_i` and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
+
+The second one is the highest grade completed corresponds to value  reported in each round of the survey (`grade`), meaning that the value can change.
+We decide to include this variable in the refreshed dataset to enrich the analysis possibly done with the data, for example if one would like to explore how the changes of highest grade completed in individuals affecting their wages.
+The highest grade completed are recorded in `new_data_qnames` as variables beginning with `Q3-4` and `HGC` with suffix of the year it was recorded.
+In addition the variables beginning with `HGCREV` contain the revised data.
+We choose to use this revised data because it seemed to have less missing and presumably has been checked.
+However, there is no revised data for 2012, 2014, 2016, and 2018, thus, we use the ordinary data for these years.
+<!--We also check the consistency between `hgc_i` and `grade`.
+Some problematic observations were found, for example, `grade` value is greater than `hgc_i`.
+We then fix this problem by replacing those `grade` that is greater than its `hgc_i`.
+-->
+
+The next step is tidying to obtain `ged`.
+Along with `hgc`, `ged` is used to subset the data to the high school dropouts as in the original data.
+The graduate equivalency status is saved as variable started with "Q3-8A" followed by the year of the survey.
+Thus, we only separate the year and the GED status.
+Although there GED status is asked in each round of the survey, we only retain the latest status of one's GED.
 
@@
 Finally, we save the resultant wage data on this cohort as `wages`.
+Note that we save `grade` variable in this dataset, instead of `demog_nlsy79` dataset because it is a longitudinal variable, while `demog_nlsy79` is cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.
 

### 2022-02-16 paper/paper.tex
Edited the paper, add information about grade.

@@
 \end{itemize}
-\begin{itemize}
-\item[$\triangleright$] Highest grade completed
-\begin{itemize}
-\item[\faCheck] Selected HGC\_EVER, which is the highest grade respondent has ever reported completed and getting credit for.
-\end{itemize}
-\end{itemize}
 \end{itemize}
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), and whether the graduate equivalency diploma is obtained (\texttt{ged}).
+In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), and whether the graduate equivalency diploma is obtained (\texttt{ged}).
 
@@
 
-The next step is tidying to obtain \texttt{hgc} and \texttt{ged}.
-Besides that \texttt{hgc} is available in the original data, along with \texttt{ged}, it is used to subset the data to the high school dropouts as in the original data.
-As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
-For \texttt{ged}, we only separate the year of survey and the GED status of the individual.
+The refreshed dataset has two types of highest grade completed.
+The first one is the highest grade completed ever reported in the database (\texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively).
+For each ID, there is only one value of \texttt{hgc} and \texttt{hgc\_i}.
+This variable is obtained from \texttt{new\_data\_qnames} with name \texttt{HGC\_EVER\_XRND} and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
+
+The second one is the highest grade completed corresponds to value reported in each round of the survey (\texttt{grade}), meaning that the value can change.
+We decide to include this variable in the refreshed dataset to enrich the analysis possibly done with the data, for example if one would like to explore how the changes of highest grade completed in individuals affecting their wages.
+The highest grade completed are recorded in \texttt{new\_data\_qnames} as variables beginning with \texttt{Q3-4} and \texttt{HGC} with suffix of the year it was recorded.
+In addition the variables beginning with \texttt{HGCREV} contain the revised data.
+We choose to use this revised data because it seemed to have less missing and presumably has been checked.
+However, there is no revised data for 2012, 2014, 2016, and 2018, thus, we use the ordinary data for these years.
+
+The next step is tidying to obtain \texttt{ged}.
+Along with \texttt{hgc}, \texttt{ged} is used to subset the data to the high school dropouts as in the original data.
+The graduate equivalency status is saved as variable started with ``Q3-8A'' followed by the year of the survey.
+Thus, we only separate the year and the GED status.
+Although there GED status is asked in each round of the survey, we only retain the latest status of one's GED.
 
@@
 Finally, we save the resultant wage data on this cohort as \texttt{wages}.
+Note that we save \texttt{grade} variable in this dataset, instead of \texttt{demog\_nlsy79} dataset because it is a longitudinal variable, while \texttt{demog\_nlsy79} is cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.
 

### 2022-02-16 paper/paper.Rmd
small fixes all over plus new check code for hgc

@@
 
-Our process of cleaning builds heavily on the `tidyverse` approach [@tidyverse].
-The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using the data pipeline and split-apply-combine approach [@plyr].
+Our approach to cleaning builds heavily on the `tidyverse` approach [@tidyverse].
+The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using a step-wise piping with split-apply-combine strategy for mutating new variables [@plyr].
 Tidy data shouldn't be confused with "tame data" which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
@@
 For example, use a single categorical race variable instead of the two binary race variables.
-We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
+We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
 We do not attempt to re-create the unemployment rate variable.
@@
 \item[\faCloudDownload] To download all 742 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download".
-We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers.
+We select the R Source code and comma-delimited datafile of selected variables with Reference Number as column headers.
 We name the filename "NLSY79" and press the download button.
@@
 
-- `NLSY79.csv`: Comma Separated Value format of the response data,
-- `NLSY79.dat`: .dat format of the response data,
-- `NLSY79.NLSY79`: Tagset of variables that can be uploaded to the website to recreate the data set, and
+- `NLSY79.csv`: comma separated value format of the response data,
+- `NLSY79.dat`: alternative text format of the response data,
+- `NLSY79.NLSY79`: tagset of variables that can be uploaded to the website to recreate the data set, and
 - `NLSY79.R`: R script for reading the data into R and converting the variables' names and label into something more sensible.

### 2022-02-16 paper/paper.tex
small fixes all over plus new check code for hgc

@@
 
-Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
+Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al.
 2019).
-The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
+The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using a step-wise piping with split-apply-combine strategy for mutating new variables (Wickham 2011).
 Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
@@
 \item
-  \texttt{NLSY79.csv}: Comma Separated Value format of the response data,
-\item
-  \texttt{NLSY79.dat}: .dat format of the response data,
-\item
-  \texttt{NLSY79.NLSY79}: Tagset of variables that can be uploaded to the website to recreate the data set, and
+  \texttt{NLSY79.csv}: comma separated value format of the response data,
+\item
+  \texttt{NLSY79.dat}: alternative text format of the response data,
+\item
+  \texttt{NLSY79.NLSY79}: tagset of variables that can be uploaded to the website to recreate the data set, and
 \item
@@
 Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
-This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
-2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham, François, et al.
+This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
+2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham et al.
 2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{/Users/cookd/numbats/yowie-paper/paper/figures/flowchart_blind}
 
@@
 2021a.
-``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
+{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
 Columbus, OH, through \url{https://www.nlsinfo.org/bibliography-citing-nls-data}.
 
-\leavevmode\hypertarget{ref-nlsy79guide}{}%
----------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.'' \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
-
-\leavevmode\hypertarget{ref-shiny}{}%
-Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.
-
-\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
-Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
-
-\leavevmode\hypertarget{ref-eliznlsy}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-nlsy79guide}{}}%
+---------. 2021b. {``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.''} \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-shiny}{}}%
+Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{{shiny: Web Application Framework for R}}. \url{https://CRAN.R-project.org/package=shiny}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chatfield1985TIEo}{}}%
+Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
+
+\leavevmode\vadjust pre{\hypertarget{ref-eliznlsy}{}}%
 Cooksey, Elizabeth C.
 2017.
-``Using the National Longitudinal Surveys of Youth (Nlsy) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
+{``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M.
 Lerner Neal Halfon Christoper B.

### 2022-02-16 data-raw/eda.R

@@ -89,7 +89,7 @@ do_ref_mod <- do %>%
 
 # XXX Code to check that the hgc matches
 do_check <- do %>%
-  mutate(hgc12 = ifelse(hgc_i < 9, "8TH", "12TH")) %>%
+  mutate(hgc12 = max(grade[between(do$year, 1979, 1994)])) %>% # ifelse(hgc_i < 9, "8TH", "12TH")) %>%
   select(id, hgc12) %>%
   distinct()
 sw_check <- sw %>%

### 2022-02-16 paper/paper.Rmd
filtering dropouts

@@
 
-An additional criteria from @SingerJudithD2003Alda is to only include males aged between 14 and 17 years old in 1979. Filters using these criteria used re-create the dropouts data.
-
+An additional criteria from @SingerJudithD2003Alda is to only include males aged between 14 and 17 years old in 1979.
 With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
@@
 2. 79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
+    i.
+35 of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded to 1).
+This suggests that they are not dropouts, and so we excluded them from the refreshed data.
+    ii. The information about `ged` is missing in 38 individuals. We decided to include them in the refreshed data. 
+    iii. 3 individuals have both diploma and GED (`ged` is coded to 3). These were kept in the refreshed data. 
+    iv. 12 individuals do not exist in `wages` data because they were participated in less than 3 rounds of survey.
 
-  + 35 of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded to 1). This suggests that they are not dropouts, and so we excluded them from the refreshed data. 
-  + The information about `ged` is missing in 38 individuals. We decided to include them in the refreshed data. 
-  + 3 individuals have both diploma and GED (`ged` is coded to 3). These were kept in the refreshed data. 
-  + 12 individuals do not exist in `wages` data because they were participated in less than 3 rounds of survey.
-
-The filtering was re-applied using these decisions. The refreshed dropout data has 863 individuals.
+The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing 863 individuals.
 

### 2022-02-16 paper/paper.tex
filtering dropouts

@@
 
-\begin{verbatim}
-#> # A tibble: 2 x 3
-#>   hgc   hgc12     n
-#>   <chr> <chr> <int>
-#> 1 12TH  8TH       1
-#> 2 8TH   12TH    243
-\end{verbatim}
-
 \begin{figure}
@@
 
-An additional criteria from Singer and Willett (2003) is to only include males aged between 14 and 17 years old in 1979. Filters using these criteria used re-create the dropouts data.
-
+An additional criteria from Singer and Willett (2003) is to only include males aged between 14 and 17 years old in 1979.
 With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
@@
   79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:
-\end{enumerate}
-
-\begin{itemize}
+
+  \begin{enumerate}
+  \def\labelenumii{\roman{enumii}.}
   \tightlist
@@
     12 individuals do not exist in \texttt{wages} data because they were participated in less than 3 rounds of survey.
-\end{itemize}
-
-The filtering was re-applied using these decisions. The refreshed dropout data has 863 individuals.
+  \end{enumerate}
+\end{enumerate}
+
+The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing 863 individuals.
 

### 2022-02-16 data-raw/eda.R

@@ -98,8 +98,8 @@ sw_check <- sw %>%
   select(id, hgc) %>%
   distinct()
 
-sw_check <- sw_check %>% left_join(do_check, by="id")
-sw_check %>% filter(hgc != hgc12) %>% count(hgc, hgc12)
+#sw_check <- sw_check %>% left_join(do_check, by="id")
+#sw_check %>% filter(hgc != hgc12) %>% count(hgc, hgc12)
 
 ## --- compare-sw-do
 sw_id <- as_tibble(sw) %>%

### 2022-02-17 response/response1.Rmd
revised fig 2 according to reviewer comment"
update response

@@
 
-* **NEED HELP**
+* The approach is for the case study only and is not meant to be extrapolated for other cases.
+We suggest using a robust mixed model as a better alternative, however, our attempt to fit such a model via `robustlmm` resulted in computational issues.
 
@@
 
-* **NEED HELP**
+* We have added in Introduction that "Section 3 presents the steps ... to find and repair anomalies.".
 
@@
 
-* **NEED HELP**
+* Thank you for your suggetion.
+We have modified the description to "To treat the extreme values in the data, we build a robust linear regression model where we use the robustness weight to determine if a value should be replaced with the fitted value from the model."
 
@@
 
-* **NEED HELP**
+* Thank you for your enquiry.
+We have re-written this section to better describe our process and intent.
+In particular, the threshold values are chosen based on extensive eye-balling of the plots of wages over time for each individual.
+We try to maintain a balance of not overly smoothing the observations but clearly repairing unlikely wage values.
+As alluded in the section, we believe that an alternative approach such as robust linear mixed models is more appropriate but we did not proceed with this as we had issues fitting the model with `robustlmm` package.
+In addition, we thought that the use of robust linear models is easier to explain in the context of teaching and the use of non-optimal approach can generate a good discussion about alternative approaches in class.
 

### 2022-02-17 paper/paper.Rmd
fig 2 text

@@
 
-```{r sample-plot, echo = FALSE, fig.cap="Longitudinal profiles of wages for a random sample of 20 individuals in the refreshed data.
-Most, but not all, individuals here experienced increasing wages over time, and several have experienced considerable fluctuation in wages.
-Some individuals are only measured for a short period.", fig.height=8, fig.width=10, out.width="100%"}
+```{r sample-plot, echo = FALSE, fig.cap="Longitudinal profiles of wages for a random sample of 36 individuals in the pre-cleaned data.
+There is considerable variation in wages.
+Some individuals  (2799, 11041, 11146) are only measured for a short period.
+Some individuals (8296, 9962) possibly have errors in wages in some years.", fig.height=8, fig.width=10, out.width="100%"}
 ```

### 2022-02-17 paper/paper.tex
fig 2 text

@@
 
-\caption{Longitudinal profiles of wages for a random sample of 20 individuals in the refreshed data.
-Most, but not all, individuals here experienced increasing wages over time, and several have experienced considerable fluctuation in wages.
-Some individuals are only measured for a short period.}\label{fig:sample-plot}
+\caption{Longitudinal profiles of wages for a random sample of 36 individuals in the pre-cleaned data.
+There is considerable variation in wages.
+Some individuals  (2799, 11041, 11146) are only measured for a short period.
+Some individuals (8296, 9962) possibly have errors in wages in some years.}\label{fig:sample-plot}
 \end{figure}

### 2022-02-17 paper/paper.Rmd
replaced rlm plot with same set of individuals as fig 2 - looks better

@@
 
-In the next step, we explore the mean hourly wage data.
-In this case, we only explore the wages data of respondents that have the highest completed grade of up to $12^{th}$ grade.
-We employ visualization techniques to perform the IDA as described next.
-
 ```{r sample-plot, echo = FALSE, fig.cap="Longitudinal profiles of wages for a random sample of 36 individuals in the pre-cleaned data.
@@
 Some individuals  (2799, 11041, 11146) are only measured for a short period.
-Some individuals (8296, 9962) possibly have errors in wages in some years.", fig.height=8, fig.width=10, out.width="100%"}
+Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation.", fig.height=8, fig.width=10, out.width="100%"}
 ```
@@
 
-We randomly take 20 samples from the data and plot them, as shown in Figure \ref{fig:sample-plot}.<!--AE comment: "It" implies...
+In the next step, we explore the mean hourly wage data of samples of individuals.
+The purpose is to examine the common patterns and check the quality.
+A random sample of 36 individuals is chosen (using the `sample_n_keys` function in `brolgar`).
+Their longitudinal profiles are plotted, faceted by `id` and using free $y$ scales, so that the individual patterns can be examined (Figure \ref{fig:sample-plot}).<!--AE comment: "It" implies...
 Please specify what "it" refers to.
@@
 "as shown in Figure 2, which shows that some respondents have high variability in wages, particularly 5, 7, and 11." However, 7 and 11 don't seem to have particularly high variability, especially when compared to 18.
-This whole paragraph needs to be fixed/updated.--> It shows that these respondents have a lot of variability in wages, for example, the IDs in panel numbers 5, 7, and 11.
-The plot also implies that the samples have a different pattern of mean hourly wages.
-Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, <!--AE comment: Not panel 9 --> the IDs in panel 9.
+This whole paragraph needs to be fixed/updated.--> There is a lot of variability from one individual to another, and substantial fluctuation in wages at different times for most individuals.
+Some individuals  (2799, 11041, 11146) are only measured for a short period.
+Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation.
+These need to be inspected more closely.
+
+ <!--AE comment: Not panel 9 --> the IDs in panel 9. 
 <!--I think there is a narrative step missing here, which is that looking at those samples led you to look at plots that show values for the wage variable over all participants and all years.
@@
 -->
-However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages, the maximum value of the wages unbelievably reached about $60,000/hour.
-Besides, some of them, for example respondents in Figure \ref {fig:feature-plot} C, had experienced unusual wages only in certain year.
+
+Figure \ref{fig:feature-plot}s shows an alternative way to check the data quality.
+Plot (A) is the spagetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to $60,000/hour) for some individuals mostly around 1990.
+Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median and maximum) for all individuals.
+This tells us that there are a number of individuals with unbelievably high maximum wages.
+Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to $1200/hour.
+This does not seem to be reasonable, and leads us to use a procedure to detect and fix these temporal anomalies.
 
@@
 
-Figure \ref{fig:flow-chart-blind} summarizes the steps taken to go from raw to input to valid data [@LooMarkvander2018Sdcw] to create a refreshed wages data set.
+There are many steps and decisions made to go from raw to input to valid data. Figure \ref{fig:flow-chart-blind} summarizes these in order to create a refreshed wages data set.
 

### 2022-02-17 paper/paper.tex
replaced rlm plot with same set of individuals as fig 2 - looks better

@@
 
-In the next step, we explore the mean hourly wage data.
-In this case, we only explore the wages data of respondents that have the highest completed grade of up to \(12^{th}\) grade.
-We employ visualization techniques to perform the IDA as described next.
-
 \begin{figure}
@@
 Some individuals  (2799, 11041, 11146) are only measured for a short period.
-Some individuals (8296, 9962) possibly have errors in wages in some years.}\label{fig:sample-plot}
+Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation.}\label{fig:sample-plot}
 \end{figure}
 
-We randomly take 20 samples from the data and plot them, as shown in Figure \ref{fig:sample-plot}.
-It shows that these respondents have a lot of variability in wages, for example, the IDs in panel numbers 5, 7, and 11.
-The plot also implies that the samples have a different pattern of mean hourly wages.
-Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.
-However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages, the maximum value of the wages unbelievably reached about \$60,000/hour.
-Besides, some of them, for example respondents in Figure \ref {fig:feature-plot} C, had experienced unusual wages only in certain year.
+In the next step, we explore the mean hourly wage data of samples of individuals.
+The purpose is to examine the common patterns and check the quality.
+A random sample of 36 individuals is chosen (using the \texttt{sample\_n\_keys} function in \texttt{brolgar}).
+Their longitudinal profiles are plotted, faceted by \texttt{id} and using free \(y\) scales, so that the individual patterns can be examined (Figure \ref{fig:sample-plot}).
+There is a lot of variability from one individual to another, and substantial fluctuation in wages at different times for most individuals.
+Some individuals (2799, 11041, 11146) are only measured for a short period.
+Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation.
+These need to be inspected more closely.
+
+the IDs in panel 9.
+
+Figure \ref{fig:feature-plot}s shows an alternative way to check the data quality.
+Plot (A) is the spagetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to \$60,000/hour) for some individuals mostly around 1990.
+Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median and maximum) for all individuals.
+This tells us that there are a number of individuals with unbelievably high maximum wages.
+Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to \$1200/hour.
+This does not seem to be reasonable, and leads us to use a procedure to detect and fix these temporal anomalies.
 
@@
 
-Figure \ref{fig:flow-chart-blind} summarizes the steps taken to go from raw to input to valid data (M. van der Loo and de Jonge 2018) to create a refreshed wages data set.
+There are many steps and decisions made to go from raw to input to valid data. Figure \ref{fig:flow-chart-blind} summarizes these in order to create a refreshed wages data set.
 

### 2022-02-17 data-raw/ida.R

@@ -157,10 +157,20 @@ spag + feature_bp + plot_high +
 set.seed(31251587)
 
 wages_cleaned <- readRDS(here::here("paper/results/wages_after.rds"))
-sample_id <- sample(unique(wages_cleaned$id), 20)
-sample <- subset(wages_cleaned, id %in% sample_id)
 
-wages_compare <- sample %>%
+# Use the same sample used in first plot
+set.seed(20220207)
+sample_id <- wages_before_tsibble %>%
+  sample_n_keys(size=36) %>%
+  as_tibble() %>%
+  select(id) %>%
+  distinct() %>%
+  pull()
+
+#sample_id <- sample(unique(wages_cleaned$id), 20)
+wages_sample <- subset(wages_cleaned, id %in% sample_id)
+
+wages_compare <- wages_sample %>%
   select(id, year, mean_hourly_wage, wages_rlm) %>%
   rename(mean_hourly_wage_rlm = wages_rlm) %>%
   pivot_longer(c(-id, -year), names_to = "type", values_to = "wages")

### 2022-02-17 paper/paper.Rmd
updated language on rlm

@@
 <!--AE comment: Awkward sentence.
-"which is the model formulation"?--> To treat the extreme values in the data, we build a robust linear regression model where we use the robustness weight to determine if a value should be replaced with the fitted value from the model.
-We build the model for each ID utilizing the `nest` and `map` function from `tidyr` [@tidyr] and `purrr` [@purrr], respectively.
-This fits a model separately for each individual, and a robust linear mixed model could be used as an alternative approach.
-We tried fitting a robust linear mixed model using `robustlmm` [@KollerManuel2016rARP], however perhaps because the data is large and imbalanced, the model was not converging.
+"which is the model formulation"?--> A robust linear regression model using the `rlm` function from `MASS` package [@mass] is used to treat the extreme values in the data.
+The robustness weight is used to determine if a value should be replaced with the fitted value from the model.
+This is constructed for each ID utilizing the `nest` and `map` function from `tidyr` [@tidyr] and `purrr` [@purrr], respectively.
+An alternative approach would be a robust linear mixed model using `robustlmm` [@KollerManuel2016rARP].
+However, when we tested this, the fit failed to convergence.
 <!--I would pull the tidy and purrr citations to the end of the sentence to make it read more smoothly-->The full code for this is shown at \url{`r yowie_pkgdown`/articles/input-to-valid-data.html} but also described in detail next.
 
-We build the model using the `rlm` function from `MASS` package [@mass].
-We set the `mean_hourly_wage` and `year` as the dependent and predictor, respectively.
+The `mean_hourly_wage` and `year` are set as the dependent and predictor, respectively.
 Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm].
 <!--AE comment: “However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.” I don’t think this sentence needs a “However” and probably should say “past which” rather than “in which”-->The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
-Moreover, it should be noted that not all the outliers are due to an error.
-Instead, it might be that one had reasonably increasing or decreasing wages in a particular period.
+It should be noted that it is not possible to determine if all outliers are errors, and it might be that an individual had abnormally high wages at a particular time.
 
@@
 
-To minimize the risk of mistakenly identifying an outlier as an "erroneous outlier", we simulate some thresholds and study how they affect the data by eye balling the results.
-Based on this eye balling, we find that 0.12 is the most reasonable value to use as the threshold to minimize the risk of overly smoothing the wage by retaining sensible spikes in the data.
-In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording.
-After deciding the threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.
-We then flag those observations in a new variable called `is_pred`.
+To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done.
+Samples of individuals were examined under the different thresholds, with an eye to smoothness of the profiles.
+Overly smooth profiles would indicate that the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
+A threshold of 0.12 was chosen, that struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
+Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.
+We then flag those observations in a new variable called `is_pred`, so that this change can be monitored in the downstream analyses.
 
@@
 
-```{r compare-plot, fig.cap="Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for a sample of individuals.
-A robust linear model prediction was used to correct mean hourly wages value.
-We can see that some extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed.
-Some spikes might remain when wage vaues are plausible.", fig.height=8, fig.width=10, out.width="90%"}
+```{r compare-plot, fig.cap="Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for same sample of individuals as shown in Figure 2.
+A robust linear model prediction was used to identify and correct mean hourly wages value.
+The extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed.", fig.height=8, fig.width=10, out.width="90%"}
 ```

### 2022-02-17 paper/paper.tex
updated language on rlm

@@
 
-To treat the extreme values in the data, we build a robust linear regression model where we use the robustness weight to determine if a value should be replaced with the fitted value from the model.
-We build the model for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively.
-This fits a model separately for each individual, and a robust linear mixed model could be used as an alternative approach.
-We tried fitting a robust linear mixed model using \texttt{robustlmm} (Koller 2016), however perhaps because the data is large and imbalanced, the model was not converging.
+A robust linear regression model using the \texttt{rlm} function from \texttt{MASS} package (Venables and Ripley 2002) is used to treat the extreme values in the data.
+The robustness weight is used to determine if a value should be replaced with the fitted value from the model.
+This is constructed for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively.
+An alternative approach would be a robust linear mixed model using \texttt{robustlmm} (Koller 2016).
+However, when we tested this, the fit failed to convergence.
 The full code for this is shown at \url{CENSORED/articles/input-to-valid-data.html} but also described in detail next.
 
-We build the model using the \texttt{rlm} function from \texttt{MASS} package (Venables and Ripley 2002).
-We set the \texttt{mean\_hourly\_wage} and \texttt{year} as the dependent and predictor, respectively.
+The \texttt{mean\_hourly\_wage} and \texttt{year} are set as the dependent and predictor, respectively.
 Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021).
 The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
-Moreover, it should be noted that not all the outliers are due to an error.
-Instead, it might be that one had reasonably increasing or decreasing wages in a particular period.
-
-To minimize the risk of mistakenly identifying an outlier as an ``erroneous outlier'', we simulate some thresholds and study how they affect the data by eye balling the results.
-Based on this eye balling, we find that 0.12 is the most reasonable value to use as the threshold to minimize the risk of overly smoothing the wage by retaining sensible spikes in the data.
-In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording.
-After deciding the threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.
-We then flag those observations in a new variable called \texttt{is\_pred}.
+It should be noted that it is not possible to determine if all outliers are errors, and it might be that an individual had abnormally high wages at a particular time.
+
+To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done.
+Samples of individuals were examined under the different thresholds, with an eye to smoothness of the profiles.
+Overly smooth profiles would indicate that the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
+A threshold of 0.12 was chosen, that struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
+Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.
+We then flag those observations in a new variable called \texttt{is\_pred}, so that this change can be monitored in the downstream analyses.
 
@@
 
-\caption{Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for a sample of individuals.
-A robust linear model prediction was used to correct mean hourly wages value.
-We can see that some extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed.
-Some spikes might remain when wage vaues are plausible.}\label{fig:compare-plot}
+\caption{Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for same sample of individuals as shown in Figure 2.
+A robust linear model prediction was used to identify and correct mean hourly wages value.
+The extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed.}\label{fig:compare-plot}
 \end{figure}

### 2022-02-17 paper/paper.Rmd
extreme values section revised

@@
 Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
-The figure shows that fluctuations in wages still exist even in the imputed data.
-However, the large spikes, which are considered "erroneous outliers", are already eliminated from the data.
-Hence, the model produces a data set with a more reasonable degree of fluctuation.
+The plot shows that fluctuations in wages remain but the large spikes (in this sample individuals 8296, 9962) which are considered implausible are replaced.
 
@@
 A robust linear model prediction was used to identify and correct mean hourly wages value.
-The extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed.", fig.height=8, fig.width=10, out.width="90%"}
+The extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years for individuals 8296 and 9962, but otherwise the profiles have not changed.", fig.height=8, fig.width=10, out.width="90%"}
 ```
 
-Further, Figure \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around $350.
-The spikes are still observed but not as extreme as the original data set.
-In Figure \ref{fig:fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value.
+Figure \ref{fig:fixed-feature-plot} shows the summary statistics after removing extremes.
+The highest wage overall is now around $1000.
+Plot (A) shows a more reasonable spagetti plot, where there are some profiles with high wages, but most profiles have wages under $300, and there is a steady increase in wages with years.
+Plot (B) shows that there are still a small number of individuals with high maximum wages.
+Plot (C) shows the profile for ID=39, after imputing the extreme value.
+The wages for this individual increase over the years, and do fluctuate some between 1900 and 2005.
 <!--AE comment: What does “It implies that the fluctuation can still be observed in the data after the treatment.” mean?
 I suspect this sentence should be replaced with something along the lines of “The figure shows that fluctuations in wages still exist even in the imputed data.”-->
-We still can see some extreme values in maximum wages, but consider it as a natural variability of the data.
-In Figure \ref{fig:fixed-feature-plot} C, we can see that after imputing the extreme value in ID=39, we can see how the the wages change over the years more clearly.
 
@@
 ```
-
-Finally, we save the imputed data and set the appropriate data type for the variables.
 

### 2022-02-17 paper/paper.tex
extreme values section revised

@@
 Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
-The figure shows that fluctuations in wages still exist even in the imputed data.
-However, the large spikes, which are considered ``erroneous outliers'', are already eliminated from the data.
-Hence, the model produces a data set with a more reasonable degree of fluctuation.
+The plot shows that fluctuations in wages remain but the large spikes (in this sample individuals 8296, 9962) which are considered implausible are replaced.
 
@@
 A robust linear model prediction was used to identify and correct mean hourly wages value.
-The extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed.}\label{fig:compare-plot}
+The extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years for individuals 8296 and 9962, but otherwise the profiles have not changed.}\label{fig:compare-plot}
 \end{figure}
 
-Further, Figure \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around \$350.
-The spikes are still observed but not as extreme as the original data set.
-In Figure \ref{fig:fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value.
-We still can see some extreme values in maximum wages, but consider it as a natural variability of the data.
-In Figure \ref{fig:fixed-feature-plot} C, we can see that after imputing the extreme value in ID=39, we can see how the the wages change over the years more clearly.
+Figure \ref{fig:fixed-feature-plot} shows the summary statistics after removing extremes.
+The highest wage overall is now around \$1000.
+Plot (A) shows a more reasonable spagetti plot, where there are some profiles with high wages, but most profiles have wages under \$300, and there is a steady increase in wages with years.
+Plot (B) shows that there are still a small number of individuals with high maximum wages.
+Plot (C) shows the profile for ID=39, after imputing the extreme value.
+The wages for this individual increase over the years, and do fluctuate some between 1900 and 2005.
 
@@
 \end{figure}
-
-Finally, we save the imputed data and set the appropriate data type for the variables.
 

### 2022-02-17 paper/paper.Rmd
modifications to the takeaways section

@@
 Secondly, in the original dataset, wages were inflation-adjusted to 1990 prices.
-This is not done for the refreshed data because we plan to keep refreshing it as the data from the newest survey period  release.
-Thus, the 1990 prices might not be relevant anymore for the future context.
-Hence, we keep it as the original value to allow the users to treat it as their needs.
+This is not done for the refreshed data because we plan to keep refreshing it as the data added to and released from the survey.
+Instead we have provided a function in the R package, `yowie`, for users to conduct the inflation adjustment when they are ready to analyse the data.
 
@@
 Either way, please clarify.-->
-It is also worth noting that the treatment of unlikely wages differ in the refreshed data.
+It is important to note that the treatment of unlikely wages differs in the refreshed data.
 In the original data by @SingerJudithD2003Alda, wages greater than $75 are set to be missing.
-However, in recent context, this value is too low to be set as the maximum threshold.
-We opted to use the weights from a robust linear regression to determine what should be regarded as outliers and imputed them with their predicted values as as described in Section \@ref(censor).
+However, this value is too low to be set as the maximum threshold, and it doesn't take into account temporal neighbors for an individual.
+We opted to use the weights from a robust linear regression to determine what should be regarded as extreme and imputed them with their predicted values as as described in Section \@ref(censor).
 

### 2022-02-17 data-raw/eda.R
Tidy the code scripts.

@@ -1,9 +1,26 @@
 ## ---- sw_wages
 
+# load the original data
 sw <- brolgar::wages %>%
   group_by(id) %>%
   mutate(index = 1:n(),
          id = as.factor(id))
+
+# load the dataset of hgc in 1979
+hgc_1979 <- hgc_1979 %>%
+  mutate(id = as.factor(id))
+
+# load the refreshed data, join it with the hgc_1979
+do <- wages_hs_do %>%
+  as_tibble() %>%
+  left_join(hgc_1979, by = "id") %>%
+  group_by(id) %>%
+  mutate(lnwage = log(wage),
+         index = 1:n())
+
+# join original and refreshed data, only for the same individuals that appears in the two data
+
+do_sw_join <- inner_join(sw, do, by = c("id", "index"))
 
 sw_wages <- sw %>%
   ggplot(aes(x = xp,
@@ -17,7 +34,7 @@ sw_wages <- sw %>%
   #theme(plot.title = element_text(size = 10)) +
   ylim(0, 5)
 
-sw_wages_mod <- sw %>%
+sw_wages_mod <- do_sw_join %>%
   as_tibble() %>%
   mutate(hgc = ifelse(high_grade < 9, "8TH", "12TH")) %>%
   mutate(race = case_when(black == 1 ~ "black",
@@ -48,11 +65,7 @@ sw_wages_mod <- sw %>%
 
 # filter do data up to 1994 since the textbook data only covers that period.
 
-do <- wages_hs_do %>% #yowie::wages_hs_do %>%
-  as_tibble() %>%
-  group_by(id) %>%
-  mutate(lnwage = log(wage),
-         index = 1:n())
+
 
 do_ref <- do %>%
   ggplot(aes(x = year,
@@ -66,8 +79,8 @@ do_ref <- do %>%
   theme(plot.title = element_text(size = 10)) +
   ylim(0, 5)
 
-do_ref_mod <- do %>%
-  mutate(hgc12 = ifelse(hgc_i < 9, "8TH", "12TH")) %>%
+do_ref_mod <- do_sw_join %>%
+  mutate(hgc12 = ifelse(high_grade < 9, "8TH", "12TH")) %>%
   ggplot(aes(x = exp,
              y = lnwage,
              linetype = hgc12)) +
@@ -198,7 +211,7 @@ ggplot(do, aes(x = yr_wforce,
 
 # ---- compare-xp-sw
 
[…]

### 2022-02-17 data-raw/ida.R

@@ -58,6 +58,7 @@ kable(as.array(summary(wages_before$mean_hourly_wage)),
   kable_styling()
 
 ## ---- sample-plot
+# take 20 sample of individuals to explore wages by ID
 wages_before_tsibble <- as_tsibble(x = wages_before,
                                      key = id,
                                      index = year,
@@ -80,13 +81,16 @@ ggplot(wages_before_tsibble,
 
 
 ## ---- feature-plot
-wages_high <- filter(wages_before, mean_hourly_wage > 500) %>%
-  as_tibble() %>%
-  head(n = 6)
 
-wages_high2 <- wages_before %>%
-  filter(id %in% wages_high$id)
+# create the summary plots of all of the individuals
+# wages_high <- filter(wages_before, mean_hourly_wage > 500) %>%
+#   as_tibble() %>%
+#   head(n = 6)
+#
+# wages_high2 <- wages_before %>%
+#   filter(id %in% wages_high$id)
 
+# the spaghetty plot of wages before imputation (raw values from database)
 spag <- wages_before %>%
   ggplot(aes(x = year,
              y = mean_hourly_wage,
@@ -101,6 +105,8 @@ spag <- wages_before %>%
   ylab("Hourly wage ($)") #+
   #theme(plot.title = element_text(size = 10))
 
+
+# three feature of wages data (mas, min, median)
 wages_three_feat <- wages_before_tsibble %>%
   features(mean_hourly_wage,
            feat_three_num
@@ -129,6 +135,8 @@ feature_bp <- ggplot(wages_feat_long,
   theme(legend.position = "none") +
   scale_color_viridis_d()
 
+# take one ID (id == 39) to illustrate that some IDs
+# only has one extremely high of wages (inconsistence with other wages )
 plot_high <- ggplot(filter(wages_high2, id == 39)) +
   geom_line(aes(x = year,
                 y = mean_hourly_wage)) +
@@ -154,6 +162,7 @@ spag + feature_bp + plot_high +
 ## ---- compare-data
 set.seed(31251587)
 
+# reading the cleaned (imputed wages)
 wages_cleaned <- readRDS(here::here("paper/results/wages_after.rds"))
 sample_id <- sample(unique(wages_cleaned$id), 20)
 sample <- subset(wages_cleaned, id %in% sample_id)
@@ -276,23 +285,10 @@ wages <- wages_clean %>%
              index = year,
              regular = FALSE)
 
-# Create a data set for demographic variables
[…]

### 2022-02-17 paper/paper.tex

@@
 Secondly, in the original dataset, wages were inflation-adjusted to 1990 prices.
-This is not done for the refreshed data because we plan to keep refreshing it as the data from the newest survey period release.
-Thus, the 1990 prices might not be relevant anymore for the future context.
-Hence, we keep it as the original value to allow the users to treat it as their needs.
-
-It is also worth noting that the treatment of unlikely wages differ in the refreshed data.
+This is not done for the refreshed data because we plan to keep refreshing it as the data added to and released from the survey.
+Instead we have provided a function in the R package, \texttt{yowie}, for users to conduct the inflation adjustment when they are ready to analyse the data.
+
+It is important to note that the treatment of unlikely wages differs in the refreshed data.
 In the original data by Singer and Willett (2003), wages greater than \$75 are set to be missing.
-However, in recent context, this value is too low to be set as the maximum threshold.
-We opted to use the weights from a robust linear regression to determine what should be regarded as outliers and imputed them with their predicted values as as described in Section \ref{censor}.
+However, this value is too low to be set as the maximum threshold, and it doesn't take into account temporal neighbors for an individual.
+We opted to use the weights from a robust linear regression to determine what should be regarded as extreme and imputed them with their predicted values as as described in Section \ref{censor}.
 

### 2022-02-17 paper/paper.Rmd
density plot instead of a scatterplot to compare experiences

@@
 
-```{r compare-xp-sw, fig.cap = "Males dropouts' work experience in the original and refreshed data.
-Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
-We can see that the correlation is reasonably high between them even there are different values for many individuals.", out.width = "70%"}
-```
-
 <!--AE comment: first sentence can be removed or rephrased.
@@
 
+```{r compare-xp-sw, fig.cap = "Density plot comparing the calculated work experience in the refreshed data, up to 1994, with that of the original data for male dropouts.
+Experience is shown on a square root scale to reduce the skewness.
+The highest concentration of observations is along the x=y axis, which means that most observations at most time points are similar.", out.width = "70%"}
+```
+
 ```{r plotting-sw-do, fig.cap = "Comparison of original textbook example (A) with refreshed data (B).
@@
 There are two aspects of the original data that make the replication with exactly the same result is not obtained.
-The first one is that the calculation of the time variable provided, which is the "work experience", is not clearly articulated.
+The first is that the calculation of the work experience is not clearly articulated.
 @SingerJudithD2003Alda mention that the temporal variable is the years of experience since the first day of work.
@@
 One of them is the weeks of worked since the last interview.
-Hence, we use this to calculate the variable.
-Even though we got different result, the correlation between the original and refreshed work experience is reasonably high as shown in Figure \ref{fig:compare-xp-sw}.
+This is used to calculate the variable.
+It produces similar results, as shown in Figure \ref{fig:compare-xp-sw}.
 

### 2022-02-17 paper/paper.tex
density plot instead of a scatterplot to compare experiences

@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.7\linewidth]{figures/compare-xp-sw-1}
-
-}
-
-\caption{Males dropouts' work experience in the original and refreshed data.
-Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
-We can see that the correlation is reasonably high between them even there are different values for many individuals.}\label{fig:compare-xp-sw}
-\end{figure}
-
 The original set contains wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994.
@@
 
+{\centering \includegraphics[width=0.7\linewidth]{figures/compare-xp-sw-1}
+
+}
+
+\caption{Density plot comparing the calculated work experience in the refreshed data, up to 1994, with that of the original data for male dropouts.
+Experience is shown on a square root scale to reduce the skewness.
+The highest concentration of observations is along the x=y axis, which means that most observations at most time points are similar.}\label{fig:compare-xp-sw}
+\end{figure}
+
+\begin{figure}
+
 {\centering \includegraphics[width=1\linewidth]{figures/plotting-sw-do-1}
@@
 There are two aspects of the original data that make the replication with exactly the same result is not obtained.
-The first one is that the calculation of the time variable provided, which is the ``work experience'', is not clearly articulated.
+The first is that the calculation of the work experience is not clearly articulated.
 Singer and Willett (2003) mention that the temporal variable is the years of experience since the first day of work.

### 2022-02-17 data-raw/eda.R

@@ -200,7 +200,7 @@ ggplot(do, aes(x = yr_wforce,
 
 do_sw_join <- left_join(sw, do, by = c("id", "index"))
 
-ggplot(do_sw_join, aes(x = xp,
+wp_compare1 <- ggplot(do_sw_join, aes(x = xp,
                y = exp)) +
   geom_point(alpha = 0.1) +
   geom_abline(intercept = 0, slope = 1) +
@@ -212,6 +212,19 @@ ggplot(do_sw_join, aes(x = xp,
   theme_bw() +
   theme(aspect.ratio=1)
 
+wp_compare2 <- ggplot(do_sw_join, aes(x = xp,
+                       y = exp)) +
+  geom_density2d_filled(contour_var = "ndensity") +
+  #geom_abline(intercept = 0, slope = 1) +
+  guides(color = "none") +
+  scale_x_sqrt(breaks = seq(0, 13, 1)) +
+  scale_y_sqrt(breaks = seq(0, 13, 1)) +
+  xlab("Work experience (Singer & Willet, 2008)") +
+  ylab("Work experience (refreshed)") +
+  theme_bw() +
+  theme(aspect.ratio=1, legend.position="none")
+
+wp_compare2
 
 # Takeaways:
 

### 2022-02-17 data-raw/eda.R
trying to compare id by id

@@ -112,14 +112,22 @@ do_id <- as_tibble(do) %>%
   count() %>%
   select(id)
 
+# Restrict do to 1994 and before
+do_1994 <- do %>%
+  filter(year <= 1994) %>%
+  filter(!is.na(exp)) %>%
+  select(id, year, wage, age_1979, gender, exp)
+duplicates(do_1994, key=id, index=exp)
+
 # check common id in data set
 sw_do <- inner_join(sw_id, do_id, by = "id")
 
-# only have 94 id that agree upon each other
 sw_agree <- filter(sw, id %in% sw_do$id)
-do_agree <- filter(do, id %in% sw_do$id)
+do_agree <- filter(do, id %in% sw_do$id) # 839 do_agree %>% count(id)
 
-sw_do_sample <- sample(sw_do$id, 12)
+sw_do_sample <- sw_agree %>%
+  mutate(id = as.integer(as.character(id))) %>%
+  sample_n_keys(size = 12)
 
 sw_agree_sample <- sw_agree %>%
   filter(id %in% sw_do_sample) %>%

### 2022-02-17 paper/paper.tex
working response

@@
 
+\begin{verbatim}
+#> # A tibble: 43 x 6
+#>    id     year  wage age_1979 gender   exp
+#>    <fct> <int> <dbl>    <int> <fct>  <dbl>
+#>  1 1663   1987  5.90       15 MALE    4.26
+#>  2 1663   1988  7          15 MALE    4.26
+#>  3 1663   1989 14          15 MALE    4.26
+#>  4 2466   1988  6.97       17 MALE    8.17
+#>  5 2466   1989  6.97       17 MALE    8.17
+#>  6 2466   1990  7.25       17 MALE    9.03
+#>  7 2466   1991  7.25       17 MALE    9.03
+#>  8 2594   1986 11          17 MALE    5.95
+#>  9 2594   1987  9.68       17 MALE    5.95
+#> 10 3087   1993 18.7        18 MALE   10.7 
+#> # ... with 33 more rows
+\end{verbatim}
+
 The original set contains wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994.

### 2022-02-17 response/response1.Rmd
working response

@@
 
-* **NEED HELP**
+* We have changed Fig 2 to show 36 individuals, that display the range of wages experience, including some with questionable values.
+The figure caption reflects this and more details are provide din the text.
 
@@
 
-* **NEED HELP**
+* This section has been re-written, and additional details provided.
 
@@
 
-* **NEED HELP**
+* Done
 
@@
 
-* **NEED HELP**
+* The abstract has been re-written to reflect the re-framing of the paper.
 
@@
 
-* **NEED HELP**
+* Paper has been re-written to change focus.
 
@@
 
-* **NEED HELP**
+* See explanation above
 
@@
 
-* We use `MASS` package instead of `robustlmm` package.
-We have tried to use `robustlmm`, but perhaps because the data is large and imbalanced, the model was not converging.
-Also, for reproducibility reason, we made it more affordable (in terms of memory and time), especially when people want to perform the stages in cleaning the data in class/teaching.
+* The `robustlmm` model would not converge. We have added some text to the paper on this.
 
@@
 
-* **NEED HELP**
+* **NEED HELP** We have added more explanation on the process used to decide on the threshold.
 
@@
 
-* **NEED HELP**
+* **NEED HELP** Figure 8 has been updated, and other comparisons of the original and refreshed data have been added.
+There are still some differences, but with the change in variables being created and better filtering to get the dropouts the data are now more similar.
 
@@
 
-* **NEED HELP**
+* NEED HELP We are really not sue what this refers to.

### 2022-02-17 data-raw/ida.R

@@ -85,12 +85,12 @@ wages_before_tsibble %>%
 ## ---- feature-plot
 
 # create the summary plots of all of the individuals
-# wages_high <- filter(wages_before, mean_hourly_wage > 500) %>%
-#   as_tibble() %>%
-#   head(n = 6)
-#
-# wages_high2 <- wages_before %>%
-#   filter(id %in% wages_high$id)
+wages_high <- filter(wages_before, mean_hourly_wage > 500) %>%
+   as_tibble() %>%
+   head(n = 6)
+
+ wages_high2 <- wages_before %>%
+   filter(id %in% wages_high$id)
 
 # the spaghetty plot of wages before imputation (raw values from database)
 spag <- wages_before %>%

### 2022-02-17 paper/paper.Rmd
done with response for now

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by @SingerJudithD2003Alda in their book, Applied longitudinal data analysis: Modeling change and event occurrence,  which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
+An example of this, is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis: Modeling change and event occurrence",  which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
@@
 Be explicit about this.-->
-Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences.
+Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences.
 Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda.
 We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the *purposes of studying racism, not race* [@racismnotrace].
+There are a number of data sets provided by @SingerJudithD2003Alda and we focus only on this one because it has captivated our attention for a number of years.
+We use it in our own teaching of longitudinal data analysis and would very much like to use data since 1994.
 <!--However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.-->
@@
 
-For gender and race, we only rename the column names.
-It is worth noting that using these two variable needs special attention.
-Gender, as reported in the data, only has two categories, which is recognised today as inadequate.
-Gender is not binary.
+It is important to discuss the two variables gender and race.
+Gender comes from the variable called sex in the database, but the reports on the data on the web site typically use gender.
+Using these two variables should be accompanied with special attention.
+Gender, as provided in the data, if self-reported.
+It only has two categories, which is recognised today as inadequate.
+Gender is not binary, and this needs to be emphasized in teaching.
 Further, race is as reported in the database.

### 2022-02-17 paper/paper.tex
done with response for now

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) in their book, Applied longitudinal data analysis: Modeling change and event occurrence, which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
+An example of this, is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis: Modeling change and event occurrence'', which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 2001).
-Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
+Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
 Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).
 We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the \emph{purposes of studying racism, not race} (Fullilove 1998).
+There are a number of data sets provided by Singer and Willett (2003) and we focus only on this one because it has captivated our attention for a number of years.
+We use it in our own teaching of longitudinal data analysis and would very much like to use data since 1994.
 

### 2022-02-17 response/response1.Rmd
done with response for now
updated response

@@
 \begin{description}
+\item[Reviewer 3:]\textcolor{violet}{The choice of 0.12 seems arbitrary at best.
+I don’t see any justification for this approach (e.g., literature, simulations, etc).
+Why not use bi-square, instead of huber, and simply impute for observations with weight 0.
+While still arbitrary, it’s decided by the bisquare model and isn’t a guesstimate.
+Further, it would be sensible for any dataset and not just this one.
+(This would correspond to a smaller threshold, which may catch too many strange outcomes).
+In any case, it’s hard to justify something so adhoc.}
+\end{description}
+
+* Similar comment to this is addressed above.
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{You’re using iteratively reweighted least squares with Huber weighting.} 
+\end{description}
+
+* Thank you, we have changed from "M estimation" to "iteratively reweighted least squares".
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{I’m not sure this is accurate.
+The robustlmm package in R fits these models rather quickly with thousands of observations.
+What are the authors using?
+Does it run out of memory (requires more RAM) or does it take too much time?}
+\end{description}
+
+* The `robustlmm` model would not converge. We have added some text to the paper on this.
+
+\begin{description}
 \item[Reviewer 2:]\textcolor{violet}{What does “It implies that the fluctuation can still be observed in the data after the treatment.” mean?
@@
 
-* **NEED HELP**
+* **NEED HELP** The paper has been fixed so that male/female is used consistently. We have also rewritten the paragraph to focus on the survey questions and shortcomings.
 
@@
 \begin{description}
-\item[Reviewer 3:]\textcolor{violet}{You’re using iteratively reweighted least squares with Huber weighting.} 
-\end{description}
-
-* **NEED HELP**
-
-\begin{description}
-\item[Reviewer 3:]\textcolor{violet}{I’m not sure this is accurate.
-The robustlmm package in R fits these models rather quickly with thousands of observations.
-What are the authors using?
-Does it run out of memory (requires more RAM) or does it take too much time?}
-\end{description}
-
-* The `robustlmm` model would not converge. We have added some text to the paper on this.
-
-\begin{description}
-\item[Reviewer 3:]\textcolor{violet}{The choice of 0.12 seems arbitrary at best.
-I don’t see any justification for this approach (e.g., literature, simulations, etc).
-Why not use bi-square, instead of huber, and simply impute for observations with weight 0.
-While still arbitrary, it’s decided by the bisquare model and isn’t a guesstimate.
-Further, it would be sensible for any dataset and not just this one.
-(This would correspond to a smaller threshold, which may catch too many strange outcomes).
-In any case, it’s hard to justify something so adhoc.}
-\end{description}
-
-* **NEED HELP** We have added more explanation on the process used to decide on the threshold.
-
-\begin{description}
 \item[Reviewer 3:]\textcolor{violet}{It’s unclear about what fluctuation or treatment the author is referring to.}
@@
 
-* NEED HELP We are really not sue what this refers to.
+* NEED HELP We are really not sure what this refers to.

### 2022-02-17 paper/paper.Rmd
updated according to reviewer comment

@@
 The `mean_hourly_wage` and `year` are set as the dependent and predictor, respectively.
-Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm].
+Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm].
 <!--AE comment: “However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.” I don’t think this sentence needs a “However” and probably should say “past which” rather than “in which”-->The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.

### 2022-02-17 paper/paper.Rmd
removed reference to book as already referenced in the intro

@@
 
-The book "Applied longitudinal data analysis" [@SingerJudithD2003Alda] used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
+@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
 This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN].

### 2022-02-17 paper/paper.Rmd
updated the takeaways

@@
 Since it is only measured since the last interview, this needs to be cumulated for each survey year.
-This produced results more similar to the original data (see Subsection \@ref(takeaways)).
+This produced results more similar to the original data (see Section \@ref(takeaways)).
 
@@
 
-There are two aspects of the original data that make the replication with exactly the same result is not obtained.
+There are two aspects of the original data that we cannot replicate.
 The first is that the calculation of the work experience is not clearly articulated.
@@
 This is used to calculate the variable.
-It produces similar results, as shown in Figure \ref{fig:compare-xp-sw}.
+It produces a highly correlated results as shown in Figure \ref{fig:compare-xp-sw}.
 
@@
 This is not done for the refreshed data because we plan to keep refreshing it as the data added to and released from the survey.
-Instead we have provided a function in the R package, `yowie`, for users to conduct the inflation adjustment when they are ready to analyse the data.
+Instead we have provided a function in the R package, `r yowie`, for users to conduct the inflation adjustment when they are ready to analyse the data.
 

### 2022-02-17 paper/paper.Rmd
removed double titling of Singer book

--- 2022-02-17 paper/paper.Rmd
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis: Modeling change and event occurrence",  which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
+An example of this, is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis",  which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
@@
 
-@SingerJudithD2003Alda used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
+@SingerJudithD2003Alda's book used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
 This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century [@MichaelRPergamit2001DWTN].

### 2022-02-17 paper/paper.tex
removed double titling of Singer book

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis: Modeling change and event occurrence'', which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
+An example of this, is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis'', which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 
-The book ``Applied longitudinal data analysis'' (Singer and Willett 2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
+Singer and Willett (2003)'s book used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
 This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al.
@@
 The \texttt{mean\_hourly\_wage} and \texttt{year} are set as the dependent and predictor, respectively.
-Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021).
+Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021).
 The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.

### 2022-02-17 paper/paper.Rmd

--- 2022-02-17 paper/paper.Rmd
@@
 Be explicit about this.-->
+
 Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences.

### 2022-02-17 paper/paper.tex

@@
 
-Singer and Willett (2003)'s book used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
+Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
 This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al.
@@
 2001).
+
 Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
@@
 Since it is only measured since the last interview, this needs to be cumulated for each survey year.
-This produced results more similar to the original data (see Subsection \ref{takeaways}).
+This produced results more similar to the original data (see Section \ref{takeaways}).
 
@@
 This is not done for the refreshed data because we plan to keep refreshing it as the data added to and released from the survey.
-Instead we have provided a function in the R package, \texttt{yowie}, for users to conduct the inflation adjustment when they are ready to analyse the data.
+Instead we have provided a function in the R package, \texttt{[CENSORED]}, for users to conduct the inflation adjustment when they are ready to analyse the data.
 

### 2022-02-17 data-raw/eda.R
comparison of two sets

@@ -10,6 +10,34 @@ sw <- brolgar::wages %>%
 hgc_1979 <- hgc_1979 %>%
   mutate(id = as.factor(id))
 
+# XXX Re-do of the comparison section
+# tabular hgc
+sw %>%
+  as_tibble() %>%
+  select(id, high_grade) %>%
+  distinct() %>%
+  count(high_grade)
+
+do_1994 <- wages_hs_do %>%
+  filter(year <= 1994) %>%
+  as_tibble() %>%
+  group_by(id) %>%
+  mutate(hgc_1994 = max(grade[between(year, 1979, 1994)], na.rm=TRUE)) %>%
+  ungroup()
+
+do_1994 %>%
+  select(id, hgc_1994) %>%
+  filter(between(hgc_1994, 6, 12)) %>%
+  distinct() %>%
+  count(hgc_1994)
+
+exp_p1 <- sw %>% ggplot(aes(x=xp)) +
+  geom_density(fill="black", alpha=0.7) +
+  xlim(c(0, 13))
+exp_p2 <- do_1994 %>% ggplot(aes(x=exp)) +
+  geom_density(fill="black", alpha=0.7) +
+  xlim(c(0, 13))
+
 # load the refreshed data, join it with the hgc_1979
 do <- wages_hs_do %>%
   as_tibble() %>%
@@ -19,7 +47,6 @@ do <- wages_hs_do %>%
          index = 1:n())
 
 # join original and refreshed data, only for the same individuals that appears in the two data
-
 do_sw_join <- inner_join(sw, do, by = c("id", "index"))
 
 sw_wages <- sw %>%

### 2022-02-17 paper/paper.tex

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis'', which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
+An example of this, is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis,'' which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 \begin{verbatim}
-#> 'data.frame':    12686 obs. of  742 variables:
+#> 'data.frame':    12686 obs. of  663 variables:
 #>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 ...
@@
 
-Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide, suggests that it should be calculated based on the variable ``number of weeks worked since last interview''.
-This would remove periods of unemployment which makes sense when measuring experience while actually working.
+Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide, suggests that it should be calculated based on the variable ``number of weeks worked since last interview.'' This would remove periods of unemployment which makes sense when measuring experience while actually working.
 Since it is only measured since the last interview, this needs to be cumulated for each survey year.
-This produced results more similar to the original data (see Section \ref{takeaways}).
+This produced results more similar to the original data (as discussed futher in Section \ref{takeaways}).
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/cookd/numbats/yowie-paper/paper/figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{/Users/emi/Dropbox/repo/yowie-paper/paper/figures/flowchart_blind}
 

### 2022-02-17 data-raw/eda.R
comparing data sets

@@ -31,12 +31,34 @@ do_1994 %>%
   distinct() %>%
   count(hgc_1994)
 
+# compare distributions of exp
 exp_p1 <- sw %>% ggplot(aes(x=xp)) +
   geom_density(fill="black", alpha=0.7) +
   xlim(c(0, 13))
 exp_p2 <- do_1994 %>% ggplot(aes(x=exp)) +
   geom_density(fill="black", alpha=0.7) +
   xlim(c(0, 13))
+
+# tabulate race
+sw %>%
+  as_tibble() %>%
+  select(id, black, hispanic) %>%
+  distinct() %>%
+  count(black, hispanic)
+
+do_1994 %>%
+  select(id, race) %>%
+  distinct() %>%
+  count(race)
+
+# compare density of wages
+wg_p1 <- sw %>% ggplot(aes(x=ln_wages)) +
+  geom_density(fill="black", alpha=0.7) +
+  xlim(c(0.5, 4.5))
+wg_p2 <- do_1994 %>% ggplot(aes(x=log(wage))) +
+  geom_density(fill="black", alpha=0.7) +
+  xlim(c(0.5, 4.5))
+
 
 # load the refreshed data, join it with the hgc_1979
 do <- wages_hs_do %>%

### 2022-02-18 paper/paper.Rmd
minor update
Add the paragraph discussing gender and sex.

@@
 
-It is important to discuss the two variables gender and race.
-Gender comes from the variable called sex in the database, but the reports on the data on the web site typically use gender.
-Using these two variables should be accompanied with special attention.
-Gender, as provided in the data, if self-reported.
-It only has two categories, which is recognised today as inadequate.
-Gender is not binary, and this needs to be emphasized in teaching.
-Further, race is as reported in the database.
-When doing analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
+It is worth noting that `gender` comes from a variable called `sex` in the database.
+We use the term `gender` as the reports on the data on the website typically use `gender`.
+Gender, as provided in the data, is self-reported and only has two categories, "male" and "female".
+The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
+Measuring gender as a binary variable has the potential to fail to capture people who do not identify themself as either male or female or people whose gender does not align with their sex classification [@kennedy2020using].
+From a statistical perspective, this sometimes difficult the adjustment of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex [@kennedy2020using].
+Hence, nowadays, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
+This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
+Similar case to race, as reported in the database.
+When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
 
@@
 Since it is only measured since the last interview, this needs to be cumulated for each survey year.
-This produced results more similar to the original data (see Section \@ref(takeaways)).
+This produced results more similar to the original data (as discussed futher in Section \@ref(takeaways)).
 

### 2022-02-18 paper/paper.tex
Added the reference related to gender and sex.

--- 2022-02-17 paper/paper.tex
@@
 
-Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al.
+Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
 2019).
@@
 As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year.
-The raw data consequently has a large number of columns (742 to be specific).
+The raw data consequently has a large number of columns (770 to be specific).
 The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars.
@@
 \begin{verbatim}
-#> 'data.frame':    12686 obs. of  742 variables:
+#> 'data.frame':    12686 obs. of  770 variables:
 #>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 ...
@@
 
-It is important to discuss the two variables gender and race.
-Gender comes from the variable called sex in the database, but the reports on the data on the web site typically use gender.
-Using these two variables should be accompanied with special attention.
-Gender, as provided in the data, if self-reported.
-It only has two categories, which is recognised today as inadequate.
-Gender is not binary, and this needs to be emphasized in teaching.
-Further, race is as reported in the database.
-When doing analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
+It is worth noting that \texttt{gender} comes from a variable called \texttt{sex} in the database.
+We use the term \texttt{gender} as the reports on the data on the website typically use \texttt{gender}.
+Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female''.
+The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
+Measuring gender as a binary variable has the potential to fail to capture people who do not identify themself as either male or female or people whose gender does not align with their sex classification ({\textbf{???}}).
+From a statistical perspective, this sometimes difficult the adjustment of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex ({\textbf{???}}).
+Hence, nowadays, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
+This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
+Similar case to race, as reported in the database.
+When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
 
@@
 Since it is only measured since the last interview, this needs to be cumulated for each survey year.
-This produced results more similar to the original data (see Section \ref{takeaways}).
+This produced results more similar to the original data (as discussed futher in Section \ref{takeaways}).
 
@@
 \section{Comparison of refreshed with the original data}\label{compare}}
+
+\begin{verbatim}
+#> # A tibble: 7 x 2
+#>   high_grade     n
+#>        <int> <int>
+#> 1          6    32
+#> 2          7    95
+#> 3          8   239
+#> 4          9   211
+#> 5         10   186
+#> 6         11   110
+#> 7         12    15
+\end{verbatim}
+
+\begin{verbatim}
+#> # A tibble: 7 x 2
+#>   hgc_1994     n
+#>      <int> <int>
+#> 1        6    10
+#> 2        7    16
+#> 3        8    92
+#> 4        9   124
+#> 5       10   112
+#> 6       11   123
+#> 7       12   318
+\end{verbatim}
+
+\begin{verbatim}
+#> # A tibble: 3 x 3
+#>   black hispanic     n
+#>   <int>    <int> <int>
+#> 1     0        0   438
+#> 2     0        1   204
+#> 3     1        0   246
+\end{verbatim}
+
+\begin{verbatim}
+#> # A tibble: 3 x 2
+#>   race                        n
+#>   <fct>                   <int>
+#> 1 HISPANIC                  202
+#> 2 BLACK                     232
+#> 3 NON-BLACK, NON-HISPANIC   430
+\end{verbatim}
 

### 2022-02-18 paper/paper.Rmd
nice paragraph on gender and race

@@
 It is worth noting that `gender` comes from a variable called `sex` in the database.
-We use the term `gender` as the reports on the data on the website typically use `gender`.
+We use the term `gender` as the reports about the data on various pages in the website more commnly use `gender`.
 Gender, as provided in the data, is self-reported and only has two categories, "male" and "female".
 The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
-Measuring gender as a binary variable has the potential to fail to capture people who do not identify themself as either male or female or people whose gender does not align with their sex classification [@kennedy2020using].
+Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification [@kennedy2020using].
 From a statistical perspective, this sometimes difficult the adjustment of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex [@kennedy2020using].
-Hence, nowadays, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
+Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
 This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
-Similar case to race, as reported in the database.
+Similar this also relates to race, as reported in the database.
 When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.

### 2022-02-18 paper/paper.tex
nice paragraph on gender and race

--- 2022-02-17 paper/paper.tex
@@
 
-It is important to discuss the two variables gender and race.
-Gender comes from the variable called sex in the database, but the reports on the data on the web site typically use gender.
-Using these two variables should be accompanied with special attention.
-Gender, as provided in the data, if self-reported.
-It only has two categories, which is recognised today as inadequate.
-Gender is not binary, and this needs to be emphasized in teaching.
-Further, race is as reported in the database.
-When doing analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
+It is worth noting that \texttt{gender} comes from a variable called \texttt{sex} in the database.
+We use the term \texttt{gender} as the reports about the data on various pages in the website more commnly use \texttt{gender}.
+Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female''.
+The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
+Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification (Kennedy et al.
+2020).
+From a statistical perspective, this sometimes difficult the adjustment of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex (Kennedy et al.
+2020).
+Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
+This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
+Similar this also relates to race, as reported in the database.
+When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
 
@@
 \begin{verbatim}
+#> # A tibble: 7 x 2
+#>   high_grade     n
+#>        <int> <int>
+#> 1          6    32
+#> 2          7    95
+#> 3          8   239
+#> 4          9   211
+#> 5         10   186
+#> 6         11   110
+#> 7         12    15
+\end{verbatim}
+
+\begin{verbatim}
+#> # A tibble: 7 x 2
+#>   hgc_1994     n
+#>      <int> <int>
+#> 1        6    10
+#> 2        7    16
+#> 3        8    92
+#> 4        9   124
+#> 5       10   112
+#> 6       11   123
+#> 7       12   318
+\end{verbatim}
+
+\begin{verbatim}
+#> # A tibble: 3 x 3
+#>   black hispanic     n
+#>   <int>    <int> <int>
+#> 1     0        0   438
+#> 2     0        1   204
+#> 3     1        0   246
+\end{verbatim}
+
+\begin{verbatim}
+#> # A tibble: 3 x 2
+#>   race                        n
+#>   <fct>                   <int>
+#> 1 HISPANIC                  202
+#> 2 BLACK                     232
+#> 3 NON-BLACK, NON-HISPANIC   430
+\end{verbatim}
+
+\begin{verbatim}
 #> # A tibble: 43 x 6
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-kennedy2020using}{}}%
+Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. {``Using Sex and Gender in Survey Adjustment.''} \url{https://arxiv.org/abs/2009.14401}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%

### 2022-02-18 paper/paper.tex
removed miscellaneous results printing that was doing checks

@@
 
-\begin{verbatim}
-#> # A tibble: 7 x 2
-#>   high_grade     n
-#>        <int> <int>
-#> 1          6    32
-#> 2          7    95
-#> 3          8   239
-#> 4          9   211
-#> 5         10   186
-#> 6         11   110
-#> 7         12    15
-\end{verbatim}
-
-\begin{verbatim}
-#> # A tibble: 7 x 2
-#>   hgc_1994     n
-#>      <int> <int>
-#> 1        6    10
-#> 2        7    16
-#> 3        8    92
-#> 4        9   124
-#> 5       10   112
-#> 6       11   123
-#> 7       12   318
-\end{verbatim}
-
-\begin{verbatim}
-#> # A tibble: 3 x 3
-#>   black hispanic     n
-#>   <int>    <int> <int>
-#> 1     0        0   438
-#> 2     0        1   204
-#> 3     1        0   246
-\end{verbatim}
-
-\begin{verbatim}
-#> # A tibble: 3 x 2
-#>   race                        n
-#>   <fct>                   <int>
-#> 1 HISPANIC                  202
-#> 2 BLACK                     232
-#> 3 NON-BLACK, NON-HISPANIC   430
-\end{verbatim}
-
-\begin{verbatim}
-#> # A tibble: 43 x 6
-#>    id     year  wage age_1979 gender   exp
-#>    <fct> <int> <dbl>    <int> <fct>  <dbl>
-#>  1 1663   1987  5.90       15 MALE    4.26
-#>  2 1663   1988  7          15 MALE    4.26
-#>  3 1663   1989 14          15 MALE    4.26
-#>  4 2466   1988  6.97       17 MALE    8.17
-#>  5 2466   1989  6.97       17 MALE    8.17
-#>  6 2466   1990  7.25       17 MALE    9.03
-#>  7 2466   1991  7.25       17 MALE    9.03
-#>  8 2594   1986 11          17 MALE    5.95
-#>  9 2594   1987  9.68       17 MALE    5.95
-#> 10 3087   1993 18.7        18 MALE   10.7 
-#> # ... with 33 more rows
-\end{verbatim}
-
 The original set contains wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994.

### 2022-02-18 paper/paper.Rmd
opening paragraph of section 4

@@
 -->
-The original set contains wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994.
-The refreshed data set for comparison is also wages on high school dropouts from 1979 to 2018.
-The original set is available in the R package `brolgar` and the refreshed data is available in the R package `r yowie`.
+The original set, containing wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994, is available in the R package `brolgar`.
+To compare the refreshed data with the original a subset needs to be matched.
+There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data, and restricting the longitudinal measurements to the same years.
+However, we decided to try to replicate the process, as suggested by the description of the original data.
+This would require first identifying individuals who dropped out of high school.
 
@@
 ```
+
+## Summaries of original with refreshed dropouts data
 

### 2022-02-18 paper/paper.tex
opening paragraph of section 4

@@
 
-The original set contains wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994.
-The refreshed data set for comparison is also wages on high school dropouts from 1979 to 2018.
-The original set is available in the R package \texttt{brolgar} and the refreshed data is available in the R package \texttt{[CENSORED]}.
+The original set, containing wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994, is available in the R package \texttt{brolgar}.
+To compare the refreshed data with the original a subset needs to be matched.
+There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data, and restricting the longitudinal measurements to the same years.
+However, we decided to try to replicate the process, as suggested by the description of the original data.
+This would require first identifying individuals who dropped out of high school.
 
@@
 \end{figure}
+
+\hypertarget{summaries-of-original-with-refreshed-dropouts-data}{%
+\subsection{Summaries of original with refreshed dropouts data}\label{summaries-of-original-with-refreshed-dropouts-data}}
 

### 2022-02-18 paper/paper.Rmd
framing the comparison

@@
 Experience is shown on a square root scale to reduce the skewness.
-The highest concentration of observations is along the x=y axis, which means that most observations at most time points are similar.", out.width = "70%"}
+The highest concentration of observations is along the x=y axis, which means that most observations at most time points are similar.", out.width = "70%", eval=FALSE}
 ```
@@
 ```{r plotting-sw-do, fig.cap = "Comparison of original textbook example (A) with refreshed data (B).
-The original data was inflation-adjusted to 1990 prices and the individual's time of collection was converted to a length of experience in the workforce, which makes it difficult to precisely compare the two sets.", fig.width = 9, fig.height = 5, out.width = "100%", cache = FALSE}
+The original data was inflation-adjusted to 1990 prices and the individual's time of collection was converted to a length of experience in the workforce, which makes it difficult to precisely compare the two sets.", fig.width = 9, fig.height = 5, out.width = "100%", cache = FALSE, eval=FALSE}
 ```
@@
 ## Summaries of original with refreshed dropouts data
+
+Because the original data does not have the year of collection, it is not possible to directly merge the two subsets.
+Merging longitudinal data requires both the key (`id`) and the index (ideally survey year).
+In the original data the experience variable is the time index, and it was not possible to exactly match this for the refreshed data.
+Thus, comparisons of the two sets has to be conducted in a two sample fashion rather than as a matched sample.
 

### 2022-02-18 paper/paper.tex
framing the comparison

@@
 
-\begin{figure}
-
-{\centering \includegraphics[width=0.7\linewidth]{figures/compare-xp-sw-1}
-
-}
-
-\caption{Density plot comparing the calculated work experience in the refreshed data, up to 1994, with that of the original data for male dropouts.
-Experience is shown on a square root scale to reduce the skewness.
-The highest concentration of observations is along the x=y axis, which means that most observations at most time points are similar.}\label{fig:compare-xp-sw}
-\end{figure}
-
-\begin{figure}
-
-{\centering \includegraphics[width=1\linewidth]{figures/plotting-sw-do-1}
-
-}
-
-\caption{Comparison of original textbook example (A) with refreshed data (B).
-The original data was inflation-adjusted to 1990 prices and the individual's time of collection was converted to a length of experience in the workforce, which makes it difficult to precisely compare the two sets.}\label{fig:plotting-sw-do}
-\end{figure}
-
 \hypertarget{summaries-of-original-with-refreshed-dropouts-data}{%
 \subsection{Summaries of original with refreshed dropouts data}\label{summaries-of-original-with-refreshed-dropouts-data}}
+
+Because the original data does not have the year of collection, it is not possible to directly merge the two subsets.
+Merging longitudinal data requires both the key (\texttt{id}) and the index (ideally survey year).
+In the original data the experience variable is the time index, and it was not possible to exactly match this for the refreshed data.
+Thus, comparisons of the two sets has to be conducted in a two sample fashion rather than as a matched sample.
 

### 2022-02-18 paper/paper.Rmd
Added list of variables in Recap section.

--- 2022-02-18 paper/paper.Rmd
@@
 
+For `gender` and `race`, we only rename the column names provided in `new_data_qnames`.
 It is worth noting that `gender` comes from a variable called `sex` in the database.
-We use the term `gender` as the reports about the data on various pages in the website more commnly use `gender`.
+We use the term `gender` as the reports on the data on the website typically use `gender`.
 Gender, as provided in the data, is self-reported and only has two categories, "male" and "female".
 The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
-Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification [@kennedy2020using].
+Measuring gender as a binary variable has the potential to fail to capture people who do not identify themself as either male or female or people whose gender does not align with their sex classification [@kennedy2020using].
 From a statistical perspective, this sometimes difficult the adjustment of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex [@kennedy2020using].
-Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
+Hence, nowadays, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
 This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
-Similar this also relates to race, as reported in the database.
+Similar case to race, as reported in the database.
 When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
@@
 
+In addition, the list of variables that we cover in those three datasets are as follows:
+
+**`demog_nlsy79`** :
+
+  1. `id`: A unique individual's ID number.
+  2. `age_1979`: The age of the individual in 1979.
+  3. `gender`: Gender of the individual, FEMALE and MALE.
+  4. `race` : Race of the individual, NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
+  5. `hgc`: The highest grade completed.
+  6. `hgc_i`: Integer value of the highest grade completed.
+  7. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+
+**`wages`** and its subset for high school dropouts cohort **`wages_hs_do`**:
+
+  1. `id`: A unique individual's ID number. This is the `key` of the data as we saved the data as a `tsibble` object. 
+  2. `year`: The year the observation was taken. This is the `index` of the data.
+  3.
+`wage`: The mean of the hourly wages the individual gets at each of their different jobs.
+The value could be a weighted or an arithmetic mean.
+The weighted mean is used when the information of hours of work as the weight is available.
+The mean hourly wage could also be a predicted value if the original value is considered influential by the robust linear regression as part of data cleaning.
+  4. `age_1979`: The age of the individual in 1979.
+  5. `gender`: Gender of the individual: FEMALE and MALE.
+  6. `race`: Race where the individual belongs to: NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
+  7. `hgc`: The highest grade completed ever.
+  8. `hgc_i`: Integer value of the highest grade completed ever.
+  9. `grade`: Integer value of the highest grade completed corresponding to `year`. 
+  10. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+  11. `njobs`: Number of jobs that an individual has.
+  12. `hours:`: The total number of hours the individual usually works per week.
+  13. `stwork`: The year when the individual starting to work.
+  14. `yr_wforce`: The length of time in the workforce in years (`year` - `stwork`).
+  15. `exp`: Work experience, i.e., the number of years of working. 
+  16. `is_wm`: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
+  17. `is_pred`: Whether the mean hourly wage is a predicted value of RLM or not.
+
 # Comparison of refreshed with the original data {#compare}

### 2022-02-18 paper/paper.tex
Added list of variables in Recap section.

--- 2022-02-18 paper/paper.tex
@@
 
+For \texttt{gender} and \texttt{race}, we only rename the column names provided in \texttt{new\_data\_qnames}.
 It is worth noting that \texttt{gender} comes from a variable called \texttt{sex} in the database.
-We use the term \texttt{gender} as the reports about the data on various pages in the website more commnly use \texttt{gender}.
+We use the term \texttt{gender} as the reports on the data on the website typically use \texttt{gender}.
 Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female''.
 The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
-Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification (Kennedy et al.
+Measuring gender as a binary variable has the potential to fail to capture people who do not identify themself as either male or female or people whose gender does not align with their sex classification (Kennedy et al.
 2020).
@@
 2020).
-Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
+Hence, nowadays, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
 This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
-Similar this also relates to race, as reported in the database.
+Similar case to race, as reported in the database.
 When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
@@
 
+In addition here is the list of variables that we cover in those three datasets:
+
+\textbf{\texttt{demog\_nlsy79}}
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \texttt{id}: A unique individual's ID number.
+\item
+  \texttt{age\_1979}: The age of the individual in 1979.
+\item
+  \texttt{gender}: Gender of the individual, FEMALE and MALE.
+\item
+  \texttt{race} : Race of the individual, NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
+\item
+  \texttt{hgc}: The highest grade completed.
+\item
+  \texttt{hgc\_i}: Integer value of the highest grade completed.
+\item
+  \texttt{ged}: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+\end{enumerate}
+
+\textbf{\texttt{wages}} and its subset for high school dropouts cohort \textbf{\texttt{wages\_hs\_do}}:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \texttt{id}: A unique individual's ID number. This is the \texttt{key} of the data as we saved the data as a \texttt{tsibble} object.
+\item
+  \texttt{year}: The year the observation was taken. This is the \texttt{index} of the data.
+\item
+  \texttt{wage}: The mean of the hourly wages the individual gets at each of their different jobs.
+The value could be a weighted or an arithmetic mean.
+The weighted mean is used when the information of hours of work as the weight is available.
+The mean hourly wage could also be a predicted value if the original value is considered influential by the robust linear regression as part of data cleaning.
+\item
+  \texttt{age\_1979}: The age of the individual in 1979.
+\item
+  \texttt{gender}: Gender of the individual: FEMALE and MALE.
+\item
+  \texttt{race}: Race where the individual belongs to: NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
+\item
+  \texttt{hgc}: The highest grade completed ever.
+\item
+  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
+\item
+  \texttt{grade}: Integer value of the highest grade completed corresponding to \texttt{year}.
+\item
+  \texttt{ged}: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+\item
+  \texttt{njobs}: Number of jobs that an individual has.
+\item
+  \texttt{hours:}: The total number of hours the individual usually works per week.
+\item
+  \texttt{stwork}: The year when the individual starting to work.
+\item
+  \texttt{yr\_wforce}: The length of time in the workforce in years (\texttt{year} - \texttt{stwork}).
+\item
+  \texttt{exp}: Work experience, i.e., the number of years of working.
+\item
+  \texttt{is\_wm}: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
+\item
+  \texttt{is\_pred}: Whether the mean hourly wage is a predicted value of RLM or not.
+\end{enumerate}
+
 \hypertarget{compare}{%
 \section{Comparison of refreshed with the original data}\label{compare}}
+
+\begin{verbatim}
+#> # A tibble: 7 x 2
+#>   high_grade     n
+#>        <int> <int>
+#> 1          6    32
+#> 2          7    95
+#> 3          8   239
+#> 4          9   211
+#> 5         10   186
+#> 6         11   110
+#> 7         12    15
+\end{verbatim}
+
+\begin{verbatim}
+#> # A tibble: 7 x 2
+#>   hgc_1994     n
+#>      <int> <int>
+#> 1        6    10
+#> 2        7    16
+#> 3        8    92
+#> 4        9   124
+#> 5       10   112
+#> 6       11   123
+#> 7       12   318
+\end{verbatim}
+
+\begin{verbatim}
+#> # A tibble: 3 x 3
+#>   black hispanic     n
+#>   <int>    <int> <int>
+#> 1     0        0   438
+#> 2     0        1   204
+#> 3     1        0   246
+\end{verbatim}
+
+\begin{verbatim}
+#> # A tibble: 3 x 2
+#>   race                        n
+#>   <fct>                   <int>
+#> 1 HISPANIC                  202
+#> 2 BLACK                     232
+#> 3 NON-BLACK, NON-HISPANIC   430
+\end{verbatim}
+
+\begin{verbatim}
+#> # A tibble: 43 x 6
+#>    id     year  wage age_1979 gender   exp
+#>    <fct> <int> <dbl>    <int> <fct>  <dbl>
+#>  1 1663   1987  5.90       15 MALE    4.26
+#>  2 1663   1988  7          15 MALE    4.26
+#>  3 1663   1989 14          15 MALE    4.26
+#>  4 2466   1988  6.97       17 MALE    8.17
+#>  5 2466   1989  6.97       17 MALE    8.17
+#>  6 2466   1990  7.25       17 MALE    9.03
+#>  7 2466   1991  7.25       17 MALE    9.03
+#>  8 2594   1986 11          17 MALE    5.95
+#>  9 2594   1987  9.68       17 MALE    5.95
+#> 10 3087   1993 18.7        18 MALE   10.7 
+#> # ... with 33 more rows
+\end{verbatim}
 
@@
 Cham: Springer.
-https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-DasuTamraparni2003Edma}{}}%
+\url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.
+
+\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
 Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.
 
-\leavevmode\vadjust pre{\hypertarget{ref-janitor}{}}%
-Firke, Sam. 2020. \emph{{janitor: Simple Tools for Examining and Cleaning Dirty Data}}. \url{https://CRAN.R-project.org/package=janitor}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-racismnotrace}{}}%
-Fullilove, M. T. 1998. {``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.
-
-\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
+\leavevmode\hypertarget{ref-janitor}{}%
+Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.
+
+\leavevmode\hypertarget{ref-racismnotrace}{}%
+Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.
+
+\leavevmode\hypertarget{ref-grimshaw}{}%
 Grimshaw, Scott D.
 2015.
-{``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14.
+``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14.
 \url{https://doi.org/10.1080/00031305.2015.1081106}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-purrr}{}}%
-Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-penguins-data}{}}%
+\leavevmode\hypertarget{ref-purrr}{}%
+Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.
+
+\leavevmode\hypertarget{ref-penguins-data}{}%
 Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
+\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
 Huebner, Marianne, Werner Vach, and Saskia le Cessie.
 2016.
-{``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ilk2004}{}}%
+``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
+
+\leavevmode\hypertarget{ref-ilk2004}{}%
 Ilk, Ozlem.
 2004.
-{``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.''} PhD thesis, Iowa State University.
+``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.'' PhD thesis, Iowa State University.
 \url{https://doi.org/10.31274/rtd-180813-11012}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-kennedy2020using}{}}%
-Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. {``Using Sex and Gender in Survey Adjustment.''} \url{https://arxiv.org/abs/2009.14401}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
+\leavevmode\hypertarget{ref-kennedy2020using}{}%
+Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. ``Using Sex and Gender in Survey Adjustment.'' \url{http://arxiv.org/abs/2009.14401}.
+
+\leavevmode\hypertarget{ref-tamedata}{}%
 Kim, A.

### 2022-02-18 paper/paper.Rmd
comparison of subsets

--- 2022-02-18 paper/paper.Rmd
@@
 
+Figure \ref{fig:compare-subsets} contains summaries of corresponding variables in the two subsets.
+
+```{r compare-subsets, fig.cap = "Comparison of original textbook example with refreshed data: (A) highest grade completed, (B) experience and (C) log wages", fig.width = 9, fig.height = 7, out.width = "100%", cache = FALSE}
+```
+
 ## The takeaways {#takeaways}

### 2022-02-18 data-raw/eda.R

@@ -10,34 +10,71 @@ sw <- brolgar::wages %>%
 hgc_1979 <- hgc_1979 %>%
   mutate(id = as.factor(id))
 
-# XXX Re-do of the comparison section
-# tabular hgc
+# Comparing the original and refreshed data
+# tabulating hgc
 sw_hgc <- sw %>%
   as_tibble() %>%
   select(id, high_grade) %>%
   distinct() %>%
-  count(high_grade)
+  count(high_grade) %>%
+  rename(hgc = high_grade, original = n)
 
 do_1994 <- wages_hs_do %>%
   filter(year <= 1994) %>%
   as_tibble() %>%
   group_by(id) %>%
-  mutate(hgc_1994 = max(grade[between(year, 1979, 1994)], na.rm=TRUE)) %>%
+  mutate(hgc_1994 = max(grade[between(year, 1979, 1994)],
+                        na.rm=TRUE)) %>%
   ungroup()
 
 d94_hgc <- do_1994 %>%
   select(id, hgc_1994) %>%
   filter(between(hgc_1994, 6, 12)) %>%
   distinct() %>%
-  count(hgc_1994)
+  count(hgc_1994) %>%
+  rename(hgc = hgc_1994, refreshed = n)
+
+hgc_join <- left_join(sw_hgc, d94_hgc) %>%
+  mutate(original = -original, hgc = factor(hgc)) %>%
+  pivot_longer(cols = c(original, refreshed),
+               names_to = "subset",
+               values_to = "count")
+
+hgc_p <- ggplot(hgc_join, aes(x=hgc, y=count, fill=subset)) +
+  geom_col() +
+  facet_wrap(~subset, scales = "free_x") +
+  coord_flip() +
+  scale_y_continuous(expand = c(0, 0)) +
+  scale_fill_brewer("", palette="Dark2") +
+  ggtitle("(A)") +
+  theme_bw() +
+  theme(panel.spacing.x = unit(0, "mm"),
+        legend.position = "none")
 
 # compare distributions of exp
-exp_p1 <- sw %>% ggplot(aes(x=xp)) +
-  geom_density(fill="black", alpha=0.7) +
-  xlim(c(0, 13))
-exp_p2 <- do_1994 %>% ggplot(aes(x=exp)) +
-  geom_density(fill="black", alpha=0.7) +
-  xlim(c(0, 13))
[…]

### 2022-02-18 paper/paper.Rmd
fixed knit error and axes in hgc barcharts nice now

@@
 
-In addition, the list of variables that we cover in those three datasets are as follows:
+The list of variables provided in the three new datasets are as follows:
 
@@
 # Comparison of refreshed with the original data {#compare}
-
-```{r sw_wages}
-```
-
-```{r do_refreshed}
-```
-
-```{r compare-sw-do}
-```
 
@@
 
-```{r compare-xp-sw, fig.cap = "Density plot comparing the calculated work experience in the refreshed data, up to 1994, with that of the original data for male dropouts.
-Experience is shown on a square root scale to reduce the skewness.
-The highest concentration of observations is along the x=y axis, which means that most observations at most time points are similar.", out.width = "70%", eval=FALSE}
-```
-
-```{r plotting-sw-do, fig.cap = "Comparison of original textbook example (A) with refreshed data (B).
-The original data was inflation-adjusted to 1990 prices and the individual's time of collection was converted to a length of experience in the workforce, which makes it difficult to precisely compare the two sets.", fig.width = 9, fig.height = 5, out.width = "100%", cache = FALSE, eval=FALSE}
-```
-
 ## Summaries of original with refreshed dropouts data
@@
 Figure \ref{fig:compare-subsets} contains summaries of corresponding variables in the two subsets.
+
+```{r calculate-summaries-on-subsets}
+```
 
@@
 
-```{r compare-xp-to-yrworkforce, fig.cap = "The association between the lenght of time in the workforce and the work experience.
-Generally, the longer someone were in the workforce, the higher the work experience.
-However, some of the respondents have much lower working experience compared to their duration in the workforce.
-This is possible when someone ", eval=FALSE}
-```
-
 <!--AE comment: “The treatment of unlikely wages differ in the refreshed data.” I think this means “The treatment of anomalous wages differs between the original and refreshed data sets.” Then you say “We opted to use the weights from a robust linear regression to determine what should be set as missing value as described in Section 3.2.1.” I don’t think you were setting missing values though, were you?
@@
 What are plausible reasons for these differences?--> In the original data there were 366 individuals with up to 8th grade and 522 9th-12th grade, as compared to 967 with 12th grade in the refreshed data.
-
-```{r summaries, eval=FALSE}
-```
-
-<!-- Hence, in this section, we will compare how close the `wages_hs_do`, henceforth refreshed data, to the textbook data.
-It is also worth noting that the textbook data is reused from @MurnaneRichardJ1999DMDB’s paper.
-Hence, it is not directly extracted from the NLSY79 database.
-@SingerJudithD2003Alda do not state the time frame covered by the data.
-However, from @MurnaneRichardJ1999DMDB’s paper, we find that the textbook data cover the survey period from 1979-1994.
-Hence, for comparison purposes, we filter the refreshed data only until 1994.
-
-In the absence of highs chool dropouts criteria in the textbook data, we create a clearer criteria of high school dropouts as mentioned in Section \@ref(#target).
-Consequently, we do not have the same ID of high-school dropouts as in the textbook data.
-The textbook data has `r nrow(sw_id)` IDs included, while the refreshed data has `r nrow(do_id)` IDs.
-Furthermore, there are only `r nrow(sw_do)` IDs included both in the textbook and the refreshed data.
-After observing @MurnaneRichardJ1999DMDB’s paper, it is mentioned that the data is randomly drawn from the low income subsample of the NLSY.
-This is probably the reason why, even using some criteria, we cannot match the dropouts in the textbook and the refreshed data.
-
-Further, the temporal variable used in the textbook data is experience, instead of year.
-The only explanation we found regarding this variable is that it is derived from the years of IDs’ experience since their first job.
-@SingerJudithD2003Alda stated that this variable reflects the specific moment in each ID labor force history associated with each wage observation.
-Since it is not clearly articulated, we decide to keep the temporal variable in the refreshed data as year.
-Accordingly, in this comparison, we only compare the same ID but not the same temporal variable.
-
-Regarding the wages, the textbook data express the wages as inflation adjusted with 1990 dollars as constant [@SingerJudithD2003Alda].
-However, it is not stated how exactly the wages are adjusted.
-Note that in the refreshed data we publish in `yowie`, we do not adjust the data with inflation because the time frame covered ranged from 1979-2018.
-Hence, we allow flexibility to the data user to adjust, not adjust,  or adjust it with different base years.
-For comparison purposes, we adjust the wages with the average Consumer Price Index For All Urban Consumers (CPI-U) we got from @fed.
--->
-
-<!-- Figure \ref{fig:plotting-sw-do} shows the comparison between the textbook and the refreshed data.
-Even though it has different temporal variables and possibly different CPI indicators used in inflation adjustment, we can still capture some of the patterns in the refreshed data.
-
-From this comparison, we learn that we cannot make an apple to apple comparison and produce the exact same data as the textbook due to the absence of clearly articulated data extraction and derivation process.
-Hence, the takeaway is that it is imperative to disclose how the data derived from the initial source to make data analysis reproducible.
-This is why in this paper, we show the practice of reproducible and responsible data cleaning workflow to make sure the longitudinal data from a continuous survey can be refreshed.-->
 

### 2022-02-18 paper/paper.tex
comparison of subsets
fixed knit error and axes in hgc barcharts nice now

--- 2022-02-18 paper/paper.tex
@@
 
+The list of variables provided in the three new datasets are as follows:
+
+\textbf{\texttt{demog\_nlsy79}} :
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \texttt{id}: A unique individual's ID number.
+\item
+  \texttt{age\_1979}: The age of the individual in 1979.
+\item
+  \texttt{gender}: Gender of the individual, FEMALE and MALE.
+\item
+  \texttt{race} : Race of the individual, NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
+\item
+  \texttt{hgc}: The highest grade completed.
+\item
+  \texttt{hgc\_i}: Integer value of the highest grade completed.
+\item
+  \texttt{ged}: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+\end{enumerate}
+
+\textbf{\texttt{wages}} and its subset for high school dropouts cohort \textbf{\texttt{wages\_hs\_do}}:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
+  \texttt{id}: A unique individual's ID number. This is the \texttt{key} of the data as we saved the data as a \texttt{tsibble} object.
+\item
+  \texttt{year}: The year the observation was taken. This is the \texttt{index} of the data.
+\item
+  \texttt{wage}: The mean of the hourly wages the individual gets at each of their different jobs.
+The value could be a weighted or an arithmetic mean.
+The weighted mean is used when the information of hours of work as the weight is available.
+The mean hourly wage could also be a predicted value if the original value is considered influential by the robust linear regression as part of data cleaning.
+\item
+  \texttt{age\_1979}: The age of the individual in 1979.
+\item
+  \texttt{gender}: Gender of the individual: FEMALE and MALE.
+\item
+  \texttt{race}: Race where the individual belongs to: NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
+\item
+  \texttt{hgc}: The highest grade completed ever.
+\item
+  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
+\item
+  \texttt{grade}: Integer value of the highest grade completed corresponding to \texttt{year}.
+\item
+  \texttt{ged}: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+\item
+  \texttt{njobs}: Number of jobs that an individual has.
+\item
+  \texttt{hours:}: The total number of hours the individual usually works per week.
+\item
+  \texttt{stwork}: The year when the individual starting to work.
+\item
+  \texttt{yr\_wforce}: The length of time in the workforce in years (\texttt{year} - \texttt{stwork}).
+\item
+  \texttt{exp}: Work experience, i.e., the number of years of working.
+\item
+  \texttt{is\_wm}: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
+\item
+  \texttt{is\_pred}: Whether the mean hourly wage is a predicted value of RLM or not.
+\end{enumerate}
+
 \hypertarget{compare}{%
@@
 Thus, comparisons of the two sets has to be conducted in a two sample fashion rather than as a matched sample.
+
+Figure \ref{fig:compare-subsets} contains summaries of corresponding variables in the two subsets.
+
+\begin{figure}
+
+{\centering \includegraphics[width=1\linewidth]{figures/compare-subsets-1}
+
+}
+
+\caption{Comparison of original textbook example with refreshed data: (A) highest grade completed, (B) experience and (C) log wages}\label{fig:compare-subsets}
+\end{figure}
 

### 2022-02-18 data-raw/eda.R

@@ -10,6 +10,7 @@ sw <- brolgar::wages %>%
 hgc_1979 <- hgc_1979 %>%
   mutate(id = as.factor(id))
 
+## ---- calculate-summaries-on-subsets
 # Comparing the original and refreshed data
 # tabulating hgc
 sw_hgc <- sw %>%
@@ -43,10 +44,13 @@ hgc_join <- left_join(sw_hgc, d94_hgc) %>%
 hgc_p <- ggplot(hgc_join, aes(x=hgc, y=count, fill=subset)) +
   geom_col() +
   facet_wrap(~subset, scales = "free_x") +
-  coord_flip() +
-  scale_y_continuous(expand = c(0, 0)) +
+  scale_y_continuous("Count",
+                     breaks = c(-300, -200, -100, 0, 100, 200, 300),
+                     labels = c(300, 200, 100, 0, 100, 200, 300),
+                     expand = c(0, 0)) +
   scale_fill_brewer("", palette="Dark2") +
   ggtitle("(A)") +
+  coord_flip() +
   theme_bw() +
   theme(panel.spacing.x = unit(0, "mm"),
         legend.position = "none")
@@ -113,244 +117,9 @@ wg_p <- ggplot(wg_join) +
   theme_bw() +
   theme(legend.position = "none")
 
-## compare-subsets
+##  ---- compare-subsets
 hgc_p + exp_p + wg_p
 
-# load the refreshed data, join it with the hgc_1979
-do <- wages_hs_do %>%
-  as_tibble() %>%
-  left_join(hgc_1979, by = "id") %>%
-  group_by(id) %>%
-  mutate(lnwage = log(wage),
-         index = 1:n())
-
-# join original and refreshed data, only for the same individuals that appears in the two data
-do_sw_join <- inner_join(sw, do, by = c("id", "index"))
-
-sw_wages <- sw %>%
-  ggplot(aes(x = xp,
-             y = ln_wages)) +
-  geom_line(aes(group = id), alpha = 0.1) +
-  geom_smooth(se = FALSE, colour = "pink") +
-  labs(tag = "(A)", title = "Original") +
-  theme_bw() +
-  ylab("Hourly wage ($, natural log)") +
-  xlab("Experience (years)") +
-  #theme(plot.title = element_text(size = 10)) +
-  ylim(0, 5)
-
-sw_wages_mod <- do_sw_join %>%
-  as_tibble() %>%
[…]

### 2022-02-18 paper/paper.Rmd
caption for figure 7

@@
 
-```{r compare-subsets, fig.cap = "Comparison of original textbook example with refreshed data: (A) highest grade completed, (B) experience and (C) log wages", fig.width = 9, fig.height = 7, out.width = "100%", cache = FALSE}
+```{r compare-subsets, fig.cap = "Comparison of original and refreshed data: (A) highest grade completed, (B) experience and (C) log wages.
+Some difference in wages would be expected because the refreshed data is not inflation-adjusted, but the two sets are reasonably similar.", fig.width = 9, fig.height = 5, out.width = "100%", cache = FALSE}
 ```

### 2022-02-18 paper/paper.tex
caption for figure 7

@@
 
-\caption{Comparison of original textbook example with refreshed data: (A) highest grade completed, (B) experience and (C) log wages}\label{fig:compare-subsets}
+\caption{Comparison of original and refreshed data: (A) highest grade completed, (B) experience and (C) log wages.
+Some difference in wages would be expected because the refreshed data is not inflation-adjusted, but the two sets are reasonably similar.}\label{fig:compare-subsets}
 \end{figure}

### 2022-02-18 data-raw/eda.R

@@ -42,7 +42,7 @@ hgc_join <- left_join(sw_hgc, d94_hgc) %>%
                values_to = "count")
 
 hgc_p <- ggplot(hgc_join, aes(x=hgc, y=count, fill=subset)) +
-  geom_col() +
+  geom_col(alpha=0.9) +
   facet_wrap(~subset, scales = "free_x") +
   scale_y_continuous("Count",
                      breaks = c(-300, -200, -100, 0, 100, 200, 300),
@@ -70,7 +70,7 @@ exp_d94 <- do_1994 %>%
 exp_join <- bind_rows(exp_sw, exp_d94)
 
 exp_p <- ggplot(exp_join) +
-  geom_density(aes(x=exp, colour=subset, fill=subset)) +
+  geom_density(aes(x=exp, colour=subset, fill=subset), alpha=0.9) +
   facet_wrap(~subset, ncol=1) +
   xlim(c(0,13)) +
   xlab("Experience") + ylab("") +
@@ -107,7 +107,7 @@ wg_d94 <- do_1994 %>%
 wg_join <- bind_rows(wg_sw, wg_d94)
 
 wg_p <- ggplot(wg_join) +
-  geom_density(aes(x=ln_wages, colour=subset, fill=subset)) +
+  geom_density(aes(x=ln_wages, colour=subset, fill=subset), alpha=0.9) +
   facet_wrap(~subset, ncol=1) +
   xlim(c(0,4)) +
   xlab("Wages (natural log)") + ylab("") +

### 2022-02-18 paper/paper.Rmd
figure 7 text

@@
 Figure \ref{fig:compare-subsets} contains summaries of corresponding variables in the two subsets.
+Plot (A) shows a back-back bar chart of the highest grade completed.
+The two sets are almost the same, but small differences remain.
+Plots (B) and (C) show stacked density plots of experience and log wages, respectively.
+The distributions are relatively close, with more differences in wages as would be expected becuase the refreshed data is not inflation-adjusted.
 

### 2022-02-18 paper/paper.Rmd
takeways section revised

@@
   4. `race` : Race of the individual, NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
-  5. `hgc`: The highest grade completed.
-  6. `hgc_i`: Integer value of the highest grade completed.
+  5. `hgc`: The highest grade completed ever.
+  6. `hgc_i`: Integer value of the highest grade completed ever.
   7. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
@@
 However, we decided to try to replicate the process, as suggested by the description of the original data.
-This would require first identifying individuals who dropped out of high school.
+This requires first identifying individuals who dropped out of high school.
 
@@
 
-There are two aspects of the original data that we cannot replicate.
-The first is that the calculation of the work experience is not clearly articulated.
-@SingerJudithD2003Alda mention that the temporal variable is the years of experience since the first day of work.
-However, this exact same variable is not explicitly available in the database.
+There are several aspects of the original data that were difficult to replicate.
+Calculation of the work experience is not clearly articulated.
+@SingerJudithD2003Alda describe the temporal variable as the years of experience since the first day of work.
+This variable is not explicitly available in the database.
 From the NLSY79 topical guide [@nlsy79guide], we find that several variables are tagged as work experience-related variables.
@@
 This is used to calculate the variable.
-It produces a highly correlated results as shown in Figure \ref{fig:compare-xp-sw}.
+It produces reasonably similar but not exactly the same values.
+Because the original data set did not include year of the survey, it cannot be precisely compared.
 
-Secondly, in the original dataset, wages were inflation-adjusted to 1990 prices.
+Highest grade completed has some confusion.
+There are several ways that this is reported, including the highest grade ever completed, and also the hgc at each survey year.
+To match the original data, it is appropriate to use the hgc while in high school.
+The documentation suggests that this variable is available but it is not actually present in the data.
+Hence, for matching the original data, we have calculated the hgc to match the hgc achieved in the years between 1979 and 1994 based on the yearly survey value.
+The result does not exactly match with the original for a few individuals.
+
+In the original dataset, wages were inflation-adjusted to 1990 prices.
 This is not done for the refreshed data because we plan to keep refreshing it as the data added to and released from the survey.
@@
 
-Figure \ref{fig:plotting-sw-do} shows a comparison of the two sets.
-(A direct ID matching is not possible.) There are 888 individuals in the original and 1,188 individuals in the refreshed data.
-This suggests some individuals were removed in the original data.
-In addition, in the original set the racial breakdown was 246 Black, 204 Hispanic and 438 White participants, while in the refreshed data there are 346 Black, 219 Hispanic, and 623 White participants, so the proportions have not been exactly reproduced.
-On education, in the refreshed data there are very few individuals with less than a 12th grade education, which is different from the original.
-<!--AE comment: How many individuals up to 8th grade are in the refreshed dataset?
-What are plausible reasons for these differences?--> In the original data there were 366 individuals with up to 8th grade and 522 9th-12th grade, as compared to 967 with 12th grade in the refreshed data.
+Lastly, the race variable is almost perfectly matched. The id's in the dropouts subset of the refreshed data all correspond to males, which is a perfect match with the original subset.
+
+ <!--AE comment: How many individuals up to 8th grade are in the refreshed dataset? What are plausible reasons for these differences?-->
 

### 2022-02-18 paper/paper.tex
takeways section revised

@@
 \item
-  \texttt{hgc}: The highest grade completed.
-\item
-  \texttt{hgc\_i}: Integer value of the highest grade completed.
+  \texttt{hgc}: The highest grade completed ever.
+\item
+  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
 \item
@@
 
-There are two aspects of the original data that we cannot replicate.
-The first is that the calculation of the work experience is not clearly articulated.
-Singer and Willett (2003) mention that the temporal variable is the years of experience since the first day of work.
-However, this exact same variable is not explicitly available in the database.
+There are several aspects of the original data that were difficult to replicate.
+Calculation of the work experience is not clearly articulated.
+Singer and Willett (2003) describe the temporal variable as the years of experience since the first day of work.
+This variable is not explicitly available in the database.
 From the NLSY79 topical guide (Bureau of Labor Statistics, U.S.
@@
 This is used to calculate the variable.
-It produces a highly correlated results as shown in Figure \ref{fig:compare-xp-sw}.
-
-Secondly, in the original dataset, wages were inflation-adjusted to 1990 prices.
+It produces reasonably similar but not exactly the same values.
+Because the original data set did not include year of the survey, it cannot be precisely compared.
+
+Highest grade completed has some confusion.
+There are several ways that this is reported, including the highest grade ever completed, and also the hgc at each survey year.
+To match the original data, it is appropriate to use the hgc while in high school.
+The documentation suggests that this variable is available but it is not actually present in the data.
+Hence, for matching the original data, we have calculated the hgc to match the hgc achieved in the years between 1979 and 1994 based on the yearly survey value.
+The result does not exactly match with the original for a few individuals.
+
+In the original dataset, wages were inflation-adjusted to 1990 prices.
 This is not done for the refreshed data because we plan to keep refreshing it as the data added to and released from the survey.
@@
 
-Figure \ref{fig:plotting-sw-do} shows a comparison of the two sets.
-(A direct ID matching is not possible.) There are 888 individuals in the original and 1,188 individuals in the refreshed data.
-This suggests some individuals were removed in the original data.
-In addition, in the original set the racial breakdown was 246 Black, 204 Hispanic and 438 White participants, while in the refreshed data there are 346 Black, 219 Hispanic, and 623 White participants, so the proportions have not been exactly reproduced.
-On education, in the refreshed data there are very few individuals with less than a 12th grade education, which is different from the original.
-In the original data there were 366 individuals with up to 8th grade and 522 9th-12th grade, as compared to 967 with 12th grade in the refreshed data.
+Lastly, the race variable is almost perfectly matched. The id's in the dropouts subset of the refreshed data all correspond to males, which is a perfect match with the original subset.
 

### 2022-02-18 paper/paper.Rmd
fixed the reviewer comment on censoring

@@
 These need to be inspected more closely.
+It is also important to note that there are some shorter profiles, indicating that some individuals have left the study before it has finished.
+Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis, to account bias induced by the inadvertent censoring.
 

### 2022-02-18 response/response1.Rmd
fixed the reviewer comment on censoring
satisfied all of the NEED HELP

@@
 This information was compared with Singer & Willet's original data.
-Comparison shown in Figure 7 shows that the information does not match up, however, there is a high correlation between the variables.
-
-* **NEED HELP in responding comments on Figure 8.**
+Comparison between the subsets has been completely revised.
 
@@
 * We have corrected all minor grammatical errors pointed out by the reviewers.
-
-* We have made the descriptions of the steps more explicit as follows [LIST OF THE CHANGES]
 
@@
 
-* **NEED HELP** The paper has been fixed so that male/female is used consistently. We have also rewritten the paragraph to focus on the survey questions and shortcomings.
+* The paper has been fixed so that male/female is used consistently. We have also rewritten the paragraph to focus on the survey questions and shortcomings.
 
@@
 
-* **NEED HELP**
+* Added some additional text on this.
 
@@
 
-* **NEED HELP** Figure 8 has been updated, and other comparisons of the original and refreshed data have been added.
-There are still some differences, but with the change in variables being created and better filtering to get the dropouts the data are now more similar.
+* The comparison of the original and refreshed data has been completely rewritten.
+There are still some differences, but with the change in variables being created and better filtering to get the dropouts the subsets are now more similar.
 
@@
 
-* NEED HELP We are really not sure what this refers to.
+* We are really not sure what this refers to. Perhaps it is solved with the revisions.

### 2022-02-18 paper/paper.Rmd
Added hgc_1979 explanation.

--- 2022-02-18 paper/paper.Rmd
@@
 In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (`gender`), race (`race`), age (`age_1979`), highest grade completed reported in each round of survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g.
-9th grade = 9, 3rd year college = 15, (`hgc_i`), and whether the graduate equivalency diploma is obtained (`ged`).
+9th grade = 9, 3rd year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).
 
+For `gender` and `race`, we only rename the column names provided in the raw data, `new_data_qnames`.
 It is worth noting that `gender` comes from a variable called `sex` in the database.
-We use the term `gender` as the reports about the data on various pages in the website more commnly use `gender`.
+We use the term `gender` as the reports about the data on various pages in the website more commonly use `gender`.
 Gender, as provided in the data, is self-reported and only has two categories, "male" and "female".
@@
 
-The raw data, `new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981` which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
+`new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981` which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 
-The refreshed dataset has two types of highest grade completed.
+The refreshed dataset has three types of highest grade completed.
 The first one is the highest grade completed ever reported in the database (`hgc` and `hgc_i`, for the factor and integer type, respectively).
@@
 
-The second one is the highest grade completed corresponds to value  reported in each round of the survey (`grade`), meaning that the value can change.
+The second one is the highest grade completed in 1979 (`hgc_1979`), the first round of the survey.
+We decide to include this variable in the dataset for comparison with the original data-purpose.
+The original data only covers the period until 1994.
+Hence, `hgc` in the refreshed data, which is the state of `hgc` in 2018, and the `hgc` in the original data will not be comparable.
+
+The third one is the highest grade completed corresponds to value  reported in each round of the survey (`grade`), meaning that the value can change.
 We decide to include this variable in the refreshed dataset to enrich the analysis possibly done with the data, for example if one would like to explore how the changes of highest grade completed in individuals affecting their wages.
@@
 Since it is only measured since the last interview, this needs to be cumulated for each survey year.
-This produced results more similar to the original data (as discussed futher in Section \@ref(takeaways)).
+This produced results more similar to the original data (as discussed further in Section \@ref(takeaways)).
 
@@
 
-The list of variables provided in the three new datasets are as follows:
+In addition, the list of variables that we cover in those three datasets are as follows:
 
@@
   4. `race` : Race of the individual, NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
-  5. `hgc`: The highest grade completed.
-  6. `hgc_i`: Integer value of the highest grade completed.
-  7. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+  5. `hgc`: The highest grade completed ever.
+  6. `hgc_i`: Integer value of the highest grade completed ever.
+  7. `hgc_1979`: The highest grade completed in 1979 (integer value). 
+  8. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
 
@@
   6. `race`: Race where the individual belongs to: NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
-  7. `hgc`: The highest grade completed ever.
-  8. `hgc_i`: Integer value of the highest grade completed ever.
-  9. `grade`: Integer value of the highest grade completed corresponding to `year`. 
-  10. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
-  11. `njobs`: Number of jobs that an individual has.
-  12. `hours:`: The total number of hours the individual usually works per week.
-  13. `stwork`: The year when the individual starting to work.
-  14. `yr_wforce`: The length of time in the workforce in years (`year` - `stwork`).
-  15. `exp`: Work experience, i.e., the number of years of working. 
-  16. `is_wm`: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
-  17. `is_pred`: Whether the mean hourly wage is a predicted value of RLM or not.
+  7. `grade`: Integer value of the highest grade completed corresponding to `year`. 
+  8. `hgc`: The highest grade completed ever.
+  9. `hgc_i`: Integer value of the highest grade completed ever.
+  10. `hgc_1979`: The highest grade completed in 1979 (integer value).
+
+  11. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+  12. `njobs`: Number of jobs that an individual has.
+  13. `hours:`: The total number of hours the individual usually works per week.
+  14. `stwork`: The year when the individual starting to work.
+  15. `yr_wforce`: The length of time in the workforce in years (`year` - `stwork`).
+  16. `exp`: Work experience, i.e., the number of years of working. 
+  17. `is_wm`: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
+  18. `is_pred`: Whether the mean hourly wage is a predicted value of RLM or not.
 
 # Comparison of refreshed with the original data {#compare}
+
+```{r sw_wages}
+```
+
+```{r do_refreshed}
+```
+
+```{r compare-sw-do}
+```
 
@@
 
+```{r compare-xp-sw, fig.cap = "Density plot comparing the calculated work experience in the refreshed data, up to 1994, with that of the original data for male dropouts.
+Experience is shown on a square root scale to reduce the skewness.
+The highest concentration of observations is along the x=y axis, which means that most observations at most time points are similar.", out.width = "70%", eval=FALSE}
+```
+
+```{r plotting-sw-do, fig.cap = "Comparison of original textbook example (A) with refreshed data (B).
+The original data was inflation-adjusted to 1990 prices and the individual's time of collection was converted to a length of experience in the workforce, which makes it difficult to precisely compare the two sets.", fig.width = 9, fig.height = 5, out.width = "100%", cache = FALSE, eval=FALSE}
+```
+
 ## Summaries of original with refreshed dropouts data
@@
 Figure \ref{fig:compare-subsets} contains summaries of corresponding variables in the two subsets.
-Plot (A) shows a back-back bar chart of the highest grade completed.
-The two sets are almost the same, but small differences remain.
-Plots (B) and (C) show stacked density plots of experience and log wages, respectively.
-The distributions are relatively close, with more differences in wages as would be expected becuase the refreshed data is not inflation-adjusted.
 
-```{r calculate-summaries-on-subsets}
-```
-
-```{r compare-subsets, fig.cap = "Comparison of original and refreshed data: (A) highest grade completed, (B) experience and (C) log wages.
-Some difference in wages would be expected because the refreshed data is not inflation-adjusted, but the two sets are reasonably similar.", fig.width = 9, fig.height = 5, out.width = "100%", cache = FALSE}
+```{r compare-subsets, fig.cap = "Comparison of original textbook example with refreshed data: (A) highest grade completed, (B) experience and (C) log wages", fig.width = 9, fig.height = 7, out.width = "100%", cache = FALSE}
 ```
@@
 
+```{r compare-xp-to-yrworkforce, fig.cap = "The association between the lenght of time in the workforce and the work experience.
+Generally, the longer someone were in the workforce, the higher the work experience.
+However, some of the respondents have much lower working experience compared to their duration in the workforce.
+This is possible when someone ", eval=FALSE}
+```
+
 <!--AE comment: “The treatment of unlikely wages differ in the refreshed data.” I think this means “The treatment of anomalous wages differs between the original and refreshed data sets.” Then you say “We opted to use the weights from a robust linear regression to determine what should be set as missing value as described in Section 3.2.1.” I don’t think you were setting missing values though, were you?
@@
 What are plausible reasons for these differences?--> In the original data there were 366 individuals with up to 8th grade and 522 9th-12th grade, as compared to 967 with 12th grade in the refreshed data.
+
+```{r summaries, eval=FALSE}
+```
+
+<!-- Hence, in this section, we will compare how close the `wages_hs_do`, henceforth refreshed data, to the textbook data.
+It is also worth noting that the textbook data is reused from @MurnaneRichardJ1999DMDB’s paper.
+Hence, it is not directly extracted from the NLSY79 database.
+@SingerJudithD2003Alda do not state the time frame covered by the data.
+However, from @MurnaneRichardJ1999DMDB’s paper, we find that the textbook data cover the survey period from 1979-1994.
+Hence, for comparison purposes, we filter the refreshed data only until 1994.
+
+In the absence of highs chool dropouts criteria in the textbook data, we create a clearer criteria of high school dropouts as mentioned in Section \@ref(#target).
+Consequently, we do not have the same ID of high-school dropouts as in the textbook data.
+The textbook data has 888 IDs included, while the refreshed data has xx IDs.
+Furthermore, there are only xx IDs included both in the textbook and the refreshed data.
+After observing @MurnaneRichardJ1999DMDB’s paper, it is mentioned that the data is randomly drawn from the low income subsample of the NLSY.
+This is probably the reason why, even using some criteria, we cannot match the dropouts in the textbook and the refreshed data.
+
+Further, the temporal variable used in the textbook data is experience, instead of year.
+The only explanation we found regarding this variable is that it is derived from the years of IDs’ experience since their first job.
+@SingerJudithD2003Alda stated that this variable reflects the specific moment in each ID labor force history associated with each wage observation.
+Since it is not clearly articulated, we decide to keep the temporal variable in the refreshed data as year.
+Accordingly, in this comparison, we only compare the same ID but not the same temporal variable.
+
+Regarding the wages, the textbook data express the wages as inflation adjusted with 1990 dollars as constant [@SingerJudithD2003Alda].
+However, it is not stated how exactly the wages are adjusted.
+Note that in the refreshed data we publish in `yowie`, we do not adjust the data with inflation because the time frame covered ranged from 1979-2018.
+Hence, we allow flexibility to the data user to adjust, not adjust,  or adjust it with different base years.
+For comparison purposes, we adjust the wages with the average Consumer Price Index For All Urban Consumers (CPI-U) we got from @fed.
+-->
+
+<!-- Figure \ref{fig:plotting-sw-do} shows the comparison between the textbook and the refreshed data.
+Even though it has different temporal variables and possibly different CPI indicators used in inflation adjustment, we can still capture some of the patterns in the refreshed data.
+
+From this comparison, we learn that we cannot make an apple to apple comparison and produce the exact same data as the textbook due to the absence of clearly articulated data extraction and derivation process.
+Hence, the takeaway is that it is imperative to disclose how the data derived from the initial source to make data analysis reproducible.
+This is why in this paper, we show the practice of reproducible and responsible data cleaning workflow to make sure the longitudinal data from a continuous survey can be refreshed.-->
 

### 2022-02-18 paper/paper.tex
Added hgc_1979 explanation.

--- 2022-02-18 paper/paper.tex
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), and whether the graduate equivalency diploma is obtained (\texttt{ged}).
-
+In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
+
+For \texttt{gender} and \texttt{race}, we only rename the column names provided in the raw data, \texttt{new\_data\_qnames}.
 It is worth noting that \texttt{gender} comes from a variable called \texttt{sex} in the database.
-We use the term \texttt{gender} as the reports about the data on various pages in the website more commnly use \texttt{gender}.
+We use the term \texttt{gender} as the reports about the data on various pages in the website more commonly use \texttt{gender}.
 Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female''.
@@
 
-The raw data, \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981} which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
+\texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981} which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 
-The refreshed dataset has two types of highest grade completed.
+The refreshed dataset has three types of highest grade completed.
 The first one is the highest grade completed ever reported in the database (\texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively).
@@
 
-The second one is the highest grade completed corresponds to value reported in each round of the survey (\texttt{grade}), meaning that the value can change.
+The second one is the highest grade completed in 1979 (\texttt{hgc\_1979}), the first round of the survey.
+We decide to include this variable in the dataset for comparison with the original data-purpose.
+The original data only covers the period until 1994.
+Hence, \texttt{hgc} in the refreshed data, which is the state of \texttt{hgc} in 2018, and the \texttt{hgc} in the original data will not be comparable.
+
+The third one is the highest grade completed corresponds to value reported in each round of the survey (\texttt{grade}), meaning that the value can change.
 We decide to include this variable in the refreshed dataset to enrich the analysis possibly done with the data, for example if one would like to explore how the changes of highest grade completed in individuals affecting their wages.
@@
 Since it is only measured since the last interview, this needs to be cumulated for each survey year.
-This produced results more similar to the original data (as discussed futher in Section \ref{takeaways}).
+This produced results more similar to the original data (as discussed further in Section \ref{takeaways}).
 
@@
 \item
-  \texttt{hgc}: The highest grade completed.
-\item
-  \texttt{hgc\_i}: Integer value of the highest grade completed.
+  \texttt{hgc}: The highest grade completed ever.
+\item
+  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
+\item
+  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
 \item
@@
 \def\labelenumi{\arabic{enumi}.}
-\tightlist
 \item
@@
 \item
+  \texttt{grade}: Integer value of the highest grade completed corresponding to \texttt{year}.
+\item
   \texttt{hgc}: The highest grade completed ever.
@@
 \item
-  \texttt{grade}: Integer value of the highest grade completed corresponding to \texttt{year}.
+  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
 \item
@@
 Figure \ref{fig:compare-subsets} contains summaries of corresponding variables in the two subsets.
-Plot (A) shows a back-back bar chart of the highest grade completed.
-The two sets are almost the same, but small differences remain.
-Plots (B) and (C) show stacked density plots of experience and log wages, respectively.
-The distributions are relatively close, with more differences in wages as would be expected becuase the refreshed data is not inflation-adjusted.
 
@@
 
-\caption{Comparison of original and refreshed data: (A) highest grade completed, (B) experience and (C) log wages.
-Some difference in wages would be expected because the refreshed data is not inflation-adjusted, but the two sets are reasonably similar.}\label{fig:compare-subsets}
+\caption{Comparison of original textbook example with refreshed data: (A) highest grade completed, (B) experience and (C) log wages}\label{fig:compare-subsets}
 \end{figure}

### 2022-02-18 data-raw/data-preprocessing.R

@@ -174,7 +174,8 @@ ged <- new_data_qnames %>%
   select(id, dip_or_ged)
 
 # join the hgc_ever and GED as education information for NLSY79 demographic data
-highest_year <- left_join(hgc_ever, ged, by = "id")
+highest_year <- left_join(hgc_ever, ged, by = "id") %>%
+  left_join(hgc_1979, by = "id")
 
 ## ---- full-demog
 
@@ -192,6 +193,7 @@ demog_nlsy79 <- full_demographics %>%
                 race,
                 hgc,
                 hgc_i,
+                hgc_1979,
                 ged = dip_or_ged) %>%
   mutate(id = as.factor(id),
          age_1979 = as.integer(age_1979),

### 2022-02-18 data-raw/eda.R

@@ -6,9 +6,7 @@ sw <- brolgar::wages %>%
   mutate(index = 1:n(),
          id = as.factor(id))
 
-# load the dataset of hgc in 1979
-hgc_1979 <- hgc_1979 %>%
-  mutate(id = as.factor(id))
+
 
 # Comparing the original and refreshed data
 # tabulating hgc
@@ -28,11 +26,11 @@ do_1994 <- wages_hs_do %>%
   ungroup()
 
 d94_hgc <- do_1994 %>%
-  select(id, hgc_1994) %>%
-  filter(between(hgc_1994, 6, 12)) %>%
+  select(id, hgc_1979) %>%
+  filter(between(hgc_1979, 6, 12)) %>%
   distinct() %>%
-  count(hgc_1994) %>%
-  rename(hgc = hgc_1994, refreshed = n)
+  count(hgc_1979) %>%
+  rename(hgc = hgc_1979, refreshed = n)
 
 hgc_join <- left_join(sw_hgc, d94_hgc) %>%
   mutate(original = -original, hgc = factor(hgc)) %>%
@@ -113,244 +111,244 @@ wg_p <- ggplot(wg_join) +
   theme_bw() +
   theme(legend.position = "none")
 
-## compare-subsets
+## ---- compare-subsets
 hgc_p + exp_p + wg_p
 
 # load the refreshed data, join it with the hgc_1979
-do <- wages_hs_do %>%
-  as_tibble() %>%
-  left_join(hgc_1979, by = "id") %>%
-  group_by(id) %>%
-  mutate(lnwage = log(wage),
-         index = 1:n())
-
-# join original and refreshed data, only for the same individuals that appears in the two data
-do_sw_join <- inner_join(sw, do, by = c("id", "index"))
-
-sw_wages <- sw %>%
-  ggplot(aes(x = xp,
-             y = ln_wages)) +
-  geom_line(aes(group = id), alpha = 0.1) +
-  geom_smooth(se = FALSE, colour = "pink") +
-  labs(tag = "(A)", title = "Original") +
-  theme_bw() +
-  ylab("Hourly wage ($, natural log)") +
-  xlab("Experience (years)") +
-  #theme(plot.title = element_text(size = 10)) +
-  ylim(0, 5)
-
-sw_wages_mod <- do_sw_join %>%
-  as_tibble() %>%
-  mutate(hgc = ifelse(high_grade < 9, "8TH", "12TH")) %>%
[…]

### 2022-02-18 data-raw/ida.R

@@ -282,7 +282,7 @@ wages_clean <- wages_cleaned %>%
 
 # rename and select the wages in tidy
 wages <- wages_clean %>%
-  select(id, year, mean_hourly_wage, age_1979, gender, race, grade, hgc, hgc_i, ged = dip_or_ged,
+  select(id, year, mean_hourly_wage, age_1979, gender, race, grade, hgc, hgc_i, hgc_1979, ged = dip_or_ged,
          number_of_jobs, total_hours, stwork, yr_wforce, exp, is_wm, is_pred) %>%
   mutate(id = as.factor(id),
          hgc = as.factor(hgc),

### 2022-02-18 paper/paper.Rmd

--- 2022-02-18 paper/paper.Rmd
@@
   6. `hgc_i`: Integer value of the highest grade completed ever.
-  7. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+  7. `hgc_1979`: The highest grade completed in 1979 (integer value). 
+  8. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
 
@@
   6. `race`: Race where the individual belongs to: NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
-  7. `hgc`: The highest grade completed ever.
-  8. `hgc_i`: Integer value of the highest grade completed ever.
-  9. `grade`: Integer value of the highest grade completed corresponding to `year`. 
-  10. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
-  11. `njobs`: Number of jobs that an individual has.
-  12. `hours:`: The total number of hours the individual usually works per week.
-  13. `stwork`: The year when the individual starting to work.
-  14. `yr_wforce`: The length of time in the workforce in years (`year` - `stwork`).
-  15. `exp`: Work experience, i.e., the number of years of working. 
-  16. `is_wm`: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
-  17. `is_pred`: Whether the mean hourly wage is a predicted value of RLM or not.
+  7. `grade`: Integer value of the highest grade completed corresponding to `year`. 
+  8. `hgc`: The highest grade completed ever.
+  9. `hgc_i`: Integer value of the highest grade completed ever.
+  10. `hgc_1979`: The highest grade completed in 1979 (integer value).
+
+  11. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+  12. `njobs`: Number of jobs that an individual has.
+  13. `hours:`: The total number of hours the individual usually works per week.
+  14. `stwork`: The year when the individual starting to work.
+  15. `yr_wforce`: The length of time in the workforce in years (`year` - `stwork`).
+  16. `exp`: Work experience, i.e., the number of years of working. 
+  17. `is_wm`: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
+  18. `is_pred`: Whether the mean hourly wage is a predicted value of RLM or not.
 

### 2022-02-18 data-raw/data-preprocessing.R
Clean unneccesary codes.

@@ -157,7 +157,7 @@ grade_ok <- left_join(demog_education, hgc_ever, by = "id") %>%
 grade <- rbind(grade_ok, grade_fixed) %>%
   select(id, year, grade)
 
-# get the highest grade completed reported in the first round of the survey
+# get the highest grade completed reported in the first round of the survey (1979)
 hgc_1979 <- grade %>%
   filter(year == 1979) %>%
   select(id, hgc_1979 = grade)

### 2022-02-18 data-raw/eda.R

@@ -6,7 +6,6 @@ sw <- brolgar::wages %>%
   mutate(index = 1:n(),
          id = as.factor(id))
 
-
 ## ---- calculate-summaries-on-subsets
 # Comparing the original and refreshed data
 # tabulating hgc
@@ -117,27 +116,6 @@ wg_p <- ggplot(wg_join) +
 ##  ---- compare-subsets
 hgc_p + exp_p + wg_p
 
-
-# Takeaways:
-
-# we don't know how exactly the criteria of drop out,
-# resulted view agreements of id being dropped out in the two dataset.
-
-# with the same id, the ln(wages) didn't agree each other, probably due to:
-
-# we don't have experience variable in the original data base.
-# we want to calculate it, but we don't know how Singer-Willet compute the experience
-# we want to compare it just with year, Singer and Willet dont have year variable
-# hence, it is hard to create key-index pair.
-
-
-# for people who work show data example (book, research, ets), it is very important to disclose how the data derived from the initial source, to make the analysis reproducible, especially for longitudinal data because other researcher often want to compare with the recent data.
-# in this case, in Singer-Willet textbook, it important to state how the experience variable derived from the database.
-# In this EDA, we found that it is difficult to make apple to apple comparison (and refresh the data in general) cause we don't know how exactly they calculate the variable (even the wages)
-
-
-# This is why in our paper, we show the practice of reproducible and responsible data cleaning work flow to make sure the data can be refreshed.
-# Especially with longitudinal data from a survey that is still being held (so it will be refreshed from year to year).
 
 
 

### 2022-02-18 data-raw/ida.R

@@ -386,33 +386,6 @@ n_obs_do_pred <- filter(wages_hs_do, is_pred == TRUE) %>%
 
 ### ---- flow-chart
 include_graphics(here::here("paper/figures/flowchart.png"))
-# grViz("digraph flowchart {
-#       node [fontname = Helvetica, shape = rectangle]
-#       tab1 [label = '@@1']
-#       tab2 [label = '@@2']
-#       tab3 [label = '@@3']
-#       tab4 [label = '@@4']
-#       tab5 [label = '@@5']
-#       tab6 [label = '@@6']
-#       tab7 [label = '@@7']
-#
-#       tab1 -> tab2;
-#       tab1 -> tab3;
-#       tab1 -> tab4 -> tab5 -> tab6;
-#       tab5 -> tab7;
-#       tab6 -> tab7;
-#       tab2 -> tab7
-#       }
-#
-#       [1]: paste0('NLSY79, n = ', a)
-#       [2]: paste0('Extract demographic variable, n = ', b)
-#       [3]: paste0('Exclude ', a - d, ' IDs whose hourly rate is missing')
-#       [4]: paste0('Eligible ID, n = ', d)
-#       [5]: paste0('Cohort whose hgc is up to 12th grade and \\n participated at least 5 years in the survey, \\n n = ', n_hgc, ', n_obs = ', n_obs_hgc, ' \\n (', n_obs_hgc_pred, ' observations are predicted value)')
-#       [6]: paste0('High school dropouts cohort, \\n n = ', n_do, ', n_obs = ', n_obs_do, ' \\n (', n_obs_do_pred, ' observations are predicted value)')
-#       [7]: 'yowie Package'
-#       ",
-#             height = 700)
 
 ### ---- flow-chart-blind
 include_graphics(here::here("paper/figures/flowchart_blind.png"))

### 2022-02-18 paper/paper.Rmd
tidied up the new section on hgc

@@
 
-`new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981` which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
+The object `new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981` which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 
-The refreshed dataset has three types of highest grade completed.
-The first one is the highest grade completed ever reported in the database (`hgc` and `hgc_i`, for the factor and integer type, respectively).
-For each ID, there is only one value of `hgc` and `hgc_i`.
-This variable is obtained from `new_data_qnames` with name `HGC_EVER_XRND` and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be `hgc_i` and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
+The next step is processing highest grade completed.
+There are several ways to define this, and this shoudl be reflected in the refreshed data, to giev some flexibility for downstream analysis.
+The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as `hgc` and `hgc_i`, for the factor and integer type, respectively.
+For each individual, there is only one value of `hgc` and `hgc_i`.
+This variable is obtained from `new_data_qnames` with name `HGC_EVER_XRND` and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as `hgc_i` and recoded as a factor to give `hgc` (e.g., 10th grade, 11th grade, and so on).
 
-The second one is the highest grade completed in 1979 (`hgc_1979`), the first round of the survey.
-We decide to include this variable in the dataset for comparison with the original data-purpose.
-The original data only covers the period until 1994.
-Hence, `hgc` in the refreshed data, which is the state of `hgc` in 2018, and the `hgc` in the original data will not be comparable.
+The second definition is the highest grade completed in 1979 (provided in the refreshed data as `hgc_1979`), corresponding to the value from the first round of the survey.
+This best reflects the grade when the individual left school, and is included in order to compare with the original data.
 
-The third one is the highest grade completed corresponds to value  reported in each round of the survey (`grade`), meaning that the value can change.
-We decide to include this variable in the refreshed dataset to enrich the analysis possibly done with the data, for example if one would like to explore how the changes of highest grade completed in individuals affecting their wages.
-The highest grade completed are recorded in `new_data_qnames` as variables beginning with `Q3-4` and `HGC` with suffix of the year it was recorded.
-In addition the variables beginning with `HGCREV` contain the revised data.
-We choose to use this revised data because it seemed to have less missing and presumably has been checked.
-However, there is no revised data for 2012, 2014, 2016, and 2018, thus, we use the ordinary data for these years.
+The third definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as `grade`).
+This value can change over time, but it should only increase.
+This has been included in the refreshed dataset to enable richer downstream analyses, for example if one would like to explore how the temporal changes in education level affects the wages of individuals.
+This is recorded in `new_data_qnames` as columns beginning with `Q3-4` and `HGC` with year as a suffix.
+In addition, it is also in the columns beginning with `HGCREV` reflecting revised data.
+We choose to use this revised values because there were less missing values indicating it had been more thoroughly checked.
+When this value was missing, 2012, 2014, 2016, and 2018, but available in the first form substituted accordingly.
 <!--We also check the consistency between `hgc_i` and `grade`.
@@
 
- <!--AE comment: Not panel 9 --> the IDs in panel 9. 
+ <!--AE comment: Not panel 9  the IDs in panel 9. -->
 <!--I think there is a narrative step missing here, which is that looking at those samples led you to look at plots that show values for the wage variable over all participants and all years.

### 2022-02-18 paper/paper.tex
tidied up the new section on hgc

--- 2022-02-18 paper/paper.tex
@@
 
-The raw data, \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981} which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
+The object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981} which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 
-The refreshed dataset has two types of highest grade completed.
-The first one is the highest grade completed ever reported in the database (\texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively).
-For each ID, there is only one value of \texttt{hgc} and \texttt{hgc\_i}.
-This variable is obtained from \texttt{new\_data\_qnames} with name \texttt{HGC\_EVER\_XRND} and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
-
-The second one is the highest grade completed corresponds to value reported in each round of the survey (\texttt{grade}), meaning that the value can change.
-We decide to include this variable in the refreshed dataset to enrich the analysis possibly done with the data, for example if one would like to explore how the changes of highest grade completed in individuals affecting their wages.
-The highest grade completed are recorded in \texttt{new\_data\_qnames} as variables beginning with \texttt{Q3-4} and \texttt{HGC} with suffix of the year it was recorded.
-In addition the variables beginning with \texttt{HGCREV} contain the revised data.
-We choose to use this revised data because it seemed to have less missing and presumably has been checked.
-However, there is no revised data for 2012, 2014, 2016, and 2018, thus, we use the ordinary data for these years.
+The next step is processing highest grade completed.
+There are several ways to define this, and this shoudl be reflected in the refreshed data, to giev some flexibility for downstream analysis.
+The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively.
+For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}.
+This variable is obtained from \texttt{new\_data\_qnames} with name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on).
+
+The second definition is the highest grade completed in 1979 (provided in the refreshed data as \texttt{hgc\_1979}), corresponding to the value from the first round of the survey.
+This best reflects the grade when the individual left school, and is included in order to compare with the original data.
+
+The third definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as \texttt{grade}).
+This value can change over time, but it should only increase.
+This has been included in the refreshed dataset to enable richer downstream analyses, for example if one would like to explore how the temporal changes in education level affects the wages of individuals.
+This is recorded in \texttt{new\_data\_qnames} as columns beginning with \texttt{Q3-4} and \texttt{HGC} with year as a suffix.
+In addition, it is also in the columns beginning with \texttt{HGCREV} reflecting revised data.
+We choose to use this revised values because there were less missing values indicating it had been more thoroughly checked.
+When this value was missing, 2012, 2014, 2016, and 2018, but available in the first form substituted accordingly.
 
@@
 Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis, to account bias induced by the inadvertent censoring.
-
-the IDs in panel 9.
 
@@
 \item
+  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
+\item
   \texttt{ged}: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.

### 2022-02-18 paper/paper.Rmd
Corrects the spelling and grammatical in Section 4.2 and 4.3.

--- 2022-02-18 paper/paper.Rmd
@@
 This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
-Similar this also relates to race, as reported in the database.
+Similarly, this also relates to race, as reported in the database.
 When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
@@
 
-Because the original data does not have the year of collection, it is not possible to directly merge the two subsets.
+Because the original data does not have the year of collection, it is not possible to merge the two subsets directly.
 Merging longitudinal data requires both the key (`id`) and the index (ideally survey year).
-In the original data the experience variable is the time index, and it was not possible to exactly match this for the refreshed data.
-Thus, comparisons of the two sets has to be conducted in a two sample fashion rather than as a matched sample.
+In the original data, the experience variable is the time index, and it was not possible to exactly match this for the refreshed data.
+Thus, comparisons of the two sets have to be conducted in a two-sample fashion rather than a matched sample.
 
@@
 Plots (B) and (C) show stacked density plots of experience and log wages, respectively.
-The distributions are relatively close, with more differences in wages as would be expected becuase the refreshed data is not inflation-adjusted.
+The distributions are relatively close, with more differences in wages as would be expected because the refreshed data is not inflation-adjusted.
 
@@
 There are several aspects of the original data that were difficult to replicate.
-Calculation of the work experience is not clearly articulated.
+The calculation of the work experience is not clearly articulated.
 @SingerJudithD2003Alda describe the temporal variable as the years of experience since the first day of work.
@@
 It produces reasonably similar but not exactly the same values.
-Because the original data set did not include year of the survey, it cannot be precisely compared.
+Because the original data set did not include the year of the survey, it cannot be precisely compared.
 
-Highest grade completed has some confusion.
-There are several ways that this is reported, including the highest grade ever completed, and also the hgc at each survey year.
-To match the original data, it is appropriate to use the hgc while in high school.
-The documentation suggests that this variable is available but it is not actually present in the data.
-Hence, for matching the original data, we have calculated the hgc to match the hgc achieved in the years between 1979 and 1994 based on the yearly survey value.
+The highest grade completed has some confusion.
+There are several ways that this is reported, including the highest grade ever completed and also the `hgc` at each survey year.
+To match the original data, it is appropriate to use the `hgc` while in high school.
+The documentation suggests that this variable is available, but it is not actually present in the data.
+Hence, for matching the original data, we have calculated the `hgc` to match the `hgc` achieved in the years between 1979 and 1994 based on the yearly survey value.
 The result does not exactly match with the original for a few individuals.
@@
 In the original dataset, wages were inflation-adjusted to 1990 prices.
-This is not done for the refreshed data because we plan to keep refreshing it as the data added to and released from the survey.
-Instead we have provided a function in the R package, `r yowie`, for users to conduct the inflation adjustment when they are ready to analyse the data.
+This is not done for the refreshed data because we plan to keep refreshing it as the data is added to and released from the survey.
+Instead, we have provided a function in the R package, `r yowie`, for users to conduct the inflation adjustment when they are ready to analyse the data.
 
@@
 However, this value is too low to be set as the maximum threshold, and it doesn't take into account temporal neighbors for an individual.
-We opted to use the weights from a robust linear regression to determine what should be regarded as extreme and imputed them with their predicted values as as described in Section \@ref(censor).
+We opted to use the weights from a robust linear regression to determine what should be regarded as extreme and imputed them with their predicted values as described in Section \@ref(censor).
 
-Lastly, the race variable is almost perfectly matched. The id's in the dropouts subset of the refreshed data all correspond to males, which is a perfect match with the original subset.
+Lastly, the race variable is almost perfectly matched. The `id`'s in the dropouts subset of the refreshed data all correspond to males, which perfectly matches the original subset.
 

### 2022-02-18 paper/paper.tex
Corrects the spelling and grammatical in Section 4.2 and 4.3.

--- 2022-02-18 paper/paper.tex
@@
 
-Because the original data does not have the year of collection, it is not possible to directly merge the two subsets.
+Because the original data does not have the year of collection, it is not possible to merge the two subsets directly.
 Merging longitudinal data requires both the key (\texttt{id}) and the index (ideally survey year).
-In the original data the experience variable is the time index, and it was not possible to exactly match this for the refreshed data.
-Thus, comparisons of the two sets has to be conducted in a two sample fashion rather than as a matched sample.
+In the original data, the experience variable is the time index, and it was not possible to exactly match this for the refreshed data.
+Thus, comparisons of the two sets have to be conducted in a two-sample fashion rather than a matched sample.
 
@@
 There are several aspects of the original data that were difficult to replicate.
-Calculation of the work experience is not clearly articulated.
+The calculation of the work experience is not clearly articulated.
 Singer and Willett (2003) describe the temporal variable as the years of experience since the first day of work.
@@
 It produces reasonably similar but not exactly the same values.
-Because the original data set did not include year of the survey, it cannot be precisely compared.
-
-Highest grade completed has some confusion.
-There are several ways that this is reported, including the highest grade ever completed, and also the hgc at each survey year.
-To match the original data, it is appropriate to use the hgc while in high school.
-The documentation suggests that this variable is available but it is not actually present in the data.
-Hence, for matching the original data, we have calculated the hgc to match the hgc achieved in the years between 1979 and 1994 based on the yearly survey value.
+Because the original data set did not include the year of the survey, it cannot be precisely compared.
+
+The highest grade completed has some confusion.
+There are several ways that this is reported, including the highest grade ever completed and also the \texttt{hgc} at each survey year.
+To match the original data, it is appropriate to use the \texttt{hgc} while in high school.
+The documentation suggests that this variable is available, but it is not actually present in the data.
+Hence, for matching the original data, we have calculated the \texttt{hgc} to match the \texttt{hgc} achieved in the years between 1979 and 1994 based on the yearly survey value.
 The result does not exactly match with the original for a few individuals.
@@
 In the original dataset, wages were inflation-adjusted to 1990 prices.
-This is not done for the refreshed data because we plan to keep refreshing it as the data added to and released from the survey.
-Instead we have provided a function in the R package, \texttt{[CENSORED]}, for users to conduct the inflation adjustment when they are ready to analyse the data.
+This is not done for the refreshed data because we plan to keep refreshing it as the data is added to and released from the survey.
+Instead, we have provided a function in the R package, \texttt{[CENSORED]}, for users to conduct the inflation adjustment when they are ready to analyse the data.
 
@@
 However, this value is too low to be set as the maximum threshold, and it doesn't take into account temporal neighbors for an individual.
-We opted to use the weights from a robust linear regression to determine what should be regarded as extreme and imputed them with their predicted values as as described in Section \ref{censor}.
-
-Lastly, the race variable is almost perfectly matched. The id's in the dropouts subset of the refreshed data all correspond to males, which is a perfect match with the original subset.
+We opted to use the weights from a robust linear regression to determine what should be regarded as extreme and imputed them with their predicted values as described in Section \ref{censor}.
+
+Lastly, the race variable is almost perfectly matched. The \texttt{id}'s in the dropouts subset of the refreshed data all correspond to males, which perfectly matches the original subset.
 

### 2022-02-18 response/response1.Rmd
Corrects the spelling and grammatical in Section 4.2 and 4.3.

@@
 Given that the stated goal is to recreate the textbook data, this seems like a glaring oversight.
-It also leads to Fig 7 (Now Figure 8) being essentially meaningless: we are unsurprised that salaries are higher with more years of experience, and we are unsurprised that salaries go up over the years with both age of employee and inflation.
+It also leads to Fig 7 being essentially meaningless: we are unsurprised that salaries are higher with more years of experience, and we are unsurprised that salaries go up over the years with both age of employee and inflation.
 The positive slope of these two different plots does not mean they are related to each other in any way.}
@@
 Singer and Willet (2003) use the natural log to store the wages data by default, although without mention their motivation of doing so.
-Thus, we also use the natural log in Figure 8 to compare the refreshed data with the original data.
+Thus, we also use the natural log in Figure 7 to compare the refreshed data with the original data.
 
@@
 * We have changed Fig 2 to show 36 individuals, that display the range of wages experience, including some with questionable values.
-The figure caption reflects this and more details are provide din the text.
+The figure caption reflects this and more details are provided in the text.
 

### 2022-02-18 paper/paper.Rmd
done with final checking.

@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis",  which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
+An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis",  which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
 
-The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, a higher level of education leads to higher wages, and that race makes a difference, on average.
 An exploratory analysis reveals, however, that the individual experience varies a great deal from the overall average.
-Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
+Some individuals experience a decline in wages the longer they are in the workforce and many experience volatility in their wages.
 The wages data was used to illustrate exploratory longitudinal data analysis in @ilk2004, and was further developed into a case study for use in the teaching of exploratory data analysis at `r censor_word("Iowa State University")`.
@@
 
-This disparity between the average and the individual is a part of statistics, as a discipline, that require more attention.
+This disparity between the average and the individual is a part of statistics, as a discipline that requires more attention.
 This particular data set is a prime example of discussing this disparity.
-Textbook data sets have longevity if there is a unresolved mystery.
+Textbook data sets have longevity if there is an unresolved mystery.
 The iris data [@iris-data] is a prime example.
 It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(A side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new and does not suffer from a connection with eugenics [@notiris].
+(A side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris].
 We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
 
-For the field of statistics, and data science by association, it is increasingly important to reach the individual.
-One might describe this as a divergence of purpose, statistics for public policy or statistics for the public.
+For the field of statistics and data science by association, it is increasingly important to reach the individual.
+One might describe this as a divergence of purpose, statistics for public policy, or statistics for the public.
 The two are not the same.
-As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual.
+As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require that statistics address the individual.
 For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic.
 The majority of people in the study do not have a pattern that is similar to the average.
-If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
-A similar tone is echoed occasionally in the public media, for example, an article published in the Sydney Morning Herald argues that there is no average Australian [@notaverage].
+If you have a bad experience, that your wages have declined over time, you are not alone; there are others like you, and more than you think.
+A similar tone is echoed occasionally in the public media.
+For example, an article published in the Sydney Morning Herald argues that there is no average Australian [@notaverage].
 
-As a textbook data set though, the wages data is outdated.
+As a textbook data set, though, the wages data is outdated.
 The most recent year in the data is 1994, 10 years prior to when @SingerJudithD2003Alda was published.
@@
 Our approach to cleaning builds heavily on the `tidyverse` approach [@tidyverse].
-The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using a step-wise piping with split-apply-combine strategy for mutating new variables [@plyr].
-Tidy data shouldn't be confused with "tame data" which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
-The resulting (tame) data is provided in a new R package called `r yowie` which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
+The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables [@plyr].
+Tidy data shouldn't be confused with "tame data", which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
+The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
 
@@
 In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research [@MichaelRPergamit2001DWTN; @eliznlsy].
-According to @eliznlsy, thousands of articles, and hundreds of book chapters and monographs have utilized this data.
+According to @eliznlsy, thousands of articles and hundreds of book chapters and monographs have utilized this data.
 Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data [@MichaelRPergamit2001DWTN].
@@
 Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda.
-We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the *purposes of studying racism, not race* [@racismnotrace].
-There are a number of data sets provided by @SingerJudithD2003Alda and we focus only on this one because it has captivated our attention for a number of years.
+We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the *purposes of studying racism, not race* [@racismnotrace].
+There are a number of data sets provided by @SingerJudithD2003Alda, and we focus only on this one because it has captivated our attention for a number of years.
 We use it in our own teaching of longitudinal data analysis and would very much like to use data since 1994.
@@
 
-There are two ways to conduct the interview of the NLSY79, which are face-to-face or by telephone interviews.
+There are two ways to conduct the interview of the NLSY79, which are face-to-face or telephone interviews.
 In recent survey years, more than 90 percent of respondents were interviewed by telephone [@eliznlsy].
@@
 6. HGC: the highest grade completed.
-7. UERATE: unemployment rate of the year of survey. When missing, the variable is set to be 7.875 (the average rate).
+7. UERATE: the unemployment rate of the year of the survey. When missing, the variable is set to be 7.875 (the average rate).
 
@@
 For example, use a single categorical race variable instead of the two binary race variables.
-We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
+We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs that the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked.
 We do not attempt to re-create the unemployment rate variable.
@@
 2. A separate table of the demographic data of the whole NLSY79 cohort.
-3. The wages data of the high school dropouts, which is closest to a refreshed version of @SingerJudithD2003Alda's data.
+3. The high school dropouts' wages data is closest to a refreshed version of @SingerJudithD2003Alda's data.
 
@@
 
-@LooMarkvander2018Sdcw in the context of official statistics describe the "statistical value chain" which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organised with correct type and identified variables) and valid data (data that has been cleaned and more accurately represents the intent of variables).
-What we have colorfully named as wild data can be considered to be raw data, and valid data could be considered to be textbook data, in the above statistical value chain.
-In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidy the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so that the resulting input data can be used downstream for validating the data as described in Section \@ref(ida).
+@LooMarkvander2018Sdcw, in the context of official statistics, describe the "statistical value chain", which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+What we have colorfully named as wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
+In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidy the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so that the resulting input data can be used downstream for validating the data as described in Section \@ref(ida) and Section \@ref(censor).
 
@@
 
-The NLSY79 data contains a large number of variables but for our purposes the scope required is limited to demographic profiles, wages data, and work experience.
-More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data and navigated as described in Figure \ref{fig:source-nav}.
+The NLSY79 data contains a large number of variables, but for our purposes, the scope required is limited to demographic profiles, wages data, and work experience.
+More specifically, we went to the NLSY79 database website at \url{https://www.nlsinfo.org/content/cohorts/nlsy79/get-data}, clicked on the direct link to NSLY79 data, and navigated as described in Figure \ref{fig:source-nav}.
 
@@
 We alter only the file path in `NLSY79.R` and run the script without any other alteration.
-This results in an initial processing of the raw data into two data sets,  `categories_qnames` (where the observations are stored in categorical/interval values) and `new_data_qnames` (the observations are stored in integer form).
+This results in the initial processing of the raw data into two data sets,  `categories_qnames` (where the observations are stored in categorical/interval values) and `new_data_qnames` (the observations are stored in integer form).
 
@@
 According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
-The raw data, `new_data_qnames`, does not comply with these rules as it is organised such that each row corresponds to an individual.
+The raw data, `new_data_qnames`, does not comply with these rules as it is organized such that each row corresponds to an individual.
 As respondents can have multiple jobs at specific years, the column names, such as `HRP1_1979`, `HRP2_1979`, `HRP1_1980` and `HRP2_1980`, contain the information about the job number up to 5 (`HRP1` = job 1, `HRP2` = job 2) and the year.
@@
 
-Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
-This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long form, with `dplyr` [@dplyr] and `stringr` [@stringr] to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
+Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
+This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long-form, with `dplyr` [@dplyr] and `stringr` [@stringr] to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (`gender`), race (`race`), age (`age_1979`), highest grade completed reported in each round of survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g.
-9th grade = 9, 3rd year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).
+In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (`gender`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).
 
@@
 
-The object `new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981` which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
+The object `new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981`, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 
-The next step is processing highest grade completed.
-There are several ways to define this, and this shoudl be reflected in the refreshed data, to giev some flexibility for downstream analysis.
+The next step is processing the highest grade completed.
+There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis.
 The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as `hgc` and `hgc_i`, for the factor and integer type, respectively.
 For each individual, there is only one value of `hgc` and `hgc_i`.
-This variable is obtained from `new_data_qnames` with name `HGC_EVER_XRND` and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as `hgc_i` and recoded as a factor to give `hgc` (e.g., 10th grade, 11th grade, and so on).
+This variable is obtained from `new_data_qnames` with the name `HGC_EVER_XRND` and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as `hgc_i` and recoded as a factor to give `hgc` (e.g., 10th grade, 11th grade, and so on).
 
-The second definition is the highest grade completed in 1979 (provided in the refreshed data as `hgc_1979`), corresponding to the value from the first round of the survey.
-This best reflects the grade when the individual left school, and is included in order to compare with the original data.
-
-The third definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as `grade`).
+The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as `grade`).
 This value can change over time, but it should only increase.
-This has been included in the refreshed dataset to enable richer downstream analyses, for example if one would like to explore how the temporal changes in education level affects the wages of individuals.
+This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals.
 This is recorded in `new_data_qnames` as columns beginning with `Q3-4` and `HGC` with year as a suffix.
 In addition, it is also in the columns beginning with `HGCREV` reflecting revised data.
-We choose to use this revised values because there were less missing values indicating it had been more thoroughly checked.
+We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked.
 When this value was missing, 2012, 2014, 2016, and 2018, but available in the first form substituted accordingly.
@@
 
+The third definition is the highest grade completed in 1979 (provided in the refreshed data as `hgc_1979`), corresponding to the value from the first round of the survey.
+This is calculated from the `grade` value in 1979.
+This best reflects the grade when the individual left school and is included in order to compare with the original data.
+
 The next step is tidying to obtain `ged`.
 Along with `hgc`, `ged` is used to subset the data to the high school dropouts as in the original data.
-The graduate equivalency status is saved as variable started with "Q3-8A" followed by the year of the survey.
+The graduate equivalency status is saved as a variable started with "Q3-8A" followed by the year of the survey.
 Thus, we only separate the year and the GED status.
-Although there GED status is asked in each round of the survey, we only retain the latest status of one's GED.
+Although the GED status is asked in each round of the survey, we only retain the latest status of one's GED.
 
@@
 From 1988 onward, we selected the total hours worked per week, including hours working from home (`QES-52D`).
-However, in 1993, this variable was missing for the first and last job so we selected to use `QES-52A` instead.
+However, in 1993, this variable was missing for the first and last job, so we selected to use `QES-52A` instead.
 In addition, 2008 only had jobs 1-4 for the `QES-52D` variable, so we use only these.
@@
 The hourly wages are in the variables beginning with `HRP` in `new_data_qnames`.
-As a respondent may have multiple jobs, the `mean_hourly_wage` is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on number of hours is available); if number of hours worked for any job is missing, then the `mean_hourly_wage` is computed as a simple average of all available hourly wages.
+As a respondent may have multiple jobs, the `mean_hourly_wage` is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on the number of hours is available); if the number of hours worked for any job is missing, then the `mean_hourly_wage` is computed as a simple average of all available hourly wages.
 Prior to computing the mean hourly wage, we undertook a number of steps to treat unusual observations as described below:
@@
 For `stwork` variable,  we only rename the column names from `new_data_qnames`.
-This variable is then used to calculate the next variable, `yr_wforce` for each survey round, which is the year of survey (`year`) minus the year of individual start working (`stwork`).
+This variable is then used to calculate the next variable, `yr_wforce` for each survey round, which is the year of survey (`year`) minus the year of individual started working (`stwork`).
 Finally, `exp` variable is derived from the number of weeks worked since the last interview indicated as variable started with `WKSWK` in `new_data_qnames`.
@@
 We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data.
-However, it is worth noting that the original data does not restrict the number of observation for each individual, i.e, there are individuals with only one and two observations.
+However, it is worth noting that the original data does not restrict the number of observations for each individual, i.e., there are individuals with only one and two observations.
 
 Finally, we save the resultant wage data on this cohort as `wages`.
-Note that we save `grade` variable in this dataset, instead of `demog_nlsy79` dataset because it is a longitudinal variable, while `demog_nlsy79` is cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.
+Note that we save `grade` variable in this dataset instead of `demog_nlsy79` dataset because it is a longitudinal variable, while `demog_nlsy79` is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.
 
@@
 
-Work experience is one of the most important variables in @SingerJudithD2003Alda as it indicates time, and makes longitudinal analysis possible.
+Work experience is one of the most important variables in @SingerJudithD2003Alda as it indicates time and makes longitudinal analysis possible.
 It is desirable to calculate this rather than using the survey year for time because it more accurately reflects a person's time in the workforce.
@@
 It represents the difference between the day an individual enters the labor force (`EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND` in the database) relative to the date of the survey, which we call `yr_wforce`.
-However, using this calculation produced numbers which don't quite match with the original data.
+However, using this calculation produced numbers that do not quite match with the original data.
 
-Reading the section titled ["Topical Guide to the Data"](https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience) in the guide, suggests that it should be calculated based on the variable "number of weeks worked since last interview".
+Reading the section titled ["Topical Guide to the Data"](https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience) in the guide suggests that it should be calculated based on the variable "number of weeks worked since the last interview".
 This would remove periods of unemployment which makes sense when measuring experience while actually working.
@@
 According to @HuebnerMariannePhD2016Asat, initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure that the data is clean, valid, and ready to be deployed in the later analyses.
-This is supported by @Chatfield1985TIEo who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
+This is supported by @Chatfield1985TIEo, who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
 
@@
 The wage anomalies should be mentioned in the introduction so as not to be a surprise at this point in the manuscript.
- --> A first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
-This is followed with graphical summaries using methods available in `ggplot2` [@ggplot2] and `brolgar` [@brolgar].
+ --> The first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
+This is followed by graphical summaries using methods available in `ggplot2` [@ggplot2] and `brolgar` [@brolgar].
 
@@
 A random sample of 36 individuals is chosen (using the `sample_n_keys` function in `brolgar`).
-Their longitudinal profiles are plotted, faceted by `id` and using free $y$ scales, so that the individual patterns can be examined (Figure \ref{fig:sample-plot}).<!--AE comment: "It" implies...
+Their longitudinal profiles are plotted, faceted by `id`, and using free $y$ scales so that the individual patterns can be examined (Figure \ref{fig:sample-plot}).<!--AE comment: "It" implies...
 Please specify what "it" refers to.
@@
 "as shown in Figure 2, which shows that some respondents have high variability in wages, particularly 5, 7, and 11." However, 7 and 11 don't seem to have particularly high variability, especially when compared to 18.
-This whole paragraph needs to be fixed/updated.--> There is a lot of variability from one individual to another, and substantial fluctuation in wages at different times for most individuals.
+This whole paragraph needs to be fixed/updated.--> There is a lot of variability from one individual to another and substantial fluctuation in wages at different times for most individuals.
 Some individuals  (2799, 11041, 11146) are only measured for a short period.
-Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation.
+Some individuals (8296, 9962) possibly have errors in wages in some years because of the extreme fluctuation.
 These need to be inspected more closely.
-It is also important to note that there are some shorter profiles, indicating that some individuals have left the study before it has finished.
-Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis, to account bias induced by the inadvertent censoring.
+It is also important to note that some shorter profiles indicate that some individuals have left the study before it has finished.
+Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis to account for the bias induced by the inadvertent censoring.
 
@@
 Figure \ref{fig:feature-plot}s shows an alternative way to check the data quality.
-Plot (A) is the spagetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to $60,000/hour) for some individuals mostly around 1990.
-Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median and maximum) for all individuals.
+Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to $60,000/hour) for some individuals, mostly around 1990.
+Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median, and maximum) for all individuals.
 This tells us that there are a number of individuals with unbelievably high maximum wages.
 Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to $1200/hour.
-This does not seem to be reasonable, and leads us to use a procedure to detect and fix these temporal anomalies.
+This does not seem to be reasonable and leads us to use a procedure to detect and fix these temporal anomalies.
 
@@
 An alternative approach would be a robust linear mixed model using `robustlmm` [@KollerManuel2016rARP].
-However, when we tested this, the fit failed to convergence.
+However, when we tested this, the fit failed to converge.
 <!--I would pull the tidy and purrr citations to the end of the sentence to make it read more smoothly-->The full code for this is shown at \url{`r yowie_pkgdown`/articles/input-to-valid-data.html} but also described in detail next.
@@
 To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done.
-Samples of individuals were examined under the different thresholds, with an eye to smoothness of the profiles.
+Samples of individuals were examined under the different thresholds, with an eye to the smoothness of the profiles.
 Overly smooth profiles would indicate that the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
-A threshold of 0.12 was chosen, that struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
+A threshold of 0.12 was chosen.
+That struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
 Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.
@@
 Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
-The plot shows that fluctuations in wages remain but the large spikes (in this sample individuals 8296, 9962) which are considered implausible are replaced.
+The plot shows that fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
 
@@
 The highest wage overall is now around $1000.
-Plot (A) shows a more reasonable spagetti plot, where there are some profiles with high wages, but most profiles have wages under $300, and there is a steady increase in wages with years.
+Plot (A) shows a more reasonable spaghetti plot, where there are some profiles with high wages, but most profiles have wages under $300, and there is a steady increase in wages with years.
 Plot (B) shows that there are still a small number of individuals with high maximum wages.
-Plot (C) shows the profile for ID=39, after imputing the extreme value.
+Plot (C) shows the profile for ID=39 after imputing the extreme value.
 The wages for this individual increase over the years, and do fluctuate some between 1900 and 2005.
@@
 "# of individuals" means the number of respondents included in each stage, while "# of observations" means the number of rows in the data.
-The color represents the stage of data cleaning in statistical value chain [@validate].
+The color represents the stage of data cleaning in the statistical value chain [@validate].
 Pink, blue, and green represent the raw, input, and valid data, respectively.
@@
   7. `hgc_1979`: The highest grade completed in 1979 (integer value). 
-  8. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+  8. `ged`: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
 
@@
 
-  11. `ged`: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+  11. `ged`: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
   12. `njobs`: Number of jobs that an individual has.
@@
 The original set, containing wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994, is available in the R package `brolgar`.
-To compare the refreshed data with the original a subset needs to be matched.
-There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data, and restricting the longitudinal measurements to the same years.
+To compare the refreshed data with the original, a subset needs to be matched.
+There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data and restricting the longitudinal measurements to the same years.
 However, we decided to try to replicate the process, as suggested by the description of the original data.
@@
 1. An individual whose highest grade completed (`hgc`) is reported to be less than 12th grade, **or**
-2. An individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and have received a GED (`ged` is code to 2).
+2. An individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and has received a GED (`ged` is code to 2).
 
-An additional criteria from @SingerJudithD2003Alda is to only include males aged between 14 and 17 years old in 1979.
+An additional criterion from @SingerJudithD2003Alda is to only include males aged between 14 and 17 years old in 1979.
 With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
 To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
-This revealed several reasons for for the disparity:
+This revealed several reasons for the disparity:
 
@@
 Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in the subset.
-Our decision is to also include them in the refreshed data as the new data contains an age variable, so analyst could filter them later.
+Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
 2. 79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
@@
     iii. 3 individuals have both diploma and GED (`ged` is coded to 3). These were kept in the refreshed data. 
-    iv. 12 individuals do not exist in `wages` data because they were participated in less than 3 rounds of survey.
+    iv. 12 individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.
 
@@
 This is not done for the refreshed data because we plan to keep refreshing it as the data is added to and released from the survey.
-Instead, we have provided a function in the R package, `r yowie`, for users to conduct the inflation adjustment when they are ready to analyse the data.
+Instead, we have provided a function in the R package, `r yowie`, for users to conduct the inflation adjustment when they are ready to analyze the data.
 
@@
 In the first stage, we showed the steps performed to get the data from the NLSY79 database.
-The data format was converted to tidy format, for more flexibility in cleaning and exploring.
-An initial data analysis was conducted to investigate and screen the quality of the data.
+The data format was converted to a tidy format for more flexibility in cleaning and exploring.
+Initial data analysis was conducted to investigate and screen the quality of the data.
 We found and provided a fix for many anomalous observations in wages using a robust linear regression model.
@@
 
-The data cleaning process is documented and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
+The data cleaning process is documented, and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
 Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a `shiny`  [@shiny] app, and the choices used in the refreshed data are documented.
@@
 
-- determining which records should be downloaded from the database.
-- calculating experience in the workforce requires comparing the date of first job with the first year the individual was recorded, both of which are available in the database. 
-- treating the extreme values since there are many unusually high hourly wages, e.g. greater than $60,000 per hour.
-- determining the dropouts subset as there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with their GED status.
-- matching ID's from the original data with those in the refreshed data, do refer to the same person based on the demographic information available.
+- Determining which records should be downloaded from the database.
+- Calculating experience in the workforce requires comparing the date of the first job with the first year the individual was recorded, both of which are available in the database. 
+- Treating the extreme values since there are many unusually high hourly wages, e.g., greater than $60,000 per hour.
+- Determining the dropouts subset as there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with their GED status.
+- Matching IDs from the original data with those in the refreshed data do refer to the same person based on the demographic information available.
 
-Ultimately, the refreshed data is reasonably similar to the original, but unsatisfactorily far from it.
+Ultimately, the refreshed data is reasonably similar to the original but unsatisfactorily far from it.
 The last step required would be to inflation-adjust wages.
-This is better to do with each wave of new data added, so that it is relative to the last date in the data.
-Our decision was to provide the raw wages, and include code to do the adjustment as part of the package.
+This is better to do with each wave of new data added so that it is relative to the last date in the data.
+Our decision was to provide the raw wages and include codes to make the adjustment as part of the package.
 
@@
 -->
-Some readers may disagree with our decisions made to produce the refreshed textbook data and may have better insight than us in producing a more appropriate textbook data.
+Some readers may disagree with our decisions made to produce the refreshed textbook data and may have better insight than us in producing more appropriate textbook data.
 We do not assert that we have produced the best textbook data, but rather we describe our journey to provide a reasonable textbook data set.
 All code and documentation are provided for transparency.
-Readers could use this to make different decisions, or provide suggestions on through package for better choices.
-Future updates of the `r yowie` package may contain additional variables, or filters of the full set, if it is deemed important.
+Readers could use this to make different decisions or provide suggestions through the package for better choices.
+Future updates of the `r yowie` package may contain additional variables, or filters of the full set if it is deemed important.
 
-For the data providers, we recommend that a better validation system with clear rules applied at data entry, and that alternative output formats, such as a tidy format would help users make better use of their resource.
-The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs.
+For the data providers, we recommend that a better validation system with clear rules applied at data entry and that alternative output formats, such as a tidy format, would help users make better use of their resources.
+The problem with many of the wages records is that there are implausible values or confusion on how to record wages for multiple jobs.
 These values can be validated with simple checks at data entry.
-Providing an open data resource also is accompanied with the responsibility that the data, especially data that is as valuable as this, is reliable.
+Providing an open data resource also is accompanied by the responsibility that the data, especially data that is as valuable as this, is reliable.
 Users need to be able to trust the data.
@@
 Why is this exercise important for teachers and students?
-This work illustrates the steps in cleaning and processing data in the preparation for it to be a textbook data set.
+This work illustrates the steps in cleaning and processing data in preparation for it to be a textbook data set.
 Choices made during the cleaning and processing can affect findings made with the data, and these should be transparent.
-Along with the textbook data, which is now provided as an R package, the documentation of the process provide some examples for teaching data cleaning, initial data analysis and exploratory data analysis.
+Along with the textbook data, which is now provided as an R package, the documentation of the process provides some examples for teaching data cleaning, initial data analysis, and exploratory data analysis.
 The refreshed textbook data provides a resource for teaching longitudinal data analysis.
@@
 Public policy is based on models, yielding averages that might vary across strata.
-Modeling the wages relative to workforce experience, with demographic covariates we learn that there are significantly different patterns.
+Modeling the wages relative to workforce experience, with demographic covariates, we learn that there are significantly different patterns.
 That more education leads to increasingly higher wages, which is a satisfying result for educators.
-It means that education makes a difference in the wage experience, and that public policy that encourages education is a data-supported action.
+It means that education makes a difference in the wage experience and that public policy that encourages education is a data-supported action.
 We would also learn, although not presented here, that race matters and that being black leads to lower wages.
-This is a disturbing finding because there is no rational for such a difference in a fair society.
+This is a disturbing finding because there is no rationale for such a difference in a fair society.
 This would provide support for action in public policy to remove this overall average effect.
@@
 
-On an individual level, one needs to know where am I in this data, and does this data relate to me.
+On an individual level, one needs to know where I am in this data and does this data relate to me.
 To do this, the individual profiles need to be explored.
@@
 For example, many individuals with lower educational attainment earn very high wages.
-From a statistics and data science educator perspective, more focus and more methodology for this type of statistics needs to be included in the curriculum.
+From a statistics and data science educator perspective, more focus and more methodology for this type of statistics need to be included in the curriculum.
 
-The above interpretations of results from analyzing the wages data, rely on trusting that the data provided is accurate and valid.
+The above interpretations of results from analyzing the wages data rely on trusting that the data provided is accurate and valid.
 The wages data is collected by a reputable organization, but we found that the data has obvious errors that should be corrected.
-This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.
+This paper has illustrated procedures and guidelines to achieve valid data and provides the code and details for it to be reproduced and modified if deemed appropriate.
 
 Having trustworthy data is imperative for statistics and data science education.
-Whenever one uses a textbook data set that is perceived as relating to the students lives, there will be interpretations made.
+Whenever one uses a textbook data set that is perceived as relating to the students' lives, there will be interpretations made.
 The wages data is an example of this.
 Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race.
-If one uses data examples that are synthetic, instilled with our own inherent prejudices (for e.g.
-gender and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
-This paper demonstrates the process to produce a trustworthy data set for teaching.
+If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., gender and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
+This paper demonstrates the process of producing a trustworthy data set for teaching.
 

### 2022-02-18 paper/paper.tex
done with final checking.

--- 2022-02-18 paper/paper.tex
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis'', which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
+An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis'', which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 
-The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, higher level of education leads to higher wages, and that race makes a difference, on average.
+The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, a higher level of education leads to higher wages, and that race makes a difference, on average.
 An exploratory analysis reveals, however, that the individual experience varies a great deal from the overall average.
-Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
+Some individuals experience a decline in wages the longer they are in the workforce and many experience volatility in their wages.
 The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at CENSORED.
 
-This disparity between the average and the individual is a part of statistics, as a discipline, that require more attention.
+This disparity between the average and the individual is a part of statistics, as a discipline that requires more attention.
 This particular data set is a prime example of discussing this disparity.
-Textbook data sets have longevity if there is a unresolved mystery.
+Textbook data sets have longevity if there is an unresolved mystery.
 The iris data (Andereson 1935) is a prime example.
 It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new and does not suffer from a connection with eugenics (Stodel 2020).
+(A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020).
 We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
 
-For the field of statistics, and data science by association, it is increasingly important to reach the individual.
-One might describe this as a divergence of purpose, statistics for public policy or statistics for the public.
+For the field of statistics and data science by association, it is increasingly important to reach the individual.
+One might describe this as a divergence of purpose, statistics for public policy, or statistics for the public.
 The two are not the same.
-As the world becomes more electronically connected combating misinformation and mitigating conspiracy theories require that statistics address the individual.
+As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require that statistics address the individual.
 For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic.
 The majority of people in the study do not have a pattern that is similar to the average.
-If you have a bad experience, that your wages have declined over time, you are not alone, there are others like you, and more than you think.
-A similar tone is echoed occasionally in the public media, for example, an article published in the Sydney Morning Herald argues that there is no average Australian (Moncrief 2015).
-
-As a textbook data set though, the wages data is outdated.
+If you have a bad experience, that your wages have declined over time, you are not alone; there are others like you, and more than you think.
+A similar tone is echoed occasionally in the public media.
+For example, an article published in the Sydney Morning Herald argues that there is no average Australian (Moncrief 2015).
+
+As a textbook data set, though, the wages data is outdated.
 The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
@@
 2019).
-The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using a step-wise piping with split-apply-combine strategy for mutating new variables (Wickham 2011).
-Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
-The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
+The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables (Wickham 2011).
+Tidy data shouldn't be confused with ``tame data'', which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
+The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
 
@@
 2001; Cooksey 2017).
-According to Cooksey (2017), thousands of articles, and hundreds of book chapters and monographs have utilized this data.
+According to Cooksey (2017), thousands of articles and hundreds of book chapters and monographs have utilized this data.
 Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al.
@@
 Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).
-We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the \emph{purposes of studying racism, not race} (Fullilove 1998).
-There are a number of data sets provided by Singer and Willett (2003) and we focus only on this one because it has captivated our attention for a number of years.
+We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the \emph{purposes of studying racism, not race} (Fullilove 1998).
+There are a number of data sets provided by Singer and Willett (2003), and we focus only on this one because it has captivated our attention for a number of years.
 We use it in our own teaching of longitudinal data analysis and would very much like to use data since 1994.
@@
 
-There are two ways to conduct the interview of the NLSY79, which are face-to-face or by telephone interviews.
+There are two ways to conduct the interview of the NLSY79, which are face-to-face or telephone interviews.
 In recent survey years, more than 90 percent of respondents were interviewed by telephone (Cooksey 2017).
@@
 \item
-  UERATE: unemployment rate of the year of survey. When missing, the variable is set to be 7.875 (the average rate).
+  UERATE: the unemployment rate of the year of the survey. When missing, the variable is set to be 7.875 (the average rate).
 \end{enumerate}
@@
 \item
-  The wages data of the high school dropouts, which is closest to a refreshed version of Singer and Willett (2003)'s data.
+  The high school dropouts' wages data is closest to a refreshed version of Singer and Willett (2003)'s data.
 \end{enumerate}
@@
 
-van der Loo and de Jonge (2018) in the context of official statistics describe the ``statistical value chain'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organised with correct type and identified variables) and valid data (data that has been cleaned and more accurately represents the intent of variables).
-What we have colorfully named as wild data can be considered to be raw data, and valid data could be considered to be textbook data, in the above statistical value chain.
-In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida}.
+van der Loo and de Jonge (2018), in the context of official statistics, describe the ``statistical value chain'', which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+What we have colorfully named as wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
+In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.
 
@@
 We alter only the file path in \texttt{NLSY79.R} and run the script without any other alteration.
-This results in an initial processing of the raw data into two data sets, \texttt{categories\_qnames} (where the observations are stored in categorical/interval values) and \texttt{new\_data\_qnames} (the observations are stored in integer form).
+This results in the initial processing of the raw data into two data sets, \texttt{categories\_qnames} (where the observations are stored in categorical/interval values) and \texttt{new\_data\_qnames} (the observations are stored in integer form).
 
 According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
-The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organised such that each row corresponds to an individual.
+The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organized such that each row corresponds to an individual.
 As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year.
@@
 
-Thus we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars and the demographic variables.
+Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
 This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
-2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long form, with \texttt{dplyr} (Wickham, François, et al.
+2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham, François, et al.
 2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
+In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
 
@@
 
-\texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981} which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
+The object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 
-The refreshed dataset has three types of highest grade completed.
-The first one is the highest grade completed ever reported in the database (\texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively).
-For each ID, there is only one value of \texttt{hgc} and \texttt{hgc\_i}.
-This variable is obtained from \texttt{new\_data\_qnames} with name \texttt{HGC\_EVER\_XRND} and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
-
-The second one is the highest grade completed in 1979 (\texttt{hgc\_1979}), the first round of the survey.
-We decide to include this variable in the dataset for comparison with the original data-purpose.
-The original data only covers the period until 1994.
-Hence, \texttt{hgc} in the refreshed data, which is the state of \texttt{hgc} in 2018, and the \texttt{hgc} in the original data will not be comparable.
-
-The third one is the highest grade completed corresponds to value reported in each round of the survey (\texttt{grade}), meaning that the value can change.
-We decide to include this variable in the refreshed dataset to enrich the analysis possibly done with the data, for example if one would like to explore how the changes of highest grade completed in individuals affecting their wages.
-The highest grade completed are recorded in \texttt{new\_data\_qnames} as variables beginning with \texttt{Q3-4} and \texttt{HGC} with suffix of the year it was recorded.
-In addition the variables beginning with \texttt{HGCREV} contain the revised data.
-We choose to use this revised data because it seemed to have less missing and presumably has been checked.
-However, there is no revised data for 2012, 2014, 2016, and 2018, thus, we use the ordinary data for these years.
+The next step is processing the highest grade completed.
+There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis.
+The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively.
+For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}.
+This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on).
+
+The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as \texttt{grade}).
+This value can change over time, but it should only increase.
+This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals.
+This is recorded in \texttt{new\_data\_qnames} as columns beginning with \texttt{Q3-4} and \texttt{HGC} with year as a suffix.
+In addition, it is also in the columns beginning with \texttt{HGCREV} reflecting revised data.
+We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked.
+When this value was missing, 2012, 2014, 2016, and 2018, but available in the first form substituted accordingly.
+
+The third definition is the highest grade completed in 1979 (provided in the refreshed data as \texttt{hgc\_1979}), corresponding to the value from the first round of the survey.
+This is calculated from the \texttt{grade} value in 1979.
+This best reflects the grade when the individual left school and is included in order to compare with the original data.
 
@@
 Along with \texttt{hgc}, \texttt{ged} is used to subset the data to the high school dropouts as in the original data.
-The graduate equivalency status is saved as variable started with ``Q3-8A'' followed by the year of the survey.
+The graduate equivalency status is saved as a variable started with ``Q3-8A'' followed by the year of the survey.
 Thus, we only separate the year and the GED status.
-Although there GED status is asked in each round of the survey, we only retain the latest status of one's GED.
+Although the GED status is asked in each round of the survey, we only retain the latest status of one's GED.
 
@@
 From 1988 onward, we selected the total hours worked per week, including hours working from home (\texttt{QES-52D}).
-However, in 1993, this variable was missing for the first and last job so we selected to use \texttt{QES-52A} instead.
+However, in 1993, this variable was missing for the first and last job, so we selected to use \texttt{QES-52A} instead.
 In addition, 2008 only had jobs 1-4 for the \texttt{QES-52D} variable, so we use only these.
@@
 The hourly wages are in the variables beginning with \texttt{HRP} in \texttt{new\_data\_qnames}.
-As a respondent may have multiple jobs, the \texttt{mean\_hourly\_wage} is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on number of hours is available); if number of hours worked for any job is missing, then the \texttt{mean\_hourly\_wage} is computed as a simple average of all available hourly wages.
+As a respondent may have multiple jobs, the \texttt{mean\_hourly\_wage} is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on the number of hours is available); if the number of hours worked for any job is missing, then the \texttt{mean\_hourly\_wage} is computed as a simple average of all available hourly wages.
 Prior to computing the mean hourly wage, we undertook a number of steps to treat unusual observations as described below:
@@
 For \texttt{stwork} variable, we only rename the column names from \texttt{new\_data\_qnames}.
-This variable is then used to calculate the next variable, \texttt{yr\_wforce} for each survey round, which is the year of survey (\texttt{year}) minus the year of individual start working (\texttt{stwork}).
+This variable is then used to calculate the next variable, \texttt{yr\_wforce} for each survey round, which is the year of survey (\texttt{year}) minus the year of individual started working (\texttt{stwork}).
 Finally, \texttt{exp} variable is derived from the number of weeks worked since the last interview indicated as variable started with \texttt{WKSWK} in \texttt{new\_data\_qnames}.
@@
 We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data.
-However, it is worth noting that the original data does not restrict the number of observation for each individual, i.e, there are individuals with only one and two observations.
+However, it is worth noting that the original data does not restrict the number of observations for each individual, i.e., there are individuals with only one and two observations.
 
 Finally, we save the resultant wage data on this cohort as \texttt{wages}.
-Note that we save \texttt{grade} variable in this dataset, instead of \texttt{demog\_nlsy79} dataset because it is a longitudinal variable, while \texttt{demog\_nlsy79} is cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.
+Note that we save \texttt{grade} variable in this dataset instead of \texttt{demog\_nlsy79} dataset because it is a longitudinal variable, while \texttt{demog\_nlsy79} is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.
 
@@
 
-Work experience is one of the most important variables in Singer and Willett (2003) as it indicates time, and makes longitudinal analysis possible.
+Work experience is one of the most important variables in Singer and Willett (2003) as it indicates time and makes longitudinal analysis possible.
 It is desirable to calculate this rather than using the survey year for time because it more accurately reflects a person's time in the workforce.
@@
 It represents the difference between the day an individual enters the labor force (\texttt{EMPLOYERS\_ALL\_STARTDATE\_ORIGINAL.01\textasciitilde{}Y\_XRND} in the database) relative to the date of the survey, which we call \texttt{yr\_wforce}.
-However, using this calculation produced numbers which don't quite match with the original data.
-
-Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide, suggests that it should be calculated based on the variable ``number of weeks worked since last interview''.
+However, using this calculation produced numbers that do not quite match with the original data.
+
+Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide suggests that it should be calculated based on the variable ``number of weeks worked since the last interview''.
 This would remove periods of unemployment which makes sense when measuring experience while actually working.
@@
 According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure that the data is clean, valid, and ready to be deployed in the later analyses.
-This is supported by Chatfield (1985) who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
+This is supported by Chatfield (1985), who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
 
 In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides.
-A first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
-This is followed with graphical summaries using methods available in \texttt{ggplot2} (Wickham 2016) and \texttt{brolgar} (Tierney, Cook, and Prvan 2020).
+The first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
+This is followed by graphical summaries using methods available in \texttt{ggplot2} (Wickham 2016) and \texttt{brolgar} (Tierney, Cook, and Prvan 2020).
 
@@
 A random sample of 36 individuals is chosen (using the \texttt{sample\_n\_keys} function in \texttt{brolgar}).
-Their longitudinal profiles are plotted, faceted by \texttt{id} and using free \(y\) scales, so that the individual patterns can be examined (Figure \ref{fig:sample-plot}).
-There is a lot of variability from one individual to another, and substantial fluctuation in wages at different times for most individuals.
+Their longitudinal profiles are plotted, faceted by \texttt{id}, and using free \(y\) scales so that the individual patterns can be examined (Figure \ref{fig:sample-plot}).
+There is a lot of variability from one individual to another and substantial fluctuation in wages at different times for most individuals.
 Some individuals (2799, 11041, 11146) are only measured for a short period.
-Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation.
+Some individuals (8296, 9962) possibly have errors in wages in some years because of the extreme fluctuation.
 These need to be inspected more closely.
-It is also important to note that there are some shorter profiles, indicating that some individuals have left the study before it has finished.
-Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis, to account bias induced by the inadvertent censoring.
-
-the IDs in panel 9.
+It is also important to note that some shorter profiles indicate that some individuals have left the study before it has finished.
+Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis to account for the bias induced by the inadvertent censoring.
 
 Figure \ref{fig:feature-plot}s shows an alternative way to check the data quality.
-Plot (A) is the spagetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to \$60,000/hour) for some individuals mostly around 1990.
-Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median and maximum) for all individuals.
+Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to \$60,000/hour) for some individuals, mostly around 1990.
+Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median, and maximum) for all individuals.
 This tells us that there are a number of individuals with unbelievably high maximum wages.
 Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to \$1200/hour.
-This does not seem to be reasonable, and leads us to use a procedure to detect and fix these temporal anomalies.
+This does not seem to be reasonable and leads us to use a procedure to detect and fix these temporal anomalies.
 
@@
 An alternative approach would be a robust linear mixed model using \texttt{robustlmm} (Koller 2016).
-However, when we tested this, the fit failed to convergence.
+However, when we tested this, the fit failed to converge.
 The full code for this is shown at \url{CENSORED/articles/input-to-valid-data.html} but also described in detail next.
@@
 The highest wage overall is now around \$1000.
-Plot (A) shows a more reasonable spagetti plot, where there are some profiles with high wages, but most profiles have wages under \$300, and there is a steady increase in wages with years.
+Plot (A) shows a more reasonable spaghetti plot, where there are some profiles with high wages, but most profiles have wages under \$300, and there is a steady increase in wages with years.
 Plot (B) shows that there are still a small number of individuals with high maximum wages.
-Plot (C) shows the profile for ID=39, after imputing the extreme value.
+Plot (C) shows the profile for ID=39 after imputing the extreme value.
 The wages for this individual increase over the years, and do fluctuate some between 1900 and 2005.
@@
 ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data.
-The color represents the stage of data cleaning in statistical value chain (van der Loo and de Jonge 2021).
+The color represents the stage of data cleaning in the statistical value chain (van der Loo and de Jonge 2021).
 Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart-blind}
@@
 \item
-  \texttt{ged}: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
 \end{enumerate}
@@
 \item
-  \texttt{ged}: Wether the respondent had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
 \item
@@
 The original set, containing wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994, is available in the R package \texttt{brolgar}.
-To compare the refreshed data with the original a subset needs to be matched.
-There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data, and restricting the longitudinal measurements to the same years.
+To compare the refreshed data with the original, a subset needs to be matched.
+There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data and restricting the longitudinal measurements to the same years.
 However, we decided to try to replicate the process, as suggested by the description of the original data.
@@
 \item
-  An individual whose highest grade completed (\texttt{hgc}) is reported to be at least 12th grade and have received a GED (\texttt{ged} is code to 2).
+  An individual whose highest grade completed (\texttt{hgc}) is reported to be at least 12th grade and has received a GED (\texttt{ged} is code to 2).
 \end{enumerate}
 
-An additional criteria from Singer and Willett (2003) is to only include males aged between 14 and 17 years old in 1979.
+An additional criterion from Singer and Willett (2003) is to only include males aged between 14 and 17 years old in 1979.
 With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
 To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
-This revealed several reasons for for the disparity:
+This revealed several reasons for the disparity:
 
@@
 Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in the subset.
-Our decision is to also include them in the refreshed data as the new data contains an age variable, so analyst could filter them later.
+Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
 \item
@@
   \item
-    12 individuals do not exist in \texttt{wages} data because they were participated in less than 3 rounds of survey.
+    12 individuals do not exist in \texttt{wages} data because they have participated in less than 3 rounds of the survey.
   \end{enumerate}
@@
 This is not done for the refreshed data because we plan to keep refreshing it as the data is added to and released from the survey.
-Instead, we have provided a function in the R package, \texttt{[CENSORED]}, for users to conduct the inflation adjustment when they are ready to analyse the data.
+Instead, we have provided a function in the R package, \texttt{[CENSORED]}, for users to conduct the inflation adjustment when they are ready to analyze the data.
 
@@
 
-The data cleaning process is documented and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
+The data cleaning process is documented, and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
 Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a \texttt{shiny} (Chang et al.
@@
 \item
-  determining which records should be downloaded from the database.
-\item
-  calculating experience in the workforce requires comparing the date of first job with the first year the individual was recorded, both of which are available in the database.
-\item
-  treating the extreme values since there are many unusually high hourly wages, e.g.~greater than \$60,000 per hour.
-\item
-  determining the dropouts subset as there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with their GED status.
-\item
-  matching ID's from the original data with those in the refreshed data, do refer to the same person based on the demographic information available.
-\end{itemize}
-
-Ultimately, the refreshed data is reasonably similar to the original, but unsatisfactorily far from it.
+  Determining which records should be downloaded from the database.
+\item
+  Calculating experience in the workforce requires comparing the date of the first job with the first year the individual was recorded, both of which are available in the database.
+\item
+  Treating the extreme values since there are many unusually high hourly wages, e.g., greater than \$60,000 per hour.
+\item
+  Determining the dropouts subset as there is no explicit variable in the database recording high school dropout, which means we needed to compare the date of 12th grade with their GED status.
+\item
+  Matching IDs from the original data with those in the refreshed data do refer to the same person based on the demographic information available.
+\end{itemize}
+
+Ultimately, the refreshed data is reasonably similar to the original but unsatisfactorily far from it.
 The last step required would be to inflation-adjust wages.
-This is better to do with each wave of new data added, so that it is relative to the last date in the data.
-Our decision was to provide the raw wages, and include code to do the adjustment as part of the package.
-
-Some readers may disagree with our decisions made to produce the refreshed textbook data and may have better insight than us in producing a more appropriate textbook data.
+This is better to do with each wave of new data added so that it is relative to the last date in the data.
+Our decision was to provide the raw wages and include codes to make the adjustment as part of the package.
+
+Some readers may disagree with our decisions made to produce the refreshed textbook data and may have better insight than us in producing more appropriate textbook data.
 We do not assert that we have produced the best textbook data, but rather we describe our journey to provide a reasonable textbook data set.
 All code and documentation are provided for transparency.
-Readers could use this to make different decisions, or provide suggestions on through package for better choices.
-Future updates of the \texttt{[CENSORED]} package may contain additional variables, or filters of the full set, if it is deemed important.
-
-For the data providers, we recommend that a better validation system with clear rules applied at data entry, and that alternative output formats, such as a tidy format would help users make better use of their resource.
-The problem with many of the wages records is that there are implausible values, or confusion on how to record wages for multiple jobs.
+Readers could use this to make different decisions or provide suggestions through the package for better choices.
+Future updates of the \texttt{[CENSORED]} package may contain additional variables, or filters of the full set if it is deemed important.
+
+For the data providers, we recommend that a better validation system with clear rules applied at data entry and that alternative output formats, such as a tidy format, would help users make better use of their resources.
+The problem with many of the wages records is that there are implausible values or confusion on how to record wages for multiple jobs.
 These values can be validated with simple checks at data entry.
-Providing an open data resource also is accompanied with the responsibility that the data, especially data that is as valuable as this, is reliable.
+Providing an open data resource also is accompanied by the responsibility that the data, especially data that is as valuable as this, is reliable.
 Users need to be able to trust the data.
@@
 For example, many individuals with lower educational attainment earn very high wages.
-From a statistics and data science educator perspective, more focus and more methodology for this type of statistics needs to be included in the curriculum.
-
-The above interpretations of results from analyzing the wages data, rely on trusting that the data provided is accurate and valid.
+From a statistics and data science educator perspective, more focus and more methodology for this type of statistics need to be included in the curriculum.
+
+The above interpretations of results from analyzing the wages data rely on trusting that the data provided is accurate and valid.
 The wages data is collected by a reputable organization, but we found that the data has obvious errors that should be corrected.
-This paper has illustrated procedures and guidelines to achieve valid data, and provides the code and details for it to be reproduced, and modified if deemed appropriate.
+This paper has illustrated procedures and guidelines to achieve valid data and provides the code and details for it to be reproduced and modified if deemed appropriate.
 
 Having trustworthy data is imperative for statistics and data science education.
-Whenever one uses a textbook data set that is perceived as relating to the students lives, there will be interpretations made.
+Whenever one uses a textbook data set that is perceived as relating to the students' lives, there will be interpretations made.
 The wages data is an example of this.
 Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race.
-If one uses data examples that are synthetic, instilled with our own inherent prejudices (for e.g.~gender and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
-This paper demonstrates the process to produce a trustworthy data set for teaching.
+If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., gender and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
+This paper demonstrates the process of producing a trustworthy data set for teaching.
 

### 2022-04-25 paper/cover-letter/cover-letter.Rmd
Add all of reviewers' comments in response file.

---
from: Dewi Amaliah
fromqualifications: Master of Business Analytics
fromWWW: "[https://numbats.github.io/yowie/](numbats.github.io/yowie/)"
fromemail: dlamaleeah@gmail.com
fromphone: "+61 435 444 025"
fromsignature: sigfile.png
toname: Prof. Nick Horton
toaddress: Editor, JSDSE 
opening: Dear Prof. Horton
closing: Regards
spacing: 1.4
output: MonashEBSTemplates::letter
---

Please consider the revised version of our paper titled "The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth" for the JDSDE Special Issue on "Teaching Reproducibility and Responsible Workflow".

The paper is co-authored by Prof. Di Cook, Dr Emi Tanaka, Kate Hyde and Dr Nick Tierney. We believe it is a good fit for the special issue because it covers various aspects of data management and curation component of data acumen, including data preparation and cleaning, missing and conflicting data. The code to reproduce the data cleaning, and make further updates of the resulting data is provided.

We also provide the response of editor and reviewers' comments. We also agree if you share this paper with the discussant for this special issue.

### 2022-04-25 paper/paper-jsdse/paper-jsdse.tex
Add all of reviewers' comments in response file.

--- 2022-02-16 paper/paper-jsdse/paper-jsdse.tex
@@
 % NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{1}
+\newcommand{\blind}{0}
 
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this, is the wages data made public by Singer and Willett (2003) in their book, Applied longitudinal data analysis: Modeling change and event occurrence, which can be used to teach generalized linear models, in addition to hierarchical, mixed effects and multilevel models.
+An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis'', which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models.
 The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 
-Our process of cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
+Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
 2019).
-The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using the data pipeline and split-apply-combine approach (Wickham 2011).
-Tidy data shouldn't be confused with ``tame data'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
-The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]} which includes the code so that the process is reproducible, and could be used to further refresh the data as new records are made available in the NLSY79 database.
+The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables (Wickham 2011).
+Tidy data shouldn't be confused with ``tame data'', which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
+The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
 
@@
 
-The book ``Applied longitudinal data analysis'' (Singer and Willett 2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
+Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
 This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al.
@@
 2001; Cooksey 2017).
-According to Cooksey (2017), thousands of articles, and hundreds of book chapters and monographs have utilized this data.
+According to Cooksey (2017), thousands of articles and hundreds of book chapters and monographs have utilized this data.
 Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al.
 2001).
-Our aim is to refresh this textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
+
+Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
 Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).
-We should also note that race is a variable in the original data set, and for compatibility it is also provided with the refreshed data, for the \emph{purposes of studying racism, not race} (Fullilove 1998).
+We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the \emph{purposes of studying racism, not race} (Fullilove 1998).
+There are a number of data sets provided by Singer and Willett (2003), and we focus only on this one because it has captivated our attention for a number of years.
+We use it in our own teaching of longitudinal data analysis and would very much like to use data since 1994.
 
@@
 For example, use a single categorical race variable instead of the two binary race variables.
-We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with diploma or with Graduate Equivalency Degree (GED), the number of jobs that the individual had in the corresponding year of survey, total number of hours the individual usually works per week, the year when individual started to work, and the number of years the individual worked.
+We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs that the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked.
 We do not attempt to re-create the unemployment rate variable.
@@
 \begin{itemize}
-\item[\faCloudDownload] To download all 663 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download".
-We select the R Source code and Comma-delimited datafile of selected variables with Reference Number as column headers.
+\item[\faCloudDownload] To download all 742 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download".
+We select the R Source code and comma-delimited datafile of selected variables with Reference Number as column headers.
 We name the filename "NLSY79" and press the download button.
@@
 We alter only the file path in \texttt{NLSY79.R} and run the script without any other alteration.
-This results in an initial processing of the raw data into two data sets, \texttt{categories\_qnames} (where the observations are stored in categorical/interval values) and \texttt{new\_data\_qnames} (the observations are stored in integer form).
+This results in the initial processing of the raw data into two data sets, \texttt{categories\_qnames} (where the observations are stored in categorical/interval values) and \texttt{new\_data\_qnames} (the observations are stored in integer form).
 
 According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
-The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organised such that each row corresponds to an individual.
+The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organized such that each row corresponds to an individual.
 As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year.
-The raw data consequently has a large number of columns (742 to be specific).
+The raw data consequently has a large number of columns (770 to be specific).
 The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars.
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed (\texttt{hgc}), highest grade completed in terms of years, e.g.~9th grade = 9, 3rd year college = 15, (\texttt{hgc\_i}), and whether the graduate equivalency diploma is obtained (\texttt{ged}).
-
-For gender and race, we only rename the column names.
-It is worth noting that using these two variable needs special attention.
-Gender, as reported in the data, only has two categories, which is recognised today as inadequate.
-Gender is not binary.
-Further, race is as reported in the database.
-When doing analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
-
-The raw data, \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981} which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
+In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
+
+For \texttt{gender} and \texttt{race}, we only rename the column names provided in the raw data, \texttt{new\_data\_qnames}.
+It is worth noting that \texttt{gender} comes from a variable called \texttt{sex} in the database.
+We use the term \texttt{gender} as the reports about the data on various pages in the website more commonly use \texttt{gender}.
+Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female''.
+The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
+Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification (Kennedy et al.
+2020).
+From a statistical perspective, this sometimes difficult the adjustment of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex (Kennedy et al.
+2020).
+Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
+This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
+Similarly, this also relates to race, as reported in the database.
+When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
+
+The object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 
-The next step is tidying to obtain \texttt{hgc} and \texttt{ged}.
-Besides that \texttt{hgc} is available in the original data, along with \texttt{ged}, it is used to subset the data to the high school dropouts as in the original data.
-As the highest grade completed is already exist in \texttt{new\_data\_qnames} and stored in year unit (e.g., 10, 11, 12, 13, and so on), we only rename the column name to be \texttt{hgc\_i} and recode it so we have the factor data type (e.g., 10th grade, 11th grade, and so on).
-For \texttt{ged}, we only separate the year of survey and the GED status of the individual.
+The next step is processing the highest grade completed.
+There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis.
+The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively.
+For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}.
+This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on).
+
+The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as \texttt{grade}).
+This value can change over time, but it should only increase.
+This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals.
+This is recorded in \texttt{new\_data\_qnames} as columns beginning with \texttt{Q3-4} and \texttt{HGC} with year as a suffix.
+In addition, it is also in the columns beginning with \texttt{HGCREV} reflecting revised data.
+We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked.
+When this value was missing, 2012, 2014, 2016, and 2018, but available in the first form substituted accordingly.
+
+The third definition is the highest grade completed in 1979 (provided in the refreshed data as \texttt{hgc\_1979}), corresponding to the value from the first round of the survey.
+This is calculated from the \texttt{grade} value in 1979.
+This best reflects the grade when the individual left school and is included in order to compare with the original data.
+
+The next step is tidying to obtain \texttt{ged}.
+Along with \texttt{hgc}, \texttt{ged} is used to subset the data to the high school dropouts as in the original data.
+The graduate equivalency status is saved as a variable started with ``Q3-8A'' followed by the year of the survey.
+Thus, we only separate the year and the GED status.
+Although the GED status is asked in each round of the survey, we only retain the latest status of one's GED.
 
@@
 We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data.
-However, it is worth noting that the original data does not restrict the number of observation for each individual, i.e, there are individuals with only one and two observations.
+However, it is worth noting that the original data does not restrict the number of observations for each individual, i.e., there are individuals with only one and two observations.
 
 Finally, we save the resultant wage data on this cohort as \texttt{wages}.
+Note that we save \texttt{grade} variable in this dataset instead of \texttt{demog\_nlsy79} dataset because it is a longitudinal variable, while \texttt{demog\_nlsy79} is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.
 
@@
 It represents the difference between the day an individual enters the labor force (\texttt{EMPLOYERS\_ALL\_STARTDATE\_ORIGINAL.01\textasciitilde{}Y\_XRND} in the database) relative to the date of the survey, which we call \texttt{yr\_wforce}.
-However, using this calculation produced numbers which don't quite match with the original data.
-
-Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide, suggests that it should be calculated based on the variable ``number of weeks worked since last interview''.
+However, using this calculation produced numbers that do not quite match with the original data.
+
+Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide suggests that it should be calculated based on the variable ``number of weeks worked since the last interview''.
 This would remove periods of unemployment which makes sense when measuring experience while actually working.
 Since it is only measured since the last interview, this needs to be cumulated for each survey year.
-This produced results more similar to the original data (see Subsection \ref{takeaways}).
+This produced results more similar to the original data (as discussed further in Section \ref{takeaways}).
 
@@
 
-\caption{Longitudinal profiles of wages for a random sample of 20 individuals in the refreshed data.
-Most, but not all, individuals here experienced increasing wages over time, and several have experienced considerable fluctuation in wages.
-Some individuals are only measured for a short period.}\label{fig:sample-plot}
+\caption{Longitudinal profiles of wages for a random sample of 36 individuals in the pre-cleaned data.
+There is considerable variation in wages.
+Some individuals  (2799, 11041, 11146) are only measured for a short period.
+Some individuals (8296, 9962) possibly have errors in wages in some years, because of the extreme fluctuation.}\label{fig:sample-plot}
 \end{figure}
 
-We randomly take 20 samples from the data and plot them, as shown in Figure \ref{fig:sample-plot}.
-It shows that these respondents have a lot of variability in wages, for example, the IDs in panel numbers 5, 7, and 11.
-The plot also implies that the samples have a different pattern of mean hourly wages.
-Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.
-However, when checking the summary plots (Figure \ref{fig:feature-plot}), we found that some observations had exceptionally high wages, the maximum value of the wages unbelievably reached about \$60,000/hour.
-Besides, some of them, for example respondents in Figure \ref {fig:feature-plot} C, had experienced unusual wages only in certain year.
+In the next step, we explore the mean hourly wage data of samples of individuals.
+The purpose is to examine the common patterns and check the quality.
+A random sample of 36 individuals is chosen (using the \texttt{sample\_n\_keys} function in \texttt{brolgar}).
+Their longitudinal profiles are plotted, faceted by \texttt{id}, and using free \(y\) scales so that the individual patterns can be examined (Figure \ref{fig:sample-plot}).
+There is a lot of variability from one individual to another and substantial fluctuation in wages at different times for most individuals.
+Some individuals (2799, 11041, 11146) are only measured for a short period.
+Some individuals (8296, 9962) possibly have errors in wages in some years because of the extreme fluctuation.
+These need to be inspected more closely.
+It is also important to note that some shorter profiles indicate that some individuals have left the study before it has finished.
+Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis to account for the bias induced by the inadvertent censoring.
+
+Figure \ref{fig:feature-plot}s shows an alternative way to check the data quality.
+Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to \$60,000/hour) for some individuals, mostly around 1990.
+Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median, and maximum) for all individuals.
+This tells us that there are a number of individuals with unbelievably high maximum wages.
+Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to \$1200/hour.
+This does not seem to be reasonable and leads us to use a procedure to detect and fix these temporal anomalies.
 
@@
 
-To treat the extreme values in the data, we build a robust linear regression model where we use the robustness weight to determine if a value should be replaced with the fitted value from the model.
-We build the model for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively.
-This fits a model separately for each individual, and a robust linear mixed model could be used as an alternative approach.
-We tried fitting a robust linear mixed model using \texttt{robustlmm} (Koller 2016), however perhaps because the data is large and imbalanced, the model was not converging.
+A robust linear regression model using the \texttt{rlm} function from \texttt{MASS} package (Venables and Ripley 2002) is used to treat the extreme values in the data.
+The robustness weight is used to determine if a value should be replaced with the fitted value from the model.
+This is constructed for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively.
+An alternative approach would be a robust linear mixed model using \texttt{robustlmm} (Koller 2016).
+However, when we tested this, the fit failed to converge.
 The full code for this is shown at \url{CENSORED/articles/input-to-valid-data.html} but also described in detail next.
 
-We build the model using the \texttt{rlm} function from \texttt{MASS} package (Venables and Ripley 2002).
-We set the \texttt{mean\_hourly\_wage} and \texttt{year} as the dependent and predictor, respectively.
-Furthermore, we use M-Estimation with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021).
+The \texttt{mean\_hourly\_wage} and \texttt{year} are set as the dependent and predictor, respectively.
+Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021).
 The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
-Moreover, it should be noted that not all the outliers are due to an error.
-Instead, it might be that one had reasonably increasing or decreasing wages in a particular period.
-
-To minimize the risk of mistakenly identifying an outlier as an ``erroneous outlier'', we simulate some thresholds and study how they affect the data by eye balling the results.
-Based on this eye balling, we find that 0.12 is the most reasonable value to use as the threshold to minimize the risk of overly smoothing the wage by retaining sensible spikes in the data.
-In other words, we keep maintaining the natural variability of the wages while minimizing anomalies because of the error in the data recording.
-After deciding the threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.
-We then flag those observations in a new variable called \texttt{is\_pred}.
+It should be noted that it is not possible to determine if all outliers are errors, and it might be that an individual had abnormally high wages at a particular time.
+
+To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done.
+Samples of individuals were examined under the different thresholds, with an eye to the smoothness of the profiles.
+Overly smooth profiles would indicate that the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
+A threshold of 0.12 was chosen.
+That struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
+Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.
+We then flag those observations in a new variable called \texttt{is\_pred}, so that this change can be monitored in the downstream analyses.
 
 Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
-The figure shows that fluctuations in wages still exist even in the imputed data.
-However, the large spikes, which are considered ``erroneous outliers'', are already eliminated from the data.
-Hence, the model produces a data set with a more reasonable degree of fluctuation.
+The plot shows that fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
 
@@
 
-\caption{Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for a sample of individuals.
-A robust linear model prediction was used to correct mean hourly wages value.
-We can see that some extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years, but otherwise the profiles are not changed.
-Some spikes might remain when wage vaues are plausible.}\label{fig:compare-plot}
+\caption{Comparison between the original (black dots) and the corrected (solid grey) mean hourly wage for same sample of individuals as shown in Figure 2.
+A robust linear model prediction was used to identify and correct mean hourly wages value.
+The extreme spikes, corresponding to implausible wages, have been replaced with values more similar to wages in neighboring years for individuals 8296 and 9962, but otherwise the profiles have not changed.}\label{fig:compare-plot}
 \end{figure}
 
-Further, Figure \ref{fig:fixed-feature-plot} A shows that after eliminating the extreme values, the highest value has decreased to be around \$350.
-The spikes are still observed but not as extreme as the original data set.
-In Figure \ref{fig:fixed-feature-plot} B, we plot the three features of mean hourly wages, namely the minimum, median, and maximum value.
-We still can see some extreme values in maximum wages, but consider it as a natural variability of the data.
-In Figure \ref{fig:fixed-feature-plot} C, we can see that after imputing the extreme value in ID=39, we can see how the the wages change over the years more clearly.
+Figure \ref{fig:fixed-feature-plot} shows the summary statistics after removing extremes.
+The highest wage overall is now around \$1000.
+Plot (A) shows a more reasonable spaghetti plot, where there are some profiles with high wages, but most profiles have wages under \$300, and there is a steady increase in wages with years.
+Plot (B) shows that there are still a small number of individuals with high maximum wages.
+Plot (C) shows the profile for ID=39 after imputing the extreme value.
+The wages for this individual increase over the years, and do fluctuate some between 1900 and 2005.
 
@@
 
-Finally, we save the imputed data and set the appropriate data type for the variables.
-
 \hypertarget{recap}{%
@@
 
-Figure \ref{fig:flow-chart-blind} summarizes the steps taken to go from raw to input to valid data (van der Loo and de Jonge 2018) to create a refreshed wages data set.
+There are many steps and decisions made to go from raw to input to valid data. Figure \ref{fig:flow-chart-blind} summarizes these in order to create a refreshed wages data set.
 
@@
 
-\hypertarget{compare}{%
-\section{Comparison of refreshed with the original data}\label{compare}}
-
-\begin{verbatim}
-#> # A tibble: 2 x 3
-#>   hgc   hgc12     n
-#>   <chr> <chr> <int>
-#> 1 12TH  8TH       1
-#> 2 8TH   12TH    243
-\end{verbatim}
-
-\begin{figure}
-
-{\centering \includegraphics[width=0.7\linewidth]{figures/compare-xp-sw-1}
-
-}
-
-\caption{Males dropouts' work experience in the original and refreshed data.
-Each point represents the work experience of individuals (identified by the ID) in the corresponding survey period.
-We can see that the correlation is reasonably high between them even there are different values for many individuals.}\label{fig:compare-xp-sw}
-\end{figure}
-
-The original set contains wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994.
-The refreshed data set for comparison is also wages on high school dropouts from 1979 to 2018.
-The original set is available in the R package \texttt{brolgar} and the refreshed data is available in the R package \texttt{[CENSORED]}.
-
-\hypertarget{filtering-determining-who-is-a-dropout}{%
-\subsection{Filtering: Determining who is a dropout}\label{filtering-determining-who-is-a-dropout}}
-
-There is no explicit explanation of how the dropouts cohort is determined in the original data. Hence, we use the high school dropouts criteria from Wolpin (2005), which are:
+The list of variables provided in the three new datasets are as follows:
+
+\textbf{\texttt{demog\_nlsy79}} :
 
@@
 \item
-  An individual whose highest grade completed (\texttt{hgc}) is reported to be less than 12th grade, \textbf{or}
-\item
-  An individual whose highest grade completed (\texttt{hgc}) is reported to be at least 12th grade and have received a GED (\texttt{ged} is code to 2).
+  \texttt{id}: A unique individual's ID number.
+\item
+  \texttt{age\_1979}: The age of the individual in 1979.
+\item
+  \texttt{gender}: Gender of the individual, FEMALE and MALE.
+\item
+  \texttt{race} : Race of the individual, NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
+\item
+  \texttt{hgc}: The highest grade completed ever.
+\item
+  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
+\item
+  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
+\item
+  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
 \end{enumerate}
 
-An additional criteria from Singer and Willett (2003) is to only include males aged between 14 and 17 years old in 1979. Filters using these criteria used re-create the dropouts data.
-
-With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
-To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
-This revealed several reasons for for the disparity:
+\textbf{\texttt{wages}} and its subset for high school dropouts cohort \textbf{\texttt{wages\_hs\_do}}:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\item
+  \texttt{id}: A unique individual's ID number. This is the \texttt{key} of the data as we saved the data as a \texttt{tsibble} object.
+\item
+  \texttt{year}: The year the observation was taken. This is the \texttt{index} of the data.
+\item
+  \texttt{wage}: The mean of the hourly wages the individual gets at each of their different jobs.
+The value could be a weighted or an arithmetic mean.
+The weighted mean is used when the information of hours of work as the weight is available.
+The mean hourly wage could also be a predicted value if the original value is considered influential by the robust linear regression as part of data cleaning.
+\item
+  \texttt{age\_1979}: The age of the individual in 1979.
+\item
+  \texttt{gender}: Gender of the individual: FEMALE and MALE.
+\item
+  \texttt{race}: Race where the individual belongs to: NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
+\item
+  \texttt{grade}: Integer value of the highest grade completed corresponding to \texttt{year}.
+\item
+  \texttt{hgc}: The highest grade completed ever.
+\item
+  \texttt{hgc\_i}: Integer value of the highest grade completed ever.
+\item
+  \texttt{hgc\_1979}: The highest grade completed in 1979 (integer value).
+\item
+  \texttt{ged}: Whether the individual had a high school diploma or Graduate Equivalency Degree (GED). 1: High school diploma; 2: GED; 3: Both.
+\item
+  \texttt{njobs}: Number of jobs that an individual has.
+\item
+  \texttt{hours:}: The total number of hours the individual usually works per week.
+\item
+  \texttt{stwork}: The year when the individual starting to work.
+\item
+  \texttt{yr\_wforce}: The length of time in the workforce in years (\texttt{year} - \texttt{stwork}).
+\item
+  \texttt{exp}: Work experience, i.e., the number of years of working.
+\item
+  \texttt{is\_wm}: Whether the mean hourly wage is weighted mean, using the hour work as the weight, or regular/arithmetic mean. TRUE = is weighted mean. FALSE = is regular mean.
+\item
+  \texttt{is\_pred}: Whether the mean hourly wage is a predicted value of RLM or not.
+\end{enumerate}
+
+\hypertarget{compare}{%
+\section{Comparison of refreshed with the original data}\label{compare}}
+
+The original set, containing wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994, is available in the R package \texttt{brolgar}.
+To compare the refreshed data with the original, a subset needs to be matched.
+There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data and restricting the longitudinal measurements to the same years.
+However, we decided to try to replicate the process, as suggested by the description of the original data.
+This requires first identifying individuals who dropped out of high school.
+
+\hypertarget{filtering-determining-who-is-a-dropout}{%
+\subsection{Filtering: Determining who is a dropout}\label{filtering-determining-who-is-a-dropout}}
+
+There is no explicit explanation of how the dropouts cohort is determined in the original data. Hence, we use the high school dropouts criteria from Wolpin (2005), which are:
 
@@
 \item
+  An individual whose highest grade completed (\texttt{hgc}) is reported to be less than 12th grade, \textbf{or}
+\item
+  An individual whose highest grade completed (\texttt{hgc}) is reported to be at least 12th grade and has received a GED (\texttt{ged} is code to 2).
+\end{enumerate}
+
+An additional criterion from Singer and Willett (2003) is to only include males aged between 14 and 17 years old in 1979.
+With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
+To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
+This revealed several reasons for the disparity:
+
+\begin{enumerate}
+\def\labelenumi{\arabic{enumi}.}
+\tightlist
+\item
   173 individuals were more than 17 years old in 1979.
 Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in the subset.
-Our decision is to also include them in the refreshed data as the new data contains an age variable, so analyst could filter them later.
+Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
 \item
   79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:
-\end{enumerate}
-
-\begin{itemize}
+
+  \begin{enumerate}
+  \def\labelenumii{\roman{enumii}.}
   \tightlist
@@
   \item
-  12 individuals do not exist in \texttt{wages} data because they were participated in less than 3 rounds of survey.
-\end{itemize}
-
-The filtering was re-applied using these decisions. The refreshed dropout data has 863 individuals.
+    12 individuals do not exist in \texttt{wages} data because they have participated in less than 3 rounds of the survey.
+  \end{enumerate}
+\end{enumerate}
+
+The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing 863 individuals.
+
+\hypertarget{summaries-of-original-with-refreshed-dropouts-data}{%
+\subsection{Summaries of original with refreshed dropouts data}\label{summaries-of-original-with-refreshed-dropouts-data}}
+
+Because the original data does not have the year of collection, it is not possible to merge the two subsets directly.
+Merging longitudinal data requires both the key (\texttt{id}) and the index (ideally survey year).
+In the original data, the experience variable is the time index, and it was not possible to exactly match this for the refreshed data.
+Thus, comparisons of the two sets have to be conducted in a two-sample fashion rather than a matched sample.
+
+Figure \ref{fig:compare-subsets} contains summaries of corresponding variables in the two subsets.
+Plot (A) shows a back-back bar chart of the highest grade completed.
+The two sets are almost the same, but small differences remain.
+Plots (B) and (C) show stacked density plots of experience and log wages, respectively.
+The distributions are relatively close, with more differences in wages as would be expected because the refreshed data is not inflation-adjusted.
 
@@
 
-{\centering \includegraphics[width=1\linewidth]{figures/plotting-sw-do-1}
+{\centering \includegraphics[width=1\linewidth]{figures/compare-subsets-1}
 
@@
 
-\caption{Comparison of original textbook example (A) with refreshed data (B).
-The original data was inflation-adjusted to 1990 prices and the individual's time of collection was converted to a length of experience in the workforce, which makes it difficult to precisely compare the two sets.}\label{fig:plotting-sw-do}
+\caption{Comparison of original and refreshed data: (A) highest grade completed, (B) experience and (C) log wages.
+Some difference in wages would be expected because the refreshed data is not inflation-adjusted, but the two sets are reasonably similar.}\label{fig:compare-subsets}
 \end{figure}
@@
 
-There are two aspects of the original data that make the replication with exactly the same result is not obtained.
-The first one is that the calculation of the time variable provided, which is the ``work experience'', is not clearly articulated.
-Singer and Willett (2003) mention that the temporal variable is the years of experience since the first day of work.
-However, this exact same variable is not explicitly available in the database.
+There are several aspects of the original data that were difficult to replicate.
+The calculation of the work experience is not clearly articulated.
+Singer and Willett (2003) describe the temporal variable as the years of experience since the first day of work.
+This variable is not explicitly available in the database.
 From the NLSY79 topical guide (Bureau of Labor Statistics, U.S.
@@
 One of them is the weeks of worked since the last interview.
-Hence, we use this to calculate the variable.
-Even though we got different result, the correlation between the original and refreshed work experience is reasonably high as shown in Figure \ref{fig:compare-xp-sw}.
-
-Secondly, in the original dataset, wages were inflation-adjusted to 1990 prices.
-This is not done for the refreshed data because we plan to keep refreshing it as the data from the newest survey period release.
-Thus, the 1990 prices might not be relevant anymore for the future context.
-Hence, we keep it as the original value to allow the users to treat it as their needs.
-
-It is also worth noting that the treatment of unlikely wages differ in the refreshed data.
+This is used to calculate the variable.
+It produces reasonably similar but not exactly the same values.
+Because the original data set did not include the year of the survey, it cannot be precisely compared.
+
+The highest grade completed has some confusion.
+There are several ways that this is reported, including the highest grade ever completed and also the \texttt{hgc} at each survey year.
+To match the original data, it is appropriate to use the \texttt{hgc} while in high school.
+The documentation suggests that this variable is available, but it is not actually present in the data.
+Hence, for matching the original data, we have calculated the \texttt{hgc} to match the \texttt{hgc} achieved in the years between 1979 and 1994 based on the yearly survey value.
+The result does not exactly match with the original for a few individuals.
+
+In the original dataset, wages were inflation-adjusted to 1990 prices.
+This is not done for the refreshed data because we plan to keep refreshing it as the data is added to and released from the survey.
+Instead, we have provided a function in the R package, \texttt{[CENSORED]}, for users to conduct the inflation adjustment when they are ready to analyze the data.
+
+It is important to note that the treatment of unlikely wages differs in the refreshed data.
 In the original data by Singer and Willett (2003), wages greater than \$75 are set to be missing.
-However, in recent context, this value is too low to be set as the maximum threshold.
-We opted to use the weights from a robust linear regression to determine what should be regarded as outliers and imputed them with their predicted values as as described in Section \ref{censor}.
-
-Figure \ref{fig:plotting-sw-do} shows a comparison of the two sets.
-(A direct ID matching is not possible.) There are 888 individuals in the original and 1,188 individuals in the refreshed data.
-This suggests some individuals were removed in the original data.
-In addition, in the original set the racial breakdown was 246 Black, 204 Hispanic and 438 White participants, while in the refreshed data there are 346 Black, 219 Hispanic, and 623 White participants, so the proportions have not been exactly reproduced.
-On education, in the refreshed data there are very few individuals with less than a 12th grade education, which is different from the original.
-In the original data there were 366 individuals with up to 8th grade and 522 9th-12th grade, as compared to 967 with 12th grade in the refreshed data.
+However, this value is too low to be set as the maximum threshold, and it doesn't take into account temporal neighbors for an individual.
+We opted to use the weights from a robust linear regression to determine what should be regarded as extreme and imputed them with their predicted values as described in Section \ref{censor}.
+
+Lastly, the race variable is almost perfectly matched. The \texttt{id}'s in the dropouts subset of the refreshed data all correspond to males, which perfectly matches the original subset.
 
@@
 
+\leavevmode\hypertarget{ref-kennedy2020using}{}%
+Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. ``Using Sex and Gender in Survey Adjustment.'' \url{http://arxiv.org/abs/2009.14401}.
+
 \leavevmode\hypertarget{ref-tamedata}{}%

### 2022-04-25 response/response2.Rmd
Add all of reviewers' comments in response file.

---
title: Response to reviewers
author: ""
fontsize: 12pt
date: "Submission ID 217815520.R1"
output: monash::memo
# if you don't have the monash package install by:
# remotes::install_github("numbats/monash")
---

We thank the Editor, Associate Editor and the three reviewers for their extensive comments, which have helped substantially to improve the manuscript. In response, we have made revisions of the main manuscript as outlined below.

\section*{Major changes:}

The editor mention that while the data are interesting, the necessary details for a pedagogical application or case study are still largely missing. The manuscript reads more like a paper about the R package, something for a software journal.

The editor agree with the general sentiment among two of three reviewers that this paper is improved over the previous version, but still does not make clear contributions to educators other than its use in class. If this was (1) a general process for demonstrating the cleaning of longitudinal data, (2) a case study for cleaning data in the classroom, or (3) a case study for using the data in the classroom the value would be clear. In its current state, however, the manuscript still partly each of the three by describing the R package they’ve created and, as a result, does not  fully address any of the three.

\begin{description}
\item[Associate Editor:] \textcolor{violet}{\textbf{The author switches between gender (man/woman/etc) and sex (male/female/etc.) throughout the manuscript. Further, the explanation here covers important ideas in teaching how to use data on gender/sex and race, but it’s just the beginning of ideas. It might be better to discuss the questions on the survey and discuss possible shortcomings, rather than make a somewhat empty gesture towards today’s standard.} The authors still incorrectly refer to male/female as gender. While it is clear the authors are trying to be thoughtful in how they address important issues with data collection on demographics they are not correctly doing so (see note about race/ethnicity below). Certainly there are potential issues with binary sex variables, this can and should be discussed but care should go into this. Please review the literature surrounding the discussion of sex and gender in data (e.g., SAGER guidelines).}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{\textbf{The discussion around highest grade completed is mechanical in what was done, but lacks substantive reasoning for those decisions and possible downstream con- sequences.} It is still unclear to me whether education is measured by year or by category (it’s described both ways) and, in either case, the decision for one or the other isn’t contrasted or justified. For example `hcg` is described as a factor and `grade` is described as only increasing (numeric?) and surely the authors mean non-decreasing. If this is meant to be a case study for the classroom, this decision is rich ground for discussion.}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{\textbf{I’m not sure this is accurate. The robustlmm package in R fits these models rather quickly with thousands of observations. What are the authors using? Does it run out of memory (requires more RAM) or does it take too much time?} The mixed effects models that robustlmm fits are identical to that of lmer. There is extensive documentation about the convergence issues (which are generally false positives). One may use the allFit() function to evaluate this.}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{\textbf{The choice of 0.12 seems arbitrary at best. I don’t see any justification for this approach (e.g., literature, simulations, etc). Why not use bi-square, instead of huber, and simply impute for observations with weight 0. While still arbitrary, it’s decided by the bisquare model and isn’t a guesstimate. Further, it would be sensible for any dataset and not just this one. (This would correspond to a smaller threshold, which may catch too many strange outcomes). In any case, it’s hard to justify something so adhoc.} I agree that it is simpler not to use robustlmm (see note above about robustlmm), and that how to decide what to do with outlandish observations would be rich for class discussion. However, this isn’t discussed as part of this data being a case study – if this is to be used for an educational purpose (beyond how the data were compiled) instructing this type of discussion and the possible avenues for exploration would be important. To teach this, we’d want to used a justified method for data imputation and outlier detection and note there may be other sensible ways (which I think this is) to do it.}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{The authors also make some problematic decisions about the race (Black or not-Black) and ethnicity (Hispanic or not-Hispanic) variables. They suggest three categories (non-Black/non- Hispanic; Black; Hispanic) and don’t discuss the process of combining them, while recent stan- dard is (non-Black/non-Hispanic; non-Black/Hispanic; Black/non-Hispanic; Black/Hispanic). It’s possible that this selection was made for backwards compatibility, lack of certain demo- graphics in the sample, or some other reason but it is not communicated. Please review the literature surrounding the discussion of race and ethnicity (e.g., Standards for the classifica- tion of federal data on race and ethnicity).}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 3-4, lines 34-20) The discussion about “the average and the individual” is underdeveloped. It seems to be tangled with a couple other ideas in these paragraphs, which muddles the point.}
\end{description}

\begin{description}
\item[Reviewer 1:] \textcolor{violet}{The manuscript is much improved. I would like to see less description of how to tidy the data in Section 3 and more discussion of lessons learned (useful for statistics and data science educators and students) in Section 5.}
\end{description}

\begin{description}
\item[Reviewer 3:] \textcolor{violet}{One thing that could be made clearer throughout the paper – especially in Section 5 – is that the ultimate goal here is to get an equivalent dataset to the original that contains the additional years of data and better respects modern social justice norms.  The authors use the terms “update”, “refresh”, and “re-create” throughout, and I think it is sometimes ambiguous whether this is a replication or update.}
\end{description}

\begin{description}
\item[Reviewer 3:] \textcolor{violet}{Here is what I think the strong takeaway in this work is:  "Any longitudinal dataset used for education should have a sufficiently reproducible process to be updated with new data.  The NLSY79 dataset is a great teaching tool that has become outdated, both because the dataset stops in 1994, and because the demographic data could be handled more delicately.  What you have done here is offered a well-documented and reproducible process that expands the dataset to modernity, and matches the original dataset decently well within its scope." This narrative doesn’t require any major structural revision; only some wordsmithing to make sure this message of your contribution hits home throughout the paper.}
\end{description}

\begin{description}
\item[Reviewer 3:] \textcolor{violet}{Page 17:  I remain slightly uncomfortable with the “hand-way-ness” of the modeling step.  I don’t think that “we tried this and it failed to converge” is a satisfactory explanation for dismissing a model that the authors themselves believe would be a better fit.  Perhaps the discussion around robustlmm could be relegated to an appendix or supplement, with a bit more detail about why this model doesn’t converge on the data.  That way it would not distract from the details of the model that was actually used.}
\end{description}

\begin{description}
\item[Reviewer 3:] \textcolor{violet}{Page 18:  Similarly, the justification of the 0.12 cutoff is much improved from the first version of this paper, but it still feels a bit strange.  This sentence in particular throws me off: “That struck a balance between maintaining the natural variability of the wageswith minimizing implausible values.” There are some big assumptions in that sentence to do with what is the “true” natural variablility of the wages and what is “truly” an implausible value.  Is there any semi-objective measure we could use to justify the choice of 0.12 beyond the “eyeball test”?
For example: maybe a plot of the variability of the imputed data for various threshold choices, showing that a low threshold leads to very high variance and a high threshold leads to low variance? Or:  Can you make an argument from predictive power of the model, i.e., that a threshold around 0.12 trained on years 1979-2016 best predicts years 2017-18? Perhaps this is too much lift at this stage, but any amount of quantifiable justification would relieve a lot of the subjectivity around that 0.12 choice.}
\end{description}

\section*{Minor changes:}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 7, figure 1). Is this figure cut off or is it just at the end of the page?}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 9, line 4-6). “HRP1 1980 and HRP2 1980, contain the information about the job number up to 5...” They only contain job number one and two – in other areas of the manuscript you use a subscript i to make the point clear and this would be a good solution here. The code here goes off the page a bit, too.}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 10, tables 1-2) I believe these tables show what the authors see in their cleaned data compared to the NLSY numbers provided on page 15. Is this correct?}
\end{description}

\begin{description}
\item[Reviewer 2:] \textcolor{violet}{On page 6, several times you use the word “plan,” as in “We also plan to include additional variables” and “The plan is to create three datasets as follows.” Has this been done in the paper, or are some of these future goals?}
\end{description}

\begin{description}
\item[Reviewer 2:] \textcolor{violet}{The word “mutate” is used several times in the manuscript. While tidyverse uses are familiar with that as function name and “verb,” I think a general audience is going to find it a bit jarring. Consider replacing with the word “create” or similar.}
\end{description}

\begin{description}
\item[Reviewer 2:] \textcolor{violet}{On page 10, I don’t know what this sentence means: “When this value was missing, 2012, 2014, 2016, and 2018, but available in the first form substituted accordingly.”}
\end{description}

\begin{description}
\item[Reviewer 2:] \textcolor{violet}{The word “that” is used a lot in this manuscript, and most instances could be removed. For example “For example, an article published in the Sydney Morning Herald argues that there is no average Australian” can be replace by “For example, an article published in the Sydney Morning Herald argues there is no average Australian.” I suggest searching for the word “that” and removing any unnecessary instances.}
\end{description}

\begin{description}
\item[Reviewer 2:] \textcolor{violet}{Some extraneous punctuation marks are present}
\begin{itemize}
  \item p1 in abstract: Both “wages textbook subset, have not” and “open source R package, called” do not need commas.
  \item p2 “high school dropouts, from 1979-1994” does not need comma
  \item p2 comma after “divergence of purpose” might be better replaced with dash.
  \item p4 the sentence “Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to 1200/hour.” needs a few edits. I suggest “Plot (C) shows the profile for an individual with a maximum wage that is not so extreme but still indicates a problem: their wages are consistently low except for one year where they earned close to 1200/hour.”
  \item p24 “predominately” does not need a hyphen.
\end{itemize}
\end{description}

\begin{description}
\item[Reviewer 2:] \textcolor{violet}{p2 missing closing parenthesis after Stodel 2020 citation.}
\end{description}

\begin{description}
\item[Reviewer 2:] \textcolor{violet}{Punctuation should go inside quotation marks, not outside}
\begin{itemize}
\item p2 comma  “Applied longitudinal data analysis”,
\item p4 comma “tame data”,
\item p6 comma “statistical value chain”,
\item p9 period “female”.
\item p13 period “number of weeks worked since the last interview”.
\end{itemize}
\end{description}

\begin{description}
\item[Reviewer 2:] \textcolor{violet}{Miscellaneous comments:} 
\begin{itemize}
\item p6 “For example, use a single categorical race variable instead of the two binary race variables.” Perhaps missing a “we” before use?
\item p6 The sentence beginning “van der Loo and de Jonge” is jarring because of the lowercase name, particularly because it starts a paragraph and section. I recommend flipping the clauses to begin “In the context of official statistics, van der Loo and de Jonge…”
\item p18 “The year when the individual starting to work.” should perhaps be “The year when the individual started to work.”
\item p24 The sentence “On an individual level, one needs to know where I am in this data and does this data relate to me.” needs edits. I’m not sure people from the longitudinal study are likely to be looking at this data, so “where I am in this data” is not quite accurate. Rather, people might want to see how their characteristics relate to those in the dataset. 
\item p26 Cooksey reference should have NLSY capitalized. 
\end{itemize}
\end{description}

\begin{description}
\item[Reviewer 1:] \textcolor{violet}{Minor comments:}
\begin{itemize}
\item P2, l46: missing right parenthesis.
\item P9, l24: missing word(s) "this sometimes difficult the adjustment"
\item P11, l43: "hours" worked
\item P12, l32: unit in "weeks"
\item P14, l38: Figure 3 shows...
\item P21, l21: "the weeks worked since" ...
\item P24, l21: The sentence "On an individual level, one needs to know where I am in this data and does this data relate to me" is awkward. Please revise.
\item P34 l 38: Figure 5 (C) is mislabeled
\end{itemize}
\end{description}

\begin{description}
\item[Reviewer 3:] \textcolor{violet}{Page 2, line 42 – Parenthesis is not closed.}
\end{description}

\begin{description}
\item[Reviewer 3:] \textcolor{violet}{This comment from the first review is unaddressed: “Black” and “Hispanic” and “White” should be capitalized.  (There is some dispute among style guides regarding “white”, but the other two are unambiguous.)}
\end{description}

\begin{description}
\item[Reviewer 3:] \textcolor{violet}{Page 7 lines 36-42:}
\begin{itemize}
\item 1. The wages data of the whole NLSY79 cohort, including females.
\item 2. A separate table of the demographic data of the whole NLSY79 cohort.
\item 3. The high school dropouts’ wages data is closest to a refreshed version of Singer and
Willett (2003)’s data.
\end{itemize}
1 and 2 are nouns and 3 is a sentence – I’m a little confused what this third dataset represents.  I think you mean that you create a subset of (1) in the scope of the original data, to see how closely it replicates?
\end{description}

### 2022-04-25 data-raw/ida.R

@@ -385,7 +385,7 @@ n_obs_do_pred <- filter(wages_hs_do, is_pred == TRUE) %>%
   nrow()
 
 ### ---- flow-chart
-include_graphics(here::here("paper/figures/flowchart.png"))
+include_graphics(here::here("paper/figures/flowchart_blind.png"))
 
 ### ---- flow-chart-blind
 include_graphics(here::here("paper/figures/flowchart_blind.png"))

### 2022-05-04 paper/paper-jsdse/paper-jsdse.tex
Adress minor comments.

@@
 % NOTE: To produce blinded version, replace "0" with "1" below.
-\newcommand{\blind}{0}
+\newcommand{\blind}{1}
 
@@
 Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future.
-Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset, have not been clearly articulated.
+Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset have not been clearly articulated.
 We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching.
-Three new data sets and the code to produce them are provided in the open source R package, called {[}CENSORED{]}.
+Three new data sets and the code to produce them are provided in the open source R package called {[}CENSORED{]}.
 \end{abstract}
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis'', which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models.
-The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
+An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis,'' which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models.
+The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
 Department of Labor 2021a).
@@
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, a higher level of education leads to higher wages, and that race makes a difference, on average.
-An exploratory analysis reveals, however, that the individual experience varies a great deal from the overall average.
-Some individuals experience a decline in wages the longer they are in the workforce and many experience volatility in their wages.
+However, an exploratory analysis reveals that the individual experience varies a great deal from the overall average.
+Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
 The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at CENSORED.
 
-This disparity between the average and the individual is a part of statistics, as a discipline that requires more attention.
+This disparity between the average and the individual is a part of statistics as a discipline that requires more attention.
 This particular data set is a prime example of discussing this disparity.
@@
 The iris data (Andereson 1935) is a prime example.
-It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(A side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020).
-We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
-
-For the field of statistics and data science by association, it is increasingly important to reach the individual.
-One might describe this as a divergence of purpose, statistics for public policy, or statistics for the public.
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis (a side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)).
+We argue the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
+
+For statistics and data science by association, it is increasingly important to reach the individual.
+One might describe this as a divergence of purpose -- statistics for public policy -- or statistics for the public.
 The two are not the same.
-As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require that statistics address the individual.
-For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic.
+As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual.
+For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is you are more than your demographic.
 The majority of people in the study do not have a pattern that is similar to the average.
@@
 A similar tone is echoed occasionally in the public media.
-For example, an article published in the Sydney Morning Herald argues that there is no average Australian (Moncrief 2015).
+For example, an article published in the Sydney Morning Herald argues there is no average Australian (Moncrief 2015).
 
@@
 EDA can be considered to encompass IDA.
-In practice, the three stages of cleaning, summarizing, and exploring are cyclical, that one often needs to do more cleaning after scrutinizing.
+In practice, the three stages of cleaning, summarizing, and exploring are cyclical, one often needs to do more cleaning after scrutinizing.
 Dasu and Johnson (2003) say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.
@@
 The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables (Wickham 2011).
-Tidy data shouldn't be confused with ``tame data'', which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
+Tidy data shouldn't be confused with ``tame data,'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
 The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
@@
 
-Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
+Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
 Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).
@@
 We also modify some variables.
-For example, use a single categorical race variable instead of the two binary race variables.
-We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs that the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked.
+For example, we use a single categorical race variable instead of the two binary race variables.
+We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked.
 We do not attempt to re-create the unemployment rate variable.
@@
 
-van der Loo and de Jonge (2018), in the context of official statistics, describe the ``statistical value chain'', which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
-What we have colorfully named as wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
+In the context of official statistics, van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
 In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.
@@
 The raw data consequently has a large number of columns (770 to be specific).
-The values in the cell under the variables that begin with \texttt{HRP} correspond to the hourly wage in dollars.
+The values in the cell under the variables begin with \texttt{HRP} correspond to the hourly wage in dollars.
 A glimpse of this data shows:
@@
 2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham, François, et al.
-2020) and \texttt{stringr} (Wickham 2019) to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
+2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 We use the term \texttt{gender} as the reports about the data on various pages in the website more commonly use \texttt{gender}.
-Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female''.
-The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
+Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female.'' The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
 Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification (Kennedy et al.
@@
 It represents the difference between the day an individual enters the labor force (\texttt{EMPLOYERS\_ALL\_STARTDATE\_ORIGINAL.01\textasciitilde{}Y\_XRND} in the database) relative to the date of the survey, which we call \texttt{yr\_wforce}.
-However, using this calculation produced numbers that do not quite match with the original data.
-
-Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide suggests that it should be calculated based on the variable ``number of weeks worked since the last interview''.
-This would remove periods of unemployment which makes sense when measuring experience while actually working.
+However, using this calculation produced numbers that do not quite match the original data.
+
+Reading the section titled \href{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}{``Topical Guide to the Data''} in the guide suggests it should be calculated based on the variable ``number of weeks worked since the last interview.'' This would remove periods of unemployment which makes sense when measuring experience while actually working.
 Since it is only measured since the last interview, this needs to be cumulated for each survey year.
@@
 
-According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure that the data is clean, valid, and ready to be deployed in the later analyses.
+According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure the data is clean, valid, and ready to be deployed in the later analyses.
 This is supported by Chatfield (1985), who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
@@
 The respondents' ages ranged from 12 to 22 when first interviewed in 1979.
-Hence, we validate whether all of the respondents were in this range in the data we extracted.
+Hence, we validate whether all respondents were in this range in the data we extracted.
 Additionally, the \href{https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction}{NLSY} also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic).
 To validate this, we used the \texttt{demog\_nlsy79}, i.e., the data with the survey years 1979 sample.
-Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest that the demographic data we had is consistent with the sample information in the database.
+Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest the demographic data we had is consistent with the sample information in the database.
 
@@
 These need to be inspected more closely.
-It is also important to note that some shorter profiles indicate that some individuals have left the study before it has finished.
+It is also important to note that some shorter profiles indicate some individuals have left the study before it has finished.
 Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis to account for the bias induced by the inadvertent censoring.
@@
 Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median, and maximum) for all individuals.
-This tells us that there are a number of individuals with unbelievably high maximum wages.
-Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to \$1200/hour.
+This tells us there are a number of individuals with unbelievably high maximum wages.
+Plot (C) shows the profile for an individual with a maximum wage that is not so extreme but still indicates a problem: their wages are consistently low except for one year where they earned close to \$1200/hour.
 This does not seem to be reasonable and leads us to use a procedure to detect and fix these temporal anomalies.
@@
 \caption{Summary plots to check the data after the tidying stage: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, and maximum wages of each individual, (C) and one individual (id=39) with an unusual wage relative to their years of data.
-It reveals that some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.
+It reveals that some values of hourly wages are unbelievable, and some individuals have extremely unusual wages in some years.
 Accordingly, more cleaning is necessary to treat these extreme values.}\label{fig:feature-plot}
@@
 
-Extremely high values were also found in the total hours of work, where some observations reported as having worked for 420 hours a week in total.
+Extremely high values were also found in the total hours of work, where some observations reported having worked for 420 hours a week in total.
 According to Pergamit et al.
-(2001), one of the flaws of the NLSY79 employment data is that the NLSY79 collects the information of the working hours since the last interview.
+(2001), one of the flaws of the NLSY79 employment data is the NLSY79 collects the information of the working hours since the last interview.
 Thus, it might be challenging for the respondents to track the within-job hours' changes between survey years, especially for the respondents with fluctuating working hours or seasonal jobs.
@@
 The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
-It should be noted that it is not possible to determine if all outliers are errors, and it might be that an individual had abnormally high wages at a particular time.
+It should be noted that it is not possible to determine if all outliers are errors, and it might be an individual indeed had abnormally high wages at a particular time.
 
 To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done.
-Samples of individuals were examined under the different thresholds, with an eye to the smoothness of the profiles.
-Overly smooth profiles would indicate that the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
+Samples of individuals were examined under the different thresholds with an eye to the smoothness of the profiles.
+Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
 A threshold of 0.12 was chosen.
@@
 Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
-The plot shows that fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
+The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
 
@@
 The highest wage overall is now around \$1000.
-Plot (A) shows a more reasonable spaghetti plot, where there are some profiles with high wages, but most profiles have wages under \$300, and there is a steady increase in wages with years.
-Plot (B) shows that there are still a small number of individuals with high maximum wages.
+Plot (A) shows a more reasonable spaghetti plot, where there are some profiles with high wages, but most profiles have wages under \$300, and there has been a steady increase in wages over the years.
+Plot (B) shows there are still a small number of individuals with high maximum wages.
 Plot (C) shows the profile for ID=39 after imputing the extreme value.
@@
 
-\caption{Re-make of the summary plots of the fully processed data suggest that it is now in a reasonable state: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, (C) and maximum wages of each individual, and one individual with an unusual wage relative to their years of data.
+\caption{Re-make of the summary plots of the fully processed data suggest it is now in a reasonable state: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, (C) and maximum wages of each individual, and one individual with an unusual wage relative to their years of data.
 }\label{fig:fixed-feature-plot}
@@
 \item
-  \texttt{stwork}: The year when the individual starting to work.
+  \texttt{stwork}: The year when the individual started to work.
 \item
@@
   173 individuals were more than 17 years old in 1979.
-Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in the subset.
+Thus, it looks like the description of the original data is not quite accurate that there are people older than 17 in the subset.
 Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
@@
   \item
-    35 of them completed at least 12th grade with a diploma instead of GED (\texttt{ged} variable is coded to 1).
-This suggests that they are not dropouts, and so we excluded them from the refreshed data.
+    35 of them completed at least 12th grade with a diploma instead of GED (\texttt{ged} variable is coded to 1). This suggests they are not dropouts, and so we excluded them from the refreshed data.
   \item
@@
 From the NLSY79 topical guide (Bureau of Labor Statistics, U.S.
-Department of Labor 2021b), we find that several variables are tagged as work experience-related variables.
+Department of Labor 2021b), we find several variables are tagged as work experience-related variables.
 One of them is the weeks of worked since the last interview.
@@
 The highest grade completed has some confusion.
-There are several ways that this is reported, including the highest grade ever completed and also the \texttt{hgc} at each survey year.
+There are several ways this is reported, including the highest grade ever completed and also the \texttt{hgc} at each survey year.
 To match the original data, it is appropriate to use the \texttt{hgc} while in high school.
-The documentation suggests that this variable is available, but it is not actually present in the data.
-Hence, for matching the original data, we have calculated the \texttt{hgc} to match the \texttt{hgc} achieved in the years between 1979 and 1994 based on the yearly survey value.
-The result does not exactly match with the original for a few individuals.
+The documentation suggests this variable is available, but it is not actually present in the data.
+Hence, to match the original data, we have calculated the \texttt{hgc} to match the \texttt{hgc} achieved in the years between 1979 and 1994 based on the yearly survey value.
+The result does not exactly match the original for a few individuals.
 
@@
 
-For the data providers, we recommend that a better validation system with clear rules applied at data entry and that alternative output formats, such as a tidy format, would help users make better use of their resources.
-The problem with many of the wages records is that there are implausible values or confusion on how to record wages for multiple jobs.
+For the data providers, we recommend a better validation system with clear rules applied at data entry and alternative output formats, such as a tidy format, which would help users better use their resources.
+The problem with many wages records is that there are implausible values or confusion on how to record wages for multiple jobs.
 These values can be validated with simple checks at data entry.
-Providing an open data resource also is accompanied by the responsibility that the data, especially data that is as valuable as this, is reliable.
+Providing an open data resource is also accompanied by the responsibility that the data, especially as valuable as this, is reliable.
 Users need to be able to trust the data.
@@
 
-The wages data provides a good opportunity to discuss the difference between statistics to use for public policy and statistics that relate to the individual.
+The wages data provides a good opportunity to discuss the difference between statistics for public policy and statistics that relate to the individual.
 Public policy is based on models, yielding averages that might vary across strata.
-Modeling the wages relative to workforce experience, with demographic covariates, we learn that there are significantly different patterns.
-That more education leads to increasingly higher wages, which is a satisfying result for educators.
-It means that education makes a difference in the wage experience and that public policy that encourages education is a data-supported action.
-We would also learn, although not presented here, that race matters and that being black leads to lower wages.
+Modeling the wages relative to workforce experience with demographic covariates, we learn that there are significantly different patterns.
+More education leads to increasingly higher wages, which is a satisfying result for educators.
+It means education makes a difference in the wage experience and that public policy that encourages education is a data-supported action.
+We would also learn, although not presented here, that race matters and being black leads to lower wages.
 This is a disturbing finding because there is no rationale for such a difference in a fair society.
@@
 
-On an individual level, one needs to know where I am in this data and does this data relate to me.
+On an individual level, people might want to see how their characteristics relate to those in the dataset.
 To do this, the individual profiles need to be explored.
-Pre-dominantly, we would learn that the variation from one individual to another is far more than the variation between demographic strata.
+Predominantly, we would learn the variation from one individual to another is far more than the variation between demographic strata.
 For example, many individuals with lower educational attainment earn very high wages.
@@
 The above interpretations of results from analyzing the wages data rely on trusting that the data provided is accurate and valid.
-The wages data is collected by a reputable organization, but we found that the data has obvious errors that should be corrected.
+The wages data is collected by a reputable organization, but we found the data has obvious errors that should be corrected.
 This paper has illustrated procedures and guidelines to achieve valid data and provides the code and details for it to be reproduced and modified if deemed appropriate.

### 2022-05-04 paper/paper.Rmd
Adress minor comments.

@@
 Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future.
-Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset, have not been clearly articulated.
+Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset have not been clearly articulated.
 We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching.
-Three new data sets and the code to produce them are provided in the open source R package, called [CENSORED].
+Three new data sets and the code to produce them are provided in the open source R package called [CENSORED].
 ---
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis",  which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models.
-The data records hourly wages of a sample of high school dropouts, from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
+An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis," which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models.
+The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
 
 The story from modeling the data (and as reported by Singer and Willett) is that wages increase with the length of time in the workforce, a higher level of education leads to higher wages, and that race makes a difference, on average.
-An exploratory analysis reveals, however, that the individual experience varies a great deal from the overall average.
-Some individuals experience a decline in wages the longer they are in the workforce and many experience volatility in their wages.
+However, an exploratory analysis reveals that the individual experience varies a great deal from the overall average.
+Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
 The wages data was used to illustrate exploratory longitudinal data analysis in @ilk2004, and was further developed into a case study for use in the teaching of exploratory data analysis at `r censor_word("Iowa State University")`.
@@
 
-This disparity between the average and the individual is a part of statistics, as a discipline that requires more attention.
+This disparity between the average and the individual is a part of statistics as a discipline that requires more attention.
 This particular data set is a prime example of discussing this disparity.
@@
 The iris data [@iris-data] is a prime example.
-It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(A side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris].
-We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis (a side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]).
+We argue the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
 
-For the field of statistics and data science by association, it is increasingly important to reach the individual.
-One might describe this as a divergence of purpose, statistics for public policy, or statistics for the public.
+For statistics and data science by association, it is increasingly important to reach the individual.
+One might describe this as a divergence of purpose -- statistics for public policy -- or statistics for the public.
 The two are not the same.
-As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require that statistics address the individual.
-For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is that you are more than your demographic.
+As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual.
+For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is you are more than your demographic.
 The majority of people in the study do not have a pattern that is similar to the average.
@@
 A similar tone is echoed occasionally in the public media.
-For example, an article published in the Sydney Morning Herald argues that there is no average Australian [@notaverage].
+For example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage].
 
@@
 EDA can be considered to encompass IDA.
-In practice, the three stages of cleaning, summarizing, and exploring are cyclical, that one often needs to do more cleaning after scrutinizing.
+In practice, the three stages of cleaning, summarizing, and exploring are cyclical, one often needs to do more cleaning after scrutinizing.
 @DasuTamraparni2003Edma say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.
@@
 The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables [@plyr].
-Tidy data shouldn't be confused with "tame data", which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
+Tidy data shouldn't be confused with "tame data," which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
 The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
@@
 Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data [@MichaelRPergamit2001DWTN].
-<!--AE comment: For example, I would like more discussion about the textbook selected as the exemplar.
-It sounds like this dataset is used in many books, why choose that one?
-The text should include the name of the book, so readers do not need to flip to the citations in order to learn what it was.
-From the name, I am guessing that the book focuses heavily on this dataset, which is why it makes sense to reproduce their subset.
-Be explicit about this.-->
 
-Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose that is consistent with @grimshaw's statistics education goal of embracing authentic data experiences.
+Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose consistent with @grimshaw's statistics education goal of embracing authentic data experiences.
 Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by @SingerJudithD2003Alda.
@@
 We also modify some variables.
-For example, use a single categorical race variable instead of the two binary race variables.
-We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs that the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked.
+For example, we use a single categorical race variable instead of the two binary race variables.
+We also include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked.
 We do not attempt to re-create the unemployment rate variable.
 
-The plan is to create three datasets as follows:
+We aim to create three datasets as follows:
 
@@
 2. A separate table of the demographic data of the whole NLSY79 cohort.
-3. The high school dropouts' wages data is closest to a refreshed version of @SingerJudithD2003Alda's data.
+3. The subset of wages data in (1), which is the wages of high school dropouts' as a refreshed version of @SingerJudithD2003Alda's data.
 
@@
 
-@LooMarkvander2018Sdcw, in the context of official statistics, describe the "statistical value chain", which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
-What we have colorfully named as wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
+In the context of official statistics, @LooMarkvander2018Sdcw describe the "statistical value chain," which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
 In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidy the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so that the resulting input data can be used downstream for validating the data as described in Section \@ref(ida) and Section \@ref(censor).
@@
 The raw data consequently has a large number of columns (`r ncol(new_data_qnames)` to be specific).
-The values in the cell under the variables that begin with `HRP` correspond to the hourly wage in dollars.
+The values in the cell under the variables begin with `HRP` correspond to the hourly wage in dollars.
 A glimpse of this data shows:
@@
 Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
-This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long-form, with `dplyr` [@dplyr] and `stringr` [@stringr] to mutate new variables from the downloaded data from the database, and code levels of factors by text wrangling.
+This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long-form, with `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 We use the term `gender` as the reports about the data on various pages in the website more commonly use `gender`.
-Gender, as provided in the data, is self-reported and only has two categories, "male" and "female".
-The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
+Gender, as provided in the data, is self-reported and only has two categories, "male" and "female." The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
 Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification [@kennedy2020using].
-From a statistical perspective, this sometimes difficult the adjustment of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex [@kennedy2020using].
+From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex [@kennedy2020using].
 Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
@@
 * If the hourly rate is recorded as 0, we set wage as missing.
-* If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hour worked as missing.
+* If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.
 
@@
 To obtain the work experience since 1979, we calculate the cumulative value.
-As the measurement unit is in week, we convert this to year.
+As the measurement unit is weeks, we convert this to year.
 
@@
 It represents the difference between the day an individual enters the labor force (`EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND` in the database) relative to the date of the survey, which we call `yr_wforce`.
-However, using this calculation produced numbers that do not quite match with the original data.
+However, using this calculation produced numbers that do not quite match the original data.
 
-Reading the section titled ["Topical Guide to the Data"](https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience) in the guide suggests that it should be calculated based on the variable "number of weeks worked since the last interview".
-This would remove periods of unemployment which makes sense when measuring experience while actually working.
+Reading the section titled ["Topical Guide to the Data"](https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience) in the guide suggests it should be calculated based on the variable "number of weeks worked since the last interview." This would remove periods of unemployment which makes sense when measuring experience while actually working.
 Since it is only measured since the last interview, this needs to be cumulated for each survey year.
@@
 
-According to @HuebnerMariannePhD2016Asat, initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure that the data is clean, valid, and ready to be deployed in the later analyses.
+According to @HuebnerMariannePhD2016Asat, initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure the data is clean, valid, and ready to be deployed in the later analyses.
 This is supported by @Chatfield1985TIEo, who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
@@
 In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides.
-<!-- AE comments: This comes as a surprise to the reader.
-What anomaly?
-The wage anomalies should be mentioned in the introduction so as not to be a surprise at this point in the manuscript.
- --> The first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
+The first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
 This is followed by graphical summaries using methods available in `ggplot2` [@ggplot2] and `brolgar` [@brolgar].
@@
 The respondents' ages ranged from 12 to 22 when first interviewed in 1979.
-Hence, we validate whether all of the respondents were in this range in the data we extracted.
+Hence, we validate whether all respondents were in this range in the data we extracted.
 Additionally, the [NLSY](https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction) also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic).
 To validate this, we used the `demog_nlsy79`, i.e., the data with the survey years 1979 sample.
-Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest that the demographic data we had is consistent with the sample information in the database.
+Tables \ref{tab:age-table} and \ref{tab:gender-race-table} suggest the demographic data we had is consistent with the sample information in the database.
 
@@
 
-<!--AE comment: With the IDA about wage data, I would appreciate a bit more context.
-Why sample 36 observations?
-I am not a longitudinal data expert, so I was a bit surprised when I flipped to the plot and found it to be small-multiple time series, although upon further consideration it makes sense.
-In that section it would be good to describe what the figure shows in a bit more detail.
-Perhaps “We randomly sample 36 respondents from the data, and plot their average wage as a time series.
-Looking at the patterns over time, we see a lot of variability in wages.
-For example, the people shown in panels 5, 7, and 11 have [explain what is interesting here, again I’m not an expert].” The next sentence here, “Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.” does not seem to provide much additional information.
-What were you looking for here?
-Overall trends over time, increasing/decreasing?
-Any places where one year’s wage looked really different?-->
-
 In the next step, we explore the mean hourly wage data of samples of individuals.
@@
 A random sample of 36 individuals is chosen (using the `sample_n_keys` function in `brolgar`).
-Their longitudinal profiles are plotted, faceted by `id`, and using free $y$ scales so that the individual patterns can be examined (Figure \ref{fig:sample-plot}).<!--AE comment: "It" implies...
-Please specify what "it" refers to.
-This could be done by adding onto the first sentence...
-"as shown in Figure 2, which shows that some respondents have high variability in wages, particularly 5, 7, and 11." However, 7 and 11 don't seem to have particularly high variability, especially when compared to 18.
-This whole paragraph needs to be fixed/updated.--> There is a lot of variability from one individual to another and substantial fluctuation in wages at different times for most individuals.
+Their longitudinal profiles are plotted, faceted by `id`, and using free $y$ scales so that the individual patterns can be examined (Figure \ref{fig:sample-plot}).
+There is a lot of variability from one individual to another and substantial fluctuation in wages at different times for most individuals.
 Some individuals  (2799, 11041, 11146) are only measured for a short period.
@@
 These need to be inspected more closely.
-It is also important to note that some shorter profiles indicate that some individuals have left the study before it has finished.
+It is also important to note that some shorter profiles indicate some individuals have left the study before it has finished.
 Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis to account for the bias induced by the inadvertent censoring.
 
- <!--AE comment: Not panel 9  the IDs in panel 9. -->
-<!--I think there is a narrative step missing here, which is that looking at those samples led you to look at plots that show values for the wage variable over all participants and all years.
-The phrase “the summary plots” was not enough description for me to know what to expect when I flipped to Figure 3.
-Please be explicit about what the three plots were, why they were made, and what you were looking for.
-I think there is another narrative step missing about Figure 3C, which is that after you saw there were outliers you tried to identify which respondents had the extreme values.
-Again, please spell this out in the text.
-When you return to the plots for the entire dataset in Figure 5, please make parallel comments.
--->
-
-Figure \ref{fig:feature-plot}s shows an alternative way to check the data quality.
+Figure \ref{fig:feature-plot} shows an alternative way to check the data quality.
 Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to $60,000/hour) for some individuals, mostly around 1990.
 Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median, and maximum) for all individuals.
-This tells us that there are a number of individuals with unbelievably high maximum wages.
-Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to $1200/hour.
+This tells us there are a number of individuals with unbelievably high maximum wages.
+Plot (C) shows the profile for an individual with a maximum wage that is not so extreme but still indicates a problem: their wages are consistently low except for one year where they earned close to $1200/hour.
 This does not seem to be reasonable and leads us to use a procedure to detect and fix these temporal anomalies.
@@
 ```
-
-<!--AE comment: Table 2 seems to have more precision than is necessary.
-Certainly 100% does not need to be reported as 100.00%, and probably all the percentages could be rounded to the nearest 1 percent.
-The caption could explain this, as there might be some rounding errors.
--->
 
@@
 
-<!--AE comment: Figure 2’s caption should include something about how that data is the raw values from the survey, before inputing anomalous values so the figure and caption can stand alone.-->
-
-<!--For both Figure 3 and Figure 5, I would prefer the letter label before the description, but that’s a matter of personal preference.-->
-
 ```{r feature-plot, fig.cap = "Summary plots to check the data after the tidying stage: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, and maximum wages of each individual, (C) and one individual (id=39) with an unusual wage relative to their years of data.
-It reveals that some values of hourly wages are unbelievable, and some individuals have extemely unusual wages in some years.
+It reveals that some values of hourly wages are unbelievable, and some individuals have extremely unusual wages in some years.
 Accordingly, more cleaning is necessary to treat these extreme values.", fig.width=10, fig.height=4, out.width="100%", cache = FALSE}
@@
 
-<!--The next paragraph, which begins “The anomalies are also found” should perhaps be “Similar anomalies were found.” Was it the same respondents with the extremely high wages that showed extremely high number of hours of work?
-I suspect not, but please be explicit.-->
-Extremely high values were also found in the total hours of work, where some observations reported as having worked for 420 hours a week in total.
-According to @MichaelRPergamit2001DWTN, one of the flaws of the NLSY79 employment data is that the NLSY79 collects the information of the working hours since the last interview.
+Extremely high values were also found in the total hours of work, where some observations reported having worked for 420 hours a week in total.
+According to @MichaelRPergamit2001DWTN, one of the flaws of the NLSY79 employment data is the NLSY79 collects the information of the working hours since the last interview.
 Thus, it might be challenging for the respondents to track the within-job hours' changes between survey years, especially for the respondents with fluctuating working hours or seasonal jobs.
-It even has been more challenging since 1994, after which respondents were only surveyed every other year and thus had to recall two full years’ job history.
+It even has been more challenging since 1994, after which respondents were only surveyed every other year and thus had to recall two full years' job history.
 This shortcoming might also contribute to the fluctuation of one's wages data.
@@
 
-<!--AE comment: Section 3.2.1 begins “As part of the IDA, which is the model formulation, we build a robust linear regression model to treat the extreme values in the data.” This could use clarification.
-In 3.2, you said that part of IDA is “ model formulation without any formal statistical inference.” To me, that sounds like using modeling as a descriptive technique, and it’s something I would consider to be more a part of EDA.
-But then in 3.2.1, while it is perhaps not using modeling in an inferential way, the model is being used for more than just description.
-I might rephrase the first sentence as follows: “In order to treat the extreme values in the data, we built a robust linear regression model.” -->
-<!--AE comment: Awkward sentence.
-"which is the model formulation"?--> A robust linear regression model using the `rlm` function from `MASS` package [@mass] is used to treat the extreme values in the data.
+A robust linear regression model using the `rlm` function from `MASS` package [@mass] is used to treat the extreme values in the data.
 The robustness weight is used to determine if a value should be replaced with the fitted value from the model.
@@
 Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm].
-<!--AE comment: “However, the challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.” I don’t think this sentence needs a “However” and probably should say “past which” rather than “in which”-->The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
-It should be noted that it is not possible to determine if all outliers are errors, and it might be that an individual had abnormally high wages at a particular time.
-
-<!--AE comment: At the bottom of page 11, where you describe the model formation, more detail would be appreciated.
-You are modeling the mean hourly wage based on year.
-Why?
-Is this standard practice for robust linear regression?
-What is a “slight” residual?
-Is it defined based on absolute magnitude, or standard deviations, or something else?
-Later, you say “To minimize the risk of mistakenly identifying an outlier as an “erroneous outlier”.” Is an erroneous outlier a technical term?
-If not, I think you are trying not to identify erroneous outliers, not mistakenly identifying outliers as erroneous outliers (that’s a double negative).
-The value of 0.12 feels very specific and out of place in a paper that has had few numerals.
-Is this a number an expert in RLR would be able to understand?
-Again, is that a raw value?
-Is it a standard deviation?
-Is there literature about how to choose a threshold?
-Overall, the sentence “We find that 0.12 is the most reasonable value to be the threshold to minimize that drawback’s risk because it still captures the sensible spikes in the data.” is vague and should be expanded upon.
-What is “that drawback’s risk”?
-What are “the sensible spikes in the data?” Be more explicit.-->
+The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
+It should be noted that it is not possible to determine if all outliers are errors, and it might be an individual indeed had abnormally high wages at a particular time.
 
 To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done.
-Samples of individuals were examined under the different thresholds, with an eye to the smoothness of the profiles.
-Overly smooth profiles would indicate that the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
+Samples of individuals were examined under the different thresholds with an eye to the smoothness of the profiles.
+Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
 A threshold of 0.12 was chosen.
@@
 The highest wage overall is now around $1000.
-Plot (A) shows a more reasonable spaghetti plot, where there are some profiles with high wages, but most profiles have wages under $300, and there is a steady increase in wages with years.
-Plot (B) shows that there are still a small number of individuals with high maximum wages.
+Plot (A) shows a more reasonable spaghetti plot, where there are some profiles with high wages, but most profiles have wages under $300, and there has been a steady increase in wages over the years.
+Plot (B) shows there are still a small number of individuals with high maximum wages.
 Plot (C) shows the profile for ID=39 after imputing the extreme value.
 The wages for this individual increase over the years, and do fluctuate some between 1900 and 2005.
-<!--AE comment: What does “It implies that the fluctuation can still be observed in the data after the treatment.” mean?
-I suspect this sentence should be replaced with something along the lines of “The figure shows that fluctuations in wages still exist even in the imputed data.”-->
 
-```{r fixed-feature-plot, fig.cap = "Re-make of the summary plots of the fully processed data suggest that it is now in a reasonable state: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, (C) and maximum wages of each individual, and one individual with an unusual wage relative to their years of data.
+```{r fixed-feature-plot, fig.cap = "Re-make of the summary plots of the fully processed data suggest it is now in a reasonable state: (A) longitudinal profiles of wages for all individuals 1979-2018, (B) boxplots of minimum, median, (C) and maximum wages of each individual, and one individual with an unusual wage relative to their years of data.
 ", fig.width=10, fig.height=4, out.width="100%", cache = FALSE}
@@
   13. `hours:`: The total number of hours the individual usually works per week.
-  14. `stwork`: The year when the individual starting to work.
+  14. `stwork`: The year when the individual started to work.
   15. `yr_wforce`: The length of time in the workforce in years (`year` - `stwork`).
@@
 173 individuals were more than 17 years old in 1979.
-Thus, it looks like the description of the original data is not quite accurate, that there are people older than 17 in the subset.
+Thus, it looks like the description of the original data is not quite accurate that there are people older than 17 in the subset.
 Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
 2. 79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
-    i.
-35 of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded to 1).
-This suggests that they are not dropouts, and so we excluded them from the refreshed data.
+    i. 35 of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded to 1). This suggests they are not dropouts, and so we excluded them from the refreshed data. 
     ii. The information about `ged` is missing in 38 individuals. We decided to include them in the refreshed data. 
@@
 This variable is not explicitly available in the database.
-From the NLSY79 topical guide [@nlsy79guide], we find that several variables are tagged as work experience-related variables.
-One of them is the weeks of worked since the last interview.
+From the NLSY79 topical guide [@nlsy79guide], we find several variables are tagged as work experience-related variables.
+One of them is the weeks worked since the last interview.
 This is used to calculate the variable.
@@
 The highest grade completed has some confusion.
-There are several ways that this is reported, including the highest grade ever completed and also the `hgc` at each survey year.
+There are several ways this is reported, including the highest grade ever completed and also the `hgc` at each survey year.
 To match the original data, it is appropriate to use the `hgc` while in high school.
-The documentation suggests that this variable is available, but it is not actually present in the data.
-Hence, for matching the original data, we have calculated the `hgc` to match the `hgc` achieved in the years between 1979 and 1994 based on the yearly survey value.
-The result does not exactly match with the original for a few individuals.
+The documentation suggests this variable is available, but it is not actually present in the data.
+Hence, to match the original data, we have calculated the `hgc` to match the `hgc` achieved in the years between 1979 and 1994 based on the yearly survey value.
+The result does not exactly match the original for a few individuals.
 
@@
 
-<!--AE comment: At the end of page 14, it would be good to explicitly say that if people don’t like your decisions, they can grab the code and adjust the parameters themselves!
-It’s implied, but should be made explicit.
--->
 Some readers may disagree with our decisions made to produce the refreshed textbook data and may have better insight than us in producing more appropriate textbook data.

### 2022-05-04 paper/paper.tex
Adress minor comments.

@@
 We use the term \texttt{gender} as the reports about the data on various pages in the website more commonly use \texttt{gender}.
-Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female''.
-The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
+Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female.'' The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
 Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification (Kennedy et al.
 2020).
-From a statistical perspective, this sometimes difficult the adjustment of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex (Kennedy et al.
+From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex (Kennedy et al.
 2020).
@@
 \item
-  If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hour worked as missing.
+  If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.
 \end{itemize}
@@
 These need to be inspected more closely.
-It is also important to note that some shorter profiles indicate that some individuals have left the study before it has finished.
+It is also important to note that some shorter profiles indicate some individuals have left the study before it has finished.
 Checking whether the demographics of the early departing are similar to those who remain in the study is an important part of any downstream analysis to account for the bias induced by the inadvertent censoring.
 
-Figure \ref{fig:feature-plot}s shows an alternative way to check the data quality.
+Figure \ref{fig:feature-plot} shows an alternative way to check the data quality.
 Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to \$60,000/hour) for some individuals, mostly around 1990.
 Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median, and maximum) for all individuals.
-This tells us that there are a number of individuals with unbelievably high maximum wages.
-Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to \$1200/hour.
+This tells us there are a number of individuals with unbelievably high maximum wages.
+Plot (C) shows the profile for an individual with a maximum wage that is not so extreme but still indicates a problem: their wages are consistently low except for one year where they earned close to \$1200/hour.
 This does not seem to be reasonable and leads us to use a procedure to detect and fix these temporal anomalies.
@@
 From the NLSY79 topical guide (Bureau of Labor Statistics, U.S.
-Department of Labor 2021b), we find that several variables are tagged as work experience-related variables.
-One of them is the weeks of worked since the last interview.
+Department of Labor 2021b), we find several variables are tagged as work experience-related variables.
+One of them is the weeks worked since the last interview.
 This is used to calculate the variable.

### 2022-05-04 response/response2.Rmd
Adress minor comments.

@@
 
+* Yes, it is cut off, we have made the size smaller, so it is not cut off.
+
 \begin{description}
@@
 
+* These have been done in the paper so we have removed it.
+
 \begin{description}
@@
 \end{description}
+
+* We have changed the word "mutate" to be "create".
 
@@
 \end{description}
+
+* We have removed unnecessary instances.
 
@@
 
+* All of the extraneous punctuation marks mentioned have been removed.
+
 \begin{description}
@@
 \end{description}
+
+* We have added the missing closing parenthesis.
 
@@
 \end{description}
+
+* Done.
 
@@
 
+* All of the comments on the list above have been addressed. (**Except for the Cooksey reference**)
+
 \begin{description}
@@
 
+* All of the comments on the list above have been addressed.
+
 \begin{description}
@@
 \end{description}
+
+* Yes, you are correct, we have edited the point 3 to improve the clarity of this sentence.

### 2022-05-04 data-raw/ida.R

@@ -201,7 +201,7 @@ ggplot(wages_compare) +
               labels = c("Before", "After")) +
   scale_x_continuous("Year",
                      breaks = seq(1980, 2020, 10),
-                     labels = c("80", "90", "00", "10", "20"),
+                     #labels = c("80", "90", "00", "10", "20"),
                      minor_breaks = seq(1980, 2020, 5)) +
   #xlab("Year") +
   ylab("Hourly wage ($)") +
@@ -217,7 +217,7 @@ spag2 <- wages_cleaned %>%
   geom_line(alpha = 0.1) +
   scale_x_continuous("Year",
                      breaks = seq(1980, 2020, 10),
-                     labels = c("80", "90", "00", "10", "20"),
+                     #labels = c("80", "90", "00", "10", "20"),
                      minor_breaks = seq(1980, 2020, 5)) +
   theme_bw() +
   labs(tag = "(A)") +
@@ -266,7 +266,7 @@ plot_high_after <- ggplot(filter(wages_cleaned, id == 39)) +
   annotate("text", x = 1985, y = 26, label = "id: 39") +
   scale_x_continuous("Year",
                      breaks = seq(1980, 2020, 10),
-                     labels = c("80", "90", "00", "10", "20"),
+                     #labels = c("80", "90", "00", "10", "20"),
                      minor_breaks = seq(1980, 2020, 5)) +
   theme_bw() +
   labs(tag = "(C)", y = "Hourly wage ($)")

### 2022-05-05 response/response2.Rmd
modifying response

@@
 
-We thank the Editor, Associate Editor and the three reviewers for their extensive comments, which have helped substantially to improve the manuscript.
-In response, we have made revisions of the main manuscript as outlined below.
+We thank the editor, associate editor and the reviewers for their feedback. Here is how we have made revisions to the manuscript to address each point.
 

### 2022-05-05 response/response2.Rmd
responses all sorted

@@
 
-The editor mention that while the data are interesting, the necessary details for a pedagogical application or case study are still largely missing.
-The manuscript reads more like a paper about the R package, something for a software journal.
+The editor says \textcolor{violet}{The paper falls squarely into the call for the special issue on `Reproducibility and responsible workflow".
+Some additional suggestions regarding framing are provided by the reviewers and the AE.} The only comments on framing appear to come from the associate editor: \textcolor{violet}{If this was (1) a general process for demonstrating the cleaning of longitudinal data, (2) a case study for cleaning data in the classroom, or (3) a case study for using the data in the classroom the value would be clear.
+In its current state, however, the manuscript still partly each of the three by describing the R package they've created and, as a result, does not  fully address any of the three.}
 
-The editor agree with the general sentiment among two of three reviewers that this paper is improved over the previous version, but still does not make clear contributions to educators other than its use in class.
-If this was (1) a general process for demonstrating the cleaning of longitudinal data, (2) a case study for cleaning data in the classroom, or (3) a case study for using the data in the classroom the value would be clear.
-In its current state, however, the manuscript still partly each of the three by describing the R package they’ve created and, as a result, does not  fully address any of the three.
+We would argue that the current form of the paper is clearly (2), and we have added some wording in the introduction and lesson learned sections to help make this clear. XXX DC to do
 
 \begin{description}
-\item[Associate Editor:] \textcolor{violet}{\textbf{The author switches between gender (man/woman/etc) and sex (male/female/etc.) throughout the manuscript.
+\item[Associate Editor:] \textcolor{violet}{\textbf{ORIGINAL COMMENT: The author switches between gender (man/woman/etc) and sex (male/female/etc.) throughout the manuscript.
 Further, the explanation here covers important ideas in teaching how to use data on gender/sex and race, but it’s just the beginning of ideas.
-It might be better to discuss the questions on the survey and discuss possible shortcomings, rather than make a somewhat empty gesture towards today’s standard.} The authors still incorrectly refer to male/female as gender.
+It might be better to discuss the questions on the survey and discuss possible shortcomings, rather than make a somewhat empty gesture towards today’s standard.} R1 COMMENT: The authors still incorrectly refer to male/female as gender.
 While it is clear the authors are trying to be thoughtful in how they address important issues with data collection on demographics they are not correctly doing so (see note about race/ethnicity below).
@@
 \end{description}
+
+We interpret this to mean we should change the variable name in the cleaned data, and so have done this, and then "sex" is uniformly used throughout the paper. XXX Dewi to do
 
@@
 \end{description}
+
+We have added more explanation of the two variables, and suggestions of where to use one vd the other in classroom analyses. XXX Dewi to do
 
@@
 \end{description}
+
+Using `rlm` with the `nest` and `map` functions from `tidyr`, efficiently and simply fit these models, and are easy to explain in a classroom setting.
+Similar result would result from using robustlmm, but it avoids memory and fitting issues.
+We have added more words to the text to explain this.
+XXX DC to do
 
@@
 
+This is more of a comment than a request to make a change, but we agree that this is an area that could be a good classroom discussion.
+We have added a couple of additional sentences to highlight the richness of this topic, and alternative approaches that might be explored.
+XXX DC to do
+
 \begin{description}
 \item[Associate Editor:] \textcolor{violet}{The authors also make some problematic decisions about the race (Black or not-Black) and ethnicity (Hispanic or not-Hispanic) variables.
-They suggest three categories (non-Black/non- Hispanic; Black; Hispanic) and don’t discuss the process of combining them, while recent stan- dard is (non-Black/non-Hispanic; non-Black/Hispanic; Black/non-Hispanic; Black/Hispanic).
-It’s possible that this selection was made for backwards compatibility, lack of certain demo- graphics in the sample, or some other reason but it is not communicated.
-Please review the literature surrounding the discussion of race and ethnicity (e.g., Standards for the classifica- tion of federal data on race and ethnicity).}
+They suggest three categories (non-Black/non-Hispanic; Black; Hispanic) and don’t discuss the process of combining them, while recent standard is (non-Black/non-Hispanic; non-Black/Hispanic; Black/non-Hispanic; Black/Hispanic).
+It's possible that this selection was made for backwards compatibility, lack of certain demographics in the sample, or some other reason but it is not communicated.
+Please review the literature surrounding the discussion of race and ethnicity (e.g., Standards for the classification of federal data on race and ethnicity).}
 \end{description}
+
+It would be nice if the database contained a broader description of race, but it only has a single variable for race.
+For this reason, our cleaned data reflects this, and has a single variable for race.
+XXX Dewi to do
 
@@
 
+XXX DC to do
+
 \begin{description}
@@
 
+We think it is best to keep this emphasis, because the hard work is on cleaning and tidying the data, in a reproducible manner, which is close to the mission of the special issue.
+We have added some additional sentences about lessons for statistics and data science educators.
+XXX DC to do (might need help from Emi or Nick)
+
 \begin{description}
@@
 \end{description}
+
+We have changed all to "refresh". XXX Dewi
 
@@
 
+Thanks for this, exactly right! We have changed the wording accordingly. XXX DC
+
 \begin{description}
@@
 \end{description}
+
+Same point as made by the AE, and how we have addressed it is described above.
 
@@
 \end{description}
+
+XXX DC to address
 
@@
 
+Done XXX Dewi
+
 \begin{description}
@@
 \end{description}
+
+XXX Dewi Add some words to make it clear what data this is summarising
 
@@
 \end{description}
+
+We have expanded this explanation XXX Dewi
 
@@
 
-* All of the comments on the list above have been addressed. (**Except for the Cooksey reference**)
+* All of the comments on the list above have been addressed.
 
@@
 
+We have put the capitalisation back in, and changed "WHITE" to be "NBH". XXX Dewi
+
 \begin{description}

### 2022-05-10 paper/paper-jsdse/paper-jsdse.Rmd
Change the blinded to unblinded version.

--- 2022-01-14 paper/paper-jsdse/paper-jsdse.Rmd
@@
     location: Nedlands, WA 6009
-    email: ""
-blind: 1 # 0 for no
+    email: "nicholas.tierney@gmail.com"
+blind: 0 # 0 for no
 bibliography: ../references.bib

### 2022-05-10 paper/paper-jsdse/paper-jsdse.tex
Change the blinded to unblinded version.

@@
 We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching.
-Three new data sets and the code to produce them are provided in the open source R package called {[}CENSORED{]}.
+Three new data sets and the code to produce them are provided in the open source R package called \texttt{yowie}.
 \end{abstract}
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at CENSORED.
+The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
 
@@
 Tidy data shouldn't be confused with ``tame data,'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
-The resulting (tame) data is provided in a new R package called \texttt{[CENSORED]}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
+The resulting (tame) data is provided in a new R package called \texttt{yowie}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
 
@@
 For example, we use a single categorical race variable instead of the two binary race variables.
-We also plan to include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked.
+We also include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked.
 We do not attempt to re-create the unemployment rate variable.
 
-The plan is to create three datasets as follows:
+We aim to create three datasets as follows:
 
@@
 \item
-  The high school dropouts' wages data is closest to a refreshed version of Singer and Willett (2003)'s data.
+  The subset of wages data in (1), which is the wages of high school dropouts' as a refreshed version of Singer and Willett (2003)'s data.
 \end{enumerate}
@@
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
-The code for tidying the data are demonstrated at \url{CENSORED/articles/raw-to-input-data.html} but also described in the subsequent subsections.
+The code for tidying the data are demonstrated at \url{https://github.com/numbats/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.
 
@@
 2020).
-From a statistical perspective, this sometimes difficult the adjustment of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex (Kennedy et al.
+From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex (Kennedy et al.
 2020).
@@
 
-Figure \ref{fig:feature-plot}s shows an alternative way to check the data quality.
+Figure \ref{fig:feature-plot} shows an alternative way to check the data quality.
 Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to \$60,000/hour) for some individuals, mostly around 1990.
@@
 However, when we tested this, the fit failed to converge.
-The full code for this is shown at \url{CENSORED/articles/input-to-valid-data.html} but also described in detail next.
+The full code for this is shown at \url{https://github.com/numbats/yowie/articles/input-to-valid-data.html} but also described in detail next.
 
@@
 
-\caption{The stages of data cleaning from the raw data to get three datasets contained in \texttt{[CENSORED]}.
+\caption{The stages of data cleaning from the raw data to get three datasets contained in \texttt{yowie}.
 ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data.
 The color represents the stage of data cleaning in the statistical value chain (van der Loo and de Jonge 2021).
-Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart-blind}
+Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart}
 \end{figure}
@@
 Department of Labor 2021b), we find several variables are tagged as work experience-related variables.
-One of them is the weeks of worked since the last interview.
+One of them is the weeks worked since the last interview.
 This is used to calculate the variable.
@@
 This is not done for the refreshed data because we plan to keep refreshing it as the data is added to and released from the survey.
-Instead, we have provided a function in the R package, \texttt{[CENSORED]}, for users to conduct the inflation adjustment when they are ready to analyze the data.
+Instead, we have provided a function in the R package, \texttt{yowie}, for users to conduct the inflation adjustment when they are ready to analyze the data.
 
@@
 The refreshed data is compared with the original set using a variety of numerical summaries and graphics.
-The current subset is made available in a new R package, called \texttt{[CENSORED]}.
+The current subset is made available in a new R package, called \texttt{yowie}.
 
@@
 Readers could use this to make different decisions or provide suggestions through the package for better choices.
-Future updates of the \texttt{[CENSORED]} package may contain additional variables, or filters of the full set if it is deemed important.
+Future updates of the \texttt{yowie} package may contain additional variables, or filters of the full set if it is deemed important.
 
@@
 \item
-  \textbf{Codes}: R script to reproduce data tidying and cleaning is available in this \href{CENSORED/articles/process-data.html}{page}.
-\item
-  \textbf{R Package}:\texttt{[CENSORED]} is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort.
-This package can be accessed \href{CENSORED}{here}.
+  \textbf{Codes}: R script to reproduce data tidying and cleaning is available in this \href{https://github.com/numbats/yowie/articles/process-data.html}{page}.
+\item
+  \textbf{R Package}:\texttt{yowie} is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort.
+This package can be accessed \href{https://github.com/numbats/yowie}{here}.
 \item
   \textbf{shiny app}: An interactive \texttt{shiny} web app to visualise the effect of selecting different weight threshold for substituting the wages data to its predicted value from a fit of the robust linear regression model.
-This app can be accessed \href{CENSORED}{here} with the source code provided \href{CENSORED/tree/master/app}{here}.
+This app can be accessed \href{https://ebsmonash.shinyapps.io/yowie_app/}{here} with the source code provided \href{https://github.com/numbats/yowie/tree/master/app}{here}.
 \end{itemize}
@@
 2017.
-``Using the National Longitudinal Surveys of Youth (Nlsy) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
+``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
 Lerner Neal Halfon Christoper B.

### 2022-05-10 paper/paper.Rmd
Change the blinded to unblinded version.

@@
 We have been diligent to provide a reproducible workflow for others to follow, which also hopefully inspires more attempts at refreshing data for teaching.
-Three new data sets and the code to produce them are provided in the open source R package called [CENSORED].
+Three new data sets and the code to produce them are provided in the open source R package called `yowie`.
 ---
@@
 ```{r censor, include = FALSE}
-CENSOR <- TRUE
+CENSOR <- FALSE
 yowie <- ifelse(CENSOR, "\\texttt{[CENSORED]}", "\\texttt{yowie}")

### 2022-05-10 response/response1.tex
Change the blinded to unblinded version.

@@
 
-We thank the Editor and the three reviewers for their comments, which have helped substantially to improve the manuscript.
-In response, we have made considerable changes to the main manuscript as outlined below.
+We thank the Editor, Associate Editor and the three reviewers for their extensive comments, which have helped substantially to improve the manuscript.
+In response, we have made a major revision of the main manuscript as outlined below.
 
@@
 
-Reviewers suggested reframing the problem space so there is more emphasis on the case study.
+The editors and reviewers suggest reframing so that readers can help their students to experience the journal from wild to textbook data and either emphasize the case study, for the paper to be a regular submission for JSDSE, or to better explain the relevance for teachers and/or students in statistics and data science to be considered for the special issue.
+We have chosen to do the latter, because we hope that the paper fits into the goals of the special issue.
 
 \begin{description}
-\item[Reviewer 1:] \textcolor{violet}{How is this manuscript relevant for teachers and/or students in statistics and data science?}
+\item[Editor and Reviewer 1:] \textcolor{violet}{How is this manuscript relevant for teachers and/or students in statistics and data science?}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  The introduction has been re-written with an emphasis on how this work should be interesting for teachers and students of data science and statistics.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 1:] \textcolor{violet}{I don't see the direct relevance of this manuscript with the readership of JSDSE.
+I think this could be a valuable paper to describe the author(s) R package and would have value for researchers using the NLSY79 dataset.
+I am not convinced that JSDSE is the appropriate venue for this.
+I do not see this manuscript being engaging enough to be used in the classroom as a case study in data cleaning.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  It is commonly said 80\% of a data analysts/scientist's work is on data cleaning/preperation so we believe that shedding more light in this aspect of work is an important for students to be aware of.
+We believe that there are often subjective decisions made in the process (as we have done) which can be used as a conversation for transparency and reproducibility in class based on a real case study.
+\end{itemize}
+
+\begin{description}
 \item[Reviewer 3:] \textcolor{violet}{The authors prominently cite (Chatfield 1985) for the definition of the term “IDA”.
@@
 It seems curious to me that the whole paper is framed as being “IDA” based on the author’s unsupported statement that data cleaning should be considered party of IDA.}
-\item[Reviewer 3:] \textcolor{violet}{My suggestion is that the authors reconfigure this paper to be presented as a pure case study.
-Ibelieve the work that has been done is valuable: the data cleaning process is often messy andunplanned, so a case study like this on a popular dataset would provide an excellent educational resource.}
 \end{description}
 
-Supporting information for the lack of data cleaning in papers.
+\begin{itemize}
+\tightlist
+\item
+  Yes, there is grey area in these three activities.
+Thanks for pointing out Chatfield's original viewpoint, which differs from Huebner et al's.
+We have revised this paragraph to de-emphasize cleaning as being part of IDA.
+We have kept the explanation of the terminology of IDA and EDA because we think this is important to clarify for teaching purposes.
+\end{itemize}
 
@@
 
-Outlier detection
+\begin{itemize}
+\tightlist
+\item
+  We agree that there is substantial literature on data cleaning.
+We have referenced one of the main works, Dasu and Johnson (2003) but don't feel it is necessary to cite more than this.
+We have similarly only pointed to one main reference for IDA and EDA.
+\end{itemize}
 
@@
 
+\begin{itemize}
+\tightlist
+\item
+  The approach is for the case study only and is not meant to be extrapolated for other cases.
+We suggest using a robust mixed model as a better alternative, however, our attempt to fit such a model via \texttt{robustlmm} resulted in computational issues.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{The authors explain the complication of the “years of job experience” variable being unavailable in the raw data.
+This information is available in the NLSY79 data accessor; it seems to simply have been not selected in the initial data download by the authors.
+Given that the stated goal is to recreate the textbook data, this seems like a glaring oversight.
+It also leads to Fig 7 being essentially meaningless: we are unsurprised that salaries are higher with more years of experience, and we are unsurprised that salaries go up over the years with both age of employee and inflation.
+The positive slope of these two different plots does not mean they are related to each other in any way.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
   We thank the reviewer for pointing out to us that the data contains information on the work experience -- this additional data is added in the extraction of the data in Figure 1.
+We also add a Subsection dedicated to explain this variable, i.e., Subsection 3.2.
+Calculated variables: work experience.
 This information was compared with Singer \& Willet's original data.
-Comparison shown in Figure X shows that the information does not match up, however, there is a high correlation between the variables.
+Comparison between the subsets has been completely revised.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:] \textcolor{violet}{My suggestion is that the authors reconfigure this paper to be presented as a pure case study.
+I believe the work that has been done is valuable: the data cleaning process is often messy and unplanned, so a case study like this on a popular dataset would provide an excellent educational resource.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have kept the case study frame of mind but have re-cast the paper so that it focuses on the importance of this type of activity for teaching of statistics and data science.
+\end{itemize}
 
@@
 \item
-  We have corrected all minor grammatical errors pointed out by the reviewers.\\
-\item
-  We have made the descriptions of the steps more explicit.
+  We have corrected all minor grammatical errors pointed out by the reviewers.
 \end{itemize}
@@
 \item
-  We have fixed the sentences and the reference.
+  We have fixed the sentences and the reference, it is reflected on the last paragraph in page 3.
 \end{itemize}
 
-\textbackslash begin\{description\}
-
-\item[Reviewer 1:]
-
-\textbackslash textcolor\{violet\}\{What is dplyr used for?
+\begin{description}
+\item[Reviewer 1:]\textcolor{violet}{What is dplyr used for?
 Are there tidyverse packages used but not mentioned?
@@
 Also, I haven't tried it in code, but I imagine that the data could be tidied as described just using pivot\_longer, at least for the job number, year, and wage data.
-In other words, I don't see why dplyr or stringr would be needed for the data described on page 6.\}
-\textbackslash end\{description\}
+In other words, I don't see why dplyr or stringr would be needed for the data described on page 6.}
+\end{description}
 
@@
 \item
-  We have split the sentences and described the use of each package mentioned.
+  We did not go into great detail about the packages used as the whole code is available as a vignette within the package.
+Also, for the blind review purpose, we have made all of the code and the raw dataset available to reviewers and it is not depend on the R package we created.
+The reviewers can see the data processing code and run it to also see the reproducibility of the code.
+Besides, in the paper, we have split the sentences and described the use of each package mentioned.
 \end{itemize}
 
-\textbackslash begin\{description\}
-
-\item[Reviewer 1:]
-
-\textbackslash textcolor\{violet\}\{please clarify.
-\enquote{If either the hourly wage or hours worked is missing, we do not tally this.} I take \enquote{this} to mean \enquote{number\_of\_jobs}.
+\begin{description}
+\item[Reviewer 1:]\textcolor{violet}{please clarify.
+"If either the hourly wage or hours worked is missing, we do not tally this." I take "this" to mean "number\_of\_jobs".
 But in row 2, total\_hours is missing, yet number\_of\_jobs is 1.
-The number of jobs (1) was tallied even though hours worked was missing.\}
-\textbackslash end\{description\}
+The number of jobs (1) was tallied even though hours worked was missing.}
+\end{description}
 
@@
 \begin{description}
-\item[Reviewer 2:] \textcolor{violet}{My primary comment is that many elements need to be more explicitly spelled out.}
+\item[Reviewer 1:]\textcolor{violet}{"to find the anomaly in the wages values" This comes as a surprise to the reader.
+What anomaly?
+The wage anomalies should be mentioned in the introduction so as not to be a surprise at this point in the manuscript.}
 \end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have added in Introduction that \enquote{Section 3 presents the steps \ldots{} to find and repair anomalies.}.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 1:]\textcolor{violet}{I don't know that Table 1 is necessary as I don't have anything to compare it to.
+If the numbers check out, a simple statement to that effect would suffice.
+Also, what does "(?CHECK)" mean?
+This needs to be resolved.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have removed \enquote{(?CHECK)} in Table 1's caption. We also have stated in a sentences after the table that Table 1 reflects the consistency between the data we had and the database.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 1:]\textcolor{violet}{for example, ID 39 experienced an unusual wage only in one year.' That's one way to fix this sentence, which needs to be rewritten.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have rewritten this sentences.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 1:]\textcolor{violet}{How many individuals up to 8th grade are in the refreshed dataset? What are plausible reasons for these differences?}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  There are 108 individuals up to 8th grade in the refreshed dataset, compared to 366 individuals in the original dataset.
+The plausible reason for this difference is that some people have had GED after the period covered in the original dataset.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 1:]\textcolor{violet}{what does the natural log have to do with the y-axis?}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  Sometimes, the natural log is used to rescale the variable in the graph to adjust with the skewness in the data.
+Singer and Willet (2003) use the natural log to store the wages data by default, although without mention their motivation of doing so.
+Thus, we also use the natural log in Figure 7 to compare the refreshed data with the original data.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{My primary comment is that many elements need to be more explicitly spelled out.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have it more explicit as per the reviewers' comments.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{I would like more discussion about the textbook selected as the exemplar.
+It sounds like this dataset is used in many books, why choose that one?
+The text should include the name of the book, so readers do not need to flip to the citations in order to learn what it was.
+From the name, I am guessing that the book focuses heavily on this dataset, which is why it makes sense to reproduce their subset.
+Be explicit about this.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have mentioned the book's title in the Introduction. The motivation of refreshing this text book data and how it is used for teaching purpose are discussed in the Introduction of the paper.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{When listing the files that come zipped from the data source, be more explicit about the file formats. Comma separated values instead of csv, etc.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have made the file format more explicit.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Before the data cleaning section I would love a glimpse into the future to see what the target dataset would look like.
+What are the observations?
+What are the variables?
+This would allow me to follow the narrative of data wrangling more easily.
+Perhaps this could be put into section 2.2, which is titled Target data.
+Then, when starting to describe the raw data in 3,1, explain the same pieces and explicitly explain why the raw data it is not tidy to begin with.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have added the explanation of the observations and the variable in Section 2.2.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{With the IDA about wage data, I would appreciate a bit more context.
+Why sample 20 observations?
+I am not a longitudinal data expert, so I was a bit surprised when I flipped to the plot and found it to be small-multiple time series, although upon further consideration it makes sense.
+In that section it would be good to describe what the figure shows in a bit more detail.
+Perhaps “We randomly sample 36 respondents from the data, and plot their average wage as a time series.
+Looking at the patterns over time, we see a lot of variability in wages.
+For example, the people shown in panels 5, 7, and 11 have [explain what is interesting here, again I’m not an expert].” The next sentence here, “Some have had flat wages for years but had a sudden increase in one particular year, then it gone down again, while the others experienced an upsurge in their wage, for instance, the IDs in panel 9.” does not seem to provide much additional information.
+What were you looking for here?
+Overall trends over time, increasing/decreasing?
+Any places where one year’s wage looked really different?}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have changed Fig 2 to show 36 individuals, that display the range of wages experience, including some with questionable values.
+The figure caption reflects this and more details are provided in the text.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{I think there is a narrative step missing here, which is that looking at those samples led you to look at plots that show values for the wage variable over all participants and all years.
+The phrase “the summary plots” was not enough description for me to know what to expect when I flipped to Figure 3.
+Please be explicit about what the three plots were, why they were made, and what you were looking for.
+I think there is another narrative step missing about Figure 3C, which is that after you saw there were outliers you tried to identify which respondents had the extreme values.
+Again, please spell this out in the text.
+When you return to the plots for the entire dataset in Figure 5, please make parallel comments.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  This section has been re-written, and additional details provided.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{
+The next paragraph, which begins “The anomalies are also found” should perhaps be “Similar anomalies were found.” Was it the same respondents with the extremely high wages that showed extremely high number of hours of work?
+I suspect not, but please be explicit.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have made this sentences more explicit that the extremely high values were also observed in the \texttt{total\ hours\ of\ work}.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Section 3.2.1 begins “As part of the IDA, which is the model formulation, we build a robust linear regression model to treat the extreme values in the data.” This could use clarification.
+In 3.2, you said that part of IDA is “ model formulation without any formal statistical inference.” To me, that sounds like using modeling as a descriptive technique, and it’s something I would consider to be more a part of EDA.
+But then in 3.2.1, while it is perhaps not using modeling in an inferential way, the model is being used for more than just description.
+I might rephrase the first sentence as follows: “In order to treat the extreme values in the data, we built a robust linear regression model.”}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  Thank you for your suggetion.
+We have modified the description to \enquote{To treat the extreme values in the data, we build a robust linear regression model where we use the robustness weight to determine if a value should be replaced with the fitted value from the model.}
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{
+At the bottom of page 11, where you describe the model formation, more detail would be appreciated.
+You are modeling the mean hourly wage based on year.
+Why?
+Is this standard practice for robust linear regression?
+What is a “slight” residual?
+Is it defined based on absolute magnitude, or standard deviations, or something else?
+Later, you say “To minimize the risk of mistakenly identifying an outlier as an “erroneous outlier”.” Is an erroneous outlier a technical term?
+If not, I think you are trying not to identify erroneous outliers, not mistakenly identifying outliers as erroneous outliers (that’s a double negative).
+The value of 0.12 feels very specific and out of place in a paper that has had few numerals.
+Is this a number an expert in RLR would be able to understand?
+Again, is that a raw value?
+Is it a standard deviation?
+Is there literature about how to choose a threshold?
+Overall, the sentence “We find that 0.12 is the most reasonable value to be the threshold to minimize that drawback’s risk because it still captures the sensible spikes in the data.” is vague and should be expanded upon.
+What is “that drawback’s risk”?
+What are “the sensible spikes in the data?” Be more explicit.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  Thank you for your enquiry.
+We have re-written this section to better describe our process and intent.
+In particular, the threshold values are chosen based on extensive eye-balling of the plots of wages over time for each individual.
+We try to maintain a balance of not overly smoothing the observations but clearly repairing unlikely wage values.
+As alluded in the section, we believe that an alternative approach such as robust linear mixed models is more appropriate but we did not proceed with this as we had issues fitting the model with \texttt{robustlmm} package.
+In addition, we thought that the use of robust linear models is easier to explain in the context of teaching and the use of non-optimal approach can generate a good discussion about alternative approaches in class.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{The choice of 0.12 seems arbitrary at best.
+I don’t see any justification for this approach (e.g., literature, simulations, etc).
+Why not use bi-square, instead of huber, and simply impute for observations with weight 0.
+While still arbitrary, it’s decided by the bisquare model and isn’t a guesstimate.
+Further, it would be sensible for any dataset and not just this one.
+(This would correspond to a smaller threshold, which may catch too many strange outcomes).
+In any case, it’s hard to justify something so adhoc.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  Similar comment to this is addressed above.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{You’re using iteratively reweighted least squares with Huber weighting.} 
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  Thank you, we have changed from \enquote{M estimation} to \enquote{iteratively reweighted least squares}.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{I’m not sure this is accurate.
+The robustlmm package in R fits these models rather quickly with thousands of observations.
+What are the authors using?
+Does it run out of memory (requires more RAM) or does it take too much time?}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  The \texttt{robustlmm} model would not converge. We have added some text to the paper on this.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{What does “It implies that the fluctuation can still be observed in the data after the treatment.” mean?
+I suspect this sentence should be replaced with something along the lines of “The figure shows that fluctuations in wages still exist even in the imputed data.”}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have replaced this sentences as per the suggestion.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{In Section 4, you talk about the variables in the original data.
+“One would expect that there is a record of the day the individual first started a job, and this is used to adjust the year of collection.” I’m not sure what this means.
+Please clarify.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We removed this sentence as we already have the \texttt{stwork} and \texttt{yr\_wforce} variables which are the year of individual start working and length of the years of individual entering the labor force, respectively.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{“The treatment of unlikely wages differ in the refreshed data.” I think this means “The treatment of anomalous wages differs between the original and refreshed data sets.” Then you say “We opted to use the weights from a robust linear regression to determine what should be set as missing value as described in Section 3.2.1.” I don’t think you were setting missing values though, were you?
+It was used to determine which values should be imputed, correct?
+Either way, please clarify.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  Yes, you are correct. We imputed the extreme wages with its predicted values instead of setting it as missing values. We have clarified this in the paper.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{The bulleted difficulties in Section 5 are a bit jarring.
+Make them parallel (i.e.
+every bullet should start with a verb -ing so it is determining, calculating, etc) or turn them into a paragraph.
+I liked the specificity of \$30,000/hour but it would be better up in the section where you discuss the anomalous values.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have added the specifity of the extremely high wages in Subsection 3.3. Initial Data Analysis. We have also edited every bullet so it started with verb -ing.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{At the end of page 14, it would be good to explicitly say that if people don’t like your decisions, they can grab the code and adjust the parameters themselves!
+It’s implied, but should be made explicit.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  Done
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Figure 2’s caption should include something about how that data is the raw values from the survey, before inputing anomalous values so the figure and caption can stand alone.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have edited the caption.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{For both Figure 3 and Figure 5, I would prefer the letter label before the description, but that’s a matter of personal preference.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have moved the letter label before the description.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{For Figure 7, why did you choose to show the entire dataset here? It would be easier to make a comparison if the plot was cut off at 1994 on the x-axis.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have filtered the x-axis of the refreshed data set until 1994, so they are comparable.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Table 1 includes the note ?CHECK which I assume was a note to self.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have removed it from the caption.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Table 2 seems to have more precision than is necessary.
+Certainly 100\% does not need to be reported as 100.00\%, and probably all the percentages could be rounded to the nearest 1 percent.
+The caption could explain this, as there might be some rounding errors.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We agree that this level of precision was not necessary. We have modified Table 2 so percentages are to the nearest 1 percent and total percentage (=100\%) is omitted.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{What does “yearly mean hourly wages with years of workforce experience” mean? Is that one variable or two?}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  These are two different variables. We have made this easier to follow in the manuscript as we add explanation on what the target data is (See Subsection 2.2: Target Data).
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{Was your goal to collect data from 1994-2018 or 1979-2018?
+Page 4 states 1994 onward, but from the figures later it appears that you collected data starting in 1979 (which makes more sense to me).}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  The original data covers 1979-1994 period, while in the refreshed data, we extended the period covered until 2018 (so that the data is from 1979-2018). We have made this explicit in Section 2.2.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 2:]\textcolor{violet}{I was confused by the following section: “We choose to use this revised May data because it seemed to have less missing and presumably has been checked.
+However, there is no revised May data for 2012, 2014, 2016, and 2018, thus, we use the ordinary May data for these years.” I was assuming the data was collected on a yearly basis, but this section suggests it was collected each month.
+Please clarify somewhere in the paper.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  The database has the variable of \texttt{the\ highest\ grade\ completed\ ever}.
+Thus, we use this variable for \texttt{hgc} instead of calculating it from the\texttt{hgc} the respondents answer for each round of the survey.
+Besides, we also add new variable \texttt{grade}, which is the highest grade completed for each round of survey, so the value changes corresponding to the year of the survey.
+According to the database, the \texttt{hgc} variable is revised as in May 1st of each survey year.
+Thus, we mentioned as the revised May data.
+In the revised version, we only mention it as revised data to avoid confusion.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Abstract: The abstract portrays this work as ”re-creating a textbook example data set.” This doesn’t appear to be quite accurate because there are several notable differences between the end result and the textbook data.
+Currently, it’s more about demonstrating a procedure for “initial data analysis.”}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  The abstract has been re-written to reflect the re-framing of the paper.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Where is IDA undervalued and neglected? Research? Teaching? Both?}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We removed this sentences, and edited the explanation about IDA in Paragraph 6 of the Introduction.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{The authors discuss refreshing an example from Singer and Willet (2003), but this isn’t ever completed.
+They compare the data, but not the longitudinal model.
+Again, it’s more about demonstrating a procedure for “initial data analysis” than it is reproducing Singer and Willet (2003).}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  Our main aim in this paper is to refresh the data, which is used as one of examples of longitudinal data and its modelling in Singer and Willet (2003).
+In other words, we only aims to refresh the data until the most recent survey period, and not the longitudinal model, so that it is more relevant for today's context, especially, if its is used for teaching.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Tidy and tame data are discussed in passing and the utility of tidy and tame data are not obvious.
+The authors later describe tidy data rather clearly (page 7) but tame data never receives the same treatment.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We put explanation on the difference between tidy and tame data in the Introduction (Paragraph 7). Tame data is sort of a result product, in which, in order to produce it, we need to tidy the data.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{A historical context of NLSY79 would be helpful.
+Specifically, the dropped subjects are mentioned but there’s no context of why.
+By the argument of the paper, these steps should be transparent and well-considered.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We have added the reason of why some subject were dropped from the interview in Section 2.1.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{The author switches between gender (man/woman/etc) and sex (male/female/etc.) throughout the manuscript.
+Further, the explanation here covers important ideas in teaching how to use data on gender/sex and race, but it’s just the beginning of ideas.
+It might be better to discuss the questions on the survey and discuss possible shortcomings, rather than make a somewhat empty gesture towards today’s standard.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  The paper has been fixed so that male/female is used consistently. We have also rewritten the paragraph to focus on the survey questions and shortcomings.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{The discussion around highest grade completed is mechanical in what was done, but lacks substantive reasoning for those decisions and possible downstream consequences.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  The variable of \texttt{highest\ grade\ completed}, besides it is available in the original data, it is also use to filter the high school dropouts (the subset of NLSY79 cohort that is contained in the original data).
+We explain how we use this variable, along with a newly added variable, GED, and possible downstream consequences in Subsection 4.1 Filtering: Determining who is a dropout.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{It appears the goal of the paper switches to just IDA and dealing with problematic points here. This is completed in the paper.} 
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  Paper has been re-written to change focus.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Gender (man/woman/etc). Sex (male/female/etc).} 
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  See explanation above
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{It is not clear why participants without a high school degree are not part of the analysis.} 
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  In the current version of the manuscript, we have included all of the cohort, regardless of their school degree in a dataset called \texttt{wages}.
+However, we also subset the data to be high school dropouts only as we aims to refresh Singer and Willet's data (only contains male high school dropouts aged 14-17 years old when first interviewed).
+Hence, for comparability, we only analyse the high school dropouts in Section 4.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Figure 2 is helpful for observing the data there is a lot of variability in both hourly wage and trends.
+A discussion about censored data might be helpful here – this is not something you impute from the data.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  Added some additional text on this.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{It’s unclear about what fluctuation or treatment the author is referring to.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  Fluctuation is the variability in the data (up and down) in the wages, while treatment is imputing the extremely high wages (anomaly) with their RLM's predicted value.
+The detail explanation of fluctation is discussed in Subsection 3.3 Initial Data Analysis, while the treatment is explained in Subsection 3.4 (Paragraph 3).
+We also have edited this sentences to avoid implicity.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{If the goal was to reproduce the textbook data, then why weren’t wages inflation-adjusted?}
+\end{description}
+
+\begin{itemize}
+\item
+  This is not done for the refreshed data because we plan to keep refreshing it as the data from the newest survey period release.
+Thus, the 1990 prices might not be relevant anymore for the future context.
+Hence, we keep it as the original value to allow the users to treat it as their needs.
+In other words, the inflation-adjustment of the wages is better to do with each wave of new data added, so that it is relative to the last date in the data.
+\item
+  This explanation exists in the manuscript in Section 4.2 and Section 5.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{Figure 7 (Figure 8 in the current manuscript) is interesting, but it wasn’t until my second time through that I realized this was the textbook example for replicating.
+There is a clear difference in analyses (different question really), but is there anything to justify a similarity, or improvement of the analysis?}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  The comparison of the original and refreshed data has been completely rewritten.
+There are still some differences, but with the change in variables being created and better filtering to get the dropouts the subsets are now more similar.
+\end{itemize}
+
+\begin{description}
+\item[Reviewer 3:]\textcolor{violet}{(page 14, line 46) Thank you for this acknowledgment.
+I think this is true for most analyses.
+It doesn’t mean that you shouldn’t transparently and fully investigate your own choices.}
+\end{description}
+
+\begin{itemize}
+\tightlist
+\item
+  We are really not sure what this refers to. Perhaps it is solved with the revisions.
+\end{itemize}
 

### 2022-05-10 paper/paper-jsdse/paper-jsdse.Rmd
DC completed the first two of her changes

--- 2022-01-14 paper/paper-jsdse/paper-jsdse.Rmd
@@
     location: Nedlands, WA 6009
-    email: ""
-blind: 1 # 0 for no
+    email: "nicholas.tierney@gmail.com"
+blind: 0 # 1 for yes
 bibliography: ../references.bib

### 2022-05-10 paper/paper-jsdse/paper-jsdse.tex
DC completed the first two of her changes

--- 2022-05-04 paper/paper-jsdse/paper-jsdse.tex
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis,'' which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models.
+An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis,'' which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models.
 The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
@@
 It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis (a side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)).
-We argue the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
+We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
 
@@
 
-Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
+Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham et al.
 2019).
@@
 
-In the context of official statistics, van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+In the context of official statistics, M.
+van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
 What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
@@
 As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year.
-The raw data consequently has a large number of columns (770 to be specific).
+The raw data consequently has a large number of columns (742 to be specific).
 The values in the cell under the variables begin with \texttt{HRP} correspond to the hourly wage in dollars.
@@
 \begin{verbatim}
-#> 'data.frame':    12686 obs. of  770 variables:
+#> 'data.frame':    12686 obs. of  742 variables:
 #>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 ...
@@
 Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
-This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
-2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham, François, et al.
+This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
+2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham et al.
 2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
@@
 An alternative approach would be a robust linear mixed model using \texttt{robustlmm} (Koller 2016).
-However, when we tested this, the fit failed to converge.
-The full code for this is shown at \url{CENSORED/articles/input-to-valid-data.html} but also described in detail next.
-
-The \texttt{mean\_hourly\_wage} and \texttt{year} are set as the dependent and predictor, respectively.
-Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021).
-The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
-It should be noted that it is not possible to determine if all outliers are errors, and it might be an individual indeed had abnormally high wages at a particular time.
-
-To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done.
-Samples of individuals were examined under the different thresholds with an eye to the smoothness of the profiles.
+
+Figure \ref{fig:compare-plot} shows the profiles for a sample of 30 individuals the mean hourly wage before and after extreme values are replaced.
+The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
+
+The challenging part of detecting an anomaly using the robustness weight is determining the weight threshold beyond which the observations are considered outliers.
+It is also important not to be too aggressive in outlier removal, because this would overly smooth the data, and ignore actual workforce experiences important wage volatility such as, abnormally high wages for a fleeting amount of time.
+To explore the risk of being overly vigorous in labeling observations as outliers, a range of threshold value were assessed.
+Numerous samples of individuals were drawn for each threshold, and the smoothness of the profiles was examined.
 Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
-A threshold of 0.12 was chosen.
-That struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
-Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.
-We then flag those observations in a new variable called \texttt{is\_pred}, so that this change can be monitored in the downstream analyses.
-
-Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
-The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
+This was achieved using a purpose written \texttt{shiny} (Chang et al.
+2020) app (provided with the code of this paper).
+Observations where a change was made are tracked with a new variable called \texttt{is\_pred}, so that this effect in the downstream analyses can be monitored.
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{../figures/flowchart_blind}
 
@@
 ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data.
-The color represents the stage of data cleaning in the statistical value chain (van der Loo and de Jonge 2021).
+The color represents the stage of data cleaning in the statistical value chain (M.
+P.
+J.
+van der Loo and de Jonge 2021).
 Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart-blind}
@@
 
-The data cleaning process is documented, and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
+The data cleaning process is documented, and the code has been made available.
+These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
 Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a \texttt{shiny} (Chang et al.
-2020) app, and the choices used in the refreshed data are documented.
+2020) app, which is provided with the code, and the choices used in the refreshed data are documented.
 
@@
 
-Why is this exercise important for teachers and students?
-This work illustrates the steps in cleaning and processing data in preparation for it to be a textbook data set.
-Choices made during the cleaning and processing can affect findings made with the data, and these should be transparent.
-Along with the textbook data, which is now provided as an R package, the documentation of the process provides some examples for teaching data cleaning, initial data analysis, and exploratory data analysis.
-The refreshed textbook data provides a resource for teaching longitudinal data analysis.
+Why is the work presented here useful for teachers and students? There are several reasons:
+
+\begin{itemize}
+\tightlist
+\item
+  It primarily illustrates the steps in cleaning and processing a well-known longitudinal data so that it can be further refreshed as a textbook data set as new data emerges.
+Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
+\item
+  Some of the methods shown, and lessons learned about the data, can also be considered generally for data cleaning, initial data analysis and exploratory data analysis, useful for working with other data sets.
+\item
+  The refreshed data can be used for teaching advanced linear modeling, using almost identical code from Singer and Willett (2003), and thus make those lessons more current in time.
+\item
+  It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience.
+The vignettes of the \texttt{brolgar} (Tierney, Cook, and Prvan 2020) package provide useful guides for the case study content.
+\end{itemize}
 
@@
   \textbf{Codes}: R script to reproduce data tidying and cleaning is available in this \href{CENSORED/articles/process-data.html}{page}.
+The code for the extreme value handling with robust linear fits is in \url{CENSORED/articles/input-to-valid-data.html}.
 \item
@@
 2021a.
-``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
+{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
 Columbus, OH, through \url{https://www.nlsinfo.org/bibliography-citing-nls-data}.
 
-\leavevmode\hypertarget{ref-nlsy79guide}{}%
----------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.'' \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
-
-\leavevmode\hypertarget{ref-shiny}{}%
-Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.
-
-\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
-Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
-
-\leavevmode\hypertarget{ref-eliznlsy}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-nlsy79guide}{}}%
+---------. 2021b. {``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.''} \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-shiny}{}}%
+Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{{shiny: Web Application Framework for R}}. \url{https://CRAN.R-project.org/package=shiny}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chatfield1985TIEo}{}}%
+Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
+
+\leavevmode\vadjust pre{\hypertarget{ref-eliznlsy}{}}%
 Cooksey, Elizabeth C.
 2017.
-``Using the National Longitudinal Surveys of Youth (Nlsy) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
+{``Using the National Longitudinal Surveys of Youth ({NLSY}) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M.
 Lerner Neal Halfon Christoper B.
@@
 Cham: Springer.
-\url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.
-
-\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
+https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-DasuTamraparni2003Edma}{}}%
 Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.
 
-\leavevmode\hypertarget{ref-janitor}{}%
-Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.
-
-\leavevmode\hypertarget{ref-racismnotrace}{}%
-Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.
-
-\leavevmode\hypertarget{ref-grimshaw}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-janitor}{}}%
+Firke, Sam. 2020. \emph{{janitor: Simple Tools for Examining and Cleaning Dirty Data}}. \url{https://CRAN.R-project.org/package=janitor}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-racismnotrace}{}}%
+Fullilove, M. T. 1998. {``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.
+
+\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
 Grimshaw, Scott D.
 2015.
-``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14.
+{``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14.
 \url{https://doi.org/10.1080/00031305.2015.1081106}.
 
-\leavevmode\hypertarget{ref-purrr}{}%
-Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.
-
-\leavevmode\hypertarget{ref-penguins-data}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-purrr}{}}%
+Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-penguins-data}{}}%
 Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
 
-\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
 Huebner, Marianne, Werner Vach, and Saskia le Cessie.
 2016.
-``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
-
-\leavevmode\hypertarget{ref-ilk2004}{}%
+{``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ilk2004}{}}%
 Ilk, Ozlem.
 2004.
-``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.'' PhD thesis, Iowa State University.
+{``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.''} PhD thesis, Iowa State University.
 \url{https://doi.org/10.31274/rtd-180813-11012}.
 
-\leavevmode\hypertarget{ref-kennedy2020using}{}%
-Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. ``Using Sex and Gender in Survey Adjustment.'' \url{http://arxiv.org/abs/2009.14401}.
-
-\leavevmode\hypertarget{ref-tamedata}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-kennedy2020using}{}}%
+Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. {``Using Sex and Gender in Survey Adjustment.''} \url{https://arxiv.org/abs/2009.14401}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
 Kim, A.
@@
 2001.
-``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
-
-\leavevmode\hypertarget{ref-R}{}%
+{``Data Watch: The National Longitudinal Surveys.''} \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
 R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
 
-\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-SingerJudithD2003Alda}{}}%
 Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.
 
-\leavevmode\hypertarget{ref-notiris}{}%
-Stodel, Megan. 2020. ``Stop Using Iris.'' \url{https://www.meganstodel.com/posts/no-to-iris/}.
-
-\leavevmode\hypertarget{ref-brolgar}{}%
-Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.
-
-\leavevmode\hypertarget{ref-tukey}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-notiris}{}}%
+Stodel, Megan. 2020. {``Stop Using Iris.''} \url{https://www.meganstodel.com/posts/no-to-iris/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-brolgar}{}}%
+Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}}. \url{https://github.com/njtierney/brolgar}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tukey}{}}%
 Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.
 
-\leavevmode\hypertarget{ref-rlm}{}%
-UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
-
-\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
-van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.
-
-\leavevmode\hypertarget{ref-validate}{}%
-van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
-
-\leavevmode\hypertarget{ref-mass}{}%
-Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
-
-\leavevmode\hypertarget{ref-tsibble}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%
+van der Loo, Mark P. J., and Edwin de Jonge. 2021. {``Data Validation Infrastructure for {R}.''} \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-LooMarkvander2018Sdcw}{}}%
+van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in r}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mass}{}}%
+Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with s}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tsibble}{}}%
 Wang, Earo, Dianne Cook, and Rob J Hyndman.
 2020.
-``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
+{``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.''} \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
 \url{https://doi.org/10.1080/10618600.2019.1695624}.
 
-\leavevmode\hypertarget{ref-plyr}{}%
-Wickham, Hadley. 2011. ``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
-
-\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
----------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.
-
-\leavevmode\hypertarget{ref-ggplot2}{}%
----------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
-
-\leavevmode\hypertarget{ref-stringr}{}%
----------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.
-
-\leavevmode\hypertarget{ref-tidyr}{}%
----------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.
-
-\leavevmode\hypertarget{ref-tidyverse}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
+Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-WickhamHadley2014TD}{}}%
+---------. 2014. {``Tidy Data.''} \emph{Journal of Statistical Software} 59 (10): 1--23.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
+---------. 2016. \emph{{ggplot2: Elegant Graphics for Data Analysis}}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-stringr}{}}%
+---------. 2019. \emph{{stringr: Simple, Consistent Wrappers for Common String Operations}}. \url{https://CRAN.R-project.org/package=stringr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tidyr}{}}%
+---------. 2020. \emph{{tidyr: Tidy Messy Data}}. \url{https://CRAN.R-project.org/package=tidyr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al.
 2019.
-``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686.
+{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686.
 \url{https://doi.org/10.21105/joss.01686}.
 
-\leavevmode\hypertarget{ref-dplyr}{}%
-Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.
-
-\leavevmode\hypertarget{ref-readr}{}%
-Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.
-
-\leavevmode\hypertarget{ref-nlsy79edu}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
+Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{{dplyr: A Grammar of Data Manipulation}}. \url{https://CRAN.R-project.org/package=dplyr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-readr}{}}%
+Wickham, Hadley, and Jim Hester. 2020. \emph{{readr: Read Rectangular Text Data}}. \url{https://CRAN.R-project.org/package=readr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nlsy79edu}{}}%
 Wolpin, Kenneth I.
 2005.
-``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Published by Bureau of Labor Statistics, U.S.
+{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Published by Bureau of Labor Statistics, U.S.
 Department of Labor.

### 2022-05-10 paper/paper.Rmd
DC completed the first two of her changes

--- 2022-05-04 paper/paper.Rmd
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis," which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models.
+An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis," which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models.
 The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
@@
 It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis (a side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]).
-We argue the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
+We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
 
@@
 An alternative approach would be a robust linear mixed model using `robustlmm` [@KollerManuel2016rARP].
-However, when we tested this, the fit failed to converge.
-<!--I would pull the tidy and purrr citations to the end of the sentence to make it read more smoothly-->The full code for this is shown at \url{`r yowie_pkgdown`/articles/input-to-valid-data.html} but also described in detail next.
+<!-- However, when we tested this, the fit failed to converge.
+I would pull the tidy and purrr citations to the end of the sentence to make it read more smoothly-->
 
-The `mean_hourly_wage` and `year` are set as the dependent and predictor, respectively.
+Figure \ref{fig:compare-plot} shows the profiles for a sample of 30 individuals the mean hourly wage before and after extreme values are replaced.
+The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
+<!--The `mean_hourly_wage` and `year` are set as the dependent and predictor, respectively.
 Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) [@rlm].
-The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
-It should be noted that it is not possible to determine if all outliers are errors, and it might be an individual indeed had abnormally high wages at a particular time.
+-->
 
-To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done.
-Samples of individuals were examined under the different thresholds with an eye to the smoothness of the profiles.
+The challenging part of detecting an anomaly using the robustness weight is determining the weight threshold beyond which the observations are considered outliers.
+It is also important not to be too aggressive in outlier removal, because this would overly smooth the data, and ignore actual workforce experiences important wage volatility such as, abnormally high wages for a fleeting amount of time.
+To explore the risk of being overly vigorous in labeling observations as outliers, a range of threshold value were assessed.
+Numerous samples of individuals were drawn for each threshold, and the smoothness of the profiles was examined.
 Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
-A threshold of 0.12 was chosen.
+This was achieved using a purpose written `shiny`  [@shiny] app (provided with the code of this paper).
+<!--A threshold of 0.12 was chosen.
 That struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
-Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.
-We then flag those observations in a new variable called `is_pred`, so that this change can be monitored in the downstream analyses.
+Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.--> Observations where a change was made are tracked with a new variable called `is_pred`, so that this effect in the downstream analyses can be monitored.
 
@@
 ```
-
-Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
-The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
 
@@
 
-The data cleaning process is documented, and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
-Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a `shiny`  [@shiny] app, and the choices used in the refreshed data are documented.
+The data cleaning process is documented, and the code has been made available.
+These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
+Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a `shiny`  [@shiny] app, which is provided with the code, and the choices used in the refreshed data are documented.
 
@@
 
-Why is this exercise important for teachers and students?
-This work illustrates the steps in cleaning and processing data in preparation for it to be a textbook data set.
-Choices made during the cleaning and processing can affect findings made with the data, and these should be transparent.
-Along with the textbook data, which is now provided as an R package, the documentation of the process provides some examples for teaching data cleaning, initial data analysis, and exploratory data analysis.
-The refreshed textbook data provides a resource for teaching longitudinal data analysis.
+Why is the work presented here useful for teachers and students? There are several reasons:
+
+- It primarily illustrates the steps in cleaning and processing a well-known longitudinal data so that it can be further refreshed as a textbook data set as new data emerges.
+Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
+- Some of the methods shown, and lessons learned about the data, can also be considered generally for data cleaning, initial data analysis and exploratory data analysis, useful for working with other data sets.
+- The refreshed data can be used for teaching advanced linear modeling, using almost identical code from @SingerJudithD2003Alda, and thus make those lessons more current in time.
+- It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience.
+The vignettes of the `brolgar` [@brolgar] package provide useful guides for the case study content.
 
@@
 - **Codes**: R script to reproduce data tidying and cleaning is available in this [page](`r yowie_pkgdown`/articles/process-data.html).
+The code for the extreme value handling with robust linear fits is in \url{`r yowie_pkgdown`/articles/input-to-valid-data.html}.
 

### 2022-05-10 paper/paper.tex
DC completed the first two of her changes

--- 2022-05-04 paper/paper.tex
@@
 An alternative approach would be a robust linear mixed model using \texttt{robustlmm} (Koller 2016).
-However, when we tested this, the fit failed to converge.
-The full code for this is shown at \url{CENSORED/articles/input-to-valid-data.html} but also described in detail next.
-
+
+Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
+The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
 The \texttt{mean\_hourly\_wage} and \texttt{year} are set as the dependent and predictor, respectively.
 Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021).
-The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
-It should be noted that it is not possible to determine if all outliers are errors, and it might be an individual indeed had abnormally high wages at a particular time.
-
+
+The challenging part of detecting an anomaly using the robustness weight is determining the weight threshold beyond which the observations are considered outliers.
+It is also important not to be too aggressive in outlier removal, because this would overly smooth the data, and ignore actual workforce experiences important wage volatility such as, abnormally high wages for a fleeting amount of time.
 To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done.
@@
 
-Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
-The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
-
 \begin{figure}

### 2022-05-10 response/response2.Rmd
DC completed the first two of her changes

@@
 
-The editor says \textcolor{violet}{The paper falls squarely into the call for the special issue on `Reproducibility and responsible workflow".
-Some additional suggestions regarding framing are provided by the reviewers and the AE.} The only comments on framing appear to come from the associate editor: \textcolor{violet}{If this was (1) a general process for demonstrating the cleaning of longitudinal data, (2) a case study for cleaning data in the classroom, or (3) a case study for using the data in the classroom the value would be clear.
+The editor says \textcolor{violet}{The paper falls squarely into the call for the special issue on "Reproducibility and responsible workflow".
+Some additional suggestions regarding framing are provided by the reviewers and the AE.}
+
+The only comments on framing appear to come from the associate editor: \textcolor{violet}{If this was (1) a general process for demonstrating the cleaning of longitudinal data, (2) a case study for cleaning data in the classroom, or (3) a case study for using the data in the classroom the value would be clear.
 In its current state, however, the manuscript still partly each of the three by describing the R package they've created and, as a result, does not fully address any of the three.}
 
-We would argue that the current form of the paper is clearly (2), and we have added some wording in the introduction and lesson learned sections to help make this clear. XXX DC to do
+It is not clear how to handle this.
+We interpret the comment from the Editor, that the paper is consistent with the call for the special issue.
+The AE would like it to be clearly one of three types.
+The paper does have of each of the three types, which we believe has value.
+There is some wording in the introduction ("An example of this is the wages data made public by Singer and Willett (2003) in their book, “Applied lon- gitudinal data analysis,” which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models.") and we have added additional text in Section 5 Summary, on lessons learned, to help the reader understand the elements that might be used for their teaching.
+XXX DONE
 
@@
 
-Using `rlm` with the `nest` and `map` functions from `tidyr`, efficiently and simply fit these models, and are easy to explain in a classroom setting.
-Similar result would result from using robustlmm, but it avoids memory and fitting issues.
-We have added more words to the text to explain this.
-XXX DC to do
+Using `rlm` with the `nest` and `map` functions from `tidyr`, efficiently and simply fit these models.
+They can be easily explained in a classroom setting, and equip students generally with advanced data handling skills.
+Similar results would be obtained from using robustlmm, though.
+We have removed reference to memory and convergence issues.
+We have also removed a specific reference to threshold used.
+Having the particular number used is not important for the explanation of the approach and as we have seen from the review, is distracting for the reader.
+This section has been modified so that the approach is more understandable.
+XXX DONE
 

### 2022-05-10 paper/paper-jsdse/paper-jsdse.tex

@@
 
-\caption{The stages of data cleaning from the raw data to get three datasets contained in \texttt{[CENSORED]}.
+\caption{The stages of data cleaning from the raw data to get three datasets contained in \texttt{yowie}.
 ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data.
@@
 van der Loo and de Jonge 2021).
-Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart-blind}
+Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart}
 \end{figure}
@@
 \item
-  \textbf{Codes}: R script to reproduce data tidying and cleaning is available in this \href{CENSORED/articles/process-data.html}{page}.
-The code for the extreme value handling with robust linear fits is in \url{CENSORED/articles/input-to-valid-data.html}.
-\item
-  \textbf{R Package}:\texttt{[CENSORED]} is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort.
-This package can be accessed \href{CENSORED}{here}.
+  \textbf{Codes}: R script to reproduce data tidying and cleaning is available at \url{https://github.com/numbats/yowie/articles/process-data.html}.
+The code for the extreme value handling with robust linear fits is in \url{https://github.com/numbats/yowie/articles/input-to-valid-data.html}.
+\item
+  \textbf{R Package}:\texttt{yowie} is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort.
+This package can be accessed from \url{https://github.com/numbats/yowie}.
 \item
   \textbf{shiny app}: An interactive \texttt{shiny} web app to visualise the effect of selecting different weight threshold for substituting the wages data to its predicted value from a fit of the robust linear regression model.
-This app can be accessed \href{CENSORED}{here} with the source code provided \href{CENSORED/tree/master/app}{here}.
+This app can be accessed at \url{https://ebsmonash.shinyapps.io/yowie_app/} with the source code provided \url{https://github.com/numbats/yowie/tree/master/app}.
 \end{itemize}

### 2022-05-10 paper/paper.Rmd

@@
 yowie_pkgdown <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie")
+yowie_code <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie-paper/data-raw")
 yowie_shiny <- ifelse(CENSOR, "CENSORED", "https://ebsmonash.shinyapps.io/yowie_app/")
@@
 
-- **Codes**: R script to reproduce data tidying and cleaning is available in this [page](`r yowie_pkgdown`/articles/process-data.html).
+- **Codes**: R script to reproduce data tidying and cleaning is available at \url{`r yowie_pkgdown`/articles/process-data.html}.
 The code for the extreme value handling with robust linear fits is in \url{`r yowie_pkgdown`/articles/input-to-valid-data.html}.
@@
 - **R Package**:`r yowie` is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort.
-This package can be accessed [here](`r yowie_repo`).
+This package can be accessed from \url{`r yowie_repo`}.
 
 - **shiny app**: An interactive `shiny` web app to visualise the effect of selecting different weight threshold for substituting the wages data to its predicted value from a fit of the robust linear regression model.
-This app can be accessed [here](`r yowie_shiny`) with the source code provided [here](`r yowie_repo`/tree/master/app).
+This app can be accessed at \url{`r yowie_shiny`} with the source code provided \url{`r yowie_repo`/tree/master/app}.
 

### 2022-05-10 paper/paper-jsdse/paper-jsdse.tex
Modify the blind option.

--- 2022-05-10 paper/paper-jsdse/paper-jsdse.tex
@@
 \usepackage{booktabs}
-\usepackage{float}
+%\usepackage{float}
 \usepackage{url} % not crucial - just used below for the URL
@@
 An alternative approach would be a robust linear mixed model using \texttt{robustlmm} (Koller 2016).
-However, when we tested this, the fit failed to converge.
-The full code for this is shown at \url{https://github.com/numbats/yowie/articles/input-to-valid-data.html} but also described in detail next.
-
-The \texttt{mean\_hourly\_wage} and \texttt{year} are set as the dependent and predictor, respectively.
-Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021).
-The challenging part of detecting the anomaly using the robustness weight is determining the weight threshold in which the observations are considered outliers.
-It should be noted that it is not possible to determine if all outliers are errors, and it might be an individual indeed had abnormally high wages at a particular time.
-
-To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done.
-Samples of individuals were examined under the different thresholds with an eye to the smoothness of the profiles.
+
+Figure \ref{fig:compare-plot} shows the profiles for a sample of 30 individuals the mean hourly wage before and after extreme values are replaced.
+The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
+
+The challenging part of detecting an anomaly using the robustness weight is determining the weight threshold beyond which the observations are considered outliers.
+It is also important not to be too aggressive in outlier removal, because this would overly smooth the data, and ignore actual workforce experiences important wage volatility such as, abnormally high wages for a fleeting amount of time.
+To explore the risk of being overly vigorous in labeling observations as outliers, a range of threshold value were assessed.
+Numerous samples of individuals were drawn for each threshold, and the smoothness of the profiles was examined.
 Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
-A threshold of 0.12 was chosen.
-That struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
-Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.
-We then flag those observations in a new variable called \texttt{is\_pred}, so that this change can be monitored in the downstream analyses.
-
-Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
-The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
+This was achieved using a purpose written \texttt{shiny} (Chang et al.
+2020) app (provided with the code of this paper).
+Observations where a change was made are tracked with a new variable called \texttt{is\_pred}, so that this effect in the downstream analyses can be monitored.
 
@@
 \item
-  \textbf{Codes}: R script to reproduce data tidying and cleaning is available in this \href{https://github.com/numbats/yowie/articles/process-data.html}{page}.
+  \textbf{Codes}: R script to reproduce data tidying and cleaning is available at \url{https://github.com/numbats/yowie/articles/process-data.html}.
+The code for the extreme value handling with robust linear fits is in \url{https://github.com/numbats/yowie/articles/input-to-valid-data.html}.
 \item
   \textbf{R Package}:\texttt{yowie} is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort.
-This package can be accessed \href{https://github.com/numbats/yowie}{here}.
+This package can be accessed from \url{https://github.com/numbats/yowie}.
 \item
   \textbf{shiny app}: An interactive \texttt{shiny} web app to visualise the effect of selecting different weight threshold for substituting the wages data to its predicted value from a fit of the robust linear regression model.
-This app can be accessed \href{https://ebsmonash.shinyapps.io/yowie_app/}{here} with the source code provided \href{https://github.com/numbats/yowie/tree/master/app}{here}.
+This app can be accessed at \url{https://ebsmonash.shinyapps.io/yowie_app/} with the source code provided \url{https://github.com/numbats/yowie/tree/master/app}.
 \end{itemize}
@@
 
-\leavevmode\hypertarget{ref-rlm}{}%
-UCLA: Statistical Consulting Group. 2021. ``Robust Regression \textbar{} R Data Analysis Examples.'' February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
-
 \leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%

### 2022-05-10 response/response2.Rmd

@@
 This is more of a comment than a request to make a change, but we agree that this is an area that could be a good classroom discussion.
-We have added a couple of additional sentences to highlight the richness of this topic, and alternative approaches that might be explored.
-XXX DC to do
+We have revised the section on handling extremes to make our explanation simpler, with links to the shiny app that we put together to assist in the judgement and choice of threshold.
+We hope that this provides fodder for discussion.
+It could definitely be worked into a teaching exercise.
+XXX DONE
 

### 2022-05-10 paper/paper-jsdse/paper-jsdse.Rmd
update of intro according to AE comment

--- 2022-05-10 paper/paper-jsdse/paper-jsdse.Rmd
@@
     thanks: ""
-    department: 
+    department: "Geospatial Health and Development"
     affiliation: Telethon Kids Institute
@@
     email: "nicholas.tierney@gmail.com"
-blind: 0 # 0 for no
+blind: 0 # 1 for yes
 bibliography: ../references.bib

### 2022-05-10 paper/paper-jsdse/paper-jsdse.tex
update of intro according to AE comment

--- 2022-05-10 paper/paper-jsdse/paper-jsdse.tex
@@
             Nicholas Tierney\\
-      , Telethon Kids Institute, }
+      Geospatial Health and Development, Telethon Kids Institute, }
         \maketitle
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
-
-This disparity between the average and the individual is a part of statistics as a discipline that requires more attention.
-This particular data set is a prime example of discussing this disparity.
-Textbook data sets have longevity if there is an unresolved mystery.
-The iris data (Andereson 1935) is a prime example.
-It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis (a side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)).
+It is for these reasons that the wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
+
+This disparity between the overall trend and the individual heterogeneity is what makes this data interesting.
+Textbook data sets have longevity if they have unresolved elements.
+The iris data (Anderson 1935) is a prime example.
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
+(It should be noted that the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)).
 We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
 
-For statistics and data science by association, it is increasingly important to reach the individual.
-One might describe this as a divergence of purpose -- statistics for public policy -- or statistics for the public.
+For statistics, and data science by association, it is increasingly important to reach the individual.
+One might think of this as a divergence of purpose -- statistics for public policy -- or statistics for the public.
 The two are not the same.
 As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual.
-For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is you are more than your demographic.
-The majority of people in the study do not have a pattern that is similar to the average.
-If you have a bad experience, that your wages have declined over time, you are not alone; there are others like you, and more than you think.
-A similar tone is echoed occasionally in the public media.
-For example, an article published in the Sydney Morning Herald argues there is no average Australian (Moncrief 2015).
+For example, with the wages data, the overall trend, across demographics, is a steady increase in wages, but the patterns among individuals is heterogeneous.
+Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year, and the heterogeneity of the patterns are evident across demographics.
+Subject-specific variation explains most of the variation in wages, and even though the overall trend is statistically significant, it is weak.
+The message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable.
+However, discussing the overall trend with an individual is misleading -- more than likely their pattern is completely different.
+It could even be depressing for them to compare themselves to the overall average, especially if the individual's wages have a declining trend or high volatility.
+It would be more useful for an individual to know what percentage of the population have profiles like theirs.
+We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation.
+We occasionally hear this voiced in the public media media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian (Moncrief 2015).
+Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.
 
@@
 Using tired old textbook data sets can imbue a misconception that the field is not current.
-The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated.
+The wages data is extracted from NSLY79, one of the best examples of open data (see details at Open Knowledge Foundation (2021)), which is constantly being updated.
 It should be possible to continuously refresh the textbook data from the data repository.
-This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) wild data to textbook data.
+This paper describes our (non-glamorous) journey from open wild data to textbook data.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-iris-data}{}}%
-Andereson, Edgar. 1935. {``The Irises of the Gaspe Peninsula.''} \emph{Bulletin of the American Iris Society} 59: 2--5.
+Anderson, Edgar. 1935. {``The Irises of the Gaspe Peninsula.''} \emph{Bulletin of the American Iris Society} 59: 2--5.
 

### 2022-05-10 paper/paper.Rmd
update of intro according to AE comment

@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-The wages data was used to illustrate exploratory longitudinal data analysis in @ilk2004, and was further developed into a case study for use in the teaching of exploratory data analysis at `r censor_word("Iowa State University")`.
+It is for these reasons that the wages data was used to illustrate exploratory longitudinal data analysis in @ilk2004, and was further developed into a case study for use in the teaching of exploratory data analysis at `r censor_word("Iowa State University")`.
 
@@
 
-This disparity between the average and the individual is a part of statistics as a discipline that requires more attention.
-This particular data set is a prime example of discussing this disparity.
-Textbook data sets have longevity if there is an unresolved mystery.
+This disparity between the overall trend and the individual heterogeneity is what makes this data interesting.
+Textbook data sets have longevity if they have unresolved elements.
 The iris data [@iris-data] is a prime example.
-It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis (a side note: the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]).
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
+(It should be noted that the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]).
 We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
 
-For statistics and data science by association, it is increasingly important to reach the individual.
-One might describe this as a divergence of purpose -- statistics for public policy -- or statistics for the public.
+For statistics, and data science by association, it is increasingly important to reach the individual.
+One might think of this as a divergence of purpose -- statistics for public policy -- or statistics for the public.
 The two are not the same.
 As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual.
-For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is you are more than your demographic.
-The majority of people in the study do not have a pattern that is similar to the average.
-If you have a bad experience, that your wages have declined over time, you are not alone; there are others like you, and more than you think.
-A similar tone is echoed occasionally in the public media.
-For example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage].
+For example, with the wages data, the overall trend, across demographics, is a steady increase in wages, but the patterns among individuals is heterogeneous.
+Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year, and the heterogeneity of the patterns are evident across demographics.
+Subject-specific variation explains most of the variation in wages, and even though the overall trend is statistically significant, it is weak.
+The message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable.
+However, discussing the overall trend with an individual is misleading -- more than likely their pattern is completely different.
+It could even be depressing for them to compare themselves to the overall average, especially if the individual's wages have a declining trend or high volatility.
+It would be more useful for an individual to know what percentage of the population have profiles like theirs.
+We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation.
+<!--this information is the appropriate message for an individual is that you as an individual are far more than your demographic.
+The majority of people in the study do not have a pattern that is similar to the overall trend.
+If you have experienced declining wages, you are not alone; there are others like you, and more than you think.--> We occasionally hear this voiced in the public media media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage].
+Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.
 
@@
 Using tired old textbook data sets can imbue a misconception that the field is not current.
-The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated.
+The wages data is extracted from NSLY79, one of the best examples of open data (see details at @opendata), which is constantly being updated.
 It should be possible to continuously refresh the textbook data from the data repository.
-This paper describes our (non-glamorous) journey from open [@opendata] wild data to textbook data.
+This paper describes our (non-glamorous) journey from open wild data to textbook data.
 

### 2022-05-10 paper/paper.tex
update of intro according to AE comment

@@
 
-For statistics and data science by association, it is increasingly important to reach the individual.
+For statistics, and data science by association, it is increasingly important to reach the individual.
 One might describe this as a divergence of purpose -- statistics for public policy -- or statistics for the public.
@@
 As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual.
-For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is you are more than your demographic.
-The majority of people in the study do not have a pattern that is similar to the average.
-If you have a bad experience, that your wages have declined over time, you are not alone; there are others like you, and more than you think.
+For example, with the wages data, the overall trend, across demographics, is a steady increase in wages, but the patterns among individuals is heterogeneous.
+Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year.
+The heterogeneous patterns occur across demographics.
+The subject-specific variation explains most of the variation in wages, even though the overall trend is statistically significant, it is weak.
+The the message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable.
+However, this information is the appropriate message for an individual is that you as an individual are far more than your demographic.
+The majority of people in the study do not have a pattern that is similar to the overall trend.
+If you have experienced declining wages, you are not alone; there are others like you, and more than you think.
 A similar tone is echoed occasionally in the public media.
@@
 
-Figure \ref{fig:compare-plot} shows the mean hourly wage before and after the extreme values are replaced.
+Figure \ref{fig:compare-plot} shows the profiles for a sample of 30 individuals the mean hourly wage before and after extreme values are replaced.
 The plot shows fluctuations in wages remain, but the large spikes (in this sample, individuals 8296, 9962), which are considered implausible, are replaced.
-The \texttt{mean\_hourly\_wage} and \texttt{year} are set as the dependent and predictor, respectively.
-Furthermore, we use iteratively reweighted least squares with Huber weighting, where the observation with a slight residual gets a weight of 1, while the larger the residual, the smaller the weight (less than 1) (UCLA: Statistical Consulting Group 2021).
 
@@
 It is also important not to be too aggressive in outlier removal, because this would overly smooth the data, and ignore actual workforce experiences important wage volatility such as, abnormally high wages for a fleeting amount of time.
-To explore the risk of being overly vigorous in labeling observations as outliers, some testing of threshold value was done.
-Samples of individuals were examined under the different thresholds with an eye to the smoothness of the profiles.
+To explore the risk of being overly vigorous in labeling observations as outliers, a range of threshold value were assessed.
+Numerous samples of individuals were drawn for each threshold, and the smoothness of the profiles was examined.
 Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
-A threshold of 0.12 was chosen.
-That struck a balance between maintaining the natural variability of the wages with minimizing implausible values.
-Using this threshold, we impute the observations whose weights are less than 0.12 with the models' predicted value.
-We then flag those observations in a new variable called \texttt{is\_pred}, so that this change can be monitored in the downstream analyses.
+This was achieved using a purpose written \texttt{shiny} (Chang et al.
+2020) app (provided with the code of this paper).
+Observations where a change was made are tracked with a new variable called \texttt{is\_pred}, so that this effect in the downstream analyses can be monitored.
 
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/cookd/numbats/yowie-paper/paper/figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{figures/flowchart_blind}
 
@@
 The refreshed data is compared with the original set using a variety of numerical summaries and graphics.
-The current subset is made available in a new R package, called \texttt{[CENSORED]}.
-
-The data cleaning process is documented, and the code has been made available. These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
+The current subset is made available in a new R package, called \texttt{yowie}.
+
+The data cleaning process is documented, and the code has been made available.
+These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
 Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a \texttt{shiny} (Chang et al.
-2020) app, and the choices used in the refreshed data are documented.
+2020) app, which is provided with the code, and the choices used in the refreshed data are documented.
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-rlm}{}}%
-UCLA: Statistical Consulting Group. 2021. {``{Robust Regression \textbar{} R Data Analysis Examples}.''} February 2021. \url{https://stats.idre.ucla.edu/r/dae/robust-regression/}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%

### 2022-05-10 paper/paper-jsdse/paper-jsdse.tex
DC done with all her changes

@@
 
-This paper has illustrated the steps and decisions made to take a particular open data set and make it a textbook data set, ready for the classroom or research.
+Any longitudinal dataset used for education should have a sufficiently reproducible process to be updated with new data.
+The NLSY79 dataset is a great teaching tool that has become outdated, both because the dataset stops in 1994, and because the demographic data could be handled more delicately.
+This paper has illustrated the necessary steps and decisions made to take a particular open data set and make it a textbook data set, ready for the classroom or research.
 In the first stage, we showed the steps performed to get the data from the NLSY79 database.
@@
 \item
-  The refreshed data can be used for teaching advanced linear modeling, using almost identical code from Singer and Willett (2003), and thus make those lessons more current in time.
+  The refreshed data can be used for teaching advanced linear modeling, using almost identical code from Singer and Willett (2003), and thus make those lessons more current for today's students.
 \item

### 2022-05-10 paper/paper.Rmd
DC done with all her changes

@@
 - Some of the methods shown, and lessons learned about the data, can also be considered generally for data cleaning, initial data analysis and exploratory data analysis, useful for working with other data sets.
-- The refreshed data can be used for teaching advanced linear modeling, using almost identical code from @SingerJudithD2003Alda, and thus make those lessons more current in time.
+- The refreshed data can be used for teaching advanced linear modeling, using almost identical code from @SingerJudithD2003Alda, and thus make those lessons more current for today's students.
 - It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience.

### 2022-05-10 response/response2.Rmd
update of intro according to AE comment
DC done with all her changes

@@
 
-XXX DC to do
+We have revised the introduction to fix this. XXX DONE
 
@@
 
-We think it is best to keep this emphasis, because the hard work is on cleaning and tidying the data, in a reproducible manner, which is close to the mission of the special issue.
-We have added some additional sentences about lessons for statistics and data science educators.
-XXX DC to do (might need help from Emi or Nick)
+We have added additional bullet points on the benefit for educators in Section 5, in response to the overview comments by the Editor and AE (top of this response).
+We think it is best to keep content on cleaning and tidying the data, in a reproducible manner, to match the mission of the special issue.
+XXX DONE
 
@@
 
-Thanks for this, exactly right! We have changed the wording accordingly. XXX DC
+Thanks for this, exactly right! We have changed the wording accordingly, in various places in the paper. XXX DC, ET, NT please check that this message comes across clearly enough.
 
@@
 
-XXX DC to address
+See response above to the associate editor, about this concern. XXX DONE
 

### 2022-05-10 paper/paper.Rmd
Edited Dewi's part, 3 comments remains.

@@
 The NLSY79 database has been continuously updated through to 2018, so new records are available.
-Here we describe our journey to re-create the wages data, and document the process so that the data can be regularly updated into the future.
+Here we describe our journey to refresh the wages data, and document the process so that the data can be regularly updated into the future.
 Our journey was difficult because the steps and decisions taken to get from the raw data to the wages textbook subset have not been clearly articulated.
@@
 3. LNW: natural logarithm of wages, adjusted with 1990's inflation rate. 
-4. BLACK: binary variable, 1 indicates black and 0 otherwise. 
-5. HISPANIC: binary variable, 1 indicates hispanic and 0 otherwise. 
+4. BLACK: binary variable, 1 indicates Black and 0 otherwise. 
+5. HISPANIC: binary variable, 1 indicates Hispanic and 0 otherwise. 
 6. HGC: the highest grade completed.
@@
 We also include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked.
-We do not attempt to re-create the unemployment rate variable.
+We do not attempt to refresh the unemployment rate variable.
 
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (`gender`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).
+In our final target data, we wish to include the demographic variables with variable names specified in brackets: sex (`sex`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).
 
-For `gender` and `race`, we only rename the column names provided in the raw data, `new_data_qnames`.
-It is worth noting that `gender` comes from a variable called `sex` in the database.
-We use the term `gender` as the reports about the data on various pages in the website more commonly use `gender`.
-Gender, as provided in the data, is self-reported and only has two categories, "male" and "female." The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
+For `sex` and `race`, we only rename the column names provided in the raw data, `new_data_qnames`.
+However, it is worth noting that the term `sex` is sometimes interchangeably used with the term `gender`.
+Therefore, when using the dataset in the classroom, the educators should emphasize this discourse that using this variables should be accompanied by special attention.
 Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification [@kennedy2020using].
@@
 Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
-This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
-Similarly, this also relates to race, as reported in the database.
-When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
 
-The object `new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981`, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
+For the `race` variable, we acknowledge a current standard of how race should be reported in a dataset [@doh].
+According to this standard, there should be at least five categories for race reported in a data, and one can be identified into more than one race category.
+Also, Hispanic or Latino is identified as an ethnic group rather than a race.
+In the refreshed data, we could not attempt to report the race and ethnicity with this standard as we did not have any additional resources to do so, and hence we only used the category as in provided in the database.
+This fact could be a suggestion for the Bureau of Labor of Statistics as the data provider to update the race reporting to today's standard.
+Similar to the "sex variable," the educators should also emphasize that when analyzing this variable, one should keep in mind that the purpose is to study racism rather than race.
+
+Further, the object `new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981`, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 This variable is obtained from `new_data_qnames` with the name `HGC_EVER_XRND` and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as `hgc_i` and recoded as a factor to give `hgc` (e.g., 10th grade, 11th grade, and so on).
+We decide to include both
 
@@
 We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked.
-When this value was missing, 2012, 2014, 2016, and 2018, but available in the first form substituted accordingly.
+In the survey years when the revised value were not reported, i.e., in 2012, 2014, 2016, and 2018, we used the unrevised version accordingly.
 <!--We also check the consistency between `hgc_i` and `grade`.
@@
 Hence, we validate whether all respondents were in this range in the data we extracted.
-Additionally, the [NLSY](https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction) also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic).
+Additionally, the [NLSY](https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction) also provides the number of the survey cohort by their sex (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic).
 To validate this, we used the `demog_nlsy79`, i.e., the data with the survey years 1979 sample.
@@
   2. `age_1979`: The age of the individual in 1979.
-  3. `gender`: Gender of the individual, FEMALE and MALE.
-  4. `race` : Race of the individual, NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
+  3. `sex`: Sex of the individual, f = Female and m = Male.
+  4. `race` : Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
   5. `hgc`: The highest grade completed ever.
@@
   4. `age_1979`: The age of the individual in 1979.
-  5. `gender`: Gender of the individual: FEMALE and MALE.
-  6. `race`: Race where the individual belongs to: NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
-  7. `grade`: Integer value of the highest grade completed corresponding to `year`. 
+  5. `sex`: Sex of the individual, f = Female and m = Male.
+  6. `race`: Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
+  7. `grade`: Integer value of the highest grade completed corresponding to the survey year. 
   8. `hgc`: The highest grade completed ever.
@@
 
-Any longitudinal dataset used for education should have a sufficiently reproducible process to be updated with new data.
+Any longitudinal dataset used for education should have a sufficiently reproducible process to be refreshed with new data.
 The NLSY79 dataset is a great teaching tool that has become outdated, both because the dataset stops in 1994, and because the demographic data could be handled more delicately.
@@
 It means education makes a difference in the wage experience and that public policy that encourages education is a data-supported action.
-We would also learn, although not presented here, that race matters and being black leads to lower wages.
+We would also learn, although not presented here, that race matters and being Black leads to lower wages.
 This is a disturbing finding because there is no rationale for such a difference in a fair society.
@@
 Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race.
-If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., gender and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
+If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., sex and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
 This paper demonstrates the process of producing a trustworthy data set for teaching.

### 2022-05-10 paper/paper.tex
Edited Dewi's part, 3 comments remains.

--- 2022-05-10 paper/paper.tex
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
-
-This disparity between the average and the individual is a part of statistics as a discipline that requires more attention.
-This particular data set is a prime example of discussing this disparity.
-Textbook data sets have longevity if there is an unresolved mystery.
-The iris data (Andereson 1935) is a prime example.
-It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis (a side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)).
-We argue the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
-
-For statistics and data science by association, it is increasingly important to reach the individual.
-One might describe this as a divergence of purpose -- statistics for public policy -- or statistics for the public.
+It is for these reasons that the wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
+
+This disparity between the overall trend and the individual heterogeneity is what makes this data interesting.
+Textbook data sets have longevity if they have unresolved elements.
+The iris data (Anderson 1935) is a prime example.
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
+(It should be noted that the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)).
+We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
+
+For statistics, and data science by association, it is increasingly important to reach the individual.
+One might think of this as a divergence of purpose -- statistics for public policy -- or statistics for the public.
 The two are not the same.
 As the world becomes more electronically connected, combating misinformation and mitigating conspiracy theories require statistics that address the individual.
-For example, with the wages data, even though the message for public policy is that demographic profile is related to different wage patterns on average, the message for the individual is you are more than your demographic.
-The majority of people in the study do not have a pattern that is similar to the average.
-If you have a bad experience, that your wages have declined over time, you are not alone; there are others like you, and more than you think.
-A similar tone is echoed occasionally in the public media.
-For example, an article published in the Sydney Morning Herald argues there is no average Australian (Moncrief 2015).
+For example, with the wages data, the overall trend, across demographics, is a steady increase in wages, but the patterns among individuals is heterogeneous.
+Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year, and the heterogeneity of the patterns are evident across demographics.
+Subject-specific variation explains most of the variation in wages, and even though the overall trend is statistically significant, it is weak.
+The message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable.
+However, discussing the overall trend with an individual is misleading -- more than likely their pattern is completely different.
+It could even be depressing for them to compare themselves to the overall average, especially if the individual's wages have a declining trend or high volatility.
+It would be more useful for an individual to know what percentage of the population have profiles like theirs.
+We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation.
+We occasionally hear this voiced in the public media media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian (Moncrief 2015).
+Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.
 
@@
 
-\caption{\label{tab:gender-race-table}Contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
+\caption{\label{tab:gender-race-table}Contingency table for sex and race for the full NLSY79 data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
 \centering
@@
 \cmidrule(l{3pt}r{3pt}){2-4}
-Gender & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
+Sex & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
 \midrule
@@
 
-Why is this exercise important for teachers and students?
-This work illustrates the steps in cleaning and processing data in preparation for it to be a textbook data set.
-Choices made during the cleaning and processing can affect findings made with the data, and these should be transparent.
-Along with the textbook data, which is now provided as an R package, the documentation of the process provides some examples for teaching data cleaning, initial data analysis, and exploratory data analysis.
-The refreshed textbook data provides a resource for teaching longitudinal data analysis.
+Why is the work presented here useful for teachers and students? There are several reasons:
+
+\begin{itemize}
+\tightlist
+\item
+  It primarily illustrates the steps in cleaning and processing a well-known longitudinal data so that it can be further refreshed as a textbook data set as new data emerges.
+Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
+\item
+  Some of the methods shown, and lessons learned about the data, can also be considered generally for data cleaning, initial data analysis and exploratory data analysis, useful for working with other data sets.
+\item
+  The refreshed data can be used for teaching advanced linear modeling, using almost identical code from Singer and Willett (2003), and thus make those lessons more current for today's students.
+\item
+  It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience.
+The vignettes of the \texttt{brolgar} (Tierney, Cook, and Prvan 2020) package provide useful guides for the case study content.
+\end{itemize}
 
@@
 \leavevmode\hypertarget{ref-iris-data}{}%
-Andereson, Edgar. 1935. ``The Irises of the Gaspe Peninsula.'' \emph{Bulletin of the American Iris Society} 59: 2--5.
+Anderson, Edgar. 1935. ``The Irises of the Gaspe Peninsula.'' \emph{Bulletin of the American Iris Society} 59: 2--5.
 

### 2022-05-10 response/response2.Rmd
Edited Dewi's part, 3 comments remains.

@@
 
-It is not clear how to handle this.
+* It is not clear how to handle this.
 We interpret the comment from the Editor, that the paper is consistent with the call for the special issue.
@@
 
-We interpret this to mean we should change the variable name in the cleaned data, and so have done this, and then "sex" is uniformly used throughout the paper. XXX Dewi to do
+* We interpret this to mean we should change the variable name in the cleaned data to be `sex`.
+We have done this and then "sex" is uniformly used throughout the paper.
+Even though we have changed the `gender` variable to be `sex`, we keep the explanation of the importance for the educators to emphasize the discussion between the usage of `sex` and `gender` because in many other cases and datasets, the usage between `sex` and `gender` are often interchangeable.
+XXX DONE
 
 \begin{description}
-\item[Associate Editor:] \textcolor{violet}{\textbf{The discussion around highest grade completed is mechanical in what was done, but lacks substantive reasoning for those decisions and possible downstream con- sequences.} It is still unclear to me whether education is measured by year or by category (it’s described both ways) and, in either case, the decision for one or the other isn’t contrasted or justified.
+\item[Associate Editor:] \textcolor{violet}{\textbf{The discussion around highest grade completed is mechanical in what was done, but lacks substantive reasoning for those decisions and possible downstream consequences.} It is still unclear to me whether education is measured by year or by category (it’s described both ways) and, in either case, the decision for one or the other isn’t contrasted or justified.
 For example `hcg` is described as a factor and `grade` is described as only increasing (numeric?) and surely the authors mean non-decreasing.
@@
 
-We have added more explanation of the two variables, and suggestions of where to use one vd the other in classroom analyses. XXX Dewi to do
+* We have added more explanation of the two variables, and suggestions of where to use one and the other in classroom analyses. XXX Dewi to do
 
@@
 
-Using `rlm` with the `nest` and `map` functions from `tidyr`, efficiently and simply fit these models.
+* Using `rlm` with the `nest` and `map` functions from `tidyr`, efficiently and simply fit these models.
 They can be easily explained in a classroom setting, and equip students generally with advanced data handling skills.
@@
 
-This is more of a comment than a request to make a change, but we agree that this is an area that could be a good classroom discussion.
-We have revised the section on handling extremes to make our explanation simpler, with links to the shiny app that we put together to assist in the judgement and choice of threshold.
+* This is more of a comment than a request to make a change, but we agree that this is an area that could be a good classroom discussion.
+We have revised the section on handling extremes to make our explanation simpler, with links to the shiny app that we put together to assist in the judgment and choice of threshold.
 We hope that this provides fodder for discussion.
@@
 
-It would be nice if the database contained a broader description of race, but it only has a single variable for race.
-For this reason, our cleaned data reflects this, and has a single variable for race.
-XXX Dewi to do
+* It would be nice if the database contained a broader description of race, but it only has a single variable for the race.
+For this reason, our cleaned data reflects this and has a single variable for the race.
+We have explained why we used these categories and suggested the data provider adjust the race variable to use the current standard.
+XXX DONE
 
@@
 
-We have revised the introduction to fix this. XXX DONE
+* We have revised the introduction to fix this. XXX DONE
 
@@
 
-We have added additional bullet points on the benefit for educators in Section 5, in response to the overview comments by the Editor and AE (top of this response).
+* We have added additional bullet points on the benefit for educators in Section 5, in response to the overview comments by the Editor and AE (top of this response).
 We think it is best to keep content on cleaning and tidying the data, in a reproducible manner, to match the mission of the special issue.
@@
 
-We have changed all to "refresh". XXX Dewi
+* We have changed all to "refresh" as it is not correct if we say it as a replication because we did not use the exactly same variable and same result (esepcially ID included in the dropouts cohort.) XXX DONE
 
@@
 
-Thanks for this, exactly right! We have changed the wording accordingly, in various places in the paper. XXX DC, ET, NT please check that this message comes across clearly enough.
+* Thanks for this, exactly right! We have changed the wording accordingly, in various places in the paper. XXX DC, ET, NT please check that this message comes across clearly enough.
 
@@
 
-Same point as made by the AE, and how we have addressed it is described above.
+* Same point as made by the AE, and how we have addressed it is described above.
 
@@
 
-See response above to the associate editor, about this concern. XXX DONE
+* See response above to the associate editor, about this concern. XXX DONE
 
@@
 
-Done XXX Dewi
+* Done XXX Dewi
 
@@
 
-XXX Dewi Add some words to make it clear what data this is summarising
+* XXX Dewi Add some words to make it clear what data this is summarising
 
@@
 
-We have expanded this explanation XXX Dewi
+* This sentence means that when the revised `hgc` variable were not reported, we only use the unrevised version of the `hgc` in that year.
+We thank the reviewer for this notice.
+We agree that this sentences is not clear, so we have modified it.
+XXX DONE
 
@@
 \begin{itemize}
-  \item p1 in abstract: Both “wages textbook subset, have not” and “open source R package, called” do not need commas.
-  \item p2 “high school dropouts, from 1979-1994” does not need comma
-  \item p2 comma after “divergence of purpose” might be better replaced with dash.
-  \item p4 the sentence “Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to 1200/hour.” needs a few edits.
-I suggest “Plot (C) shows the profile for an individual with a maximum wage that is not so extreme but still indicates a problem: their wages are consistently low except for one year where they earned close to 1200/hour.”
-  \item p24 “predominately” does not need a hyphen.
+  \item \textcolor{violet}{p1 in abstract: Both “wages textbook subset, have not” and “open source R package, called” do not need commas.}
+  \item \textcolor{violet}{p2 “high school dropouts, from 1979-1994” does not need comma}
+  \item \textcolor{violet}{p2 comma after “divergence of purpose” might be better replaced with dash.}
+  \item \textcolor{violet}{p4 the sentence “Plot (C) shows the profile for an individual, with not such a high maximum wage but still indicates a problem: their wages are consistently low except for one year where they earned close to 1200/hour.” needs a few edits.
+I suggest “Plot (C) shows the profile for an individual with a maximum wage that is not so extreme but still indicates a problem: their wages are consistently low except for one year where they earned close to 1200/hour.”}
+  \item \textcolor{violet}{p24 “predominately” does not need a hyphen.}
 \end{itemize}
@@
 \begin{itemize}
-\item p2 comma  “Applied longitudinal data analysis”,
-\item p4 comma “tame data”,
-\item p6 comma “statistical value chain”,
-\item p9 period “female”.
-\item p13 period “number of weeks worked since the last interview”.
+\item \textcolor{violet}{p2 comma  “Applied longitudinal data analysis”,}
+\item \textcolor{violet}{p4 comma “tame data”,}
+\item \textcolor{violet}{p6 comma “statistical value chain”,}
+\item \textcolor{violet}{p9 period “female”.}
+\item \textcolor{violet}{p13 period “number of weeks worked since the last interview”.}
 \end{itemize}
@@
 \begin{itemize}
-\item p6 “For example, use a single categorical race variable instead of the two binary race variables.” Perhaps missing a “we” before use?
-\item p6 The sentence beginning “van der Loo and de Jonge” is jarring because of the lowercase name, particularly because it starts a paragraph and section.
-I recommend flipping the clauses to begin “In the context of official statistics, van der Loo and de Jonge…”
-\item p18 “The year when the individual starting to work.” should perhaps be “The year when the individual started to work.”
-\item p24 The sentence “On an individual level, one needs to know where I am in this data and does this data relate to me.” needs edits.
+\item \textcolor{violet}{p6 “For example, use a single categorical race variable instead of the two binary race variables.” Perhaps missing a “we” before use?}
+\item \textcolor{violet}{p6 The sentence beginning “van der Loo and de Jonge” is jarring because of the lowercase name, particularly because it starts a paragraph and section.
+I recommend flipping the clauses to begin “In the context of official statistics, van der Loo and de Jonge…”}
+\item \textcolor{violet}{p18 “The year when the individual starting to work.” should perhaps be “The year when the individual started to work.”}
+\item \textcolor{violet}{p24 The sentence “On an individual level, one needs to know where I am in this data and does this data relate to me.” needs edits.
 I’m not sure people from the longitudinal study are likely to be looking at this data, so “where I am in this data” is not quite accurate.
-Rather, people might want to see how their characteristics relate to those in the dataset.
-\item p26 Cooksey reference should have NLSY capitalized. 
+Rather, people might want to see how their characteristics relate to those in the dataset.}
+\item \textcolor{violet}{p26 Cooksey reference should have NLSY capitalized.} 
 \end{itemize}
@@
 \begin{itemize}
-\item P2, l46: missing right parenthesis.
-\item P9, l24: missing word(s) "this sometimes difficult the adjustment"
-\item P11, l43: "hours" worked
-\item P12, l32: unit in "weeks"
-\item P14, l38: Figure 3 shows...
-\item P21, l21: "the weeks worked since" ...
-\item P24, l21: The sentence "On an individual level, one needs to know where I am in this data and does this data relate to me" is awkward. Please revise.
-\item P34 l 38: Figure 5 (C) is mislabeled
+\item \textcolor{violet}{P2, l46: missing right parenthesis.}
+\item \textcolor{violet}{P9, l24: missing word(s) "this sometimes difficult the adjustment"}
+\item \textcolor{violet}{P11, l43: "hours" worked}
+\item \textcolor{violet}{P12, l32: unit in "weeks"}
+\item \textcolor{violet}{P14, l38: Figure 3 shows...}
+\item \textcolor{violet}{P21, l21: "the weeks worked since" ...}
+\item \textcolor{violet}{P24, l21: The sentence "On an individual level, one needs to know where I am in this data and does this data relate to me" is awkward. Please revise.}
+\item  \textcolor{violet}{P34 l 38: Figure 5 (C) is mislabeled}
 \end{itemize}
@@
 
-We have put the capitalisation back in, and changed "WHITE" to be "NBH". XXX Dewi
+* We have put the capitalisation back in. Besides, we did not use the term "White" in the dataset as we used the same term reported in the database, which is "Non-Black, Non-Hispanic". XXX Done
 
@@
 \begin{itemize}
-\item 1. The wages data of the whole NLSY79 cohort, including females.
-\item 2. A separate table of the demographic data of the whole NLSY79 cohort.
-\item 3. The high school dropouts’ wages data is closest to a refreshed version of Singer and
-Willett (2003)’s data.
+\item \textcolor{violet}{1. The wages data of the whole NLSY79 cohort, including females.}
+\item \textcolor{violet}{2. A separate table of the demographic data of the whole NLSY79 cohort.}
+\item \textcolor{violet}{3. The high school dropouts’ wages data is closest to a refreshed version of Singer and Willett (2003)’s data.}
 \end{itemize}
-1 and 2 are nouns and 3 is a sentence – I’m a little confused what this third dataset represents.
-I think you mean that you create a subset of (1) in the scope of the original data, to see how closely it replicates?
+\textcolor{violet}{1 and 2 are nouns and 3 is a sentence – I’m a little confused what this third dataset represents.
+I think you mean that you create a subset of (1) in the scope of the original data, to see how closely it replicates?}
 \end{description}

### 2022-05-10 data-raw/data-preprocessing.R

@@ -57,7 +57,12 @@ has_dob_conflict <- any(dob_tidy$dob_conflict, na.rm = TRUE)
 demog_tidy <- categories_qnames %>%
   select(id = CASEID_1979,
          race = SAMPLE_RACE_78SCRN,
-         gender = SAMPLE_SEX_1979)
+         sex = SAMPLE_SEX_1979) %>%
+  mutate(race = as.factor(case_when(race == "NON-BLACK, NON-HISPANIC" ~ "NBH",
+                                    race == "HISPANIC" ~ "H",
+                                    race == "BLACK" ~ "B")),
+         sex = as.factor(case_when(sex == "FEMALE" ~ "f",
+                                   sex == "MALE" ~ "m")))
 
 ## ---- demog-ed
 # tidy highest grade completed
@@ -189,7 +194,7 @@ demog_nlsy79 <- full_demographics %>%
   mutate(age_1979 = 1979 - (dob_year + 1900)) %>%
   dplyr::select(id,
                 age_1979,
-                gender,
+                sex,
                 race,
                 hgc,
                 hgc_i,

### 2022-05-10 data-raw/ida.R

@@ -30,18 +30,18 @@ freq_formatting <- function(x) {
     mutate(across(where(is.numeric), ~scales::comma(.x, 1)))
 }
 
-gender_race_table <- full_demographics %>%
-  tabyl(gender, race) %>%
+sex_race_table <- full_demographics %>%
+  tabyl(sex, race) %>%
   adorn_totals(c("row", "col")) %>%
   adorn_percentages("row") %>%
   adorn_pct_formatting(digits = 0) %>%
   adorn_ns(., position = "front", ns = freq_formatting(.)) %>%
-  mutate(gender = str_to_title(gender),
+  mutate(sex = str_to_title(sex),
          Total = str_replace(Total, "\\([1|2]00\\%\\)", ""))
 
-kable(gender_race_table,
-      caption = "Contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1\\%) is out of the total corresponding to row.",
-      col.names = c("Gender", "Hispanic", "Black", "Non-Black, Non-Hispanic", "Total"),
+kable(sex_race_table,
+      caption = "Contingency table for sex and race for the full NLSY79 data. The percentage (rounded to closest 1\\%) is out of the total corresponding to row.",
+      col.names = c("Sex", "Hispanic", "Black", "Non-Black, Non-Hispanic", "Total"),
       booktabs = TRUE,
       linesep = "",
       align = "lrrrr") %>%
@@ -282,7 +282,7 @@ wages_clean <- wages_cleaned %>%
 
 # rename and select the wages in tidy
 wages <- wages_clean %>%
-  select(id, year, mean_hourly_wage, age_1979, gender, race, grade, hgc, hgc_i, hgc_1979, ged = dip_or_ged,
+  select(id, year, mean_hourly_wage, age_1979, sex, race, grade, hgc, hgc_i, hgc_1979, ged = dip_or_ged,
          number_of_jobs, total_hours, stwork, yr_wforce, exp, is_wm, is_pred) %>%
   mutate(id = as.factor(id),
          hgc = as.factor(hgc),
@@ -303,7 +303,7 @@ wages_hs_do <- wages %>%
   # based on two criteria mentioned in the paper
   filter(hgc_i < 12 | (hgc_i >= 12 & ged != 1)) %>%
   filter(age_1979 <= 17,
-         gender == "MALE") %>%
+         sex == "MALE") %>%
   as_tsibble(key = id,
              index = year,
              regular = FALSE)

### 2022-05-11 paper/paper.Rmd
working on gender/sex explanation

@@
 
-According to @WickhamHadley2014TD, tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
-The raw data, `new_data_qnames`, does not comply with these rules as it is organized such that each row corresponds to an individual.
-As respondents can have multiple jobs at specific years, the column names, such as `HRP1_1979`, `HRP2_1979`, `HRP1_1980` and `HRP2_1980`, contain the information about the job number up to 5 (`HRP1` = job 1, `HRP2` = job 2) and the year.
+To get the data into tidy form [@WickhamHadley2014TD], it needs to comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
+The raw data, `new_data_qnames`, violates these rules (i) and (ii) because information about an individual's multiple jobs over different years are in multiple columns.
 The raw data consequently has a large number of columns (`r ncol(new_data_qnames)` to be specific).
 The values in the cell under the variables begin with `HRP` correspond to the hourly wage in dollars.
-A glimpse of this data shows:
+A glimpse of this data show the format problems:
 
@@
 
-Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
+Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
 This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long-form, with `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: sex (`sex`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).
+Our final target data will include the demographic variables (variable names): sex (`sex`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).
 
-For `sex` and `race`, we only rename the column names provided in the raw data, `new_data_qnames`.
-However, it is worth noting that the term `sex` is sometimes interchangeably used with the term `gender`.
-Therefore, when using the dataset in the classroom, the educators should emphasize this discourse that using this variables should be accompanied by special attention.
+For `sex` and `race`, we have simplified the original names provided in the raw data, `SAMPLE_SEX_1979` and `SAMPLE_RACE_78SCRN`, respectively.
+However, it is important to note that the use of "sex" is probably not correct.
+This information is provided by the individual, and although only has two categories, is more consistent with "gender" as defined in  Therefore, when using the dataset in the classroom, the educators should emphasize this discourse that using this variables should be accompanied by special attention.
 Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification [@kennedy2020using].

### 2022-05-11 paper/paper-jsdse/paper-jsdse.tex
revised sex and race paragraphs

@@
 \item
-  BLACK: binary variable, 1 indicates black and 0 otherwise.
-\item
-  HISPANIC: binary variable, 1 indicates hispanic and 0 otherwise.
+  BLACK: binary variable, 1 indicates Black and 0 otherwise.
+\item
+  HISPANIC: binary variable, 1 indicates Hispanic and 0 otherwise.
 \item
@@
 We also modify some variables.
-For example, we use a single categorical race variable instead of the two binary race variables.
+For example, we use a single categorical race variable instead of the two binary race variables, for reasons detailed below.
 We also include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked.
-We do not attempt to re-create the unemployment rate variable.
+We do not attempt to refresh the unemployment rate variable.
 
@@
 
-According to Wickham (2014), tidy data sets comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
-The raw data, \texttt{new\_data\_qnames}, does not comply with these rules as it is organized such that each row corresponds to an individual.
-As respondents can have multiple jobs at specific years, the column names, such as \texttt{HRP1\_1979}, \texttt{HRP2\_1979}, \texttt{HRP1\_1980} and \texttt{HRP2\_1980}, contain the information about the job number up to 5 (\texttt{HRP1} = job 1, \texttt{HRP2} = job 2) and the year.
+To get the data into tidy form (Wickham 2014), it needs to comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
+The raw data, \texttt{new\_data\_qnames}, violates these rules (i) and (ii) because information about an individual's multiple jobs over different years are in multiple columns.
 The raw data consequently has a large number of columns (742 to be specific).
 The values in the cell under the variables begin with \texttt{HRP} correspond to the hourly wage in dollars.
-A glimpse of this data shows:
+A glimpse of this data show the format problems:
 
@@
 
-Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
+Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
 This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
-
-For \texttt{gender} and \texttt{race}, we only rename the column names provided in the raw data, \texttt{new\_data\_qnames}.
-It is worth noting that \texttt{gender} comes from a variable called \texttt{sex} in the database.
-We use the term \texttt{gender} as the reports about the data on various pages in the website more commonly use \texttt{gender}.
-Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female.'' The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
-Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification (Kennedy et al.
-2020).
-From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex (Kennedy et al.
-2020).
-Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
-This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
-Similarly, this also relates to race, as reported in the database.
-When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
-
-The object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
+Our final target data will include the demographic variables (variable names): sex (\texttt{sex}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
+
+For \texttt{sex} and \texttt{race}, we have simplified the original names provided in the raw data, \texttt{SAMPLE\_SEX\_1979} and \texttt{SAMPLE\_RACE\_78SCRN}, respectively.
+However, it is important to note that the use of ``sex'' is probably not correct.
+This information is provided by the individual, and although only has two categories, is more consistent with ``gender'' as defined in Heidari et al.
+(2016).
+When using the dataset in the classroom, the educators might include discourse on the use of terms ``sex'' and ``gender''.
+Particularly, measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or create measurement error for people whose gender does not align with their sex classification.
+From a statistical perspective, this can make it difficult to adjust survey statistics to the population when gender is measured in different categorization in different data sources.
+Kennedy et al.
+(2020) provides a helpful discussion on these topics.
+
+Similarly, there are issues for the \texttt{race} variable.
+The current US Federal guidelines (Office of Management and Budget 1997) state that there should be at least five categories for race, and an individual should be able to identify as more than one race category.
+in addition, ``Hispanic'' or ``Latino'' should be considered to be an ethnic group rather than a race.
+This level of detail is not available in the database.
+There is one variable with only three categories, and an individual can be a member of only one.
+Thus, the \texttt{race} variable is inadequate by today's standards, and educators should point to the current Federal guidelines when using this variable, and keep in mind that the purpose of any analysis is to study racism rather than race.
+
+Further, the object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on).
+We decide to include both
 
@@
 We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked.
-When this value was missing, 2012, 2014, 2016, and 2018, but available in the first form substituted accordingly.
+In the survey years when the revised value were not reported, i.e., in 2012, 2014, 2016, and 2018, we used the unrevised version accordingly.
 
@@
 Hence, we validate whether all respondents were in this range in the data we extracted.
-Additionally, the \href{https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction}{NLSY} also provides the number of the survey cohort by their gender (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic).
+Additionally, the \href{https://www.nlsinfo.org/content/cohorts/nlsy79/intro-to-the-sample/nlsy79-sample-introduction}{NLSY} also provides the number of the survey cohort by their sex (6,403 males and 6,283 females) and race (7,510 Non-Black/Non-Hispanic; 3,174 Black; 2,002 Hispanic).
 To validate this, we used the \texttt{demog\_nlsy79}, i.e., the data with the survey years 1979 sample.
@@
 \cmidrule(l{3pt}r{3pt}){2-4}
-Gender & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
+Sex & Hispanic & Black & Non-Black, Non-Hispanic & Total\\
 \midrule
-Male & 1,000 (16\%) & 1,613 (25\%) & 3,790 (59\%) & 6,403\\
-Female & 1,002 (16\%) & 1,561 (25\%) & 3,720 (59\%) & 6,283\\
+F & 1,561 (25\%) & 1,002 (16\%) & 3,720 (59\%) & 6,283\\
+M & 1,613 (25\%) & 1,000 (16\%) & 3,790 (59\%) & 6,403\\
 \midrule
-Total & 2,002 (16\%) & 3,174 (25\%) & 7,510 (59\%) & 12,686\\
+Total & 3,174 (25\%) & 2,002 (16\%) & 7,510 (59\%) & 12,686\\
 \bottomrule
@@
 \item
-  \texttt{gender}: Gender of the individual, FEMALE and MALE.
-\item
-  \texttt{race} : Race of the individual, NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
+  \texttt{sex}: Sex of the individual, f = Female and m = Male.
+\item
+  \texttt{race} : Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
 \item
@@
 \item
-  \texttt{gender}: Gender of the individual: FEMALE and MALE.
-\item
-  \texttt{race}: Race where the individual belongs to: NON-BLACK,NON-HISPANIC; HISPANIC; BLACK.
-\item
-  \texttt{grade}: Integer value of the highest grade completed corresponding to \texttt{year}.
+  \texttt{sex}: Sex of the individual, f = Female and m = Male.
+\item
+  \texttt{race}: Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
+\item
+  \texttt{grade}: Integer value of the highest grade completed corresponding to the survey year.
 \item
@@
 It means education makes a difference in the wage experience and that public policy that encourages education is a data-supported action.
-We would also learn, although not presented here, that race matters and being black leads to lower wages.
+We would also learn, although not presented here, that wages on average differ by race category.
 This is a disturbing finding because there is no rationale for such a difference in a fair society.
@@
 {``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14.
-\url{https://doi.org/10.1080/00031305.2015.1081106}.
+https://doi.org/\url{https://doi.org/10.1080/00031305.2015.1081106}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-SAGER}{}}%
+Heidari, S., T.
+F.
+Babor, P.
+De Castro, S.
+Tort, and M.
+Curno.
+2016.
+{``Sex and Gender Equity in Research: Rationale for the {SAGER} Guidelines and Recommended Use.''} \emph{Research Integrity and Peer Review} 1 (2).
+https://doi.org/\url{https://doi.org/10.1186/s41073-016-0007-6}.
 
@@
 {``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1).
-\url{https://doi.org/10.5070/T511103589}.
+https://doi.org/\url{https://doi.org/10.5070/T511103589}.
 
@@
 Moncrief, Marc. 2015. {``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.''} \url{https://bit.ly/smh-not-normal}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-doh}{}}%
+Office of Management and Budget.
+1997.
+{``Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity.''} \url{https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf}.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
-Wickham, Hadley. 2011. {``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
+Wickham, Hadley.
+2011.
+{``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29.
+https://doi.org/\url{https://doi.org/10.18637/jss.v040.i01}.
 

### 2022-05-11 paper/paper.Rmd
revised sex and race paragraphs

@@
 We also modify some variables.
-For example, we use a single categorical race variable instead of the two binary race variables.
+For example, we use a single categorical race variable instead of the two binary race variables, for reasons detailed below.
 We also include additional variables, some for the purpose of providing more options for data exploration in teaching examples: year of the survey, age of individual in 1979, whether the individual completed high school with a diploma or with a graduate equivalency degree (GED), the highest grade completed in the corresponding year of survey, the number of jobs the individual had in the corresponding year of survey, the total number of hours the individual usually works per week, the year when the individual started to work, and the number of years the individual worked.
@@
 However, it is important to note that the use of "sex" is probably not correct.
-This information is provided by the individual, and although only has two categories, is more consistent with "gender" as defined in  Therefore, when using the dataset in the classroom, the educators should emphasize this discourse that using this variables should be accompanied by special attention.
-Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification [@kennedy2020using].
-From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex [@kennedy2020using].
-Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
+This information is provided by the individual, and although only has two categories, is more consistent with "gender" as defined in @SAGER.
+When using the dataset in the classroom, the educators might include discourse on the use of terms "sex" and "gender".
+Particularly, measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or create measurement error for people whose gender does not align with their sex classification.
+From a statistical perspective, this can make it difficult to adjust survey statistics to the population when gender is measured in different categorization in different data sources.
+@kennedy2020using provides a helpful discussion on these topics.
+<!--Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
+-->
 
-For the `race` variable, we acknowledge a current standard of how race should be reported in a dataset [@doh].
-According to this standard, there should be at least five categories for race reported in a data, and one can be identified into more than one race category.
-Also, Hispanic or Latino is identified as an ethnic group rather than a race.
-In the refreshed data, we could not attempt to report the race and ethnicity with this standard as we did not have any additional resources to do so, and hence we only used the category as in provided in the database.
-This fact could be a suggestion for the Bureau of Labor of Statistics as the data provider to update the race reporting to today's standard.
-Similar to the "sex variable," the educators should also emphasize that when analyzing this variable, one should keep in mind that the purpose is to study racism rather than race.
+Similarly, there are issues for the `race` variable.
+The current US Federal guidelines [@doh] state that there should be at least five categories for race, and an individual should be able to identify as more than one race category.
+in addition, "Hispanic" or "Latino" should be considered to be  an ethnic group rather than a race.
+This level of detail is not available in the database.
+There is one variable with only three categories, and an individual can be a member of only one.
+Thus, the `race` variable is inadequate by today's standards, and educators should point to the current Federal guidelines when using this variable, and keep in mind that the purpose of any analysis is to study racism rather than race.
 
@@
 It means education makes a difference in the wage experience and that public policy that encourages education is a data-supported action.
-We would also learn, although not presented here, that race matters and being Black leads to lower wages.
+We would also learn, although not presented here, that wages on average differ by race category.
 This is a disturbing finding because there is no rationale for such a difference in a fair society.

### 2022-05-11 response/response2.Rmd
revised sex and race paragraphs

@@
 
-* We interpret this to mean we should change the variable name in the cleaned data to be `sex`.
-We have done this and then "sex" is uniformly used throughout the paper.
-Even though we have changed the `gender` variable to be `sex`, we keep the explanation of the importance for the educators to emphasize the discussion between the usage of `sex` and `gender` because in many other cases and datasets, the usage between `sex` and `gender` are often interchangeable.
-XXX DONE
+* We have revised the language, made a note for educators on usage, and referred to the SAGER guidelines. XXX DONE
 
@@
 
-* It would be nice if the database contained a broader description of race, but it only has a single variable for the race.
-For this reason, our cleaned data reflects this and has a single variable for the race.
-We have explained why we used these categories and suggested the data provider adjust the race variable to use the current standard.
-XXX DONE
+* We have revised the language, referenced the Federal standard, and explained the reasoning for using a single variable.  XXX DONE
 

### 2022-05-11 paper/paper-jsdse/paper-jsdse.tex
Finished address all of the comments.

@@
 
-In the context of official statistics, M.
-van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+In the context of official statistics, van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
 What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
@@
 Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
-This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
-2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham et al.
+This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
+2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham, François, et al.
 2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
@@
 The current US Federal guidelines (Office of Management and Budget 1997) state that there should be at least five categories for race, and an individual should be able to identify as more than one race category.
-in addition, ``Hispanic'' or ``Latino'' should be considered to be an ethnic group rather than a race.
+In addition, ``Hispanic'' or ``Latino'' should be considered to be an ethnic group rather than a race.
 This level of detail is not available in the database.
@@
 This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on).
-We decide to include both
+We decide to include both the factor and integer values to be used as examples in various classroom demonstrations, such as to explain different data types.
+Besides, it can also be used as the demonstration of different type of data visualization which is determined by the data type.
+It can also be used for teaching the modeling of cross sectional data.
 
@@
 This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals.
+Hence, different from \texttt{hgc}, this variable could be used to demonstrate longitudinal and time series analysis in the classroom.
 This is recorded in \texttt{new\_data\_qnames} as columns beginning with \texttt{Q3-4} and \texttt{HGC} with year as a suffix.
@@
 
-\caption{\label{tab:age-table}Frequency table of the age at the start of the survey in NSLY79 cohort}
+\caption{\label{tab:age-table}Frequency table of the age at the start of the survey in NSLY79 cohort in the extracted data.}
 \centering
@@
 
-\caption{\label{tab:gender-race-table}Contingency table for sex and race for the full NLSY79 data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
+\caption{\label{tab:gender-race-table}Contingency table for sex and race for the extracted NLSY79 demographic data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
 \centering
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{../figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart}
 
@@
 ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data.
-The color represents the stage of data cleaning in the statistical value chain (M.
-P.
-J.
-van der Loo and de Jonge 2021).
+The color represents the stage of data cleaning in the statistical value chain (van der Loo and de Jonge 2021).
 Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart}
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-iris-data}{}}%
-Anderson, Edgar. 1935. {``The Irises of the Gaspe Peninsula.''} \emph{Bulletin of the American Iris Society} 59: 2--5.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
+\leavevmode\hypertarget{ref-iris-data}{}%
+Anderson, Edgar. 1935. ``The Irises of the Gaspe Peninsula.'' \emph{Bulletin of the American Iris Society} 59: 2--5.
+
+\leavevmode\hypertarget{ref-nlsy79}{}%
 Bureau of Labor Statistics, U.S.
@@
 2021a.
-{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
+``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
 Columbus, OH, through \url{https://www.nlsinfo.org/bibliography-citing-nls-data}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlsy79guide}{}}%
----------. 2021b. {``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.''} \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-shiny}{}}%
-Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{{shiny: Web Application Framework for R}}. \url{https://CRAN.R-project.org/package=shiny}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Chatfield1985TIEo}{}}%
-Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
-
-\leavevmode\vadjust pre{\hypertarget{ref-eliznlsy}{}}%
+\leavevmode\hypertarget{ref-nlsy79guide}{}%
+---------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.'' \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
+
+\leavevmode\hypertarget{ref-shiny}{}%
+Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.
+
+\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
+Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
+
+\leavevmode\hypertarget{ref-eliznlsy}{}%
 Cooksey, Elizabeth C.
 2017.
-{``Using the National Longitudinal Surveys of Youth ({NLSY}) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M.
+``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
 Lerner Neal Halfon Christoper B.
@@
 Cham: Springer.
-https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-DasuTamraparni2003Edma}{}}%
+\url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.
+
+\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
 Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.
 
-\leavevmode\vadjust pre{\hypertarget{ref-janitor}{}}%
-Firke, Sam. 2020. \emph{{janitor: Simple Tools for Examining and Cleaning Dirty Data}}. \url{https://CRAN.R-project.org/package=janitor}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-racismnotrace}{}}%
-Fullilove, M. T. 1998. {``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.
-
-\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
+\leavevmode\hypertarget{ref-janitor}{}%
+Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.
+
+\leavevmode\hypertarget{ref-racismnotrace}{}%
+Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.
+
+\leavevmode\hypertarget{ref-grimshaw}{}%
 Grimshaw, Scott D.
 2015.
-{``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14.
-https://doi.org/\url{https://doi.org/10.1080/00031305.2015.1081106}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-SAGER}{}}%
+``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14.
+\url{https://doi.org/https://doi.org/10.1080/00031305.2015.1081106}.
+
+\leavevmode\hypertarget{ref-SAGER}{}%
 Heidari, S., T.
@@
 2016.
-{``Sex and Gender Equity in Research: Rationale for the {SAGER} Guidelines and Recommended Use.''} \emph{Research Integrity and Peer Review} 1 (2).
-https://doi.org/\url{https://doi.org/10.1186/s41073-016-0007-6}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-purrr}{}}%
-Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-penguins-data}{}}%
+``Sex and Gender Equity in Research: Rationale for the SAGER Guidelines and Recommended Use.'' \emph{Research Integrity and Peer Review} 1 (2).
+\url{https://doi.org/https://doi.org/10.1186/s41073-016-0007-6}.
+
+\leavevmode\hypertarget{ref-purrr}{}%
+Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.
+
+\leavevmode\hypertarget{ref-penguins-data}{}%
 Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
+\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
 Huebner, Marianne, Werner Vach, and Saskia le Cessie.
 2016.
-{``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ilk2004}{}}%
+``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
+
+\leavevmode\hypertarget{ref-ilk2004}{}%
 Ilk, Ozlem.
 2004.
-{``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.''} PhD thesis, Iowa State University.
+``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.'' PhD thesis, Iowa State University.
 \url{https://doi.org/10.31274/rtd-180813-11012}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-kennedy2020using}{}}%
-Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. {``Using Sex and Gender in Survey Adjustment.''} \url{https://arxiv.org/abs/2009.14401}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
+\leavevmode\hypertarget{ref-kennedy2020using}{}%
+Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. ``Using Sex and Gender in Survey Adjustment.'' \url{http://arxiv.org/abs/2009.14401}.
+
+\leavevmode\hypertarget{ref-tamedata}{}%
 Kim, A.
@@
 2018.
-{``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1).
-https://doi.org/\url{https://doi.org/10.5070/T511103589}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-KollerManuel2016rARP}{}}%
-Koller, Manuel. 2016. {``{robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models}.''} \emph{Journal of Statistical Software} 75 (6): 1--24.
-
-\leavevmode\vadjust pre{\hypertarget{ref-notaverage}{}}%
-Moncrief, Marc. 2015. {``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.''} \url{https://bit.ly/smh-not-normal}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-doh}{}}%
+``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1).
+\url{https://doi.org/https://doi.org/10.5070/T511103589}.
+
+\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
+Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.
+
+\leavevmode\hypertarget{ref-notaverage}{}%
+Moncrief, Marc. 2015. ``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.'' \url{https://bit.ly/smh-not-normal}.
+
+\leavevmode\hypertarget{ref-doh}{}%
 Office of Management and Budget.
 1997.
-{``Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity.''} \url{https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-opendata}{}}%
-Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-patchwork}{}}%
-Pedersen, Thomas Lin. 2020. \emph{{patchwork: The Composer of Plots}}. \url{https://CRAN.R-project.org/package=patchwork}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-MichaelRPergamit2001DWTN}{}}%
+``Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity.'' \url{https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf}.
+
+\leavevmode\hypertarget{ref-opendata}{}%
+Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.
+
+\leavevmode\hypertarget{ref-patchwork}{}%
+Pedersen, Thomas Lin. 2020. \emph{patchwork: The Composer of Plots}. \url{https://CRAN.R-project.org/package=patchwork}.
+
+\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
 Pergamit, Michael R., Charles R.
@@
 2001.
-{``Data Watch: The National Longitudinal Surveys.''} \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
-
-\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
+``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
+
+\leavevmode\hypertarget{ref-R}{}%
 R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-SingerJudithD2003Alda}{}}%
+\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
 Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.
 
-\leavevmode\vadjust pre{\hypertarget{ref-notiris}{}}%
-Stodel, Megan. 2020. {``Stop Using Iris.''} \url{https://www.meganstodel.com/posts/no-to-iris/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-brolgar}{}}%
-Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}}. \url{https://github.com/njtierney/brolgar}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tukey}{}}%
+\leavevmode\hypertarget{ref-notiris}{}%
+Stodel, Megan. 2020. ``Stop Using Iris.'' \url{https://www.meganstodel.com/posts/no-to-iris/}.
+
+\leavevmode\hypertarget{ref-brolgar}{}%
+Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.
+
+\leavevmode\hypertarget{ref-tukey}{}%
 Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.
 
-\leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%
-van der Loo, Mark P. J., and Edwin de Jonge. 2021. {``Data Validation Infrastructure for {R}.''} \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-LooMarkvander2018Sdcw}{}}%
-van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in r}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-mass}{}}%
-Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with s}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tsibble}{}}%
+\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
+van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.
+
+\leavevmode\hypertarget{ref-validate}{}%
+van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
+
+\leavevmode\hypertarget{ref-mass}{}%
+Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
+
+\leavevmode\hypertarget{ref-tsibble}{}%
 Wang, Earo, Dianne Cook, and Rob J Hyndman.
 2020.
-{``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.''} \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
+``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
 \url{https://doi.org/10.1080/10618600.2019.1695624}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
+\leavevmode\hypertarget{ref-plyr}{}%
 Wickham, Hadley.
 2011.
-{``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29.
-https://doi.org/\url{https://doi.org/10.18637/jss.v040.i01}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-WickhamHadley2014TD}{}}%
----------. 2014. {``Tidy Data.''} \emph{Journal of Statistical Software} 59 (10): 1--23.
-
-\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
----------. 2016. \emph{{ggplot2: Elegant Graphics for Data Analysis}}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-stringr}{}}%
----------. 2019. \emph{{stringr: Simple, Consistent Wrappers for Common String Operations}}. \url{https://CRAN.R-project.org/package=stringr}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tidyr}{}}%
----------. 2020. \emph{{tidyr: Tidy Messy Data}}. \url{https://CRAN.R-project.org/package=tidyr}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
+``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29.
+\url{https://doi.org/https://doi.org/10.18637/jss.v040.i01}.
+
+\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
+---------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.
+
+\leavevmode\hypertarget{ref-ggplot2}{}%
+---------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
+
+\leavevmode\hypertarget{ref-stringr}{}%
+---------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.
+
+\leavevmode\hypertarget{ref-tidyr}{}%
+---------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.
+
+\leavevmode\hypertarget{ref-tidyverse}{}%
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al.
 2019.
-{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686.
+``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686.
 \url{https://doi.org/10.21105/joss.01686}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
-Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{{dplyr: A Grammar of Data Manipulation}}. \url{https://CRAN.R-project.org/package=dplyr}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-readr}{}}%
-Wickham, Hadley, and Jim Hester. 2020. \emph{{readr: Read Rectangular Text Data}}. \url{https://CRAN.R-project.org/package=readr}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlsy79edu}{}}%
+\leavevmode\hypertarget{ref-dplyr}{}%
+Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.
+
+\leavevmode\hypertarget{ref-readr}{}%
+Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.
+
+\leavevmode\hypertarget{ref-nlsy79edu}{}%
 Wolpin, Kenneth I.
 2005.
-{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Published by Bureau of Labor Statistics, U.S.
+``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Published by Bureau of Labor Statistics, U.S.
 Department of Labor.

### 2022-05-11 paper/paper.Rmd
Finished address all of the comments.

@@
 The current US Federal guidelines [@doh] state that there should be at least five categories for race, and an individual should be able to identify as more than one race category.
-in addition, "Hispanic" or "Latino" should be considered to be  an ethnic group rather than a race.
+In addition, "Hispanic" or "Latino" should be considered to be  an ethnic group rather than a race.
 This level of detail is not available in the database.
@@
 This variable is obtained from `new_data_qnames` with the name `HGC_EVER_XRND` and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as `hgc_i` and recoded as a factor to give `hgc` (e.g., 10th grade, 11th grade, and so on).
-We decide to include both
+We decide to include both the factor and integer values to be used as examples in various classroom demonstrations, such as to explain different data types.
+Besides, it can also be used as the demonstration of different type of data visualization which is determined by the data type.
+It can also be used for teaching the modeling of cross sectional data.
 
@@
 This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals.
+Hence, different from `hgc`, this variable could be used to demonstrate longitudinal and time series analysis in the classroom.
 This is recorded in `new_data_qnames` as columns beginning with `Q3-4` and `HGC` with year as a suffix.
@@
 
-```{r flow-chart-blind, fig.cap="(ref:flow-chart)", out.width="85%", eval = CENSOR}
+```{r flow-chart-blind, fig.cap="(ref:flowchart-blind)", out.width="85%", eval = CENSOR}
 ```

### 2022-05-11 paper/paper.tex
Finished address all of the comments.

@@
 
-Thus, we re-arrange and wrangle the data into tidy data form, columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
+Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
 This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
-
-For \texttt{gender} and \texttt{race}, we only rename the column names provided in the raw data, \texttt{new\_data\_qnames}.
-It is worth noting that \texttt{gender} comes from a variable called \texttt{sex} in the database.
-We use the term \texttt{gender} as the reports about the data on various pages in the website more commonly use \texttt{gender}.
-Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female.'' The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
-Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification (Kennedy et al.
-2020).
-From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex (Kennedy et al.
-2020).
-Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
-This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
-Similarly, this also relates to race, as reported in the database.
-When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
-
-The object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
+Our final target data will include the demographic variables (variable names): sex (\texttt{sex}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
+
+For \texttt{sex} and \texttt{race}, we have simplified the original names provided in the raw data, \texttt{SAMPLE\_SEX\_1979} and \texttt{SAMPLE\_RACE\_78SCRN}, respectively.
+However, it is important to note that the use of ``sex'' is probably not correct.
+This information is provided by the individual, and although only has two categories, is more consistent with ``gender'' as defined in Heidari et al.
+(2016).
+When using the dataset in the classroom, the educators might include discourse on the use of terms ``sex'' and ``gender''.
+Particularly, measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or create measurement error for people whose gender does not align with their sex classification.
+From a statistical perspective, this can make it difficult to adjust survey statistics to the population when gender is measured in different categorization in different data sources.
+Kennedy et al.
+(2020) provides a helpful discussion on these topics.
+
+Similarly, there are issues for the \texttt{race} variable.
+The current US Federal guidelines (Office of Management and Budget 1997) state that there should be at least five categories for race, and an individual should be able to identify as more than one race category.
+In addition, ``Hispanic'' or ``Latino'' should be considered to be an ethnic group rather than a race.
+This level of detail is not available in the database.
+There is one variable with only three categories, and an individual can be a member of only one.
+Thus, the \texttt{race} variable is inadequate by today's standards, and educators should point to the current Federal guidelines when using this variable, and keep in mind that the purpose of any analysis is to study racism rather than race.
+
+Further, the object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on).
+We decide to include both the factor and integer values to be used as examples in various classroom demonstrations, such as to explain different data types.
+Besides, it can also be used as the demonstration of different type of data visualization which is determined by the data type.
+It can also be used for teaching the modeling of cross sectional data.
 
@@
 \midrule
-Male & 1,000 (16\%) & 1,613 (25\%) & 3,790 (59\%) & 6,403\\
-Female & 1,002 (16\%) & 1,561 (25\%) & 3,720 (59\%) & 6,283\\
+F & 1,561 (25\%) & 1,002 (16\%) & 3,720 (59\%) & 6,283\\
+M & 1,613 (25\%) & 1,000 (16\%) & 3,790 (59\%) & 6,403\\
 \midrule
-Total & 2,002 (16\%) & 3,174 (25\%) & 7,510 (59\%) & 12,686\\
+Total & 3,174 (25\%) & 2,002 (16\%) & 7,510 (59\%) & 12,686\\
 \bottomrule
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart}
 
@@
 ``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14.
-\url{https://doi.org/10.1080/00031305.2015.1081106}.
+\url{https://doi.org/https://doi.org/10.1080/00031305.2015.1081106}.
+
+\leavevmode\hypertarget{ref-SAGER}{}%
+Heidari, S., T.
+F.
+Babor, P.
+De Castro, S.
+Tort, and M.
+Curno.
+2016.
+``Sex and Gender Equity in Research: Rationale for the SAGER Guidelines and Recommended Use.'' \emph{Research Integrity and Peer Review} 1 (2).
+\url{https://doi.org/https://doi.org/10.1186/s41073-016-0007-6}.
 
@@
 ``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1).
-\url{https://doi.org/10.5070/T511103589}.
+\url{https://doi.org/https://doi.org/10.5070/T511103589}.
 
@@
 Moncrief, Marc. 2015. ``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.'' \url{https://bit.ly/smh-not-normal}.
+
+\leavevmode\hypertarget{ref-doh}{}%
+Office of Management and Budget.
+1997.
+``Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity.'' \url{https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf}.
 
@@
 \leavevmode\hypertarget{ref-plyr}{}%
-Wickham, Hadley. 2011. ``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29. \url{https://doi.org/10.18637/jss.v040.i01}.
+Wickham, Hadley.
+2011.
+``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29.
+\url{https://doi.org/https://doi.org/10.18637/jss.v040.i01}.
 

### 2022-05-11 response/response2.Rmd
Finished address all of the comments.

@@
 
-* We have added more explanation of the two variables, and suggestions of where to use one and the other in classroom analyses. XXX Dewi to do
+* We have added more explanation of the two variables, and suggestions of where to use one and the other in classroom analyses. XXX DONE
 
@@
 
-* Done XXX Dewi
+* We have modified the discussion around HRP to be more concise about what we have done with the variable.
+On page 9, we use the HRP to demonstrate the untidy form of the data.
+Hence, the sentence “HRP1 1980 and HRP2 1980, contain the information about the job number up to 5...” has been removed.
+In another part of the manuscript, we explain that if a respondent has multiple jobs, the `mean_hourly_wage` is computed as a weighted average of the hourly wage (`HRP`) for each job with the number of hours worked for each job as weights (provided that the information on the number of hours is available); if the number of hours worked for any job is missing, then the `mean_hourly_wage` is computed as a simple average of all available hourly wages.
+We hope that our modification makes the manuscript clearer.
+XXX DONE
 
@@
 
-* XXX Dewi Add some words to make it clear what data this is summarising
+* The tables show the number by characteristics (age, sex, and race) in the input data (referring the term in the de Jonge and van der Loo's statistical value chain), i.e., the data we have extracted and tidied from the database.
+We are not saying it as a clean data as we have not treat the extreme values.
+We have modified the tables' caption to make this clearer.
+XXX DONE
 

### 2022-05-11 data-raw/ida.R

@@ -15,7 +15,7 @@ age_table <- full_demographics %>%
 age_table %>%
   mutate(n = scales::comma(n)) %>%
   kable(
-      caption = "Frequency table of the age at the start of the survey in NSLY79 cohort",
+      caption = "Frequency table of the age at the start of the survey in NSLY79 cohort in the extracted data.",
       col.names = c("Age", "Number of individuals"),
       booktabs = TRUE,
       linesep = "",
@@ -40,7 +40,7 @@ sex_race_table <- full_demographics %>%
          Total = str_replace(Total, "\\([1|2]00\\%\\)", ""))
 
 kable(sex_race_table,
-      caption = "Contingency table for sex and race for the full NLSY79 data. The percentage (rounded to closest 1\\%) is out of the total corresponding to row.",
+      caption = "Contingency table for sex and race for the extracted NLSY79 demographic data. The percentage (rounded to closest 1\\%) is out of the total corresponding to row.",
       col.names = c("Sex", "Hispanic", "Black", "Non-Black, Non-Hispanic", "Total"),
       booktabs = TRUE,
       linesep = "",
@@ -301,9 +301,9 @@ wages <- wages_clean %>%
 # Create a data set for the high school dropouts cohort
 wages_hs_do <- wages %>%
   # based on two criteria mentioned in the paper
-  filter(hgc_i < 12 | (hgc_i >= 12 & ged != 1)) %>%
+  filter(hgc_i < 12 | (hgc_i >= 12 & ged == 2)) %>%
   filter(age_1979 <= 17,
-         sex == "MALE") %>%
+         sex == "m") %>%
   as_tsibble(key = id,
              index = year,
              regular = FALSE)
@@ -385,10 +385,10 @@ n_obs_do_pred <- filter(wages_hs_do, is_pred == TRUE) %>%
   nrow()
 
 ### ---- flow-chart
-include_graphics(here::here("paper/figures/flowchart_blind.png"))
+include_graphics(here::here("paper/figures/flowchart.png"))
 
 ### ---- flow-chart-blind
-include_graphics(here::here("paper/figures/flowchart_blind.png"))
+include_graphics(here::here("paper/figures/flowchart-blind.png"))
 
 ## The end of data refreshing procedure
 ## Go to "data-raw/eda" to see the code of comparison between original and refreshed data.

### 2022-05-11 paper/paper-jsdse/paper-jsdse.tex
intro done

--- 2022-05-11 paper/paper-jsdse/paper-jsdse.tex
@@
 The iris data (Anderson 1935) is a prime example.
-It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(It should be noted that the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)).
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis\footnote{It should be noted that the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)}.
 We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
@@
 As a textbook data set, though, the wages data is outdated.
-The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
+The most recent year in the data is 1994, 9 years prior to when Singer and Willett (2003) was published.
 Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students.
 Using tired old textbook data sets can imbue a misconception that the field is not current.
-The wages data is extracted from NSLY79, one of the best examples of open data (see details at Open Knowledge Foundation (2021)), which is constantly being updated.
+The wages data is extracted from NSLY79, one of the best examples of open data (see details at Open Knowledge Foundation 2021), which is constantly being updated.
 It should be possible to continuously refresh the textbook data from the data repository.
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{../figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{../figures/flowchart}
 

### 2022-05-11 paper/paper.Rmd
intro done

@@
 The iris data [@iris-data] is a prime example.
-It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis.
-(It should be noted that the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]).
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis^[It should be noted that the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]].
 We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
@@
 As a textbook data set, though, the wages data is outdated.
-The most recent year in the data is 1994, 10 years prior to when @SingerJudithD2003Alda was published.
+The most recent year in the data is 1994, 9 years prior to when @SingerJudithD2003Alda was published.
 Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students.
 Using tired old textbook data sets can imbue a misconception that the field is not current.
-The wages data is extracted from NSLY79, one of the best examples of open data (see details at @opendata), which is constantly being updated.
+The wages data is extracted from NSLY79, one of the best examples of open data [see details at @opendata], which is constantly being updated.
 It should be possible to continuously refresh the textbook data from the data repository.

### 2022-05-11 paper/paper.Rmd
section 2 done

@@
 We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the *purposes of studying racism, not race* [@racismnotrace].
-There are a number of data sets provided by @SingerJudithD2003Alda, and we focus only on this one because it has captivated our attention for a number of years.
-We use it in our own teaching of longitudinal data analysis and would very much like to use data since 1994.
+There are a number of data sets provided by @SingerJudithD2003Alda, but we focus only on wages data because it has captivated our attention for a number of years in our own teaching of longitudinal data analysis.
 <!--However, we cannot create the exact same data set as published in their book since we do not have the information of what age threshold they used to determine the high school dropouts.-->
@@
 
-1. The wages data of the whole NLSY79 cohort, including females.
-2. A separate table of the demographic data of the whole NLSY79 cohort.
+1. The wages data of the whole NLSY79 cohort, including females,
+2. A separate table of the demographic data of the whole NLSY79 cohort, and 
 3. The subset of wages data in (1), which is the wages of high school dropouts' as a refreshed version of @SingerJudithD2003Alda's data.
@@
 
-In the context of official statistics, @LooMarkvander2018Sdcw describe the "statistical value chain," which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+In the context of official statistics, @LooMarkvander2018Sdcw describe the "statistical value chain," which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
 What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.

### 2022-05-11 paper/paper-jsdse/paper-jsdse.tex
section 3 done

@@
 We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the \emph{purposes of studying racism, not race} (Fullilove 1998).
-There are a number of data sets provided by Singer and Willett (2003), and we focus only on this one because it has captivated our attention for a number of years.
-We use it in our own teaching of longitudinal data analysis and would very much like to use data since 1994.
+There are a number of data sets provided by Singer and Willett (2003), but we focus only on wages data because it has captivated our attention for a number of years in our own teaching of longitudinal data analysis.
 
@@
 \item
-  The wages data of the whole NLSY79 cohort, including females.
-\item
-  A separate table of the demographic data of the whole NLSY79 cohort.
+  The wages data of the whole NLSY79 cohort, including females,
+\item
+  A separate table of the demographic data of the whole NLSY79 cohort, and
 \item
@@
 In the context of official statistics, M.
-van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
 What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
@@
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
-The code for tidying the data are demonstrated at \url{https://github.com/numbats/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.
+The code for tidying the data are demonstrated at \url{https://numbats.github.io/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.
 
@@
 
-Further, the object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
+Furthermore, the object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis.
-The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively.
+The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, encoded as a factor and integer variable, respectively.
 For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}.
 This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on).
-We decide to include both the factor and integer values to be used as examples in various classroom demonstrations, such as to explain different data types.
-Besides, it can also be used as the demonstration of different type of data visualization which is determined by the data type.
-It can also be used for teaching the modeling of cross sectional data.
+We decide to include both the factor and integer variables to be used as examples in various classroom demonstrations to explain different data types say.
+Besides this, it can also serve as a demonstration of different type of data visualization ascribed by the data type and for teaching the modeling of cross sectional data.
 
@@
 As a respondent may have multiple jobs, the \texttt{mean\_hourly\_wage} is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on the number of hours is available); if the number of hours worked for any job is missing, then the \texttt{mean\_hourly\_wage} is computed as a simple average of all available hourly wages.
-Prior to computing the mean hourly wage, we undertook a number of steps to treat unusual observations as described below:
+Prior to computing the mean hourly wage, we undertook a number of steps to treat extreme observations as described below:
 
@@
 \item
-  If the hourly rate is recorded as 0, we set wage as missing.
-\item
-  If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.
+  If the hourly rate is recorded as 0, we set wage as missing, and
+\item
+  If the total hours of worked per week for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.
 \end{itemize}
@@
 
-There are many steps and decisions made to go from raw to input to valid data. Figure \ref{fig:flow-chart-blind} summarizes these in order to create a refreshed wages data set.
+There are many steps and decisions made to go from raw to input to valid data. Figure \ref{fig:flow-chart} summarizes these in order to create a refreshed wages data set.
 
@@
 \item
-  \texttt{hours:}: The total number of hours the individual usually works per week.
+  \texttt{hours}: The total number of hours the individual usually works per week.
 \item
@@
 \item
-  \textbf{Codes}: R script to reproduce data tidying and cleaning is available at \url{https://github.com/numbats/yowie/articles/process-data.html}.
-The code for the extreme value handling with robust linear fits is in \url{https://github.com/numbats/yowie/articles/input-to-valid-data.html}.
+  \textbf{Codes}: R script to reproduce data tidying and cleaning is available at \url{https://numbats.github.io/yowie/articles/process-data.html}.
+The code for the extreme value handling with robust linear fits is in \url{https://numbats.github.io/yowie/articles/input-to-valid-data.html}.
 \item

### 2022-05-11 paper/paper.Rmd
section 3 done

@@
 yowie_repo <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie")
-yowie_pkgdown <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie")
+yowie_pkgdown <- ifelse(CENSOR, "CENSORED", "https://numbats.github.io/yowie")
 yowie_code <- ifelse(CENSOR, "CENSORED", "https://github.com/numbats/yowie-paper/data-raw")
@@
 
-Further, the object `new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981`, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
+Furthermore, the object `new_data_qnames`, contains the variables `Q1-3_A~Y_1979` and `Q1-3_A~Y_1981`, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (`Q1-3_A~M_1979` and `Q1-3_A~M_1981`).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis.
-The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as `hgc` and `hgc_i`, for the factor and integer type, respectively.
+The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as `hgc` and `hgc_i`, encoded as a factor and integer variable, respectively.
 For each individual, there is only one value of `hgc` and `hgc_i`.
 This variable is obtained from `new_data_qnames` with the name `HGC_EVER_XRND` and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as `hgc_i` and recoded as a factor to give `hgc` (e.g., 10th grade, 11th grade, and so on).
-We decide to include both the factor and integer values to be used as examples in various classroom demonstrations, such as to explain different data types.
-Besides, it can also be used as the demonstration of different type of data visualization which is determined by the data type.
-It can also be used for teaching the modeling of cross sectional data.
+We decide to include both the factor and integer variables to be used as examples in various classroom demonstrations to explain different data types say.
+Besides this, it can also serve as a demonstration of different type of data visualization ascribed by the data type and for teaching the modeling of cross sectional data.
 
@@
 As a respondent may have multiple jobs, the `mean_hourly_wage` is computed as a weighted average of the hourly wage for each job with the number of hours worked for each job as weights (provided that the information on the number of hours is available); if the number of hours worked for any job is missing, then the `mean_hourly_wage` is computed as a simple average of all available hourly wages.
-Prior to computing the mean hourly wage, we undertook a number of steps to treat unusual observations as described below:
+Prior to computing the mean hourly wage, we undertook a number of steps to treat extreme observations as described below:
 
-* If the hourly rate is recorded as 0, we set wage as missing.
-* If the total hours of worked for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.
+* If the hourly rate is recorded as 0, we set wage as missing, and
+* If the total hours of worked per week for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.
 
@@
   12. `njobs`: Number of jobs that an individual has.
-  13. `hours:`: The total number of hours the individual usually works per week.
+  13. `hours`: The total number of hours the individual usually works per week.
   14. `stwork`: The year when the individual started to work.

### 2022-05-12 paper/paper.Rmd
section 4 done

@@
 The original set, containing wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994, is available in the R package `brolgar`.
-To compare the refreshed data with the original, a subset needs to be matched.
-There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data and restricting the longitudinal measurements to the same years.
-However, we decided to try to replicate the process, as suggested by the description of the original data.
-This requires first identifying individuals who dropped out of high school.
+To compare the refreshed data with the original, we need to subset the data into a matching cohort of high school dropouts.
+There are multiple ways to match a comparative cohort from the updated data -- the simplest approach is to subset to individuals with matching IDs in the original data and restricting the longitudinal measurements to the same years; another approach is to replicate the process as suggested by the description of the original data thus identifying a new (and possibly removal of old) set of individuals that is included in the refreshed data.
+We use the latter approach and as the first step we identify individuals who dropped out of high school.
 
@@
 1. An individual whose highest grade completed (`hgc`) is reported to be less than 12th grade, **or**
-2. An individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and has received a GED (`ged` is code to 2).
+2. An individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and has received a GED (i.e. `ged` code is 2).
 
@@
 2. 79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
-    i. 35 of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded to 1). This suggests they are not dropouts, and so we excluded them from the refreshed data. 
+    i. 35 of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded as 1). This suggests they are not dropouts, and so we excluded them from the refreshed data. 
     ii. The information about `ged` is missing in 38 individuals. We decided to include them in the refreshed data. 
-    iii. 3 individuals have both diploma and GED (`ged` is coded to 3). These were kept in the refreshed data. 
+    iii. 3 individuals have both diploma and GED (`ged` is coded as 3). These were kept in the refreshed data. 
     iv. 12 individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.
@@
 Figure \ref{fig:compare-subsets} contains summaries of corresponding variables in the two subsets.
-Plot (A) shows a back-back bar chart of the highest grade completed.
+Plot (A) shows a back-to-back bar chart of the highest grade completed.
 The two sets are almost the same, but small differences remain.

### 2022-05-12 paper/paper-jsdse/paper-jsdse.tex
fixed all of Kate changes

@@
 We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation.
-We occasionally hear this voiced in the public media media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian (Moncrief 2015).
+We occasionally hear this voiced in the public media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian (Moncrief 2015).
 Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.
@@
 
-\begin{figure}[t]
-
-\begin{tcolorbox}[title = Navigating the data source]
+\begin{figure*}[p]
+
+\begin{tcolorbox}[title = Navigating the data source, halign=center,fontupper=\linespread{.66}\selectfont]{\fontsize{10pt}{10pt}}
 \faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
@@
 \caption{Documented steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
-\end{figure}
+\end{figure*}
 
@@
 This is better to do with each wave of new data added so that it is relative to the last date in the data.
-Our decision was to provide the raw wages and include codes to make the adjustment as part of the package.
+Our decision was to provide the raw wages and include code to make the adjustment as part of the package.
 

### 2022-05-12 paper/paper.Rmd
fixed all of Kate changes

--- 2022-05-11 paper/paper.Rmd
@@
 The majority of people in the study do not have a pattern that is similar to the overall trend.
-If you have experienced declining wages, you are not alone; there are others like you, and more than you think.--> We occasionally hear this voiced in the public media media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage].
+If you have experienced declining wages, you are not alone; there are others like you, and more than you think.--> We occasionally hear this voiced in the public media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage].
 Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.
@@
 
-\begin{figure}[t]
+\begin{figure*}[p]
 
-\begin{tcolorbox}[title = Navigating the data source]
+\begin{tcolorbox}[title = Navigating the data source, halign=center,fontupper=\linespread{.66}\selectfont]{\fontsize{10pt}{10pt}}
 \faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\

### 2022-05-12 paper/paper-jsdse/paper-jsdse.tex
section 5 done

--- 2022-05-11 paper/paper-jsdse/paper-jsdse.tex
@@
 The original set, containing wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994, is available in the R package \texttt{brolgar}.
-To compare the refreshed data with the original, a subset needs to be matched.
-There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data and restricting the longitudinal measurements to the same years.
-However, we decided to try to replicate the process, as suggested by the description of the original data.
-This requires first identifying individuals who dropped out of high school.
+To compare the refreshed data with the original, we need to subset the data into a matching cohort of high school dropouts.
+There are multiple ways to match a comparative cohort from the updated data -- the simplest approach is to subset to individuals with matching IDs in the original data and restricting the longitudinal measurements to the same years; another approach is to replicate the process as suggested by the description of the original data thus identifying a new (and possibly removal of old) set of individuals that is included in the refreshed data.
+We use the latter approach and as the first step we identify individuals who dropped out of high school.
 
@@
 \item
-  An individual whose highest grade completed (\texttt{hgc}) is reported to be at least 12th grade and has received a GED (\texttt{ged} is code to 2).
+  An individual whose highest grade completed (\texttt{hgc}) is reported to be at least 12th grade and has received a GED (i.e.~\texttt{ged} code is 2).
 \end{enumerate}
@@
   \item
-    35 of them completed at least 12th grade with a diploma instead of GED (\texttt{ged} variable is coded to 1). This suggests they are not dropouts, and so we excluded them from the refreshed data.
+    35 of them completed at least 12th grade with a diploma instead of GED (\texttt{ged} variable is coded as 1). This suggests they are not dropouts, and so we excluded them from the refreshed data.
   \item
@@
   \item
-    3 individuals have both diploma and GED (\texttt{ged} is coded to 3). These were kept in the refreshed data.
+    3 individuals have both diploma and GED (\texttt{ged} is coded as 3). These were kept in the refreshed data.
   \item
@@
 The wages data is an example of this.
-Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race.
-If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., sex and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
+Students can take away multiple interpretations from the data (e.g.~wages increase with experience, education, and race) based on the teaching focus to learn about the world.
+If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., sex and race), or data that has been poorly processed containing errors, we, as educators, are being irresponsible because the societal message taught to students may be flawed.
 This paper demonstrates the process of producing a trustworthy data set for teaching.

### 2022-05-12 paper/paper.Rmd
section 5 done

--- 2022-05-12 paper/paper.Rmd
@@
 The wages data is an example of this.
-Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race.
-If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., sex and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
+Students can take away multiple interpretations from the data (e.g.
+wages increase with experience, education, and race) based on the teaching focus to learn about the world.
+If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., sex and race), or data that has been poorly processed containing errors, we, as educators, are being irresponsible because the societal message taught to students may be flawed.
 This paper demonstrates the process of producing a trustworthy data set for teaching.

### 2022-05-12 paper/paper-jsdse/paper-jsdse.tex
format of figure 1

@@
 
-\begin{tcolorbox}[title = Navigating the data source, halign=center,fontupper=\linespread{.66}\selectfont]{\fontsize{10pt}{10pt}}
-\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
-\vspace{1mm}
+\begin{tcolorbox}[title = Navigating the data source, fontupper=\linespread{.8}\selectfont]
+\vspace{3mm}
+\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\\\
+\vspace{3mm}
 \faCheck\ The CASEID will be always be selected.  \\
 \vspace{1mm}
-\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
-\vspace{1mm}
-For the remaining variables, we went to the "Variable Search" tab and select variables as follows
+\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  For the remaining variables, we went to the ``Variable Search" tab and select variables as follows
 \begin{itemize}
@@
 \begin{itemize}
-\item[\faCloudDownload] To download all 742 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download".
+\item[\faCloudDownload] To download all 742 variables selected, we then navigate to the tab ``Save / Download" then select the tab ``Advanced Download".
 We select the R Source code and comma-delimited datafile of selected variables with Reference Number as column headers.
-We name the filename "NLSY79" and press the download button.
+We name the filename ``NLSY79" and press the download button.
 There are also options to get control or dictionary files for SAS, SPSS and STATA.
@@
 \item
-  \texttt{NLSY79.R}: R script for reading the data into R and converting the variables' names and label into something more sensible.
+  \texttt{NLSY79.R}: R script for reading the data into R and converting the variable names and label into something more sensible.
 \end{itemize}
@@
 The wages data is an example of this.
-Students can take away multiple interpretations from the data (e.g.~wages increase with experience, education, and race) based on the teaching focus to learn about the world.
-If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., sex and race), or data that has been poorly processed containing errors, we, as educators, are being irresponsible because the societal message taught to students may be flawed.
+Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race.
+If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., sex and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
 This paper demonstrates the process of producing a trustworthy data set for teaching.

### 2022-05-12 paper/paper.Rmd
format of figure 1

--- 2022-05-12 paper/paper.Rmd
@@
 - `NLSY79.NLSY79`: tagset of variables that can be uploaded to the website to recreate the data set, and
-- `NLSY79.R`: R script for reading the data into R and converting the variables' names and label into something more sensible.
+- `NLSY79.R`: R script for reading the data into R and converting the variable names and label into something more sensible.
 

### 2022-05-12 paper/paper.tex
format of figure 1

--- 2022-05-10 paper/paper.tex
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-The wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
-
-This disparity between the average and the individual is a part of statistics as a discipline that requires more attention.
-This particular data set is a prime example of discussing this disparity.
-Textbook data sets have longevity if there is an unresolved mystery.
-The iris data (Andereson 1935) is a prime example.
-It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis (a side note: the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)).
+It is for these reasons that the wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
+
+This disparity between the overall trend and the individual heterogeneity is what makes this data interesting.
+Textbook data sets have longevity if they have unresolved elements.
+The iris data (Anderson 1935) is a prime example.
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis\footnote{It should be noted that the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)}.
 We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
@@
 For statistics, and data science by association, it is increasingly important to reach the individual.
-One might describe this as a divergence of purpose -- statistics for public policy -- or statistics for the public.
+One might think of this as a divergence of purpose -- statistics for public policy -- or statistics for the public.
 The two are not the same.
@@
 For example, with the wages data, the overall trend, across demographics, is a steady increase in wages, but the patterns among individuals is heterogeneous.
-Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year.
-The heterogeneous patterns occur across demographics.
-The subject-specific variation explains most of the variation in wages, even though the overall trend is statistically significant, it is weak.
-The the message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable.
-However, this information is the appropriate message for an individual is that you as an individual are far more than your demographic.
-The majority of people in the study do not have a pattern that is similar to the overall trend.
-If you have experienced declining wages, you are not alone; there are others like you, and more than you think.
-A similar tone is echoed occasionally in the public media.
-For example, an article published in the Sydney Morning Herald argues there is no average Australian (Moncrief 2015).
+Some individuals see a decline in their wages, some an increase, some have dramatic ups and downs from year to year, and the heterogeneity of the patterns are evident across demographics.
+Subject-specific variation explains most of the variation in wages, and even though the overall trend is statistically significant, it is weak.
+The message for public policy is that demographic profile is related to different wage patterns on average and some structural social change is desirable.
+However, discussing the overall trend with an individual is misleading -- more than likely their pattern is completely different.
+It could even be depressing for them to compare themselves to the overall average, especially if the individual's wages have a declining trend or high volatility.
+It would be more useful for an individual to know what percentage of the population have profiles like theirs.
+We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation.
+We occasionally hear this voiced in the public media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian (Moncrief 2015).
+Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.
 
 As a textbook data set, though, the wages data is outdated.
-The most recent year in the data is 1994, 10 years prior to when Singer and Willett (2003) was published.
+The most recent year in the data is 1994, 9 years prior to when Singer and Willett (2003) was published.
 Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students.
 Using tired old textbook data sets can imbue a misconception that the field is not current.
-The wages data is extracted from NSLY79, one of the best examples of open data, which is constantly being updated.
+The wages data is extracted from NSLY79, one of the best examples of open data (see details at Open Knowledge Foundation 2021), which is constantly being updated.
 It should be possible to continuously refresh the textbook data from the data repository.
-This paper describes our (non-glamorous) journey from open (Open Knowledge Foundation 2021) wild data to textbook data.
+This paper describes our (non-glamorous) journey from open wild data to textbook data.
 
@@
 
-\begin{figure}[t]
-
-\begin{tcolorbox}[title = Navigating the data source]
+\begin{figure*}[p]
+
+\begin{tcolorbox}[title = Navigating the data source, fontupper=\linespread{.66}\selectfont]{\fontsize{10pt}{10pt}}
 \faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
@@
 \vspace{1mm}
-For the remaining variables, we went to the "Variable Search" tab and select variables as follows
+For the remaining variables, we went to the ``Variable Search" tab and select variables as follows
 \begin{itemize}
@@
 
-In our final target data, we wish to include the demographic variables with variable names specified in brackets: gender (\texttt{gender}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
-
-For \texttt{gender} and \texttt{race}, we only rename the column names provided in the raw data, \texttt{new\_data\_qnames}.
-It is worth noting that \texttt{gender} comes from a variable called \texttt{sex} in the database.
-We use the term \texttt{gender} as the reports about the data on various pages in the website more commonly use \texttt{gender}.
-Gender, as provided in the data, is self-reported and only has two categories, ``male'' and ``female.'' The interchangeable use of these terms reflects that gender and sex are sometimes regarded as similar in surveys, although they are not.
-Measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or people whose gender does not align with their sex classification (Kennedy et al.
-2020).
-From a statistical perspective, this sometimes difficult the adjustments of survey statistics to the population when gender is measured, for example, with three categories, and the census measured it with only two categories and used the term sex (Kennedy et al.
-2020).
-Hence, for modern societal studies, it is highly recommended to separate sex and gender in the survey question and provide more options to respond to questions on gender.
-This discourse should also be emphasized in teaching that using these two variables should be accompanied by special attention.
-Similarly, this also relates to race, as reported in the database.
-When doing an analysis with this variable, one should keep in mind that the purpose is to study racism rather than race.
-
-The object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
+Our final target data will include the demographic variables (variable names): sex (\texttt{sex}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
+
+For \texttt{sex} and \texttt{race}, we have simplified the original names provided in the raw data, \texttt{SAMPLE\_SEX\_1979} and \texttt{SAMPLE\_RACE\_78SCRN}, respectively.
+However, it is important to note that the use of ``sex'' is probably not correct.
+This information is provided by the individual, and although only has two categories, is more consistent with ``gender'' as defined in Heidari et al.
+(2016).
+When using the dataset in the classroom, the educators might include discourse on the use of terms ``sex'' and ``gender''.
+Particularly, measuring gender as a binary variable has the potential to fail to capture people who do not identify themselves as either male or female or create measurement error for people whose gender does not align with their sex classification.
+From a statistical perspective, this can make it difficult to adjust survey statistics to the population when gender is measured in different categorization in different data sources.
+Kennedy et al.
+(2020) provides a helpful discussion on these topics.
+
+Similarly, there are issues for the \texttt{race} variable.
+The current US Federal guidelines (Office of Management and Budget 1997) state that there should be at least five categories for race, and an individual should be able to identify as more than one race category.
+In addition, ``Hispanic'' or ``Latino'' should be considered to be an ethnic group rather than a race.
+This level of detail is not available in the database.
+There is one variable with only three categories, and an individual can be a member of only one.
+Thus, the \texttt{race} variable is inadequate by today's standards, and educators should point to the current Federal guidelines when using this variable, and keep in mind that the purpose of any analysis is to study racism rather than race.
+
+Furthermore, the object \texttt{new\_data\_qnames}, contains the variables \texttt{Q1-3\_A\textasciitilde{}Y\_1979} and \texttt{Q1-3\_A\textasciitilde{}Y\_1981}, which records two versions of the birth year of the respondent; this is also the case for the record of birth month (\texttt{Q1-3\_A\textasciitilde{}M\_1979} and \texttt{Q1-3\_A\textasciitilde{}M\_1981}).
 The record contains two versions of birth year and birth month as the survey recorded this in 1979 and 1981.
@@
 There are several ways to define this, and this should be reflected in the refreshed data to give some flexibility for downstream analysis.
-The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, for the factor and integer type, respectively.
+The first one is the highest grade ever completed is reported in the database and provided in the refreshed data as \texttt{hgc} and \texttt{hgc\_i}, encoded as a factor and integer variable, respectively.
 For each individual, there is only one value of \texttt{hgc} and \texttt{hgc\_i}.
 This variable is obtained from \texttt{new\_data\_qnames} with the name \texttt{HGC\_EVER\_XRND} and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as \texttt{hgc\_i} and recoded as a factor to give \texttt{hgc} (e.g., 10th grade, 11th grade, and so on).
+We decide to include both the factor and integer variables to be used as examples in various classroom demonstrations to explain different data types say.
+Besides this, it can also serve as a demonstration of different type of data visualization ascribed by the data type and for teaching the modeling of cross sectional data.
 
@@
 
-\caption{\label{tab:gender-race-table}Contingency table for gender and race for the full NLSY79 data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
+\caption{\label{tab:gender-race-table}Contingency table for sex and race for the extracted NLSY79 demographic data. The percentage (rounded to closest 1\%) is out of the total corresponding to row.}
 \centering
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{figures/flowchart_blind}
+{\centering \includegraphics[width=0.85\linewidth]{figures/flowchart}
 
@@
 
-This paper has illustrated the steps and decisions made to take a particular open data set and make it a textbook data set, ready for the classroom or research.
+Any longitudinal dataset used for education should have a sufficiently reproducible process to be refreshed with new data.
+The NLSY79 dataset is a great teaching tool that has become outdated, both because the dataset stops in 1994, and because the demographic data could be handled more delicately.
+This paper has illustrated the necessary steps and decisions made to take a particular open data set and make it a textbook data set, ready for the classroom or research.
 In the first stage, we showed the steps performed to get the data from the NLSY79 database.

### 2022-05-12 paper/paper-jsdse/paper-jsdse.tex
Change the number to r inline code.

@@
 
-In the context of official statistics, M.
-van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+In the context of official statistics, van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
 What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{../figures/flowchart}
+{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart}
 

### 2022-05-12 paper/paper.Rmd
Change the number to r inline code.

@@
 An additional criterion from @SingerJudithD2003Alda is to only include males aged between 14 and 17 years old in 1979.
-With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
+With this filtering, we obtained `r ori_do_nrow` individuals in the refreshed data compared to `r sw_do_nrow` individuals in the original data.
 To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
@@
 1.
-173 individuals were more than 17 years old in 1979.
+`r nrow(join_more_17)` individuals were more than 17 years old in 1979.
 Thus, it looks like the description of the original data is not quite accurate that there are people older than 17 in the subset.
 Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
-2. 79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
-    i. 35 of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded as 1). This suggests they are not dropouts, and so we excluded them from the refreshed data. 
-    ii. The information about `ged` is missing in 38 individuals. We decided to include them in the refreshed data. 
-    iii. 3 individuals have both diploma and GED (`ged` is coded as 3). These were kept in the refreshed data. 
-    iv. 12 individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.
+2. `r nrow(join_less_17)` individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
+    i.
+`r nrow(join_not_elig)` of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded as 1).
+This suggests they are not dropouts, and so we excluded them from the refreshed data.
+    ii. The information about `ged` is missing in `r nrow(join_missing_ged)` individuals. We decided to include them in the refreshed data. 
+    iii. `r nrow(join_ged_both)` individuals have both diploma and GED (`ged` is coded as 3). These were kept in the refreshed data. 
+    iv. `r nrow(join_not_exist)` individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.
 
-The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing 863 individuals.
+The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing `r n_do` individuals.
 

### 2022-05-12 paper/paper.tex
Change the number to r inline code.

--- 2022-05-11 paper/paper.tex
@@
 
-In the context of official statistics, van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (data in the initial that it arrives), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+In the context of official statistics, van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
 What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
@@
 
-\begin{figure}[t]
-
-\begin{tcolorbox}[title = Navigating the data source]
-\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
-\vspace{1mm}
+\begin{figure*}[p]
+
+\begin{tcolorbox}[title = Navigating the data source, fontupper=\linespread{.8}\selectfont]
+\vspace{3mm}
+\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\\\
+\vspace{3mm}
 \faCheck\ The CASEID will be always be selected.  \\
 \vspace{1mm}
-\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
-\vspace{1mm}
-For the remaining variables, we went to the "Variable Search" tab and select variables as follows
+\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  For the remaining variables, we went to the ``Variable Search" tab and select variables as follows
 \begin{itemize}

### 2022-05-12 data-raw/ida.R

@@ -311,7 +311,7 @@ wages_hs_do <- wages %>%
 
 # investigate the dropouts IDs whether it agrees with the IDs in the original data
 
-# number of IDs in the original data
+# number of IDs in the refreshed data
 ori_do_nrow <- wages_hs_do %>% as_tibble() %>% count(id) %>% nrow()
 
 # create a negation function
@@ -319,6 +319,8 @@ ori_do_nrow <- wages_hs_do %>% as_tibble() %>% count(id) %>% nrow()
 
 # read the original data
 sw <- brolgar::wages
+sw_do_nrow <- sw %>% as_tibble() %>% count(id) %>% nrow()
+
 
 # filter the IDs in the refreshed data set that do not exist in the original data
 not_in_sw <- sw %>% filter(id %!in% wages_hs_do$id) %>% distinct(id, .keep_all = TRUE)

### 2022-05-12 paper/paper-jsdse/paper-jsdse.tex
Rewrite last paragraph in section 4.

@@
 
-Lastly, the race variable is almost perfectly matched. The \texttt{id}'s in the dropouts subset of the refreshed data all correspond to males, which perfectly matches the original subset.
+Lastly, the individuals in the dropouts subset of the refreshed data (as matched by their \texttt{id}s) belong to the same race as the individuals in the original data.
+This is also the case in the \texttt{sex} variables, which all correspond to males.
 

### 2022-05-12 paper/paper.Rmd
Rewrite last paragraph in section 4.

@@
 
-Lastly, the race variable is almost perfectly matched. The `id`'s in the dropouts subset of the refreshed data all correspond to males, which perfectly matches the original subset.
+Lastly, the individuals in the dropouts subset of the refreshed data (as matched by their `id`s) belong to the same race as the individuals in the original data.
+This is also the case in the `sex` variables, which all correspond to males.
 

### 2022-05-12 paper/paper.Rmd
changed first data print so that it fits into page, and the paragraph on race matching

--- 2022-05-12 paper/paper.Rmd
@@
 
-Lastly, the race variable is almost perfectly matched. The `id`'s in the dropouts subset of the refreshed data all correspond to males, which perfectly matches the original subset.
+Lastly, the matching of individuals in the dropouts subset of the refreshed data with the original data was done using their `id`s.
+(Only males were in the original data.) It was comforting to see that all of the individuals in both sets do match on sex (all are male!), and race.
 

### 2022-05-12 paper/paper.Rmd
code not codes

--- 2022-05-12 paper/paper.Rmd
@@
 The majority of people in the study do not have a pattern that is similar to the overall trend.
-If you have experienced declining wages, you are not alone; there are others like you, and more than you think.--> We occasionally hear this voiced in the public media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage].
+If you have experienced declining wages, you are not alone; there are others like you, and more than you think.--> We occasionally hear this voiced in the public media media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage].
 Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.
@@
 
-\begin{figure*}[p]
+\begin{figure}[t]
 
-\begin{tcolorbox}[title = Navigating the data source, fontupper=\linespread{.8}\selectfont]
-\vspace{3mm}
-\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\\\
-\vspace{3mm}
+\begin{tcolorbox}[title = Navigating the data source]
+\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
+\vspace{1mm}
 \faCheck\ The CASEID will be always be selected.  \\
 \vspace{1mm}
-\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  For the remaining variables, we went to the ``Variable Search" tab and select variables as follows
+\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
+\vspace{1mm}
+For the remaining variables, we went to the "Variable Search" tab and select variables as follows
 \begin{itemize}
@@
 \begin{itemize}
-\item[\faCloudDownload] To download all 742 variables selected, we then navigate to the tab ``Save / Download" then select the tab ``Advanced Download".
+\item[\faCloudDownload] To download all 742 variables selected, we then navigate to the tab "Save / Download" then select the tab "Advanced Download".
 We select the R Source code and comma-delimited datafile of selected variables with Reference Number as column headers.
-We name the filename ``NLSY79" and press the download button.
+We name the filename "NLSY79" and press the download button.
 There are also options to get control or dictionary files for SAS, SPSS and STATA.
@@
 \caption{Documented steps taken to select variables of interest and download the raw data.\label{fig:source-nav}}
-\end{figure*}
+\end{figure}
 
@@
 - `NLSY79.NLSY79`: tagset of variables that can be uploaded to the website to recreate the data set, and
-- `NLSY79.R`: R script for reading the data into R and converting the variable names and label into something more sensible.
+- `NLSY79.R`: R script for reading the data into R and converting the variables' names and label into something more sensible.
 
@@
 The original set, containing wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994, is available in the R package `brolgar`.
-To compare the refreshed data with the original, we need to subset the data into a matching cohort of high school dropouts.
-There are multiple ways to match a comparative cohort from the updated data -- the simplest approach is to subset to individuals with matching IDs in the original data and restricting the longitudinal measurements to the same years; another approach is to replicate the process as suggested by the description of the original data thus identifying a new (and possibly removal of old) set of individuals that is included in the refreshed data.
-We use the latter approach and as the first step we identify individuals who dropped out of high school.
+To compare the refreshed data with the original, a subset needs to be matched.
+There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data and restricting the longitudinal measurements to the same years.
+However, we decided to try to replicate the process, as suggested by the description of the original data.
+This requires first identifying individuals who dropped out of high school.
 
@@
 1. An individual whose highest grade completed (`hgc`) is reported to be less than 12th grade, **or**
-2. An individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and has received a GED (i.e. `ged` code is 2).
+2. An individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and has received a GED (`ged` is code to 2).
 
 An additional criterion from @SingerJudithD2003Alda is to only include males aged between 14 and 17 years old in 1979.
-With this filtering, we obtained `r ori_do_nrow` individuals in the refreshed data compared to `r sw_do_nrow` individuals in the original data.
+With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
 To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
@@
 1.
-`r nrow(join_more_17)` individuals were more than 17 years old in 1979.
+173 individuals were more than 17 years old in 1979.
 Thus, it looks like the description of the original data is not quite accurate that there are people older than 17 in the subset.
 Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
-2. `r nrow(join_less_17)` individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
-    i.
-`r nrow(join_not_elig)` of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded as 1).
-This suggests they are not dropouts, and so we excluded them from the refreshed data.
-    ii. The information about `ged` is missing in `r nrow(join_missing_ged)` individuals. We decided to include them in the refreshed data. 
-    iii. `r nrow(join_ged_both)` individuals have both diploma and GED (`ged` is coded as 3). These were kept in the refreshed data. 
-    iv. `r nrow(join_not_exist)` individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.
+2. 79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
+    i. 35 of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded to 1). This suggests they are not dropouts, and so we excluded them from the refreshed data. 
+    ii. The information about `ged` is missing in 38 individuals. We decided to include them in the refreshed data. 
+    iii. 3 individuals have both diploma and GED (`ged` is coded to 3). These were kept in the refreshed data. 
+    iv. 12 individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.
 
-The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing `r n_do` individuals.
+The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing 863 individuals.
 
@@
 Figure \ref{fig:compare-subsets} contains summaries of corresponding variables in the two subsets.
-Plot (A) shows a back-to-back bar chart of the highest grade completed.
+Plot (A) shows a back-back bar chart of the highest grade completed.
 The two sets are almost the same, but small differences remain.
@@
 This is better to do with each wave of new data added so that it is relative to the last date in the data.
-Our decision was to provide the raw wages and include code to make the adjustment as part of the package.
+Our decision was to provide the raw wages and include codes to make the adjustment as part of the package.
 
@@
 The wages data is an example of this.
-Students can take away multiple interpretations from the data (e.g.
-wages increase with experience, education, and race) based on the teaching focus to learn about the world.
-If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., sex and race), or data that has been poorly processed containing errors, we, as educators, are being irresponsible because the societal message taught to students may be flawed.
+Students will take away the interpretations from whatever is taught with the data, that wages increase with experience, education, race.
+If one uses data examples that are synthetic, instilled with our own inherent prejudices (e.g., sex and race), or data that has been poorly processed containing errors, we as educators are being irresponsible because the societal message taught to students may be flawed.
 This paper demonstrates the process of producing a trustworthy data set for teaching.
@@
 
-- **Codes**: R script to reproduce data tidying and cleaning is available at \url{`r yowie_pkgdown`/articles/process-data.html}.
+- **Code**: R script to reproduce data tidying and cleaning is available at \url{`r yowie_pkgdown`/articles/process-data.html}.
 The code for the extreme value handling with robust linear fits is in \url{`r yowie_pkgdown`/articles/input-to-valid-data.html}.

### 2022-05-12 paper/paper.Rmd
typo

@@
 Numerous samples of individuals were drawn for each threshold, and the smoothness of the profiles was examined.
-Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
+Overly smooth profiles would indicate the replacement of values was too severe, resulting in removing the very interesting volatility of wages seen in many individuals.
 This was achieved using a purpose written `shiny`  [@shiny] app (provided with the code of this paper).

### 2022-05-12 paper/paper.Rmd
minor typos

@@
 -->
-The original set, containing wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994, is available in the R package `brolgar`.
+The original set, containing wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994, is available in the R package `brolgar` [@brolgar].
 To compare the refreshed data with the original, a subset needs to be matched.
@@
 More education leads to increasingly higher wages, which is a satisfying result for educators.
-It means education makes a difference in the wage experience and that public policy that encourages education is a data-supported action.
+It means education makes a difference in the wage experience, so public policy encouraging education is a data-supported action.
 We would also learn, although not presented here, that wages on average differ by race category.
@@
 
-- **R Package**:`r yowie` is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort.
+- **R Package**: `r yowie` is an R data package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort.
 This package can be accessed from \url{`r yowie_repo`}.

### 2022-05-12 paper/paper-jsdse/paper-jsdse.tex

--- 2022-05-12 paper/paper-jsdse/paper-jsdse.tex
@@
 \begin{verbatim}
-#> 'data.frame':    12686 obs. of  742 variables:
-#>  $ CASEID_1979                               : int  1 2 3 4 5 6 7 8 ...
-#>  $ HRP1_1979                                 : int  328 385 365 NA 310 NA NA NA ...
-#>  $ HRP2_1979                                 : int  NA NA NA NA 375 NA NA NA ...
-#>  $ HRP3_1979                                 : int  NA NA 275 NA NA NA NA NA ...
-#>  $ HRP4_1979                                 : int  NA NA NA NA NA 250 NA NA ...
-#>  $ HRP5_1979                                 : int  NA NA NA NA NA NA NA NA ...
-#>  $ HRP1_1980                                 : int  NA 457 397 NA 333 275 300 394 ...
-#>  $ HRP2_1980                                 : int  NA NA 367 NA NA NA NA NA ...
-#>  $ HRP3_1980                                 : int  NA NA 380 NA NA NA 290 NA ...
-#>  $ HRP4_1980                                 : int  NA NA NA NA NA NA NA NA ...
-#>   [list output truncated]
+#> Rows: 12,686
+#> Columns: 11
+#> $ CASEID_1979 <int> 1, 2, 3, 4, 5, 6, 7, 8, 9, 10, 11, 12, 13, 14, 1~
+#> $ HRP1_1979   <int> 328, 385, 365, NA, 310, NA, NA, NA, 214, NA, 337~
+#> $ HRP2_1979   <int> NA, NA, NA, NA, 375, NA, NA, NA, NA, NA, 300, NA~
+#> $ HRP3_1979   <int> NA, NA, 275, NA, NA, NA, NA, NA, NA, NA, NA, NA,~
+#> $ HRP4_1979   <int> NA, NA, NA, NA, NA, 250, NA, NA, NA, NA, NA, NA,~
+#> $ HRP5_1979   <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, ~
+#> $ HRP1_1980   <int> NA, 457, 397, NA, 333, 275, 300, 394, 200, 318, ~
+#> $ HRP2_1980   <int> NA, NA, 367, NA, NA, NA, NA, NA, NA, NA, NA, NA,~
+#> $ HRP3_1980   <int> NA, NA, 380, NA, NA, NA, 290, NA, NA, NA, NA, NA~
+#> $ HRP4_1980   <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, ~
+#> $ HRP5_1980   <int> NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, NA, ~
 \end{verbatim}
@@
 An additional criterion from Singer and Willett (2003) is to only include males aged between 14 and 17 years old in 1979.
-With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
+With this filtering, we obtained 0 individuals in the refreshed data compared to 888 individuals in the original data.
 To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
@@
 \item
-  79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:
+  715 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:
 
@@
   \item
-    The information about \texttt{ged} is missing in 38 individuals. We decided to include them in the refreshed data.
+    The information about \texttt{ged} is missing in 341 individuals. We decided to include them in the refreshed data.
   \item
@@
 
-The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing 863 individuals.
+The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing 865 individuals.
 
@@
 
-Lastly, the race variable is almost perfectly matched. The \texttt{id}'s in the dropouts subset of the refreshed data all correspond to males, which perfectly matches the original subset.
+Lastly, the matching of individuals in the dropouts subset of the refreshed data with the original data was done using their \texttt{id}s.
+(Only males were in the original data.) It was comforting to see that all of the individuals in both sets do match on sex (all are male!), and race.
 

### 2022-05-12 paper/paper.Rmd

@@
 
-\begin{figure}[t]
+\begin{figure*}[p]
 
-\begin{tcolorbox}[title = Navigating the data source]
-\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\
-\vspace{1mm}
+\begin{tcolorbox}[title = Navigating the data source, fontupper=\linespread{.8}\selectfont]
+\vspace{3mm}
+\faDatabase\ NLSY79 (\url{https://www.nlsinfo.org/investigator/pages/search?s=NLSY79})\\\\
+\vspace{3mm}
 \faCheck\ The CASEID will be always be selected.  \\
 \vspace{1mm}
-\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  \\
-\vspace{1mm}
-For the remaining variables, we went to the "Variable Search" tab and select variables as follows
+\faCheck\ The 3 recommended demographic variable (sample ID, race and sex) were selected.  For the remaining variables, we went to the ``Variable Search" tab and select variables as follows
 \begin{itemize}
@@
 -->
+
 The original set, containing wages of high school dropouts [@SingerJudithD2003Alda] from 1979 through to 1994, is available in the R package `brolgar` [@brolgar].
@@
 1. An individual whose highest grade completed (`hgc`) is reported to be less than 12th grade, **or**
-2. An individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and has received a GED (`ged` is code to 2).
+2. An individual whose highest grade completed (`hgc`) is reported to be at least 12th grade and has received a GED (i.e. `ged` code is 2).
 
 An additional criterion from @SingerJudithD2003Alda is to only include males aged between 14 and 17 years old in 1979.
-With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
+With this filtering, we obtained `r ori_do_nrow` individuals in the refreshed data compared to `r sw_do_nrow` individuals in the original data.
 To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
@@
 1.
-173 individuals were more than 17 years old in 1979.
+`r nrow(join_more_17)` individuals were more than 17 years old in 1979.
 Thus, it looks like the description of the original data is not quite accurate that there are people older than 17 in the subset.
 Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
-2. 79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
-    i. 35 of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded to 1). This suggests they are not dropouts, and so we excluded them from the refreshed data. 
-    ii. The information about `ged` is missing in 38 individuals. We decided to include them in the refreshed data. 
-    iii. 3 individuals have both diploma and GED (`ged` is coded to 3). These were kept in the refreshed data. 
-    iv. 12 individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.
+2. `r nrow(join_less_17)` individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because: 
+    i.
+`r nrow(join_not_elig)` of them completed at least 12th grade with a diploma instead of GED (`ged` variable is coded as 1).
+This suggests they are not dropouts, and so we excluded them from the refreshed data.
+    ii. The information about `ged` is missing in `r nrow(join_missing_ged)` individuals. We decided to include them in the refreshed data. 
+    iii. `r nrow(join_ged_both)` individuals have both diploma and GED (`ged` is coded as 3). These were kept in the refreshed data. 
+    iv. `r nrow(join_not_exist)` individuals do not exist in `wages` data because they have participated in less than 3 rounds of the survey.
 
-The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing 863 individuals.
+The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing `r n_do` individuals.
 

### 2022-05-12 paper/paper-jsdse/paper-jsdse.tex
removed DONE from response

@@
 Numerous samples of individuals were drawn for each threshold, and the smoothness of the profiles was examined.
-Overly smooth profiles would indicate the replacement of values was too severe, resulting in the removing the very interesting volatility of wages seen in many individuals.
+Overly smooth profiles would indicate the replacement of values was too severe, resulting in removing the very interesting volatility of wages seen in many individuals.
 This was achieved using a purpose written \texttt{shiny} (Chang et al.
@@
 
-The original set, containing wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994, is available in the R package \texttt{brolgar}.
-To compare the refreshed data with the original, we need to subset the data into a matching cohort of high school dropouts.
-There are multiple ways to match a comparative cohort from the updated data -- the simplest approach is to subset to individuals with matching IDs in the original data and restricting the longitudinal measurements to the same years; another approach is to replicate the process as suggested by the description of the original data thus identifying a new (and possibly removal of old) set of individuals that is included in the refreshed data.
-We use the latter approach and as the first step we identify individuals who dropped out of high school.
+The original set, containing wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994, is available in the R package \texttt{brolgar} (Tierney, Cook, and Prvan 2020).
+To compare the refreshed data with the original, a subset needs to be matched.
+There are numerous ways to do this, with the simplest being to extract the individuals based on their id being part of the original data and restricting the longitudinal measurements to the same years.
+However, we decided to try to replicate the process, as suggested by the description of the original data.
+This requires first identifying individuals who dropped out of high school.
 
@@
 \item
-  \textbf{Codes}: R script to reproduce data tidying and cleaning is available at \url{https://numbats.github.io/yowie/articles/process-data.html}.
+  \textbf{Code}: R script to reproduce data tidying and cleaning is available at \url{https://numbats.github.io/yowie/articles/process-data.html}.
 The code for the extreme value handling with robust linear fits is in \url{https://numbats.github.io/yowie/articles/input-to-valid-data.html}.
 \item
-  \textbf{R Package}:\texttt{yowie} is a data container R package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort.
+  \textbf{R Package}: \texttt{yowie} is an R data package that contains 3 datasets, namely the high school mean hourly wage data, high school dropouts mean hourly wage data, and demographic data of the NLSY79 cohort.
 This package can be accessed from \url{https://github.com/numbats/yowie}.

### 2022-05-12 response/response2.Rmd
removed DONE from response

@@
 There is some wording in the introduction ("An example of this is the wages data made public by Singer and Willett (2003) in their book, “Applied lon- gitudinal data analysis,” which can be used to teach generalized linear models, in addition to hierarchical, mixed effects, and multilevel models.") and we have added additional text in Section 5 Summary, on lessons learned, to help the reader understand the elements that might be used for their teaching.
-XXX DONE
 
@@
 
-* We have revised the language, made a note for educators on usage, and referred to the SAGER guidelines. XXX DONE
+* We have revised the language, made a note for educators on usage, and referred to the SAGER guidelines.
 
@@
 
-* We have added more explanation of the two variables, and suggestions of where to use one and the other in classroom analyses. XXX DONE
+* We have added more explanation of the two variables, and suggestions of where to use one and the other in classroom analyses.
 
@@
 This section has been modified so that the approach is more understandable.
-XXX DONE
 
@@
 It could definitely be worked into a teaching exercise.
-XXX DONE
 
@@
 
-* We have revised the language, referenced the Federal standard, and explained the reasoning for using a single variable.  XXX DONE
+* We have revised the language, referenced the Federal standard, and explained the reasoning for using a single variable.
 
@@
 
-* We have revised the introduction to fix this. XXX DONE
+* We have revised the introduction to fix this.
 
@@
 We think it is best to keep content on cleaning and tidying the data, in a reproducible manner, to match the mission of the special issue.
-XXX DONE
 
@@
 
-* We have changed all to "refresh" as it is not correct if we say it as a replication because we did not use the exactly same variable and same result (esepcially ID included in the dropouts cohort.) XXX DONE
+* We have changed all to "refresh" as it is not correct if we say it as a replication because we did not use the exactly same variable and same result (esepcially ID included in the dropouts cohort.)
 
@@
 
-* See response above to the associate editor, about this concern. XXX DONE
+* See response above to the associate editor, about this concern.
 
@@
 We hope that our modification makes the manuscript clearer.
-XXX DONE
 
@@
 We have modified the tables' caption to make this clearer.
-XXX DONE
 
@@
 We agree that this sentences is not clear, so we have modified it.
-XXX DONE
 
@@
 
-* We have put the capitalisation back in. Besides, we did not use the term "White" in the dataset as we used the same term reported in the database, which is "Non-Black, Non-Hispanic". XXX Done
+* We have put the capitalisation back in. Besides, we did not use the term "White" in the dataset as we used the same term reported in the database, which is "Non-Black, Non-Hispanic".
 

### 2022-05-12 paper/paper-jsdse/paper-jsdse.tex
nicks comment on year
Incorporate Nick's suggestions.

--- 2022-05-12 paper/paper-jsdse/paper-jsdse.tex
@@
 \item
-  EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
+  EXPER: stands for experience, the temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
 \item
@@
 
-Our final target data will include the demographic variables (variable names): sex (\texttt{sex}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
+Our final target data will include the demographic variables (variable names): sex (\texttt{sex}), race (\texttt{race}), age (\texttt{age\_1979}), highest grade completed reported in each round of the survey (\texttt{grade}), highest grade completed ever reported (\texttt{hgc}), highest grade completed in terms of years (as an integers), e.g., 9th grade = 9, 3rd-year college = 15, (\texttt{hgc\_i}), highest grade completed in 1979 (\texttt{hgc\_1979}) and whether the graduate equivalency diploma is obtained (\texttt{ged}).
 
@@
 To obtain the work experience since 1979, we calculate the cumulative value.
-As the measurement unit is weeks, we convert this to year.
+As the measurement unit is weeks, we convert this to fractional year.
 
@@
 
-Lastly, the individuals in the dropouts subset of the refreshed data (as matched by their \texttt{id}s) belong to the same race as the individuals in the original data.
-This is also the case in the \texttt{sex} variables, which all correspond to males.
+Lastly, the matching of individuals in the dropouts subset of the refreshed data with the original data was done using their \texttt{id}s.
+(Only males were in the original data.) It was comforting to see that all of the individuals in both sets do match on sex (all are male!), and race.
 

### 2022-05-12 paper/paper.Rmd
nicks comment on year
Incorporate Nick's suggestions.

@@
 1. ID: the respondents' ID. 
-2. EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
+2. EXPER: stands for experience, the temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
 3. LNW: natural logarithm of wages, adjusted with 1990's inflation rate. 
@@
 
-Our final target data will include the demographic variables (variable names): sex (`sex`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).
+Our final target data will include the demographic variables (variable names): sex (`sex`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years (as an integers), e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).
 

### 2022-05-13 paper/paper-monash/paper-monash.tex
monash preprint also

--- 2022-05-12 paper/paper.tex
@@
-\documentclass{article}
-
-\usepackage{arxiv}
-
-\usepackage[utf8]{inputenc} % allow utf-8 input
-\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
-\usepackage{lmodern}        % https://github.com/rstudio/rticles/issues/343
-\usepackage{hyperref}       % hyperlinks
-\usepackage{url}            % simple URL typesetting
-\usepackage{booktabs}       % professional-quality tables
-\usepackage{amsfonts}       % blackboard math symbols
-\usepackage{nicefrac}       % compact symbols for 1/2, etc.
-\usepackage{microtype}      % microtypography
-\usepackage{lipsum}
-\usepackage{graphicx}
+\documentclass[11pt,a4paper,]{article}
+\usepackage{lmodern}
+
+\usepackage{amssymb,amsmath}
+\usepackage{ifxetex,ifluatex}
+\usepackage{fixltx2e} % provides \textsubscript
+\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
+  \usepackage[T1]{fontenc}
+  \usepackage[utf8]{inputenc}
+\else % if luatex or xelatex
+  \usepackage{unicode-math}
+  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
+\fi
+% use upquote if available, for straight quotes in verbatim environments
+\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
+% use microtype if available
+\IfFileExists{microtype.sty}{%
+\usepackage[]{microtype}
+\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
+}{}
+\PassOptionsToPackage{hyphens}{url} % url is loaded by hyperref
+\usepackage[unicode=true]{hyperref}
+\hypersetup{
+            pdftitle={A Journey from Wild to Textbook Data to Reproducibly Refresh the Wages Data from the National Longitudinal Survey of Youth Database},
+            pdfkeywords={Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;},
+            pdfborder={0 0 0},
+            breaklinks=true}
+\urlstyle{same}  % don't use monospace font for urls
+\usepackage{geometry}
+\geometry{left=2.5cm,right=2.5cm,top=2.5cm,bottom=2.5cm}
+\usepackage[style=authoryear-comp,]{biblatex}
+\addbibresource{../references.bib}
+\usepackage{longtable,booktabs}
+% Fix footnotes in tables (requires footnote package)
+\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{long table}}{}
+\IfFileExists{parskip.sty}{%
+\usepackage{parskip}
+}{% else
+\setlength{\parindent}{0pt}
+\setlength{\parskip}{6pt plus 2pt minus 1pt}
+}
+\setlength{\emergencystretch}{3em}  % prevent overfull lines
+\providecommand{\tightlist}{%
+  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
+\setcounter{secnumdepth}{5}
+
+% set default figure placement to htbp
+\makeatletter
+\def\fps@figure{htbp}
+\makeatother
 
@@
 
-\author{
-    Dewi Amaliah
-   \\
-    Department of Econometrics and Business Statistics \\
-    Monash University \\
-  Clayton, VIC 3800 \\
-  \texttt{\href{mailto:dlamaleeah@gmail.com}{\nolinkurl{dlamaleeah@gmail.com}}} \\
-   \And
-    Dianne Cook
-   \\
-    Department of Econometrics and Business Statistics \\
-    Monash University \\
-  Clayton, VIC 3800 \\
-  \texttt{\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}} \\
-   \And
-    Emi Tanaka
-   \\
-    Department of Econometrics and Business Statistics \\
-    Monash University \\
-  Clayton, VIC 3800 \\
-  \texttt{\href{mailto:emi.tanaka@monash.edu}{\nolinkurl{emi.tanaka@monash.edu}}} \\
-   \And
-    Kate Hyde
-   \\
-    Department of Econometrics and Business Statistics \\
-    Monash University \\
-  Clayton, VIC 3800 \\
-  \texttt{\href{mailto:hyde.kate.a@gmail.com}{\nolinkurl{hyde.kate.a@gmail.com}}} \\
-   \And
-    Nicholas Tierney
-   \\
-    Telethon Kids Institute \\
-  Nedlands, WA 6009 \\
-  \texttt{\href{mailto:nicholas.tierney@gmail.com}{\nolinkurl{nicholas.tierney@gmail.com}}} \\
-  }
-
-% Pandoc citation processing
-
+%% MONASH STUFF
+
+%% CAPTIONS
+\RequirePackage{caption}
+\DeclareCaptionStyle{italic}[justification=centering]
+ {labelfont={bf},textfont={it},labelsep=colon}
+\captionsetup[figure]{style=italic,format=hang,singlelinecheck=true}
+\captionsetup[table]{style=italic,format=hang,singlelinecheck=true}
+
+%% FONT
+\RequirePackage{bera}
+\RequirePackage{mathpazo}
+
+%% HEADERS AND FOOTERS
+\RequirePackage{fancyhdr}
+\pagestyle{fancy}
+\rfoot{\Large\sffamily\raisebox{-0.1cm}{\textbf{\thepage}}}
+\makeatletter
+\lhead{\textsf{\expandafter{\@title}}}
+\makeatother
+\rhead{}
+\cfoot{}
+\setlength{\headheight}{15pt}
+\renewcommand{\headrulewidth}{0.4pt}
+\renewcommand{\footrulewidth}{0.4pt}
+\fancypagestyle{plain}{%
+\fancyhf{} % clear all header and footer fields
+\fancyfoot[C]{\sffamily\thepage} % except the center
+\renewcommand{\headrulewidth}{0pt}
+\renewcommand{\footrulewidth}{0pt}}
+
+%% MATHS
+\RequirePackage{bm,amsmath}
+\allowdisplaybreaks
+
+%% GRAPHICS
+\RequirePackage{graphicx}
+\setcounter{topnumber}{2}
+\setcounter{bottomnumber}{2}
+\setcounter{totalnumber}{4}
+\renewcommand{\topfraction}{0.85}
+\renewcommand{\bottomfraction}{0.85}
+\renewcommand{\textfraction}{0.15}
+\renewcommand{\floatpagefraction}{0.8}
+
+%\RequirePackage[section]{placeins}
+
+%% SECTION TITLES
+\RequirePackage[compact,sf,bf]{titlesec}
+\titleformat{\section}[block]
+  {\fontsize{15}{17}\bfseries\sffamily}
+  {\thesection}
+  {0.4em}{}
+\titleformat{\subsection}[block]
+  {\fontsize{12}{14}\bfseries\sffamily}
+  {\thesubsection}
+  {0.4em}{}
+\titlespacing{\section}{0pt}{*5}{*1}
+\titlespacing{\subsection}{0pt}{*2}{*0.2}
+
+%% TITLE PAGE
+\def\Date{\number\day}
+\def\Month{\ifcase\month\or
+ January\or February\or March\or April\or May\or June\or
+ July\or August\or September\or October\or November\or December\fi}
+\def\Year{\number\year}
+
+\makeatletter
+\def\wp#1{\gdef\@wp{#1}}\def\@wp{??/??}
+\def\jel#1{\gdef\@jel{#1}}\def\@jel{??}
+\def\showjel{{\large\textsf{\textbf{JEL classification:}}~\@jel}}
+\def\nojel{\def\showjel{}}
+\def\addresses#1{\gdef\@addresses{#1}}\def\@addresses{??}
+\def\cover{{\sffamily\setcounter{page}{0}
+        \thispagestyle{empty}
+        \placefig{2}{1.5}{width=5cm}{monash2}
+        \placefig{16.9}{1.5}{width=2.1cm}{MBusSchool}
+        \begin{textblock}{4}(16.9,4)ISSN 1440-771X\end{textblock}
+        \begin{textblock}{7}(12.7,27.9)\hfill
+        \includegraphics[height=0.7cm]{AACSB}~~~
+        \includegraphics[height=0.7cm]{EQUIS}~~~
+        \includegraphics[height=0.7cm]{AMBA}
+        \end{textblock}
+        \vspace*{2cm}
+        \begin{center}\Large
+        Department of Econometrics and Business Statistics\\[.5cm]
+        \footnotesize http://monash.edu/business/ebs/research/publications
+        \end{center}\vspace{2cm}
+        \begin{center}
+        \fbox{\parbox{14cm}{\begin{onehalfspace}\centering\Huge\vspace*{0.3cm}
+                \textsf{\textbf{\expandafter{\@title}}}\vspace{1cm}\par
+                \LARGE\@author\end{onehalfspace}
+        }}
+        \end{center}
+        \vfill
+                \begin{center}\Large
+                \Month~\Year\\[1cm]
+                Working Paper \@wp
+        \end{center}\vspace*{2cm}}}
+\def\pageone{{\sffamily\setstretch{1}%
+        \thispagestyle{empty}%
+        \vbox to \textheight{%
+        \raggedright\baselineskip=1.2cm
+     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}
+        \vspace{2cm}\par
+        \hspace{1cm}\parbox{14cm}{\sffamily\large\@addresses}\vspace{1cm}\vfill
+        \hspace{1cm}{\large\Date~\Month~\Year}\\[1cm]
+        \hspace{1cm}\showjel\vss}}}
+\def\blindtitle{{\sffamily
+     \thispagestyle{plain}\raggedright\baselineskip=1.2cm
+     {\fontsize{24.88}{30}\sffamily\textbf{\expandafter{\@title}}}\vspace{1cm}\par
+        }}
+\def\titlepage{{\cover\newpage\pageone\newpage\blindtitle}}
+
+\def\blind{\def\titlepage{{\blindtitle}}\let\maketitle\blindtitle}
+\def\titlepageonly{\def\titlepage{{\pageone\end{document}}}}
+\def\nocover{\def\titlepage{{\pageone\newpage\blindtitle}}\let\maketitle\titlepage}
+\let\maketitle\titlepage
+\makeatother
+
+%% SPACING
+\RequirePackage{setspace}
+\spacing{1.5}
+
+%% LINE AND PAGE BREAKING
+\sloppy
+\clubpenalty = 10000
+\widowpenalty = 10000
+\brokenpenalty = 10000
+\RequirePackage{microtype}
+
+%% PARAGRAPH BREAKS
+\setlength{\parskip}{1.4ex}
+\setlength{\parindent}{0em}
+
+%% HYPERLINKS
+\RequirePackage{xcolor} % Needed for links
+\definecolor{darkblue}{rgb}{0,0,.6}
+\RequirePackage{url}
+
+\makeatletter
+\@ifpackageloaded{hyperref}{}{\RequirePackage{hyperref}}
+\makeatother
+\hypersetup{
+     citecolor=0 0 0,
+     breaklinks=true,
+     bookmarksopen=true,
+     bookmarksnumbered=true,
+     linkcolor=darkblue,
+     urlcolor=blue,
+     citecolor=darkblue,
+     colorlinks=true}
+
+%% KEYWORDS
+\newenvironment{keywords}{\par\vspace{0.5cm}\noindent{\sffamily\textbf{Keywords:}}}{\vspace{0.25cm}\par\hrule\vspace{0.5cm}\par}
+
+%% ABSTRACT
+\renewenvironment{abstract}{\begin{minipage}{\textwidth}\parskip=1.4ex\noindent
+\hrule\vspace{0.1cm}\par{\sffamily\textbf{\abstractname}}\newline}
+  {\end{minipage}}
+
+\usepackage[T1]{fontenc}
+\usepackage[utf8]{inputenc}
+
+\usepackage[showonlyrefs]{mathtools}
+\usepackage[no-weekday]{eukdate}
+
+%% BIBLIOGRAPHY
+
+\makeatletter
+\@ifpackageloaded{biblatex}{}{\usepackage[style=authoryear-comp, backend=biber, natbib=true]{biblatex}}
+\makeatother
+\ExecuteBibliographyOptions{bibencoding=utf8,minnames=1,maxnames=3, maxbibnames=99,dashed=false,terseinits=true,giveninits=true,uniquename=false,uniquelist=false,doi=false, isbn=false,url=true,sortcites=false, date=year}
+
+\DeclareFieldFormat{url}{\texttt{\url{#1}}}
+\DeclareFieldFormat[article]{pages}{#1}
+\DeclareFieldFormat[inproceedings]{pages}{\lowercase{pp.}#1}
+\DeclareFieldFormat[incollection]{pages}{\lowercase{pp.}#1}
+\DeclareFieldFormat[article]{volume}{\mkbibbold{#1}}
+\DeclareFieldFormat[article]{number}{\mkbibparens{#1}}
+\DeclareFieldFormat[article]{title}{\MakeCapital{#1}}
+\DeclareFieldFormat[inproceedings]{title}{#1}
+\DeclareFieldFormat{shorthandwidth}{#1}
+% No dot before number of articles
+\usepackage{xpatch}
+\xpatchbibmacro{volume+number+eid}{\setunit*{\adddot}}{}{}{}
+% Remove In: for an article.
+\renewbibmacro{in:}{%
+  \ifentrytype{article}{}{%
+  \printtext{\bibstring{in}\intitlepunct}}}
+
+\makeatletter
+\DeclareDelimFormat[cbx@textcite]{nameyeardelim}{\addspace}
+\makeatother
+\renewcommand*{\finalnamedelim}{%
+  %\ifnumgreater{\value{liststop}}{2}{\finalandcomma}{}% there really should be no funny Oxford comma business here
+  \addspace\&\space}
+
+\wp{no/yr}
+\jel{C10,C14,C22}
+
+\RequirePackage[absolute,overlay]{textpos}
+\setlength{\TPHorizModule}{1cm}
+\setlength{\TPVertModule}{1cm}
+\def\placefig#1#2#3#4{\begin{textblock}{.1}(#1,#2)\rlap{\includegraphics[#3]{#4}}\end{textblock}}
+
+\author{}
+\addresses{}
+
+\date{\sf\Date~\Month~\Year}
+\makeatletter
+ \lfoot{\sf : \@date}
+\makeatother
+
+%% Any special functions or other packages can be loaded here.
 \usepackage{tcolorbox}
@@
 
+% Adjust headwidth in case user has changed geometry in header-includes
+\renewcommand{\headwidth}{\textwidth}
+
 \begin{document}
 \maketitle
-
-\def\tightlist{}
-
 \begin{abstract}
@@
 \end{abstract}
-
-\keywords{
+\begin{keywords}
 Data cleaning; Data tidying; Reproducible workflow; Longitudinal data; NLSY79; Initial data analysis;
-  }
+\end{keywords}
 
@@
 Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
-An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis,'' which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models.
-The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
-Department of Labor 2021a).
+An example of this is the wages data made public by \textcite{SingerJudithD2003Alda} in their book, ``Applied longitudinal data analysis,'' which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models.
+The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) \autocite{nlsy79}.
 
@@
 Some individuals experience a decline in wages the longer they are in the workforce, and many experience volatility in their wages.
-It is for these reasons that the wages data was used to illustrate exploratory longitudinal data analysis in Ilk (2004), and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
+It is for these reasons that the wages data was used to illustrate exploratory longitudinal data analysis in \textcite{ilk2004}, and was further developed into a case study for use in the teaching of exploratory data analysis at Iowa State University.
 
@@
 Textbook data sets have longevity if they have unresolved elements.
-The iris data (Anderson 1935) is a prime example.
-It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis\footnote{It should be noted that the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)}.
+The iris data \autocite{iris-data} is a prime example.
+It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis\footnote{It should be noted that the iris data is best replaced today with the penguins data \autocite{penguins-data}, which has similar qualities, is new, and does not suffer from a connection with eugenics \autocite{notiris}}.
 We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
@@
 We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation.
-We occasionally hear this voiced in the public media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian (Moncrief 2015).
+We occasionally hear this voiced in the public media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian \autocite{notaverage}.
 Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.
@@
 As a textbook data set, though, the wages data is outdated.
-The most recent year in the data is 1994, 9 years prior to when Singer and Willett (2003) was published.
+The most recent year in the data is 1994, 9 years prior to when \textcite{SingerJudithD2003Alda} was published.
 Teachers of statistics need contemporary data sets to show how techniques are relevant for today's students.
 Using tired old textbook data sets can imbue a misconception that the field is not current.
-The wages data is extracted from NSLY79, one of the best examples of open data (see details at Open Knowledge Foundation 2021), which is constantly being updated.
+The wages data is extracted from NSLY79, one of the best examples of open data \autocite[see details at][]{opendata}, which is constantly being updated.
 It should be possible to continuously refresh the textbook data from the data repository.
@@
 
-This paper demonstrates the steps of cleaning data, including subjective decisions made on dealing with anomalies, and documents the process, as recommended by Huebner, Vach, and Cessie (2016).
+This paper demonstrates the steps of cleaning data, including subjective decisions made on dealing with anomalies, and documents the process, as recommended by \textcite{HuebnerMariannePhD2016Asat}.
 They emphasize that making the data cleaning process accountable and transparent is imperative and essential for the integrity of downstream statistical analyses and model building.
-Clean data often then goes through an ``initial data analysis'' (IDA) (Chatfield 1985), where one would summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling.
-This stage is related to exploratory data analysis (EDA), coined by Tukey (1977) with a focus on learning from data.
+Clean data often then goes through an ``initial data analysis'' (IDA) \autocite{Chatfield1985TIEo}, where one would summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling.
+This stage is related to exploratory data analysis (EDA), coined by \textcite{tukey} with a focus on learning from data.
 EDA can be considered to encompass IDA.
 In practice, the three stages of cleaning, summarizing, and exploring are cyclical, one often needs to do more cleaning after scrutinizing.
-Dasu and Johnson (2003) say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.
-
-Our approach to cleaning builds heavily on the \texttt{tidyverse} approach (Wickham, Averick, et al.
-2019).
-The data is first organized into ``tidy data'' (Wickham 2014) and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables (Wickham 2011).
-Tidy data shouldn't be confused with ``tame data,'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
+\textcite{DasuTamraparni2003Edma} say that data cleaning and exploration is a difficult task and typically consumes a large percentage of the time spent in analyzing data.
+
+Our approach to cleaning builds heavily on the \texttt{tidyverse} approach \autocite{tidyverse}.
+The data is first organized into ``tidy data'' \autocite{WickhamHadley2014TD} and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables \autocite{plyr}.
+Tidy data shouldn't be confused with ``tame data,'' which \textcite{tamedata} coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
 The resulting (tame) data is provided in a new R package called \texttt{yowie}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
@@
 
-Singer and Willett (2003) used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
-This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century (Pergamit et al.
-2001).
-In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research (Pergamit et al.
-2001; Cooksey 2017).
-According to Cooksey (2017), thousands of articles and hundreds of book chapters and monographs have utilized this data.
-Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data (Pergamit et al.
-2001).
-
-Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose consistent with Grimshaw (2015)'s statistics education goal of embracing authentic data experiences.
-Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by Singer and Willett (2003).
-We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the \emph{purposes of studying racism, not race} (Fullilove 1998).
-There are a number of data sets provided by Singer and Willett (2003), but we focus only on wages data because it has captivated our attention for a number of years in our own teaching of longitudinal data analysis.
+\textcite{SingerJudithD2003Alda} used the wages and other variables of high school dropouts from the NLSY79 data as an example data set to illustrate longitudinal data modeling of wages on workforce experience, with covariates education and race.
+This data has been playing an important role in research in various disciplines, including but not limited to economics, sociology, education, public policy, and public health for more than a quarter of the century \autocite{MichaelRPergamit2001DWTN}.
+In addition, this is considered a carefully designed longitudinal survey with high retention rates, making it suitable for life course research \autocite{MichaelRPergamit2001DWTN,eliznlsy}.
+According to \textcite{eliznlsy}, thousands of articles and hundreds of book chapters and monographs have utilized this data.
+Moreover, the NLSY79 is considered the most widely used and most important cohort in the survey data \autocite{MichaelRPergamit2001DWTN}.
+
+Our aim is to refresh the wages textbook data and append it with data from 1994 through to the latest data reported in 2018, a purpose consistent with \textcite{grimshaw}'s statistics education goal of embracing authentic data experiences.
+Here, we investigate the process of getting from the raw NLSY79 data to a textbook data set as similar as possible to that provided by \textcite{SingerJudithD2003Alda}.
+We should also note that race is a variable in the original data set, and for compatibility, it is also provided with the refreshed data for the \emph{purposes of studying racism, not race} \autocite{racismnotrace}.
+There are a number of data sets provided by \textcite{SingerJudithD2003Alda}, but we focus only on wages data because it has captivated our attention for a number of years in our own teaching of longitudinal data analysis.
 
@@
 
-The NLSY79 is a longitudinal survey administered by the U.S Bureau of Labor Statistics that follows the lives of a sample of American youth born between 1957-1964 (Bureau of Labor Statistics, U.S.
-Department of Labor 2021a).
+The NLSY79 is a longitudinal survey administered by the U.S Bureau of Labor Statistics that follows the lives of a sample of American youth born between 1957-1964 \autocite{nlsy79}.
 The cohort originally included 12,686 respondents aged 14-22 when first interviewed in 1979.
@@
 There are two ways to conduct the interview of the NLSY79, which are face-to-face or telephone interviews.
-In recent survey years, more than 90 percent of respondents were interviewed by telephone (Cooksey 2017).
+In recent survey years, more than 90 percent of respondents were interviewed by telephone \autocite{eliznlsy}.
 
@@
 
-The NLSY79 data used in Singer and Willett (2003) contains the longitudinal records of male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994.
+The NLSY79 data used in \textcite{SingerJudithD2003Alda} contains the longitudinal records of male high school dropouts who first participated in the study at age 14-17 years from 1979 through to 1994.
 This dataset contains several variables as follows:
@@
 \item
-  The subset of wages data in (1), which is the wages of high school dropouts' as a refreshed version of Singer and Willett (2003)'s data.
+  The subset of wages data in (1), which is the wages of high school dropouts' as a refreshed version of \textcite{SingerJudithD2003Alda}'s data.
 \end{enumerate}
@@
 
-In the context of official statistics, van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+In the context of official statistics, \textcite{LooMarkvander2018Sdcw} describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
 What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
@@
 
-To get the data into tidy form (Wickham 2014), it needs to comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
+To get the data into tidy form \autocite{WickhamHadley2014TD}, it needs to comply with three rules: (i) each variable forms a column, (ii) each observation forms a row, and (iii) each type of observational unit forms a table.
 The raw data, \texttt{new\_data\_qnames}, violates these rules (i) and (ii) because information about an individual's multiple jobs over different years are in multiple columns.
@@
 Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
-This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
-2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham, François, et al.
-2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
+This is done using the \texttt{tidyverse} suite of packages \autocite{tidyverse}: \texttt{tidyr} \autocite{tidyr} to pivot the data into long-form, with \texttt{dplyr} \autocite{dplyr} and \texttt{stringr} \autocite{stringr} to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
 The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
@@
 However, it is important to note that the use of ``sex'' is probably not correct.
-This information is provided by the individual, and although only has two categories, is more consistent with ``gender'' as defined in Heidari et al.
-(2016).
+This information is provided by the individual, and although only has two categories, is more consistent with ``gender'' as defined in \textcite{SAGER}.
 When using the dataset in the classroom, the educators might include discourse on the use of terms ``sex'' and ``gender''.
@@
 From a statistical perspective, this can make it difficult to adjust survey statistics to the population when gender is measured in different categorization in different data sources.
-Kennedy et al.
-(2020) provides a helpful discussion on these topics.
+\textcite{kennedy2020using} provides a helpful discussion on these topics.
 
 Similarly, there are issues for the \texttt{race} variable.
-The current US Federal guidelines (Office of Management and Budget 1997) state that there should be at least five categories for race, and an individual should be able to identify as more than one race category.
+The current US Federal guidelines \autocite{doh} state that there should be at least five categories for race, and an individual should be able to identify as more than one race category.
 In addition, ``Hispanic'' or ``Latino'' should be considered to be an ethnic group rather than a race.
@@
 The employment and demographic variables are then joined.
-These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in Singer and Willett (2003).
+These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in \textcite{SingerJudithD2003Alda}.
 We also opt to restrict the minimum number of observations so that the data can be used to demonstrate within-person variation when it is used to teach longitudinal data.
@@
 
-Work experience is one of the most important variables in Singer and Willett (2003) as it indicates time and makes longitudinal analysis possible.
+Work experience is one of the most important variables in \textcite{SingerJudithD2003Alda} as it indicates time and makes longitudinal analysis possible.
 It is desirable to calculate this rather than using the survey year for time because it more accurately reflects a person's time in the workforce.
@@
 It is not straightforward.
-We start with the definition of experience in Singer and Willett (2003).
+We start with the definition of experience in \textcite{SingerJudithD2003Alda}.
 
@@
 
-According to Huebner, Vach, and Cessie (2016), initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure the data is clean, valid, and ready to be deployed in the later analyses.
-This is supported by Chatfield (1985), who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
+According to \textcite{HuebnerMariannePhD2016Asat}, initial data analysis (IDA) is the step of inspecting and screening the data after collection to ensure the data is clean, valid, and ready to be deployed in the later analyses.
+This is supported by \textcite{Chatfield1985TIEo}, who argues that the two main objectives of IDA are data description, which is to assess the structure and the quality of the data, and model formulation without any formal statistical inference.
 
@@
 The first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
-This is followed by graphical summaries using methods available in \texttt{ggplot2} (Wickham 2016) and \texttt{brolgar} (Tierney, Cook, and Prvan 2020).
+This is followed by graphical summaries using methods available in \texttt{ggplot2} \autocite{ggplot2} and \texttt{brolgar} \autocite{brolgar}.
 
@@
 Extremely high values were also found in the total hours of work, where some observations reported having worked for 420 hours a week in total.
-According to Pergamit et al.
-(2001), one of the flaws of the NLSY79 employment data is the NLSY79 collects the information of the working hours since the last interview.
+According to \textcite{MichaelRPergamit2001DWTN}, one of the flaws of the NLSY79 employment data is the NLSY79 collects the information of the working hours since the last interview.
 Thus, it might be challenging for the respondents to track the within-job hours' changes between survey years, especially for the respondents with fluctuating working hours or seasonal jobs.
@@
 
-A robust linear regression model using the \texttt{rlm} function from \texttt{MASS} package (Venables and Ripley 2002) is used to treat the extreme values in the data.
+A robust linear regression model using the \texttt{rlm} function from \texttt{MASS} package \autocite{mass} is used to treat the extreme values in the data.
 The robustness weight is used to determine if a value should be replaced with the fitted value from the model.
-This is constructed for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} (Wickham 2020) and \texttt{purrr} (Henry and Wickham 2020), respectively.
-An alternative approach would be a robust linear mixed model using \texttt{robustlmm} (Koller 2016).
+This is constructed for each ID utilizing the \texttt{nest} and \texttt{map} function from \texttt{tidyr} \autocite{tidyr} and \texttt{purrr} \autocite{purrr}, respectively.
+An alternative approach would be a robust linear mixed model using \texttt{robustlmm} \autocite{KollerManuel2016rARP}.
 
@@
 Overly smooth profiles would indicate the replacement of values was too severe, resulting in removing the very interesting volatility of wages seen in many individuals.
-This was achieved using a purpose written \texttt{shiny} (Chang et al.
-2020) app (provided with the code of this paper).
+This was achieved using a purpose written \texttt{shiny} \autocite{shiny} app (provided with the code of this paper).
 Observations where a change was made are tracked with a new variable called \texttt{is\_pred}, so that this effect in the downstream analyses can be monitored.
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart}
+{\centering \includegraphics[width=0.85\linewidth]{../figures/flowchart}
 
@@
 ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data.
-The color represents the stage of data cleaning in the statistical value chain (van der Loo and de Jonge 2021).
+The color represents the stage of data cleaning in the statistical value chain \autocite{validate}.
 Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart}
@@
 
-The original set, containing wages of high school dropouts (Singer and Willett 2003) from 1979 through to 1994, is available in the R package \texttt{brolgar} (Tierney, Cook, and Prvan 2020).
+The original set, containing wages of high school dropouts \autocite{SingerJudithD2003Alda} from 1979 through to 1994, is available in the R package \texttt{brolgar} \autocite{brolgar}.
 To compare the refreshed data with the original, a subset needs to be matched.
@@
 
-There is no explicit explanation of how the dropouts cohort is determined in the original data. Hence, we use the high school dropouts criteria from Wolpin (2005), which are:
+There is no explicit explanation of how the dropouts cohort is determined in the original data. Hence, we use the high school dropouts criteria from \textcite{nlsy79edu}, which are:
 
@@
 
-An additional criterion from Singer and Willett (2003) is to only include males aged between 14 and 17 years old in 1979.
+An additional criterion from \textcite{SingerJudithD2003Alda} is to only include males aged between 14 and 17 years old in 1979.
 With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
@@
 The calculation of the work experience is not clearly articulated.
-Singer and Willett (2003) describe the temporal variable as the years of experience since the first day of work.
+\textcite{SingerJudithD2003Alda} describe the temporal variable as the years of experience since the first day of work.
 This variable is not explicitly available in the database.
-From the NLSY79 topical guide (Bureau of Labor Statistics, U.S.
-Department of Labor 2021b), we find several variables are tagged as work experience-related variables.
+From the NLSY79 topical guide \autocite{nlsy79guide}, we find several variables are tagged as work experience-related variables.
 One of them is the weeks worked since the last interview.
@@
 It is important to note that the treatment of unlikely wages differs in the refreshed data.
-In the original data by Singer and Willett (2003), wages greater than \$75 are set to be missing.
+In the original data by \textcite{SingerJudithD2003Alda}, wages greater than \$75 are set to be missing.
 However, this value is too low to be set as the maximum threshold, and it doesn't take into account temporal neighbors for an individual.
@@
 These provide the opportunity to again refresh the textbook data as new data is published into the NLSY79 database.
-Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a \texttt{shiny} (Chang et al.
-2020) app, which is provided with the code, and the choices used in the refreshed data are documented.
+Determining an appropriate robustness weight from which to threshold unusual observations was conducted using a \texttt{shiny} \autocite{shiny} app, which is provided with the code, and the choices used in the refreshed data are documented.
 
@@
 \item
-  The refreshed data can be used for teaching advanced linear modeling, using almost identical code from Singer and Willett (2003), and thus make those lessons more current for today's students.
+  The refreshed data can be used for teaching advanced linear modeling, using almost identical code from \textcite{SingerJudithD2003Alda}, and thus make those lessons more current for today's students.
 \item
   It provides the data to develop a case study for teaching exploratory data analysis of longitudinal data, focusing on the individual experience.
-The vignettes of the \texttt{brolgar} (Tierney, Cook, and Prvan 2020) package provide useful guides for the case study content.
+The vignettes of the \texttt{brolgar} \autocite{brolgar} package provide useful guides for the case study content.
 \end{itemize}
@@
 
-The entire analysis is conducted using \texttt{R} (R Core Team 2020) in RStudio IDE using these packages: \texttt{tidyverse} (Wickham, Averick, et al.
-2019), \texttt{ggplot2} (Wickham 2016), \texttt{dplyr} (Wickham, François, et al.
-2020), \texttt{readr} (Wickham and Hester 2020), \texttt{tidyr} (Wickham 2020), \texttt{stringr} (Wickham 2019), \texttt{purrr} (Henry and Wickham 2020), \texttt{brolgar} (Tierney, Cook, and Prvan 2020), \texttt{patchwork} (Pedersen 2020), \texttt{kableExtra} (Zhu 2019), \texttt{MASS} (Venables and Ripley 2002), \texttt{janitor} (Firke 2020), and \texttt{tsibble} (Wang, Cook, and Hyndman 2020).
-The paper was generated using \texttt{knitr} (Xie 2014) and \texttt{rmarkdown} (Xie, Dervieux, and Riederer 2020).
+The entire analysis is conducted using \texttt{R} \autocite{R} in RStudio IDE using these packages: \texttt{tidyverse} \autocite{tidyverse}, \texttt{ggplot2} \autocite{ggplot2}, \texttt{dplyr} \autocite{dplyr}, \texttt{readr} \autocite{readr}, \texttt{tidyr} \autocite{tidyr}, \texttt{stringr} \autocite{stringr}, \texttt{purrr} \autocite{purrr}, \texttt{brolgar} \autocite{brolgar}, \texttt{patchwork} \autocite{patchwork}, \texttt{kableExtra} \autocite{kableExtra}, \texttt{MASS} \autocite{mass}, \texttt{janitor} \autocite{janitor}, and \texttt{tsibble} \autocite{tsibble}.
+The paper was generated using \texttt{knitr} \autocite{knitr} and \texttt{rmarkdown} \autocite{rmarkdown}.
 
@@
 
-\hypertarget{references}{%
-\section*{References}\label{references}}
-\addcontentsline{toc}{section}{References}
-
-\hypertarget{refs}{}
-\leavevmode\hypertarget{ref-iris-data}{}%
-Anderson, Edgar. 1935. ``The Irises of the Gaspe Peninsula.'' \emph{Bulletin of the American Iris Society} 59: 2--5.
-
-\leavevmode\hypertarget{ref-nlsy79}{}%
-Bureau of Labor Statistics, U.S.
-Department of Labor.
-2021a.
-``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
-Columbus, OH, through \url{https://www.nlsinfo.org/bibliography-citing-nls-data}.
-
-\leavevmode\hypertarget{ref-nlsy79guide}{}%
----------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.'' \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
-
-\leavevmode\hypertarget{ref-shiny}{}%
-Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.
-
-\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
-Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
-
-\leavevmode\hypertarget{ref-eliznlsy}{}%
-Cooksey, Elizabeth C.
-2017.
-``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
-Lerner Neal Halfon Christoper B.
-Forrest, 561--77.
-Cham: Springer.
-\url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.
-
-\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
-Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.
-
-\leavevmode\hypertarget{ref-janitor}{}%
-Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.
-
-\leavevmode\hypertarget{ref-racismnotrace}{}%
-Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.
-
-\leavevmode\hypertarget{ref-grimshaw}{}%
-Grimshaw, Scott D.
-2015.
-``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14.
-\url{https://doi.org/https://doi.org/10.1080/00031305.2015.1081106}.
-
-\leavevmode\hypertarget{ref-SAGER}{}%
-Heidari, S., T.
-F.
-Babor, P.
-De Castro, S.
-Tort, and M.
-Curno.
-2016.
-``Sex and Gender Equity in Research: Rationale for the SAGER Guidelines and Recommended Use.'' \emph{Research Integrity and Peer Review} 1 (2).
-\url{https://doi.org/https://doi.org/10.1186/s41073-016-0007-6}.
-
-\leavevmode\hypertarget{ref-purrr}{}%
-Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.
-
-\leavevmode\hypertarget{ref-penguins-data}{}%
-Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
-
-\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
-Huebner, Marianne, Werner Vach, and Saskia le Cessie.
-2016.
-``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
-
-\leavevmode\hypertarget{ref-ilk2004}{}%
-Ilk, Ozlem.
-2004.
-``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.'' PhD thesis, Iowa State University.
-\url{https://doi.org/10.31274/rtd-180813-11012}.
-
-\leavevmode\hypertarget{ref-kennedy2020using}{}%
-Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. ``Using Sex and Gender in Survey Adjustment.'' \url{http://arxiv.org/abs/2009.14401}.
-
-\leavevmode\hypertarget{ref-tamedata}{}%
-Kim, A.
-Y, C.
-Ismay, and J.
-Chunn.
-2018.
-``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1).
-\url{https://doi.org/https://doi.org/10.5070/T511103589}.
-
-\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
-Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.
-
-\leavevmode\hypertarget{ref-notaverage}{}%
-Moncrief, Marc. 2015. ``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.'' \url{https://bit.ly/smh-not-normal}.
-
-\leavevmode\hypertarget{ref-doh}{}%
-Office of Management and Budget.
-1997.
-``Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity.'' \url{https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf}.
-
-\leavevmode\hypertarget{ref-opendata}{}%
-Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.
-
-\leavevmode\hypertarget{ref-patchwork}{}%
-Pedersen, Thomas Lin. 2020. \emph{patchwork: The Composer of Plots}. \url{https://CRAN.R-project.org/package=patchwork}.
-
-\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
-Pergamit, Michael R., Charles R.
-Pierret, Donna S.
-Rothstein, and Jonathan R.
-Veum.
-2001.
-``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
-
-\leavevmode\hypertarget{ref-R}{}%
-R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
-
-\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
-Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.
-
-\leavevmode\hypertarget{ref-notiris}{}%
-Stodel, Megan. 2020. ``Stop Using Iris.'' \url{https://www.meganstodel.com/posts/no-to-iris/}.
-
-\leavevmode\hypertarget{ref-brolgar}{}%
-Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.
-
-\leavevmode\hypertarget{ref-tukey}{}%
-Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.
-
-\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
-van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.
-
-\leavevmode\hypertarget{ref-validate}{}%
-van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
-
-\leavevmode\hypertarget{ref-mass}{}%
-Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
-
-\leavevmode\hypertarget{ref-tsibble}{}%
-Wang, Earo, Dianne Cook, and Rob J Hyndman.
-2020.
-``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
-\url{https://doi.org/10.1080/10618600.2019.1695624}.
-
-\leavevmode\hypertarget{ref-plyr}{}%
-Wickham, Hadley.
-2011.
-``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29.
-\url{https://doi.org/https://doi.org/10.18637/jss.v040.i01}.
-
-\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
----------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.
-
-\leavevmode\hypertarget{ref-ggplot2}{}%
----------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
-
-\leavevmode\hypertarget{ref-stringr}{}%
----------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.
-
-\leavevmode\hypertarget{ref-tidyr}{}%
----------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.
-
-\leavevmode\hypertarget{ref-tidyverse}{}%
-Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al.
-2019.
-``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686.
-\url{https://doi.org/10.21105/joss.01686}.
-
-\leavevmode\hypertarget{ref-dplyr}{}%
-Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.
-
-\leavevmode\hypertarget{ref-readr}{}%
-Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.
-
-\leavevmode\hypertarget{ref-nlsy79edu}{}%
-Wolpin, Kenneth I.
-2005.
-``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Published by Bureau of Labor Statistics, U.S.
-Department of Labor.
-\url{https://www.bls.gov/opub/mlr/2005/02/art3full.pdf}.
-
-\leavevmode\hypertarget{ref-knitr}{}%
-Xie, Yihui.
-2014.
-``Knitr: A Comprehensive Tool for Reproducible Research in R.'' In \emph{Implementing Reproducible Computational Research}, edited by Victoria Stodden, Friedrich Leisch, and Roger D.
-Peng.
-Chapman; Hall/CRC.
-\url{http://www.crcpress.com/product/isbn/9781466561595}.
-
-\leavevmode\hypertarget{ref-rmarkdown}{}%
-Xie, Yihui, Christophe Dervieux, and Emily Riederer. 2020. \emph{R Markdown Cookbook}. Boca Raton, Florida: Chapman; Hall/CRC. \url{https://bookdown.org/yihui/rmarkdown-cookbook}.
-
-\leavevmode\hypertarget{ref-kableExtra}{}%
-Zhu, Hao. 2019. \emph{kableExtra: Construct Complex Table with 'kable' and Pipe Syntax}. \url{https://CRAN.R-project.org/package=kableExtra}.
-
-\bibliographystyle{unsrt}
-\bibliography{references.bib}
+\printbibliography[title=References]
 

### 2022-06-07 paper/cover-letter/cover-letter.Rmd
Add response to the editors.

@@
 
-Please consider the revised version of our paper titled "The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth" for the JDSDE Special Issue on "Teaching Reproducibility and Responsible Workflow".
+Please consider the second revision of our paper titled "The Journey from Wild to Textbook Data: A Case Study from the National Longitudinal Survey of Youth" for the JDSDE Special Issue on "Teaching Reproducibility and Responsible Workflow".
 
 The paper is co-authored by Prof.
-Di Cook, Dr Emi Tanaka, Kate Hyde and Dr Nick Tierney.
+Di Cook, Dr Emi Tanaka, Kate Hyde, and Dr Nick Tierney.
 We believe it is a good fit for the special issue because it covers various aspects of data management and curation component of data acumen, including data preparation and cleaning, missing and conflicting data.
 The code to reproduce the data cleaning, and make further updates of the resulting data is provided.
+We also believe that this data fits for teaching longitudinal data-purpose.
 
-We also provide the response of editor and reviewers' comments. We also agree if you share this paper with the discussant for this special issue.
+We also provide the response of editor and reviewers' comments.

### 2022-06-07 response/response3.Rmd
Add response to the editors.

---
title: Response to reviewers
author: ""
fontsize: 12pt
date: "Submission ID 217815520.R2"
output: monash::memo
# if you don't have the monash package install by:
# remotes::install_github("numbats/monash")
---

We thank the editor, associate editor and the reviewers for their feedback. Here is how we have made revisions to the manuscript to address each point.

The editor says \textcolor{violet}{The reviews are in general quite favorable and suggest that, subject to the minor revisions that the paper would be suitable for publication as part of the special issue on reproducibility and responsible workflow.}

Reviewer 3 also comments that \textcolor{violet}{Thank you for this revision.  My minor comments and corrections are all well addressed, and I particularly commend the authors for their changes to Section 3.4.  It was a great idea to create a Shiny app showing the exploratory process that lead to your choice of threshold; this feels far more justifiable and reproducible than reporting the final number alone.}

\section*{Major changes:}

The only comments on the pedagogical application of the data appear to come from the associate editor: \textcolor{violet}{While the data are interesting, the necessary details for a pedagogical application or case study are still largely missing. The manuscript reads more like a paper about the R package, something for a software journal. In my last review, I had missed that this was for a special issue, still I agree with the reviewer that the clear contributions to educators other than its use in class are plenty, but not fully described.}

\section*{Specific comments:}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{The abstract promises a discussion of the process for cleaning data and exploratory analyses for longitudinal data. The latter is not provided, nor is a discussion about how that might be done in a class activity. This is a place for adding another clear contribution to educators. (Alternatively it can be deleted).}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 10, line 18) “these” is errant}
\end{description}

* Dewi to do

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 11, line 14) Providing one or two sentences about the use of pivot\_longer() would help readers gain insight and motivate clicking the link to the code.}
\end{description}

* Dewi to do

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 11, line 35) The sentence beginning “This information is provided” is missing at least one word.}
\end{description}

* Dewi to do

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 12, line 49) There is still the error stating that education should only increase. Certainly, you mean non-decreasing.}
\end{description}

* Dewi to do

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 13, line 24) I’m just checking that all Q... variables are correct. The formatting is inconsistent; e.g., Q3-8A versus Q1-3\_A}
\end{description}

* Yes it is correct, the formatting of all Q... variables uses the same format as downloaded data from the database.

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 13, line 36) The sentence beginning “Our target variables” is missing at least one word.}
\end{description}

* Dewi to do

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 14, line 20) Why is 84 hours the cut-off? This seems to be an arbitrary choice.}
\end{description}

* It is assumed that people will not work more than 12 hours a day. Dewi to do.

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 15, line 3) The sentence beginning “For stwork variable” is missing at least one word.}
\end{description}

* Dewi to do

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 15, line 3)  It might be helpful for readers if you refer to variables as what they are and put the column title in parentheses.
}
\end{description}

* Dewi to do

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 15, line 15) Providing one or two sentences about the use of join() would help readers gain insight into what the authors are doing here.}
\end{description}

* Dewi to do

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 22, line 5) The default for rlm is huber weighting which never applies a weight of 0. You may have meant bisquare, but that change should be discussed.}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 28, figure 7) This is a helpful graph, but perhaps making the barplot “position dodge” and the density plots overlapping with transparency would sell your point better. This graphic makes the point, but the suggestions I made may produce more compelling evidence.}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 29, line 21) The sentence “The highest grade completed has some confusion” should be rewritten.}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 29, line 37) Naming the function would be helpful for readers interested in using it.}
\end{description}

* Dewi to do

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 31, line 9) Certainly you don’t mean to say that you’ve created data ”unsatisfactorily far from” the original.}
\end{description}

\begin{description}
\item[Associate Editor:] \textcolor{violet}{(page 32, line 18-30) Here you describe some modeling (see first note). Re-framing this as a possible activity could be helpful to the reader. As it is written now it sounds like a summary (the section title) of what was presented – instead you can make it a path forward for the readers.}
\end{description}

### 2022-06-07 paper/paper.tex
response

@@
 
-In the context of official statistics, van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
+In the context of official statistics, M.
+van der Loo and de Jonge (2018) describe the ``statistical value chain,'' which includes various production stages of the data cleaning process as raw data (initial data as delivered originally), input data (data organized with correct type and identified variables), and valid data (data that has been cleaned and more accurately represents the intent of variables).
 What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
@@
 Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
-This is done using the \texttt{tidyverse} suite of packages (Wickham, Averick, et al.
-2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham, François, et al.
+This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
+2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham et al.
 2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
@@
 
-{\centering \includegraphics[width=0.85\linewidth]{/Users/deeamaliah/Documents/yowie-paper/paper/figures/flowchart}
+{\centering \includegraphics[width=0.85\linewidth]{figures/flowchart}
 
@@
 ``\# of individuals'' means the number of respondents included in each stage, while ``\# of observations'' means the number of rows in the data.
-The color represents the stage of data cleaning in the statistical value chain (van der Loo and de Jonge 2021).
+The color represents the stage of data cleaning in the statistical value chain (M.
+P.
+J.
+van der Loo and de Jonge 2021).
 Pink, blue, and green represent the raw, input, and valid data, respectively.}\label{fig:flow-chart}
@@
 \hypertarget{refs}{}
-\leavevmode\hypertarget{ref-iris-data}{}%
-Anderson, Edgar. 1935. ``The Irises of the Gaspe Peninsula.'' \emph{Bulletin of the American Iris Society} 59: 2--5.
-
-\leavevmode\hypertarget{ref-nlsy79}{}%
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-iris-data}{}}%
+Anderson, Edgar. 1935. {``The Irises of the Gaspe Peninsula.''} \emph{Bulletin of the American Iris Society} 59: 2--5.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nlsy79}{}}%
 Bureau of Labor Statistics, U.S.
@@
 2021a.
-``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
+{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Produced and distributed by the Center for Human Resource Research (CHRR), The Ohio State University.
 Columbus, OH, through \url{https://www.nlsinfo.org/bibliography-citing-nls-data}.
 
-\leavevmode\hypertarget{ref-nlsy79guide}{}%
----------. 2021b. ``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.'' \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
-
-\leavevmode\hypertarget{ref-shiny}{}%
-Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{shiny: Web Application Framework for R}. \url{https://CRAN.R-project.org/package=shiny}.
-
-\leavevmode\hypertarget{ref-Chatfield1985TIEo}{}%
-Chatfield, C. 1985. ``The Initial Examination of Data.'' \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
-
-\leavevmode\hypertarget{ref-eliznlsy}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-nlsy79guide}{}}%
+---------. 2021b. {``National Longitudinal Survey of Youth 1979 Cohort, Topical Guide to the Data.''} \url{https://www.nlsinfo.org/content/cohorts/nlsy79/topical-guide/employment/work-experience}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-shiny}{}}%
+Chang, Winston, Joe Cheng, JJ Allaire, Yihui Xie, and Jonathan McPherson. 2020. \emph{{shiny: Web Application Framework for R}}. \url{https://CRAN.R-project.org/package=shiny}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Chatfield1985TIEo}{}}%
+Chatfield, C. 1985. {``The Initial Examination of Data.''} \emph{Journal of the Royal Statistical Society. Series A. General} 148 (3): 214--53.
+
+\leavevmode\vadjust pre{\hypertarget{ref-eliznlsy}{}}%
 Cooksey, Elizabeth C.
 2017.
-``Using the National Longitudinal Surveys of Youth (NLSY) to Conduct Life Course Analyses.'' In \emph{Handbook of Life Course Health Development}, edited by Richard M.
+{``Using the National Longitudinal Surveys of Youth ({NLSY}) to Conduct Life Course Analyses.''} In \emph{Handbook of Life Course Health Development}, edited by Richard M.
 Lerner Neal Halfon Christoper B.
@@
 Cham: Springer.
-\url{https://doi.org/https://doi.org/10.1007/978-3-319-47143-3_23}.
-
-\leavevmode\hypertarget{ref-DasuTamraparni2003Edma}{}%
+https://doi.org/\url{https://doi.org/10.1007/978-3-319-47143-3_23}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-DasuTamraparni2003Edma}{}}%
 Dasu, Tamraparni, and Theodore Johnson. 2003. \emph{Exploratory Data Mining and Data Cleaning}. Wiley Series in Probability and Statistics. Hoboken: WILEY.
 
-\leavevmode\hypertarget{ref-janitor}{}%
-Firke, Sam. 2020. \emph{janitor: Simple Tools for Examining and Cleaning Dirty Data}. \url{https://CRAN.R-project.org/package=janitor}.
-
-\leavevmode\hypertarget{ref-racismnotrace}{}%
-Fullilove, M. T. 1998. ``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.'' \emph{American Journal of Public Health} 88 (9): 1297--8.
-
-\leavevmode\hypertarget{ref-grimshaw}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-janitor}{}}%
+Firke, Sam. 2020. \emph{{janitor: Simple Tools for Examining and Cleaning Dirty Data}}. \url{https://CRAN.R-project.org/package=janitor}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-racismnotrace}{}}%
+Fullilove, M. T. 1998. {``Comment: Abandoning ``Race" as a Variable in Public Health Research--an Idea Whose Time Has Come.''} \emph{American Journal of Public Health} 88 (9): 1297--98.
+
+\leavevmode\vadjust pre{\hypertarget{ref-grimshaw}{}}%
 Grimshaw, Scott D.
 2015.
-``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.'' \emph{The American Statistician} 69 (4): 307--14.
-\url{https://doi.org/https://doi.org/10.1080/00031305.2015.1081106}.
-
-\leavevmode\hypertarget{ref-SAGER}{}%
+{``A Framework for Infusing Authentic Data Experiences Within Statistics Courses.''} \emph{The American Statistician} 69 (4): 307--14.
+https://doi.org/\url{https://doi.org/10.1080/00031305.2015.1081106}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-SAGER}{}}%
 Heidari, S., T.
@@
 2016.
-``Sex and Gender Equity in Research: Rationale for the SAGER Guidelines and Recommended Use.'' \emph{Research Integrity and Peer Review} 1 (2).
-\url{https://doi.org/https://doi.org/10.1186/s41073-016-0007-6}.
-
-\leavevmode\hypertarget{ref-purrr}{}%
-Henry, Lionel, and Hadley Wickham. 2020. \emph{purrr: Functional Programming Tools}. \url{https://CRAN.R-project.org/package=purrr}.
-
-\leavevmode\hypertarget{ref-penguins-data}{}%
+{``Sex and Gender Equity in Research: Rationale for the {SAGER} Guidelines and Recommended Use.''} \emph{Research Integrity and Peer Review} 1 (2).
+https://doi.org/\url{https://doi.org/10.1186/s41073-016-0007-6}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-purrr}{}}%
+Henry, Lionel, and Hadley Wickham. 2020. \emph{{purrr: Functional Programming Tools}}. \url{https://CRAN.R-project.org/package=purrr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-penguins-data}{}}%
 Horst, Allison Marie, Alison Presmanes Hill, and Kristen B Gorman. 2020. \emph{Palmerpenguins: Palmer Archipelago (Antarctica) Penguin Data}. \url{https://doi.org/10.5281/zenodo.3960218}.
 
-\leavevmode\hypertarget{ref-HuebnerMariannePhD2016Asat}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-HuebnerMariannePhD2016Asat}{}}%
 Huebner, Marianne, Werner Vach, and Saskia le Cessie.
 2016.
-``A Systematic Approach to Initial Data Analysis Is Good Research Practice.'' \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
-
-\leavevmode\hypertarget{ref-ilk2004}{}%
+{``A Systematic Approach to Initial Data Analysis Is Good Research Practice.''} \emph{The Journal of Thoracic and Cardiovascular Surgery} 151 (1): 25--27.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ilk2004}{}}%
 Ilk, Ozlem.
 2004.
-``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.'' PhD thesis, Iowa State University.
+{``Exploratory Multivariate Longitudinal Data Analysis and Models for Multivariate Longitudinal Binary Data.''} PhD thesis, Iowa State University.
 \url{https://doi.org/10.31274/rtd-180813-11012}.
 
-\leavevmode\hypertarget{ref-kennedy2020using}{}%
-Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. ``Using Sex and Gender in Survey Adjustment.'' \url{http://arxiv.org/abs/2009.14401}.
-
-\leavevmode\hypertarget{ref-tamedata}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-kennedy2020using}{}}%
+Kennedy, Lauren, Katharine Khanna, Daniel Simpson, and Andrew Gelman. 2020. {``Using Sex and Gender in Survey Adjustment.''} \url{https://arxiv.org/abs/2009.14401}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tamedata}{}}%
 Kim, A.
@@
 2018.
-``The Fivethirtyeight R Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.'' \emph{Technology Innovations in Statistics Education} 11 (1).
-\url{https://doi.org/https://doi.org/10.5070/T511103589}.
-
-\leavevmode\hypertarget{ref-KollerManuel2016rARP}{}%
-Koller, Manuel. 2016. ``robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models.'' \emph{Journal of Statistical Software} 75 (6): 1--24.
-
-\leavevmode\hypertarget{ref-notaverage}{}%
-Moncrief, Marc. 2015. ``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.'' \url{https://bit.ly/smh-not-normal}.
-
-\leavevmode\hypertarget{ref-doh}{}%
+{``The Fivethirtyeight {R} Package: ``Tame Data" Principles for Introductory Statistics and Data Science Courses.''} \emph{Technology Innovations in Statistics Education} 11 (1).
+https://doi.org/\url{https://doi.org/10.5070/T511103589}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-KollerManuel2016rARP}{}}%
+Koller, Manuel. 2016. {``{robustlmm: An R Package for Robust Estimation of Linear Mixed-Effects Models}.''} \emph{Journal of Statistical Software} 75 (6): 1--24.
+
+\leavevmode\vadjust pre{\hypertarget{ref-notaverage}{}}%
+Moncrief, Marc. 2015. {``By the Numbers - the Average Australian Doesn't Exist ... Not a Single One of Us Is 'Normal'.''} \url{https://bit.ly/smh-not-normal}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-doh}{}}%
 Office of Management and Budget.
 1997.
-``Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity.'' \url{https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf}.
-
-\leavevmode\hypertarget{ref-opendata}{}%
-Open Knowledge Foundation. 2021. ``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.'' 2021. \url{http://opendefinition.org/od/2.1/en/}.
-
-\leavevmode\hypertarget{ref-patchwork}{}%
-Pedersen, Thomas Lin. 2020. \emph{patchwork: The Composer of Plots}. \url{https://CRAN.R-project.org/package=patchwork}.
-
-\leavevmode\hypertarget{ref-MichaelRPergamit2001DWTN}{}%
+{``Revisions to the Standards for the Classification of Federal Data on Race and Ethnicity.''} \url{https://www.govinfo.gov/content/pkg/FR-1997-10-30/pdf/97-28653.pdf}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-opendata}{}}%
+Open Knowledge Foundation. 2021. {``Open Definition. Defining Open in Open Data, Open Content, and Open Knowledge.''} 2021. \url{http://opendefinition.org/od/2.1/en/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-patchwork}{}}%
+Pedersen, Thomas Lin. 2020. \emph{{patchwork: The Composer of Plots}}. \url{https://CRAN.R-project.org/package=patchwork}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-MichaelRPergamit2001DWTN}{}}%
 Pergamit, Michael R., Charles R.
@@
 2001.
-``Data Watch: The National Longitudinal Surveys.'' \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
-
-\leavevmode\hypertarget{ref-R}{}%
+{``Data Watch: The National Longitudinal Surveys.''} \emph{The Journal of Economic Perspectives} 15 (2): 239--53.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R}{}}%
 R Core Team. 2020. \emph{R: A Language and Environment for Statistical Computing}. Vienna, Austria: R Foundation for Statistical Computing. \url{https://www.R-project.org/}.
 
-\leavevmode\hypertarget{ref-SingerJudithD2003Alda}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-SingerJudithD2003Alda}{}}%
 Singer, Judith D, and John B Willett. 2003. \emph{Applied Longitudinal Data Analysis: Modeling Change and Event Occurrence}. Oxford u.a: Oxford Univ. Pr.
 
-\leavevmode\hypertarget{ref-notiris}{}%
-Stodel, Megan. 2020. ``Stop Using Iris.'' \url{https://www.meganstodel.com/posts/no-to-iris/}.
-
-\leavevmode\hypertarget{ref-brolgar}{}%
-Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}. \url{https://github.com/njtierney/brolgar}.
-
-\leavevmode\hypertarget{ref-tukey}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-notiris}{}}%
+Stodel, Megan. 2020. {``Stop Using Iris.''} \url{https://www.meganstodel.com/posts/no-to-iris/}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-brolgar}{}}%
+Tierney, Nicholas, Di Cook, and Tania Prvan. 2020. \emph{{brolgar: BRowse Over Longitudinal data Graphically and Analytically in R}}. \url{https://github.com/njtierney/brolgar}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tukey}{}}%
 Tukey, John W. (John Wilder). 1977. \emph{Exploratory Data Analysis}. Addison-Wesley Series in Behavioral Science. Reading, Mass.: Addison-Wesley Pub. Co.
 
-\leavevmode\hypertarget{ref-LooMarkvander2018Sdcw}{}%
-van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in R}.
-
-\leavevmode\hypertarget{ref-validate}{}%
-van der Loo, Mark P. J., and Edwin de Jonge. 2021. ``Data Validation Infrastructure for R.'' \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
-
-\leavevmode\hypertarget{ref-mass}{}%
-Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with S}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
-
-\leavevmode\hypertarget{ref-tsibble}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%
+van der Loo, Mark P. J., and Edwin de Jonge. 2021. {``Data Validation Infrastructure for {R}.''} \emph{Journal of Statistical Software} 97 (10): 1--31. \url{https://doi.org/10.18637/jss.v097.i10}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-LooMarkvander2018Sdcw}{}}%
+van der Loo, Mark, and Edwin de Jonge. 2018. \emph{Statistical Data Cleaning with Applications in r}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-mass}{}}%
+Venables, W. N., and B. D. Ripley. 2002. \emph{Modern Applied Statistics with s}. Fourth. New York: Springer. \url{http://www.stats.ox.ac.uk/pub/MASS4}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tsibble}{}}%
 Wang, Earo, Dianne Cook, and Rob J Hyndman.
 2020.
-``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.'' \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
+{``A New Tidy Data Structure to Support Exploration and Modeling of Temporal Data.''} \emph{Journal of Computational and Graphical Statistics} 29 (3): 466--78.
 \url{https://doi.org/10.1080/10618600.2019.1695624}.
 
-\leavevmode\hypertarget{ref-plyr}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-plyr}{}}%
 Wickham, Hadley.
 2011.
-``The Split-Apply-Combine Strategy for Data Analysis.'' \emph{Journal of Statistical Software, Articles} 40 (1): 1--29.
-\url{https://doi.org/https://doi.org/10.18637/jss.v040.i01}.
-
-\leavevmode\hypertarget{ref-WickhamHadley2014TD}{}%
----------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software} 59 (10): 1--23.
-
-\leavevmode\hypertarget{ref-ggplot2}{}%
----------. 2016. \emph{ggplot2: Elegant Graphics for Data Analysis}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
-
-\leavevmode\hypertarget{ref-stringr}{}%
----------. 2019. \emph{stringr: Simple, Consistent Wrappers for Common String Operations}. \url{https://CRAN.R-project.org/package=stringr}.
-
-\leavevmode\hypertarget{ref-tidyr}{}%
----------. 2020. \emph{tidyr: Tidy Messy Data}. \url{https://CRAN.R-project.org/package=tidyr}.
-
-\leavevmode\hypertarget{ref-tidyverse}{}%
+{``The Split-Apply-Combine Strategy for Data Analysis.''} \emph{Journal of Statistical Software, Articles} 40 (1): 1--29.
+https://doi.org/\url{https://doi.org/10.18637/jss.v040.i01}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-WickhamHadley2014TD}{}}%
+---------. 2014. {``Tidy Data.''} \emph{Journal of Statistical Software} 59 (10): 1--23.
+
+\leavevmode\vadjust pre{\hypertarget{ref-ggplot2}{}}%
+---------. 2016. \emph{{ggplot2: Elegant Graphics for Data Analysis}}. Springer-Verlag New York. \url{https://ggplot2.tidyverse.org}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-stringr}{}}%
+---------. 2019. \emph{{stringr: Simple, Consistent Wrappers for Common String Operations}}. \url{https://CRAN.R-project.org/package=stringr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tidyr}{}}%
+---------. 2020. \emph{{tidyr: Tidy Messy Data}}. \url{https://CRAN.R-project.org/package=tidyr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-tidyverse}{}}%
 Wickham, Hadley, Mara Averick, Jennifer Bryan, Winston Chang, Lucy D'Agostino McGowan, Romain François, Garrett Grolemund, et al.
 2019.
-``Welcome to the tidyverse.'' \emph{Journal of Open Source Software} 4 (43): 1686.
+{``Welcome to the {tidyverse}.''} \emph{Journal of Open Source Software} 4 (43): 1686.
 \url{https://doi.org/10.21105/joss.01686}.
 
-\leavevmode\hypertarget{ref-dplyr}{}%
-Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{dplyr: A Grammar of Data Manipulation}. \url{https://CRAN.R-project.org/package=dplyr}.
-
-\leavevmode\hypertarget{ref-readr}{}%
-Wickham, Hadley, and Jim Hester. 2020. \emph{readr: Read Rectangular Text Data}. \url{https://CRAN.R-project.org/package=readr}.
-
-\leavevmode\hypertarget{ref-nlsy79edu}{}%
+\leavevmode\vadjust pre{\hypertarget{ref-dplyr}{}}%
+Wickham, Hadley, Romain François, Lionel Henry, and Kirill Müller. 2020. \emph{{dplyr: A Grammar of Data Manipulation}}. \url{https://CRAN.R-project.org/package=dplyr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-readr}{}}%
+Wickham, Hadley, and Jim Hester. 2020. \emph{{readr: Read Rectangular Text Data}}. \url{https://CRAN.R-project.org/package=readr}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nlsy79edu}{}}%
 Wolpin, Kenneth I.
 2005.
-``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).'' Published by Bureau of Labor Statistics, U.S.
+{``National Longitudinal Survey of Youth 1979 Cohort, 1979-2016 (Rounds 1-28).''} Published by Bureau of Labor Statistics, U.S.
 Department of Labor.

### 2022-06-07 response/response3.Rmd
response

@@
 title: Response to reviewers
-author: ""
+author: "Amaliah, Cook, Tanaka, Hyde, Tierney"
 fontsize: 12pt
@@
 
-The editor says \textcolor{violet}{The reviews are in general quite favorable and suggest that, subject to the minor revisions that the paper would be suitable for publication as part of the special issue on reproducibility and responsible workflow.}
+Editor: \textcolor{violet}{The reviews are in general quite favorable and suggest that, subject to the minor revisions that the paper would be suitable for publication as part of the special issue on reproducibility and responsible workflow.}
 
-Reviewer 3 also comments that \textcolor{violet}{Thank you for this revision.
+Reviewer 3: \textcolor{violet}{Thank you for this revision.
 My minor comments and corrections are all well addressed, and I particularly commend the authors for their changes to Section 3.4.
@@
 
-\section*{Major changes:}
+Thank you!
 
-The only comments on the pedagogical application of the data appear to come from the associate editor: \textcolor{violet}{While the data are interesting, the necessary details for a pedagogical application or case study are still largely missing.
+\section*{Associate editor}
+
+\textcolor{violet}{While the data are interesting, the necessary details for a pedagogical application or case study are still largely missing.
 The manuscript reads more like a paper about the R package, something for a software journal.
@@
 
-\section*{Specific comments:}
+\textcolor{violet}{The main item left to be addressed is polish. This twice-revised manuscript still contains several places where authors are missing words, awkward phrasing, or grammatical errors.}
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{The abstract promises a discussion of the process for cleaning data and exploratory analyses for longitudinal data.
+\textcolor{violet}{The abstract promises a discussion of the process for cleaning data and exploratory analyses for longitudinal data.
 The latter is not provided, nor is a discussion about how that might be done in a class activity.
@@
 (Alternatively it can be deleted).}
-\end{description}
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 10, line 18) “these” is errant}
-\end{description}
+\textcolor{violet}{(page 10, line 18) “these” is errant}
 
@@
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 11, line 14) Providing one or two sentences about the use of pivot\_longer() would help readers gain insight and motivate clicking the link to the code.}
-\end{description}
+\textcolor{violet}{(page 11, line 14) Providing one or two sentences about the use of pivot\_longer() would help readers gain insight and motivate clicking the link to the code.}
 
@@
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 11, line 35) The sentence beginning “This information is provided” is missing at least one word.}
-\end{description}
+\textcolor{violet}{(page 11, line 35) The sentence beginning “This information is provided” is missing at least one word.}
 
@@
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 12, line 49) There is still the error stating that education should only increase. Certainly, you mean non-decreasing.}
-\end{description}
+\textcolor{violet}{(page 12, line 49) There is still the error stating that education should only increase. Certainly, you mean non-decreasing.}
 
@@
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 13, line 24) I’m just checking that all Q... variables are correct. The formatting is inconsistent; e.g., Q3-8A versus Q1-3\_A}
-\end{description}
+\textcolor{violet}{(page 13, line 24) I’m just checking that all Q... variables are correct. The formatting is inconsistent; e.g., Q3-8A versus Q1-3\_A}
 
@@
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 13, line 36) The sentence beginning “Our target variables” is missing at least one word.}
-\end{description}
+\textcolor{violet}{(page 13, line 36) The sentence beginning “Our target variables” is missing at least one word.}
 
@@
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 14, line 20) Why is 84 hours the cut-off? This seems to be an arbitrary choice.}
-\end{description}
+\textcolor{violet}{(page 14, line 20) Why is 84 hours the cut-off? This seems to be an arbitrary choice.}
 
@@
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 15, line 3) The sentence beginning “For stwork variable” is missing at least one word.}
-\end{description}
+\textcolor{violet}{(page 15, line 3) The sentence beginning “For stwork variable” is missing at least one word.}
 
@@
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 15, line 3)  It might be helpful for readers if you refer to variables as what they are and put the column title in parentheses.
+\textcolor{violet}{(page 15, line 3)  It might be helpful for readers if you refer to variables as what they are and put the column title in parentheses.
 }
-\end{description}
 
@@
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 15, line 15) Providing one or two sentences about the use of join() would help readers gain insight into what the authors are doing here.}
-\end{description}
+\textcolor{violet}{(page 15, line 15) Providing one or two sentences about the use of join() would help readers gain insight into what the authors are doing here.}
 
@@
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 22, line 5) The default for rlm is huber weighting which never applies a weight of 0.
-You may have meant bisquare, but that change should be discussed.}
-\end{description}
+\textcolor{violet}{(page 22, line 5) The default for rlm is huber weighting which never applies a weight of 0. You may have meant bisquare, but that change should be discussed.}
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 28, figure 7) This is a helpful graph, but perhaps making the barplot “position dodge” and the density plots overlapping with transparency would sell your point better.
+\textcolor{violet}{(page 28, figure 7) This is a helpful graph, but perhaps making the barplot “position dodge” and the density plots overlapping with transparency would sell your point better.
 This graphic makes the point, but the suggestions I made may produce more compelling evidence.}
-\end{description}
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 29, line 21) The sentence “The highest grade completed has some confusion” should be rewritten.}
-\end{description}
+\textcolor{violet}{(page 29, line 21) The sentence “The highest grade completed has some confusion” should be rewritten.}
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 29, line 37) Naming the function would be helpful for readers interested in using it.}
-\end{description}
+\textcolor{violet}{(page 29, line 37) Naming the function would be helpful for readers interested in using it.}
 
@@
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 31, line 9) Certainly you don’t mean to say that you’ve created data ”unsatisfactorily far from” the original.}
-\end{description}
+\textcolor{violet}{(page 31, line 9) Certainly you don’t mean to say that you’ve created data ”unsatisfactorily far from” the original.}
 
-\begin{description}
-\item[Associate Editor:] \textcolor{violet}{(page 32, line 18-30) Here you describe some modeling (see first note).
+\textcolor{violet}{(page 32, line 18-30) Here you describe some modeling (see first note).
 Re-framing this as a possible activity could be helpful to the reader.
 As it is written now it sounds like a summary (the section title) of what was presented – instead you can make it a path forward for the readers.}
-\end{description}

### 2022-06-07 response/response3.Rmd
addressed the AE comment on the abstract - not sure what they are reading but they are wrong

@@
 
+Nothing to respond to here.
+
 \textcolor{violet}{The main item left to be addressed is polish. This twice-revised manuscript still contains several places where authors are missing words, awkward phrasing, or grammatical errors.}
+
+Corrected as follows
 
@@
 (Alternatively it can be deleted).}
+
+The abstract does NOT promise an exploratory analysis for longitudinal data.
+It mentions that one of the purposes of this textbook data is to use it to teach exploratory analysis of longitudinal data.
+Similarly, it also states that the data is useful for teaching modeling.
+Both of these are beyond the scope of this paper.
+They are still important to keep because it makes the reader aware of why this data is useful and important.
+No change to abstract was made.
 

### 2022-06-07 paper/paper.Rmd
figure 7

@@
 ```{r compare-subsets, fig.cap = "Comparison of original and refreshed data: (A) highest grade completed, (B) experience and (C) log wages.
-Some difference in wages would be expected because the refreshed data is not inflation-adjusted, but the two sets are reasonably similar.", fig.width = 9, fig.height = 5, out.width = "100%", cache = FALSE}
+Some difference in wages would be expected because the refreshed data is not inflation-adjusted, but the two sets are reasonably similar.", fig.width = 9, fig.height = 4, out.width = "100%", cache = FALSE}
 ```

### 2022-06-07 data-raw/eda.R

@@ -66,8 +66,9 @@ exp_d94 <- do_1994 %>%
 exp_join <- bind_rows(exp_sw, exp_d94)
 
 exp_p <- ggplot(exp_join) +
-  geom_density(aes(x=exp, colour=subset, fill=subset), alpha=0.9) +
-  facet_wrap(~subset, ncol=1) +
+  geom_density(aes(x=exp, colour=subset, fill=subset),
+               alpha=0.7) +
+  #facet_wrap(~subset, ncol=1) +
   xlim(c(0,13)) +
   xlab("Experience") + ylab("") +
   scale_fill_brewer("", palette="Dark2") +
@@ -103,8 +104,9 @@ wg_d94 <- do_1994 %>%
 wg_join <- bind_rows(wg_sw, wg_d94)
 
 wg_p <- ggplot(wg_join) +
-  geom_density(aes(x=ln_wages, colour=subset, fill=subset), alpha=0.9) +
-  facet_wrap(~subset, ncol=1) +
+  geom_density(aes(x=ln_wages, colour=subset, fill=subset),
+               alpha=0.7) +
+  #facet_wrap(~subset, ncol=1) +
   xlim(c(0,4)) +
   xlab("Wages (natural log)") + ylab("") +
   scale_fill_brewer("", palette="Dark2") +

### 2022-06-07 paper/paper-jsdse/paper-jsdse.tex
hgc confusion
unsatisfactory to disappointing

@@
 
-The highest grade completed has some confusion.
+The reporting of highest grade completed is confusing.
 There are several ways this is reported, including the highest grade ever completed and also the \texttt{hgc} at each survey year.
 To match the original data, it is appropriate to use the \texttt{hgc} while in high school.
-The documentation suggests this variable is available, but it is not actually present in the data.
+The documentation suggests this variable is available, but it is actually not.
 Hence, to match the original data, we have calculated the \texttt{hgc} to match the \texttt{hgc} achieved in the years between 1979 and 1994 based on the yearly survey value.
@@
 
-Ultimately, the refreshed data is reasonably similar to the original but unsatisfactorily far from it.
+Ultimately, the refreshed data is reasonably similar to the original but disappointingly far from it.
 The last step required would be to inflation-adjust wages.

### 2022-06-07 paper/paper.Rmd
hgc confusion
unsatisfactory to disappointing

@@
 
-The highest grade completed has some confusion.
+The reporting of highest grade completed is confusing.
 There are several ways this is reported, including the highest grade ever completed and also the `hgc` at each survey year.
 To match the original data, it is appropriate to use the `hgc` while in high school.
-The documentation suggests this variable is available, but it is not actually present in the data.
+The documentation suggests this variable is available, but it is actually not.
 Hence, to match the original data, we have calculated the `hgc` to match the `hgc` achieved in the years between 1979 and 1994 based on the yearly survey value.

### 2022-06-07 response/response3.Rmd
figure 7
hgc confusion
unsatisfactory to disappointing
last AE comment

@@
 
+We have done as you suggest for the density plots, but feel the back-to-back bar chart is better for the hgc comparison.
+
 \textcolor{violet}{(page 29, line 21) The sentence “The highest grade completed has some confusion” should be rewritten.}
+
+Done
 
@@
 
+We have changed "unsatisfactorily" to "disappointingly"
+
 \textcolor{violet}{(page 32, line 18-30) Here you describe some modeling (see first note).
@@
 As it is written now it sounds like a summary (the section title) of what was presented – instead you can make it a path forward for the readers.}
+
+This bullet list is a path forward.
+It is clearly not a summary of what was presented.
+We did not discuss teaching modeling.
+We simply used one type of model for pre-rpcessing the data to handle unusual observations.
+No change made.

### 2022-06-07 paper/paper-jsdse/paper-jsdse.tex
AE's comments addressed.

@@
 This is done using the \texttt{tidyverse} suite of packages (Wickham et al.
-2019): \texttt{tidyr} (Wickham 2020) to pivot the data into long-form, with \texttt{dplyr} (Wickham et al.
-2020) and \texttt{stringr} (Wickham 2019) to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
-The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
+2019).
+We use \texttt{pivot\_longer} in \texttt{tidyr} (Wickham 2020) to pivot the data into long-form.
+For example, we use it to pull the hourly wage (\texttt{HRPjob\_year}), which is stored in a column corresponding to each job and each year of survey, into columns: job and rate per hour.
+After that, \texttt{separate} from \texttt{tidyr} is also used to get the job and survey year in two separate columns.
+Further \texttt{dplyr} (Wickham et al.
+2020) is used and \texttt{stringr} (Wickham 2019) are used to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
+The long form of the data makes it possible to do data transformations efficiently, and it is an intermediate step towards the final target data.
 The code for tidying the data are demonstrated at \url{https://numbats.github.io/yowie/articles/raw-to-input-data.html} but also described in the subsequent subsections.
@@
 However, it is important to note that the use of ``sex'' is probably not correct.
-This information is provided by the individual, and although only has two categories, is more consistent with ``gender'' as defined in Heidari et al.
+This information is provided by the individual, and although it only has two categories, it is more consistent with ``gender'' as defined in Heidari et al.
 (2016).
@@
 The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as \texttt{grade}).
-This value can change over time, but it should only increase.
+This value can change over time, but it is non-decreasing.
 This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals.
@@
 
-Our target variables for the employment are to obtain respondent's mean hourly wage (\texttt{wage}), the number of jobs (\texttt{njobs}), the total hours of work per week (\texttt{hours}) for each survey year, the year when individual starting to work (\texttt{stwork}), the length of time (years) in workforce (\texttt{yr\_wforce}), and work experience measured as the number of years worked (\texttt{exp}).
+Our target for the employment-related variables is to obtain the respondent's mean hourly wage (\texttt{wage}), the number of jobs (\texttt{njobs}), the total hours of work per week (\texttt{hours}) for each survey year, the year when individual starting to work (\texttt{stwork}), the length of time (in years) in the workforce (\texttt{yr\_wforce}), and work experience measured as the number of years worked (\texttt{exp}).
 As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.
@@
 
-For \texttt{stwork} variable, we only rename the column names from \texttt{new\_data\_qnames}.
-This variable is then used to calculate the next variable, \texttt{yr\_wforce} for each survey round, which is the year of survey (\texttt{year}) minus the year of individual started working (\texttt{stwork}).
-Finally, \texttt{exp} variable is derived from the number of weeks worked since the last interview indicated as variable started with \texttt{WKSWK} in \texttt{new\_data\_qnames}.
+For the year of individual started working (\texttt{stwork}), we only rename the column names, which is \texttt{EMPLOYERS\_ALL\_STARTDATE\_ORIGINAL.01\textasciitilde{}Y\_XRND}, from \texttt{new\_data\_qnames}.
+This variable is then used to calculate the next variable, the length of time in the workforce (\texttt{yr\_wforce}) for each survey round, which is the year of survey (\texttt{year}) minus \texttt{stwork}.
+Finally, experience (\texttt{exp}) variable is derived from the number of weeks worked since the last interview indicated as variable started with \texttt{WKSWK} in \texttt{new\_data\_qnames}.
 To obtain the work experience since 1979, we calculate the cumulative value.
@@
 
-The employment and demographic variables are then joined.
+The employment and demographic variables are then merged into one dataset as in the original data.
 These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in Singer and Willett (2003).
@@
 
-Experience is years after entering the labor force.
-It represents the difference between the day an individual enters the labor force (\texttt{EMPLOYERS\_ALL\_STARTDATE\_ORIGINAL.01\textasciitilde{}Y\_XRND} in the database) relative to the date of the survey, which we call \texttt{yr\_wforce}.
+Experience is defined as years after entering the labor force.
+It represents the difference between the day an individual enters the labor force (\texttt{stwork}) relative to the date of the survey, which we call \texttt{yr\_wforce}.
 However, using this calculation produced numbers that do not quite match the original data.
@@
 An additional criterion from Singer and Willett (2003) is to only include males aged between 14 and 17 years old in 1979.
-With this filtering, we obtained 0 individuals in the refreshed data compared to 888 individuals in the original data.
+With this filtering, we obtained 670 individuals in the refreshed data compared to 888 individuals in the original data.
 To investigate the reason for the difference, individuals from the original and refreshed dataset were matched by id.
@@
 \item
-  715 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:
+  79 individuals were less than or equal to 17 years old in 1979. However, they were not captured in the refreshed data because:
 
@@
   \item
-    The information about \texttt{ged} is missing in 341 individuals. We decided to include them in the refreshed data.
+    The information about \texttt{ged} is missing in 38 individuals. We decided to include them in the refreshed data.
   \item
@@
 
-The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing 865 individuals.
+The filtering was re-applied using these decisions, resulting in a refreshed dropout subset containing 863 individuals.
 
@@
 This is not done for the refreshed data because we plan to keep refreshing it as the data is added to and released from the survey.
-Instead, we have provided a function in the R package, \texttt{yowie}, for users to conduct the inflation adjustment when they are ready to analyze the data.
+Instead, we have provided a function called \texttt{adj\_inflation} in the R package, \texttt{yowie}, for users to conduct the inflation adjustment when they are ready to analyze the data.
 

### 2022-06-07 paper/paper.Rmd
AE's comments addressed.

@@
 Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
-This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long-form, with `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
-The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
+This is done using the `tidyverse` suite of packages [@tidyverse].
+We use `pivot_longer` in `tidyr` [@tidyr] to pivot the data into long-form.
+For example, we use it to pull the hourly wage (`HRPjob_year`), which is stored in a column corresponding to each job and each year of survey, into columns: job and rate per hour.
+After that, `separate` from `tidyr` is also used to get the job and survey year in two separate columns.
+Further `dplyr` [@dplyr] is used and `stringr` [@stringr] are used to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
+The long form of the data makes it possible to do data transformations efficiently, and it is an intermediate step towards the final target data.
 The code for tidying the data are demonstrated at \url{`r yowie_pkgdown`/articles/raw-to-input-data.html} but also described in the subsequent subsections.
@@
 However, it is important to note that the use of "sex" is probably not correct.
-This information is provided by the individual, and although only has two categories, is more consistent with "gender" as defined in @SAGER.
+This information is provided by the individual, and although it only has two categories, it is more consistent with "gender" as defined in @SAGER.
 When using the dataset in the classroom, the educators might include discourse on the use of terms "sex" and "gender".
@@
 The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as `grade`).
-This value can change over time, but it should only increase.
+This value can change over time, but it is non-decreasing.
 This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals.
@@
 
-Our target variables for the employment are to obtain respondent's mean hourly wage (`wage`), the number of jobs (`njobs`), the total hours of work per week (`hours`) for each survey year, the year when individual starting to work (`stwork`), the length of time (years) in workforce (`yr_wforce`), and work experience measured as the number of years worked (`exp`).
+Our target for the employment-related variables is to obtain the respondent's mean hourly wage (`wage`), the number of jobs (`njobs`), the total hours of work per week (`hours`) for each survey year, the year when individual starting to work (`stwork`), the length of time (in years) in the workforce (`yr_wforce`), and work experience measured as the number of years worked (`exp`).
 As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.
@@
 
-For `stwork` variable,  we only rename the column names from `new_data_qnames`.
-This variable is then used to calculate the next variable, `yr_wforce` for each survey round, which is the year of survey (`year`) minus the year of individual started working (`stwork`).
-Finally, `exp` variable is derived from the number of weeks worked since the last interview indicated as variable started with `WKSWK` in `new_data_qnames`.
+For the year of individual started working (`stwork`),  we only rename the column names, which is `EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND`, from `new_data_qnames`.
+This variable is then used to calculate the next variable, the length of time in the workforce (`yr_wforce`) for each survey round, which is the year of survey (`year`) minus `stwork`.
+Finally, experience (`exp`) variable is derived from the number of weeks worked since the last interview indicated as variable started with `WKSWK` in `new_data_qnames`.
 To obtain the work experience since 1979, we calculate the cumulative value.
@@
 
-The employment and demographic variables are then joined.
+The employment and demographic variables are then merged into one dataset as in the original data.
 These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in @SingerJudithD2003Alda.
@@
 
-Experience is years after entering the labor force.
-It represents the difference between the day an individual enters the labor force (`EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND` in the database) relative to the date of the survey, which we call `yr_wforce`.
+Experience is defined as years after entering the labor force.
+It represents the difference between the day an individual enters the labor force (`stwork`) relative to the date of the survey, which we call `yr_wforce`.
 However, using this calculation produced numbers that do not quite match the original data.
@@
 This is not done for the refreshed data because we plan to keep refreshing it as the data is added to and released from the survey.
-Instead, we have provided a function in the R package, `r yowie`, for users to conduct the inflation adjustment when they are ready to analyze the data.
+Instead, we have provided a function called `adj_inflation` in the R package, `r yowie`, for users to conduct the inflation adjustment when they are ready to analyze the data.
 

### 2022-06-07 response/response3.Rmd
rlm comment on Huber, ignore it
AE's comments addressed.

--- 2022-06-07 response/response3.Rmd
@@
 
-Thank you!
+* Thank you!
 
@@
 
-Nothing to respond to here.
+* Nothing to respond to here.
 
@@
 
-Corrected as follows
+* Corrected as follows.
 
@@
 
-The abstract does NOT promise an exploratory analysis for longitudinal data.
+* The abstract does NOT promise an exploratory analysis for longitudinal data.
 It mentions that one of the purposes of this textbook data is to use it to teach exploratory analysis of longitudinal data.
@@
 
-* Dewi to do
+* It has been removed from the sentence.
 
@@
 
-* Dewi to do
+* Done.
 
@@
 
-* Dewi to do
+* This sentence has been modified.
 
@@
 
-* Dewi to do
+* We have modified this sentence and change the word to be non-decreasing.
 
@@
 
-* Yes it is correct, the formatting of all Q... variables uses the same format as downloaded data from the database.
+* Yes it is correct, the formatting of all Q... variables uses the same format as the downloaded data from the database.
 
@@
 
-* Dewi to do
+* We have modified this sentence.
 
@@
 
-* It is assumed that people will not work more than 12 hours a day. Dewi to do.
+* It is assumed that people usually do not work more than 12 hours a day. Moreover, this value is more than the cut-off of upper value if we used the IQR method, which is only 67 hours.
 
@@
 
-* Dewi to do
+* We do not mention the `join()` function in the manuscript.
+It is true that we mentioned "the employment and demographic variables are then joined".
+However, in the current version, we have changed "join" to be "merged" so that it could reach people who are not familiar with `dplyr`.
 
 \textcolor{violet}{(page 22, line 5) The default for rlm is huber weighting which never applies a weight of 0. You may have meant bisquare, but that change should be discussed.}
+
+* We are really not sure what you are requesting here. The section was re-written in the revision. We don't refer to huber weighting, and it makes no sense to refer to bisquare.
 
@@
 
+* We have done as you suggest for the density plots, but feel the back-to-back bar chart is better for the `hgc` comparison.
+
 \textcolor{violet}{(page 29, line 21) The sentence “The highest grade completed has some confusion” should be rewritten.}
+
+* Done.
 
@@
 
-* Dewi to do
+* Done.
 
 \textcolor{violet}{(page 31, line 9) Certainly you don’t mean to say that you’ve created data ”unsatisfactorily far from” the original.}
+
+* We have changed "unsatisfactorily" to "disappointingly".
 
@@
 As it is written now it sounds like a summary (the section title) of what was presented – instead you can make it a path forward for the readers.}
+
+* This bullet list is a path forward.
+It is clearly not a summary of what was presented.
+We did not discuss teaching modeling.
+We simply used one type of model for pre-processing the data to handle unusual observations.
+No change made.

### 2022-06-07 paper/paper-jsdse/paper-jsdse.tex
clarified the 84 hours per week

@@
 \item
-  If the total hours of worked per week for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.
+  If the total hours of worked per week for the corresponding job is greater than 84 hours (12 hours/day every day), we set the wage and hours worked as missing.
 \end{itemize}

### 2022-06-07 paper/paper.Rmd
clarified the 84 hours per week

@@
 * If the hourly rate is recorded as 0, we set wage as missing, and
-* If the total hours of worked per week for the corresponding job is greater than 84 hours, we set the wage and hours worked as missing.
+* If the total hours of worked per week for the corresponding job is greater than 84 hours (12 hours/day every day), we set the wage and hours worked as missing.
 

### 2022-06-07 response/response3.Rmd
clarified the 84 hours per week

@@
 
-* Corrected as follows.
+* Corrected as detailed below.
 
@@
 Both of these are beyond the scope of this paper.
-They are still important to keep because it makes the reader aware of why this data is useful and important.
+The mention is still important to keep because it makes the reader aware of why this data are useful and important for education.
 No change to abstract was made.
@@
 
-* It is assumed that people usually do not work more than 12 hours a day. Moreover, this value is more than the cut-off of upper value if we used the IQR method, which is only 67 hours.
+* This corresponds to 12 hours per day, each day of the week, which is a reasonable upper limit on the expectation of the maximum amount of hours worked per week.
+It is quite a high upper bound, and higher than the IQR of 67.
+We have added clarification in the text.
 

### 2022-06-07 paper/paper-jsdse/paper-jsdse.tex
small fixes suggested by mervyn silverpulle
incorprated Unwins suggestion about reference to iris data

@@
 Subsets of NLSY79, including the wages data, can be found in supplementary files from numerous textbooks and research articles.
-The NLSY79 database has been continuously updated through to 2018, so new records are available.
+The NLSY79 database has been continually updated through to 2018, so new records are available.
 Here we describe our journey to refresh the wages data, and document the process so that the data can be regularly updated into the future.
@@
 The iris data (Anderson 1935) is a prime example.
-It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis\footnote{It should be noted that the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)}.
+Although the origin of the commonly available data today is uncertain and the original purpose was different\textasciitilde{}(Unwin and Kleinman 2021), it's use as a textbook data has grown and persisted.
+The three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis\footnote{It should be noted that the iris data is best replaced today with the penguins data (Horst, Hill, and Gorman 2020), which has similar qualities, is new, and does not suffer from a connection with eugenics (Stodel 2020)}.
 We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
@@
 \hypertarget{summary}{%
-\section{Summary}\label{summary}}
+\section{Summary and Discussion}\label{summary}}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-iris-unwin}{}}%
+Unwin, Antony, and Kim Kleinman. 2021. {``The Iris Data Set: In Search of the Source of Virginica.''} \emph{Significance} 18 (6): 26--29. https://doi.org/\url{https://doi.org/10.1111/1740-9713.01589}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-validate}{}}%

### 2022-06-07 paper/paper.Rmd
small fixes suggested by mervyn silverpulle
incorprated Unwins suggestion about reference to iris data

@@
 The iris data [@iris-data] is a prime example.
-It has withstood the test of time because the three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis^[It should be noted that the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]].
+Although the origin of the commonly available data today is uncertain and the original purpose was different~[@iris-unwin], it's use as a textbook data has grown and persisted.
+The three species cannot be perfectly classified, and so it continues to challenge researchers and instructors to do better in the analysis^[It should be noted that the iris data is best replaced today with the penguins data [@penguins-data], which has similar qualities, is new, and does not suffer from a connection with eugenics [@notiris]].
 We argue that the wages data is in this class of textbook data, too, because it presents a challenge for longitudinal data analysis: how can we better summarize and explain the individual experience?
@@
 
-# Summary {#summary}
+# Summary and Discussion {#summary}
 

### 2022-06-08 paper/paper.Rmd
various minor changes for final revision

--- 2022-06-07 paper/paper.Rmd
@@
 
-Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
+Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples on how to apply different techniques.
 An example of this is the wages data made public by @SingerJudithD2003Alda in their book, "Applied longitudinal data analysis," which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models.
-The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
+The data records hourly wages of a sample of high school dropouts from 1979-1994, along with demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) [@nlsy79].
 
@@
 It would be more useful for an individual to know what percentage of the population have profiles like theirs.
-We argue that different summaries are needed to communicate with the public, and ones where they might be able to individually act on to change their own situation.
+We argue that different summaries are needed to communicate with the public, so they might be able to individually act on to change their own situation.
 <!--this information is the appropriate message for an individual is that you as an individual are far more than your demographic.
 The majority of people in the study do not have a pattern that is similar to the overall trend.
-If you have experienced declining wages, you are not alone; there are others like you, and more than you think.--> We occasionally hear this voiced in the public media, also, for example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage].
+If you have experienced declining wages, you are not alone; there are others like you, and more than you think.--> We occasionally hear this voiced in the public media, for example, an article published in the Sydney Morning Herald argues there is no average Australian [@notaverage].
 Thus the wages data provides educators with a challenge, from a methodological and a philosophical perspective, and wage experience is a topic of interest to many.
@@
 Clean data often then goes through an "initial data analysis" (IDA) [@Chatfield1985TIEo], where one would summarize and scrutinize the data, especially to check if the data is consistent with assumptions required for modeling.
-This stage is related to exploratory data analysis (EDA), coined by @tukey with a focus on learning from data.
+This stage is related to exploratory data analysis (EDA), coined by @tukey, with a focus on learning from data.
 EDA can be considered to encompass IDA.
@@
 Our approach to cleaning builds heavily on the `tidyverse` approach [@tidyverse].
-The data is first organized into "tidy data" [@WickhamHadley2014TD] and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables [@plyr].
+The data is first organized into "tidy data" [@WickhamHadley2014TD], and then further wrangled using step-wise piping with a split-apply-combine strategy for mutating new variables [@plyr].
 Tidy data shouldn't be confused with "tame data," which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
-The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
+The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
 
@@
 1. ID: the respondents' ID. 
-2. EXPER: stands for experience, the temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
+2. EXPER: temporal scale, i.e., the length of time (years) in the workforce, starting on the respondents' first day at work.
 3. LNW: natural logarithm of wages, adjusted with 1990's inflation rate. 
@@
 What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
-In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidy the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so that the resulting input data can be used downstream for validating the data as described in Section \@ref(ida) and Section \@ref(censor).
+In this section, we outline the steps to download the raw data (Section \@ref(getdata)) and then tidy the raw data into input data, specifically for the demographic variables (Section \@ref(tidydemog)) and the employment variables (Section \@ref(tidyemp)), so the resulting input data can be used downstream for validating the data as described in Section \@ref(ida) and Section \@ref(censor).
 
@@
 Thus, we re-arrange and wrangle the data into tidy data form, with columns corresponding to individual ID, year, job number, wage in dollars, and the demographic variables.
-This is done using the `tidyverse` suite of packages [@tidyverse].
-We use `pivot_longer` in `tidyr` [@tidyr] to pivot the data into long-form.
-For example, we use it to pull the hourly wage (`HRPjob_year`), which is stored in a column corresponding to each job and each year of survey, into columns: job and rate per hour.
-After that, `separate` from `tidyr` is also used to get the job and survey year in two separate columns.
-Further `dplyr` [@dplyr] is used and `stringr` [@stringr] are used to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
-The long form of the data makes it possible to do data transformations efficiently, and it is an intermediate step towards the final target data.
+This is done using the `tidyverse` suite of packages [@tidyverse]: `tidyr` [@tidyr] to pivot the data into long-form, with `dplyr` [@dplyr] and `stringr` [@stringr] to create new variables from the downloaded data from the database, and code levels of factors by text wrangling.
+The long form of the data makes it possible to do these data transformations efficiently, and it is an intermediate step towards the final target data.
 The code for tidying the data are demonstrated at \url{`r yowie_pkgdown`/articles/raw-to-input-data.html} but also described in the subsequent subsections.
@@
 
-Our final target data will include the demographic variables (variable names): sex (`sex`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years (as an integers), e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).
+Our final target data will include the demographic variables (variable names): sex (`sex`), race (`race`), age (`age_1979`), highest grade completed reported in each round of the survey (`grade`), highest grade completed ever reported (`hgc`), highest grade completed in terms of years, e.g., 9th grade = 9, 3rd-year college = 15, (`hgc_i`), highest grade completed in 1979 (`hgc_1979`) and whether the graduate equivalency diploma is obtained (`ged`).
 
@@
 However, it is important to note that the use of "sex" is probably not correct.
-This information is provided by the individual, and although it only has two categories, it is more consistent with "gender" as defined in @SAGER.
+This information is provided by the individual, and although only has two categories, is more consistent with "gender" as defined in @SAGER.
 When using the dataset in the classroom, the educators might include discourse on the use of terms "sex" and "gender".
@@
 This variable is obtained from `new_data_qnames` with the name `HGC_EVER_XRND` and stored in year units (e.g., 10, 11, 12, 13, and so on), so the transformation is simply giving a new column name as `hgc_i` and recoded as a factor to give `hgc` (e.g., 10th grade, 11th grade, and so on).
-We decide to include both the factor and integer variables to be used as examples in various classroom demonstrations to explain different data types say.
+We decide to include both the factor and integer variables to be used as examples in various classroom demonstrations to explain different data types.
 Besides this, it can also serve as a demonstration of different type of data visualization ascribed by the data type and for teaching the modeling of cross sectional data.
@@
 The second definition is the highest grade completed that is reported in each round of the survey (provided in the refreshed data as `grade`).
-This value can change over time, but it is non-decreasing.
-This has been included in the refreshed dataset to enable richer downstream analyses, for example, if one would like to explore how the temporal changes in education level affect the wages of individuals.
+This value can change over time, but it should only increase.
+This has been included in the refreshed dataset to enable richer downstream analyses.
+For example, if one would like to explore how the temporal changes in education level affect the wages of individuals.
 Hence, different from `hgc`, this variable could be used to demonstrate longitudinal and time series analysis in the classroom.
@@
 In addition, it is also in the columns beginning with `HGCREV` reflecting revised data.
-We chose to use these revised values because there were fewer missing values indicating it had been more thoroughly checked.
+We chose to use these revised values because there were fewer missing values, indicating it had been more thoroughly checked.
 In the survey years when the revised value were not reported, i.e., in 2012, 2014, 2016, and 2018, we used the unrevised version accordingly.
@@
 
-Our target for the employment-related variables is to obtain the respondent's mean hourly wage (`wage`), the number of jobs (`njobs`), the total hours of work per week (`hours`) for each survey year, the year when individual starting to work (`stwork`), the length of time (in years) in the workforce (`yr_wforce`), and work experience measured as the number of years worked (`exp`).
+Our target variables to extract for employment are the respondent's mean hourly wage (`wage`), the number of jobs (`njobs`), the total hours of work per week (`hours`) for each survey year, the year when individual starting to work (`stwork`), the length of time (years) in workforce (`yr_wforce`), and work experience measured as the number of years worked (`exp`).
 As the data only reports up to 5 jobs for each respondent, the maximum number of jobs is capped at 5.
@@
 
-For the year of individual started working (`stwork`),  we only rename the column names, which is `EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND`, from `new_data_qnames`.
-This variable is then used to calculate the next variable, the length of time in the workforce (`yr_wforce`) for each survey round, which is the year of survey (`year`) minus `stwork`.
-Finally, experience (`exp`) variable is derived from the number of weeks worked since the last interview indicated as variable started with `WKSWK` in `new_data_qnames`.
+For the `stwork` variable,  we only rename the column names from `new_data_qnames`.
+This variable is then used to calculate the next variable, `yr_wforce` for each survey round, which is the year of survey (`year`) minus the year of individual started working (`stwork`).
+Finally, the `exp` variable is derived from the number of weeks worked since the last interview indicated as variable started with `WKSWK` in `new_data_qnames`.
 To obtain the work experience since 1979, we calculate the cumulative value.
-As the measurement unit is weeks, we convert this to fractional year.
+As the measurement unit is weeks, we convert this to year.
 
-The employment and demographic variables are then merged into one dataset as in the original data.
+The employment and demographic variables are then joined.
 These data are further filtered to the cohort who participated in at least three rounds in the survey, the minimum observation recommended for longitudinal data in @SingerJudithD2003Alda.
@@
 Finally, we save the resultant wage data on this cohort as `wages`.
-Note that we save `grade` variable in this dataset instead of `demog_nlsy79` dataset because it is a longitudinal variable, while `demog_nlsy79` is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.
+Note that we save the `grade` variable in this dataset instead of the `demog_nlsy79` dataset because it is a longitudinal variable, while `demog_nlsy79` is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.
 
@@
 
-Experience is defined as years after entering the labor force.
-It represents the difference between the day an individual enters the labor force (`stwork`) relative to the date of the survey, which we call `yr_wforce`.
+Experience is years after entering the labor force.
+It represents the difference between the day an individual enters the labor force (`EMPLOYERS_ALL_STARTDATE_ORIGINAL.01~Y_XRND` in the database) relative to the date of the survey, which we call `yr_wforce`.
 However, using this calculation produced numbers that do not quite match the original data.
@@
 In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides.
-The first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
+The first step is validating that the numerical summaries of the raw data are the same as reported by NLSY79.
 This is followed by graphical summaries using methods available in `ggplot2` [@ggplot2] and `brolgar` [@brolgar].
@@
 Figure \ref{fig:feature-plot} shows an alternative way to check the data quality.
-Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to $60,000/hour) for some individuals, mostly around 1990.
+Plot (A) is the spaghetti plot where all profiles are shown, and it can be seen that there are unbelievably high wage values (up to \$60,000/hour) for some individuals, mostly around 1990.
 Plot (B) shows side-by-side boxplots of the three number summaries (minimum, median, and maximum) for all individuals.
 This tells us there are a number of individuals with unbelievably high maximum wages.
-Plot (C) shows the profile for an individual with a maximum wage that is not so extreme but still indicates a problem: their wages are consistently low except for one year where they earned close to $1200/hour.
+Plot (C) shows the profile for an individual with a maximum wage that is not so extreme but still indicates a problem: their wages are consistently low except for one year where they earned close to \$1200/hour.
 This does not seem to be reasonable and leads us to use a procedure to detect and fix these temporal anomalies.
@@
 The challenging part of detecting an anomaly using the robustness weight is determining the weight threshold beyond which the observations are considered outliers.
-It is also important not to be too aggressive in outlier removal, because this would overly smooth the data, and ignore actual workforce experiences important wage volatility such as, abnormally high wages for a fleeting amount of time.
+It is also important not to be too aggressive in outlier removal, because this would overly smooth the data, and ignore actual workforce experiences important to wage volatility, such as abnormally high wages for a fleeting amount of time.
 To explore the risk of being overly vigorous in labeling observations as outliers, a range of threshold value were assessed.
@@
 Figure \ref{fig:fixed-feature-plot} shows the summary statistics after removing extremes.
-The highest wage overall is now around $1000.
-Plot (A) shows a more reasonable spaghetti plot, where there are some profiles with high wages, but most profiles have wages under $300, and there has been a steady increase in wages over the years.
+The highest wage overall is now around \$1000.
+Plot (A) shows a more reasonable spaghetti plot, where there are some profiles with high wages, but most profiles have wages under \$300, and there has been a steady increase in wages over the years.
 Plot (B) shows there are still a small number of individuals with high maximum wages.
@@
   5. `sex`: Sex of the individual, f = Female and m = Male.
-  6. `race`: Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
+  6. `race`: Race the individual belongs to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
   7. `grade`: Integer value of the highest grade completed corresponding to the survey year. 
@@
 `r nrow(join_more_17)` individuals were more than 17 years old in 1979.
-Thus, it looks like the description of the original data is not quite accurate that there are people older than 17 in the subset.
+Thus, it looks like the description of the original data, that there are people older than 17 in the subset, is not quite accurate.
 Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
@@
 ```{r compare-subsets, fig.cap = "Comparison of original and refreshed data: (A) highest grade completed, (B) experience and (C) log wages.
-Some difference in wages would be expected because the refreshed data is not inflation-adjusted, but the two sets are reasonably similar.", fig.width = 9, fig.height = 4, out.width = "100%", cache = FALSE}
+Some difference in wages would be expected because the refreshed data is not inflation-adjusted, but the two sets are reasonably similar.", fig.width = 9, fig.height = 5, out.width = "100%", cache = FALSE}
 ```
@@
 
-The reporting of highest grade completed is confusing.
+The variables from "highest grade completed" have some confusion.
 There are several ways this is reported, including the highest grade ever completed and also the `hgc` at each survey year.
 To match the original data, it is appropriate to use the `hgc` while in high school.
-The documentation suggests this variable is available, but it is actually not.
+The documentation suggests this variable is available, but it is not actually present in the data.
 Hence, to match the original data, we have calculated the `hgc` to match the `hgc` achieved in the years between 1979 and 1994 based on the yearly survey value.
@@
 This is not done for the refreshed data because we plan to keep refreshing it as the data is added to and released from the survey.
-Instead, we have provided a function called `adj_inflation` in the R package, `r yowie`, for users to conduct the inflation adjustment when they are ready to analyze the data.
+Instead, we have provided a function in the R package, `r yowie`, for users to conduct the inflation adjustment when they are ready to analyze the data.
 
@@
 
-Ultimately, the refreshed data is reasonably similar to the original but disappointingly far from it.
+Ultimately, the refreshed data is reasonably similar to the original but unsatisfactorily far from it.
 The last step required would be to inflation-adjust wages.
@@
 Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
-- Some of the methods shown, and lessons learned about the data, can also be considered generally for data cleaning, initial data analysis and exploratory data analysis, useful for working with other data sets.
+- Some of the methods shown, and lessons learned about the data, can also be considered when working with other datasets, and more generally for data cleaning, IDA, and EDA.
 - The refreshed data can be used for teaching advanced linear modeling, using almost identical code from @SingerJudithD2003Alda, and thus make those lessons more current for today's students.

### 2022-06-08 paper/paper-jsdse/paper-jsdse.tex
incorporated nicks suggestions

@@
 
-Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples about how to apply different techniques.
+Statistics and data science education relies on cleaned and simplified data, suitably called textbook data, for clear examples on how to apply different techniques.
 An example of this is the wages data made public by Singer and Willett (2003) in their book, ``Applied longitudinal data analysis,'' which can be used to teach generalized linear models, and additionally hierarchical, mixed effects, and multilevel models.
-The data records hourly wages of a sample of high school dropouts from 1979-1994, along with the demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
+The data records hourly wages of a sample of high school dropouts from 1979-1994, along with demographic variables, such as education and race, taken from the National Longitudinal Survey of Youth (NLSY79) (Bureau of Labor Statistics, U.S.
 Department of Labor 2021a).
@@
 Tidy data shouldn't be confused with ``tame data,'' which Kim, Ismay, and Chunn (2018) coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
-The resulting (tame) data is provided in a new R package called \texttt{yowie}, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
+The resulting (tame) data is provided in a new R package called \texttt{yowie}, which includes the code so the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
 
@@
 What we have colorfully named wild data can be considered to be raw data, and valid data could be considered to be textbook data in the above statistical value chain.
-In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so that the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.
+In this section, we outline the steps to download the raw data (Section \ref{getdata}) and then tidy the raw data into input data, specifically for the demographic variables (Section \ref{tidydemog}) and the employment variables (Section \ref{tidyemp}), so the resulting input data can be used downstream for validating the data as described in Section \ref{ida} and Section \ref{censor}.
 
@@
 Finally, we save the resultant wage data on this cohort as \texttt{wages}.
-Note that we save \texttt{grade} variable in this dataset instead of \texttt{demog\_nlsy79} dataset because it is a longitudinal variable, while \texttt{demog\_nlsy79} is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.
+Note that we save the \texttt{grade} variable in this dataset instead of the \texttt{demog\_nlsy79} dataset because it is a longitudinal variable, while \texttt{demog\_nlsy79} is a cross-sectional dataset reflecting the state of individuals corresponding to the most recent round of the survey.
 
@@
 In this paper, we conduct an IDA or a preliminary data analysis to assess the validity of the variable values in the cohort of data that the NLSY provides.
-The first step is validating numerical summaries of the raw data are the same as reported by NLSY79.
+The first step is validating that the numerical summaries of the raw data are the same as reported by NLSY79.
 This is followed by graphical summaries using methods available in \texttt{ggplot2} (Wickham 2016) and \texttt{brolgar} (Tierney, Cook, and Prvan 2020).
@@
 \item
-  \texttt{race} : Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
+  \texttt{race} : Race the individual belongs to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
 \item
@@
   173 individuals were more than 17 years old in 1979.
-Thus, it looks like the description of the original data is not quite accurate that there are people older than 17 in the subset.
+Thus, it looks like the description of the original data, that there are people older than 17 in the subset, is not quite accurate.
 Our decision is to also include them in the refreshed data as the new data contains an age variable, so analysts could filter them later.
@@
 
-The reporting of highest grade completed is confusing.
+The variables from ``highest grade completed'' have some confusion.
 There are several ways this is reported, including the highest grade ever completed and also the \texttt{hgc} at each survey year.
@@
 \item
-  Some of the methods shown, and lessons learned about the data, can also be considered generally for data cleaning, initial data analysis and exploratory data analysis, useful for working with other data sets.
+  Some of the methods shown, and lessons learned about the data, can also be considered to be useful for working with other data sets, and also apply generally for data cleaning, initial data analysis and exploratory data analysis.
 \item

### 2022-06-08 paper/paper.Rmd
incorporated nicks suggestions

--- 2022-06-07 paper/paper.Rmd
@@
 Tidy data shouldn't be confused with "tame data," which @tamedata coined to refer to textbook data sets suitable for teaching, particularly teaching statistics.
-The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so that the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
+The resulting (tame) data is provided in a new R package called `r yowie`, which includes the code so the process is reproducible and could be used to further refresh the data as new records are made available in the NLSY79 database.
 
@@
   3. `sex`: Sex of the individual, f = Female and m = Male.
-  4. `race` : Race of the individual belong to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
+  4. `race` : Race the individual belongs to, NBH = Non-Black, Non-Hispanic; H = Hispanic; B = Black.
   5. `hgc`: The highest grade completed ever.
@@
 
-The reporting of highest grade completed is confusing.
+The variables from "highest grade completed" have some confusion.
 There are several ways this is reported, including the highest grade ever completed and also the `hgc` at each survey year.
@@
 Choices made during the cleaning and processing can affect findings made with the data, are transparent, and can be a basis for classroom discussions.
-- Some of the methods shown, and lessons learned about the data, can also be considered generally for data cleaning, initial data analysis and exploratory data analysis, useful for working with other data sets.
+- Some of the methods shown, and lessons learned about the data, can also be considered to be useful for working with other data sets, and also apply generally for data cleaning, initial data analysis and exploratory data analysis.
 - The refreshed data can be used for teaching advanced linear modeling, using almost identical code from @SingerJudithD2003Alda, and thus make those lessons more current for today's students.
